### 2016-06-05 trunk/redi/paper/paper.tex
Some old stuff I never committed. Fixed a bug in CopyFrom, too

\documentclass[twocolumn]{article}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{textcomp}
% \usepackage{inlinebib}

\pagestyle{empty}

\newcommand\comment[1]{}

\newcommand\st{$^{\mathrm{st}}$}
\newcommand\nd{$^{\mathrm{nd}}$}
\newcommand\rd{$^{\mathrm{rd}}$}
\renewcommand\th{$^{\mathrm{th}}$}
\newcommand\tm{$^{\mbox{\tiny \textsc{tm}}}$}

% nice fractions
\newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}

\newcommand\citef[1]{\addtocounter{footnote}{1}\footnotetext{\cite{#1}}\ensuremath{^{\mbox{\footnotesize [\thefootnote]}}}}

% XXX TODO: before submit, search ``red eye'' and ``reduction'' vs removal
% we -> I

\begin{document}

\title{Red i removal with artificial retinal networks}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2015 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2015 with the increasingly askance visage of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
0.00 Australian Neo-Dollars}
}

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{1 April 2015}

\maketitle

\begin{abstract}
We present a GPU-accelerated means for red i removal in photographs.
\end{abstract}

\vspace{1em}
{\noindent \small {\bf Keywords}:
  computational photography, image processing, generalized photoshop, artificial retinal networks, types
}

\section{Introduction}

When light---such as the bright flashbulb of a camera---strikes the
human eye, it illuminates the retina. Some of that light bounces back
out of the eye, but most of it stimulates neurons in the retina to
produce electrical signals. These signals stimulate other neurons to
which they are connected, and so on, until the brain (which is
technically part of the eye) perceives an image, as a two-dimensional
array of neurons with different activation levels. Humans often use
these images to sense the world, for example, in reading research papers.

This research paper concerns a particular feature of this process,
which is that humans are able to view an image and ignore certain
details of it. For example, Figure~\ref{fig:cbs} contains a printout
of an image file of a photograph of a television displaying a recorded
video of an actor. The video contains a superimposed eye in the
corner, the logo of the network CBS. Most viewers are not tormented by
this everpresent eye staring at them! In fact, most viewers are able
to completely ignore the eye, and view the scene as though it didn't
contain the stimulus, even if details such as the actor's sweatshirt's
collar pass beneath the stimulus and are occluded by it.

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.85 \linewidth]{cbs}
\end{center}
\caption{Q.~Who watches the TV~watchers?
A.~CBS's all-seeing eye.} \label{fig:cbs}
\end{figure}

Some stimulus is more everpresent than others. The Clay Mathematics
Institute lists among its unsolved {\em Millennium Prize} problems the
``red i removal problem.'' This concerns the removal of stimulus (a
red letter ``i'') from images (Figure~\ref{fig:redi}). The problem
is particularly difficult because the information occluded by the
i is completely gone, and because the authors of papers about the
problem are persistently agitated because it seems like the letter
should be capitalized.

In this paper I show how red i removal can be solved in certain
specialized cases, using an artificial retinal network patterned after
the brain contained within the human eye. Training this artificial
retinal network is feasible on a single powerful desktop machine. Both
training and execution of the model (a mere 400 megabytes) are GPU
accelerated. The model presented in this paper was trained in about 3
days, and executing it in parallel on a suite of images takes about
100 milliseconds per image.\!\footnote{Source code is available on
the World Wide Web at:
{\tt http://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/redi}}

\begin{figure}[tb]
\begin{center}
\includegraphics[width=0.65 \linewidth]{rotor}
\end{center}
\caption{An image of an Enigma machine rotor with a red i
 superimposed. Solving this instance of the red i removal problem
 would mean producing an image without the red i. One way to do this
 would be to steal a floppy disk containing an original, unimposed
 image of the rotor, from someone in possession of it, for example
 the paper's author.} \label{fig:redi}
\end{figure}

\section{Artificial retinal networks}

As I expertly foreshadowed in the previous section, an artificial
retinal network works just like the brain inside a human eye. The
retina is itself a rectangular 2D array of neurons, which turn photons
into IEEE-754 floating point values between 0.0f and 1.0f. Behind this
is another 2D array of exactly the same size, except the pixels are in
a weird jumbled order, and then another layer. This series of layers
is known as the optic nerve. Finally, the brain perceives the image as
an array of pixels, again of the same size (Figure~\ref{fig:retinal}).

It is easy to reconceptualize this process as an array of pixels
undergoing several transformations. Obviously the story is more
complicated: Humans see in color, so each pixel is actually three
different nodes, one for red, green and blue. In fact, since some
scientists hypothesize of certain ``superseers''---that is, people who
can perceive more than just the three wavelengths of light---we
actually allow an arbitrary number of nodes per pixel. In this work,
we used $N=4$.

In a real human eye, each node is fed inputs from every node in the
previous layer. For computational efficiency, in this work we allow
only 64 inputs to each node from the previous layer. Because we
suspect that layers are spatially related, a node is always connected
to its {\em neighborhood} in the previous layer (each node from the 9
pixels within Manhattan distance 1). The rest of the inputs are
selected randomly from a Gaussian distribution, as long as the samples
fall within the image (using rejection sampling---the sides and
corners do not ``wrap around''). By the way, the images are always
256x256, because numbers that are a power of two are
faster.\!\footnote{This is true on computers, because computers count
in binary. In the human eye, powers of ten are faster, because humans
have ten fingers.} The connection from one node to another is
modulated by a {\em weight}, again an IEEE-754 floating point number.
A node outputs the sum of its input values, passed through a
smoothulator, specifically the one found in Gray's~Anatomy (the book,
not the TV show),

$$ \frac{1}{1 + e^{-v}} $$

\begin{figure}[tb]
\includegraphics[width=0.95 \linewidth]{retinal}
\caption{ How eyes work, and thus, artificial retinal networks. } \label{fig:retinal}
\end{figure}

This function is {\em biologically plausible}.

We learn by backpropagated stochastic gradient descent, like babies
do. Specifically: The network is presented with an image on its
retina, and then we run the floating point values through the layers,
summing them up and applying the smoothulator function, to produce a
final image within the brain. Like a baby, it compares this image to
what it expected to see, node by node. Where each node does not agree
with the image, the error is computed. The baby propagates the partial
derivative of the change in error with respect to the change in
stimulus to the previous layer proportional to its weighted impact;
fortunately the smoothulator has a simple derivative that is easily
computed at a point from its output values. Error is not propagated
into the real world (i.e., by sending light off of the retina back to
the physical object that created the stimulus; that would be
ridiculous).

\subsection{Training data}

One of the insights of this paper is that although artificial retinal
networks require a lot of data to train, for certain problems the
training data can be easily {\em generated}. For the red i removal
problem, we begin with a corpus of about 4,000 images that were
scraped from Google Image Search. Scraping images is easy; one just
needs to list a bunch of queries for things that babies would want to
view in order to learn what the world looks like. In this experiment I
used terms such as [snakes], [dog on skateboard], [guitar],
[stonehenge] and [superyacht]. Following this, I manually cleaned the
training data. I deleted images that were not photographic (drawings,
etc.; for example, most images of guitars are actually 3D rendered
cartoon guitars, fantasy images of guitars on fire, the Guitar~Hero logo,
and so on) or that were {\em too}, uh, pornographic (most queries for
things that can have sex, like tapirs, contain prominent images of the
things having sex). These are not appropriate for babies.

All images are cropped to a square and resized to 256x256. Then we
generate training instances: An input image and the expected result.
An image that does not contain a red i should just be transformed into
the image itself (indeed, when we peer directly into the brain of a
baby looking at a TV show, we find a region of the brain where the TV
show is clearly visible). It is also easy to generate instances of the
red i removal problem along with their solutions---we simply put a red
i randomly on the source image and keep the destination image
unchanged. In this way, we can easily generate a large amount of
training instances (in actual practice, this procedure had a small bug;
see Section~\ref{sec:evaluation}). One unexpected phenomenon is that
I had to be careful to remove images that already contained a red i,
like many images of casinos, which are often called ``CASINO''.

% notes on corpus:
% - my own picture of library in there

In order to coax networks into recognizing the i, we also place an i
into the 4th color channel in the same position in the {\em expected
output}. This is an invisible color channel which we discard, and
which is always zero in the input. In essence we giving a hint to the
eye's brain that it should not just remove the red i, but it should
also {\em perceive} it. I have not performed enough experiments to
know if this is helpful.

For repeatability's sake, important constants used in this experiment:
There were 2 hidden layers. Gaussan samples were produced with a
standard deviation of 16 pixels. The red i was rendered in Comic Sans,
at a height of 80 pixels. I used a variety of learning rates,
including an expontentially decreasing one (the standard advice of
0.05 is too large for constant learning rates on this kind of task, and
limits the sharpness of resultant images).

\subsection{CUDA, SHUDA, WUDA}

Because we are working with graphical data, we should use the Graphics
Processing Unit of the computer, not its Central Processing Unit (we
are not processing centers). I implemented high-performance OpenCL
kernels for each phase of training: The {\em forward} pass (signals
flowing from the retina to the eye's brain), the {\em backpropagation}
step (when the eye computes the error and partial derivatives) and the
{\em weight update} step (when the eye rewires its neurons so that it
sees the right thing next time). The phases have different parallelism
constraints. Because the connectivity is sparse, we represent both
forward and inverted index maps, which are decoded on the GPU. We take
care to only load a single layer of the retinal network into the GPU's
RAM at once, to enable very large models, but we run many training
instances in parallel for a single round. Some other operations, like
the preparation of training data, are performed on the CPU. These are
also frequently done in parallel, using C++11's new {\tt std::thread}
with some crazy-ass wrappers to allow them to function in mingw32's
64-bit gcc port. On a good day, training uses all 6 CPU cores and all
2800 GPU cores and about 14 GB of RAM and warms the home office like a
1kW space heater.

\section{Results}

After 4 rounds of training, the network produces an excellent-looking
image that could be a {\em Cure} album cover, regardless of the stimulus
cast upon its retina (Figure~\ref{fig:cure}).

\begin{figure}[bth]
\begin{center}
\includegraphics[width=0.65 \linewidth]{cure}
\end{center}
\caption{Result after 4 rounds of training. Looks great, and there is no
red i to be seen, but it loses some points for not resembling the
input image at all.} \label{fig:cure}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{frosted}
\end{center}
\caption{Result after 80 rounds of training, with the input image at the top
and the signal proceeding downward through two hidden layers. The hidden layers
aspire to crazy noise-terror glitch art versions of the stimulus as well.} \label{fig:frosted}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{instagram}
\end{center}
\caption{After about 9,000 rounds. Top row is the input image, the
second row is the output (no longer showing hidden layers because
they all just look like firefly raves); the bottom row shows 4x
magnified detail of the region formerly containing the red i.
Images are somewhat desaturated and blurry, but the red i is removed.
Note how in the right image, the retinal network successfully continued
both the horizontal and vertical bookshelves into the occluded region.
This is not a trick.} \label{fig:instagram}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{eval}
\end{center}
\caption{Evaluation on new images after 30,000 rounds of training. Top row
is the input image, the second row is the output, and third is 4x magnified
defail. The first image (no i) shows the high amount of detail preserved.
In the latter two, the i is successfully removed; the quality of the replacement
is not perfect, but certainly reasonable.} \label{fig:eval}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.95 \linewidth]{jpeg}
\end{center}
\caption{Evaluation of an early model for J peg dequantization. The model
still contains a lot of noise pixels, which sometimes take a long time to
converge, but it is already easy to see how quantization artifacts have been reduced (left).
Actually, there is no reason why such dequantization must only be applied to J pegs;
the right column shows it working on a nice rainbow picture.
} \label{fig:jpeg}
\end{figure}

It's not long before the network learns that it should not produce the
same result for every input, and the output starts to mimic the
input. These images look sort of like the world viewed through frosted
glass (Figure~\ref{fig:frosted}), simulating how a baby first learns
to see the world through the 1cm bulletproof Lexan of its translucent
BabyLearn incubation cylinder.

Soon thereafter, the network begins to converge on something like the
identity function, as this drastically reduces error (even if some
error is incurred by the persistence of the red i). Left overnight
(about 9,000 rounds), we start to see the network both produce images
much like the original (perhaps through a hip ``vintage'' Instagram
filter), as well as removing the red i stimulus (Figure~\ref{fig:instagram}).

\subsection{Evaluation} \label{sec:evaluation}

With a further 30,000 rounds of training, the output images sharpen
and lose their Instagram quality (maybe only a small amount of
``grain''), and the i is still successfully removed. However, since
we've now made many passes over each image in the training set (and
the model has about 100,000 degrees of freedom), it is certainly
possible that we've simply overfit to this set of images (that is,
that the baby's eye's brain is simply memorizing the i-free images and
then recalling the one that looks closest to the stimulus). To evaluate
fairly, we need to apply the model to totally new images that it was
not trained on. These are called ``eval'' images.

Firing this up, I observed that the model successfully reproduced eval
images that did not contain a red i; this is good because it means
that it is not simply memorizing the training set images. I then
started placing red i stimulus on the images with the mouse, and my
heart sank: It wasn't removing the red i at all! Dejected, I tried
loading up the training images and putting a red i on them---it also
did not remove the i, which did not make sense! Even for babies!
Eventually, I discovered that the red i would be removed, as expected,
but only when the i was in a handful of very specific locations. This
was found to be a bug in the random i placing code; can you find it too?

\smallskip
\noindent \verb+        uint8 x_dice = seed & 0x255;+ \\
          \verb+        seed >>= 8;+ \\
          \verb+        uint8 y_dice = seed & 0x255;+
\smallskip

As a result, there are only 16 different possible $x$ coordinates, and same
for $y$. Nonetheless, this is still 256 different i locations that work,
which implies considerable generality is possible. Due to Draconian SIGBOVIK
deadlines, I have not yet been able to test a debugged training procedure.

Once the evaluation code only places an i at expected locations, the artificial
retinal network works well (Figure~\ref{fig:eval})!

\section{Conclusions}

We find that the supposedly impossible red i removal problem is in
fact solvable, at least in some forms, using artificial retinal
networks. There are some limitations of the current model:
\begin{itemize}
\item It has only been tested to remove the letter i when it is rendered
  in bright red, in 30 point Comic Sans.
\item It probably also removes letters like j, but maybe also in some
  other fonts, which is a sword that cuts both ways.
\item Due to a bug, the red i must be at a position whose coordinates are exactly
  $\langle 12 + x_0 + 4x_1 + 16x_2 + 64x_3, 12 + y_0 + 4y_1 + 16y_2 + 64y_3 \rangle$,
for $x_j$ and $y_j$ in $\{0, 1\}$.
\item It automatically and non-optionally applies Instagram-style filters.
\end{itemize}

This technique can probably be applied to other image processing
problems, for example, J peg dequantization. Here, we take an image
and badly quantize it (for example, to 4 bits per color channel), and
the training instance consists of the quantized image as input and the
original image as the expected output; the retinal network learns how
to fill in detail. Figure~\ref{fig:jpeg} shows the early stages (about
4000 rounds) of training such a model.

A related, still unsolved problem is ``red i reduction''; here we do
simply remove the i but replace it with a smaller i. For example, we
could replace a capital I with a lowercase one, or replace a lowercase
30pt Comic Sans i with a lowercase 29pt Comic Sans i. This is an
offshoot of the text ure compression field, which seeks to make the
text ``ure'' smaller wherever it appears.

\medskip
Biologically-inspired computer algorithms hold many wonders for those
that seek to tap into the limitless potential of the 85\% of the human
eye's brain that is currently unused. Perhaps humans even contain
graphics processing units!

\bigskip
For higher-fidelity images and source code, please consult {\tt
http://tom7.org/redi}.

\end{document}

### 2016-06-05 trunk/redi/clutil.h

@@ -45,6 +45,12 @@ struct CL {
     queue = clCreateCommandQueue(context, devices[0], 0, nullptr);
   }
 
+  cl_command_queue NewCommandQueue(bool out_of_order = true) {
+    return clCreateCommandQueue(context, devices[0], 
+				out_of_order ? CL_QUEUE_OUT_OF_ORDER_EXEC_MODE_ENABLE : 0,
+				nullptr);
+  }
+
   pair<cl_program, cl_kernel> BuildOneKernel(const string &kernel_src,
 					     const string &function_name) {
     Timer gpu_compile;

### 2016-06-05 trunk/redi/jpeg.cc

@@ -747,6 +747,33 @@ static void UpdateWeights(const float learning_rate,
 	       }, 12);
 }
 
+struct RoundOnGPU {
+  explicit RoundOnGPU(CL *cl) : cl(cl) {}
+
+  void SetupForward(Network *net) {
+    indices.resize(net->num_layers);
+    weights.resize(net->num_layers);
+    biases.resize(net->num_layers);
+
+    for (int layer = 0; layer < net->num_layers; layer++) {
+      indices[layer] = MoveMemoryToGPUConst(cl->context, cl->queue, net->indices[layer]);
+      weights[layer] = MoveMemoryToGPU(cl->context, cl->queue, false, &net->weights[layer]);
+      biases[layer] = MoveMemoryToGPU(cl->context, cl->queue, false, &net->biases[layer]);
+    }
+  }
+
+  ~RoundOnGPU() {
+    for (int i = 0; i < indices.size(); i++) {
+      CHECK_SUCCESS(clReleaseMemObject(indices[i]));
+      CHECK_SUCCESS(clReleaseMemObject(weights[i]));
+      CHECK_SUCCESS(clReleaseMemObject(biases[i]));
+    }
+  }
+
+  CL *cl;
+  vector<cl_mem> indices, weights, biases;
+};
+
 struct ForwardLayerCL {
   explicit ForwardLayerCL(CL *cl) : cl(cl) {
     const string kernel_src = 
@@ -757,12 +784,8 @@ struct ForwardLayerCL {
   }
 
   struct ForwardContext {
-    ForwardContext(ForwardLayerCL *parent, const Network &net, int layer) :
-      parent(parent), net(net), layer(layer) {
-      CL *cl = parent->cl;
-      indices = MoveMemoryToGPUConst(cl->context, cl->queue, net.indices[layer]);
-      weights = MoveMemoryToGPUConst(cl->context, cl->queue, net.weights[layer]);
-      biases = MoveMemoryToGPUConst(cl->context, cl->queue, net.biases[layer]);
+    ForwardContext(ForwardLayerCL *parent, RoundOnGPU *round, int layer) :
+      parent(parent), round(round), layer(layer) {
     }
 
     // TODO: Do we really want to share the same command queue across threads?
@@ -786,9 +809,12 @@ struct ForwardLayerCL {
 	MutexLock ml(&parent->m);
[…]

### 2016-11-26 abc/paper/paper.tex
Writes broken EXE. Stub for paper, more notes, INSTALL notes

--- 2016-06-05 trunk/redi/paper/paper.tex
@@
 
+\usepackage{supertabular}
+\usepackage{amsthm}
+\usepackage{relsize}
 \usepackage{amsmath}
 \usepackage{amssymb}
+% \usepackage{code}
 \usepackage{graphicx}
@@
 \usepackage{textcomp}
-% \usepackage{inlinebib}
 
@@
 
-% XXX TODO: before submit, search ``red eye'' and ``reduction'' vs removal
-% we -> I
+\usepackage{ulem}
+% go back to italics for emphasis, though
+\normalem
 
@@
 
-\title{Red i removal with artificial retinal networks}
+\title{This is the Title}
 \author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
-Copyright \copyright\ 2015 the Regents of the Wikiplia
-Foundation. Appears in SIGBOVIK 2015 with the increasingly askance visage of the
+    Copyright \copyright\ 2017 the Regents of the Wikiplia Foundation.
+    Appears in SIGBOVIK 2017 with the utmost concern of the
     Association for Computational Heresy; {\em IEEEEEE!} press,
-Verlag-Verlag volume no.~0x40-2A.
-0.00 Australian Neo-Dollars}
-}
+    Verlag--Verlag volume no.~0x2017. BMD 0.00} }
 
@@
 
-\date{1 April 2015}
+\date{1 April 2017}
 
@@
 
-\begin{abstract}
-We present a GPU-accelerated means for red i removal in photographs.
-\end{abstract}
+\section*{Abstract}
+This is the abstract
 
@@
 {\noindent \small {\bf Keywords}:
-  computational photography, image processing, generalized photoshop, artificial retinal networks, types
+  keywords
 }
@@
 
-When light---such as the bright flashbulb of a camera---strikes the
-human eye, it illuminates the retina. Some of that light bounces back
-out of the eye, but most of it stimulates neurons in the retina to
-produce electrical signals. These signals stimulate other neurons to
-which they are connected, and so on, until the brain (which is
-technically part of the eye) perceives an image, as a two-dimensional
-array of neurons with different activation levels. Humans often use
-these images to sense the world, for example, in reading research papers.
+% ``characters'' ``words'' and ``paragraphs'' and ``pages''
 
-This research paper concerns a particular feature of this process,
-which is that humans are able to view an image and ignore certain
-details of it. For example, Figure~\ref{fig:cbs} contains a printout
-of an image file of a photograph of a television displaying a recorded
-video of an actor. The video contains a superimposed eye in the
-corner, the logo of the network CBS. Most viewers are not tormented by
-this everpresent eye staring at them! In fact, most viewers are able
-to completely ignore the eye, and view the scene as though it didn't
-contain the stimulus, even if details such as the actor's sweatshirt's
-collar pass beneath the stimulus and are occluded by it.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=0.85 \linewidth]{cbs}
-\end{center}
-\caption{Q.~Who watches the TV~watchers?
-A.~CBS's all-seeing eye.} \label{fig:cbs}
-\end{figure}
-
-Some stimulus is more everpresent than others. The Clay Mathematics
-Institute lists among its unsolved {\em Millennium Prize} problems the
-``red i removal problem.'' This concerns the removal of stimulus (a
-red letter ``i'') from images (Figure~\ref{fig:redi}). The problem
-is particularly difficult because the information occluded by the
-i is completely gone, and because the authors of papers about the
-problem are persistently agitated because it seems like the letter
-should be capitalized.
-
-In this paper I show how red i removal can be solved in certain
-specialized cases, using an artificial retinal network patterned after
-the brain contained within the human eye. Training this artificial
-retinal network is feasible on a single powerful desktop machine. Both
-training and execution of the model (a mere 400 megabytes) are GPU
-accelerated. The model presented in this paper was trained in about 3
-days, and executing it in parallel on a suite of images takes about
-100 milliseconds per image.\!\footnote{Source code is available on
-the World Wide Web at:
-{\tt http://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/redi}}
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics[width=0.65 \linewidth]{rotor}
-\end{center}
-\caption{An image of an Enigma machine rotor with a red i
- superimposed. Solving this instance of the red i removal problem
- would mean producing an image without the red i. One way to do this
- would be to steal a floppy disk containing an original, unimposed
- image of the rotor, from someone in possession of it, for example
- the paper's author.} \label{fig:redi}
-\end{figure}
-
-\section{Artificial retinal networks}
-
-As I expertly foreshadowed in the previous section, an artificial
-retinal network works just like the brain inside a human eye. The
-retina is itself a rectangular 2D array of neurons, which turn photons
-into IEEE-754 floating point values between 0.0f and 1.0f. Behind this
-is another 2D array of exactly the same size, except the pixels are in
-a weird jumbled order, and then another layer. This series of layers
-is known as the optic nerve. Finally, the brain perceives the image as
-an array of pixels, again of the same size (Figure~\ref{fig:retinal}).
-
-It is easy to reconceptualize this process as an array of pixels
-undergoing several transformations. Obviously the story is more
-complicated: Humans see in color, so each pixel is actually three
-different nodes, one for red, green and blue. In fact, since some
-scientists hypothesize of certain ``superseers''---that is, people who
-can perceive more than just the three wavelengths of light---we
-actually allow an arbitrary number of nodes per pixel. In this work,
-we used $N=4$.
-
-In a real human eye, each node is fed inputs from every node in the
-previous layer. For computational efficiency, in this work we allow
-only 64 inputs to each node from the previous layer. Because we
-suspect that layers are spatially related, a node is always connected
-to its {\em neighborhood} in the previous layer (each node from the 9
-pixels within Manhattan distance 1). The rest of the inputs are
-selected randomly from a Gaussian distribution, as long as the samples
-fall within the image (using rejection sampling---the sides and
-corners do not ``wrap around''). By the way, the images are always
-256x256, because numbers that are a power of two are
-faster.\!\footnote{This is true on computers, because computers count
-in binary. In the human eye, powers of ten are faster, because humans
-have ten fingers.} The connection from one node to another is
-modulated by a {\em weight}, again an IEEE-754 floating point number.
-A node outputs the sum of its input values, passed through a
-smoothulator, specifically the one found in Gray's~Anatomy (the book,
-not the TV show),
-
-$$ \frac{1}{1 + e^{-v}} $$
-
-\begin{figure}[tb]
-\includegraphics[width=0.95 \linewidth]{retinal}
-\caption{ How eyes work, and thus, artificial retinal networks. } \label{fig:retinal}
-\end{figure}
-
-This function is {\em biologically plausible}.
-
-We learn by backpropagated stochastic gradient descent, like babies
-do. Specifically: The network is presented with an image on its
-retina, and then we run the floating point values through the layers,
-summing them up and applying the smoothulator function, to produce a
-final image within the brain. Like a baby, it compares this image to
-what it expected to see, node by node. Where each node does not agree
-with the image, the error is computed. The baby propagates the partial
-derivative of the change in error with respect to the change in
-stimulus to the previous layer proportional to its weighted impact;
-fortunately the smoothulator has a simple derivative that is easily
-computed at a point from its output values. Error is not propagated
-into the real world (i.e., by sending light off of the retina back to
-the physical object that created the stimulus; that would be
-ridiculous).
-
-\subsection{Training data}
-
-One of the insights of this paper is that although artificial retinal
-networks require a lot of data to train, for certain problems the
-training data can be easily {\em generated}. For the red i removal
-problem, we begin with a corpus of about 4,000 images that were
-scraped from Google Image Search. Scraping images is easy; one just
-needs to list a bunch of queries for things that babies would want to
-view in order to learn what the world looks like. In this experiment I
-used terms such as [snakes], [dog on skateboard], [guitar],
-[stonehenge] and [superyacht]. Following this, I manually cleaned the
-training data. I deleted images that were not photographic (drawings,
-etc.; for example, most images of guitars are actually 3D rendered
-cartoon guitars, fantasy images of guitars on fire, the Guitar~Hero logo,
-and so on) or that were {\em too}, uh, pornographic (most queries for
-things that can have sex, like tapirs, contain prominent images of the
-things having sex). These are not appropriate for babies.
-
-All images are cropped to a square and resized to 256x256. Then we
-generate training instances: An input image and the expected result.
-An image that does not contain a red i should just be transformed into
-the image itself (indeed, when we peer directly into the brain of a
-baby looking at a TV show, we find a region of the brain where the TV
-show is clearly visible). It is also easy to generate instances of the
-red i removal problem along with their solutions---we simply put a red
-i randomly on the source image and keep the destination image
-unchanged. In this way, we can easily generate a large amount of
-training instances (in actual practice, this procedure had a small bug;
-see Section~\ref{sec:evaluation}). One unexpected phenomenon is that
-I had to be careful to remove images that already contained a red i,
-like many images of casinos, which are often called ``CASINO''.
-
-% notes on corpus:
-% - my own picture of library in there
-
-In order to coax networks into recognizing the i, we also place an i
-into the 4th color channel in the same position in the {\em expected
-output}. This is an invisible color channel which we discard, and
-which is always zero in the input. In essence we giving a hint to the
-eye's brain that it should not just remove the red i, but it should
-also {\em perceive} it. I have not performed enough experiments to
-know if this is helpful.
-
-For repeatability's sake, important constants used in this experiment:
-There were 2 hidden layers. Gaussan samples were produced with a
-standard deviation of 16 pixels. The red i was rendered in Comic Sans,
-at a height of 80 pixels. I used a variety of learning rates,
-including an expontentially decreasing one (the standard advice of
-0.05 is too large for constant learning rates on this kind of task, and
-limits the sharpness of resultant images).
-
-\subsection{CUDA, SHUDA, WUDA}
-
-Because we are working with graphical data, we should use the Graphics
-Processing Unit of the computer, not its Central Processing Unit (we
-are not processing centers). I implemented high-performance OpenCL
-kernels for each phase of training: The {\em forward} pass (signals
-flowing from the retina to the eye's brain), the {\em backpropagation}
-step (when the eye computes the error and partial derivatives) and the
-{\em weight update} step (when the eye rewires its neurons so that it
-sees the right thing next time). The phases have different parallelism
-constraints. Because the connectivity is sparse, we represent both
-forward and inverted index maps, which are decoded on the GPU. We take
-care to only load a single layer of the retinal network into the GPU's
-RAM at once, to enable very large models, but we run many training
-instances in parallel for a single round. Some other operations, like
-the preparation of training data, are performed on the CPU. These are
-also frequently done in parallel, using C++11's new {\tt std::thread}
-with some crazy-ass wrappers to allow them to function in mingw32's
-64-bit gcc port. On a good day, training uses all 6 CPU cores and all
-2800 GPU cores and about 14 GB of RAM and warms the home office like a
-1kW space heater.
-
-\section{Results}
-
-After 4 rounds of training, the network produces an excellent-looking
-image that could be a {\em Cure} album cover, regardless of the stimulus
-cast upon its retina (Figure~\ref{fig:cure}).
-
-\begin{figure}[bth]
-\begin{center}
-\includegraphics[width=0.65 \linewidth]{cure}
-\end{center}
-\caption{Result after 4 rounds of training. Looks great, and there is no
-red i to be seen, but it loses some points for not resembling the
-input image at all.} \label{fig:cure}
-\end{figure}
-
-\begin{figure}[htb]
-\begin{center}
-\includegraphics[width=0.95 \linewidth]{frosted}
-\end{center}
-\caption{Result after 80 rounds of training, with the input image at the top
-and the signal proceeding downward through two hidden layers. The hidden layers
-aspire to crazy noise-terror glitch art versions of the stimulus as well.} \label{fig:frosted}
-\end{figure}
-
-\begin{figure}[htb]
-\begin{center}
-\includegraphics[width=0.95 \linewidth]{instagram}
-\end{center}
-\caption{After about 9,000 rounds. Top row is the input image, the
-second row is the output (no longer showing hidden layers because
-they all just look like firefly raves); the bottom row shows 4x
-magnified detail of the region formerly containing the red i.
-Images are somewhat desaturated and blurry, but the red i is removed.
-Note how in the right image, the retinal network successfully continued
-both the horizontal and vertical bookshelves into the occluded region.
-This is not a trick.} \label{fig:instagram}
-\end{figure}
-
-\begin{figure}[htb]
-\begin{center}
-\includegraphics[width=0.95 \linewidth]{eval}
-\end{center}
-\caption{Evaluation on new images after 30,000 rounds of training. Top row
-is the input image, the second row is the output, and third is 4x magnified
-defail. The first image (no i) shows the high amount of detail preserved.
-In the latter two, the i is successfully removed; the quality of the replacement
-is not perfect, but certainly reasonable.} \label{fig:eval}
-\end{figure}
-
-\begin{figure}[htb]
-\begin{center}
-\includegraphics[width=0.95 \linewidth]{jpeg}
-\end{center}
-\caption{Evaluation of an early model for J peg dequantization. The model
-still contains a lot of noise pixels, which sometimes take a long time to
-converge, but it is already easy to see how quantization artifacts have been reduced (left).
-Actually, there is no reason why such dequantization must only be applied to J pegs;
-the right column shows it working on a nice rainbow picture.
-} \label{fig:jpeg}
-\end{figure}
-
-It's not long before the network learns that it should not produce the
-same result for every input, and the output starts to mimic the
-input. These images look sort of like the world viewed through frosted
-glass (Figure~\ref{fig:frosted}), simulating how a baby first learns
-to see the world through the 1cm bulletproof Lexan of its translucent
-BabyLearn incubation cylinder.
-
-Soon thereafter, the network begins to converge on something like the
-identity function, as this drastically reduces error (even if some
-error is incurred by the persistence of the red i). Left overnight
-(about 9,000 rounds), we start to see the network both produce images
-much like the original (perhaps through a hip ``vintage'' Instagram
-filter), as well as removing the red i stimulus (Figure~\ref{fig:instagram}).
-
-\subsection{Evaluation} \label{sec:evaluation}
-
-With a further 30,000 rounds of training, the output images sharpen
-and lose their Instagram quality (maybe only a small amount of
-``grain''), and the i is still successfully removed. However, since
-we've now made many passes over each image in the training set (and
-the model has about 100,000 degrees of freedom), it is certainly
-possible that we've simply overfit to this set of images (that is,
-that the baby's eye's brain is simply memorizing the i-free images and
-then recalling the one that looks closest to the stimulus). To evaluate
-fairly, we need to apply the model to totally new images that it was
-not trained on. These are called ``eval'' images.
-
-Firing this up, I observed that the model successfully reproduced eval
-images that did not contain a red i; this is good because it means
-that it is not simply memorizing the training set images. I then
-started placing red i stimulus on the images with the mouse, and my
-heart sank: It wasn't removing the red i at all! Dejected, I tried
-loading up the training images and putting a red i on them---it also
-did not remove the i, which did not make sense! Even for babies!
-Eventually, I discovered that the red i would be removed, as expected,
-but only when the i was in a handful of very specific locations. This
-was found to be a bug in the random i placing code; can you find it too?
-
-\smallskip
-\noindent \verb+        uint8 x_dice = seed & 0x255;+ \\
-          \verb+        seed >>= 8;+ \\
-          \verb+        uint8 y_dice = seed & 0x255;+
-\smallskip
-
-As a result, there are only 16 different possible $x$ coordinates, and same
-for $y$. Nonetheless, this is still 256 different i locations that work,
-which implies considerable generality is possible. Due to Draconian SIGBOVIK
-deadlines, I have not yet been able to test a debugged training procedure.
-
-Once the evaluation code only places an i at expected locations, the artificial
-retinal network works well (Figure~\ref{fig:eval})!
-
-\section{Conclusions}
-
-We find that the supposedly impossible red i removal problem is in
-fact solvable, at least in some forms, using artificial retinal
-networks. There are some limitations of the current model:
-\begin{itemize}
-\item It has only been tested to remove the letter i when it is rendered
-  in bright red, in 30 point Comic Sans.
-\item It probably also removes letters like j, but maybe also in some
-  other fonts, which is a sword that cuts both ways.
-\item Due to a bug, the red i must be at a position whose coordinates are exactly
-  $\langle 12 + x_0 + 4x_1 + 16x_2 + 64x_3, 12 + y_0 + 4y_1 + 16y_2 + 64y_3 \rangle$,
-for $x_j$ and $y_j$ in $\{0, 1\}$.
-\item It automatically and non-optionally applies Instagram-style filters.
-\end{itemize}
-
-This technique can probably be applied to other image processing
-problems, for example, J peg dequantization. Here, we take an image
-and badly quantize it (for example, to 4 bits per color channel), and
-the training instance consists of the quantized image as input and the
-original image as the expected output; the retinal network learns how
-to fill in detail. Figure~\ref{fig:jpeg} shows the early stages (about
-4000 rounds) of training such a model.
-
-A related, still unsolved problem is ``red i reduction''; here we do
-simply remove the i but replace it with a smaller i. For example, we
-could replace a capital I with a lowercase one, or replace a lowercase
-30pt Comic Sans i with a lowercase 29pt Comic Sans i. This is an
-offshoot of the text ure compression field, which seeks to make the
-text ``ure'' smaller wherever it appears.
-
-\medskip
-Biologically-inspired computer algorithms hold many wonders for those
-that seek to tap into the limitless potential of the 85\% of the human
-eye's brain that is currently unused. Perhaps humans even contain
-graphics processing units!
-
-\bigskip
-For higher-fidelity images and source code, please consult {\tt
-http://tom7.org/redi}.
+\bibliographystyle{IEEEtran}
+% \bibliographycomment{} % nothing
+\bibliography{paper}
 

### 2017-02-20 trunk/abc/global.c
Loads of stuff while sourceforge was down. Installation/hacking notes. Cleanup in dos_execute. Allow globals to have some initializing bytes. During 'optimization', move global initialization to a new initialization stub so that assembly doesn't need to deal with it. Assembling now outputs an initial data segment. Optimize PLUS when arguments are known. New test of globals. Move genvar to cil-util; should rally all variable generation around this struct so that we don't get accidental collisions. Fix a few cases where exceptions were inaccessible because they were generated inside some functor

@@ -1,0 +1,8 @@
+int toplevel_global = 0;
+char toplevel_printable = 'c';
+char toplevel_global_simplified = (1 + 2) ? (char)'a' : (char)'b';
+
+int main (int argc, char *argv) {
+  toplevel_global ++;
+  return (int)toplevel_printable;
+}

### 2017-02-20 abc/paper/paper.tex
Loads of stuff while sourceforge was down. Installation/hacking notes. Cleanup in dos_execute. Allow globals to have some initializing bytes. During 'optimization', move global initialization to a new initialization stub so that assembly doesn't need to deal with it. Assembling now outputs an initial data segment. Optimize PLUS when arguments are known. New test of globals. Move genvar to cil-util; should rally all variable generation around this struct so that we don't get accidental collisions. Fix a few cases where exceptions were inaccessible because they were generated inside some functor

@@
 
+% texticutable
+
 % ``characters'' ``words'' and ``paragraphs'' and ``pages''
+
+% Literate executable, knuth
+% In the dead space of the binary, put this paper or part of it
+
+% ``PISC'' = Printable Instruction Set Computer, etc.
+
+It's a common experience that each time you program a solution to a
+hard problem, you do it a different way, due to a lack of complete
+satisfaction with the way you did it last time. True to form, the
+ABC compiler exercises this technique *within the same compiler*,
+using a different approach to represent each of the intermediate
+languages.
+
+TODO: Convert byte table thing into ascii. Unfortunately can't fit
+the whole 256x256 thing, so maybe 128x128 with a small amount of
+cropping, then a 1:1 zoom on the sierpinski triangle part?
 

### 2017-02-20 abc/dosbox/src/dos/dos_execute.cpp

@@ -347,7 +347,8 @@ bool DOS_Execute(char * name,PhysPt block_pt,Bit8u flags) {
 		Bit16u minsize,maxsize;Bit16u maxfree=0xffff;
 		// pspseg ultimately becomes the data segment for EXE files. -tom7
 		DOS_AllocateMemory(&pspseg,&maxfree);
-		LOG(LOG_EXEC,LOG_NORMAL)("pspseg: %lx = %lu", (long unsigned int)pspseg,
+		LOG(LOG_EXEC,LOG_NORMAL)("pspseg: %lx = %lu",
+					 (long unsigned int)pspseg,
 					 (long unsigned int)pspseg);
 		if (iscom) {
 			minsize=0x1000;maxsize=0xffff;

### 2017-03-03 abc/paper/paper.tex
Some notes: I think perhaps JumpInd should pop from the stack.
Delete aforementioned badideas. ASM program needs to record where local frames start. Add x86 encoding to acc. Pipe calls through to individual command translation. Implement Init instruction. Add eicar to paper notes
Separate segment out into its own files. Make fallthrough explicit in tox86. Get into the xblocks layout phase for real. (DOES NOT WORK)

@@
 
+Thank you for playing wing commander.
+
 TODO: Convert byte table thing into ascii. Unfortunately can't fit
@@
 cropping, then a 1:1 zoom on the sierpinski triangle part?
+
+this page confirms the wrap-around with jumps btw:
+http://www.felixcloutier.com/x86/Jcc.html
+(Probably already in the x86 manual actually?)
+
+Another possibility is to include readable debug information (e.g.
+frame layouts) in a separate section.
+
+BTW here is eicar:
+
+X5O!P%@AP[4\PZX54(P^)7CC)7}$EICAR-STANDARD-ANTIVIRUS-TEST-FILE!$H+H*
+
+It does indeed use self-modifying code to create the interrupt
+instructions, which it uses to print the string and exit.
+
+histogram of actual bytes used in code segment?
 

### 2017-03-04 trunk/abc/paper/paper.tex
Some notes, etc. from RIVERCITY
Attempted to write some authoritative description of the layout strategy and calling convention; add 32-bit literals since those will be easy
Make more of the EXE filler be printable. Break out X86 encoding and functorize it as per the perfomance TODO. This helped tremendously--9x the speed with mlton
Fix bugs and add some 16-bit addressing modes (note possible large code size improvement from using these for temp addressing, etc.). builtin and simpcall tests both work, whee

--- 2016-11-26 abc/paper/paper.tex
@@
 
+% texticutable
+
 % ``characters'' ``words'' and ``paragraphs'' and ``pages''
+
+% Literate executable, knuth
+% In the dead space of the binary, put this paper or part of it
+
+% ``PISC'' = Printable Instruction Set Computer, etc.
+
+It's a common experience that each time you program a solution to a
+hard problem, you do it a different way, due to a lack of complete
+satisfaction with the way you did it last time. True to form, the
+ABC compiler exercises this technique *within the same compiler*,
+using a different approach to represent each of the intermediate
+languages.
+
+Thank you for playing wing commander.
+
+TODO: Convert byte table thing into ascii. Unfortunately can't fit
+the whole 256x256 thing, so maybe 128x128 with a small amount of
+cropping, then a 1:1 zoom on the sierpinski triangle part?
+
+this page confirms the wrap-around with jumps btw:
+http://www.felixcloutier.com/x86/Jcc.html
+(Probably already in the x86 manual actually?)
+
+Another possibility is to include readable debug information (e.g.
+frame layouts) in a separate section.
+
+BTW here is eicar:
+
+X5O!P%@AP[4\PZX54(P^)7CC)7}$EICAR-STANDARD-ANTIVIRUS-TEST-FILE!$H+H*
+
+It does indeed use self-modifying code to create the interrupt
+instructions, which it uses to print the string and exit.
+
+histogram of actual bytes used in code segment?
+
+interesting bugs:
+  DI <- [DI]
 

### 2017-03-06 trunk/abc/builtin.c
Fix a few bad bugs: Failure to clear knowledge of registers when they change leads to corrupt calculations, and cil initializing declaration would just initialize a different variable, ugh. Made tests one-liners so more fit on the screen.

@@ -3,6 +3,6 @@ int _putc(int);
 int main(int argc, char **argv) {
   _putc('o');
   _putc('k');
-  _putc('\n');
+  _putc('!');
   return 0;
 }

### 2017-03-06 trunk/abc/callargs.c

@@ -7,6 +7,6 @@ int Other(int a, int b, int c) {
 }
 
 int main(int argc, char **argv) {
-  Other('\n', 'o', 'k');
+  Other('!', 'o', 'k');
   return 0;
 }

### 2017-03-06 abc/eqzero.c
Fix a few bad bugs: Failure to clear knowledge of registers when they change leads to corrupt calculations, and cil initializing declaration would just initialize a different variable, ugh. Made tests one-liners so more fit on the screen.

@@ -5,17 +5,22 @@ int main(int argc, char **argv) {
   if (x) {
     _putc('X');
   }
-  /*
+
   if (argc == (int)1 || argc == (int)2) {
     _putc('X');
   }
 
-  if (argc == (int)0) {
+  if (!x) {
     _putc('o');
-    _putc('k');
-    _putc('\n');
   }
-  */
+
+  if (x == 0) {
+    _putc('k');
+  }
+
+  if (argc == (int)0) {
+    _putc('!');
+  }
 
   return 0;
 }

### 2017-03-06 abc/globals.c

@@ -4,11 +4,11 @@ int global_o = 'o';
 int global_k;
 
 int main(int argc, char **argv) {
-  static int global_nl;
-  global_nl = '\n';
+  static int global_bang;
+  global_bang = '!';
   _putc(global_o);
   global_k = 'k';
   _putc(global_k);
-  _putc(global_nl);
+  _putc(global_bang);
   return 0;
 }

### 2017-03-07 abc/paper/paper.tex
Put relocations after CS, so that it's not corrupting code, and do so using a printable word. Some debugging code and simple seg:off solver

@@
 
+interesting bugs:
+  DI <- [DI]
+
+write about load16 tmp <- tmp; it's amazing now.
+(benchmark history shows performance gains..)
+
+words contained in this paper
+
+style points for:
+- put relocations a little later after CS, so that we can mark
+  with text the spot where they go.
+
+opportunity for style points, like making your phone number into
+a word:
+
+ j ? X 4 ?   (4 can alternatively be $ or ,)
+
+ NEED TO FIX:
+  - jump at least 0x20
+  - store16 (not printable)
+  - Not bug
+  - some kind of output other than \_putc
+
+dosbox patches, wing commander
+
+ascii art in paper
+
 \bibliographystyle{IEEEtran}

### 2017-03-07 abc/dosbox/src/cpu/core_normal.cpp

@@ -167,7 +167,6 @@ restart_opcode:
 				Bitu len=(GETIP-reg_eip);
 				LOADIP;
 				if (len>16) len=16;
-				LOG(LOG_CPU,LOG_NORMAL)("core_normal");
 				if (len == 1 && mem_readb(core.cseip) == 0x63) {
 				  LOG(LOG_CPU,LOG_NORMAL)("Thank you for playing Wing Commander!");
 				  core.cseip++;

### 2017-03-07 abc/dosbox/src/dos/dos_execute.cpp

@@ -259,7 +259,7 @@ bool DOS_Execute(char * name,PhysPt block_pt,Bit8u flags) {
 	EXE_Header head;Bitu i;
 	Bit16u fhandle;Bit16u len;Bit32u pos;
 	Bit16u pspseg,envseg,loadseg,memsize,readsize;
-	PhysPt loadaddress;RealPt relocpt;
+	PhysPt loadaddress;
 	Bitu headersize=0,imagesize=0;
 	DOS_ParamBlock block(block_pt);
 
@@ -467,14 +467,23 @@ bool DOS_Execute(char * name,PhysPt block_pt,Bit8u flags) {
 		LOG(LOG_EXEC,LOG_NORMAL)("Looking for relocations at %u.", pos);
 		DOS_SeekFile(fhandle,&pos,0);
 		for (i=0;i<head.relocations;i++) {
+		        RealPt relocpt;
 			readsize=4;DOS_ReadFile(fhandle,(Bit8u *)&relocpt,&readsize);
 			relocpt=host_readd((HostPt)&relocpt);		//Endianize
+			// So this is like a 32-bit address (presumably, little-endian),
+			// but we break it into seg:offset, then add the loadseg
+			// PhysMake(seg,off) is just (seg<<4) + off.
 			PhysPt address=PhysMake(RealSeg(relocpt)+loadseg,RealOff(relocpt));
-			if (i == 0 || address != 147689) { // XXX
-			  LOG(LOG_EXEC,LOG_NORMAL)("%d. Reloc (ld %u) (reloc @$%8x) addr %u (+%u)",
+			// Only prints the first relocation since there are
+			// many (all should be the same in abc code).
+			if (i == 0) {
+			  LOG(LOG_EXEC,LOG_NORMAL)("%d. Reloc (ld %u) (reloc @$%8x = %x:%x) "
+						   "addr %x (+%u)",
 						   i,
-						   loadseg, relocpt,
-						   address, relocate);
+						   loadseg,
+						   relocpt, RealSeg(relocpt), RealOff(relocpt),
+						   address,
+						   relocate);
 			}
 
 			mem_writew(address,mem_readw(address)+relocate);

### 2017-03-08 trunk/abc/paper/paper.tex
Lovely and absurd trick for load16 tmp <- tmp
Implement 'Yet' CIL operator as improvement for dos2unix eqzero.c in translation of and/or boolean ops. Note a case where I'm generating non-printable bytes. Implement EqZero and NeZero throughout. Find that there's a bug somewhere, and made a good test case with eqzero.c
Fix a few bad bugs: Failure to clear knowledge of registers when they change leads to corrupt calculations, and cil initializing declaration would just initialize a different variable, ugh. Made tests one-liners so more fit on the screen.
Implement argc/argv, though they can't rea

@@
 
+write about load16 tmp <- tmp; it's amazing now.
+(benchmark history shows performance gains..)
+
+words contained in this paper
+
+``interrupt is a way of telling the operating system: Check this out!''
+
+style points for:
+- put relocations a little later after CS, so that we can mark
+  with text the spot where they go.
+
+opportunity for style points, like making your phone number into
+a word:
+
+ j ? X 4 ?   (4 can alternatively be $ or ,)
+
+ NEED TO FIX:
+  - some kind of output other than \_putc
+
+ TOP PERFORMANCE IMPROVEMENTS:
+  - don't jump at end of every block
+  - don't assume a block needs to be split just because it's 0x7e-0x20.
+    measure from the last jump!
+
+dosbox patches, wing commander
+
+ascii art in paper
+
 \bibliographystyle{IEEEtran}

### 2017-03-08 abc/echoing.c
Implement argc/argv, though they can't really be used because load8 isn't implemented, and also I didn't test it so it might also not work

@@ -1,0 +1,14 @@
+int _putc(int);
+
+int main(int argc, char **argv) {
+  int i = 0;
+  char *cmdline = *argv;
+  for (i = 0; i < (int)(0xFF - 0x81); i++) {
+    char c = cmdline[i];
+    //    if (c == 0x0D)
+    // return 0;
+    // _putc((int)c);
+  }
+  _putc('\n');
+  return 0;
+}

### 2017-03-10 trunk/abc/text2pdf.c
OK, twocolumn is working, so now what remains are to write the insertable bits and include them in the right place.

@@ -95,7 +95,7 @@ void WriteHeader(char *title){
   writestr("1 0 obj\n");
   writestr("<<\n");
   sprintf(buf, "/CreationDate (%s)\n", datestring); writestr(buf);
-  sprintf(buf, "/Producer (%s (\\251 Phil Smith, 1996))\n", appname); writestr(buf);
+  sprintf(buf, "/Producer (ABC)\n"); writestr(buf);
   if (title) {sprintf(buf, "/Title (%s)\n", title); writestr(buf);}
   writestr(">>\n");
   writestr("endobj\n");

### 2017-03-10 abc/paper/paper.tex
Make minimum displacement 0x20. Print out instruction bytes that aren't printable; in the current tests, the only remaining violations are INT 21 (expected, for tests) and the known bug in store16.
Fix store16 to be printable.
OK, twocolumn is working, so now what remains are to write the insertable bits and include them in the right place.

@@
 
+``interrupt is a way of telling the operating system: Check this out!''
+
 style points for:
@@
  NEED TO FIX:
-  - jump at least 0x20
-  - store16 (not printable)
-  - Not bug
   - some kind of output other than \_putc
+
+ TOP PERFORMANCE IMPROVEMENTS:
+  - don't jump at end of every block
+  - don't assume a block needs to be split just because it's 0x7e-0x20.
+    measure from the last jump!
 
@@
 
+The CISC Ridiculous
+
 ascii art in paper
+
+awesome text art headings?
+
+Truly self-documenting
+Giving new meaning to "program committee" "evaluating" my paper
+For good reasons that I will explain later...
+
+RISC ROLL
 

### 2017-03-10 abc/paper.c

@@ -1,0 +1,47 @@
+/* Here's the C source code for the program that is
+   the SIGBOVIK paper.
+
+   (Highly in-progress as of 10 Mar 2017!)
+*/
+
+int _putc(int); // XXX don't use in paper!
+int _out8(int, int);
+
+int Adlib(int reg, int value) {
+  int i;
+  _out8((int)0x0388, (int)reg);
+  for (i = 0; i < (int)12; i++) {}
+  _out8((int)0x0389, (int)value);
+  for (i = 0; i < (int)84; i++) {}
+  return 0;
+}
+
+int Quiet() {
+  int port;
+
+  // Zero all registers to clear sound card.
+  for (port = (int)0x01; port <= (int)0xF5; port++) {
+    // _putc('.');
+    Adlib((int)port, (int)0x00);
+  }
+}
+
+int main(int argc, char **argv) {
+  int i;
+  Quiet();
+
+  Adlib((int)0x20, (int)0x01); // Set the modulator's multiple to 1
+  Adlib((int)0x40, (int)0x10); // Set the modulator's level to about 40 dB
+  Adlib((int)0x60, (int)0xF0); // Modulator attack: quick; decay: long
+  Adlib((int)0x80, (int)0x77); // Modulator sustain: medium; release: medium
+  Adlib((int)0xA0, (int)0x98); // Set voice frequency's LSB (it'll be a D#)
+  Adlib((int)0x23, (int)0x01); // Set the carrier's multiple to 1
+  Adlib((int)0x43, (int)0x00); // Set the carrier to max volume (about 47 dB)
+  Adlib((int)0x63, (int)0xF0); // Carrier attack: quick; decay: long
+  Adlib((int)0x83, (int)0x77); // Carrier sustain: medium; release: medium
+  Adlib((int)0xB0, (int)0x31); // Turn voice on; set the octave and freq MSB
+
+  for (i = 0; i < (int)1000; i++) {}
+  Quiet();
+  return 0;
+}

### 2017-03-11 trunk/abc/paper.c
Some more on the paper; fix typo

@@ -4,7 +4,7 @@
    (Highly in-progress as of 10 Mar 2017!)
 */
 
-int _putc(int); // XXX don't use in paper!
+int _putc(int); // XXX non-printable! don't use in paper!
 int _out8(int, int);
 
 // Adlib uses two bytes to do a "note-on", and the notes are specified

### 2017-03-12 trunk/abc/paper/paper.tex
Almost filled postreloc. Lunch time

@@
 % Literate executable, knuth
-% In the dead space of the binary, put this paper or part of it
 
@@
 
-TODO: Convert byte table thing into ascii. Unfortunately can't fit
-the whole 256x256 thing, so maybe 128x128 with a small amount of
-cropping, then a 1:1 zoom on the sierpinski triangle part?
-
-this page confirms the wrap-around with jumps btw:
-http://www.felixcloutier.com/x86/Jcc.html
-(Probably already in the x86 manual actually?)
-
 Another possibility is to include readable debug information (e.g.
 frame layouts) in a separate section.
-
-BTW here is eicar:
-
-X5O!P%@AP[4\PZX54(P^)7CC)7}$EICAR-STANDARD-ANTIVIRUS-TEST-FILE!$H+H*
-
-It does indeed use self-modifying code to create the interrupt
-instructions, which it uses to print the string and exit.
 
@@
 
-write about load16 tmp <- tmp; it's amazing now.
-(benchmark history shows performance gains..)
-
 words contained in this paper
-
-``interrupt is a way of telling the operating system: Check this out!''
 
@@
 
+The CISC Ridiculous
+
 ascii art in paper
+
+awesome text art headings?
+
+Truly self-documenting
+Giving new meaning to "program committee" "evaluating" my paper
+For good reasons that I will explain later...
+
+I was kidding when I said the x86 architecture is elegant.
+
+Dr. Dobbs Journal typing in the program
+
+ES segment must be used for IN?
+
+TODO: X86 vs x86
+
+(Idea: ladder is actually interspersed in code, by putting some long instruction
+byte (that takes several arguments) before DEC SP JNZ +7e? We can jump directly
+to the jump, but if executed in a normal synchronized flow, then it executes some
+instruction that does nothing (load immediate?).)
+
+(Idea: Once we've gone past EIP=0xFFFF, we can still execute code from our binary.
+It may be possible to get back to CS by doing a jump with the address/operand size prefix?
+Neither one has an effect in DOSBox, which may be a bug? Anyway, if an interrupt
+is triggered while we're up here, it seems like it only does a 16-bit RET, so that's
+kinda busted.)
+
+missing condition; ``left-facing''
+
+RISC ROLL
+
+video: bingo of tom 7 video themes (``retro'', ``alphabet,''
+``fractal geometry'')
 

### 2017-03-15 trunk/abc/sizeof.c
Implement sizeof; more paper writing

@@ -1,0 +1,8 @@
+int _putc(int);
+
+int main(int argc, char **argv) {
+  _putc('m' + sizeof(int));
+  _putc('j' + sizeof(unsigned char));
+  _putc('!');
+  return 0;
+}

### 2017-03-15 abc/paper/paper.tex
Some more on the paper; fix typo
Wrote more in paper. Insert diagram in data segment
Write a bunch more, add topclearance param, generalize pdf2ascii a bit, add printer drawing

@@
 
+I was kidding when I said the x86 architecture is elegant.
+
+Dr. Dobbs Journal typing in the program
+
+ES segment must be used for IN?
+
+TODO: X86 vs x86
+
+(Idea: ladder is actually interspersed in code, by putting some long instruction
+byte (that takes several arguments) before DEC SP JNZ +7e? We can jump directly
+to the jump, but if executed in a normal synchronized flow, then it executes some
+instruction that does nothing (load immediate?).)
+
+(Idea: Once we've gone past EIP=0xFFFF, we can still execute code from our binary.
+It may be possible to get back to CS by doing a jump with the address/operand size prefix?
+Neither one has an effect in DOSBox, which may be a bug? Anyway, if an interrupt
+is triggered while we're up here, it seems like it only does a 16-bit RET, so that's
+kinda busted.)
+
 RISC ROLL
+
+In video: show error in intel manual http:\\www.intel.com
+
+video: bingo of tom 7 video themes (``retro'', ``alphabet,''
+``fractal geometry'')
+
+Ideas for improvement before paper's submitted:
+ - Implement 'void' return type
+ - Make int literals actually 16 bit for now?
+ - Implement multiplication and division
+ - Implement switch
+ - in paper.c, support multichannel sound.
+ - implement sbrk primitive (just returns 0x8000?)
 

### 2017-03-15 abc/struct.c
Thread context through tocil routines. Elaborate struct types, but not actually supporting them yet. Remove 'width' parameter from translvalue continuation, since it is redundant. Some paper hacking. Extremely simple structs test.

@@ -1,0 +1,14 @@
+int _putc(int);
+
+
+
+int main(int argc, char **argv) {
+  struct {
+    int x;
+    char c;
+  } s;
+
+  _putc(s.x);
+
+  return 0;
+}

### 2017-03-15 trunk/abc/paper/paper.tex
Implement sizeof; more paper writing
Thread context through tocil routines. Elaborate struct types, but not actually supporting them yet. Remove 'width' parameter from translvalue continuation, since it is redundant. Some paper hacking. Extremely simple structs test.

@@
 Truly self-documenting
-Giving new meaning to "program committee" "evaluating" my paper
 For good reasons that I will explain later...
@@
 
-missing condition; ``left-facing''
+RISC ROLL
 
-RISC ROLL
+songs to do: https://www.youtube.com/watch?v=EJkJbUbvYJA
+
+In video: show error in intel manual http:\\www.intel.com
 
@@
 ``fractal geometry'')
+
+XXX: Should really write about PSP.
+could write about optimizing the value table
+
+Ideas for improvement before paper's submitted:
+ - Implement 'void' return type
+ - Make int literals actually 16 bit for now?
+ - Implement multiplication and division
+ - Implement switch
+ - in paper.c, support multichannel sound.
+ - implement sbrk primitive (just returns 0x8000?)
 

### 2017-03-15 abc/paper/paper.tex
Point out the spot that keeps getting relocated. This changes the string, which maybe makes a flower joke/misdirection possible?

@@
 
-Thank you for playing wing commander.
-
 Another possibility is to include readable debug information (e.g.
@@
 
-style points for:
-- put relocations a little later after CS, so that we can mark
-  with text the spot where they go.
-
 opportunity for style points, like making your phone number into
@@
  j ? X 4 ?   (4 can alternatively be $ or ,)
-
- NEED TO FIX:
-  - some kind of output other than \_putc
 
@@
 
-dosbox patches, wing commander
-
 The CISC Ridiculous
-
-ascii art in paper
 
@@
 
+songs to do: https://www.youtube.com/watch?v=EJkJbUbvYJA
+
 In video: show error in intel manual http:\\www.intel.com
@@
 ``fractal geometry'')
+
+could write about optimizing the value table
 

### 2017-03-16 trunk/abc/paper/paper.tex
Add weird histo idea

@@
 TODO: X86 vs x86
+TODO: make sure to ctrl-u meta-q every paragraph
 
@@
 
-(Idea: Once we've gone past EIP=0xFFFF, we can still execute code from our binary.
-It may be possible to get back to CS by doing a jump with the address/operand size prefix?
-Neither one has an effect in DOSBox, which may be a bug? Anyway, if an interrupt
-is triggered while we're up here, it seems like it only does a 16-bit RET, so that's
-kinda busted.)
-
 RISC ROLL
@@
 songs to do: https://www.youtube.com/watch?v=EJkJbUbvYJA
+ - mario theme.
 
@@
 
-XXX: Should really write about PSP.
 could write about optimizing the value table

### 2017-03-16 abc/multiply.c
Add weird histo idea

@@ -1,0 +1,9 @@
+int _putc(int);
+
+int main(int argc, char **argv) {
+  int x = (((int)0x4320 + argc) * (int)0x229 + (int)0x22) & (int)0xFF;
+  _putc((int)'o' * (int)argc);
+  _putc(x);
+  _putc((int)11 * (int)3);
+  return 0;
+}

### 2017-04-21 trunk/smeight/paper/paper.tex
Fix typo in paper

\documentclass[twocolumn]{article}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

\usepackage{supertabular}
\usepackage{amsthm}
\usepackage{relsize}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{code}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{textcomp}

\pagestyle{empty}

\newcommand\comment[1]{}

\newcommand\st{$^{\mathrm{st}}$}
\newcommand\nd{$^{\mathrm{nd}}$}
\newcommand\rd{$^{\mathrm{rd}}$}
\renewcommand\th{$^{\mathrm{th}}$}
\newcommand\tm{$^{\mbox{\tiny \textsc{tm}}}$}

% nice fractions
\newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}

\newcommand\citef[1]{\addtocounter{footnote}{1}\footnotetext{\cite{#1}}\ensuremath{^{\mbox{\footnotesize [\thefootnote]}}}}

\usepackage{ulem}
% go back to italics for emphasis, though
\normalem

\begin{document}

\title{The glEnd() of Zelda}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
    Copyright \copyright\ 2016 the Regents of the Wikiplia Foundation.
    Appears in SIGBOVIK 2016 with the danger of going alone of the
    Association for Computational Heresy; {\em IEEEEEE!} press,
    Verlag--Verlag volume no.~0x2016. 255 Rupees} }

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{1 April 2016}

\maketitle

\section*{Abstract}
3D ZELDA

\vspace{1em}
{\noindent \small {\bf Keywords}:
  small keys, boss keys, dungeon keys
}

\section{Introduction}

{\bf 1986. Hyrule.}\quad The Legend of {\it frickin'}~Zelda for the
Nintendo {\it freakin'}~Entertainment System. Need I say more? A {\it
  god damn} \uline{gold cartridge}. Fortunes made just from melting
the gold cartridge down to make gold teeth grilles, after carefully
extracting and preserving the even more precious ROM inside. A die-cut hole in the
box so that you could get a peek of the cartridge and presage that you
were getting some solid gold. A die cut little window that you could
palpate through the wrapping paper on Christmas~Eve, presaging some
epic thumb blisters in store for the coming weeks. Koji {\it freckin'}
Kondo. Koji Kondo whipping up a nice 8-bit arrangement of Bol\'ero as
the theme music until realizing at the last minute that this music was
copyrighted\footnote{Perhaps ironic, since Ravel's {Bol\'ero} itself
  was composed as a result of Ravel getting
  cop-blocked.\cite{wikipedia2016bolero} And surely some Zelda
  knock-off since then has included a Muzak ersatz of the Zelda theme!
  What is the longest documented sequence of compositions due to
  Copyright restrictions?} and so instead composing its epic theme in
{\it one day}??

\begin{figure}[h!t]
\begin{center}
\includegraphics[width=\linewidth]{link3d}
\end{center}\vspace{-0.1in}
\caption{Technical diagram of mathematical equations.} \label{fig:link3d}
% \smallskip
% caption can go here...
% } 
\end{figure}

A gold cartridge that contained ROMs and a little swallowable battery
to keep the onboard SRAM powered up so that it could retain your epic
save game. A battery designed to last 70~years. Nothing could cause
you to lose your save game, even once you were half way through the
Second Quest. Unless your little brother starts a completely new game
and saves over your slot, earning him one of the most righteously
deserved clobberings this side of Inigo Montoya. Saves right on top of
your slot with a player called just \verb+A +. Right over your slot,
erasing it, and hasn't even picked up the {\it sword} yet. All you
have to do is step on the black square in the first room and the guy
gives you the sword and then dies. Jeez pooleaze. Saves right over
yours with no sword even though there are TWO OTHER UNUSED SLOTS and
you even wisely put your game in the third slot {\it exactly to avoid
  this kind of calamity}.

\medskip
Need I say more? Shall I say it? Apparently so. KIDS THESE DAYS. Kids
these days and their three dimensions. You can certainly make a
respectable case that the Zeldest of the series was {\it Legend of
  Zelda III: A Link To The Past}, or even {\it Legend of Game Boy
  Zelda: I forgot What It's Called}. But there seems to be an
established consensus that {\it Legend of Zelda LXIV: The Ocarina of
  Time} is the true masterpiece; that while the NES Original indeed
invented the genre, the technical limitations of the system
handicapped it. You're talking to me about technical {\it frockin'}
limitations? Like somehow appreciating the subtle wonder of the
universe and its mechanics requires me to look at 3D tree people with
tetrahedric clubs for hands. And they say this of the 64\th\ game in
the series? Please jouize.

A fine game, don't get me wrong. But was its theme song composed in
one day? Did it come in a gold cartridge?

And the thing is: The original Zelda was built with enough forethought
that it {\em can} be rendered in first-person 3D. So here I will make
the game 3D so that it can be enjoyed by these kids and they can get
off my lawn, tapdancing around playing their titular Ocarinas. I will
make it possible to play the game in glorious HD, or even 4K
resolution, and then some (Figure~\ref{fig:link3d}). And while
we're at it, why not try to do it in a way that can play loads of NES
games in three {\it fr\"ockin'} dimensions? What's the worst that
could happen?

\section{Technical limitations}

\begin{figure}[h!t]
\begin{center}
\includegraphics[width= \linewidth]{ppu1} \\
\vspace{1em}
\includegraphics[width= \linewidth]{ppu2}
\end{center}\vspace{-0.1in}
\caption{Pattern tables for the Zelda overworld. In Zelda, the top tiles
are used for sprites, and the bottom are used for the background graphics.
All tiles are 8x8 and use 4 color values (color 0 is transparent). These
color values index into 4 sprite palettes and 4 background palettes, with
some limitations.
} \label{fig:ppu}
\end{figure}

As presaged, the NES does have certain technical limitations, which I
prefer to think of as constraints. Some basic details are important
for understanding how this works.

The NES has a modest CPU that executes game logic. It's a little 8-bit
baby puppy with access to 2 kilobytes of RAM. If you think about it,
the NES's 256x240 pixel screen is already 61,440 pixels, which if it
used the entire 2k of RAM to represent it, we'd have only
$0.2\overline{6}$ bits of RAM per pixel, which makes no sense. At
1.7~MhZ, and a generous 1 instruction per pixel, we'd only be able to
render 28 frames per second! This doesn't make sense, and this is
because the CPU does not render the graphics on a NES.

The NES also contains a custom Picture Processing Unit, or PPU, the
8-bit baby puppy's baby ppupy brother. This thing outputs 60 frames of
NTSC video per second. In fact, since the PPU also drives a CPU
interrupt for each scanline and vblank, and most games use
this for timing,%
%
\footnote{Fun fact: NTSC (``Never Twice the Same Color''), the video
  standard used in the Americas, has a $\sim\!\!60$ Hz frame rate. But
  PAL (``Pointless and Lousy''), used in Europe and Asia, has a $\sim\!\!50$
  Hz refresh rate. This means that many games run 16\% slower in
  Europe than the US, so if you grew up playing there, you were
  playing on {\em Easy} mode. {\tt\ ;-)}}
%
in many ways the CPU is subservient to the PPU, kind of like the PPU
is the {\it true} pure heart of the NES. In the context of this paper, you
could say that the CPU is the Ocarina of Time and the PPU is the
Legend of Zelda. It's just an analogy. The PPU is very complicated and
has tetrahedra for hands, like clubs. The CPU communicates with
the PPU not by drawing pixels, but by giving a high-level description
of the screen (by writing it into PPU memory). The chief insight of
this work is that the PPU representation can be interpreted and
rendered in 3D instead of 2D. This allows us to use the exact original
game logic and only change the way it's viewed. Sometimes this even
allows us to see aspects of the game that are actually present but
were not visible on NES hardware. Often it makes the game more
difficult by making things invisible (enemies behind the player or
behind walls) and inducing motion sickness.

\subsubsection{Background}
The PPU has two major drawing facilities: Background and Sprites. Both
are made of ``tiles'', which are 8x8 graphics that are
usually\footnote{Some games trick the NES hardware by remapping tile
memory (e.g. in response to a memory read), even during a scanline. I
do not allow for such chicanery in this work.} stored in ROM. There
are two tile sets, each with 256 tiles. Figure~\ref{fig:ppu} shows the
tile sets for the starting location in the Zelda overworld.

The NES background is described by the ``nametable'', which is an
array of 32x30 bytes giving the tile numbers to fill the screen with.
It's two rows shy of 256 probably for some reason having to do with
the aspect ratio of TVs or number of actual NTSC scanlines, but also
to make room for 64 bytes of attribute data. There are just two bits
per four tiles. These two bits select which of the four palettes to
use for each 2x2 group of tiles on the screen. It's an interesting hobby to
look for the tricks that artists use to work around the constraint on
the number of background colors. % XXX figure showing this off?

Because of this array of tiles, NES games have to be built of
high-level, block-like structures. In Zelda, this structure is
particularly (but not unusually) regular. Since each 2x2 tile group
(16x16 pixels) shares the same palette, the screens are built around
16x16 pixel grid-aligned blocks.

\subsubsection{Sprites} \label{sec:sprites}

Sprites, named in honor of the forest pixies whose homes were razed in
order to make space inside the PPU for them, are more complicated.

There are 64 sprites, which are all turned on or off together. Each
sprite is described by four bytes: Its $x$ and $y$ coordinates (these
can be placed at arbitrary integer pixel locations), the index of
the tile to draw, and an attribute byte:

\begin{tabular}{rrl}
    & bit \# & bit deets \\
\hline
MSB & 7      & Vertical flip \\
    & 6      & Horizontal flip \\
    & 5      & something weird \\
    & 4      & I don't know \\
    & 3      & unused? who can say \\
    & 2      & bit 2; could be anything \\
LSB & 1,0    & Palette index \\
\end{tabular}

Additionally, there are some global sprite settings. The only one that
is important for this presentation is the ``tall sprites'' flag. If
it's set, then sprites are 8x16 (drawn of two consecutive tiles)
instead of 8x8. Sprites are drawn in a particular order. Barring some
bizarre mischief, all of the sprites are drawn; you can't turn them
off individually. Instead, you position the ones you don't want off
the screen or make them be completely transparent. In order to make
things more challenging, if there are exactly 8 sprites on a
particular scan line, some really weird stuff happens. % XXX explain

Sprite Zero has some special properties. I don't want to tell you
about the properties, I just wanted to make a soda joke.

\section[Three to D---$u$ and $v$]{Three to D---$u$ and $v$\footnote{Spoiler alert because I think
this joke may be too abstruse. ``Free to Be You and Me'' was a
semi-famous kids record (here meaning actual vinyl) by 
Marlo~Thomas. D is ``dimensions.'' $u$ and $v$ are common names of
the axes for texture coordinates in computer graphics.}}

The game runs in an emulator, a version of FCEUX\cite{fceux} that I
have hacked up almost beyond recognition. In normal operation, it

\begin{itemize}
\item Emulates a complete frame of the NES. This includes the CPU and
  PPU operation for all 240 scanlines. The execution of a frame depends
  on the current controller input (which doesn't change as it runs);
\item Extracts scene geometry and textures. This consists of two
  pieces: Boxes and Sprites;
\item Extracts the player location and orientation;
\item Moves texture data to the GPU;
\item Transforms the scene geometry into a textured model, sets the camera
  based on the player location, and renders.
\end{itemize}

I'll describe the interesting parts of these.

\paragraph{Emulation.} 
Well, this is not that interesting. I give the emulator its controller
input and it does its thing, modifying the RAM and PPU and CPU state.
I have been working on improvements to emulator technology but these
are independent of this paper and better described elsewhere.

\paragraph{Extracting boxes.} 
Yeah! Okay! Boxes are the three-dimensional version of squares, and
are what give the game its 3Dness. I read the nametable in chunks of
2x2 tiles, which is the size of one overworld tree (and the same size
as Link). Recall that Zelda uses this grid size, probably because
color attributes can only be set at this scale. These become $2x2x2$
cubes. The $x$ and $y$ coordinates are obvious (pixel positions
divided by 16, since each 2x2 tile block is 16x16 pixels), but what
should $z$ be? Zelda and other games tend to use tiles quite
consistently to represent walls, obstacles, floor, empty space, and so
on. This makes sense because a tree graphic, for example, suggests
something to the player. So we simply build a file,
\verb+zelda.tiles+, which maps tiles (by index) to one of \verb+wall+,
\verb+floor+, \verb+rut+, or \verb+unknown+. Floor blocks are placed
at $z=0$, which will be under the player's feet (we typically only see
the top face). Walls are placed at $z=2$, so that their bottoms are
coplanar with the floor. Rut blocks are placed at $z=-0.5$; this is
used for blocks that are part of the ``floor'' but that can't be
walked on, like water. Because each box is actually made of 4 tiles,
we just use the maximum $z$ value for the constituent tiles (they should
normally be the same). It would make sense to develop heuristics for
when games use tile-level resolution, even to generate primitives
other than cubes.

This approach is a bit unsatisfying because it requires us to have
specific knowledge of the game as well as do some manual labor.
However, there really aren't that many tiles, and it's kind of fun.
(In fact, you really only need to pick out a handful of floor tiles
before the game is totally playable.) There is another problem, which
is that many games switch tile sets, which may change whether tile
\verb+0x84+ (``desert dots'' in the overworld) is floor or wall. What
happens if you go into a dungeon in Zelda, for example? I didn't even
try that yet but I bet it's nuts! This can easily be fixed by
identifying distinct tile sets (for example by using the source ROM
address, or a hash of their content) and mapping them all.

However, there's some hope that this can be done in a much more
automatic way. I describe an idea in Section~\ref{sec:cast}.

\paragraph{Box textures.}
The game is pretty understandable with just TRON-cubes, but it is
better to texture them. We put the same texture on each face of a box;
that texture is just the 16x16 pixel graphic consisting of the 4 tiles
that made it, in the appropriate palette. It does {\it not} work to
generate a bunch of tile-sized boxes (with the individual tile texture
on all sides); these will only look right on the top face. Indeed, it
is possible to get weird texture mismatches between these 2x2x2 boxes
for basically the same reason. It usually looks pretty good, though.

Conceivably, every box on the screen could have a different texture,
and conceivably these all change every frame. It is inefficient to
move 960 distinct textures to video memory every frame, so instead we
actually just move the rendered background (256x240) itself and record
with each box its texture coordinates within. This is {\it much}
faster, though it can create artifacts at the edges of boxes as an
adjacent pixel peeks into the texture. I checked out Ocarina of Time
and it is {\it loaded} with texture seam issues, so I figure this is
how they like it.

\paragraph{Extracting sprites.}
Sprites are the most difficult. Aside from just being more fiddly to
implement, it is tricky to recover an important high-level structure.
Although we know the location of each individual sprite and the two tiles
(Zelda uses ``tall' 8x16 sprites; see Section~\ref{sec:sprites}) that
make it up, most game objects (like enemies) are in fact drawn with
{\it two} sprites. These sprites could be in any of the 64 sprite
slots (it's not at all obvious how Zelda allocates these, or that it
would be similar for other games). Adding to the complexity, sprites
like Link's sword or the pulsing balls of plasma that frogmen shoot
at you from the water {\it just for being an elf} are just a single
sprite, and bosses can be like 6 or more sprites (Figure~\ref{fig:spider}). If we
promote sprites to 3D without knowing which ones are supposed to be
part of the same object, then the enemies get cut up along their
sprite seems and it looks super wrong.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{spider}
\end{center}\vspace{-0.1in}
\caption{Link's spider nemesis, made of six 8x16 sprites. The sprites
have adjacent edges, so they end up being fused into one merged sprite.
} \label{fig:spider}
\end{figure}

To fix this, we perform {\it sprite fusion}. We say that two sprites
are part of the same object if they share a complete edge
(transitively). Since this allows the construction of very tall ($64
\times 16$) or wide ($64 \times 8$) sprites, we have to make sure that
our textures are large enough to contain large sprites. Rather than
try to pack them into a texture at runtime (computationally expensive
since they can also be weird shapes), we just pre-allocate one texture
of the maximum size for each sprite slot (16MB video ram; nothing).
Sprite fusion, computation of the sprite size, drawing the sprites,
extracting the geometry, and moving the textures to the video card are
all intertwined. The simplified algorithm is

\begin{itemize}
\item Create 64 ``pre-sprites'', which are baby sprites. These know
  their coordinates and have the index of a ``parent'' pre-sprite
  (initialized to $-1$).
\item While filling these in, consult a hash-map to see if there are
  already pre-sprites exactly aligned in each of the cardinal
  directions. If so, use a union-find--like algorithm\cite{tarjan1975efficiency}
  to efficiently update its parent index; each fused sprite has one
  canonical representative whose parent index is $-1$. Add the
  pre-sprite to the hash-map.
\item Another pass pushes the min/max bounds to the canonical representative,
  so that we know how big the fused sprite is.
\item Another pass draws all sprites into the texture data for their
  canonical representative, at the appropriate location. This pass must
  happen in the same order as on the NES hardware, so that overlay effects
  are correct.
\item Finally, create proper sprites to return for each of the fused sprites.
  Copy the used region of the texture data for these to the video card.
\end{itemize}

% XXX diagram of this algorithm would be nice

Two types of sprite rendering are supported: in-plane and billboard.
In-plane just draws the sprites on the floor or some other $xy$ plane
of your choice. (This does not actually need sprite fusion, but it is
not harmful.) Billboard sprites are drawn so that they face the
player, like in Wolfenstein~3D. These are positioned at their $xy$
centroid and positioned in $z$ so that the bottom of the sprite is
touching the floor.

\paragraph{Extracting location.}

Most NES games use very simple layouts for their game facts (i.e.,
global variables at fixed memory locations), since the processor is
simple and memory light.\cite{murphy2013first} In Zelda, Link's
$x$ coordinate is at \verb+0x70+ and $y$ at \verb+0x84+. Since this
is his top-left corner, we add 8 to each to get to the middle of the
16x16 pixel boy.

We also have to face the camera in the same direction as Link, to
make the perspective first-person. Link's orientation is at memory
location \verb+0x98+.

Link moves smoothly (with pixel resolution) around the screen, but
turns abruptly. This is sickening to watch. I apply some smoothing to
Link's angle, by treating the location from memory as a ``target''
angle and twisting towards it (along the shortest angular path) at a
limited speed.

Knowing what memory locations hold the player's position and
orientation also involves some manual work. You could imagine
heuristics for trying to determine this, by seeing which bytes tend to
go up when pressing right on the controller, and so on. However, it's
easy to find these values manually or to even look them up for popular
games.

\paragraph{Rendering.}

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\linewidth]{bridgewide}
\end{center}\vspace{-0.1in}
\caption{A representative view of the Legend of Zelda in glorious
  3D.} \label{fig:bridgewide}
\end{figure*}

Finally, we can render the scene. I used ``classic'' OpenGL interface,
even though it is terrible, for the sake of making good on the
excellent pun that is this paper's title. The final result looks like
Figures~\ref{fig:bridgewide} and~\ref{fig:alone}.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{alone}
\end{center}\vspace{-0.1in}
\caption{The first cave, where this man gives you a sword and then dies,
rendered in 3D. Letters are treated as walls so you can see 'em upright,
but then of course the second line is in front of the first. The 2D NES
screen is superimposed on the top left for comparison.
} \label{fig:alone}
\end{figure}

Rendering with OpenGL is quite straightforward given the geometry
extracted from the previous sections. I try to preserve the pixel
aesthetic by using \verb+GL_TEXTURE_MAG_FILTER+ = \verb+GL_NEAREST+,
but at this point we can use whatever approach to 3D graphics most
titillates the youth. Full-screen multisample anti-aliasing? Sure.
Screen Space Ambient Occlusion? I hear that this creates physically
realistic shadows on the cheap, sure. Stencil buffer shadows?
Volumetric fog? God rays? Depth-of-field? Lens flares? You got it.

\subsection{Quirks}
There are some problems or quirks with the current approach.

\paragraph{Extramural geometry.}
In Zelda, some of the tiles on screen are used to draw a menu at the
top, which is not part of the gameplay area. This still gets
interpreted as part of the geometry of the world, of course
(Figure~\ref{fig:offscreen}). It would be a simple matter to crop the
region of interest (especially if done manually), and it would even be
natural to overlay it on a heads-up display. But I like it over there.
Why shouldn't menus be 3D as well?

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{offscreen}
\end{center}\vspace{-0.1in}
\caption{ The game menu visible to the North because the ocean does
  not get in its way. Did you know that if you were able to see beyond
  the cluttering trees and mountains at the horizon of Earth, you'd
  see your own status menu in real life? } \label{fig:offscreen}
\end{figure}

\paragraph{Positive self-image.}
We place the camera at Link's position and orientation, so that the
player ``becomes'' Link. However, Link is still represented by a
sprite in the game, and we do not treat it specially. This means that
the fused Link billboard is always in the scene geometry right at the
location of the camera. The upshot is that whenever the camera is
facing south, an enormous Link graphic fills the entire screen,
reminding you that he has always been following a few pixels behind
you, haunting you (Figure~\ref{fig:megalink}). This sprite could be
rejected by seeing that it is coincident with the player's location
(which we extract) or by its particular content. We could also set the
``near plane'' so that nearby sprites are not visible. However, the
everpresent looming spectre of a ghost mimicking our every move is a
good reminder of our mortality, so I kept it.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{megalink}
\end{center}\vspace{-0.1in}
\caption{Look who's behind you! WE ARE FRIENDS FOREVER {\tt:-) :-)}
} \label{fig:megalink}
\end{figure}

\paragraph{Surprising sprite fusion.}
Sprite fusion works well for merging sprites that are part of the same
logical object in Zelda. (It may not work as well in other games where
e.g. a monster's legs are moved independently from its body.) However,
it occasionally mistakenly fuses two sprites that just incidentally
share an edge (Figure~\ref{fig:badfusion}). This usually persists for
just a tiny moment since the sprites are moving. But who's to say that
as those monsters high-five, they aren't forming a momentary totem?
Only Shigeru {\it froakin'} Miyamoto can say that isn't what he intended.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{bad-sprite-fusion}
\end{center}\vspace{-0.1in}
\caption{Two Octoroks that were incorrectly fused because they have
constituent sprites that have adjacent edges (the fused sprite shape is
not rectangular). The formation is a Hexadecirok.
} \label{fig:badfusion}
\end{figure}

\paragraph{Texture orientation.} The textures of both backgrounds and sprites
often have strange and inconsistent perspective ideas in video games.
For example, the Octorok points its nose in the direction it's moving,
because it is always viewed ``from above''. However, when Link is
facing south, you see his front (not the top of his head facing down),
and when facing west you see his side. This doesn't make sense, and
makes it impossible to apply any consistent rule for mapping sprites
into world orientation. This mostly just produces humorous results,
like when Link throws his sword and it moves away from him in first
person view: It may be pointing towards the sky, towards the floor, or
to either side, but never in the direction it's thrown. Even if we
were to render the sword ``in plane'', its handle should always be
oriented towards the camera (which would not be the case). No single
policy will work for all cases, but perhaps heuristics can look for
the use of the horizontal and vertical flip bits to figure out what
sprites are rotatable, and then rotate them relative to the camera.

\section{Future work}

This work is literally in progress, and due to Draconian SIGBOVIK
deadlines, this paper is figuratively already out of date by the time
you read it. You can check out the project's home page

\medskip
\begin{center}
\url{http://tom7.org/zelda}
\end{center}
\medskip

and maybe download it and play it, and probably watch a video
demonstration. There may be some new features. Depends what I
get to, you know?

\subsection{Control}
Currently, the game is viewed first person but played with native
controls; this means that pressing ``up'' always moves link to the
North, turning the camera if necessary. This is super weird. However,
it is pretty easy to remap first person controls into native controls,
as long as there is a way of implementing that intent (for example, it
is not possible to ``strafe'' in Zelda). We can change the camera
angle arbitrary, implementing ``mouselook'' or virtual reality
support. Since the z axis is otherwise unused, we could allow the player
to jump, though this would do nothing in the game itself.

\subsection{Other games}
This program was built in such a way that it should take minimal
effort make other NES games 3D, perhaps with humorous results. There
are some issues that need to be dealt with. A common type of game is
the side-scrolling platformer, like Super Mario Bros.. At least two
important issues are not yet handled: Mario has a different coordinate
system than Zelda (it is ``side view'' rather than ``top down''), so
we need to support that rendering style and control/camera scheme. No
big deal. Second, this game scrolls the screen rather than using
single-screen layouts; this uses the PPU in a different and somewhat
more complex way. Many games will use advanced PPU tricks that cannot
be interpreted in a generic high-level way.

\subsection{A Link to the cast} \label{sec:cast}

One dissatisfying problem with the current approach is that it
requires some manual effort to map different tile types to geometry.
One could imagine various heuristics (even a image classifier) to
provide defaults or to remove the manual input entirely. An exciting
angle is to use the game's own understanding of these tiles. The main
thing that differentiates a ``wall'' from a ``floor'' is whether the
player can walk over it. Clearly somewhere in the game's code it
``knows'' what is floor and what is wall. We could extract that
directly, but this would require a different manual step.
Alternatively, we could determine it experimentally. A simple
heuristic would keep track of what tiles the player has walked on. But
we are not limited to what the player actually does; since we are
emulating, we can also emulate counterfactual scenarios. Imagine that
as we draw a first person view, we cast lines out from the camera
towards the geometry, as in a ray tracer. Except that instead of
tracing light rays until an intersection with scene geometry, we trace
a hypothetical player path by emulating the player moving in that
direction, until Link can move no more. This only requires
understanding the bytes that correspond to the player's location
(which we already need) and how the controller works.\footnote{Most
  rays are diagonal. We can implement a north-northwest diagonal by
  repeatedly moving two pixels north, then one west. We could also set
  the player's position directly to a putative destination and then
  move in cardinal directions to test solidness.} Testing contact with
sprites might give us useful heuristics about how they should be
rendered as well.

There are many ways we can imagine this going wrong, but probably
spectacularly so. Interestingly, this would allow us to render parts
of the game that are not even visible in the PPU yet, since emulating
the player's path ahead far enough would cause the game to switch
to future scenes.

\section{Conclusion}

In this paper I made Zelda three-dimensional so that it can be enjoyed
by the kids. I am fully aware that these ``kids'' are actually also
full-grown adults who themselves resent a slightly younger generation
who inexplicably think of Legend of Zelda Twilight Princess HD as the
canonical episode; these people should get to work at enhancing Ocarina
of Time to add Amiibo support. While you're at it, get rid of the
really long section where you have to be a dog. I didn't like that part.

\bibliographystyle{IEEEtran}
% \bibliographycomment{} % nothing
\bibliography{paper}

\end{document}

### 2018-03-06 trunk/citations/paper.tex
Version that works, produces graph

strange authors:

capinha para celular horizonte artificial iphone

### 2018-03-06 trunk/citations/authorstats.cc

@@ -1,0 +1,195 @@
+
+#ifdef __MINGW32__
+#include <windows.h>
+#undef ARRAYSIZE
+#endif
+
+#include "base/logging.h"
+#include "base/stringprintf.h"
+#include "util.h"
+#include "threadutil.h"
+#include "re2/re2.h"
+#include "textsvg.h"
+
+#include <cmath>
+#include <vector>
+#include <string>
+#include <map>
+#include <unordered_map>
+
+using namespace std;
+
+template<class K, class C>
+inline bool ContainsKey(const C &container, const K &key) {
+  return container.find(key) != container.end();
+}
+
+struct AuthorStats {
+  int64 articles = 0;
+  int64 citations = 0;
+};
+
+template<class F>
+static void LocalForEachLine(const string &filename, F f) {
+  // printf("Readfiletolines %s\n", filename.c_str());
+  vector<string> lines = Util::ReadFileToLines(filename);
+  // printf("%d lines\n", (int)lines.size());
+  for (int i = 0; i < lines.size(); i++) {
+    const string &line = lines[i];
+    if (0 == i % 250000) printf("%.2f%%\n", (i * 100.0) / lines.size());
+    f(line);
+  }
+}
+
+template<class T>
+void Reverse(vector<T> *v) {
+  vector<T> ret;
+  ret.reserve(v->size());
+  for (int i = v->size() - 1; i >= 0; i--)
+    ret.push_back(std::move((*v)[i]));
+  v->swap(ret);
+}
+
+bool Dictionaryize(string author, string *dict) {
+  vector<string> tokens;
+  while (!author.empty()) {
+    string tok = Util::chop(author);
+    if (!tok.empty()) tokens.push_back(Util::lcase(tok));
+  }
+
+  // Totally whitespace name?
+  if (tokens.empty())
+    return false;
+
+  Reverse(&tokens);
+  
+  // Make sure it fits somewhere in alphabetical order.
+  if (tokens[0][0] >= 'a' && tokens[0][0] <= 'z') {
+    dict->clear();
+    for (const string &tok : tokens) {
+      if (dict->empty()) {
+	*dict = tok;
+      } else {
+	*dict += " ";
+	*dict += tok;
+      }
+    }
+    return true;
+  } else {
+    return false;
+  }
+}
+
+static string Rtos(double d) {
+  if (std::isnan(d)) return "NaN";
+  char out[16];
[…]

### 2018-03-06 trunk/citations/merge.cc

@@ -1,0 +1,99 @@
+
+#ifdef __MINGW32__
+#include <windows.h>
+#undef ARRAYSIZE
+#endif
+
+#include "base/logging.h"
+#include "base/stringprintf.h"
+#include "util.h"
+#include "threadutil.h"
+#include "re2/re2.h"
+
+#include <vector>
+#include <string>
+#include <map>
+#include <unordered_map>
+
+using namespace std;
+
+template<class K, class C>
+inline bool ContainsKey(const C &container, const K &key) {
+  return container.find(key) != container.end();
+}
+
+struct AuthorStats {
+  int64 articles = 0;
+  int64 citations = 0;
+};
+
+template<class F>
+static void LocalForEachLine(const string &filename, F f) {
+  // printf("Readfiletolines %s\n", filename.c_str());
+  vector<string> lines = Util::ReadFileToLines(filename);
+  // printf("%d lines\n", (int)lines.size());
+  for (int i = 0; i < lines.size(); i++) {
+    const string &line = lines[i];
+    // if (0 == i % 1000) printf("%d\n", i);
+    f(line);
+  }
+}
+
+int main(int argc, char **argv) {
+  #ifdef __MINGW32__
+  if (!SetPriorityClass(GetCurrentProcess(), BELOW_NORMAL_PRIORITY_CLASS)) {
+    LOG(FATAL) << "Unable to go to BELOW_NORMAL priority.\n";
+  }
+  #endif
+
+  string outfile = "authors.txt";
+  
+  vector<string> filenames;
+  for (int i = 1; i < argc; i++) {
+    string arg = (string)argv[i];
+    if (arg == "-o") {
+      CHECK(i < argc - 1);
+      outfile = argv[i + 1];
+      i++;
+    } else {
+      filenames.push_back(arg);
+    }
+  }
+  
+  if (filenames.empty()) {
+    fprintf(stderr, "Pass JSON paper files on the command line.\n");
+    return -1;
+  }
+  
+  // Number of articles authored by each name.
+  std::unordered_map<string, AuthorStats> author_stats;
+  RE2 line_re{"([^\t]*)\t([\\d]+)\t([\\d]+)"};  
+  auto OneFile = 
+    [&author_stats, &line_re](const string &filename) {
+    printf("%s\n", filename.c_str());
+    LocalForEachLine(filename,
+		     [&author_stats, &line_re](string line) {
+      string author;
+      int64 articles = 0LL, citations = 0LL;
[…]

### 2018-03-10 citations/paper.tex
Also pull out title words; a little work on paper

\documentclass[twocolumn]{article}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

% \usepackage{supertabular}
\usepackage{amsthm}
% \usepackage{relsize}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{code}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{textcomp}

\pagestyle{empty}

\newcommand\comment[1]{}

\newcommand\st{$^{\mathrm{st}}$}
\newcommand\nd{$^{\mathrm{nd}}$}
\newcommand\rd{$^{\mathrm{rd}}$}
\renewcommand\th{$^{\mathrm{th}}$}
\newcommand\tm{$^{\mbox{\tiny \textsc{tm}}}$}

% nice fractions
\newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}

% \newcommand\citef[1]{\addtocounter{footnote}{1}\footnotetext{\cite{#1}}\ensuremath{^{\mbox{\footnotesize [\thefootnote]}}}}

\newcommand\dirham{\protect \includegraphics[width=1em]{dirham}}

% \usepackage{ulem}
% go back to italics for emphasis, though
% \normalem

\begin{document}

% Academic Advancement Advice: Adopt A A Authorship

\title{Want to succeed in academia? Take the name A A}
\author{A.~A. \\
  (tom7.org foundation)
  \thanks{
    Copyright \copyright\ 2018 the Regents of the Wikiplia Foundation.
    Appears in SIGBOVIK 2018 with the abecedarian advantage of the
    Association for Aomputational Aeresy; {\em IEEEEEE!} press,
    Verlag--Verlag volume no.~{\tt std::numeric\_limits<float>::lowest()}.
    1 \dirham
} }

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{1 April 2018}

\maketitle

\section*{Abstract}
proof at last
    1\ \dirham

\vspace{1em}
{\noindent \small {\bf Keywords}:
  A A A A A A
}

\section{Introduction}

Those with names falling in the dead-middle of the alphabet---like as
a random example, Murphy---have suspected since early life a
disadvantage. For example, when the teacher lines up the students for
pie, in order to be completely fair and avoid bias, orders them by the
perfectly objective criterion of alphabetical order. Occasionally, the
teacher realizes that the same students always get pie first, and
corrects this bias by using {\em reverse} alphabetical order.
Sometimes the teacher uses alphabetical or reverse-alphabetical by
first name, or sometimes last name. All variations of this idea are of
course unfair, as is clear to first- and second-graders, who are
powerless to rebel without (a) status in the K-12 hierarchy (b) the
statistical language in which to formalize their objection and (c)
access to a prestigious conference with permissive publication
standards.

This paper demonstrates that even within academic paper publishing
(which one could argue is obsessed with bias, prestigious conferences,
statistical language for formalizing objections, and status in the
K-12 hierarchy), this unfairness persists. Specifically, we find that
authors whose names come alphabetically earlier have higher citation
rates.

\section{The Data}

I obtained a FISA warrant based solely on a salacious and unverified
memo funded by SIGBOVIK's political opponents. The warrant was for the
AMiner Open Academic Graph, which consists of summary information for
154,771,162 papers in about 130 gigabytes of JSON data. Then I embarked
on one of the Classic Journeys of Computer Science.

\subsection{Classic Journey of Computer Science}

The task is apparently simple: Parse 130 gigabytes of JSON data to
extract the author names and citations from the papers.

multiple hypothesized causes:

 - When doing a

300 GB of json woes

strange authors:

capinha para celular horizonte artificial iphone

\end{document}

### 2018-03-10 citations/citation-util.h

@@ -1,0 +1,21 @@
+
+#include "util.h"
+
+#include <string>
+#include <vector>
+
+using namespace std;
+
+template<class K, class C>
+inline bool ContainsKey(const C &container, const K &key) {
+  return container.find(key) != container.end();
+}
+
+template<class F>
+static void LocalForEachLine(const string &filename, F f) {
+  vector<string> lines = Util::ReadFileToLines(filename);
+  for (int i = 0; i < lines.size(); i++) {
+    const string &line = lines[i];
+    f(line);
+  }
+}

### 2018-03-10 trunk/citations/words.cc
Also pull out title words; a little work on paper

@@ -1,0 +1,170 @@
+
+#ifdef __MINGW32__
+#include <windows.h>
+#undef ARRAYSIZE
+#endif
+
+#include "rapidjson/document.h"
+#include "base/logging.h"
+#include "base/stringprintf.h"
+#include "util.h"
+#include "threadutil.h"
+#include "citation-util.h"
+
+#include <vector>
+#include <string>
+#include <map>
+#include <unordered_map>
+
+using namespace rapidjson;
+using namespace std;
+
+struct WordStats {
+  int64 articles = 0;
+  int64 citations = 0;
+};
+
+string Normalize(string w) {
+  // Lowercase ASCII letters.
+  for (char &c : w) {
+    if (c >= 'A' && c <= 'Z') c |= 32;
+  }
+
+  // Nonstandard quotation marks
+  w = Util::Replace(std::move(w), "”", "\"");
+  w = Util::Replace(std::move(w), "“", "\"");
+  w = Util::Replace(std::move(w), "‘", "'");
+  w = Util::Replace(std::move(w), "’", "'");
+  
+  // Remove punctuation from the end of words.
+  [&w]() {
+    while (!w.empty()) {
+      switch (w.back()) {
+      case ':':
+      case ',':
+      case '.':
+      case '?':
+      case ';':
+      case '!':
+      case '\'':
+      case '\"':
+      case ')':
+      case ']':
+	w.resize(w.size() - 1);
+	break;
+      default:
+	return;
+      }
+    }
+  }();
+
+  // And a few things from the front of words:
+  w = [](const string &w) -> string {
+    for (int i = 0; i < w.size(); i++) {
+      switch (w[i]) {
+      case '(':
+      case '\'':
+      case '\"':
+      case '[':
+	break;
+      default:
+	return w.substr(i, string::npos);
+      }
+    }
+    return "";
+  }(w);
+  
+  // U+2013 EN DASH becomes hyphen
+  // w = Util::Replace(std::move(w), "\u2013", "-");
+  w = Util::Replace(std::move(w), "–", "-");
+  w = Util::Replace(std::move(w), "—", "-");
+  return w;
+}
+
+int main(int argc, char **argv) {
+  #ifdef __MINGW32__
+  if (!SetPriorityClass(GetCurrentProcess(), BELOW_NORMAL_PRIORITY_CLASS)) {
+    LOG(FATAL) << "Unable to go to BELOW_NORMAL priority.\n";
+  }
+  #endif
+
+  vector<string> filenames;
+  string outfile;
[…]

### 2018-03-11 trunk/citations/paper.tex
a little on paper

@@
+\documentclass[twocolumn]{article}
+\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}
+
+% \usepackage{supertabular}
+\usepackage{amsthm}
+% \usepackage{relsize}
+\usepackage{amsmath}
+\usepackage{amssymb}
+% \usepackage{code}
+\usepackage{graphicx}
+\usepackage{fancyvrb}
+\usepackage{url}
+\usepackage{textcomp}
+
+\pagestyle{empty}
+
+\newcommand\comment[1]{}
+
+\newcommand\st{$^{\mathrm{st}}$}
+\newcommand\nd{$^{\mathrm{nd}}$}
+\newcommand\rd{$^{\mathrm{rd}}$}
+\renewcommand\th{$^{\mathrm{th}}$}
+\newcommand\tm{$^{\mbox{\tiny \textsc{tm}}}$}
+
+% nice fractions
+\newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}
+
+% \newcommand\citef[1]{\addtocounter{footnote}{1}\footnotetext{\cite{#1}}\ensuremath{^{\mbox{\footnotesize [\thefootnote]}}}}
+
+\newcommand\dirham{\protect \includegraphics[width=1em]{dirham}}
+
+% \usepackage{ulem}
+% go back to italics for emphasis, though
+% \normalem
+
+\begin{document}
+
+% Academic Advancement Advice: Adopt A A Authorship
+% Academic Advancement Advice: Author Articles as A A
+
+\title{Academic Advancement Advice: Author Articles as A.~A.}
+\author{A.~A. \\
+  (tom7.org foundation)
+  \thanks{
+    Copyright \copyright\ 2018 the Regents of the Wikiplia Foundation.
+    Appears in SIGBOVIK 2018 with the abecedarian advantage of the
+    Association for Aomputational Aeresy; {\em IEEEEEE!} press,
+    Verlag--Verlag volume no.~{\tt std::numeric\_limits<float>::lowest()}.
+    1 \dirham
+} }
+
+\renewcommand\>{$>$}
+\newcommand\<{$<$}
+
+\date{1 April 2018}
+
+\maketitle
+
+\section*{Abstract}
+proof at last
+    1\ \dirham
+
+\vspace{1em}
+{\noindent \small {\bf Keywords}:
+  A A A A A A
+}
+
+\section{Introduction}
+
+Those with names falling in the dead-middle of the alphabet---like as
+a random example, Murphy---have suspected since early life a
+disadvantage. For example, when the teacher lines up the students for
+pie, in order to be completely fair and avoid bias, orders them by the
+perfectly objective criterion of alphabetical order. Occasionally, the
+teacher realizes that the same students always get pie first, and
+corrects this bias by using {\em reverse} alphabetical order.
+Sometimes the teacher uses alphabetical or reverse-alphabetical by
+first name, or sometimes last name. All variations of this idea are of
+course unfair, as is clear to first- and second-graders, who are
+powerless to rebel without (a) status in the K-12 hierarchy (b) the
+statistical language in which to formalize their objection and (c)
+access to a prestigious conference with permissive publication
+standards.
+
+This paper demonstrates that even within academic paper publishing
+(which one could argue is obsessed with bias, prestigious conferences,
+statistical language for formalizing objections, and status in the
+K-12 hierarchy), this unfairness persists. Specifically, we find that
+authors whose names come alphabetically earlier have higher citation
+rates.
+
+\section{The Data}
+
+I obtained a FISA warrant based solely on a salacious and unverified
+memo funded by SIGBOVIK's political opponents. The warrant was for the
+AMiner Open Academic Graph, which consists of summary information for
+154,771,162 papers in about 130 gigabytes of JSON data. Then I embarked
+on one of the Classic Journeys of Computer Science.
+
+\subsection{Relatable Computer Science Activities}
+
+The task is apparently simple: Parse 130 gigabytes of JSON data to
+extract the author names and citations from the papers and count them.
+To begin my journey, I selected a C++ JSON library from 39
+alternatives listed on a benchmark page.\cite{yip2018native} The first
+criterion I ordered by was ``correctness,'' for which only two parsers
+achieved the highest correctness score, $100\%$. I selected the slower
+(i.e., more methodical) of the two parsers, since I did not want to do
+shoddy work.
+
+Next I spent a nice Saturday morning integrating this library into my
+private build environment, making sure its open source license was
+properly declared and so on, while I waited for the 130 gigabytes of
+JSON to download. Miraculously, this was straightforward. I was able
+to finish writing the code to process the articles before they even
+finished downloading. However, upon unzipping, I found that each file
+contained one million articles, not as a single JSON array but as a
+series of adjacent objects. The JSON library allowed for reading JSON
+from files, but {\em not} e.g. for reading a single record from a file pointer
+repeatedly. So, I needed to do my own ad hoc parsing before feeding
+each record to the JSON library as strings. OK. Now when I run it, it
+prints
+\begin{verbatim}
+  [Segmentation fault
+\end{verbatim}
+
+This is not my first segmentation fault---let me tell you!---so I knew
+what to do. I debugged the program, adding assertions using a
+\verb+CHECK+ macro that aborts if its condition is violated. I
+eventually determined that the bug must be inside the JSON library
+itself, since the segmentation fault happened between two
+\verb+CHECK+s where the only code being run was JSON library code. So,
+deciding that perhaps its $100\%$ correctness benchmark score was
+overstated, I downloaded and repeated the process with the other,
+faster (i.e., more slapdash) parser. Of course this parser has a
+different API so I needed to rewrite my program. (The program is very
+simple so this was the fastest step; however, getting the library
+properly incorporated and compiling and then learning its API was
+still a computer science chore.) Once my program was rid of the old
+crashy library, I ran it, and it produced this result
+\begin{verbatim}
+  [Segmentation fault
+\end{verbatim}
+Since these two libraries produced the same result, either
+segmentation violation must be part of the JSON specification, or the
+bug is instead elsewhere in the program. After reducing the program to
+a fairly simple test case, I determined that it must be some kind of
+``Heisenbug'' whereby an earlier piece of code was corrupting the heap
+or similar, causing a later crash only when present. This was not my
+first Heisenbug---let me tell you!---so I knew what to do. However,
+not being able to use Valgrind on Windows (did I mention that I'm
+using Cygwin and mingw64 on the eponymous Windows~7 by the way? Why am
+I doing that?), I booted Linux in a virtual machine, installed the 87
+security updates that had been released since I last booted this
+virtual machine a few months ago, and compiled my program there. I
+then needed to move one of the JSON files to the virtual computer to
+test, without having to download 130 GB from the internet again, but
+the file was apparently too large to transfer using the mysterious
+``Drag and drop to virtual machine'' functionality. I recalled that
+VirtualBox can mount a shared folder from the host machine if you edit
+the \verb+/etc/fstab+ file to contain the correct arcane magic, and
+that I had performed this ritual before, but then had commented-out
+that line in ``rescue mode'' because I also learned that having this
+line present during the boot process causes a kernel panic. So I
+temporarily reinstated the line and mounted the drive and copied the
+gigantic file into the virtual machine and then removed the dangerous
+magic from the f-stab for f-safety.
+
+Next I ran Valgrind on this minimal test case and it worked as
+expected with no unclean behavior detected and no crashing.
+
+Since Linux was no help debugging, I returned to Windows~7 and tried
+single-step debugging. Here I discovered that the crash in my minimal
+test case was occurring when using \verb+cerr+. Specifically, the
+second thing sent to \verb+cerr+ would produce a segmentation fault.
+I then recalled that the \verb+CHECK+ macro uses \verb+cerr+ and
+also found code I had written in the Year 2014
+
+The original ``bug'' was a violated assumption: I assumed,
+optimistically, that if the article had an {\tt author} field then
+the author would be a string. I accounted for a missing author,
+an author that was the empty string, but not {\tt author: null}.
+
+multiple hypothesized causes:
+
+ - When doing a
+
+300 GB of json woes
+
+\subsection{Alliterative Articles}
+
+Adopting an all-'A' affect also advances articles.
+Awesome!
+
 strange authors:
@@
 capinha para celular horizonte artificial iphone
+
+\bibliographystyle{IEEEtran}
+% \bibliographycomment{} % nothing
+\bibliography{paper}
+
+\end{document}

### 2018-03-11 citations/words.cc
a little on paper

@@ -8,21 +8,16 @@
 #include "base/logging.h"
 #include "base/stringprintf.h"
 #include "util.h"
-#include "threadutil.h"
 #include "citation-util.h"
 
 #include <vector>
 #include <string>
 #include <map>
 #include <unordered_map>
+#include <unordered_set>
 
 using namespace rapidjson;
 using namespace std;
-
-struct WordStats {
-  int64 articles = 0;
-  int64 citations = 0;
-};
 
 string Normalize(string w) {
   // Lowercase ASCII letters.
@@ -30,6 +25,24 @@ string Normalize(string w) {
     if (c >= 'A' && c <= 'Z') c |= 32;
   }
 
+  // Characters that shall not occur.
+  {
+    string other;
+    for (char c : w) {
+      switch (c) {
+      case '\n':
+      case '\r':
+      case '\t':
+      case '\0':
+	break;
+      default:
+	other += c;
+      }
+    }
+    other.swap(w);
+  }
+    
+  
   // Nonstandard quotation marks
   w = Util::Replace(std::move(w), "”", "\"");
   w = Util::Replace(std::move(w), "“", "\"");
@@ -113,7 +126,7 @@ int main(int argc, char **argv) {
   int64 counter = 0LL, no_title = 0LL;
 
   // Stats for each title word.
-  std::unordered_map<string, WordStats> word_stats;
+  std::unordered_map<string, CiteStats> word_stats;
 
   for (const string &filename : filenames) {
     LocalForEachLine(filename,
@@ -131,7 +144,7 @@ int main(int argc, char **argv) {
 	}
 	
 	auto Count = [n_citation, &word_stats](const string &word) {
-	  WordStats &stats = word_stats[word];
+	  CiteStats &stats = word_stats[word];
 	  stats.articles++;
 	  stats.citations += n_citation;
 	};
@@ -139,10 +152,17 @@ int main(int argc, char **argv) {
 	const Value &title = article["title"];
 	if (title.IsString()) {
 	  string words = title.GetString();
+	  // Non-breaking space?
+	  words = Util::Replace(std::move(words), " ", " ");
+	  // Only count a word once per title.
+	  std::unordered_set<string> boolean_words;
 	  while (!words.empty()) {
 	    string word = Util::chop(words);
-	    if (!word.empty()) Count(Normalize(std::move(word)));
+	    if (!word.empty()) {
+	      boolean_words.insert(Normalize(std::move(word)));
+	    }
[…]

### 2018-03-11 trunk/citations/authorstats.cc
more analysis, writing

@@ -18,16 +18,7 @@
 #include <map>
 #include <unordered_map>
 
-template<class T>
-void Reverse(vector<T> *v) {
-  vector<T> ret;
-  ret.reserve(v->size());
-  for (int i = v->size() - 1; i >= 0; i--)
-    ret.push_back(std::move((*v)[i]));
-  v->swap(ret);
-}
-
-bool Dictionaryize(string author, string *dict) {
+static bool Dictionaryize(string author, string *dict) {
   vector<string> tokens;
   while (!author.empty()) {
     string tok = Util::chop(author);
@@ -55,15 +46,6 @@ bool Dictionaryize(string author, string *dict) {
   } else {
     return false;
   }
-}
-
-static string Rtos(double d) {
-  if (std::isnan(d)) return "NaN";
-  char out[16];
-  sprintf(out, "%.5f", d);
-  char *o = out;
-  while (*o == '0') o++;
-  return string{o};
 }
 
 int main(int argc, char **argv) {
@@ -136,6 +118,7 @@ int main(int argc, char **argv) {
     citations_cdf.reserve(1 << 23);
     
     int64 articles = 0LL, citations = 0LL;
+    bool ahead = false;
     for (int i = 0; i < rows.size(); i++) {
       articles += rows[i].second.articles;
       citations += rows[i].second.citations;
@@ -148,6 +131,16 @@ int main(int argc, char **argv) {
 	  StringPrintf("%s,%s ", Rtos(x * XSCALE).c_str(), Rtos(ay * YSCALE).c_str());
 	citations_cdf +=
 	  StringPrintf("%s,%s ", Rtos(x * XSCALE).c_str(), Rtos(cy * YSCALE).c_str());
+
+	if (cy >= ay && !ahead) {
+	  printf("Crossover to cy >= ay (%.4f <= %.4f) at %d. %s\n",
+		 cy, ay, i, rows[i].first.c_str());
+	  ahead = true;
+	} else if (cy <= ay && ahead) {
+	  printf("Crossover to cy <= ay (%.4f <= %.4f) at %d. %s\n",
+		 cy, ay, i, rows[i].first.c_str());
+	  ahead = false;
+	}
       }
     }
 
@@ -158,17 +151,19 @@ int main(int argc, char **argv) {
     Util::WriteFile(outfile, svg);
     printf("Wrote %s\n", outfile.c_str());
   }
-  
-  string sorted_outfile = infile + "sorted.txt";
[…]

### 2018-03-11 citations/paper.tex
more analysis, writing

@@
 % Academic Advancement Advice: Adopt A A Authorship
+% Academic Advancement Advice: Author Articles as A A
 
-\title{Want to succeed in academia? Take the name A A}
+\title{Academic Advancement Advice: Author Articles as A.~A.}
 \author{A.~A. \\
@@
 
-\subsection{Classic Journey of Computer Science}
+\subsection{Relatable Computer Science Activities}
 
 The task is apparently simple: Parse 130 gigabytes of JSON data to
-extract the author names and citations from the papers.
+extract the author names and citations from the papers and count them.
+To begin my journey, I selected a C++ JSON library from 39
+alternatives listed on a benchmark page.\cite{yip2018native} The first
+criterion I ordered by was ``correctness,'' for which only two parsers
+achieved the highest correctness score, $100\%$. I selected the slower
+(i.e., more methodical) of the two parsers, since I did not want to do
+shoddy work.
+
+Next I spent a nice Saturday morning integrating this library into my
+private build environment, making sure its open source license was
+properly declared and so on, while I waited for the 130 gigabytes of
+JSON to download. Miraculously, this was straightforward. I was able
+to finish writing the code to process the articles before they even
+finished downloading. However, upon unzipping, I found that each file
+contained one million articles, not as a single JSON array but as a
+series of adjacent objects. The JSON library allowed for reading JSON
+from files, but {\em not} e.g. for reading a single record from a file pointer
+repeatedly. So, I needed to do my own ad hoc parsing before feeding
+each record to the JSON library as strings. OK. Now when I run it, it
+prints
+\begin{verbatim}
+  [Segmentation fault
+\end{verbatim}
+
+This is not my first segmentation fault---let me tell you!---so I knew
+what to do. I debugged the program, adding assertions using a
+\verb+CHECK+ macro that aborts if its condition is violated. I
+eventually determined that the bug must be inside the JSON library
+itself, since the segmentation fault happened between two
+\verb+CHECK+s where the only code being run was JSON library code. So,
+deciding that perhaps its $100\%$ correctness benchmark score was
+overstated, I downloaded and repeated the process with the other,
+faster (i.e., more slapdash) parser. Of course this parser has a
+different API so I needed to rewrite my program. (The program is very
+simple so this was the fastest step; however, getting the library
+properly incorporated and compiling and then learning its API was
+still a computer science chore.) Once my program was rid of the old
+crashy library, I ran it, and it produced this result
+\begin{verbatim}
+  [Segmentation fault
+\end{verbatim}
+Since these two libraries produced the same result, either
+segmentation violation must be part of the JSON specification, or the
+bug is instead elsewhere in the program. After reducing the program to
+a fairly simple test case, I determined that it must be some kind of
+``Heisenbug'' whereby an earlier piece of code was corrupting the heap
+or similar, causing a later crash only when present. This was not my
+first Heisenbug---let me tell you!---so I knew what to do. However,
+not being able to use Valgrind on Windows (did I mention that I'm
+using Cygwin and mingw64 on the eponymous Windows~7 by the way? Why am
+I doing that?), I booted Linux in a virtual machine, installed the 87
+security updates that had been released since I last booted this
+virtual machine a few months ago, and compiled my program there. I
+then needed to move one of the JSON files to the virtual computer to
+test, without having to download 130 GB from the internet again, but
+the file was apparently too large to transfer using the mysterious
+``Drag and drop to virtual machine'' functionality. I recalled that
+VirtualBox can mount a shared folder from the host machine if you edit
+the \verb+/etc/fstab+ file to contain the correct code words, and
+that I had performed this ritual before, but then had commented-out
+that line in ``rescue mode'' because I also learned that having this
+line present during the boot process causes a kernel panic. So I
+temporarily reinstated the line and mounted the drive and copied the
+gigantic file into the virtual machine and then removed the dangerous
+magic from the f-stab for f-safety.
+
+Next I ran Valgrind on this minimal test case and it worked as
+expected with no unclean behavior detected and no crashing.
+
+Since Linux was no help debugging, I returned to Windows~7 and tried
+single-step debugging. Here I discovered that the crash in my minimal
+test case was occurring when using \verb+cerr+. Specifically, the
+second thing sent to \verb+cerr+ would produce a segmentation fault. I
+then recalled that the \verb+CHECK+ macro uses \verb+cerr+ and now saw that
+the \verb+[+ in \verb+[Segmentation fault+ was probative, for this
+was the location of the original crash:
+\begin{Verbatim}[fontsize=\small]
+    std::cerr << "[" << pretty_date_.HumanDate() << "] "
+              << file << ":" << line << ": ";
+\end{Verbatim}
+
+Of course, inside \verb+HumanDate()+ I found a hasty patch I had
+installed in A.D.~2014 called ``make logging not spit garbage on
+mingw,'' suggesting that I had encountered and misdiagnosed this issue
+in the past. Although this led me on a minor herring (e.g. is
+\verb+__MINGW32__+ defined when cross-compiling for x86\_64?)
+HumanDate just returned \verb+"mingw"+, so this really was a problem
+with \verb+std::cerr+ crashing on the second thing printed with it
+(i.e., with the returned \verb+ostream+).
+
+Minimizing the test case, this bug only occurred after reading a large
+file all in one syscall (for ``performance'', even though doing these
+huge reads basically locks up my machine). Specifically the file was
+2,105,319,548 bytes, which is $98\%$ of $2^{31}$, suggesting the
+possibility of integer overflow in some buffer-sizing code in the
+weird Cygwin or mingw wrappers for \verb+fread+, \verb+fstat+, etc.
+This could plausibly cause corruption of the \verb+cerr+ ostream,
+which is probably part of the same code.
+
+Preparing to at least make a bug report if not try to fix it, I
+upgraded my compiler and runtime from GCC version 5.1.0 to 6.4.0,
+and the problem completely went away. Since Valgrind also agreed
+that there was no issue, I chalked this one up to simply an internal
+C runtime bug that someone else found and fixed in the last 2 years.
+Thanks!
+
+With \verb+cerr+ fixed, I could now easily see the original ``bug'',
+which was a violated assumption: I expected (and \verb+CHECK+ed),
+optimistically, that if the article had an {\tt author} field then the
+author would be a string. I accounted for a missing author, an author
+that was the empty string, but not {\tt author: null}. Thanks!
+
+Finally, I could run the author and citation tallier on all of the
+JSON files. However, two of the files appeared to be empty. Although
+this was only a small percentage of the articles, I did not want to
+skip any because it's possible that they are arranged in a non-random
+order (e.g. alphabetically!) and that this would bias the results. I
+learned that although the AMiner database is helpfully split into
+files each containing one million articles, for ease of handling, two
+of these files are nonetheless slightly larger than $2^{31}$ bytes. Such
+files cannot be cleanly sized with \verb+fstat+ (the \verb+st_size+
+field has type \verb+off_t+, a signed 32-bit type in POSIX) and even
+if you follow the standard advice of just treating it as unsigned
+anyway, \verb+fread+ will simply fail when asked to read such a large
+chunk\footnote{
+%
+  Curious why I am even using {\tt fstat} and {\tt fread}? I have my
+  own library routine for reading a file into a string. When working
+  with large files it's useful to avoid copying, because the contents
+  of the file may be a significant fraction of all RAM---copying is
+  not only expensive, but might temporarily exceed the available RAM.
+  So, you need to allocate the buffer ahead of time and avoid resizing
+  it (which often also performs a copy). It seems like something that
+  everyone would want to do, but it's ridiculously hard to do portably
+  in C or C++. Specifically, there seems to be no way to actually know
+  the size of a file so that you can size the buffer ahead of time.
+  For example, {\tt fseek} to {\tt SEEK\_END} and {\tt ftell} does
+  {\em not} work (it has {\em undefined behavior}). The {\tt st\_size}
+  field from {\tt stat} also does not give correct results e.g. when
+  reading from the {\tt /proc} filesystem. My routine attempts to
+  guess the size of the file and perform a single read in the case
+  that the guess, while still usually avoiding copying or needless
+  work in the case that it's an under- or over-estimate.
+%
+} despite the fact that it takes \verb+size_t+, an unsigned type. (?!)
+So, another afternoon spent doing Computer Science. But all you need
+to do of course is perform multiple sequential reads.
+
+Finally, the trivial task of counting the number of articles and
+citations per author could be completed.
+
+\section{The Data}
+
+This citation database is very noisy. About $1.6\%$ of articles have
+no authors, and another $1\%$ have a blank (or {\tt null}) author. I
+verified the comprehensiveness of the dataset by finding my own author
+record(s)\footnote{This paper is published under the pseudonym A.~A.,
+  standing for Awesome Author, obviously in order top increase its
+  citation rate. Normally I publish as Tom Murphy VII, which can be
+  found in citations with many broken variations, such as ``Vii, T.''}
+and 275 citations ({\bf Yeah!!}). I also spot-checked that the database
+includes prestigious conferences, which it does, at least including
+the ``article'' {\em A Record of the Proceedings of SIGBOVIK 2008},
+with author {\em Pennsylvania USA}. It was necessary to normalize
+author names to remove punctuation, weird unicode stuff like non-breaking
+spaces, javascript-encoded newlines and nul-bytes, and so on. There
+are many strange authors in the database, such as
+\begin{itemize}
+\item capinha para celular horizonte artificial iphone \quad(no citations)
+\item +0aaaaaabablaaaaaea\&\#xa \quad(no citations)
+\item a h poop \quad(4 citations)
+\item john mcm anus\footnote{Presumably a hostile citation of John McManus.} \quad(2 citations)
+\item coffee hour \quad(0 citations)
+\item nominate com registering dad domains since \quad(0 citations)
+\item a a \quad(195 citations)
+\item a a a a m islam \quad(6 citations)
+\end{itemize}
+
+On the other hand, from this hand-picked sample we can already see the
+citation advantage of having a name that's alphabetically early. Some
+authors seem to have already intuited this fact.
+
+\section{Analysis}
+
+Authors names appear in both ``Firstname Lastname'' order and
+``Lastname Firstname'' (when this grammar is even applicable).
+Therefore, I considered alphabetizing from front to back (by
+space-separated token) as well as back to front. This produces two
+different graphs. I also rejected a large number entries when I could
+not place the name in alphabetical order because its first character
+was not ASCII. Later non-ASCII charactes are OK, and are just sorted
+by their UTF-8 encodings (or whatever garbage is in the file). This of
+course produces a bias (excluding authors who write their names in
+e.g. Cyrillic and CJK scripts), although it's not clear how to assess
+the alphabetical hypothesis for them.
+
+To visualize the data, I produced a cumulative distribution function
+(CDF) for both articles and citations. Considering each author in
+alphabetical order, I keep a running total of how many articles and
+how many citations have been seen so far. The x-coordinate is the rank
+of the author (its index in the sorted list) and the y-coordinates are
+the fraction of articles (or citations) seen so far. The CDFs for the
+forward and backward token orders are in Figure~\ref{fig:authorcdf}.
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=0.44 \textwidth]{authorstats-forward}
+  \includegraphics[width=0.44 \textwidth]{authorstats-backward}
+  \end{center}
+  \caption{ Author CDFs for (a) forward token order and (b) backward
+    token order. (For example, (a) treats {\tt tom murphy vii} as
+    coming between {\tt tom marphy} and {\tt tom myrphie vxx} and (b)
+    treats it as coming between {\tt zzz murphx vii} and {\tt a
+      viii}.) The x-axis is the author's position alphabetically among
+    all authors. The blue lines are the fraction of all citations seen
+    so far, and the red lines the same for articles. The lines must
+    meet at $(0,0)$ and $(1,1)$ by definition. When the blue line is
+    above the red, the number of citations is outpacing the number of
+    articles for authors alphabetically before this point. Both show a
+    bias towards the early alphabet, with the forward direction being
+    more dramatic (for basically the entire range). The backward range
+    briefly shows the opposite bias (crossing over at author
+    \#5,400,000, ``ojars biskaps''), suggesting that the best names
+    may be alphabetically early but in the {\em B}s, not {\em A}s. The
+    poorer performance of A-names may be due to authors attempting to
+    ``game the system'' by publishing with names like ``a a.''
+  } \label{fig:authorcdf}
+\end{figure}
 
@@
 
- - When doing a
+- When doing a last-minute
 
-300 GB of json woes
+- Some academics try to \verb+fread+ all papers into memory, but
+  it fails on the file that's just slightly larger than $2^{31}$ bytes,
+  and so the papers therein don't get cited, and those papers are
+  alphabetically in the middle or end of the alphabet.
+
+\subsection{Alliterative Articles}
+
+Adopting an all-'A' affect also advances articles.
+Awesome!
 
@@
 
-capinha para celular horizonte artificial iphone
+u fucks
+a a fucks
+
+\bibliographystyle{IEEEtran}
+% \bibliographycomment{} % nothing
+\bibliography{paper}
 

### 2018-03-11 citations/bestworst.cc

@@ -1,0 +1,112 @@
+
+#ifdef __MINGW32__
+#include <windows.h>
+#undef ARRAYSIZE
+#endif
+
+#include "base/logging.h"
+#include "base/stringprintf.h"
+#include "util.h"
+#include "threadutil.h"
+#include "re2/re2.h"
+#include "citation-util.h"
+
+#include <cmath>
+#include <vector>
+#include <string>
+#include <map>
+#include <unordered_map>
+
+struct Stats {
+  int64 articles = 0;
+  int64 citations = 0;
+  // citations per article, [XXX but assuming one free article and one free citation?]
+  double citation_rate = 0.0;
+};
+
+int main(int argc, char **argv) {
+  #ifdef __MINGW32__
+  if (!SetPriorityClass(GetCurrentProcess(), BELOW_NORMAL_PRIORITY_CLASS)) {
+    LOG(FATAL) << "Unable to go to BELOW_NORMAL priority.\n";
+  }
+  #endif
+
+  CHECK_EQ(argc, 4) << "bestworst.exe infile.txt output-best.txt ouput-worst.txt\n";
+  string infile = argv[1];
+  string bestfile = argv[2];
+  string worstfile = argv[3];
+
+  std::unordered_map<string, Stats> cite_stats;
+  int64 articles_kept = 0LL, citations_kept = 0LL;
+  RE2 line_re{"([^\t]*)\t([\\d]+)\t([\\d]+)"};  
+  LocalForEachLine(infile,
+		   [&cite_stats, &line_re,
+		    &articles_kept, &citations_kept](string line) {
+    string author;
+    int64 articles = 0LL, citations = 0LL;
+    CHECK(RE2::FullMatch(line, line_re, &author, &articles, &citations)) << line;
+
+    Stats &stats = cite_stats[author];
+    stats.articles += articles;
+    stats.citations += citations;
+    articles_kept += articles;
+    citations_kept += citations;
+  });
+
+  for (auto &p : cite_stats) {
+    p.second.citation_rate = (double)(p.second.citations + 1LL) /
+      (double)(p.second.articles + 1LL);
+  }
+  
+  printf("Got stats for %lld keys.\n"
+	 "Total kept articles: %lld  and citations: %lld\n"
+	 "Average citations per paper: %.04f\n",
+	 (int64)cite_stats.size(), 
+	 articles_kept, citations_kept,
+	 (double)citations_kept / articles_kept);
+  
[…]

### 2018-03-12 citations/authorstats.cc
More in paper, tourney, figures

@@ -18,36 +18,6 @@
 #include <map>
 #include <unordered_map>
 
-static bool Dictionaryize(string author, string *dict) {
-  vector<string> tokens;
-  while (!author.empty()) {
-    string tok = Util::chop(author);
-    if (!tok.empty()) tokens.push_back(Util::lcase(tok));
-  }
-
-  // Totally whitespace name?
-  if (tokens.empty())
-    return false;
-
-  Reverse(&tokens);
-  
-  // Make sure it fits somewhere in alphabetical order.
-  if (tokens[0][0] >= 'a' && tokens[0][0] <= 'z') {
-    dict->clear();
-    for (const string &tok : tokens) {
-      if (dict->empty()) {
-	*dict = tok;
-      } else {
-	*dict += " ";
-	*dict += tok;
-      }
-    }
-    return true;
-  } else {
-    return false;
-  }
-}
-
 int main(int argc, char **argv) {
   #ifdef __MINGW32__
   if (!SetPriorityClass(GetCurrentProcess(), BELOW_NORMAL_PRIORITY_CLASS)) {
@@ -69,10 +39,11 @@ int main(int argc, char **argv) {
 		    &articles_kept, &citations_kept](string line) {
     string author;
     int64 articles = 0LL, citations = 0LL;
-    CHECK(RE2::FullMatch(line, line_re, &author, &articles, &citations)) << line;
+    CHECK(RE2::FullMatch(line, line_re, &author, &articles, &citations)) <<
+      line;
 
     string dict;
-    if (Dictionaryize(author, &dict)) {
+    if (Dictionaryize<true>(author, &dict)) {
       CiteStats &stats = cite_stats[dict];
       stats.articles += articles;
       stats.citations += citations;
@@ -84,7 +55,8 @@ int main(int argc, char **argv) {
     }
   });
 
-  printf("Got stats for %lld keys, with %lld rejected (%lld articles rejected)\n"
+  printf("Got stats for %lld keys, "
+	 "with %lld rejected (%lld articles rejected)\n"
 	 "Total kept articles: %lld  and citations: %lld\n",
 	 (int64)cite_stats.size(), authors_bad, articles_bad,
 	 articles_kept, citations_kept);
@@ -116,6 +88,9 @@ int main(int argc, char **argv) {
[…]

### 2018-03-12 trunk/citations/paper.tex
More in paper, tourney, figures

@@
 \usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}
+% \usepackage[utf8]{inputenc}
 
@@
 VirtualBox can mount a shared folder from the host machine if you edit
-the \verb+/etc/fstab+ file to contain the correct arcane magic, and
+the \verb+/etc/fstab+ file to contain the correct code words, and
 that I had performed this ritual before, but then had commented-out
@@
 test case was occurring when using \verb+cerr+. Specifically, the
-second thing sent to \verb+cerr+ would produce a segmentation fault.
-I then recalled that the \verb+CHECK+ macro uses \verb+cerr+ and
-also found code I had written in the Year 2014
+second thing sent to \verb+cerr+ would produce a segmentation fault. I
+then recalled that the \verb+CHECK+ macro uses \verb+cerr+ and now saw that
+the \verb+[+ in \verb+[Segmentation fault+ was probative, for this
+was the location of the original crash:
+\begin{Verbatim}[fontsize=\small]
+    std::cerr << "[" << pretty_date_.HumanDate() << "] "
+              << file << ":" << line << ": ";
+\end{Verbatim}
 
-The original ``bug'' was a violated assumption: I assumed,
-optimistically, that if the article had an {\tt author} field then
-the author would be a string. I accounted for a missing author,
-an author that was the empty string, but not {\tt author: null}.
+Of course, inside \verb+HumanDate()+ I found a hasty patch I had
+installed in A.D.~2014 called ``make logging not spit garbage on
+mingw,'' suggesting that I had encountered and misdiagnosed this issue
+in the past. Although this led me on a minor herring (e.g. is
+\verb+__MINGW32__+ defined when cross-compiling for x86\_64?)
+HumanDate just returned \verb+"mingw"+, so this really was a problem
+with \verb+std::cerr+ crashing on the second thing printed with it
+(i.e., with the returned \verb+ostream+).
 
-multiple hypothesized causes:
+Minimizing the test case, this bug only occurred after reading a large
+file all in one syscall (for ``performance'', even though doing these
+huge reads basically locks up my machine). Specifically the file was
+2,105,319,548 bytes, which is $98\%$ of $2^{31}$, suggesting the
+possibility of integer overflow in some buffer-sizing code in the
+weird Cygwin or mingw wrappers for \verb+fread+, \verb+fstat+, etc.
+This could plausibly cause corruption of the \verb+cerr+ ostream,
+which is probably part of the same code.
 
- - When doing a
+Preparing to at least make a bug report if not try to fix it, I
+upgraded my compiler and runtime from GCC version 5.1.0 to 6.4.0,
+and the problem completely went away. Since Valgrind also agreed
+that there was no issue, I chalked this one up to simply an internal
+C runtime bug that someone else found and fixed in the last 2 years.
+Thanks!
 
-300 GB of json woes
+With \verb+cerr+ fixed, I could now easily see the original ``bug'',
+which was a violated assumption: I expected (and \verb+CHECK+ed),
+optimistically, that if the article had an {\tt author} field then the
+author would be a string. I accounted for a missing author, an author
+that was the empty string, but not {\tt author: null}. Thanks!
+
+Finally, I could run the author and citation tallier on all of the
+JSON files. However, two of the files appeared to be empty. Although
+this was only a small percentage of the articles, I did not want to
+skip any because it's possible that they are arranged in a non-random
+order (e.g. alphabetically!) and that this would bias the results. I
+learned that although the AMiner database is helpfully split into
+files each containing one million articles, for ease of handling, two
+of these files are nonetheless slightly larger than $2^{31}$ bytes. Such
+files cannot be cleanly sized with \verb+fstat+ (the \verb+st_size+
+field has type \verb+off_t+, a signed 32-bit type in POSIX) and even
+if you follow the standard advice of just treating it as unsigned
+anyway, \verb+fread+ will simply fail when asked to read such a large
+chunk\footnote{
+%
+  Curious why I am even using {\tt fstat} and {\tt fread}? I have my
+  own library routine for reading a file into a string. When working
+  with large files it's useful to avoid copying, because the contents
+  of the file may be a significant fraction of all RAM---copying is
+  not only expensive, but might temporarily exceed the available RAM.
+  So, you need to allocate the buffer ahead of time and avoid resizing
+  it (which often also performs a copy). It seems like something that
+  everyone would want to do, but it's ridiculously hard to do portably
+  in C or C++. Specifically, there seems to be no way to actually know
+  the size of a file so that you can size the buffer ahead of time.
+  For example, {\tt fseek} to {\tt SEEK\_END} and {\tt ftell} does
+  {\em not} work (it has {\em undefined behavior}). The {\tt st\_size}
+  field from {\tt stat} also does not give correct results e.g. when
+  reading from the {\tt /proc} filesystem. My routine attempts to
+  guess the size of the file and perform a single read in the case
+  that the guess is correct, while still usually avoiding copying or
+  needless work in the case that it's an under- or over-estimate.
+%
+} despite the fact that it takes \verb+size_t+, an unsigned type. (?!)
+So, another afternoon spent doing Computer Science. But all you need
+to do of course is perform multiple sequential reads.
+
+Finally, the trivial task of counting the number of articles and
+citations per author could be completed.
+
+\section{The Data}
+
+This citation database is very noisy. About $1.6\%$ of articles have
+no authors, and another $1\%$ have a blank (or {\tt null}) author. I
+verified the comprehensiveness of the dataset by finding my own author
+record(s)\footnote{This paper is published under the pseudonym A.~A.,
+  standing for Awesome Author, obviously in order top increase its
+  citation rate. Normally I publish as Tom Murphy VII, which can be
+  found in citations with many broken variations, such as ``Vii, T.''}
+and 275 citations ({\bf Yeah!!}). I also spot-checked that the database
+includes prestigious conferences, which it does, at least including
+the ``article'' {\em A Record of the Proceedings of SIGBOVIK 2008},
+with author {\em Pennsylvania USA}. It was necessary to normalize
+author names to remove punctuation, weird unicode stuff like non-breaking
+spaces, javascript-encoded newlines and nul-bytes, and so on. There
+are many strange authors in the database, such as
+\begin{itemize}
+\item capinha para celular horizonte artificial iphone \quad(no citations)
+\item +0aaaaaabablaaaaaea\&\#xa \quad(no citations)
+\item a h poop \quad(4 citations)
+\item john mcm anus\footnote{Presumably a hostile citation of John McManus.} \quad(2 citations)
+\item coffee hour \quad(0 citations)
+\item nominate com registering dad domains since \quad(0 citations)
+\item a a \quad(195 citations)
+\item a a a a m islam \quad(6 citations)
+\end{itemize}
+
+On the other hand, from this hand-picked sample we can already see the
+citation advantage of having a name that's alphabetically early. Some
+authors seem to have already intuited this fact.
+
+\section{Analysis}
+
+Author names appear in both ``Firstname Lastname'' order and
+``Lastname Firstname'' (when this grammar is even applicable).
+Therefore, I analyzed alphabetizing from front to back (by
+space-separated token) as well as back to front. I also rejected a
+large number entries when I could not place the name in alphabetical
+order because its first character was not ASCII. Later non-ASCII
+characters are OK, and are just sorted by their UTF-8 encodings (or
+whatever garbage is in the file). This of course produces a bias
+(excluding authors who write their names in e.g. Cyrillic and CJK
+scripts), although it's not clear how to assess the alphabetical
+hypothesis for them.
+
+To visualize the data, I produced a cumulative distribution function
+(CDF) for both articles and citations. Considering each author in
+alphabetical order, I keep a running total of how many articles and
+how many citations have been seen so far. The x-coordinate is the rank
+of the author (its index in the sorted list) and the y-coordinates are
+the fraction of articles (or citations) seen so far. The CDFs for the
+forward and backward token orders are in Figure~\ref{fig:authorcdf}.
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=0.44 \textwidth]{authorstats-forward}
+  \includegraphics[width=0.44 \textwidth]{authorstats-backward}
+  \end{center}
+  \caption{ Author CDFs for (a) forward token order and (b) backward
+    token order. (For example, (a) treats {\tt tom murphy vii} as
+    coming between {\tt tom marphy} and {\tt tom myrphie vxx} and (b)
+    treats it as coming between {\tt zzz murphx vii} and {\tt a
+      viii}.) The x-axis is the author's position alphabetically among
+    all authors. The blue lines are the fraction of all citations seen
+    so far, and the red lines the same for articles. The lines must
+    meet at $(0,0)$ and $(1,1)$ by definition. When the blue line is
+    above the red, the number of citations is outpacing the number of
+    articles for authors alphabetically before this point. Both show a
+    bias towards the early alphabet, with the forward direction being
+    more dramatic (for basically the entire range). The backward range
+    briefly shows the opposite bias (crossing over at author
+    \#5,400,000, ``ojars biskaps''), suggesting that the best names
+    may be alphabetically early but in the {\em B}s, not {\em A}s. The
+    poorer performance of A-names may be due to authors attempting to
+    ``game the system'' by publishing with names like ``a a.''
+  } \label{fig:authorcdf}
+\end{figure}
+
+These CDFs show that---as expected!---there is a bias towards citing
+authors alphabetically early, whether we order alphabetically by
+``first name'' or ``last name.''
+
+There are multiple hypothesized causes:
+
+- When writing a last-minute ``related work'' section, authors seek
+  out work
+
+- Some academics try to \verb+fread+ all papers into memory, but
+  it fails on the file that's just slightly larger than $2^{31}$ bytes,
+  and so the papers therein don't get cited, and those papers are
+  alphabetically in the middle or end of the alphabet.
+
+% 
+% 313779.  書評	29582	1	0.000068
+% 313778.  インタビュー	20547	1	0.000097
+% 313777.  facharzt	85315	9	0.000117
+% 313776.  f3d	125257	15	0.000128
+% 313775.  recenzja	63568	8	0.000142
+% 313774.  newhaven	13296	1	0.000150
+% 313773.  secrétaire	12267	1	0.000163
+% 313772.  zahnarzt	17958	2	0.000167
+% 313771.  libguides	389727	72	0.000187
+% 313770.  硏究	12961	2	0.000231
+% 313769.  그	15984	3	0.000250
+% 313768.  kitas	7713	1	0.000259
+% 313767.  シンポジウム	18539	4	0.000270
+% 313766.  特集	878011	240	0.000274
+%
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=0.475 \textwidth]{wordstats}
+  \end{center}
+  \caption{ CDF for words arranged alphabetically. This is just as
+    in Figure~\ref{fig:authorcdf}, but since each entry is a single
+    word, there is no need to treat forward and backward differently.
+    Again, we see a small alphabetical bias, with citations leading
+    articles for most of the range. In fact, papers that include the
+    very first word, ``a'', are cited $1.3\times$ as often as the
+    average paper. Other common and advantageous words are visible
+    as clear bumps. (See Section XXX)
+  } \label{fig:wordcdf}
+\end{figure}
+
+% Tourney -- what is wrong here?
+%  1000.000 Mrounds:
+%  (69.66 vs. 72.69 Gcit) (6.72 vs. 7.20 Gart) (0.281 vs. 0.295 Gwin)
+%  (0.958361 L:R cit) (0.933736 L:R art) (0.280867 vs. 0.295327 win rate)
 
@@
 
-Adopting an all-'A' affect also advances articles.
+Adopting an all-'A' affect also advances articles. Although affairs
+are abbreviated, adequate abstracts abound: Astronomy, asthma,
+atmospheric acclimatization, autobiography, Azerbaijani accents,
+acceleration, and abundant alternatives.
+
+% arranged alphabetically.
+
 Awesome!
@@
 
-capinha para celular horizonte artificial iphone
+u fucks
+a a fucks
 

### 2018-03-12 trunk/citations/citation-util.h

@@ -120,3 +120,35 @@ inline static string Normalize(string w) {
   w = Util::Replace(std::move(w), "—", "-");
   return w;
 }
+
+template<bool backward_order>
+static bool Dictionaryize(string author, string *dict) {
+  vector<string> tokens;
+  while (!author.empty()) {
+    string tok = Util::chop(author);
+    if (!tok.empty()) tokens.push_back(Util::lcase(tok));
+  }
+
+  // Totally whitespace name?
+  if (tokens.empty())
+    return false;
+
+  if (backward_order)
+    Reverse(&tokens);
+  
+  // Make sure it fits somewhere in alphabetical order.
+  if (tokens[0][0] >= 'a' && tokens[0][0] <= 'z') {
+    dict->clear();
+    for (const string &tok : tokens) {
+      if (dict->empty()) {
+	*dict = tok;
+      } else {
+	*dict += " ";
+	*dict += tok;
+      }
+    }
+    return true;
+  } else {
+    return false;
+  }
+}

### 2018-03-12 trunk/citations/tourney.cc

@@ -1,0 +1,129 @@
+
+#ifdef __MINGW32__
+#include <windows.h>
+#undef ARRAYSIZE
+#endif
+
+#include "base/logging.h"
+#include "base/stringprintf.h"
+#include "util.h"
+#include "threadutil.h"
+#include "re2/re2.h"
+#include "textsvg.h"
+#include "citation-util.h"
+#include "randutil.h"
+#include "arcfour.h"
+
+#include <cmath>
+#include <vector>
+#include <string>
+#include <map>
+#include <unordered_map>
+
+int main(int argc, char **argv) {
+  #ifdef __MINGW32__
+  if (!SetPriorityClass(GetCurrentProcess(), BELOW_NORMAL_PRIORITY_CLASS)) {
+    LOG(FATAL) << "Unable to go to BELOW_NORMAL priority.\n";
+  }
+  #endif
+
+  CHECK_EQ(argc, 2) << "tourney.exe infile.txt\n";
+  string infile = argv[1];
+
+  // Number of articles authored by each name.
+  std::unordered_map<string, CiteStats> cite_stats;
+  int64 authors_bad = 0LL, articles_bad = 0LL;
+  int64 articles_kept = 0LL, citations_kept = 0LL;
+  RE2 line_re{"([^\t]*)\t([\\d]+)\t([\\d]+)"};  
+  LocalForEachLine(infile,
+		   [&cite_stats, &line_re, &authors_bad, &articles_bad,
+		    &articles_kept, &citations_kept](string line) {
+    string author;
+    int64 articles = 0LL, citations = 0LL;
+    CHECK(RE2::FullMatch(line, line_re, &author, &articles, &citations)) <<
+      line;
+
+    string dict;
+    if (Dictionaryize<false>(author, &dict)) {
+      CiteStats &stats = cite_stats[dict];
+      stats.articles += articles;
+      stats.citations += citations;
+      articles_kept += articles;
+      citations_kept += citations;
+    } else {
+      authors_bad++;
+      articles_bad += articles;
+    }
+  });
+
+  printf("Got stats for %lld keys, "
+	 "with %lld rejected (%lld articles rejected)\n"
+	 "Total kept articles: %lld  and citations: %lld\n",
+	 (int64)cite_stats.size(), authors_bad, articles_bad,
+	 articles_kept, citations_kept);
+  
+  vector<std::pair<string, CiteStats>> rows;
+  rows.reserve(cite_stats.size());
[…]

### 2018-03-14 trunk/citations/paper.tex
Compute worst papers

@@
 There are multiple hypothesized causes:
+\begin{itemize}
 
-- When writing a last-minute ``related work'' section, authors seek
-  out work
+\item When writing a last-minute ``related work'' section, authors
+  sometimes find papers to cite in alphabetical order, for example, by
+  reading another paper's bibliography, which is sometimes
+  alphabetized.
 
-- Some academics try to \verb+fread+ all papers into memory, but
+\item Some academics try to \verb+fread+ all papers into memory, but
   it fails on the file that's just slightly larger than $2^{31}$ bytes,
@@
 
-% 
+\item When applying for grad school, tenure track faculty jobs, grants,
+  and so on, packets are alphabetized ``for fairness,'' and the
+  disposition of the committee becomes more irritable as they make
+  progress through the set.
+
+\end{itemize}
+
 % 313779.  書評	29582	1	0.000068

### 2018-03-14 citations/titles.cc
Compute worst papers

@@ -1,0 +1,104 @@
+
+#ifdef __MINGW32__
+#include <windows.h>
+#undef ARRAYSIZE
+#endif
+
+#include "rapidjson/document.h"
+#include "base/logging.h"
+#include "base/stringprintf.h"
+#include "util.h"
+#include "citation-util.h"
+
+#include <vector>
+#include <string>
+#include <map>
+#include <unordered_map>
+#include <unordered_set>
+
+using namespace rapidjson;
+using namespace std;
+
+int main(int argc, char **argv) {
+  #ifdef __MINGW32__
+  if (!SetPriorityClass(GetCurrentProcess(), BELOW_NORMAL_PRIORITY_CLASS)) {
+    LOG(FATAL) << "Unable to go to BELOW_NORMAL priority.\n";
+  }
+  #endif
+
+  vector<string> filenames;
+  string outfile;
+  for (int i = 1; i < argc; i++) {
+    string arg = (string)argv[i];
+    if (arg == "-o") {
+      CHECK(i < argc - 1);
+      outfile = argv[i + 1];
+      i++;
+    } else {
+      filenames.push_back(arg);
+    }
+  }
+
+  if (outfile.empty()) {
+    outfile = filenames[0] + ".titles";
+  }
+  
+  if (filenames.empty()) {
+    fprintf(stderr, "Pass JSON paper files on the command line.\n");
+    return -1;
+  }
+  
+  int64 counter = 0LL, num_output = 0LL, not_ascii = 0LL;
+
+  
+  FILE *out = fopen(outfile.c_str(), "wb");
+
+  for (const string &filename : filenames) {
+    LocalForEachLine(filename,
+		     [&counter, &num_output, &not_ascii,
+		      out](string j) {
+      counter++;
+      Document article;
+      CHECK(!article.Parse(j.c_str()).HasParseError());
+      CHECK(article.IsObject());
+
+      // Have to have a title or there's no way to count it.
+      CHECK(article.HasMember("id")) << j;
+      const Value &id_value = article["id"];
+      CHECK(id_value.IsString()) << j;
+      const string id = id_value.GetString();
+      
+      if (article.HasMember("title")) {
+	int64 n_citation = 0LL;
+	if (article.HasMember("n_citation") &&
+	    article["n_citation"].IsInt()) {
+	  n_citation = article["n_citation"].GetInt();
+	}
+
+	// Papers with citations can't be the least-citable.
[…]

### 2018-03-14 trunk/citations/titles.cc
grep thing, restrict to english titles

@@ -25,7 +25,7 @@ int main(int argc, char **argv) {
     LOG(FATAL) << "Unable to go to BELOW_NORMAL priority.\n";
   }
   #endif
-
+  
   vector<string> filenames;
   string outfile;
   for (int i = 1; i < argc; i++) {
@@ -48,14 +48,15 @@ int main(int argc, char **argv) {
     return -1;
   }
   
-  int64 counter = 0LL, num_output = 0LL, not_ascii = 0LL;
-
+  int64 counter = 0LL, num_output = 0LL, not_ascii = 0LL, not_en = 0LL;
+  int64 has_citations = 0LL;
   
   FILE *out = fopen(outfile.c_str(), "wb");
 
   for (const string &filename : filenames) {
     LocalForEachLine(filename,
-		     [&counter, &num_output, &not_ascii,
+		     [&counter, &has_citations,
+		      &num_output, &not_ascii, &not_en,
 		      out](string j) {
       counter++;
       Document article;
@@ -76,9 +77,23 @@ int main(int argc, char **argv) {
 	}
 
 	// Papers with citations can't be the least-citable.
-	if (n_citation > 0)
+	if (n_citation > 0) {
+	  has_citations++;
 	  return;
+	}
 
+	// Require them to be in English.
+	if (!article.HasMember("lang")) {
+	  not_en++;
+	  return;
+	}
+	const Value &lang = article["lang"];
+	if (!lang.IsString() ||
+	    (string)lang.GetString() != "en") {
+	  not_en++;
+	  return;
+	}
+	
 	const Value &title = article["title"];
 	if (title.IsString()) {
 	  string words = LightNormalization(title.GetString());
@@ -93,8 +108,9 @@ int main(int argc, char **argv) {
       }
     });
 
-    printf("%lld articles. %lld not ascii. %lld kept\n",
-	   counter, not_ascii, num_output);
+    printf("%lld articles. %lld has citations.\n"
+	   "%lld not ascii. %lld not en. %lld kept\n",
+	   counter, has_citations, not_ascii, not_en, num_output);
   }
   
   printf("Wrote %s\n", outfile.c_str());

### 2018-03-14 citations/paper.tex
grep thing, restrict to english titles

@@
   guess the size of the file and perform a single read in the case
-  that the guess, while still usually avoiding copying or needless
-  work in the case that it's an under- or over-estimate.
+  that the guess is correct, while still usually avoiding copying or
+  needless work in the case that it's an under- or over-estimate.
 %
@@
 
-Authors names appear in both ``Firstname Lastname'' order and
+Author names appear in both ``Firstname Lastname'' order and
 ``Lastname Firstname'' (when this grammar is even applicable).
-Therefore, I considered alphabetizing from front to back (by
-space-separated token) as well as back to front. This produces two
-different graphs. I also rejected a large number entries when I could
-not place the name in alphabetical order because its first character
-was not ASCII. Later non-ASCII charactes are OK, and are just sorted
-by their UTF-8 encodings (or whatever garbage is in the file). This of
-course produces a bias (excluding authors who write their names in
-e.g. Cyrillic and CJK scripts), although it's not clear how to assess
-the alphabetical hypothesis for them.
+Therefore, I analyzed alphabetizing from front to back (by
+space-separated token) as well as back to front. I also rejected a
+large number entries when I could not place the name in alphabetical
+order because its first character was not ASCII. Later non-ASCII
+characters are OK, and are just sorted by their UTF-8 encodings (or
+whatever garbage is in the file). This of course produces a bias
+(excluding authors who write their names in e.g. Cyrillic and CJK
+scripts), although it's not clear how to assess the alphabetical
+hypothesis for them.
 
@@
 
-multiple hypothesized causes:
+These CDFs show that---as expected!---there is a bias towards citing
+authors alphabetically early, whether we order alphabetically by
+``first name'' or ``last name.''
 
-- When doing a last-minute
+There are multiple hypothesized causes:
+\begin{itemize}
 
-- Some academics try to \verb+fread+ all papers into memory, but
+\item When writing a last-minute ``related work'' section, authors
+  sometimes find papers to cite in alphabetical order, for example, by
+  reading another paper's bibliography, which is sometimes
+  alphabetized.
+
+\item Some academics try to \verb+fread+ all papers into memory, but
   it fails on the file that's just slightly larger than $2^{31}$ bytes,
@@
 
+\item When applying for grad school, tenure track faculty jobs, grants,
+  and so on, packets are alphabetized ``for fairness,'' and the
+  disposition of the committee becomes more irritable as they make
+  progress through the set.
+
+\item Decreased access to pie and  unfair situations in
+  youth cause a lifelong predisposition to failure.
+
+\end{itemize}
+
+% 313779.  書評	29582	1	0.000068
+% 313778.  インタビュー	20547	1	0.000097
+% 313777.  facharzt	85315	9	0.000117
+% 313776.  f3d	125257	15	0.000128
+% 313775.  recenzja	63568	8	0.000142
+% 313774.  newhaven	13296	1	0.000150
+% 313773.  secrétaire	12267	1	0.000163
+% 313772.  zahnarzt	17958	2	0.000167
+% 313771.  libguides	389727	72	0.000187
+% 313770.  硏究	12961	2	0.000231
+% 313769.  그	15984	3	0.000250
+% 313768.  kitas	7713	1	0.000259
+% 313767.  シンポジウム	18539	4	0.000270
+% 313766.  特集	878011	240	0.000274
+%
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=0.475 \textwidth]{wordstats}
+  \end{center}
+  \caption{ CDF for words arranged alphabetically. This is just as
+    in Figure~\ref{fig:authorcdf}, but since each entry is a single
+    word, there is no need to treat forward and backward differently.
+    Again, we see a small alphabetical bias, with citations leading
+    articles for most of the range. In fact, papers that include the
+    very first word, ``a'', are cited $1.3\times$ as often as the
+    average paper. Other common and advantageous words are visible
+    as clear bumps. (See Section XXX)
+  } \label{fig:wordcdf}
+\end{figure}
+
+% Tourney -- what is wrong here?
+%  1000.000 Mrounds:
+%  (69.66 vs. 72.69 Gcit) (6.72 vs. 7.20 Gart) (0.281 vs. 0.295 Gwin)
+%  (0.958361 L:R cit) (0.933736 L:R art) (0.280867 vs. 0.295327 win rate)
+
 \subsection{Alliterative Articles}
 
-Adopting an all-'A' affect also advances articles.
+Adopting an all-'A' affect also advances articles. Although affairs
+are abbreviated, adequate abstracts abound: Astronomy, asthma,
+atmospheric acclimatization, autobiography, Azerbaijani accents,
+acceleration, and abundant alternatives.
+
+% arranged alphabetically.
+
 Awesome!

### 2018-03-14 citations/worstpapers.cc

@@ -67,6 +67,7 @@ int main(int argc, char **argv) {
 
   static constexpr int64 MIN_ARTICLES = 100;
   static constexpr int64 MIN_CITATIONS = 1;
+  static constexpr int64 MAX_TITLE_WORDS = 20;
   
   // Now, probabilities for all words.
   printf("Reading words...");
@@ -113,14 +114,21 @@ int main(int argc, char **argv) {
       [&titles, &word_stats](int title_idx) {
 	string title = titles[title_idx].title;
 	double prob = 1.0;
+	int words = 0;
 	while (!title.empty()) {
 	  string token = Normalize(Util::chop(title));
+	  words++;
 	  auto it = word_stats.find(token);
 	  if (it == word_stats.end())
 	    continue;
 	  prob *= it->second.citation_multiplier;
 	}
-	titles[title_idx].cite_probability = prob;
+	if (words <= MAX_TITLE_WORDS) {
+	  titles[title_idx].cite_probability = prob;
+	} else {
+	  // This will cause it to be deleted below.
+	  titles[title_idx].cite_probability = -1.0;
+	}
       },
       12);
   printf("Scored titles.\n");
@@ -134,13 +142,21 @@ int main(int argc, char **argv) {
 
   printf("Sorted.\n");
 
+  int64 too_long = 0;
+  while (!titles.empty() && titles.back().cite_probability < 0.0) {
+    titles.pop_back();
+    too_long++;
+  }
+
+  printf("Dropped %lld titles that were too long.\n", too_long);
+  
   // Now write output files.
   {
     FILE *out = fopen(bestfile.c_str(), "wb");
     CHECK(out != nullptr) << bestfile.c_str();
     for (int i = 0; i < 5000 && i < titles.size(); i++) {
       const Title &title = titles[i];
-      fprintf(out, "%d.  %.9g\t%s\t%s\n",
+      fprintf(out, "%d.  %.17g\t%s\t%s\n",
 	      i,
 	      title.cite_probability,
 	      title.id.c_str(), title.title.c_str());
@@ -154,7 +170,7 @@ int main(int argc, char **argv) {
     CHECK(out != nullptr) << worstfile.c_str();
     for (int i = titles.size() - 1; i >= 0 && i >= titles.size() - 5000; i--) {
       const Title &title = titles[i];
-      fprintf(out, "%d.  %.9g\t%s\t%s\n",
+      fprintf(out, "%d.  %.17g\t%s\t%s\n",
[…]

### 2018-03-15 trunk/citations/paper.tex
oy, almost there...

@@
 memo funded by SIGBOVIK's political opponents. The warrant was for the
-AMiner Open Academic Graph, which consists of summary information for
-154,771,162 papers in about 130 gigabytes of JSON data. Then I embarked
-on one of the Classic Journeys of Computer Science.
+AMiner Open Academic Graph~\cite{Tang08ArnetMiner}, which consists of
+summary information for 154,771,162 papers in about 130 gigabytes of
+JSON data. Then I performed computer science for longer than it should
+have taken.
 
@@
 
+This subsection contains a relatable tale of computer science. It can
+be safely skipped, like all content in this paper.
+
+\smallskip
 The task is apparently simple: Parse 130 gigabytes of JSON data to
@@
 properly declared and so on, while I waited for the 130 gigabytes of
-JSON to download. Miraculously, this was straightforward. I was able
-to finish writing the code to process the articles before they even
-finished downloading. However, upon unzipping, I found that each file
-contained one million articles, not as a single JSON array but as a
-series of adjacent objects. The JSON library allowed for reading JSON
-from files, but {\em not} e.g. for reading a single record from a file pointer
-repeatedly. So, I needed to do my own ad hoc parsing before feeding
-each record to the JSON library as strings. OK. Now when I run it, it
-prints
+JSON to download.\footnote{I assure the reader that I have the most
+  premium optical-fiber-based Internet. However, the Microsoft Content
+  Distribution Network throttled these downloads mightily, presumably
+  because of Net Neutrality.} Miraculously, this was straightforward.
+I was able to finish writing the code to process the articles before
+they even finished downloading. However, upon unzipping, I found that
+each file contained one million articles, not as a single JSON array
+but as a series of adjacent objects. AdJSONt objects. The JSON library
+allowed for reading JSON from files, but {\em not} e.g. for reading a
+single record from a file pointer repeatedly. So, I needed to do my
+own ad hoc parsing before feeding each record to the JSON library as
+strings. OK. Now when I run it, it prints
 \begin{verbatim}
@@
 simple so this was the fastest step; however, getting the library
-properly incorporated and compiling and then learning its API was
-still a computer science chore.) Once my program was rid of the old
-crashy library, I ran it, and it produced this result
+properly incorporated and compiling and then learning its API still
+required me to apply fundamental principles from my advanced computer
+science degree.) Once my program was rid of the old crashy library, I
+ran it, and it produced this result
 \begin{verbatim}
@@
 I doing that?), I booted Linux in a virtual machine, installed the 87
-security updates that had been released since I last booted this
-virtual machine a few months ago, and compiled my program there. I
-then needed to move one of the JSON files to the virtual computer to
+critical security updates that had been released since I last booted
+this virtual machine a few months ago, and compiled my program there.
+I then needed to move one of the JSON files to the virtual computer to
 test, without having to download 130 GB from the internet again, but
@@
 VirtualBox can mount a shared folder from the host machine if you edit
-the \verb+/etc/fstab+ file to contain the correct code words, and
-that I had performed this ritual before, but then had commented-out
-that line in ``rescue mode'' because I also learned that having this
-line present during the boot process causes a kernel panic. So I
+the \verb+/etc/fstab+ file to contain the correct code words, and that
+I had performed this ritual before, but then had commented-out that
+line in ``rescue mode'' because I also learned that having this line
+present during the boot process causes a kernel panic. So I
 temporarily reinstated the line and mounted the drive and copied the
@@
 %
-  Curious why I am even using {\tt fstat} and {\tt fread}? I have my
-  own library routine for reading a file into a string. When working
-  with large files it's useful to avoid copying, because the contents
-  of the file may be a significant fraction of all RAM---copying is
-  not only expensive, but might temporarily exceed the available RAM.
-  So, you need to allocate the buffer ahead of time and avoid resizing
-  it (which often also performs a copy). It seems like something that
+  Curious why I am even using {\tt fstat} and {\tt fread} like some
+  dramatically bearded guy from the 1980s? I have my own library
+  routine for reading a file into a string. When working with large
+  files it's useful to avoid copying, because the contents of the file
+  may be a significant fraction of all RAM---copying is not only
+  expensive, but might temporarily exceed the available RAM. So, you
+  need to allocate the buffer ahead of time and avoid resizing it
+  (which often also performs a copy). It seems like something that
   everyone would want to do, but it's ridiculously hard to do portably
@@
 record(s)\footnote{This paper is published under the pseudonym A.~A.,
-  standing for Awesome Author, obviously in order top increase its
+  standing for Awesome Author, obviously in order to increase its
   citation rate. Normally I publish as Tom Murphy VII, which can be
@@
 
-\section{Analysis}
+\section{Author Analysis} \label{sec:authoranalysis}
 
@@
 how many citations have been seen so far. The x-coordinate is the rank
-of the author (its index in the sorted list) and the y-coordinates are
-the fraction of articles (or citations) seen so far. The CDFs for the
-forward and backward token orders are in Figure~\ref{fig:authorcdf}.
+of the author (its index in the alphabetical list) and the
+y-coordinates are the fraction of articles (or citations) seen so far.
+The CDFs for the forward and backward token orders are in
+Figure~\ref{fig:authorcdf}.
 
-\begin{figure}
+\begin{figure*}[t]
   \begin{center}
@@
     above the red, the number of citations is outpacing the number of
-    articles for authors alphabetically before this point. Both show a
-    bias towards the early alphabet, with the forward direction being
-    more dramatic (for basically the entire range). The backward range
-    briefly shows the opposite bias (crossing over at author
-    \#5,400,000, ``ojars biskaps''), suggesting that the best names
-    may be alphabetically early but in the {\em B}s, not {\em A}s. The
-    poorer performance of A-names may be due to authors attempting to
-    ``game the system'' by publishing with names like ``a a.''
-  } \label{fig:authorcdf}
-\end{figure}
+    articles for authors alphabetically before this point. Since you
+    may be reading in old timey blacke \& whyte, I will also tell you
+    that the blue line is more or less above the red line in both
+    pictures. This means that both show a bias towards the early
+    alphabet, with the forward direction being more dramatic (for
+    basically the entire range). The backward range briefly shows the
+    opposite bias (crossing over at author \#5,400,000, ``ojars
+    biskaps''), suggesting that the best names may be alphabetically
+    early but in the {\em B}s, not {\em A}s. The poorer performance of
+    A-names may be due to authors attempting to ``game the system'' by
+    publishing with names like ``a a.'' } \label{fig:authorcdf}
+\end{figure*}
 
@@
 
+\item Sometimes the last page of a manuscript is ``lost,'' truncating
+  its alphabetical bibliography to some prefix.
+
 \item Some academics try to \verb+fread+ all papers into memory, but
@@
   and so on, packets are alphabetized ``for fairness,'' and the
-  disposition of the committee becomes more irritable as they make
+  mood of the committee becomes more irritable as they make
   progress through the set.
+
+\item Decreased access to pie and exposure to unfair situations in
+  youth cause a lifelong predisposition to failure.
 
@@
 
-% 313779.  書評	29582	1	0.000068
-% 313778.  インタビュー	20547	1	0.000097
-% 313777.  facharzt	85315	9	0.000117
-% 313776.  f3d	125257	15	0.000128
-% 313775.  recenzja	63568	8	0.000142
-% 313774.  newhaven	13296	1	0.000150
-% 313773.  secrétaire	12267	1	0.000163
-% 313772.  zahnarzt	17958	2	0.000167
-% 313771.  libguides	389727	72	0.000187
-% 313770.  硏究	12961	2	0.000231
-% 313769.  그	15984	3	0.000250
-% 313768.  kitas	7713	1	0.000259
-% 313767.  シンポジウム	18539	4	0.000270
-% 313766.  特集	878011	240	0.000274
-%
+\section{I now bore of the titular topic. Let us discuss new topics.}
+
+Specifically, let's discuss titles. Some have suggested that the
+titles of papers also affect their citation rates, and it is
+reasonable to suspect that an alphabetical bias could apply here as
+well. After applying some light normalization, I extracted each of the
+words that appear in the titles of all papers, counting both the
+articles (an article is only counted once per word, even if that word
+appears multiple times in its title) and citations. This tells us
+what the expected number of citations for an article is (assuming
+independence, which is---as usual---completely false) when its
+title contains a word. Considering only words that appeared in
+at least 100 articles, here are the top by expectation:
+
+\begin{tabular}{l|l|l|l}
+{\bf word} & {\bf articles} & {\bf citations} & {\bf E = c/a} \\
+  \hline
+folin & 172 & 209,642 & 1,211.8 \\
+power/knowledge & 111 & 55,992 & 499.9 \\
+\{s\} & 101 & 38,241 & 374.9 \\
+1972-1977 & 171 & 54,917 & 319.2 \\
+\textbackslash sqrt\{s\} & 126 & 33,253 & 261.8 \\
+eigenfaces & 176 & 45,284 & 255.8 \\
+america''s & 119 & 30,478 & 253.9 \\
+gromacs & 125 & 30,865 & 244.9 \\
+neo-ffi & 102 & 21,992 & 213.5 \\
+esh & 156 & 33,142 & 211.1 \\
+\end{tabular}
+
+These are mostly explained by being relatively rare words that appear
+in extremely popular papers. For example, {\em folin} comes from an
+1951 paper\cite{lowry1951protein} describing a procedure for using
+this reagent, which I guess everyone who uses this technique cites
+(Google scholar has this paper at 215,329 citations). {\em
+  Power/knowledge} and {\em 1972-1977} both come from the title of a
+Foucault book\cite{foucault1980power} with 31,590 citations.
+
+Considering just the alphabetizable words, these too show a bias
+towards early words (Figure~\ref{fig:wordcdf}). Bibliographies that
+are alphabetized by title could exhibit the same effects discussed in
+Section~\ref{sec:authoranalysis}. Additionally, since some scholars
+learn English by reading the dictionary front to back, they may
+simply be more practiced at early words and find them more attractive
+or easier to understand.
+
+\medskip
+It is a bit easier to interpret the expected citation numbers if we
+normalize them. The average citation rate of all articles in this set
+is 5.1261. This may seem high, especially considering the 90 million
+articles with no citations, but keep in mind that most papers have
+bibliographies in which they cite several others. Data quality issues
+aside, this is basically the same as saying that the averge
+bibliography length is 5.1261. We can assign each word a ``citation
+multiplier'' effect now, defined as simply
+$$
+\mathrm{multiplier}_w = \frac{\mathrm{citations}_w / \mathrm{articles}_w}{5.1261}
+$$
+and now numbers greater than 1 improve your chances at citation, and
+numbers less than 1 diminish it. Here are the most common words:
+
+\begin{tabular}{l|l|l|l}
+\,  & artices & citations & multiplier \\
+  \hline
+of & 64322278 & 366756237 & 1.207822 \\
+and & 42576889 & 276526184 & 1.375780 \\
+in & 39685539 & 230534759 & 1.230526 \\
+the & 38873614 & 216491707 & 1.179704 \\
+a   & 22666824 & 139244700 & 1.301292 \\
+for & 19048050 & 114197875 & 1.269972 \\
+on & 15194275 & 68904896 & 0.960631 \\
+with & 10825547 & 60752792 & 1.188784 \\
+to & 10202749 & 58590695 & 1.216461 \\
+de & 7753205 & 3283786 & 0.089718 \\
+by & 7259598 & 42234470 & 1.232370 \\
+from & 5389957 & 35999433 & 1.414806 \\
+an & 5368923 & 33088995 & 1.305519 \\
+la & 4417389 & 1471023 & 0.070541 \\
+study & 4370395 & 22022172 & 1.067397 \\
+\end{tabular}
+
+Several of these words are visible in Figure~\ref{fig:wordcdf}, in
+fact; they are so common that they cause large jumps in both article
+and citation count. Curiously, most of the common words have a
+multiplier greater than 1; they {\em increase} the expected number of
+citations. {\em of} sounds smart (e.g. {\em Of Mice And Men}), {\em
+  and} makes sense (paper contains at least two results). On the other
+hand, {\em on} actually reduces the citation count, probably because
+it implies equivocation (e.g. {\em On the other hand}). Dramatically,
+common non-English words have very low multipliers; Spanish words {\em
+  de} and {\em la} reduce citations by a factor of more than ten. This
+may be due to problems in the data set, but it is certainly easy to
+imagine real cultural bias!
+
+Finally, here are the words with the worst multipliers:
+\begin{tabular}{l|l|l|l}
+313779.  \includegraphics[width=2em]{bookreview} & 29582 & 1 & 0.000068 \\
+313778.  \includegraphics[width=5em]{interview} & 20547 & 1 & 0.000097 \\
+313777.  facharzt & 85315 & 9 & 0.000117 \\
+313776.  f3d & 125257 & 15 & 0.000128 \\
+313775.  recenzja & 63568 & 8 & 0.000142 \\
+313774.  newhaven & 13296 & 1 & 0.000150 \\
+313773.  secr\'etaire & 12267 & 1 & 0.000163 \\
+313772.  zahnarzt & 17958 & 2 & 0.000167 \\
+313771.  libguides & 389727 & 72 & 0.000187 \\
+313770.  \includegraphics[width=2em]{research} & 12961 & 2 & 0.000231 \\
+313769.  \includegraphics[width=1em]{that} & 15984 & 3 & 0.000250 \\
+313768.  kitas & 7713 & 1 & 0.000259 \\
+% 313767.  シンポジウム & 18539 & 4 & 0.000270 \\
+% 313766.  特集 & 878011 & 240 & 0.000274 \\
+\end{tabular}
+
+For this set, I only considered words that appeared in the title of at
+least one paper that was actually cited. The worst multiplier is
+\includegraphics[width=2em]{bookreview}, which is Chinese for ``book
+review;'' it reduces your chances of citation by more than ten
+thousandfold. Close by is \includegraphics[width=5em]{interview},
+Japanese for ``interview.'' {\em Facharzt} is like (medical)
+``specialist'' in German, {\em recenzja} is ``review'' in Polish, and
+{\em newhaven} is a misspelling of a city in Connecticut where I grew
+up and was unfairly and repeatedly placed in the middle of lines. But
+most of these are just generic non-English words
+(\includegraphics[width=2em]{research} is Chinese for ``research'' and
+\includegraphics[width=1em]{that} Korean for ``that'').
 
@@
     average paper. Other common and advantageous words are visible
-    as clear bumps. (See Section XXX)
+    as clear bumps.
   } \label{fig:wordcdf}
 \end{figure}
+
+\section{What is the least citable paper?}
+
+Thinking about first and last, one could reasonably ask: Of all
+papers, what are the least citable ones? Of course, the paper should
+have never actually been cited, but in the database there at least 90
+million articles with no citations, all tied for this distinction. Can
+we rank among these papers to break the tie?
+
+(Of course I am just poking fun, not seriously declaring these
+articles somehow ``worst.'' Many good articles are never cited. For
+example, as far as I know, the paper you are reading right now has no
+citations.)
+
+One way to do this is to try to estimate the number of citations that
+a paper will get based on the words in its title. Since we already
+collected the words that are most citable, this is an apparently
+simple affair. We can estimate the citation rate by just taking the
+product of the multipliers for the words. We treat words for which
+we don't have data (because they didn't meet thresholds) as having
+multiplier 1.
+
+This first cut has some problems. As an illustration, the article
+with the highest estimated citation rate (among those with zero actual
+citations) is a paper\cite{leigland2009grid} called
+\begin{quote}
+Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid Grid
+\end{quote}
+which is expected to have 5,262,751,498,750,000,000,000,000 citations.
+This is because the word {\em grid} has a favorable multiplier and is
+repeated 121 times.
+
+So I limited it to articles with no more than 20 words in the title
+(arbitrary). Now the least citable articles all look like this:
+
+\begin{quote}
+  IMPLEMENTASI PERATURAN MAHKAMAH AGUNG NOMOR 1 TAHUN 2008 TENTANG
+  PROSEDUR MEDIASI DI PENGADILAN (STUDI PUTUSAN MEDIASI DI PENGADILAN
+  NEGERI SAMARINDA)\footnote{Translated: IMPLEMENTATION OF REGULATION
+    OF THE SUPREME COURTS NUMBER 1 OF 2008 ABOUT MEDIATION
+    PROCEDURES IN THE COURT (STUDY OF DECISION MEDIATION IN THE COURT
+    SAMARINDA COUNTRY)}
+\end{quote}
+
+Which appears to be a real dissertation\cite{putri2009implementasi}
+(bachelor's degree of law) from a student in Indonesia, but there are
+literally thousands of similar looking papers. Given its company, it
+may very well be truly the least citable paper, or it may be a systematic
+data problem, or even spam. But I was hoping for some English language
+articles so that we could appreciate the joke together in the native
+tongue.
+
+Even filtering for articles with \verb+"lang"+ explicitly set to
+\verb+"en"+, I needed to manually create a blacklist of hundreds of
+non-English words that commonly appeared in these article titles.
+After a few rounds of blacklisting, some ``English'' articles started
+to appear:
+
+% 78371002.  5.3718855409784041e-027	53e9a089b7602d970296e0eb	Office Hours: Monday, Wednesday, Thursday and Friday 9:00 a.m. to 5:00 p.m.; Tuesday 9:00 a.m. to 6:45 p.m
+
+% 78370998.  6.049184011159267e-027	56d81919dabfae2eee86f82a	Thomas Aquinas: Theologian By Thomas F. O'Meara, O.P. Notre Dame, University of Notre Dame Press, 1997. 302 pp. $16.95
+
+The first is probably not the title of an article, although we can
+agree that it is unlikely to be cited. The second is a bad entry, with
+the entire citation packed into the title field. This book actually
+does have 94 citations (or DOES IT?~\cite{o1997thomas}) on Google
+Scholar and an Amazon Sales Rank of \#10,837 in {\em Christian
+  Denominations \& Sects}. In order to prevent such data problems from
+affecting our results, I then added a requirement that the article
+have basic metadata (either a Digital DOI Identifier field or {\tt venue})
+for it to be considered.
+
+% 48654527.
+5.8084026066881388e-026	56d918ecdabfae2eee6c6965	Eight contemporary poets : Charles Tomlinson, Donald Davie, R.S.
+Thomas, Philip Larkin, Ted Hughes, Thomas Kinsella, Stevie Smith, W.S.
+Graham
+
+% 48654111.  1.7552310525171227e-019	53e99952b7602d9702190ac1	The 80th Birthday of Sir Henry Dale, O.M., G.B.E., M.D., F.R.C.P., F.R.S.: Salute to Henry Hallett Dale
+
+% 48654054.  2.7492425189773547e-019	56d8ecc5dabfae2eee5b30f5	Narratives Unsettled: Digression in Robert Walser, Thomas Bernhard, and Adalbert Stifter by Samuel Frederick (review)
+
+% 48552586.  5.0644175742872403e-019	56d9266ddabfae2eeebdd19a	Catherine Mowry LaCugna's Trinitarian Theology: III- The Ecumenical Implications of Catherine Mowry Lacugna's Trinitarian Theology
+
+% 48552476.  1.7011818225031437e-018	56d85cd0dabfae2eee5cb5f4	Texts Illustrating the Causality of the Sacraments from William of Melitona, Assisi Bibl. Comm. 182, and Brussels Bibl. Royale 1542
+
+% 48552391.  3.4094058998675013e-018	53e9bce1b7602d9704965e64	Shakespeare's Titus Andronicus and Bandello's Novelle as Sources for the Munera Episode in Spenser's Faerie Queene, Book 5, Canto 2
+
+1. memorials for dead guys
+2. reviews of books
+3. articles not in english
+4. broken citations
+
+\largeskip
+Of course, since these papers are all notable for being so extremely
+uncitable, the true least citable papers are the next least citable,
+according to this analysis.
 

### 2018-03-15 trunk/citations/paper.tex
More work on paper, clean up titles

@@
 \newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}
+
+\def\signed#1{{\unskip\nobreak\hfil\penalty50
+    \hskip2em\hbox{}\nobreak\hfil\sl#1
+    \parfillskip=0pt \finalhyphendemerits=0 \par}}
 
@@
 
-% 78371002.  5.3718855409784041e-027	53e9a089b7602d970296e0eb	Office Hours: Monday, Wednesday, Thursday and Friday 9:00 a.m. to 5:00 p.m.; Tuesday 9:00 a.m. to 6:45 p.m
+\begin{quote}
+ Office Hours: Monday, Wednesday, Thursday and Friday 9:00 a.m. to 5:00 p.m.; Tuesday 9:00 a.m. to 6:45 p.m \\
+  \signed{(citation multiplier $5.37\!\times\!10^{-27}$)}
+\end{quote}
 
-% 78370998.  6.049184011159267e-027	56d81919dabfae2eee86f82a	Thomas Aquinas: Theologian By Thomas F. O'Meara, O.P. Notre Dame, University of Notre Dame Press, 1997. 302 pp. $16.95
+\begin{quote}
+	Thomas Aquinas: Theologian By Thomas F. O'Meara, O.P. Notre Dame, University of Notre Dame Press, 1997. 302 pp. \$16.95 \\
+  \signed{(citation multiplier $6.05\!\times\!10^{-27}$)}
+\end{quote}
 
-The first is probably not the title of an article, although we can
-agree that it is unlikely to be cited. The second is a bad entry, with
-the entire citation packed into the title field. This book actually
-does have 94 citations (or DOES IT?~\cite{o1997thomas}) on Google
-Scholar and an Amazon Sales Rank of \#10,837 in {\em Christian
-  Denominations \& Sects}. In order to prevent such data problems from
-affecting our results, I then added a requirement that the article
-have basic metadata (either a Digital DOI Identifier field or {\tt venue})
-for it to be considered.
+The first is probably not the title of an article\cite{officehours},
+although we can agree that it is unlikely to be cited. The second is
+probably a bad entry, with the entire citation packed into the title
+field. This book actually does have 94 citations (or DOES
+IT?~\cite{o1997thomas}) on Google Scholar and an Amazon Sales Rank of
+\#10,837 in {\em Christian Denominations \& Sects}.\footnote{ On the
+  other hand, the entry does have a different
+  author\cite{stevenson2000thomas}, so it could possibly be a review
+  of this book whose title is literally the string above, including
+  the price?} In order to prevent such data problems from affecting
+our results, I then added a requirement that the article have basic
+metadata (either a Digital DOI Identifier field or {\tt venue}) for it
+to be considered. There are still many broken entries, but picking through
+the results I'm able to find what appear to be actual articles:
 
-% 48654527.
-5.8084026066881388e-026	56d918ecdabfae2eee6c6965	Eight contemporary poets : Charles Tomlinson, Donald Davie, R.S.
-Thomas, Philip Larkin, Ted Hughes, Thomas Kinsella, Stevie Smith, W.S.
-Graham
+\begin{quote}
+  % 48654527.  5.8084026066881388e-026	56d918ecdabfae2eee6c6965
+  Eight contemporary poets : Charles Tomlinson, Donald Davie, R.S. Thomas, Philip Larkin, Ted Hughes, Thomas Kinsella, Stevie Smith, W.S. Graham
+  \signed{(citation multiplier $5.8\!\times\!10^{-26}$)}
+\end{quote}
 
@@
 
+Lessons learned:
+The least-cited works take the following typical forms. Acade
 1. memorials for dead guys
@@
 
-\largeskip
+\medskip
 Of course, since these papers are all notable for being so extremely
@@
 atmospheric acclimatization, autobiography, Azerbaijani accents,
-acceleration, and abundant alternatives.
+acceleration, and abundant alternatives. After achieving academic
+accomplishments, acknowledge A.A.'s {\it Academic Advancement Advice:
+  Author Articles as A.~A}!
 
 % arranged alphabetically.
-
-Awesome!
 
@@
 
+\begin{verbatim}  
+{"id": "56d8981adabfae2eee20314b", "title": "IMPLEMENTASI PERATURAN MAHKAMAH AGUNG NOMOR 1 TAHUN 2008 TENTANG PROSEDUR MEDIASI DI PENGADILAN (STUDI PUTUSAN MEDIASI DI PENGADILAN NEGERI SAMARINDA)", "authors": [{"name": "frianur ivan zairani lisi insan tajali nur"}], "year": 2013, "lang": "en"}
+{"id": "56d8a690dabfae2eee9057b0", "title": "Pengaruh Pengawasan Inspektorat Dan Partisipasi Dalam Penyusunan Anggaran Terhadap Kinerja Manajerial SKPD (Studi Empiris Pada Instansi Pemerintah Kabupaten Tanah Datar)", "authors": [{"name": "retno nabila sari"}], "year": 2015, "lang": "en"}
+{"id": "56d8ecc5dabfae2eee5b30f5", "title": "Narratives Unsettled: Digression in Robert Walser, Thomas Bernhard, and Adalbert Stifter by Samuel Frederick (review)", "authors": [{"name": "tove holmes"}], "venue": "Monatshefte", "year": 2015, "lang": "en"}
+{"id": "56d918ecdabfae2eee6c6965", "title": "Eight contemporary poets : Charles Tomlinson, Donald Davie, R.S.
+Thomas, Philip Larkin, Ted Hughes, Thomas Kinsella, Stevie Smith, W.S.
+Graham", "authors": [{"name": "calvin bedient"}], "venue": "Modern Language Review", "year": 1975, "lang": "en"}
+{"id": "56d9266ddabfae2eeebdd19a", "title": "Catherine Mowry LaCugna's Trinitarian Theology: III- The Ecumenical Implications of Catherine Mowry Lacugna's Trinitarian Theology", "authors": [{"name": "alan j torrance", "org": "university of st andrews"}], "venue": "Horizons", "year": 2000, "lang": "en", "doi": "10.1017/S036096690003262X", "url": ["http://dx.doi.org/10.1017/S036096690003262X"]}
+{"id": "56d85cd0dabfae2eee5cb5f4", "title": "Texts Illustrating the Causality of the Sacraments from William of Melitona, Assisi Bibl.
+Comm.
+182, and Brussels Bibl.
+Royale 1542", "authors": [{"name": "f o f m kilian lynch"}], "year": 1957, "lang": "en", "doi": "10.1353/frc.1957.0003", "url": ["http://dx.doi.org/10.1353/frc.1957.0003"]}
+\end{verbatim}
+
 \bibliographystyle{IEEEtran}

### 2018-03-15 trunk/citations/paper.tex
Needs proofreading, but maybe paper is done?

--- 2018-03-15 trunk/citations/paper.tex
@@
 
-\section*{Abstract}
-proof at last
-    1\ \dirham
+% \section*{Abstract}
 
@@
 authors whose names come alphabetically earlier have higher citation
-rates.
+rates. We also study some related subjects like, ``What is the most
+uncitable paper?''
 
@@
 
-\subsection{Relatable Computer Science Activities}
+\subsection{Relatable Computer Science Tale}
 
@@
 
+First I checked for overfull hbox, but this was not the problem.
 Of course, inside \verb+HumanDate()+ I found a hasty patch I had
@@
 author would be a string. I accounted for a missing author, an author
-that was the empty string, but not {\tt author: null}. Thanks!
+that was the empty string, but not \verb+author: null+. Thanks!
 
@@
 These are mostly explained by being relatively rare words that appear
-in extremely popular papers. For example, {\em folin} comes from an
+in extremely popular papers. For example, {\em folin} comes from a
 1951 paper\cite{lowry1951protein} describing a procedure for using
@@
 \verb+"en"+, I needed to manually create a blacklist of hundreds of
-non-English words that commonly appeared in these article titles.
-After a few rounds of blacklisting, some ``English'' articles started
-to appear:
+non-English words (mostly Indonesian and Polish) that commonly
+appeared in these article titles. After a few rounds of blacklisting,
+some ``English'' articles started to appear:
 
-% 78371002.  5.3718855409784041e-027	53e9a089b7602d970296e0eb	Office Hours: Monday, Wednesday, Thursday and Friday 9:00 a.m. to 5:00 p.m.; Tuesday 9:00 a.m. to 6:45 p.m
+\begin{quote}
+ Office Hours: Monday, Wednesday, Thursday and Friday 9:00 a.m. to 5:00 p.m.; Tuesday 9:00 a.m. to 6:45 p.m \\
+  \signed{(citation multiplier $5.37\!\times\!10^{-27}$)}
+\end{quote}
 
-% 78370998.  6.049184011159267e-027	56d81919dabfae2eee86f82a	Thomas Aquinas: Theologian By Thomas F. O'Meara, O.P. Notre Dame, University of Notre Dame Press, 1997. 302 pp. $16.95
+\begin{quote}
+	Thomas Aquinas: Theologian By Thomas F. O'Meara, O.P. Notre Dame, University of Notre Dame Press, 1997. 302 pp. \$16.95 \\
+  \signed{(citation multiplier $6.05\!\times\!10^{-27}$)}
+\end{quote}
 
-The first is probably not the title of an article, although we can
-agree that it is unlikely to be cited. The second is a bad entry, with
-the entire citation packed into the title field. This book actually
-does have 94 citations (or DOES IT?~\cite{o1997thomas}) on Google
-Scholar and an Amazon Sales Rank of \#10,837 in {\em Christian
-  Denominations \& Sects}. In order to prevent such data problems from
-affecting our results, I then added a requirement that the article
-have basic metadata (either a Digital DOI Identifier field or {\tt venue})
-for it to be considered.
+The first is probably not the title of an article\cite{officehours},
+although we can agree that it is unlikely to be cited. The second is
+probably a bad entry, with the entire citation packed into the title
+field. This book~\cite{o1997thomas} actually does have 94 citations
+(or DOES IT?) on Google Scholar and an Amazon Sales Rank of \#10,837
+in {\em Christian Denominations \& Sects}.\footnote{ On the other
+  hand, the entry does have a different
+  author\cite{stevenson2000thomas}, so it could possibly be a review
+  of this book whose title is literally the string above, including
+  the price?} In order to prevent such data problems from affecting
+our results, I then added a requirement that the article have basic
+metadata (either a Digital DOI Identifier field or {\tt venue}) for it
+to be considered. There vast majority still appear to be broken
+entries, but picking through the results I'm finally able to find what
+appear to be actual articles:
 
-% 48654527.
-5.8084026066881388e-026	56d918ecdabfae2eee6c6965	Eight contemporary poets : Charles Tomlinson, Donald Davie, R.S.
-Thomas, Philip Larkin, Ted Hughes, Thomas Kinsella, Stevie Smith, W.S.
-Graham
+\begin{quote}
+  % 48654527.  5.8084026066881388e-026	56d918ecdabfae2eee6c6965
+  Eight contemporary poets : Charles Tomlinson, Donald Davie, R.S. Thomas, Philip Larkin, Ted Hughes, Thomas Kinsella, Stevie Smith, W.S. Graham
+  \signed{(citation multiplier $5.81\!\times\!10^{-26}$)}
+\end{quote}
 
-% 48654111.  1.7552310525171227e-019	53e99952b7602d9702190ac1	The 80th Birthday of Sir Henry Dale, O.M., G.B.E., M.D., F.R.C.P., F.R.S.: Salute to Henry Hallett Dale
+\begin{quote}
+  % 48654111.  1.7552310525171227e-019	53e99952b7602d9702190ac1
+  The 80th Birthday of Sir Henry Dale, O.M., G.B.E., M.D., F.R.C.P., F.R.S.: Salute to Henry Hallett Dale
+  \signed{(citation multiplier $1.75\!\times\!10^{-19}$)}
+\end{quote}
 
-% 48654054.  2.7492425189773547e-019	56d8ecc5dabfae2eee5b30f5	Narratives Unsettled: Digression in Robert Walser, Thomas Bernhard, and Adalbert Stifter by Samuel Frederick (review)
+\begin{quote}
+  % 48654054.  2.7492425189773547e-019	56d8ecc5dabfae2eee5b30f5
+  Narratives Unsettled: Digression in Robert Walser, Thomas Bernhard, and Adalbert Stifter by Samuel Frederick (review)
+  \signed{(citation multiplier $2.74\!\times\!10^{-19}$)}
+\end{quote}
 
-% 48552586.  5.0644175742872403e-019	56d9266ddabfae2eeebdd19a	Catherine Mowry LaCugna's Trinitarian Theology: III- The Ecumenical Implications of Catherine Mowry Lacugna's Trinitarian Theology
+\begin{quote}
+  % 48552586.  5.0644175742872403e-019	56d9266ddabfae2eeebdd19a
+  Catherine Mowry LaCugna's Trinitarian Theology: III- The Ecumenical Implications of Catherine Mowry Lacugna's Trinitarian Theology
+  \signed{(citation multiplier $5.06\!\times\!10^{-19}$)}
+\end{quote}
 
-% 48552476.  1.7011818225031437e-018	56d85cd0dabfae2eee5cb5f4	Texts Illustrating the Causality of the Sacraments from William of Melitona, Assisi Bibl. Comm. 182, and Brussels Bibl. Royale 1542
+\begin{quote}
+  % 48552476.  1.7011818225031437e-018	56d85cd0dabfae2eee5cb5f4
+  Texts Illustrating the Causality of the Sacraments from William of Melitona, Assisi Bibl. Comm. 182, and Brussels Bibl. Royale 1542
+  \signed{(citation multiplier $1.70\!\times\!10^{-18}$)}
+\end{quote}
 
-% 48552391.  3.4094058998675013e-018	53e9bce1b7602d9704965e64	Shakespeare's Titus Andronicus and Bandello's Novelle as Sources for the Munera Episode in Spenser's Faerie Queene, Book 5, Canto 2
+\begin{quote}
+  % 48552391.  3.4094058998675013e-018	53e9bce1b7602d9704965e64
+  Shakespeare's Titus Andronicus and Bandello's Novelle as Sources for the Munera Episode in Spenser's Faerie Queene, Book 5, Canto 2
+  \signed{(citation multiplier $3.41\!\times\!10^{-18}$)}
+\end{quote}
 
-1. memorials for dead guys
-2. reviews of books
-3. articles not in english
-4. broken citations
+The first is definitely a real article~\cite{bedient1974eight}---which
+apparently has 95 citations on Google Scholar---so the estimate failed
+us here. It's easy to see why it has such a bad score, though; it's
+packed with people's names and initials, just like the scores of
+uncited book reviews and other broken citations. The second is also a
+real tribute to this Nobel prize winner~\cite{loewi195580th} with 5
+citations that Google knows of. Again it contains names and
+appelations that look like initials.
 
-\largeskip
+The third is a book review~\cite{holmes2015narratives} of a
+book~\cite{frederick2012narratives} with itself only 11 citations
+(makes sense, as the title contains three people's names) and indeed
+has no citations that Google knows of, so it is a solid contender.
+After that some theological
+stuff~\cite{torrance2000catherine,lynch1957texts} each with 3
+citations on Google, and another literary
+analysis~\cite{fitzpatrick2005shakespeare} which has 2.
+
+Lessons learned:
+\begin{itemize}
+\item Taking a noisy database of tens of millions of entries and then
+  trying to dig through the bottom of the barrel for something
+  ``funny'' or even ``interesting'' is tough going. Maybe this should
+  have been obvious.
+\item The least-cited works take on some recurring forms:
+  \begin{itemize}
+  \item Memorials for dead or old guys with lots of honorifics
+  \item Articles with people's names in the title
+  \item Reviews of books, or books about other literary work, or reviews of books about other literary work
+  \item Articles about God stuff or God People
+  \item Articles not in English
+  \item Broken citations
+  \item Combinations of the above.
+  \end{itemize}
+  Enterprising academics would do well to avoid such areas of research, in addition
+  to using alphabetically early words and author names in their publications.
+
+\item As one downside, this approach does not really work, given
+  that several of these articles have a few actual citations upon
+  comparing to another database.
+\end{itemize}
+
+\medskip
 Of course, since these papers are all notable for being so extremely
-uncitable, the true least citable papers are the next least citable,
-according to this analysis.
+uncitable, the true least citable papers are the {\bf next least
+  citable}, according to this analysis. To avoid paradoxes, the
+reader is discouraged from trying to reproduce these results.
 
@@
 atmospheric acclimatization, autobiography, Azerbaijani accents,
-acceleration, and abundant alternatives.
+acceleration, and abundant alternatives. After achieving academic
+accomplishments, acknowledge A.A.'s {\it Academic Advancement Advice:
+  Author Articles as A.~A}!
 
-% arranged alphabetically.
-
-Awesome!
-
-strange authors:
-
-u fucks
-a a fucks
+\section{Bibliography}
 
 \bibliographystyle{IEEEtran}
-% \bibliographycomment{} % nothing
 \bibliography{paper}

### 2018-04-01 trunk/citations/paper.tex
Submitted version

@@
 \usepackage{textcomp}
+
+\usepackage{nopageno}
 
@@
 esh & 156 & 33,142 & 211.1 \\
+eqs & 128 & 25,504 & 197.7 \\
+charmm & 175 & 31,579 & 179.4 \\
+position-specific & 378 & 67,250 & 177.4 \\
+kegg & 201 & 32,989 & 163.3 \\
+arlequin & 161 & 26,318 & 162.4 \\
+praat & 158 & 25,217 & 158.6 \\
 \end{tabular}
@@
   Power/knowledge} and {\em 1972-1977} both come from the title of a
-Foucault book\cite{foucault1980power} with 31,590 citations.
+Foucault book\cite{foucault1980power} with 31,590 citations. {\tt
+  america''s} is not a LaTeX disaster; it really has two apostrophes
+in it, and it is weird that this typo appears in 119 articles.
 
@@
   \hline
-of & 64322278 & 366756237 & 1.207822 \\
-and & 42576889 & 276526184 & 1.375780 \\
-in & 39685539 & 230534759 & 1.230526 \\
-the & 38873614 & 216491707 & 1.179704 \\
-a   & 22666824 & 139244700 & 1.301292 \\
-for & 19048050 & 114197875 & 1.269972 \\
-on & 15194275 & 68904896 & 0.960631 \\
-with & 10825547 & 60752792 & 1.188784 \\
-to & 10202749 & 58590695 & 1.216461 \\
-de & 7753205 & 3283786 & 0.089718 \\
-by & 7259598 & 42234470 & 1.232370 \\
-from & 5389957 & 35999433 & 1.414806 \\
-an & 5368923 & 33088995 & 1.305519 \\
-la & 4417389 & 1471023 & 0.070541 \\
-study & 4370395 & 22022172 & 1.067397 \\
+1. of & 64,322,278 & 366,756,237 & 1.207822 \\
+2. and & 42,576,889 & 276,526,184 & 1.375780 \\
+3. in & 39,685,539 & 230,534,759 & 1.230526 \\
+4. the & 38,873,614 & 216,491,707 & 1.179704 \\
+5. a   & 22,666,824 & 139,244,700 & 1.301292 \\
+6. for & 19,048,050 & 114,197,875 & 1.269972 \\
+7. on & 15,194,275 & 68,904,896 & 0.960631 \\
+8. with & 10,825,547 & 60,752,792 & 1.188784 \\
+9. to & 10,202,749 & 58,590,695 & 1.216461 \\
+10. de & 7,753,205 & 3,283,786 & 0.089718 \\
+11. by & 7,259,598 & 42,234,470 & 1.232370 \\
+12. from & 5,389,957 & 35,999,433 & 1.414806 \\
+13. an & 5,368,923 & 33,088,995 & 1.305519 \\
+14. la & 4,417,389 & 1,471,023 & 0.070541 \\
+15. study & 4,370,395 & 22,022,172 & 1.067397 \\
 \end{tabular}
@@
 multiplier greater than 1; they {\em increase} the expected number of
-citations. {\em of} sounds smart (e.g. {\em Of Mice And Men}), {\em
+citations. {\em Of} sounds smart (e.g. {\em Of Mice And Men}), {\em
   and} makes sense (paper contains at least two results). On the other
@@
 \begin{tabular}{l|l|l|l}
-313779.  \includegraphics[width=2em]{bookreview} & 29582 & 1 & 0.000068 \\
-313778.  \includegraphics[width=5em]{interview} & 20547 & 1 & 0.000097 \\
-313777.  facharzt & 85315 & 9 & 0.000117 \\
-313776.  f3d & 125257 & 15 & 0.000128 \\
-313775.  recenzja & 63568 & 8 & 0.000142 \\
-313774.  newhaven & 13296 & 1 & 0.000150 \\
-313773.  secr\'etaire & 12267 & 1 & 0.000163 \\
-313772.  zahnarzt & 17958 & 2 & 0.000167 \\
-313771.  libguides & 389727 & 72 & 0.000187 \\
-313770.  \includegraphics[width=2em]{research} & 12961 & 2 & 0.000231 \\
-313769.  \includegraphics[width=1em]{that} & 15984 & 3 & 0.000250 \\
-313768.  kitas & 7713 & 1 & 0.000259 \\
+313779.  \includegraphics[width=2em]{bookreview} & 29,582 & 1 & 0.000068 \\
+313778.  \includegraphics[width=5em]{interview} & 20,547 & 1 & 0.000097 \\
+313777.  facharzt & 85,315 & 9 & 0.000117 \\
+313776.  f3d & 125,257 & 15 & 0.000128 \\
+313775.  recenzja & 63,568 & 8 & 0.000142 \\
+313774.  newhaven & 13,296 & 1 & 0.000150 \\
+313773.  secr\'etaire & 12,267 & 1 & 0.000163 \\
+313772.  zahnarzt & 17,958 & 2 & 0.000167 \\
+313771.  libguides & 389,727 & 72 & 0.000187 \\
+313770.  \includegraphics[width=2em]{research} & 12,961 & 2 & 0.000231 \\
+313769.  \includegraphics[width=1em]{that} & 15,984 & 3 & 0.000250 \\
+313768.  kitas & 7,713 & 1 & 0.000259 \\
 % 313767.  シンポジウム & 18539 & 4 & 0.000270 \\
@@
 
-(Of course I am just poking fun, not seriously declaring these
-articles somehow ``worst.'' Many good articles are never cited. For
-example, as far as I know, the paper you are reading right now has no
-citations.)
+(Of course: In case someone who wrote one of these papers finds
+themselves here, please note that I am just poking fun, not seriously
+declaring these articles somehow ``worst.'' Many good articles are
+never cited. For example, as far as I know, the paper you are reading
+right now has no citations.)
 
@@
 a paper will get based on the words in its title. Since we already
-collected the words that are most citable, this is an apparently
+collected citation multipliers for all words, this is an apparently
 simple affair. We can estimate the citation rate by just taking the
-product of the multipliers for the words. We treat words for which
-we don't have data (because they didn't meet thresholds) as having
+product of the multipliers for the words. We treat words for which we
+don't have data (because they didn't meet thresholds) as having
 multiplier 1.
@@
 \end{quote}
-which is expected to have 5,262,751,498,750,000,000,000,000 citations.
+which according to the analysis is expected to have 5,262,751,498,750,000,000,000,000 citations.
 This is because the word {\em grid} has a favorable multiplier and is
@@
 probably a bad entry, with the entire citation packed into the title
-field. This book~\cite{o1997thomas} actually does have 94 citations
-(or DOES IT?) on Google Scholar and an Amazon Sales Rank of \#10,837
+field. This book actually does have 94 citations
+(or DOES IT?~\cite{o1997thomas}) on Google Scholar and an Amazon Sales Rank of \#10,837
 in {\em Christian Denominations \& Sects}.\footnote{ On the other
@@
 
-The first is definitely a real article~\cite{bedient1974eight}---which
+I don't know about you but I fell asleep just reading the titles. The
+first is definitely a real article~\cite{bedient1974eight}---which
 apparently has 95 citations on Google Scholar---so the estimate failed

### 2018-11-03 chess/papers/survival.tex
Tested on hundreds of millions of games so I think it's likely correct now, at least for the sake of accepting legal moves. Add skeleton of paper with my guesses prior to getting any results. Some performance improvements. Finally, track fates of pieces as necessary for the computation.

\begin{document}

Survival in Chessland

In formal chess, the king can never be captured: The game ends when
the king is attacked but cannot move, and it is illegal to make a move
that leaves the king attacked. The king's death is implied, of course,
but it is seen as more poetic to end the game prior to this point.

For the sake of this question, we'll consider the the white king to
``die'' if white loses (i.e., is checkmated), and likewise for black.
Otherwise, of course, the best chances of survival would trivially be
with the two kings, since they always survive. Loss includes
resignation, since most high-level games actually end once the
defeated player agrees that loss is inevitable. We can think of this
common case like king seppuku. Many games also end in time forfeit,
which is like the king's poor diet and lifestyle choices leading to
a death by natural causes.

Neither side has a decisive advantage, and many game end in a draw,
with both kings surviving. So the survival chances of a king are
clearly greater than 50\%---pretty decent odds. Is it possible that
any other piece has even better chances?

Here are my guesses:

\begin{itemize}
\item Black and white are probably not significantly different. That
  is, the a1 and a8 rooks probably have about the same survival
  chances (it's known that white has a slight statistical advantage
  but it is probably only around 1\%). So these guesses will be
  written about white's pieces.
\item The d2 and e2 pawns are very active in common openings, and
  are frequently captured as part of those openings. I think they
  are the least likely to survive overall.
\item Bishops and knights are often involved in the opening and
  midgame, and often exchanged nonchalantly. I think they all have
  relatively low survival chances.
\item Although the queen is very valuable, a queen exchange is often
  forced for games that enter the endgame.
\item Rooks tend to be late-game pieces, because they are difficult
  to get out of their corners (and at most one can be activated
  by the fastest method, castling) and are relatively valuable.
\item This leaves the non-central pawns. These are the hardest to
  predict, and they are hard to think about (at least for me) because
  when e.g. the a2 pawn recaptures the b3 pawn that it supported, I
  just think of this as the b3 pawn. Of these pawns, b2 and g2 are
  somewhat weak because they are undefended once the bishop is
  developed (cf. the famous ``poison pawn'' at b2). On the other hand,
  in the fianchetto configuration, this pawn is very strong and often
  survives the entire game without leaving the third rank. Since pawn
  chains usually progress towards the middle of the board, the a2 pawn
  is more likely to be supporting than supported. This both leaves it
  weak to capture, but prone to recapturing. Outside pawns block one's
  own rook, although for this same reason they often clear the file by
  capturing (and so survive). They are also commonly used to push into
  a well-defended king's territory (e.g. in the fianchetto); kingside
  castling is more common, so this means that the h pawns are often
  lost to this fate.
\end{itemize}

This leaves my final ranking, from least surviving to most surviving,
as:

16. d2,
15. e2,
14. b1 knight,
13. g1 knight,
12. c1 bishop,
11. f1 bishop,
10. queen,
9. king,
8. a1 rook,
7. h1 rook,
6. b2,
5. h2
4. a2
3. g2
2. c2
1. f2

% Ben, David, Jim also wagered guesses

\end{document}

### 2018-11-03 chess/rungames.cc

@@ -12,11 +12,19 @@
 #include "base/logging.h"
 #include "pgn.h"
 #include "util.h"
+#include "city.h"
 
 using namespace std;
 using int64 = int64_t;
+using uint64 = uint64_t;
 
 using Move = Position::Move;
+
+// static constexpr int64 MAX_GAMES = 1000000;
+static constexpr int64 MAX_GAMES = 0LL;
+
+// #define SELF_CHECK true
+#undef SELF_CHECK
 
 // TODO: To threadutil, but note that this is C++17.
 struct ReadMutexLock {
@@ -32,9 +40,44 @@ struct WriteMutexLock {
   std::shared_mutex *m;
 };
 
-template<class W, class F>
+static constexpr const char *const PIECE_NAME[32] = {
+  "a8 rook",
+  "b8 knight",
+  "c8 bishop",
+  "d8 queen",
+  "e8 king",
+  "f8 bishop",
+  "g8 knight",
+  "h8 rook",
+  "a7 pawn", "b7 pawn", "c7 pawn", "d7 pawn",
+  "e7 pawn", "f7 pawn", "g7 pawn", "h7 pawn",
+  // white
+  "a2 pawn", "b2 pawn", "c2 pawn", "d2 pawn",
+  "e2 pawn", "f2 pawn", "g2 pawn", "h2 pawn",
+  "a1 rook",
+  "b1 knight",
+  "c1 bishop",
+  "d1 queen",
+  "e1 king",
+  "f1 bishop",
+  "g1 knight",
+  "h1 rook", };
+
+// TODO: This does not parallelize enough. Only one of the numa nodes
+// succeeds in getting significant work scheduled. (Maybe actually
+// what happens is the N threads are all scheduled onto the numa node
+// that's closest to the data?) Would be pretty simple to have K
+// distinct deques and round-robin assignment of work to them. When
+// one is empty we could only then start work stealing.
+//
+// An alternate simple experiment would be to grab larger batches
+// of work when lots is available. Like if there are N threads, always
+// be willing to claim close to (1/N) * pending work units.
+template<class W, class F, int64 max_chunk>
 struct WorkQueue {
-  WorkQueue(F f, int num_workers) : f(std::move(f)) {
+  static_assert(max_chunk > 0LL, "Must take positive chunks!");
+  WorkQueue(F f, int num_workers) : f(std::move(f)),
[…]

### 2018-11-03 trunk/chess/pgn.h
Tested on hundreds of millions of games so I think it's likely correct now, at least for the sake of accepting legal moves. Add skeleton of paper with my guesses prior to getting any results. Some performance improvements. Finally, track fates of pieces as necessary for the computation.

@@ -6,9 +6,9 @@
 #include <vector>
 #include <string>
 
+#include "re2.h"
+
 struct PGN {
-  // Parses a subset of the PGN language. Returns false upon failure.
-  static bool Parse(const std::string &s, PGN *pgn);
 
   enum Result {
     WHITE_WINS,
@@ -16,6 +16,10 @@ struct PGN {
     DRAW,
     OTHER,
   };
+
+  // If you are parsing a large number of PGNs, it is slightly
+  // faster to make a PGNParser instance and reuse it.
+  static bool Parse(const std::string &s, PGN *pgn);
   
   std::unordered_map<std::string, std::string> meta;
   struct Move {
@@ -28,6 +32,15 @@ struct PGN {
   // Does not include the terminating event like 1-0.
   std::vector<Move> moves;
   Result result;
+};
+
+struct PGNParser {
+  // Parses a subset of the PGN language. Returns false upon failure.
+  bool Parse(const std::string &s, PGN *pgn) const;
+  PGNParser();
+  
+private:
+  const RE2 meta_line_re, move_re, end_re;
 };
 
 #endif

### 2018-11-15 trunk/chess/papers/survival.tex
Footnote 1

--- 2018-11-03 chess/papers/survival.tex
@@
 Survival in Chessland
+
+If you are forced to play chess to the death, you are in trouble,
+because most people are not good at chess (for example, the author)
+and yet want to live.\footnote{It is easy for two players to
+  collaborate to produce a draw, especially by simply agreeing to a
+  draw at the outset of the game (if allowed). Some tournament formats
+  forbid the players from agreeing to a draw verbally before a certain
+  point (e.g. 30 moves). There are always other routes to a draw, for
+  example by stalemate or repeating the same position three times.
+  Collaboratively producing such situations is easy, but this strategy
+  is not likely a stable equilibrium: Players can often gain a
+  substantial advantage by going ``off script'' and instead trying to
+  win the game. Additionally, sometimes the terms of
+  chess-to-the-death do not allow the players to communicate at all
+  beforehand, nor during the game. If this is the case, then it may be
+  difficult to agree on the approach to drawing, let alone establish
+  that this is both player's desire. Since the rules of
+  chess-to-the-death can't forbid us from colluding at this moment, I
+  hereby declare that the following is the correct approach: 
+%
+  1. Nf3. This is a reasonable opening move for white (``Zukertort opening'')
+  which can transpose into several common systems (e.g. King's
+  Indian). Since the knight can move back to g1 on the next move,
+  knight moves are the fastest route to a draw by repetition. This
+  has a good chance of signaling to a wise player that a draw is
+  desired. The player should make this move after pondering carefully
+  for some time, and then looking meaningfully into the other player's
+  eyes. 
+  1. ... Nf6. This is both a strong response for black in a real game,
+  and simultaneously a signal that a draw is desired. The other
+  advantage is that very weak players may simply copy what white does.
+  In doing so, they will also play this move. 
+  2. Ng1?!. This is a terrible move for white, but clearly signals
+  the intention to draw.
+  2. ... Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
+  for a second time.
+  3. Nf3 Nf6. At this point the game should clearly continue repeating
+  the sequence, although since we are in the start position, white has
+  any number of strong opening moves available. Signaling the draw line
+  and then 3. d4!? may be pyschologically devastating.
+  4. Ng1 Ng8 1/2-1/2. The starting position is reached for the third
+  time, which by rule is a draw.
+
+  Since this is one of the shortest possible routes to a draw, I hereby
+  dub this line the ``Fool's Draw,'' by analogy with the Fool's Mate.
+
+  If 1. ... Nc6 or another Knight's move, white can also consider
+  continuing in the obvious way. However after 1. ... d5, black
+  has refused or not noticed the draw. Fortunately, white is still
+  in a good position to play the game normally. White can try
+  to be more obvious with 2. Ng1, but if black is choosing to just
+  play normally, white takes a distinct handicap by doing so.
+
+  The biggest risk for white is that black does not play 2. ... Ng8
+  but rather a normal move like 2. ... g6 (``Fool's Draw Betrayed'').
+  This can happen if black is not metagaming at all (for Nf6 is a
+  normal response to the normal Nf3), or if black is an exceptionally
+  shrewd metagamer (tricking white into wasting two tempos with Ng1
+  by pretending to be cooperating).
+
+  Of course, this all relies on the assumption that if
+  chess-to-the-death ends in a draw, the players are spared or allowed
+  to repeat indefinitely. If both players are actually executed, then
+  this line is truly a Fool's Draw! }
+%
+But what if you are forced to {\it be one of the chess pieces} to the
+death? That is, your little soul inhabits one of the 32 pieces or
+pawns and your soul is vanquished if that piece is eliminated. Now it
+doesn't matter whether you're good or bad at chess, but
 
@@
 
-Neither side has a decisive advantage, and many game end in a draw,
+Neither side has a decisive advantage, and many games end in a draw,
 with both kings surviving. So the survival chances of a king are
@@
 
+% Battle chess
+% Feminist game
+
 \end{document}

### 2018-11-18 trunk/chess/papers/survival.tex
Allow games to have illegal moves in them (and skip them). Some old games in the lichess database have some illegal castling.

@@
+\documentclass[twocolumn]{article}
+\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}
+
+\usepackage{url}
+% \usepackage{code}
+% \usepackage{cite}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{chessboard}
+
+\pagestyle{empty}
+
+\usepackage{ulem}
+% go back to italics for emphasis, though
+\normalem
+
+\usepackage{natbib}
+
+% \newcommand\comment[1]{}
+\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
+
 \begin{document}
 
-Survival in Chessland
+\title{Survival in chessland}
+\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
+Copyright \copyright\ 2018 the Regents of the Wikiplia
+Foundation. Appears in SIGBOVIK 2018 with the threefold
+repetition of the Association for Computational Heresy; 
+{\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
+53 Centipawns} }
+
+\setchessboard{showmover=false}
+
+\newcommand\checkmate{\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}
+
+\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
+\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
+\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
+\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
+\newcommand\at{\ensuremath{\scriptstyle @}}
+
+\date{1 April 2018}
+
+\maketitle \thispagestyle{empty}
+
+\begin{abstract}
+CHESSMATE.
+\end{abstract}
+
+\section*{Introduction}
 
@@
 pawns and your soul is vanquished if that piece is eliminated. Now it
-doesn't matter whether you're good or bad at chess, but
+doesn't matter whether you're good or bad at chess, because you don't
+get to pick what happens in the game. What matters is that your piece
+lives to the end of the game, when all surviving pieces are set free.
+Which piece should you want to be?
 
@@
 
-Here are my guesses:
+\section{Hypotheses}
+
+Like all good scientific research, I clearly laid out my hypothesis
+and wrote down the motivation before performing the study. This helps
+prevent presentation bias where the results appear more satisfactory
+because they are framed as a natural conclusion from the idea that
+motivated the research in the first place (when in fact, of course,
+if you write the motivation after witnessing the results, ..). 
+It is also much more exciting. I literally don't know the answer
+as I'm writing this, nor whether it is interesting in any way!
+
+Here are my guesses.
 
@@
 
-This leaves my final ranking, from least surviving to most surviving,
-as:
+This leaves my final ranking, from most surviving to most dead:
 
-16. d2,
-15. e2,
-14. b1 knight,
-13. g1 knight,
-12. c1 bishop,
-11. f1 bishop,
-10. queen,
-9. king,
-8. a1 rook,
-7. h1 rook,
-6. b2,
-5. h2
-4. a2
-3. g2
-2. c2
-1. f2
+\begin{enumerate}
+\item  f2
+\item  c2
+\item  g2
+\item  a2
+\item  h2
+\item  b2,
+\item  h1 rook,
+\item  a1 rook,
+\item  king,
+\item  queen,
+\item  f1 bishop,
+\item  c1 bishop,
+\item  g1 knight,
+\item  b1 knight,
+\item  e2,
+\item  d2,
+\end{enumerate}
 
-% Ben, David, Jim also wagered guesses
+As already copped to, while the author is an aficionado and also knows
+how to spell the difficult word aficionado without spell-check, he is
+not good at chess. A few drinking buddies with varying chessperience were
+also consulted for their wagers.
+
+\subsection{Ben}
+
+edge pawns almost never played til endgame let alone traded off
+
+ph
+pa
+
+not quite sure where these should go (pb more likely to see play in
+queenside minority attacks in k-side castle games?)
+
+pf
+pg
+pb
+
+rook play more likely to be active on q side than on k side (also the
+classic Nxc7 fork in low rank play), but overall more likely to stay
+tucked away compared to q
+
+Rh
+Ra
+
+i think IQP positions are more likely than not saccing e in e4 openings
+but on the other hand d is often traded off in e4 openings while vice
+versa is not as true
+
+pe
+pd
+
+q probably involved in many checkmates (low ranked play) or resignations
+before traded off (high ranked play)
+
+Q
+
+just randomly guessing k dies in about 1/3 of games, times 1/2 for 2
+sides
+
+K
+
+this pawn is a super goner (sicilian, QGA, ...)
+
+pc
+
+most doomed seem to be the minor pieces as i'd guess at least half of
+them get traded off on c/f/3/6 or e/d/4/5 in near every game so
+
+Bc
+Bf
+Nb
+Ng
+
+\subsection{Jim}
+
+Most-to-least-survival hero tier list for chess (patch 1.0):
+
+1.: King
+ -- If I estimate that about 2/3rds of all regular pieces are captured in an average game, and the probability of any non-king piece being captured is uniform, then the king is clearly the most likely to survive.
+(I'm going to break symmetry here and rank black king less likely to survive than white king.)
+
+2. Both Rooks
+ -- Kept in reserve for castling purposes.
+
+3. The A, B, G, and H pawns
+ -- maybe people will forget to move them because they are far from the center.
+
+4. Both Knights
+ -- They are slippery, but they often get deep into enemy territory quickly.
+
+5. The C,D,E, F pawns
+ -- Moved forward to release various more important pieces => more likely to die.
+
+6. Both Bishops
+ -- https://www.youtube.com/watch?v=gDnE-5lD7w8
+
+7. The Queen
+ -- A high-value target, seems unlikely to survive.
+
+This list was assembled for chess.report by ix, who only barely understands the rules of chess and rarely plays. This is probably why the justifications get increasingly nonsensical.
+
+\subsection{David}
+
+Here's my ranking, along with the probabilities of survival that I guess for each piece:
+
+white g-pawn: 0.72
+black g-pawn: 0.72
+white b-pawn: 0.69
+black a-pawn: 0.66
+black b-pawn: 0.65
+white king: 0.65
+white a-pawn: 0.64
+black king: 0.55
+black f-pawn: 0.54
+white f-pawn: 0.54
+black h-pawn: 0.53
+white h-pawn: 0.53
+white kingside rook: 0.52
+black kingside rook: 0.52
+white queenside rook: 0.51
+black queenside rook: 0.51
+white c-pawn: 0.49
+black c-pawn: 0.44
+white light-squared bishop: 0.33
+black light-squared bishop: 0.33
+white dark-squared bishop: 0.32
+black dark-squared bishop: 0.32
+white queen: 0.31
+black queen: 0.30
+white kingside knight: 0.29
+black kingside knight: 0.29
+white queenside knight: 0.28
+black queenside knight: 0.28
+white e-pawn: 0.19
+black d-pawn: 0.18
+white d-pawn: 0.17
+black e-pawn: 0.16
+
+\subsection{William}
+
+Okay, this is pretty much off the cuff, but let’s say, just for white, from most to least likely to survive:
+
+Queenside Rook
+B Pawn
+A Pawn
+Kingside Bishop
+C Pawn
+Queen
+King
+H Pawn
+Queenside Bishop
+F Pawn
+G Pawn
+Queenside Knight
+Kingside Rook
+Kingside Knight
+D Pawn
+E Pawn
+
+Some motivated reasoning:
+
+I figure the king has got to be somewhere near the middle of the pack since he dies in half of games featuring a winner — but with slightly higher-than-even odds of surviving, since some games end in a draw.
+I’m probably mixing up means and medians here somehow..
+
+...
+King
+...
+
+I’m gonna assume castling happens more often on the King’s side, so let’s give Kingside Rook and F, G, and H Pawns a better shot than their fellows on the left.
+But maybe it should actually be worse, since if they die, it’s because they failed to protect the king.
+Plus, having heard the tip about C Pawn loud and clear, I’m gonna assume that bad boy most often becomes a new Queen, which means he gets more survival points than the real Queen herself.
+
+...
+C Pawn
+Queen
+A Pawn
+B Pawn
+Queenside Rook
+King
+F Pawn
+G Pawn
+H Pawn
+Kingside Rook
+...
+
+D and E Pawn are nothing but pawns, and they mostly sacrifice themselves to the cause.
+
+...
+D Pawn
+E Pawn
+
+Randomizing within these constraints gives us our starting point.  Then the wildcard Bishops and Knights get randomly distributed through what remains to come up with this final answer shown above.
 
@@
 
+\nocite{chesstego}
+
+\bibliography{chess}{}
+\bibliographystyle{plain}
+
 \end{document}

### 2018-12-09 chess/papers/survival.tex
Work on paper, maketable

@@
+\documentclass[twocolumn]{article}
+\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}
+
+\usepackage{url}
+% \usepackage{code}
+% \usepackage{cite}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{chessboard}
+
+% lets me explicitly set a. or 1. etc. as enum label
+\usepackage{enumitem}
+
+\pagestyle{empty}
+
+\usepackage{ulem}
+% go back to italics for emphasis, though
+\normalem
+
+\usepackage{natbib}
+
+\setlength{\footnotesep}{2em}
+
+% \newcommand\comment[1]{}
+\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
+
 \begin{document}
 
-Survival in Chessland
+\title{Survival in chessland}
+\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
+Copyright \copyright\ 2018 the Regents of the Wikiplia
+Foundation. Appears in SIGBOVIK 2018 with the threefold
+repetition of the Association for Computational Heresy; 
+{\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
+53 Centipawns} }
+
+\setchessboard{showmover=false}
+
+\newcommand\checkmate{\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}
+
+\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
+\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
+\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
+\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
+\newcommand\at{\ensuremath{\scriptstyle @}}
+
+\date{1 April 2018}
+
+\maketitle \thispagestyle{empty}
+
+\begin{abstract}
+CHESSMATE.
+\end{abstract}
+
+\section*{Introduction}
+
+If you are forced to play chess to the death, you are in trouble,
+because most people are not good at chess (for example, the author)
+and yet want to live.\footnote{It is easy for two players to
+  collaborate to produce a draw, especially by simply agreeing to a
+  draw at the outset of the game (if allowed). Some tournament formats
+  forbid the players from agreeing to a draw verbally before a certain
+  point (e.g. 30 moves). There are always other routes to a draw, for
+  example by stalemate or repeating the same position three times.
+  Collaboratively producing such situations is easy, but this strategy
+  is not likely a stable equilibrium: Players can often gain a
+  substantial advantage by going ``off script'' and instead trying to
+  win the game. Additionally, sometimes the terms of
+  chess-to-the-death do not allow the players to communicate at all
+  beforehand, nor during the game. If this is the case, then it may be
+  difficult to agree on the approach to drawing, let alone establish
+  that this is both player's desire. Since the rules of
+  chess-to-the-death can't forbid us from colluding at this moment, I
+  hereby declare that the following is the correct approach: 
+\begin{enumerate}[label=\arabic*.]
+  \item[1.] Nf3. This is a reasonable opening move for white (begins the R\'eti)
+    which can transpose into several common systems (e.g. King's
+    Indian). Since the knight can move back to g1 on the next move,
+    knight moves are the fastest route to a draw by repetition. This
+    has a good chance of signaling to a wise player that a draw is
+    desired. The player should make this move after pondering carefully
+    for some time, and then looking meaningfully into the other player's
+    eyes. 
+  \item[1.] ... Nf6. This is both a strong response for black in a real game,
+    and simultaneously a signal that a draw is desired. The other
+    advantage is that very weak players may simply copy what white does.
+    In doing so, they will also play this move. 
+  \item[2.] Ng1?!. White moves the knight back to its starting square.
+    This is a terrible move for white, but clearly signals the intention to
+    draw.
+  \item[2.] ... Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
+    for a second time.
+  \item[3.] Nf3 Nf6. At this point the game should clearly continue repeating
+    the sequence, although since we are in the start position, white has
+    any number of strong opening moves available. Signaling the draw line
+    and then 3. d4!? may be pyschologically devastating.
+  \item[4.] Ng1 Ng8 1/2-1/2. The starting position is reached for the third
+    time, which by rule\footnotemark\ is a draw.
+\end{enumerate}
+
+  Since this is one of the shortest possible routes to a draw, I hereby
+  dub this line the ``Fool's Draw,'' by analogy with the Fool's Mate.
+
+  If 1. ... Nc6 or another Knight's move, white can also consider
+  continuing in the obvious way. However after 1. ... d5, black has
+  refused or not noticed the draw. Fortunately, white is still in a
+  good position to play the game normally (this is the main line of
+  the R\'eti opening, followed by 2. c4). White can try to be more
+  obvious with 2. Ng1, but if black is choosing to just play normally,
+  white takes a distinct handicap by doing so.
+
+  The biggest risk for white is that black does not play 2. ... Ng8
+  but rather a normal move like 2. ... g6 (``Fool's Draw Betrayed'').
+  This can happen if black is not metagaming at all (for Nf6 is a
+  normal response to the normal Nf3), or if black is an exceptionally
+  shrewd metagamer (tricking white into wasting two tempos with Ng1
+  by pretending to be cooperating).
+
+  Of course, this all relies on the assumption that if
+  chess-to-the-death ends in a draw, the players are spared or allowed
+  to repeat indefinitely. If both players are actually executed, then
+  this line is truly a Fool's Draw!
+}
+%
+\footnotetext{
+  But is it? The lichess implementation (although known
+  to be buggy\cite{kingme}) does not permit a threefold repetition
+  claim in this situation, which got me thinking that maybe there is
+  some subtlety here. Is the starting position special somehow, not
+  counting as having occurred? The relevant statute, from the FIDE
+  Laws of Chess\cite{fiderules}:
+  \begin{quote}
+    {\bf 9.2}.\quad The game is drawn, upon a correct claim by a
+    player having the move, when the same position for at least the
+    third time (not necessarily by a repetition of moves):
+    \begin{enumerate}[label=\alph*.]
+    \item is about to appear, if he first writes his move, which
+      cannot be changed, on his scoresheet and declares to the arbiter
+      his intention to make this move, or
+    \item has just appeared, and the player claiming the draw has the
+      move.
+  \end{enumerate}
+
+  Positions are considered the same if and only if the same player has
+  the move, pieces of the same kind and colour occupy the same squares
+  and the possible moves of all the pieces of both players are the
+  same. Thus positions are not the same if:
+  \begin{enumerate}[label=\arabic*.]
+  \item at the start of the sequence a pawn could have been captured
+    en passant.
+  \item a king or rook had castling rights, but forfeited these after
+    moving. The castling rights are lost only after the king or rook
+    is moved.
+  \end{enumerate}
+  \end{quote}
+
+  So the question is, has the starting position ``appeared'' before
+  white's first move? The rules are not totally clear on this point.
+  Note that ``positions are considered the same'' only when the
+  same player ``has the move.'' FIDE defines ``have the move'' as
+  \begin{quote}
+    1.3.\quad A player is said to 'have the move' when his opponent's
+    move has been 'made'.
+  \end{quote}
+  A strong case can therefore be made that white does not 'have the
+  move' in the formal sense at the beginning of the game, since black
+  has not made a move!
+
+  Nonetheless, it does seem clear that white can claim a draw by
+  9.2.a, by committing the move 5. Nf3 and declaring to the arbiter
+  that the position is now {\it about to appear} for the third time.
+  This seems unambiguously legal. }
+%
+But what if you are forced to {\it be one of the chess pieces} to the
+death? That is, your little soul inhabits one of the 32 pieces or
+pawns and your soul is vanquished if that piece is eliminated. Now it
+doesn't matter whether you're good or bad at chess, because you don't
+get to pick what happens in the game. What matters is that your piece
+lives to the end of the game, when all surviving pieces are set free.
+Which piece should you want to be?
 
@@
 
-Here are my guesses:
+\section{Hypotheses}
+
+Like all good scientific research, I clearly laid out my hypothesis
+and wrote down the motivation before performing the study. This helps
+prevent presentation bias where the results appear more satisfactory
+because they are framed as a natural conclusion from the idea that
+motivated the research in the first place (when in fact, of course, if
+you write the motivation after witnessing the results, backflow is
+inevitable). It is also much more exciting. I literally don't know the
+answer as I'm writing this, nor whether it is interesting in any way!
+
+Here are my guesses.
 
@@
 
-This leaves my final ranking, from least surviving to most surviving,
-as:
+This leaves my final ranking, from most surviving to most dead:
 
-16. d2,
-15. e2,
-14. b1 knight,
-13. g1 knight,
-12. c1 bishop,
-11. f1 bishop,
-10. queen,
-9. king,
-8. a1 rook,
-7. h1 rook,
-6. b2,
-5. h2
-4. a2
-3. g2
-2. c2
-1. f2
+\begin{enumerate}
+\item  f2
+\item  c2
+\item  g2
+\item  a2
+\item  h2
+\item  b2,
+\item  h1 rook,
+\item  a1 rook,
+\item  king,
+\item  queen,
+\item  f1 bishop,
+\item  c1 bishop,
+\item  g1 knight,
+\item  b1 knight,
+\item  e2,
+\item  d2,
+\end{enumerate}
 
-% Ben, David, Jim also wagered guesses
+As already copped to, while the author is an aficionado and also knows
+how to spell the difficult word aficionado without spell-check, he is
+not good at chess. A few drinking buddies with varying chessperience were
+also consulted for their wagers.
+
+\subsection{Ben}\footnote{Ben does not prefer to use the shift key.}
+
+edge pawns almost never played til endgame let alone traded off
+
+ph
+pa
+
+not quite sure where these should go (pb more likely to see play in
+queenside minority attacks in k-side castle games?)
+
+pf
+pg
+pb
+
+rook play more likely to be active on q side than on k side (also the
+classic Nxc7 fork in low rank play), but overall more likely to stay
+tucked away compared to q
+
+Rh
+Ra
+
+i think IQP positions are more likely than not saccing e in e4 openings
+but on the other hand d is often traded off in e4 openings while vice
+versa is not as true
+
+pe
+pd
+
+q probably involved in many checkmates (low ranked play) or resignations
+before traded off (high ranked play)
+
+Q
+
+just randomly guessing k dies in about 1/3 of games, times 1/2 for 2
+sides
+
+K
+
+this pawn is a super goner (sicilian, QGA, ...)
+
+pc
+
+most doomed seem to be the minor pieces as i'd guess at least half of
+them get traded off on c/f/3/6 or e/d/4/5 in near every game so
+
+Bc
+Bf
+Nb
+Ng
+
+\subsection{Jim}
+
+Most-to-least-survival hero tier list for chess (patch 1.0):
+
+1.: King
+ -- If I estimate that about 2/3rds of all regular pieces are captured in an average game, and the probability of any non-king piece being captured is uniform, then the king is clearly the most likely to survive.
+(I'm going to break symmetry here and rank black king less likely to survive than white king.)
+
+2. Both Rooks
+ -- Kept in reserve for castling purposes.
+
+3. The A, B, G, and H pawns
+ -- maybe people will forget to move them because they are far from the center.
+
+4. Both Knights
+ -- They are slippery, but they often get deep into enemy territory quickly.
+
+5. The C,D,E,F pawns
+ -- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.
+
+6. Both Bishops
+ -- https://youtu.be/gDnE-5lD7w8
+
+7. The Queen
+ -- A high-value target, seems unlikely to survive.
+
+This list was assembled for chess.report by ix, who only barely
+understands the rules of chess and rarely plays. This is probably why
+the justifications get increasingly nonsensical.
+
+\subsection{David}
+
+Here's my ranking, along with the probabilities of survival that I guess for each piece:
+
+white g-pawn: 0.72
+black g-pawn: 0.72
+white b-pawn: 0.69
+black a-pawn: 0.66
+black b-pawn: 0.65
+white king: 0.65
+white a-pawn: 0.64
+black king: 0.55
+black f-pawn: 0.54
+white f-pawn: 0.54
+black h-pawn: 0.53
+white h-pawn: 0.53
+white kingside rook: 0.52
+black kingside rook: 0.52
+white queenside rook: 0.51
+black queenside rook: 0.51
+white c-pawn: 0.49
+black c-pawn: 0.44
+white light-squared bishop: 0.33
+black light-squared bishop: 0.33
+white dark-squared bishop: 0.32
+black dark-squared bishop: 0.32
+white queen: 0.31
+black queen: 0.30
+white kingside knight: 0.29
+black kingside knight: 0.29
+white queenside knight: 0.28
+black queenside knight: 0.28
+white e-pawn: 0.19
+black d-pawn: 0.18
+white d-pawn: 0.17
+black e-pawn: 0.16
+
+\subsection{William}
+
+Okay, this is pretty much off the cuff, but let’s say, just for white,
+from most to least likely to survive:
+
+Queenside Rook
+B Pawn
+A Pawn
+Kingside Bishop
+C Pawn
+Queen
+King
+H Pawn
+Queenside Bishop
+F Pawn
+G Pawn
+Queenside Knight
+Kingside Rook
+Kingside Knight
+D Pawn
+E Pawn
+
+Some motivated reasoning:
+
+I figure the king has got to be somewhere near the middle of the pack
+since he dies in half of games featuring a winner — but with slightly
+higher-than-even odds of surviving, since some games end in a draw.
+I’m probably mixing up means and medians here somehow..
+
+...
+King
+...
+
+I’m gonna assume castling happens more often on the King’s side, so
+let’s give Kingside Rook and F, G, and H Pawns a better shot than
+their fellows on the left. But maybe it should actually be worse,
+since if they die, it’s because they failed to protect the king. Plus,
+having heard the tip about C Pawn loud and clear, I’m gonna assume
+that bad boy most often becomes a new Queen, which means he gets more
+survival points than the real Queen herself.
+
+...
+C Pawn
+Queen
+A Pawn
+B Pawn
+Queenside Rook
+King
+F Pawn
+G Pawn
+H Pawn
+Kingside Rook
+...
+
+D and E Pawn are nothing but pawns, and they mostly sacrifice
+themselves to the cause.
+
+...
+D Pawn
+E Pawn
+
+Randomizing within these constraints gives us our starting point. Then
+the wildcard Bishops and Knights get randomly distributed through what
+remains to come up with this final answer shown above.
+
+\section{Methodology}
+
+To compute the chances for survival, I legally acquired
+
+% Battle chess
+% Feminist game
+
+\nocite{chesstego}
+
+\bibliography{chess}{}
+\bibliographystyle{plain}
 

### 2018-12-09 chess/maketable.cc

@@ -59,6 +59,10 @@ static uint8 PiecePiece(int p) {
   return startpos.PieceAt(prow, pcol);
 }
 
+static string PieceCol(int p) {
+  return StringPrintf("%c", 'a' + (p % 8));
+}
+
 void ReadStats(const string &filename, Stats *stat_buckets) {
   ifstream s;
   s.open(filename);
@@ -213,9 +217,27 @@ void GenReport(Stats *stat_buckets) {
     FILE *f = fopen("survival.svg", "wb");
     fprintf(f, "%s", TextSVG::Header(WIDTH, HEIGHT).c_str());
 
-    
+
+    // Just keep a set of used rectangles. Place the piece at
+    // "the first" x position such that it does not overlap any
+    // existing rectangle. This has bad asymptotic complexity,
+    // but there are only 32 pieces!
     
     for (int i = 0; i < pieces.size(); i++) {
+      auto PieceName =
+	[](int x) {
+	  const int p = PiecePiece(x);
+	  const string ent = Position::HTMLEntity(p);
+	  const string col = PieceCol(x);
+	  switch (p & Position::TYPE_MASK) {
+	  case Position::QUEEN:
+	  case Position::KING:
+	    return ent;
+	  default:
+	    return ent + col;
+	  }
+	};
+
       const int p = pieces[i];
       const double x = MARGIN + i * ((WIDTH - MARGIN * 2.0) /
 				     (pieces.size() + 1.0));
@@ -230,9 +252,7 @@ void GenReport(Stats *stat_buckets) {
 
       fprintf(f, "%s\n",
 	      TextSVG::Text(x + 6.0, cy + 3.2, "sans-serif",
-			    11.0, {{"#000",
-				    Position::HTMLEntity(PiecePiece(p))
-				    }}).c_str());
+			    11.0, {{"#000", PieceName(p)}}).c_str());
       
       #if 0
       // TODO: Compute "confidence interval" here.
@@ -261,7 +281,8 @@ void GenReport(Stats *stat_buckets) {
     FILE *f = fopen("report.html", "wb");
 
     fprintf(f, "<!doctype html>\n"
-	    "<link href=\"report.css\" rel=\"stylesheet\" type=\"text/css\">\n");
+	    "<link href=\"report.css\" rel=\"stylesheet\" "
+	    "type=\"text/css\">\n");
 
     fprintf(f, "<h1>Overall chances survival (%lld games)</h1>\n"
 	    "<table>"

### 2018-12-15 trunk/chess/papers/survival.tex
Improvements to survival table, paper text

@@
 
+% \usepackage{chessfs}
+\usepackage{adjustbox}
+
+% Define black versions of pieces for inline use. Gross, but it works.
+\newcommand{\Pawn}[1][1.3ex]{%
+\adjustbox{Trim=4.3pt 2.6pt 4.3pt 0pt,width=#1,margin=0.2ex 0ex 0.2ex 0ex}{\BlackPawnOnWhite}%
+}%
+\newcommand{\Rook}[1][1.58ex]{%
+\adjustbox{Trim=3.2pt 2.2pt 3.2pt 0pt,width=#1,raise=0ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackRookOnWhite}%
+}%
+\newcommand{\Knight}[1][1.85ex]{%
+\adjustbox{Trim=2.3pt 2.35pt 2.5pt 0pt,width=#1,raise=-0.03ex,margin=0.14ex 0ex 0.14ex 0ex}{\BlackKnightOnWhite}%
+}%
+\newcommand{\Bishop}[1][1.79ex]{%
+\adjustbox{Trim=2.3pt 2pt 2.3pt 0pt,width=#1,raise=-0.12ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackBishopOnWhite}%
+}%
+\newcommand{\Queen}[1][2.05ex]{%
+\adjustbox{Trim=1.2pt 2.2pt 1.2pt 0pt,width=#1,raise=-0.08ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackQueenOnWhite}%
+}%
+\newcommand{\King}[1][1.95ex]{%
+\adjustbox{Trim=2pt 2pt 2pt 0pt,width=#1,raise=-0.06ex,margin=0.13ex 0ex 0.13ex 0ex}{\BlackKingOnWhite}%
+}%
+
+% lets me explicitly set a. or 1. etc. as enum label
+\usepackage{enumitem}
+
 \pagestyle{empty}
@@
 \usepackage{natbib}
+
+\setlength{\footnotesep}{2em}
 
@@
   hereby declare that the following is the correct approach: 
-%
-  1. Nf3. This is a reasonable opening move for white (``Zukertort opening'')
+\begin{enumerate}[label=\arabic*.]
+  \item[1.] Nf3. This is a reasonable opening move for white (begins the R\'eti)
     which can transpose into several common systems (e.g. King's
@@
     eyes. 
-  1. ... Nf6. This is both a strong response for black in a real game,
+  \item[1.] ... Nf6. This is both a strong response for black in a real game,
     and simultaneously a signal that a draw is desired. The other
@@
     In doing so, they will also play this move. 
-  2. Ng1?!. This is a terrible move for white, but clearly signals
-  the intention to draw.
-  2. ... Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
+  \item[2.] Ng1?!. White moves the knight back to its starting square.
+    This is a terrible move for white, but clearly signals the intention to
+    draw.
+  \item[2.] ... Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
     for a second time.
-  3. Nf3 Nf6. At this point the game should clearly continue repeating
+  \item[3.] Nf3 Nf6. At this point the game should clearly continue repeating
     the sequence, although since we are in the start position, white has
@@
     and then 3. d4!? may be pyschologically devastating.
-  4. Ng1 Ng8 1/2-1/2. The starting position is reached for the third
-  time, which by rule is a draw.
+  \item[4.] Ng1 Ng8 1/2-1/2. The starting position is reached for the third
+    time, which by rule\footnotemark\ is a draw.
+\end{enumerate}
 
@@
   If 1. ... Nc6 or another Knight's move, white can also consider
-  continuing in the obvious way. However after 1. ... d5, black
-  has refused or not noticed the draw. Fortunately, white is still
-  in a good position to play the game normally. White can try
-  to be more obvious with 2. Ng1, but if black is choosing to just
-  play normally, white takes a distinct handicap by doing so.
+  continuing in the obvious way. However after 1. ... d5, black has
+  refused or not noticed the draw. Fortunately, white is still in a
+  good position to play the game normally (this is the main line of
+  the R\'eti opening, followed by 2. c4). White can try to be more
+  obvious with 2. Ng1, but if black is choosing to just play normally,
+  white takes a distinct handicap by doing so.
 
@@
   to repeat indefinitely. If both players are actually executed, then
-  this line is truly a Fool's Draw! }
+  this line is truly a Fool's Draw!
+}
+%
+\footnotetext{
+  But is it? The lichess implementation (although known
+  to be buggy\cite{kingme}) does not permit a threefold repetition
+  claim in this situation, which got me thinking that maybe there is
+  some subtlety here. Is the starting position special somehow, not
+  counting as having occurred? The relevant statute, from the FIDE
+  Laws of Chess\cite{fiderules}:
+  \begin{quote}
+    {\bf 9.2}.\quad The game is drawn, upon a correct claim by a
+    player having the move, when the same position for at least the
+    third time (not necessarily by a repetition of moves):
+    \begin{enumerate}[label=\alph*.]
+    \item is about to appear, if he first writes his move, which
+      cannot be changed, on his scoresheet and declares to the arbiter
+      his intention to make this move, or
+    \item has just appeared, and the player claiming the draw has the
+      move.
+  \end{enumerate}
+
+  Positions are considered the same if and only if the same player has
+  the move, pieces of the same kind and colour occupy the same squares
+  and the possible moves of all the pieces of both players are the
+  same. Thus positions are not the same if:
+  \begin{enumerate}[label=\arabic*.]
+  \item at the start of the sequence a pawn could have been captured
+    en passant.
+  \item a king or rook had castling rights, but forfeited these after
+    moving. The castling rights are lost only after the king or rook
+    is moved.
+  \end{enumerate}
+  \end{quote}
+
+  So the question is, has the starting position ``appeared'' before
+  white's first move? The rules are not totally clear on this point.
+  Note that ``positions are considered the same'' only when the
+  same player ``has the move.'' FIDE defines ``have the move'' as
+  \begin{quote}
+    1.3.\quad A player is said to 'have the move' when his opponent's
+    move has been 'made'.
+  \end{quote}
+  A strong case can therefore be made that white does not 'have the
+  move' in the formal sense at the beginning of the game, since black
+  has not made a move!
+
+  Nonetheless, it does seem clear that white can claim a draw by
+  9.2.a, by committing the move 5. Nf3 and declaring to the arbiter
+  that the position is now {\it about to appear} for the third time.
+  This seems unambiguously legal. }
 %
@@
 because they are framed as a natural conclusion from the idea that
-motivated the research in the first place (when in fact, of course,
-if you write the motivation after witnessing the results, ..). 
-It is also much more exciting. I literally don't know the answer
-as I'm writing this, nor whether it is interesting in any way!
+motivated the research in the first place (when in fact, of course, if
+you write the motivation after witnessing the results, backflow is
+inevitable). It is also much more exciting. I literally don't know the
+answer as I'm writing this, nor whether it is interesting in any way!
 
@@
 
-\subsection{Ben}
+\subsection{Ben}\footnote{Ben does not prefer to use the shift key.}
 
@@
 
-5. The C,D,E, F pawns
- -- Moved forward to release various more important pieces => more likely to die.
+5. The C,D,E,F pawns
+ -- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.
 
 6. Both Bishops
- -- https://www.youtube.com/watch?v=gDnE-5lD7w8
+ -- https://youtu.be/gDnE-5lD7w8
 
@@
 
-This list was assembled for chess.report by ix, who only barely understands the rules of chess and rarely plays. This is probably why the justifications get increasingly nonsensical.
+This list was assembled for chess.report by ix, who only barely
+understands the rules of chess and rarely plays. This is probably why
+the justifications get increasingly nonsensical.
 
@@
 
-Okay, this is pretty much off the cuff, but let’s say, just for white, from most to least likely to survive:
+Okay, this is pretty much off the cuff, but let’s say, just for white,
+from most to least likely to survive:
 
@@
 
-I figure the king has got to be somewhere near the middle of the pack since he dies in half of games featuring a winner — but with slightly higher-than-even odds of surviving, since some games end in a draw.
+I figure the king has got to be somewhere near the middle of the pack
+since he dies in half of games featuring a winner — but with slightly
+higher-than-even odds of surviving, since some games end in a draw.
 I’m probably mixing up means and medians here somehow..
@@
 
-I’m gonna assume castling happens more often on the King’s side, so let’s give Kingside Rook and F, G, and H Pawns a better shot than their fellows on the left.
-But maybe it should actually be worse, since if they die, it’s because they failed to protect the king.
-Plus, having heard the tip about C Pawn loud and clear, I’m gonna assume that bad boy most often becomes a new Queen, which means he gets more survival points than the real Queen herself.
+I’m gonna assume castling happens more often on the King’s side, so
+let’s give Kingside Rook and F, G, and H Pawns a better shot than
+their fellows on the left. But maybe it should actually be worse,
+since if they die, it’s because they failed to protect the king. Plus,
+having heard the tip about C Pawn loud and clear, I’m gonna assume
+that bad boy most often becomes a new Queen, which means he gets more
+survival points than the real Queen herself.
 
@@
 
-D and E Pawn are nothing but pawns, and they mostly sacrifice themselves to the cause.
+D and E Pawn are nothing but pawns, and they mostly sacrifice
+themselves to the cause.
 
@@
 
-Randomizing within these constraints gives us our starting point.  Then the wildcard Bishops and Knights get randomly distributed through what remains to come up with this final answer shown above.
+Randomizing within these constraints gives us our starting point. Then
+the wildcard Bishops and Knights get randomly distributed through what
+remains to come up with this final answer shown above.
+
+\section{Methodology}
+
+To compute the chances for survival, I legally acquired 506,000,416
+chess games from lichess.org. This is all of the standard variant,
+rated games from Jan 2013 to November 2018, in any time format. Only
+games that are completed and valid are included (about 200,000 games
+did not meet this criteria). The total data size is 875 gigabytes, so
+processing these took some care for efficiency and parallelism.
+
+Other than that, I simply parsed and simulated each game. For each
+of the 32 pieces in the starting position, I tracked its current
+location, and whether it is alive; multiple dead pieces could
+occupy the same square. At the end of the game, one of the kings is
+killed if his side has lost.
+
+For a piece, there is a single factual survival rate in these games,
+given just by $\frac{\mathrm{num~survived}}{\mathrm{num~games}}$. What
+we're really interested in, however, is estimating the underlying true
+survival probability for each piece. In order to do this with
+reasonable efficency, we divide the games into 32 separate buckets,
+and count statistics separately for each. From these samples we can
+then estimate variance, for example. The games are bucketized by a
+deterministic hash of the White player's username. This way, if there
+exist some players who are highly unusual (perhaps automated
+accounts), their games are grouped together and pessimistically
+represented in the variance estimate. This also helps account for
+different opening preferences; the chosen opening certainly affects
+the survival chances.
+
+The basic survival chances appear in Figure~\ref{fig:piece-survival}.
+Indeed, many pieces are more likely to survive than the kings. Even as
+black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
+survival rate. Across the board, the survival chances for a white
+piece and its black twin are similar, usually with a small edge to
+white. Notable exceptions are the \knight g (the overall most doomed
+piece), and both white bishops, which die more than their black
+counterparts. The \pawn e is vastly more dead than \Pawn e. Note
+somewhat satisfyingly that the \pawn c has the highest variance; this
+was the most controversial among the drinking buddies
+(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed pawn
+  in the popular Queen's Gambit (1. d4 d5 2. c4); variance may come
+  from individual players having a strong preference to accept or
+  decline.\footnote{XXX actually this can't be true because we group
+    by the white player, but the black player chooses what to do.}
+
+% XXX
+
+\begin{figure}[htb]
+  \begin{center}
+    % To generate this pdf, I ran maketable to make
+    % piece-survival.svg, then printed to PDF using Chrome, then
+    % cropped it in Illustrator. Illustrator can open the SVG but
+    % doesn't have the unicode chess pieces in any font I could find.
+    \includegraphics[width=0.95 \linewidth]{piece-survival-export}
+  \end{center}\vspace{-0.1in}
+  \caption{Survival probabilities for each of the 32 pieces in standard
+    chess, in 500 million games. The number is the mean survival rate
+    across all samples. The vertical position of the dot is this mean
+    rate, with a line drawing the span between the smallest and largest
+    sample bucket (this is usually a very tiny range).
+    Horizontal position is purely presentational, to avoid overlap.}
+  \label{fig:piece-survival}
+\end{figure}
 

### 2018-12-15 chess/papers/survival.tex
Generate and import survival maps for each piece

@@
 \documentclass[twocolumn]{article}
-\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}
+\usepackage[top=0.5in, left=0.45in, right=0.45in, bottom=0.5in]{geometry}
 
@@
 \usepackage{chessboard}
+
+% \usepackage{chessfs}
+\usepackage{adjustbox}
+
+% Define black versions of pieces for inline use. Gross, but it works.
+\newcommand{\Pawn}[1][1.3ex]{%
+\adjustbox{Trim=4.3pt 2.6pt 4.3pt 0pt,width=#1,margin=0.2ex 0ex 0.2ex 0ex}{\BlackPawnOnWhite}%
+}%
+\newcommand{\Rook}[1][1.58ex]{%
+\adjustbox{Trim=3.2pt 2.2pt 3.2pt 0pt,width=#1,raise=0ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackRookOnWhite}%
+}%
+\newcommand{\Knight}[1][1.85ex]{%
+\adjustbox{Trim=2.3pt 2.35pt 2.5pt 0pt,width=#1,raise=-0.03ex,margin=0.14ex 0ex 0.14ex 0ex}{\BlackKnightOnWhite}%
+}%
+\newcommand{\Bishop}[1][1.79ex]{%
+\adjustbox{Trim=2.3pt 2pt 2.3pt 0pt,width=#1,raise=-0.12ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackBishopOnWhite}%
+}%
+\newcommand{\Queen}[1][2.05ex]{%
+\adjustbox{Trim=1.2pt 2.2pt 1.2pt 0pt,width=#1,raise=-0.08ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackQueenOnWhite}%
+}%
+\newcommand{\King}[1][1.95ex]{%
+\adjustbox{Trim=2pt 2pt 2pt 0pt,width=#1,raise=-0.06ex,margin=0.13ex 0ex 0.13ex 0ex}{\BlackKingOnWhite}%
+}%
 
@@
 
-To compute the chances for survival, I legally acquired
+To compute the chances for survival, I legally acquired 506,000,416
+chess games from lichess.org. This is all of the standard variant,
+rated games from Jan 2013 to November 2018, in any time format. Only
+games that are completed and valid are included (about 200,000 games
+did not meet this criteria). The total data size is 875 gigabytes, so
+processing these took some care for efficiency and parallelism.
+
+Other than that, I simply parsed and simulated each game. For each
+of the 32 pieces in the starting position, I tracked its current
+location, and whether it is alive; multiple dead pieces could
+occupy the same square. At the end of the game, one of the kings is
+killed if his side has lost.
+
+For a piece, there is a single factual survival rate in these games,
+given just by $\frac{\mathrm{num~survived}}{\mathrm{num~games}}$. What
+we're really interested in, however, is estimating the underlying true
+survival probability for each piece. In order to do this with
+reasonable efficency, we divide the games into 32 separate buckets,
+and count statistics separately for each. From these samples we can
+then estimate variance, for example. The games are bucketized by a
+deterministic hash of the White player's username. This way, if there
+exist some players who are highly unusual (perhaps automated
+accounts), their games are grouped together and pessimistically
+represented in the variance estimate. This also helps account for
+different opening preferences; the chosen opening certainly affects
+the survival chances.
+
+\begin{figure}[htb]
+  \begin{center}
+    % To generate this pdf, I ran maketable to make
+    % piece-survival.svg, then printed to PDF using Chrome, then
+    % cropped it in Illustrator. Illustrator can open the SVG but
+    % doesn't have the unicode chess pieces in any font I could find.
+    \includegraphics[width=0.95 \linewidth]{piece-survival-export}
+  \end{center}\vspace{-0.1in}
+  \caption{Survival probabilities for each of the 32 pieces in standard
+    chess, in 500 million games. The number is the mean survival rate
+    across all samples. The vertical position of the dot is this mean
+    rate, with a line drawing the span between the smallest and largest
+    sample bucket (this is usually a very tiny range).
+    Horizontal position is purely presentational, to avoid overlap.}
+  \label{fig:piece-survival}
+\end{figure}
+
+The basic survival chances appear in Figure~\ref{fig:piece-survival}.
+Indeed, many pieces are more likely to survive than the kings. Even as
+black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
+survival rate. Across the board, the survival chances for a white
+piece and its black twin are similar, usually with a small edge to
+white. Notable exceptions are the \knight g (the overall most doomed
+piece), and both white bishops, which die more than their black
+counterparts. The \pawn e is vastly more dead than \Pawn e. Note
+somewhat satisfyingly that the \pawn c has the highest variance; this
+was the most controversial among the drinking buddies
+(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed pawn
+in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a good
+example of a piece that has very different survival rates in different
+opening preferences. Since the variance is otherwise very low, I did
+not attempt to
+
+Despite my impression that many games end in a draw, ties are actually
+rare in the lichess database. In January 2018, only 3.8\% of games
+were drawn;
+% (8922275 white, 8329221 black, 682042 draw)
+as a result, the survival chances for the kings are both close to
+50\%.
+
+\section{Safest spaces}
+
+The fate of each piece is to either survive or die, and it does so on
+one of the 64 squares. With the same replay of the 500 million games
+I also kept statistics on the fates of each piece. If being a chess
+piece to the death, and possessing some influence over the
 
@@
 % Feminist game
+
+Other than the bishops---which have no legal way to reach half of the
+squares---every piece reaches every square in at least a thousand
+games. So we have enough samples to have reasonable confidence in our
+statistics, even for the most unlikely odysseys. The least mobile
+pieces are the pawns, who can technically reach any square by
+promoting, but are usually confined to cones emanating from their
+start squares. The overall rarest fate is for \pawn f2 to die on the
+a7 square, which only happened 1,244 times (however, it survived on
+this square 31,438 times). This square is actually reachable without
+promoting, but it would need to capture 5 times in order to get
+there, which seems quite unlikely! There may even be a hidden
+achievement for reaching this square this way!
+% TODO: Perhaps could find a game where this happened?!
+Aside from pawns, the weirdest fate is for \knight g to die on h1,
+which happened 47,307 times. Corners are of course fairly garbage for
+the knight, although it is twice as likely to survive on this square.
+
+% There may even be hidden achievements for reaching these squares!
+
+That said, there are characteristic patterns for each of the pieces,
+which make sense given their starting positions and patterns of
+movement. You could probably guess the piece just by looking at one of
+the heat maps below, although---spoiler alert---the piece is just
+listed right there and they are in order. In each map, the background
+color gives the
+
+\newcommand{\chessmap}[2]{\begin{center} \includegraphics[width=0.65 \linewidth]{#1} {#2} \end{center}\vspace{-0.1in}}
+
+\chessmap{piece0}{\Rook a8}
+\chessmap{piece1}{\Knight b8}
+\chessmap{piece2}{\Bishop c8}
+\chessmap{piece3}{\Queen}
+\chessmap{piece4}{\King}
+\chessmap{piece5}{\Bishop f8}
+\chessmap{piece6}{\Knight g8}
+\chessmap{piece7}{\Rook h8}
+
+\chessmap{piece8}{\Pawn a7}
+\chessmap{piece9}{\Pawn b7}
+\chessmap{piece10}{\Pawn c7}
+\chessmap{piece11}{\Pawn d7}
+\chessmap{piece12}{\Pawn e7}
+\chessmap{piece13}{\Pawn f7}
+\chessmap{piece14}{\Pawn g7}
+\chessmap{piece15}{\Pawn h7}
+
+\chessmap{piece16}{\pawn a2}
+\chessmap{piece17}{\pawn b2}
+\chessmap{piece18}{\pawn c2}
+\chessmap{piece19}{\pawn d2}
+\chessmap{piece20}{\pawn e2}
+\chessmap{piece21}{\pawn f2}
+\chessmap{piece22}{\pawn g2}
+\chessmap{piece23}{\pawn h2}
+
+\chessmap{piece24}{\rook a1}
+\chessmap{piece25}{\knight b1}
+\chessmap{piece26}{\bishop c1}
+\chessmap{piece27}{\queen}
+\chessmap{piece28}{\king}
+\chessmap{piece29}{\bishop f1}
+\chessmap{piece30}{\knight g1}
+\chessmap{piece31}{\rook h1}
 

### 2018-12-15 chess/maketable.cc

@@ -191,7 +191,7 @@ struct RectSet {
     for (int x = 0; /* returns in loop */; x++) {
       r.x0 = x;
       r.x1 = x + w;
-      fprintf(stderr, "Try %d,%d-%d,%d\n", r.x0, r.y0, r.x1, r.y1);
+      // fprintf(stderr, "Try %d,%d-%d,%d\n", r.x0, r.y0, r.x1, r.y1);
       if (!AnyOverlap(r))
 	return r;
     }
@@ -208,6 +208,116 @@ struct RectSet {
   // uses here we're talking about like 32-64 rectangles.
   vector<Rect> rects;
 };
+
+
+// SVG version. This one shows the total rate of a piece ending
+// on each square, as well as its survival odds (given that it's
+// there).
+void GenFateMap(Stats *stat_buckets) {
+  for (int p = 0; p < 32; p++) {
+    FILE *f = fopen(StringPrintf("piece%d.svg", p).c_str(), "wb");
+    Ratio ended_on[64] = {}, died_on[64] = {}, survived_on[64] = {};
+
+    // For each square index, its rank in terms of the
+    // overall probability of the piece's fate ending there.
+    vector<double> ranks;
+
+    for (int b = 0; b < NUM_BUCKETS; b++) {
+      const Stats &stats = stat_buckets[b];
+      for (int i = 0; i < 64; i++) {
+	died_on[i].denom[b] = stats.num_games;
+	survived_on[i].denom[b] = stats.num_games;
+	ended_on[i].denom[b] = stats.num_games;
+	
+	died_on[i].numer[b] = stats.pieces[p].died_on[i];
+	survived_on[i].numer[b] = stats.pieces[p].survived_on[i];
+	ended_on[i].numer[b] = stats.pieces[p].died_on[i] +
+	  stats.pieces[p].survived_on[i];
+
+	ranks.push_back(ended_on[i].Mean());
+      }
+    }
+
+    std::sort(ranks.begin(), ranks.end());
+
+    auto GetRank =
+      [](const vector<double> &ranks, double value) {
+	// PERF can be done with binary search, obv...
+	for (int i = 0; i < ranks.size(); i++) {
+	  if (value <= ranks[i]) return i / (double)ranks.size();
+	}
+	return 1.0;
+      };
+
+    // XXX PiecePiece?
+    const bool black = p < 16;
[…]

### 2018-12-15 trunk/chess/rungames.cc
Generate and import survival maps for each piece

@@ -199,6 +199,7 @@ static void ReadLargePGN(const char *filename) {
   auto DoWork = 
     [&bad_games_m, &bad_games,
      &stat_buckets, &parser](const string &pgn_text) {
+      printf("DoWork[%s]\n", pgn_text.c_str());
       PGN pgn;
       CHECK(parser.Parse(pgn_text, &pgn));
 
@@ -337,7 +338,7 @@ static void ReadLargePGN(const char *filename) {
       }
 
       const int bucket = bucket_hash & NUM_BUCKETS_MASK;
-      if (false) {
+      if (true) {
 	fprintf(stderr, "Fates:\n");
 	for (int i = 0; i < 32; i++) {
 	  fprintf(stderr, "%d (%s). %s on %c%c.\n",
@@ -381,40 +382,41 @@ static void ReadLargePGN(const char *filename) {
     };
   
   string game;
-  int c = 0;
   int blank_lines = 0;
   int lastc = 0;
-  while ( EOF != (c = BigGetC()) ) {
-    game += c;
-    if (c == '\n' && lastc == '\n') {
+  int c = 0;
+  do {
+    c = BigGetC();
+    if (c == '\n' && lastc == '\n')
       blank_lines++;
-      if (blank_lines == 2) {
-	work_queue->Add(std::move(game));
-	game.clear();
-	game.reserve(1024);
-	num_read++;
-	if (num_read % 20000LL == 0) {
-	  int64 done, in_progress, pending;
-	  work_queue->Stats(&done, &in_progress, &pending);
-	  fprintf(stderr,
-		  "[Still reading; %lld games at %.1f/sec] %lld %lld %lld\n",
-		  num_read,
-		  num_read / (double)(time(nullptr) - start),
-		  done, in_progress, pending);
-	  fflush(stderr);
-	}
 
-	if (MAX_GAMES > 0 && num_read >= MAX_GAMES)
-	  break;
-	
-	blank_lines = 0;
-	lastc = 0;
-	continue;
+    // Don't require double blank lines at end of file.
+    if (blank_lines == 2 || c == EOF) {
+      work_queue->Add(std::move(game));
+      game.clear();
+      game.reserve(1024);
+      num_read++;
+      if (num_read % 20000LL == 0) {
+	int64 done, in_progress, pending;
+	work_queue->Stats(&done, &in_progress, &pending);
+	fprintf(stderr,
[…]

### 2018-12-16 trunk/chess/papers/survival.tex
Underline safest squares, paper text

@@
 
+\interfootnotelinepenalty=0
+
 % lets me explicitly set a. or 1. etc. as enum label
@@
 \title{Survival in chessland}
-\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
+\author{Dr.~Tom~Murphy~VII~Ph.D.\footnote{
 Copyright \copyright\ 2018 the Regents of the Wikiplia
@@
 }
-%
+
+But what if you are forced to {\it be one of the chess pieces} to the
+death? That is, your little soul inhabits one of the 32 pieces or
+pawns and your soul is vanquished if that piece is eliminated. Now it
+doesn't matter whether you're good or bad at chess, because you don't
+get to pick what happens in the game. What matters is that your piece
+lives to the end of the game, when all surviving pieces are set free.
+Which piece should you want to be?
+
+In formal chess, the king can never be captured: The game ends when
+the king is attacked but cannot move, and it is illegal to make a move
+that leaves the king attacked. The king's death is implied, of course,
+but it is seen as more poetic to end the game prior to this point.
+
+For the sake of this question, we'll consider the the white king to
+``die'' if white loses (i.e., is checkmated), and likewise for black.
+Otherwise, of course, the best chances of survival would trivially be
+with the two kings, since they always survive. Loss includes
+resignation, since most high-level games actually end once the
+defeated player agrees that loss is inevitable. We can think of this
+common case like king seppuku. Many games also end in time forfeit,
+which is like the king's poor diet and lifestyle choices leading to
+a death by natural causes.
+
+Neither side has a decisive advantage, and many games end in a draw,
+with both kings surviving. So the survival chances of a king are
+clearly greater than 50\%---pretty decent odds. Is it possible that
+any other piece has even better chances?
+
 \footnotetext{
@@
   same player ``has the move.'' FIDE defines ``have the move'' as
-  \begin{quote}
-    1.3.\quad A player is said to 'have the move' when his opponent's
+  \begin{enumerate}[label=\arabic*.3.] % hack to get 1.3.
+    \item A player is said to 'have the move' when his opponent's
     move has been 'made'.
-  \end{quote}
+  \end{enumerate}
   A strong case can therefore be made that white does not 'have the
@@
   This seems unambiguously legal. }
-%
-But what if you are forced to {\it be one of the chess pieces} to the
-death? That is, your little soul inhabits one of the 32 pieces or
-pawns and your soul is vanquished if that piece is eliminated. Now it
-doesn't matter whether you're good or bad at chess, because you don't
-get to pick what happens in the game. What matters is that your piece
-lives to the end of the game, when all surviving pieces are set free.
-Which piece should you want to be?
-
-In formal chess, the king can never be captured: The game ends when
-the king is attacked but cannot move, and it is illegal to make a move
-that leaves the king attacked. The king's death is implied, of course,
-but it is seen as more poetic to end the game prior to this point.
-
-For the sake of this question, we'll consider the the white king to
-``die'' if white loses (i.e., is checkmated), and likewise for black.
-Otherwise, of course, the best chances of survival would trivially be
-with the two kings, since they always survive. Loss includes
-resignation, since most high-level games actually end once the
-defeated player agrees that loss is inevitable. We can think of this
-common case like king seppuku. Many games also end in time forfeit,
-which is like the king's poor diet and lifestyle choices leading to
-a death by natural causes.
-
-Neither side has a decisive advantage, and many games end in a draw,
-with both kings surviving. So the survival chances of a king are
-clearly greater than 50\%---pretty decent odds. Is it possible that
-any other piece has even better chances?
 
@@
 answer as I'm writing this, nor whether it is interesting in any way!
+
+% XXX: Move this stuff to the appendix.
 
@@
 
-The basic survival chances appear in Figure~\ref{fig:piece-survival}.
-Indeed, many pieces are more likely to survive than the kings. Even as
-black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
-survival rate. Across the board, the survival chances for a white
-piece and its black twin are similar, usually with a small edge to
-white. Notable exceptions are the \knight g (the overall most doomed
-piece), and both white bishops, which die more than their black
-counterparts. The \pawn e is vastly more dead than \Pawn e. Note
-somewhat satisfyingly that the \pawn c has the highest variance; this
-was the most controversial among the drinking buddies
-(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed pawn
-  in the popular Queen's Gambit (1. d4 d5 2. c4); variance may come
-  from individual players having a strong preference to accept or
-  decline.\footnote{XXX actually this can't be true because we group
-    by the white player, but the black player chooses what to do.}
-
-% XXX
-
 \begin{figure}[htb]
@@
 
+The basic survival chances appear in Figure~\ref{fig:piece-survival}.
+Indeed, many pieces are more likely to survive than the kings. Even as
+black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
+survival rate. Across the board, the survival chances for a white
+piece and its black twin are similar, usually with a small edge to
+white. Notable exceptions are the \knight g (the overall most doomed
+piece), and both white bishops, which die more than their
+Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
+Note somewhat satisfyingly that the \pawn c has the highest variance;
+this was the most controversial among the drinking buddies
+(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
+pawn in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a
+good example of a piece that has very different survival rates in
+different opening preferences. Since the variance is otherwise very
+low, I did not attempt to
+
+Despite my impression that many games end in a draw, ties are actually
+rare in the lichess database. In January 2018, only 3.8\% of games
+were drawn;
+% (8922275 white, 8329221 black, 682042 draw)
+as a result, the survival chances for the kings are both close to
+50\%.
+
+\section{Safest spaces}
+
+The fate of each piece is to either survive or die, and it does so on
+one of the 64 squares. With the same replay of the 500 million games
+I also kept statistics on the fates of each piece. If being a chess
+piece to the death, and possessing some influence over where your piece
+moves, it may be helpful to know where to go. Even without influence,
+such knowledge could help calibrate your anxiety.
+
+Other than the bishops---which have no legal way to reach half of the
+squares---every piece reaches every square in at least a thousand
+games. So we have enough samples to have reasonable confidence in our
+statistics, even for the most unlikely odysseys. The least mobile
+pieces are the pawns, who can technically reach any square by
+promoting, but are usually confined to cones emanating from their
+start squares. The overall rarest fate is for \pawn f2 to die on the
+a7 square, which only happened 1,244 times (however, it survived on
+this square 31,438 times). This square is actually reachable without
+promoting, but it would need to capture 5 times in order to get
+there, which seems quite unlikely! There may even be a hidden
+achievement for reaching this square this way!
+% TODO: Perhaps could find a game where this happened?!
+Aside from pawns, the weirdest fate is for \knight g to die on h1,
+which happened 47,307 times. Corners are of course fairly garbage for
+the knight, although it is twice as likely to survive on this square.
+
+% There may even be hidden achievements for reaching these squares!
+
+That said, there are characteristic patterns for each of the pieces,
+which make sense given their starting positions and patterns of
+movement. You could probably guess the piece just by looking at one of
+the heat maps below, although---spoiler alert---the piece is just
+listed right there and they are in order. Two independent things are
+communicated in these graphics: The chance that the piece ends the
+game on some square (alive or dead), and its survival chances there.
+In each map, a darker background color indicates that the piece ends
+on that square more often. The shade is based on the rank (64/64 black
+is most common, 63/64 black is next most common, etc.) rather than the
+absolute probability, since otherwise the graphic looks boring. The
+number on the square is the percentage survival of the piece when it
+is last seen (either being captured or surviving to the end of the
+game) on that square. The four squares with the highest survival rates
+are underlined for your convenience.
+
+\newcommand{\chessmap}[2]{\begin{center} \includegraphics[width=0.65 \linewidth]{#1} {#2} \end{center}\vspace{-0.1in}}
+
+\chessmap{piece0}{\Rook a8}
+\chessmap{piece1}{\Knight b8}
+\chessmap{piece2}{\Bishop c8}
+\chessmap{piece3}{\Queen d8}
+\chessmap{piece4}{\King e8}
+\chessmap{piece5}{\Bishop f8}
+\chessmap{piece6}{\Knight g8}
+\chessmap{piece7}{\Rook h8}
+
+\chessmap{piece8}{\Pawn a7}
+\chessmap{piece9}{\Pawn b7}
+\chessmap{piece10}{\Pawn c7}
+\chessmap{piece11}{\Pawn d7}
+\chessmap{piece12}{\Pawn e7}
+\chessmap{piece13}{\Pawn f7}
+\chessmap{piece14}{\Pawn g7}
+\chessmap{piece15}{\Pawn h7}
+
+\chessmap{piece16}{\pawn a2}
+\chessmap{piece17}{\pawn b2}
+\chessmap{piece18}{\pawn c2}
+\chessmap{piece19}{\pawn d2}
+\chessmap{piece20}{\pawn e2}
+\chessmap{piece21}{\pawn f2}
+\chessmap{piece22}{\pawn g2}
+\chessmap{piece23}{\pawn h2}
+
+\chessmap{piece24}{\rook a1}
+\chessmap{piece25}{\knight b1}
+\chessmap{piece26}{\bishop c1}
+\chessmap{piece27}{\queen d1}
+\chessmap{piece28}{\king e1}
+\chessmap{piece29}{\bishop f1}
+\chessmap{piece30}{\knight g1}
+\chessmap{piece31}{\rook h1}
+
 % Battle chess

### 2018-12-16 trunk/chess/maketable.cc

@@ -1,6 +1,7 @@
 
 #include "chess.h"
 
+#include <cmath>
 #include <stdio.h>
 #include <unistd.h>
 
@@ -220,8 +221,12 @@ void GenFateMap(Stats *stat_buckets) {
 
     // For each square index, its rank in terms of the
     // overall probability of the piece's fate ending there.
-    vector<double> ranks;
-
+    vector<double> ranks, sranks;
+    for (int i = 0; i < 64; i++) {
+      ranks.push_back(0.0);
+      sranks.push_back(0.0);
+    }
+    
     for (int b = 0; b < NUM_BUCKETS; b++) {
       const Stats &stats = stat_buckets[b];
       for (int i = 0; i < 64; i++) {
@@ -233,30 +238,38 @@ void GenFateMap(Stats *stat_buckets) {
 	survived_on[i].numer[b] = stats.pieces[p].survived_on[i];
 	ended_on[i].numer[b] = stats.pieces[p].died_on[i] +
 	  stats.pieces[p].survived_on[i];
-
-	ranks.push_back(ended_on[i].Mean());
       }
     }
 
+    for (int i = 0; i < 64; i++) {
+      ranks[i] = ended_on[i].Numer();
+      if (ended_on[i].Numer() > 0) {
+	sranks[i] = survived_on[i].Numer() /
+	  (double)ended_on[i].Numer();
+      }
+    }
+    
     std::sort(ranks.begin(), ranks.end());
-
+    std::sort(sranks.begin(), sranks.end());
+    printf("Num sranks: %d\n", (int)sranks.size());
+    
+    // PERF: These can be done with binary search, obviously
+    auto GetRankRevIdx = 
+      [](const vector<double> &ranks, double value) {
+	int rank = 0;
+	for (int i = ranks.size() - 1; i >= 0; i--) {
+	  if (value >= ranks[i]) return rank;
+	  rank++;
+	}
+	return rank;
+      };
+    
     auto GetRank =
       [](const vector<double> &ranks, double value) {
-	// PERF can be done with binary search, obv...
-	for (int i = 0; i < ranks.size(); i++) {
+	for (int i = 0; i < ranks.size(); i++)
 	  if (value <= ranks[i]) return i / (double)ranks.size();
-	}
 	return 1.0;
[…]

### 2018-12-16 chess/papers/survival.tex
Fork rungames for 'chessreduce', and refactor some shared code.

@@
 white. Notable exceptions are the \knight g (the overall most doomed
-piece), and both white bishops, which die more than their black
-counterparts. The \pawn e is vastly more dead than \Pawn e. Note
-somewhat satisfyingly that the \pawn c has the highest variance; this
-was the most controversial among the drinking buddies
-(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed pawn
-in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a good
-example of a piece that has very different survival rates in different
-opening preferences. Since the variance is otherwise very low, I did
-not attempt to
+piece), and both white bishops, which die more than their
+Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
+Note somewhat satisfyingly that the \pawn c has the highest variance;
+this was the most controversial among the drinking buddies
+(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
+pawn in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a
+good example of a piece that has very different survival rates in
+different opening preferences. Since the variance is otherwise
+extremely low, I only report means for the remainder of the paper.
 
@@
 I also kept statistics on the fates of each piece. If being a chess
-piece to the death, and possessing some influence over the
-
-% Battle chess
-% Feminist game
+piece to the death, and possessing some influence over where your piece
+moves, it may be helpful to know where to go. Even without influence,
+such knowledge could help calibrate your anxiety.
 
 Other than the bishops---which have no legal way to reach half of the
-squares---every piece reaches every square in at least a thousand
+squares---every piece ends on every square in at least a thousand
 games. So we have enough samples to have reasonable confidence in our
@@
 the heat maps below, although---spoiler alert---the piece is just
-listed right there and they are in order. In each map, the background
-color gives the
+listed right there and they are in order. Two independent things are
+communicated in these graphics: The chance that the piece ends the
+game on some square (alive or dead), and its survival chances there.
+In each map, a darker background color indicates that the piece ends
+on that square more often. The shade is based on the rank (64/64 black
+is most common, 63/64 black is next most common, etc.) rather than the
+absolute probability, since otherwise the graphic looks boring. The
+number on the square is the percentage survival of the piece when it
+is last seen (either being captured or surviving to the end of the
+game) on that square. The four squares with the highest survival rates
+are underlined for your convenience.
 
@@
 \chessmap{piece2}{\Bishop c8}
-\chessmap{piece3}{\Queen}
-\chessmap{piece4}{\King}
+\chessmap{piece3}{\Queen d8}
+\chessmap{piece4}{\King e8}
 \chessmap{piece5}{\Bishop f8}
@@
 \chessmap{piece26}{\bishop c1}
-\chessmap{piece27}{\queen}
-\chessmap{piece28}{\king}
+\chessmap{piece27}{\queen d1}
+\chessmap{piece28}{\king e1}
 \chessmap{piece29}{\bishop f1}
@@
 \chessmap{piece31}{\rook h1}
+
+% Battle chess
+% Feminist game
 

### 2018-12-16 trunk/chess/chessreduce.cc
Fork rungames for 'chessreduce', and refactor some shared code.

@@ -17,6 +17,7 @@
 #include "gamestats.h"
 #include "bigchess.h"
 
+
 using namespace std;
 using int64 = int64_t;
 using uint64 = uint64_t;
@@ -28,31 +29,6 @@ static constexpr int64 MAX_GAMES = 0LL;
 
 // #define SELF_CHECK true
 #undef SELF_CHECK
-
-
-static constexpr const char *const PIECE_NAME[32] = {
-  "a8 rook",
-  "b8 knight",
-  "c8 bishop",
-  "d8 queen",
-  "e8 king",
-  "f8 bishop",
-  "g8 knight",
-  "h8 rook",
-  "a7 pawn", "b7 pawn", "c7 pawn", "d7 pawn",
-  "e7 pawn", "f7 pawn", "g7 pawn", "h7 pawn",
-  // white
-  "a2 pawn", "b2 pawn", "c2 pawn", "d2 pawn",
-  "e2 pawn", "f2 pawn", "g2 pawn", "h2 pawn",
-  "a1 rook",
-  "b1 knight",
-  "c1 bishop",
-  "d1 queen",
-  "e1 king",
-  "f1 bishop",
-  "g1 knight",
-  "h1 rook", };
-
 
 static void ReadLargePGN(const char *filename) {
   std::shared_mutex bad_games_m;
@@ -203,16 +179,6 @@ static void ReadLargePGN(const char *filename) {
       }
 
       const int bucket = bucket_hash & NUM_BUCKETS_MASK;
-      if (true) {
-	fprintf(stderr, "Fates:\n");
-	for (int i = 0; i < 32; i++) {
-	  fprintf(stderr, "%d (%s). %s on %c%c.\n",
-		  i, PIECE_NAME[i],
-		  (GameStats::DIED & gs.fates[i]) ? "DIED" : "Survived",
-		  'a' + (gs.fates[i] & 7),
-		  '1' + (7 - ((gs.fates[i] & GameStats::POS_MASK) >> 3)));
-	}
-      }
       stat_buckets[bucket].AddGame(gs);
     };
 
@@ -260,32 +226,16 @@ static void ReadLargePGN(const char *filename) {
 
   fprintf(stderr, "Done! Join threads...\n");
   work_queue.reset(nullptr);
-
-  for (int bucket = 0; bucket < NUM_BUCKETS; bucket++) {
-    const Stats &s = stat_buckets[bucket];
-    printf("%lld\n", s.num_games);
-    for (int i = 0; i < 32; i++) {
[…]

### 2019-01-26 trunk/nand/paper.tex
Cleanup, optimize, notes

flinf flops and NaN gates
flip FLOPs and NaN gates?

Some useful primitives:

        These are called minNum and maxNum in IEE 754-2008 (5.3.1, p19)

        "If exactly one argument is NaN, they return the other.
        If both are NaN they return NaN.

        NAND  0 1     MAX  nan inf
             +---         +-------
           0 |1 1     nan |nan inf
           1 |1 0     inf |inf inf

                    HYPOT  nan inf
                          +-------
                      nan |nan inf
                      inf |inf inf

        So if 0 = nan and 1 = inf, then MAX is OR.

        We also have AND, since only inf*inf (aka 1 AND 1)
        returns inf in that truth table.

        Is AND and OR together complete? (I don't think so?)
        Do we have NOT?

YES!

Here are some:
Min(-x, -1.0) + inf
hypot(nan, max(1/x, -inf))
inf - max(x, 1)
sqrt(copysign(inf, -x))

Now direct search for some gates:
NAND: inf - max(x + y, -inf)
AND: x + y
OR: hypot(x, y)
XOR: Abs(max(y, -inf) - max(x, -inf))
     Abs(min(y, -x) + max(y, -inf))
     (Wouldn't be surprised if there's a simpler one,
     but my search does not find it before it gets
     out of hand...)

NOR: -inf / max(y, max(x, -1))

And how about if we leave out min, max? They are certainly
reasonable since they are defined in IEEE 754, but perhaps
not the best joke since they are not as math-y as e.g. hypot

NOR: sqrt(copysign(-inf, -hypot(y, x)))
(Here, hypot is just computing OR.)

So another NOT is:
sqrt(copysign(-inf, -x))  (actually can be inf)

XOR: abs(copysign(-inf, y) - copysign(-inf, x))
OR: hypot(y, x)
AND: x + y
NAND: inf - copysign(-inf, x + y)

ok cool, but copysign kinda has the same problems as min,max.
maybe without that?

the database saturates with 36 entries if we remove
copysign and min, max. They are these:

[nan, nan, nan, nan] = NaN
[nan, nan, nan, ~inf] = Minus(Neg(y), x)
[nan, nan, nan, ~2.0] = Plus(Tanh(Minus(Neg(y), x)), -1)
[nan, nan, nan, ~1.0] = Tanh(Minus(Neg(y), x))
[nan, nan, nan, 0.0] = Pow(y, Neg(x))
[nan, nan, nan, 1.0] = Exp(Pow(y, Neg(x)))
[nan, nan, nan, 2.0] = Plus(Exp(Pow(y, Neg(x))), 1)
[nan, nan, nan, inf] = Plus(y, x)
[nan, nan, ~inf, ~inf] = Neg(y)
[nan, nan, ~2.0, ~2.0] = Plus(Tanh(Neg(y)), -1)
[nan, nan, ~1.0, ~1.0] = Tanh(Neg(y))
[nan, nan, ~0.0, ~0.0] = Div(-1, y)
[nan, nan, 1.0, 1.0] = Exp(Div(-1, y))
[nan, nan, 2.0, 2.0] = Minus(1, Tanh(Neg(y)))
[nan, nan, inf, inf] = y
[nan, ~inf, nan, ~inf] = Neg(x)
[nan, ~inf, ~inf, ~inf] = Neg(Hypot(y, x))
[nan, ~2.0, nan, ~2.0] = Plus(Tanh(Neg(x)), -1)
[nan, ~2.0, ~2.0, ~2.0] = Plus(Tanh(Neg(Hypot(y, x))), -1)
[nan, ~1.0, nan, ~1.0] = Tanh(Neg(x))
[nan, ~1.0, ~1.0, ~1.0] = Tanh(Neg(Hypot(y, x)))
[nan, ~0.0, nan, ~0.0] = Div(-1, x)
[nan, ~0.0, ~0.0, ~0.0] = Div(-1, Hypot(y, x))
[nan, 1.0, nan, 1.0] = Exp(Div(-1, x))
[nan, 1.0, 1.0, 1.0] = Exp(Div(-1, Hypot(y, x)))
[nan, 2.0, nan, 2.0] = Minus(1, Tanh(Neg(x)))
[nan, 2.0, 2.0, 2.0] = Minus(1, Tanh(Neg(Hypot(y, x))))
[nan, inf, nan, inf] = x
[nan, inf, inf, inf] = Hypot(y, x)
[~inf, ~inf, ~inf, ~inf] = NegInf
[~2.0, ~2.0, ~2.0, ~2.0] = Plus(-1, -1)
[~1.0, ~1.0, ~1.0, ~1.0] = -1
[0.0, 0.0, 0.0, 0.0] = 0
[1.0, 1.0, 1.0, 1.0] = 1
[2.0, 2.0, 2.0, 2.0] = Plus(1, 1)
[inf, inf, inf, inf] = Inf

### 2019-01-27 nand/paper.tex
just write the paper man

\documentclass[twocolumn]{article}
\usepackage[top=0.5in, left=0.45in, right=0.45in, bottom=0.5in]{geometry}

\usepackage{url}
% \usepackage{code}
% \usepackage{cite}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

\usepackage{nopageno}

\interfootnotelinepenalty=0

% lets me explicitly set a. or 1. etc. as enum label
\usepackage{enumitem}

\pagestyle{empty}

\usepackage{ulem}
% go back to italics for emphasis, though
\normalem

\usepackage{natbib}

\setlength{\footnotesep}{2em}

\newcommand\comment[1]{}
\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}

\newcommand\nan{\textsf{NaN}}
\renewcommand\inf{\textsf{inf}}

\newcommand\plusminus{\pm}

\begin{document}

% flInf flops?
\title{NaN gates and flip FLOPS}
\author{Dr.~Tom~Murphy~VII~Ph.D.\footnote{
Copyright \copyright\ 2019 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2019 with the
XXX JOKE
of the Association for Computational Heresy; 
{\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
\$-0.00 } }

\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
\newcommand\at{\ensuremath{\scriptstyle @}}

\date{1 April 2019}

\maketitle \thispagestyle{empty}

\begin{abstract}
what do I put here
\end{abstract}

\section*{Introduction}

Mathematics is fundamental to computer science, and the foundation of
mathematics is the real numbers; this is obvious from the name. One of
computing's dirtiest secrets, however, is that computers themselves
are not based on real numbers---rather, they are based on so-called
``ones'' and ``zeroes'' combined with ``logic gates'' simulated with
transistors. While this suffices for most practical purposes, it is
unsatisfying from a theoretical perspective.

Recently, some progress has been made by human geniuses on completely
replacing integer calculations with calculations on real
numbers\cite{fluint8}. While this removes many of the hacks present in
modern software, there are still many components of the computer (e.g.
RAM, registers, the {\it scroll lock} LED, a tiny USB-powered fan that
can cool you on hot summer days or during particularly strenuous
programming sessions) % XXX cite skymall etc?
that are not integer-based.

In this paper I give a new foundation for computing based solely on
real numbers. This ...

I begin with a brief reminder of the definition of real numbers,
although the reader is expected to be familiar as these are pretty
fundamental to everything. The approach of the paper is then to
identify a pair of real numbers that have nice properties
(Section~\ref{sec:distinguished}), and then to give mathematical
operations on these numbers that parallel the logical operations
typically used in the construction of computers
(Section~\ref{sec:logical}). I then discuss how these operations
can be implemented efficiently (Section~\ref{sec:representations}).
I conclude with some wild speculation.

\section{Real numbers}
The real numbers are described by IEEE 754, most recently revised in
2008\cite{ieee754}. Every real number has a sign, a mantissa, and an
exponent. Actually, this understates the elegance of real numbers,
since there are a number of numbers, such as \nan\ (``not a number'')
which are not of this form; \nan\ nas no sign nor mantissa nor
exponent. We also have \inf\ and $-\inf$, which do have a sign, but no
mantissa nor exponent. These are the infinite numbers that you get if
you count very high or very low. Excitingly, we also have both
positive and negative versions of 0.

The real numbers have an equality operation {\tt ==}. This operation
has some very exciting properties: It is not reflexive (\nan\ {\tt ==}
\nan\ is false), and does not obey substitution (for $+0$ {\tt ==} $-0$
is true, but $\sfrac{1}{+0}$ {\tt ==} $\sfrac{1}{-0}$ is false).

Real numbers are an absolute joy to work with.

\subsection{Choosing some distinguished values} \label{sec:distinguished}

use nan and inf because other real numbers can be ``easily confused''
with one another. And do we really want ordering relationships on
our fundamental particles, like $0 < 1$?

% really lean into the joke that we choose the "number" NAN

\nocite{fluint8}

ha okay now there is just notes

Some useful primitives:

        These are called minNum and maxNum in IEE 754-2008 (5.3.1, p19)

        "If exactly one argument is NaN, they return the other.
        If both are NaN they return NaN.

        NAND  0 1     MAX  nan inf
             +---         +-------
           0 |1 1     nan |nan inf
           1 |1 0     inf |inf inf

                    HYPOT  nan inf
                          +-------
                      nan |nan inf
                      inf |inf inf

        So if 0 = nan and 1 = inf, then MAX is OR.

        We also have AND, since only inf*inf (aka 1 AND 1)
        returns inf in that truth table.

        Is AND and OR together complete? (I don't think so?)
        Do we have NOT?

YES!

Here are some:
Min(-x, -1.0) + inf
hypot(nan, max(1/x, -inf))
inf - max(x, 1)
sqrt(copysign(inf, -x))

Now direct search for some gates:
NAND: inf - max(x + y, -inf)
AND: x + y
OR: hypot(x, y)
XOR: Abs(max(y, -inf) - max(x, -inf))
     Abs(min(y, -x) + max(y, -inf))
     (Wouldn't be surprised if there's a simpler one,
     but my search does not find it before it gets
     out of hand...)

NOR: -inf / max(y, max(x, -1))

And how about if we leave out min, max? They are certainly
reasonable since they are defined in IEEE 754, but perhaps
not the best joke since they are not as math-y as e.g. hypot

NOR: sqrt(copysign(-inf, -hypot(y, x)))
(Here, hypot is just computing OR.)

So another NOT is:
sqrt(copysign(-inf, -x))  (actually can be inf)

XOR: abs(copysign(-inf, y) - copysign(-inf, x))
OR: hypot(y, x)
AND: x + y
NAND: inf - copysign(-inf, x + y)

ok cool, but copysign kinda has the same problems as min,max.
maybe without that?

the database saturates with 36 entries if we remove
copysign and min, max. They are these:

[nan, nan, nan, nan] = NaN
[nan, nan, nan, ~inf] = Minus(Neg(y), x)
[nan, nan, nan, ~2.0] = Plus(Tanh(Minus(Neg(y), x)), -1)
[nan, nan, nan, ~1.0] = Tanh(Minus(Neg(y), x))
[nan, nan, nan, 0.0] = Pow(y, Neg(x))
[nan, nan, nan, 1.0] = Exp(Pow(y, Neg(x)))
[nan, nan, nan, 2.0] = Plus(Exp(Pow(y, Neg(x))), 1)
[nan, nan, nan, inf] = Plus(y, x)
[nan, nan, ~inf, ~inf] = Neg(y)
[nan, nan, ~2.0, ~2.0] = Plus(Tanh(Neg(y)), -1)
[nan, nan, ~1.0, ~1.0] = Tanh(Neg(y))
[nan, nan, ~0.0, ~0.0] = Div(-1, y)
[nan, nan, 1.0, 1.0] = Exp(Div(-1, y))
[nan, nan, 2.0, 2.0] = Minus(1, Tanh(Neg(y)))
[nan, nan, inf, inf] = y
[nan, ~inf, nan, ~inf] = Neg(x)
[nan, ~inf, ~inf, ~inf] = Neg(Hypot(y, x))
[nan, ~2.0, nan, ~2.0] = Plus(Tanh(Neg(x)), -1)
[nan, ~2.0, ~2.0, ~2.0] = Plus(Tanh(Neg(Hypot(y, x))), -1)
[nan, ~1.0, nan, ~1.0] = Tanh(Neg(x))
[nan, ~1.0, ~1.0, ~1.0] = Tanh(Neg(Hypot(y, x)))
[nan, ~0.0, nan, ~0.0] = Div(-1, x)
[nan, ~0.0, ~0.0, ~0.0] = Div(-1, Hypot(y, x))
[nan, 1.0, nan, 1.0] = Exp(Div(-1, x))
[nan, 1.0, 1.0, 1.0] = Exp(Div(-1, Hypot(y, x)))
[nan, 2.0, nan, 2.0] = Minus(1, Tanh(Neg(x)))
[nan, 2.0, 2.0, 2.0] = Minus(1, Tanh(Neg(Hypot(y, x))))
[nan, inf, nan, inf] = x
[nan, inf, inf, inf] = Hypot(y, x)
[~inf, ~inf, ~inf, ~inf] = NegInf
[~2.0, ~2.0, ~2.0, ~2.0] = Plus(-1, -1)
[~1.0, ~1.0, ~1.0, ~1.0] = -1
[0.0, 0.0, 0.0, 0.0] = 0
[1.0, 1.0, 1.0, 1.0] = 1
[2.0, 2.0, 2.0, 2.0] = Plus(1, 1)
[inf, inf, inf, inf] = Inf

IEEE 754, 3.3
minimal encoding size:
b, radix, 2
p, precision, ?
emax, maximum exponent, ?
emin = always 1 - emax

I deduce:
p = t + 1
    (because you get one implied bit by normalization)
t = 1
% using the table
emax = $2^{k-p-1} - 1$, which is $2^{4-2-1}-1 = 2^1-1 = 1$
bias = emax
emin = 1 - emax = 0

I think there's still some ambiguity about whether we have
emax = 1, emin = 0,
or
emax = 0, emin = -1,
probably both work, but I think the former has more support
in the docs.

\comment{
a normal value is
(-1)^sign x b^{exp} x significand
(I think significand = mantissa?)

by standard, need:
- sign bit
- exponent could maybe be empty, though it's sort of implied
that you just have emax=0, emin= -1, so one bit
- m could probably be empty (no bits mantissa = 0)

DID YOU KNOW?
``every finite floating point numbe is an integral multiple
of the smallest subnormal magnitude b^{emin} * b^{1 - p}''

fewest exponent bits w?
The range of the encoding's biased exponent E shall include:
- every integer between 1 and 2^{w}−2, inclusive, to encode normal numbers
- the reserved value 0 to encode +/-0 and subnormal numbers
- the reserved value 2^{w}−1 to encode +/-inf and NaNs

so we need at least 0,
and a reserved value for inf & nan,
so if w = 1, we'd have 0 and 2^1-1 = 1,
but this seems to violate the spec because you need
``every integer between 1 and 2^w-2 inclusive'', and here 2^w-2 is 0,
so either this interval is inverted (arguable?) and that's why it
doesn't include its endpoints, or it's in violation
but with w = 2,

2^2 - 1 = 0b11 = 3 (inf, nan)
1 to 2^2 - 2 = 0b01, 0b10 = 1,2 (normals)
0 = 0b00 = 0 (zero, subnormals)

(Note this is the biased exponent E, not e)

2 bits looks clearly legal.

 s       w bits           t bits
[sign][ exponent ][  trailing significand T   ]

when e=0b11, t determines inf or nan.
- nan is any nonzero sign is ignored.
(recall, nans can carry info, so that is why they get
the nonzero case)
- inf is zero

so we clearly need 1 t-bit to encode inf vs nan,
but it seems that this is enough

so probable case is four bits:

 - sign
 - two exponent bits
 - one mantissa bit

The sign is always 0 (positive infinity; doesn't matter for nan)
and the exponent is always 11 (since we always encode nan or
positive inf). Unless we do any intermediate calculations
(could check what all the intermediate values are in the above
expressions and make sure they fit in 4 bits), a hardware
realization of this could just hard-wire 3 of 4 bits? Is that
better or worse?

so another version of this joke is that a NAND gate is a super
fast FPU for calculating inf - max(x + y, -inf) on four-bit
IEEE 754 where the top 3 MSBs are fixed (although be careful,
nan = 1 and inf = 0 in this representation. Is that maybe a
better choice for this presentation? it just dualizes some
gates)

}

\section{The binary4 representation}

IEEE 754 natively defines several bit widths for floating-point
values, such as the 32-bit binary32 (aka ``single-precision float'')
and 64-bit binary64 (aka ``double-precision float''). The
specification is parameterized to allow other bit widths; for example,
half-precision 16-bit floats are common in GPU code for machine
learning applications (XXX cite). Smaller floats sacrifice precision,
but require less space and allow faster calculations. For our purposes
in this paper, since we only need to represent the two values \nan\ and
\inf, we are interested in the smallest possible representation.

% TODO: what if each bit in the float is itself represented as
% inf or NaN?

This section describes the binary4 representation, a four-bit floating
point number that is clearly allowed by the IEEE 754 standard.

The representation of any floating-point number has a single sign bit,
some number $w$ of exponent bits, and some number $t$ of mantissa
bits. For binary32, $w = 8$ and $t = 23$; and with the sign bit we
have $23 + 8 + 1 = 32$ bits as expected. We need at least a sign bit,
but what are the smallest permissible values of $w$ and $t$?

The most stringent constraint on $w$ comes in IEEE-754-2008~3.4, which
states
\begin{quote}
  The range of the encoding's biased exponent E shall include:
  \begin{enumerate}[label=---]
    \item every integer between 1 and $2^w - 2$, inclusive, to encode
      normal numbers
    \item the reserved value 0 to encode $\plusminus 0$ and subnormal
      numbers
    \item the reserved value $2^w - 1$ to encode $\plusminus \infty$
      and NaNs.
  \end{enumerate}
\end{quote}

$E$ is the binary number encoded by $w$. It must include at least the
two special values consisting of all zeroes and all ones (second and
third clause). A conservative reading of ``every integer between 1 and
$2^w - 2$'' seems to require that $1 \leq 2^w - 2$ (otherwise how
could the interval be inclusive of its endpoints?), which would imply
that $w$ is at least 2. (However, see Section~\ref{sec:binary3} for
the hypothesized case where $w=1$.)

The representation of \nan\ and \inf\ are distinguished by the value
of $t$ when $E$ is all ones. We certainly need to distinguish these,
so $t = 1$ is the minimal size.

We have one sign bit, two exponent bits, and one mantissa bit, for a
total of four. Since ``single precision'' is 32 bits, ``half
precision'' is 16, 4 bits is ``eighth precision.'' Given how nicely
all this works out, shouldn't there be a {\tt eighth} base type in
most modern programming languages and GPUs? Since there are so few
values representable, it would be practical for all the standard
operations to be done in constant time via table lookups. All 16
possible values are given in Figure~\ref{fig:binary4}.

\begin{figure*}[tp]
\begin{tabular}{|l|c|c|c|c|c|c|}
  \hline
  parameter & {\bf binary4} & binary16 & binary32 & binary64 & binary128 & binary$_k$ \\
  \hline
  $k$, storage in bits & {\bf 4} & 16 & 32 & 64 & 128 & multiple of 32 \\
  $p$, precision in bits & {\bf 2} & 11 & 24 & 53 & 113 & $k$ - round(4 * log$_2$(k)) + 13 \\
  $emax$, maximum exponent e & {\bf 1} & 15 & 127 & 1023 & 16383 & $2^{k-p-1}$ - 1 \\
  $bias$ = $E$ - $e$ & {\bf 1} & 15 & 127 & 1023 & 16383 & emax \\
  $sign bits$ & {\bf 1} & 1 & 1 & 1 & 1 & 1 \\
  $w$, exponent width & {\bf 2} & 5 & 8 & 11 & 15 & round(4 * log$_2$(k)) - 13 \\
  $t$, trailing significand width & {\bf 1} & 10 & 23 & 52 & 112 & k - w - 1 \\
  $k$, storage width & {\bf 4} & 16 & 32 & 64 & 128 & 1 + w + t \\
\hline
\end{tabular}
\caption{Parameters for the newly-introduced {\bf binary4} encoding
  for IEEE-754, compared to the standard widths (see Table~3.5 in the
  standard\cite{ieee754}). } \label{fig:binary4}
\end{figure*}

% bias = 1

\begin{figure}[h]
\begin{tabular}{|l@{\,}c@{\,}l|p{2.5in}|}
\hline
  $s$ & $E$ & $T$ & value \\
  \hline
0 & 00 & 0 &   $+0$ \\
0 & 00 & 1 &    subnormal: % $2^{emin} * 2^{1-p} * T$ =
                $2^0 * 2^{1-2} * 1$ =
                $1 * \sfrac{1}{2} * 1$ = 0.5 \\
0 & 01 & 0 &    normal: % $2^{E - bias} * (1 + 2^{1-p} * T)$
                $2^0 * (1 + \sfrac{1}{2} * 0)$ = 1 \\
0 & 01 & 1 &    $2^0 * (1 + \sfrac{1}{2} * 1)$ = 1.5 \\
0 & 10 & 0 &    $2^1 * (1 + \sfrac{1}{2} * 0)$ = 2 \\
0 & 10 & 1 &    $2^1 * (1 + \sfrac{1}{2} * 1)$ = 3 \\
0 & 11 & 0 &   $+\infty$ \\
0 & 11 & 1 &    \nan \\
1 & 00 & 0 &   $-0$ \\
1 & 00 & 1 &   $-0.5$ \\
1 & 01 & 0 &   $-1$ \\ 
1 & 01 & 1 &   $-1.5$ \\
1 & 10 & 0 &   $-2$ \\
1 & 10 & 1 &   $-3$ \\ 
1 & 11 & 0 &   $-\infty$ \\
1 & 11 & 1 &    \nan \\
\hline
\end{tabular}
\caption{All 16 values representable in binary4 floating-point.
  The format works reasonably well even at this very low precision,
  although note how many of the values are not finite.} \label{fig:allvalues4}
\end{figure}

\subsection{The hypothesized binary3 format} \label{sec:binary3}
The IEEE-754 representation clearly requires a sign bit, and for this
purpose we need at least one bit for the mantissa in order to
distinguish \nan\ and \inf. It is perhaps a stretch of the wording,
but arguably the spec permits a 1-bit exponent ($w = 1$). To
rationalize this we need to interpret the phrase ``every integer
between $1$ and $2^w-2$ inclusive'' (that is, between 1 and 0 inclusive)
as denoting the empty set.

With one bit for sign, exponent, and mantissa, we can represent just 8
different values. Here $emax$ is $0$, and the standard clearly
requires $emin = 1 - emax$, so $emin = 1$. Certainly fishy for $emin$
to be larger than $emax$, but we can just not stress out about it; the
representable values are all reasonable-looking
(Figure~\ref{fig:allvalues3}).

% here, precision is still 2 (just one mantissa bit)
% so emax is 2^{k-p-1}-1, which is 2^{3-2-1}-1 = 0
% bias = 0
% emin is 1 - emax = 1 (!?)
% this is kind of absurd, it may still "work"?

\begin{figure}[h]
\begin{tabular}{|l@{\,}c@{\,}l|p{2.5in}|}
\hline
  $s$ & $E$ & $T$ & value \\
  \hline
0 & 0 & 0 &   $+0$ \\
0 & 0 & 1 &    subnormal: % $2^{emin} * 2^{1-p} * T$ =
                $2^1 * 2^{1-2} * 1$ =
                $2 * \sfrac{1}{2} * 1$ = 1 \\
0 & 1 & 0 &   $+\infty$ \\
0 & 1 & 1 &    \nan \\
1 & 0 & 0 &   $-0$ \\
1 & 0 & 1 &   $-1$ \\
1 & 1 & 0 &   $-\infty$ \\
1 & 1 & 1 &    \nan \\
\hline
\end{tabular}
\caption{All 8 values of the hypothetical binary3 representation.
  There are no normal values; the only finite values are the
  positive and negative zero and a single subnormal which
  denotes $1$ (or $-1$).
} \label{fig:allvalues3}
\end{figure}

% TODO: We could do without a mantissa if we didn't care about
% distinguishing NaN and inf. With an empty t, we should
% consider it to have value 0 (true? maybe it is defined as
% a product so the empty string is the multiplicative unit 1?)
% which means that we encode +/- inf, and nan cannot be encoded.
% Now the operations will not be closed since e.g. 0/0 needs to
% result in nan.

% TODO: What if we have no mantissa or exponent, just a sign bit?
% When E=0 and T=0, we can only encode +/- zero. This definitely
% involves some leaps (and is not closed as above), but could
% we find expressions that work for just +/- 0?

\nocite{ieee754}
\bibliography{nand}{}
\bibliographystyle{plain}

\end{document}

### 2019-01-30 chess/papers/survival.tex
Start on tournament wrapper

@@
 \author{Dr.~Tom~Murphy~VII~Ph.D.\footnote{
-Copyright \copyright\ 2018 the Regents of the Wikiplia
-Foundation. Appears in SIGBOVIK 2018 with the threefold
+Copyright \copyright\ 2019 the Regents of the Wikiplia
+Foundation. Appears in SIGBOVIK 2019 with the threefold
 repetition of the Association for Computational Heresy; 
@@
 
-\date{1 April 2018}
+\date{1 April 2019}
 
@@
   forbid the players from agreeing to a draw verbally before a certain
-  point (e.g. 30 moves). There are always other routes to a draw, for
-  example by stalemate or repeating the same position three times.
+  point (e.g. 30 moves), or without the arbiter's consent, and FIDE
+  rules technically do not allow a draw until both players have made a
+  move (5.2.3). There are always other routes to a draw, for example
+  by stalemate or repeating the same position three times.
   Collaboratively producing such situations is easy, but this strategy

### 2019-01-30 trunk/chess/bigchess.h
Start on tournament wrapper

@@ -5,8 +5,15 @@
 
 #include <shared_mutex>
 #include <cstdint>
+#include <thread>
+#include <deque>
+#include <vector>
+
+#include "../cc-lib/base/logging.h"
+
 
 using int64 = int64_t;
+using uint8 = uint8_t;
 
 template<class K, class C>
 inline bool ContainsKey(const C &container, const K &key) {
@@ -81,7 +88,7 @@ struct WorkQueue {
 	  } else {
 	    if /* constexpr */ (max_chunk > 1LL) {
 	      int64 get_up_to = std::max(1LL, pending / this->num_workers);
-	      vector<W> local_work;
+	      std::vector<W> local_work;
 	      local_work.reserve(get_up_to);
 	      while (!todo.empty() && get_up_to--) {
 		local_work.emplace_back(todo.front());
@@ -180,7 +187,7 @@ struct PGNTextStream {
     CHECK(f != nullptr) << filename;
   }
 
-  bool NextPGN(string *pgn_text) {
+  bool NextPGN(std::string *pgn_text) {
     if (f == nullptr)
       return false;
   

### 2019-01-30 trunk/chess/chess.h

@@ -139,6 +139,17 @@ struct Position {
 
   // Assuming the move is legal, is it an en passant capture?
   bool IsEnPassant(Move m) const;
+
+  // The following two are used in 50/75-move draw rules.
+  // Assuming the move is legal, is it a pawn move?
+  bool IsPawnMove(Move m) const {
+    return (PieceAt(m.src_row, m.src_col) & TYPE_MASK) == PAWN;
+  }
+  // Assuming the move is legal, is it a capturing move?
+  bool IsCapturing(Move m) const {
+    return IsEnPassant(m) ||
+      PieceAt(m.dst_row, m.dst_col) != EMPTY;
+  }
   
   // Is the move legal in this current board state? The move must be
   // well-formed (positions within the board).
@@ -179,11 +190,15 @@ struct Position {
     }
   }
   
+  // True if the current player is in check.
+  bool IsInCheck();
+
   // True if the current player is checkmated.
   bool IsMated();
 
   // Return true if there are any legal moves. This is faster than
-  // enumerating them all.
+  // enumerating them all. If there are none, then the player is
+  // either checkmated (IsInCheck) or stalemated.
   bool HasLegalMoves();
   // Get all the legal moves.
   std::vector<Move> GetLegalMoves();
@@ -197,7 +212,7 @@ struct Position {
   // on that square, not considering check. Does not include en passant.
   // The square is typically unoccupied.
   bool Attacked(int r, int c) const;
-  
+
   template<class F>
   auto SetExcursion(int r, int c, uint8 piece, const F &f) -> decltype(f()) {
     const uint8 old = PieceAt(r, c);
@@ -207,20 +222,30 @@ struct Position {
     return ret;
   }
 
-  #if 0
   // Apply the move to the current board state, and execute the
   // function with that state applied. Return the return value
   // of the function after undoing the applied move. As above,
   // the move must be legal.
   template<class F>
   auto MoveExcursion(Move m, const F &f) -> decltype(f()) {
-    // TODO
-    return f();
+    // PERF: Modify in place.
+    Position copy = *this;
+    ApplyMove(m);
+    auto ret = f();
+    *this = copy;
+    return ret;
   }
-  #endif
 
   bool BlackMove() const {
     return !!(bits & BLACK_MOVE);
+  }
+
+  void SetBlackMove(bool black_move) {
+    if (black_move) {
+      bits |= BLACK_MOVE;
+    } else {
+      bits &= ~BLACK_MOVE;
+    }
   }
   
  private:

### 2019-01-30 trunk/chess/player.cc

@@ -1,0 +1,235 @@
+
+#include "player.h"
+
+#include <string>
+#include <mutex>
+#include <cstdint>
+
+#include "../cc-lib/arcfour.h"
+#include "../cc-lib/randutil.h"
+#include "../cc-lib/base/logging.h"
+#include "../cc-lib/base/stringprintf.h"
+
+using namespace std;
+using int64 = int64_t;
+using Move = Position::Move;
+
+namespace {
+
+template<class T, class F>
+static const T &GetBest(const std::vector<T> &v, F f) {
+  CHECK(!v.empty());
+  int best_i = 0;
+  for (int i = 1; i < v.size(); i++) {
+    // f is <, so this means a strict improvement
+    if (f(v[i], v[best_i])) {
+      best_i = i;
+    }
+  }
+  return v[best_i];
+}
+
+#if 0
+static bool IsMating(Position *pos, const Move &m) {
+  return pos->MoveExcursion(m, [pos]() { return pos->IsMated(); });
+}
+
+static bool IsChecking(Position *pos, const Move &m) {
+  return pos->MoveExcursion(m, [pos]() { return pos->IsInCheck(); });
+}
+#endif
+
+struct FirstMovePlayer : public Player {
+
+  // TODO: Flip this when playing as black, right?
+  static int GetCode(const Move &m) {
+    int src = (int)m.src_row * 8 + (int)m.src_col;
+    int dst = (int)m.dst_row * 8 + (int)m.dst_row;
+    return (src * 64 + dst) * 8 + (int)m.promote_to;
+  }
+  
+  virtual Move MakeMove(const Position &orig_pos) override {
+    Position pos = orig_pos;
+    std::vector<Move> legal = pos.GetLegalMoves();
+
+    return GetBest(legal, 
+		   [](const Move &a, const Move &b) {
+		     return GetCode(a) < GetCode(b);
+		   });
+  }
+  
+  const char *Name() const override { return "first_move"; }
+  const char *Desc() const override {
+    return "Makes the lexicographically first legal move.";
+  }
+};
+
+static std::mutex seed_m;
+static string GetSeed() {
+  static int64 counter = 0LL;
+  int64 c = 0LL;
+  seed_m.lock();
+  c = counter++;
+  seed_m.unlock();
+
+  int64 t = time(nullptr);
+  return StringPrintf("s%lld.%lld", c, t);
+}
+
+struct RandomPlayer : public Player {
+  RandomPlayer() : rc(GetSeed()) {
[…]

### 2019-01-30 trunk/chess/tournament.cc

@@ -1,0 +1,133 @@
+
+// Run a tournament of players to determine their Elo ratings.
+
+#include <memory>
+
+#include "../cc-lib/base/logging.h"
+
+#include "chess.h"
+#include "player.h"
+
+#define TESTING true
+
+using Move = Position::Move;
+using namespace std;
+
+// Under FIDE rules, after 50 moves without a pawn move or capture, a
+// player may CLAIM a draw. We don't allow these computer players to
+// claim draws, so the relevant rule is actually 9.6.2, where the game
+// is automatically a draw after *75* such moves.
+// Similarly, threefold repetition allows a claim, and fivefold
+// repetition forces the draw.
+
+enum class Result {
+  WHITE_WINS,
+  BLACK_WINS,
+  // Different forms of draw.
+  DRAW_STALEMATE,
+  DRAW_75MOVES,
+  // DRAW_REPETITION,
+  // DRAW_INSUFFICIENT,
+};
+
+Result PlayGame(Player *white, Player *black) {
+  Position pos;
+
+  int white_stale_moves = 0;
+  int black_stale_moves = 0;
+
+  // TODO: Draw by five instances of the same position.
+  // TODO: Draw by insufficient material. (But we can
+  // be guaranteed that these eventually end in draws by
+  // the 75-move rule.)
+  
+  auto DrawByRule =
+    [&white_stale_moves,
+     &black_stale_moves]() {
+      return (white_stale_moves >= 75 &&
+	      black_stale_moves >= 75);
+    };
+  
+  for (;;) {
+    if (TESTING) { CHECK(!pos.BlackMove()); }
+
+    {
+      Move m = white->MakeMove(pos);
+      if (TESTING) { CHECK(pos.IsLegal(m)); }
+      if (pos.IsCapturing(m) ||
+	  pos.IsPawnMove(m)) {
+	white_stale_moves = 0;
+      } else {
+	white_stale_moves++;
+      }
+
+      pos.ApplyMove(m);
+    
+      // Checkmate takes precedence over draw by
+      // 75 moves.
+
+      if (!pos.HasLegalMoves()) {
+	if (pos.IsInCheck())
+	  return Result::WHITE_WINS;
+
+	return Result::DRAW_STALEMATE;
+      }
+
+      if (DrawByRule())
+	return Result::DRAW_75MOVES;
[…]

### 2019-02-09 chess/chess.cc
Add huddle and swarm players

@@ -1423,6 +1423,9 @@ struct CountMC {
   int count = 0;
 };
 
+// If true, movegen stops iterating moves after it reaches another
+// piece. On the macbook this was worse, but on mingw/threadripper it
+// is faster.
 #define MOVEGEN_STOP_PIECE true
 
 template<class MoveCollector>
@@ -1504,17 +1507,53 @@ static void CollectLegalMoves(Position &pos, MoveCollector &mc) {
 
 	  // Horizontal.
 	  m.dst_row = srcr;
+	  #if MOVEGEN_STOP_PIECE
+	  // Left.
+	  for (int dstc = srcc - 1; dstc >= 0; dstc--) {
+	    m.dst_col = dstc;
+	    TRY_MOVE(m);
+	    if (pos.PieceAt(srcr, dstc) != Position::EMPTY)
+	      break;
+	  }
+	  // Right.
+	  for (int dstc = srcc + 1; dstc < 8; dstc++) {
+	    m.dst_col = dstc;
+	    TRY_MOVE(m);
+	    if (pos.PieceAt(srcr, dstc) != Position::EMPTY)
+	      break;
+	  }
+
+          #else
 	  for (int dstc = 0; dstc < 8; dstc++) {
 	    m.dst_col = dstc;
 	    TRY_MOVE(m);
 	  }
+	  #endif
 
 	  // Vertical.
 	  m.dst_col = srcc;
+	  #if MOVEGEN_STOP_PIECE
+	  // Left.
+	  for (int dstr = srcr - 1; dstr >= 0; dstr--) {
+	    m.dst_row = dstr;
+	    TRY_MOVE(m);
+	    if (pos.PieceAt(dstr, srcc) != Position::EMPTY)
+	      break;
+	  }
+	  // Right.
+	  for (int dstr = srcr + 1; dstr < 8; dstr++) {
+	    m.dst_row = dstr;
+	    TRY_MOVE(m);
+	    if (pos.PieceAt(dstr, srcc) != Position::EMPTY)
+	      break;
+	  }
+
+	  #else
 	  for (int dstr = 0; dstr < 8; dstr++) {
 	    m.dst_row = dstr;
 	    TRY_MOVE(m);
 	  }
+	  #endif
 
 	  break;
 	  
@@ -1568,11 +1607,29 @@ static void CollectLegalMoves(Position &pos, MoveCollector &mc) {
 
[…]

### 2019-02-09 chess/player.cc

@@ -416,6 +416,64 @@ struct ReverseStartingPlayer : public EvalResultPlayer {
   }
 };
 
+struct HuddlePlayer : public EvalResultPlayer {
+  int64 PositionPenalty(Position *pos) override {
+    // (Reversed since we are on the opposite move now.)
+    const bool black = !pos->BlackMove();
+    const uint8 my_mask = black ? Position::BLACK : Position::WHITE;
+
+    int kr, kc;
+    std::tie(kr, kc) = pos->GetKing(black);
+    
+    int64 dist = 0LL;
+    for (int r = 0; r < 8; r++) {
+      for (int c = 0; c < 8; c++) {
+	uint8 piece = pos->PieceAt(r, c);
+	if (piece != Position::EMPTY &&
+	    (piece & Position::COLOR_MASK) == my_mask) {
+	  dist += std::max(std::abs(r - kr),
+			   std::abs(c - kc));
+	}
+      }
+    }
+    return dist;
+  }
+  
+  string Name() const override { return "huddle"; }
+  string Desc() const override {
+    return "Try to move pieces to surround our own king.";
+  }
+};
+
+struct SwarmPlayer : public EvalResultPlayer {
+  int64 PositionPenalty(Position *pos) override {
+    // (Reversed since we are on the opposite move now.)
+    const bool black = !pos->BlackMove();
+    const uint8 my_mask = black ? Position::BLACK : Position::WHITE;
+
+    // Same as Huddle, but target the enemy's king.
+    int kr, kc;
+    std::tie(kr, kc) = pos->GetKing(!black);
+    
+    int64 dist = 0LL;
+    for (int r = 0; r < 8; r++) {
+      for (int c = 0; c < 8; c++) {
+	uint8 piece = pos->PieceAt(r, c);
+	if (piece != Position::EMPTY &&
+	    (piece & Position::COLOR_MASK) == my_mask) {
+	  dist += std::max(std::abs(r - kr),
+			   std::abs(c - kc));
+	}
+      }
+    }
+    return dist;
+  }
+  
+  string Name() const override { return "swarm"; }
+  string Desc() const override {
+    return "Try to move pieces to surround the opponent's king.";
+  }
+};
 
 struct GenerousPlayer : public EvalResultPlayer {
   int64 PositionPenalty(Position *p) override {
@@ -734,7 +792,6 @@ struct SinglePlayerPlayer : public Player {
[…]

### 2019-02-09 chess/player.h

@@ -35,6 +35,8 @@ Player *CreatePacifist();
 Player *CreateMinOpponentMoves();
 Player *CreateSuicideKing();
 Player *CreateReverseStarting();
+Player *CreateHuddle();
+Player *CreateSwarm();
 Player *CreateGenerous();
 Player *CreateNoIInsist();
 Player *CreateSameColor();

### 2019-02-09 trunk/chess/papers/weak.tex
Add huddle and swarm players

% Dramatic failure of the "suicide king" strategy against
% CCCP.
1. d4 g5 2. Bxg5 Nc6 3. Bxe7 Kxe7 4. d5 Kd6
5. dxc6+ Kc5 6. Qd6+ Kc4 7. e4#

% miraculous win for "generous" strategy as black:
% playing against "suicide king". White forces black
% to mate him:
1. a4 b5 2. d3 Nh6 3. Kd2 Ng8 4. Kc3 Nh6
5. Kd4 c6 6. Ke5 f5 7. a5 Qb6 8. Bg5 Qe3+
9. Bxe3 Na6 10. f3 Nc5 11. Ra2 Ne4 12. f4 Nd2
13. Bf2 Nf3+ 14. exf3 Ng4+ 15. Kxf5 g5 16. Nh3 e5
17. Bh4 Ba3 18. b4 e4 19. Nf2 c5 20. c4 Ba6
21. g3 Rc8 22. Be2 Kf7 23. fxe4 d5 24. Qc1 dxe4
25. Rg1 bxc4 26. Rd1 cxd3 27. Qxc5 Rhf8 28. Nd2 Bc4
29. Kxg5 Rc7 30. Bf3 Ke8 31. Qe5+ Kf7 32. Qe6+ Kg7
33. Qb6 Rxf4 34. Nb3 Bb5 35. Qb7 d2 36. Nh3 Bc1
37. Qxa7 Be2 38. Nd4 h5 39. Ra1 Nf2 40. Raxc1 Nh1
41. Rf1 d1=Q 42. Qxc7+ Kh8 43. Kg6 Bc4 44. Bxd1 e3
45. Qf7 Bb3 46. Rf3 Nf2 47. Rc6 e2 48. Rc7 Rf5
49. Rc4 Rf4 50. Nb5 e1=Q 51. Rc2 Qc3 52. g4 Rf5
53. Ng1 Qc7 54. Rcc3 Nd3 55. Rc5 Nf2 56. Nc3 Ba2
57. Qh7+

// Good site for some computer chess programs with ELO:
// http://www.computerchess.org.uk/ccrl/404/

// citation for knights distance:
// Counting the Number of Squares Reachable in k
// Knight's Moves
// Amanda M. Miller, David L. Farnsworth
// Open Journal of Discrete Mathematics, 2013, 3, 151-154
// http://dx.doi.org/10.4236/ojdm.2013.33027

### 2019-02-09 trunk/chess/bench.cc

@@ -11,6 +11,7 @@
 
 #include "chess.h"
 #include "player.h"
+#include "blind/timer.h"
 
 using Move = Position::Move;
 using namespace std;
@@ -126,7 +127,8 @@ static void RunBenchmark() {
   int white_wins = 0, black_wins = 0, draws = 0;
   int total_moves = 0;
 
-  const int64 start_time = time(nullptr);
+  // const int64 start_time = time(nullptr);
+  Timer bench_timer;
   for (int game = 0; game < NUM_GAMES; game++) {
     vector<Move> moves;
     switch (PlayGame(white.get(), black.get(), &moves)) {
@@ -143,7 +145,8 @@ static void RunBenchmark() {
     }
     total_moves += moves.size();
   }
-  const double seconds = time(nullptr) - start_time;
+  // const double seconds = time(nullptr) - start_time;
+  const double seconds = bench_timer.MS() / 1000.0;
   printf("Total time: %.2fs (%.2f moves/sec)\n",
 	 seconds, total_moves / seconds);
 }

### 2019-02-09 trunk/chess/tournament.cc

@@ -13,7 +13,7 @@
 #include "chess.h"
 #include "player.h"
 // #include "stockfish.h"
-// #include "stockfish-player.h"
+#include "stockfish-player.h"
 
 #define TESTING true
 
@@ -23,8 +23,8 @@ using namespace std;
 // Number of round-robin rounds.
 // (Maybe should be based on the total number of games we want
 // to simulate?)
-static constexpr int TOTAL_ROUNDS = 40; // 250;
-static constexpr int THREADS = 1;
+static constexpr int TOTAL_ROUNDS = 108; // 250;
+static constexpr int THREADS = 54;
 static constexpr int ROUNDS_PER_THREAD = TOTAL_ROUNDS / THREADS;
 
 typedef Player *(*Entrant)();
@@ -34,7 +34,7 @@ const vector<Entrant> &GetEntrants() {
     new vector<Entrant>{
 			// CreateWorstfish,
 			// CreateRandom,
-			CreateFirstMove,
+			// CreateFirstMove,
 			// CreateAlphabetical,
 			// CreateCCCP,
 			// CreateNoIInsist,
@@ -45,11 +45,13 @@ const vector<Entrant> &GetEntrants() {
 			CreateSymmetry180,
 			// CreateOppositeColor,
 			// CreateMinOpponentMoves,
-			CreateSuicideKing,
+			// CreateSuicideKing,
 			// CreateReverseStarting,
-			CreateGenerous,
+			CreateHuddle,
+			CreateSwarm,
+			// CreateGenerous,
 			// CreateSinglePlayer,
-			// CreateStockfish0,
+			CreateStockfish0,
 			// CreateStockfish5,
 			// CreateStockfish10,
 			// CreateStockfish15,
@@ -202,12 +204,12 @@ static void TournamentThread(int thread_id,
 	  vector<Move> as_white;
 	  switch (PlayGame(entrants[row], entrants[col], &as_white)) {
 	  case Result::WHITE_WINS:
-	    if (cell->row_wins == 0)
+	    if (cell->example_white_win.empty())
 	      cell->example_white_win = as_white;
 	    cell->row_wins++;
 	    break;
 	  case Result::BLACK_WINS:
-	    if (cell->row_losses == 0)
+	    if (cell->example_white_loss.empty())
 	      cell->example_white_loss = as_white;
 	    cell->row_losses++;
 	    break;
[…]

### 2019-02-10 chess/chessmaster.h
Better progress updates in Tournament. About to do incremental rewrite

@@ -47,5 +47,7 @@ private:
 
 // Level 1. Makes random move if chessmaster fails for some reason.
 Player *CreateChessmaster1();
+// Level 2.
+Player *CreateChessmaster2();
 
 #endif

### 2019-02-11 trunk/chess/tournament.cc
Tournament now accumulates results to database. Separated elo/table code out into its own program.

@@ -22,6 +22,7 @@
 #include "stockfish-player.h"
 #include "uci-player.h"
 #include "chessmaster.h"
+#include "tournament-db.h"
 
 #define TESTING true
 
@@ -46,8 +47,8 @@ using namespace std;
 // Number of round-robin rounds.
 // (Maybe should be based on the total number of games we want
 // to simulate?)
-static constexpr int TOTAL_ROUNDS = 50; // 250;
-static constexpr int THREADS = 50;
+static constexpr int TOTAL_ROUNDS = 108; // 250;
+static constexpr int THREADS = 54;
 static constexpr int ROUNDS_PER_THREAD = TOTAL_ROUNDS / THREADS;
 
 typedef Player *(*Entrant)();
@@ -55,29 +56,29 @@ typedef Player *(*Entrant)();
 const vector<Entrant> &GetEntrants() {
   static vector<Entrant> *entrants =
     new vector<Entrant>{
-			// CreateWorstfish,
+			CreateWorstfish,
 			CreateRandom,
-			CreateFirstMove,
-			CreateAlphabetical,
+			// CreateFirstMove,
+			// CreateAlphabetical,
 			CreateCCCP,
 			CreateNoIInsist,
 			CreatePacifist,
 			CreateSameColor,
-			CreateMirrorYSymmetry,
-			CreateMirrorXSymmetry,
+			// CreateMirrorYSymmetry,
+			// CreateMirrorXSymmetry,
 			CreateSymmetry180,
 			CreateOppositeColor,
 			CreateSuicideKing,
 			CreateReverseStarting,
-			CreateMinOpponentMoves,
-			CreateHuddle,
+			// CreateMinOpponentMoves,
+			// CreateHuddle,
 			CreateSwarm,
 			CreateGenerous,
 			// CreateTopple1M,
 			// CreateTopple10K,
 			// CreateSinglePlayer,
-			CreateChessmaster1,
-			CreateChessmaster2,
+			// CreateChessmaster1,
+			// CreateChessmaster2,
 			// CreateStockfish0,
 			// CreateStockfish5,
 			// CreateStockfish10,
@@ -200,12 +201,6 @@ Result PlayGame(Player *white, Player *black, vector<Move> *moves) {
   }
 }
 
-struct Cell {
-  int row_wins = 0, row_losses = 0, draws = 0;
-  vector<Move> example_white_win, example_white_loss, example_white_draw;
[…]

### 2019-02-11 chess/elo.cc
Tournament now accumulates results to database. Separated elo/table code out into its own program.

@@ -1,0 +1,253 @@
+
+// Reads the tournament database and produces the html table with elo ratings.
+
+#include <mutex>
+#include <memory>
+#include <unordered_map>
+#include <cstdint>
+#include <vector>
+#include <string>
+
+#include "../cc-lib/base/logging.h"
+#include "../cc-lib/base/stringprintf.h"
+#include "../cc-lib/threadutil.h"
+#include "../cc-lib/util.h"
+#include "../cc-lib/arcfour.h"
+#include "../cc-lib/randutil.h"
+
+#include "chess.h"
+#include "tournament-db.h"
+
+using namespace std;
+
+using Move = Position::Move;
+using int64 = int64_t;
+
+static constexpr double ELO_START = 1000.0;
+
+struct Elo {
+  double elo = ELO_START;
+  // Used to determine k-factor.
+  int games = 0;
+  // Totals.
+  int wins = 0, losses = 0, draws = 0;
+};
+
+// With the win/loss/draw matrix, compute elo ratings for each player.
+static vector<Elo> ComputeElo(int num_entrants, const std::vector<Cell> &orig_outcomes) {
+  CHECK(orig_outcomes.size() == num_entrants * num_entrants);
+  // We don't need the examples, but it's easiest to just copy
+  // everything. We will modify wins/losses/draws to replay the games.
+  std::vector<Cell> outcomes = orig_outcomes;
+  
+  vector<Elo> elos;
+  elos.resize(num_entrants);
+  
+  ArcFour rc(StringPrintf("elo.%lld", (int64)time(nullptr)));
+  // Perform iterative updates in random order.
+  vector<pair<int, int>> nonempty_matchups;
+  // Initialize, ignoring self-play.
+  for (int white = 0; white < num_entrants; white++) {
+    for (int black = 0; black < num_entrants; black++) {
+      if (white != black) {
+	Cell *cell = &outcomes[white * num_entrants + black];
+	if (cell->white_wins != 0 ||
+	    cell->white_losses != 0 ||
+	    cell->draws != 0) {
+	  nonempty_matchups.emplace_back(white, black);
+	}
+      }
+    }
+  }
+
+  auto ClaimResult =
+    [&rc](Cell *cell) {
+      // Pick uniformly in proportion to the number of games
+      // played, not the three (non-empty) categories.
[…]

### 2019-02-14 trunk/chess/blind/getlegal.cc
Experiment to predict legal moves. Packed representation of moves. Mode for tournament where we only compute one row/col. Add stockfish node option and Stockfish1M player

@@ -1,0 +1,2267 @@
+
+// See unblind.cc for history.
+
+// In this experiment, I try to predict (from the blinded 64-bit
+// bitmask) which of the 64^2 = 4096 moves are legal. We represent
+// these as source/destination pairs (as in Position::Move),
+// assuming for simplicity that any promotion is promotion to Q.
+//
+// 4096 is a reasonable size for the output layer, but note that many
+// source-destination pairs are always impossible. A piece cannot move
+// to the square it's already occupying. No piece on A1 could ever
+// move to B8, B7, B6, etc. So we can put a bit less pressure on the
+// model by only asking it to learn these pairs. There are 1792 of
+// them. Pack(src_row, src_col, dst_row, dst_col) gives the canonical
+// move number in [0, 1792) and Unpack(idx) is the inverse.
+
+#include "SDL.h"
+#include "SDL_main.h"
+#include "../cc-lib/sdl/sdlutil.h"
+#include "../cc-lib/sdl/font.h"
+
+#include <CL/cl.h>
+
+#include <string.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <time.h>
+
+#include <cmath>
+#include <chrono>
+#include <algorithm>
+#include <tuple>
+#include <utility>
+#include <set>
+#include <vector>
+#include <map>
+#include <unordered_set>
+#include <deque>
+#include <shared_mutex>
+
+#include "../cc-lib/base/stringprintf.h"
+#include "../cc-lib/base/logging.h"
+#include "../cc-lib/arcfour.h"
+#include "../cc-lib/util.h"
+#include "../cc-lib/stb_image.h"
+#include "../cc-lib/stb_image_write.h"
+#include "../cc-lib/vector-util.h"
+#include "../cc-lib/threadutil.h"
+#include "../cc-lib/randutil.h"
+#include "../cc-lib/base/macros.h"
+#include "../cc-lib/color-util.h"
+#include "../cc-lib/image.h"
+
+#include "../chess.h"
+#include "../pgn.h"
+#include "../bigchess.h"
+#include "../pack.h"
+
+#include "loadpositions.h"
+#include "network.h"
+
+#include "clutil.h"
+#include "timer.h"
+
+#define FONTCHARS " ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz0123456789`-=[]\\;',./~!@#$%^&*()_+{}|:\"<>?" /* removed icons */
+#define FONTSTYLES 7
+
[…]

### 2019-02-14 trunk/chess/pack.cc

@@ -1,0 +1,688 @@
+
+#include "pack.h"
+
+#include <cstdint>
+#include <tuple>
+
+static constexpr int16_t pack_table[4096] = {
+  -1, 0, 1, 2, 3, 4, 5, 6,
+  7, 8, 9, -1, -1, -1, -1, -1,
+  10, 11, 12, -1, -1, -1, -1, -1,
+  13, -1, -1, 14, -1, -1, -1, -1,
+  15, -1, -1, -1, 16, -1, -1, -1,
+  17, -1, -1, -1, -1, 18, -1, -1,
+  19, -1, -1, -1, -1, -1, 20, -1,
+  21, -1, -1, -1, -1, -1, -1, 22,
+  23, -1, 24, 25, 26, 27, 28, 29,
+  30, 31, 32, 33, -1, -1, -1, -1,
+  34, 35, 36, 37, -1, -1, -1, -1,
+  -1, 38, -1, -1, 39, -1, -1, -1,
+  -1, 40, -1, -1, -1, 41, -1, -1,
+  -1, 42, -1, -1, -1, -1, 43, -1,
+  -1, 44, -1, -1, -1, -1, -1, 45,
+  -1, 46, -1, -1, -1, -1, -1, -1,
+  47, 48, -1, 49, 50, 51, 52, 53,
+  54, 55, 56, 57, 58, -1, -1, -1,
+  59, 60, 61, 62, 63, -1, -1, -1,
+  -1, -1, 64, -1, -1, 65, -1, -1,
+  -1, -1, 66, -1, -1, -1, 67, -1,
+  -1, -1, 68, -1, -1, -1, -1, 69,
+  -1, -1, 70, -1, -1, -1, -1, -1,
+  -1, -1, 71, -1, -1, -1, -1, -1,
+  72, 73, 74, -1, 75, 76, 77, 78,
+  -1, 79, 80, 81, 82, 83, -1, -1,
+  -1, 84, 85, 86, 87, 88, -1, -1,
+  89, -1, -1, 90, -1, -1, 91, -1,
+  -1, -1, -1, 92, -1, -1, -1, 93,
+  -1, -1, -1, 94, -1, -1, -1, -1,
+  -1, -1, -1, 95, -1, -1, -1, -1,
+  -1, -1, -1, 96, -1, -1, -1, -1,
+  97, 98, 99, 100, -1, 101, 102, 103,
+  -1, -1, 104, 105, 106, 107, 108, -1,
+  -1, -1, 109, 110, 111, 112, 113, -1,
[…]

### 2019-02-14 trunk/chess/stockfish-player.h

@@ -9,6 +9,7 @@ Player *CreateStockfish5();
 Player *CreateStockfish10();
 Player *CreateStockfish15();
 Player *CreateStockfish20();
+Player *CreateStockfish1M();
 
 Player *CreateWorstfish();
 

### 2019-02-14 trunk/chess/stockfish.h

@@ -5,6 +5,7 @@
 #include <string>
 #include <memory>
 #include <mutex>
+#include <cstdint>
 
 struct Subprocess;
 
@@ -13,9 +14,11 @@ struct Stockfish {
   // a child process.
 
   // Level in [0, 20] with 20 being strongest.
+  // Nodes is the number of nodes to search (if nonzero). 1 million takes about
+  // one second per move.
   // Note that engine loading is lazy; errors like missing stockfish.exe
   // won't occur until the first call to GetMove.
-  Stockfish(int level);
+  Stockfish(int level, int64_t nodes);
 
   struct Score {
     bool is_mate = false;
@@ -34,6 +37,7 @@ private:
   // Must hold lock.
   void InitEngine();
   const int level;
+  const int64_t nodes;
   std::mutex subprocess_m;
   std::unique_ptr<Subprocess> subprocess;
 };

### 2019-02-14 trunk/chess/tournament.cc

@@ -4,6 +4,8 @@
 #include <mutex>
 #include <memory>
 #include <unordered_map>
+#include <cstdlib>
+#include <unordered_set>
 
 #ifdef __MINGW32__
 #include <windows.h>
@@ -47,46 +49,52 @@ using namespace std;
 // Number of round-robin rounds.
 // (Maybe should be based on the total number of games we want
 // to simulate?)
-static constexpr int TOTAL_ROUNDS = 108; // 250;
 static constexpr int THREADS = 54;
-static constexpr int ROUNDS_PER_THREAD = TOTAL_ROUNDS / THREADS;
+static constexpr int ROUNDS_PER_THREAD = 1;
+static constexpr int TOTAL_ROUNDS = THREADS * ROUNDS_PER_THREAD;
 
 typedef Player *(*Entrant)();
-// using Entrant = Player *(*)();
+
+// 			CreateSinglePlayer,
 const vector<Entrant> &GetEntrants() {
   static vector<Entrant> *entrants =
     new vector<Entrant>{
 			CreateWorstfish,
 			CreateRandom,
-			// CreateFirstMove,
-			// CreateAlphabetical,
-			CreateCCCP,
-			CreateNoIInsist,
+			CreateFirstMove,
+			CreateAlphabetical,
 			CreatePacifist,
-			CreateSameColor,
-			// CreateMirrorYSymmetry,
-			// CreateMirrorXSymmetry,
+			CreateGenerous,
+			CreateNoIInsist,
+			CreateMirrorYSymmetry,
+			CreateMirrorXSymmetry,
 			CreateSymmetry180,
+			CreateSameColor,
 			CreateOppositeColor,
+			CreateHuddle,
+			CreateSwarm,
 			CreateSuicideKing,
 			CreateReverseStarting,
-			// CreateMinOpponentMoves,
-			// CreateHuddle,
-			CreateSwarm,
-			CreateGenerous,
-			// CreateTopple1M,
-			// CreateTopple10K,
-			// CreateSinglePlayer,
-			// CreateChessmaster1,
-			// CreateChessmaster2,
-			// CreateStockfish0,
-			// CreateStockfish5,
-			// CreateStockfish10,
-			// CreateStockfish15,
-			// CreateStockfish20,
+			CreateMinOpponentMoves,
+			CreateCCCP,
+			CreateTopple10K,
+			CreateTopple1M,
+			CreateChessmaster1,
+			CreateChessmaster2,
[…]

### 2019-02-14 chess/chessmaster.cc
Experiment to predict legal moves. Packed representation of moves. Mode for tournament where we only compute one row/col. Add stockfish node option and Stockfish1M player

@@ -335,9 +335,11 @@ Move Chessmaster::GetMove(const Position &pos) {
       }
 
       if (sr == -1 || sc == -1 || dr == -1 || dc == -1) {
-	if (VERBOSE) printf("no move %d %d %d %d\n",
-			    sr, sc, dr, dc);
-	Screenshot("no-move.png");
+	if (VERBOSE) {
+	  printf("no move %d %d %d %d\n",
+		 sr, sc, dr, dc);
+	  Screenshot("no-move.png");
+	}
 	return move;
       }
 

### 2019-02-14 chess/pairs.cc

@@ -1,0 +1,85 @@
+
+// Simple ditty to enumerate the pairs of square that can possibly
+// have legal moves between them. The answer is 1792, which is about
+// 43% of the dense 64^2.
+// Used to generate pack.cc.
+
+#include <cmath>
+#include <cstdio>
+#include <vector>
+
+// #include "chess.h"
+
+using namespace std;
+
+bool CouldMove(int sr, int sc, int dr, int dc) {
+  // No self-moves.
+  if (sr == dr && sc == dc)
+    return false;
+
+  // Horizontal and vertical ok.
+  if (sr == dr || sc == dc)
+    return true;
+
+  // Diagonal ok
+  if (std::abs(dr - sr) == std::abs(dc - sc))
+    return true;
+
+  // Within 2 squares. This covers king (1 square, and
+  // castling is represented as moving two spaces),
+  // pawn (double moves, captures, en passant) and knight
+  // moves. This does not overcount, because within this
+  // box, queen and knight cover the entire thing (in fact
+  // with no overlap!)
+  return std::abs(dr - sr) <= 2 && std::abs(dc - sc) <= 2;
+}
+
+static void PrintVec(const vector<int> &v, int w) {
+  int row_left = w;
+  printf("  ");
+  for (int i : v) {
+    printf("%d,", i);
+    row_left--;
+    if (row_left == 0) {
+      printf("\n  ");
+      row_left = w;
+    } else {
+      printf(" ");
+    }
+  }
+}
+
+int main (int argc, char **argv) {
+  int count = 0;
+  // index is sr * 512 + sc * 64 + dr * 8 + dc
+  vector<int> pack;
+  vector<int> unpack;
+  
+  for (int sr = 0; sr < 8; sr++) {
+    for (int sc = 0; sc < 8; sc++) {
+      for (int dr = 0; dr < 8; dr++) {
+	for (int dc = 0; dc < 8; dc++) {
+	  if (CouldMove(sr, sc, dr, dc)) {
+	    pack.push_back(count);
+	    // unpack.emplace_back(sr, sc, dr, dc);
+	    unpack.push_back(sr * 512 + sc * 64 + dr * 8 + dc);
+	    count++;
+	  } else {
+	    pack.push_back(-1);
+	  }
+	}
+      }
+    }
+  }
+
+  printf("constexpr int16 pack_table[4096] = {\n");
+  PrintVec(pack, 8);
[…]

### 2019-02-14 chess/stockfish-player.cc

@@ -1,6 +1,7 @@
 
 #include <string>
 #include <memory>
+#include <cstdint>
 
 #include "../cc-lib/base/logging.h"
 #include "../cc-lib/base/stringprintf.h"
@@ -11,14 +12,17 @@
 #include "player-util.h"
 #include "subprocess.h"
 
+using int64 = int64_t;
+
 using Move = Position::Move;
 using namespace std;
 
 namespace {
 struct StockfishPlayer : public Player {
 
-  StockfishPlayer(int level) : level(level) {
-    fish.reset(new Stockfish(level));
+  StockfishPlayer(int level, int nodes, const string &name) :
+    level(level), nodes(nodes), name(name) {
+    fish.reset(new Stockfish(level, nodes));
     CHECK(fish.get());
   }
   
@@ -43,22 +47,24 @@ struct StockfishPlayer : public Player {
   }
   
   string Name() const override {
-    return StringPrintf("stockfish%d", level);
+    return name;
   }
   
   string Desc() const override {
-    return StringPrintf("Stockfish engine, no tables, level %d",
-			level);
+    return StringPrintf("Stockfish engine, no tables, level %d, nodes %lld",
+			level, nodes);
   }
   
   const int level;
+  const int64 nodes;
+  const string name;
   std::unique_ptr<Stockfish> fish;
 };
 
 struct WorstfishPlayer : public EvalResultPlayer {
 
   WorstfishPlayer(int level) : EvalResultPlayer(), level(level) {
-    fish.reset(new Stockfish(level));
+    fish.reset(new Stockfish(level, 0));
     CHECK(fish.get());
   }
 
@@ -127,23 +133,27 @@ struct WorstfishPlayer : public EvalResultPlayer {
 
 
 Player *CreateStockfish0() {
-  return new StockfishPlayer(0);
+  return new StockfishPlayer(0, 0, "stockfish0");
 }
 
 Player *CreateStockfish5() {
-  return new StockfishPlayer(5);
+  return new StockfishPlayer(5, 0, "stockfish5");
 }
 
 Player *CreateStockfish10() {
-  return new StockfishPlayer(10);
+  return new StockfishPlayer(10, 0, "stockfish10");
 }
 
 Player *CreateStockfish15() {
-  return new StockfishPlayer(15);
+  return new StockfishPlayer(15, 0, "stockfish15");
 }
 
 Player *CreateStockfish20() {
[…]

### 2019-02-17 chess/blind/getlegal.cc
Paper editing. Code to make permutation tables. Some other miscellaney, apparently

@@ -1152,6 +1152,7 @@ static void RandomizeNetwork(ArcFour *rc, Network *net) {
 // These must be initialized before starting the UI thread!
 static constexpr int NUM_VIDEO_STIMULATIONS = 6;
 static constexpr int EXPORT_EVERY = 16;
+static constexpr int LOG_EXPORT_EVERY = 25;
 // static constexpr int EXPORT_EVERY = 1;
 static std::shared_mutex video_export_m;
 static int current_round = 0;
@@ -1628,11 +1629,13 @@ LoadPositions load_positions(
 static void TrainThread() {
   Timer setup_timer; 
 
+  int exports_until_log = 1; // LOG_EXPORT_EVERY;
+  
   // Number of training examples per round of training.
   // XXX This is probably still way too low. These are very small for chess (a
   // stimulation just needs the node activation values, so it's much smaller than
   // the network itself, which has to store weights for each incoming edge).
-  static constexpr int EXAMPLES_PER_ROUND = 64; // 4096;
+  static constexpr int EXAMPLES_PER_ROUND = 4096; // 4096;
   static constexpr int EXAMPLE_QUEUE_TARGET = std::max(EXAMPLES_PER_ROUND * 2, 1024);
   // On a verbose round, we write a network checkpoint and maybe some
   // other stuff to disk. XXX: Do this based on time, since rounds speed can vary
@@ -2052,7 +2055,24 @@ static void TrainThread() {
 	  total_error += fabs(d);
 	}
       }
-      ExportTotalErrorToVideo(total_error / (double)examples.size());
+      double avg_error = total_error / (double)examples.size();
+      ExportTotalErrorToVideo(avg_error);
+
+      exports_until_log--;
+      if (exports_until_log <= 0) {
+
+	FILE *log = fopen("move-train-log.tsv", "ab");
+	if (log) {
+	  fprintf(log, "%lld\t%d\t%lld\t%.6f\t%.6f\n",
+		  net->rounds,
+		  EXAMPLES_PER_ROUND,
+		  net->examples,
+		  round_learning_rate, avg_error);
+	  fclose(log);
+	}
+
+	exports_until_log = LOG_EXPORT_EVERY;
+      }
     }
     
     const int num_examples = examples.size();

### 2019-02-17 trunk/chess/papers/survival.tex
Paper editing. Code to make permutation tables. Some other miscellaney, apparently

@@
   difficult to agree on the approach to drawing, let alone establish
-  that this is both player's desire. Since the rules of
-  chess-to-the-death can't forbid us from colluding at this moment, I
-  hereby declare that the following is the correct approach: 
+  that this is both players' desire. Since the rules of
+  chess-to-the-death can't forbid us from colluding right now as you
+  read this paper, I hereby declare that the following is the correct
+  approach:
 \begin{enumerate}[label=\arabic*.]
@@
     eyes. 
-  \item[1.] ... Nf6. This is both a strong response for black in a real game,
+  \item[1.] \ldots Nf6. This is both a strong response for black in a real game,
     and simultaneously a signal that a draw is desired. The other
     advantage is that very weak players may simply copy what white does.
+    % XXX potential citation to 'elo world' if players are stateful..
     In doing so, they will also play this move. 
@@
     draw.
-  \item[2.] ... Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
+  \item[2.] \ldots Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
     for a second time.
@@
 
-  If 1. ... Nc6 or another Knight's move, white can also consider
-  continuing in the obvious way. However after 1. ... d5, black has
+  If 1. \ldots Nc6 or another Knight's move, white can also consider
+  continuing in the obvious way. However after 1. \ldots d5, black has
   refused or not noticed the draw. Fortunately, white is still in a
@@
 
-  The biggest risk for white is that black does not play 2. ... Ng8
-  but rather a normal move like 2. ... g6 (``Fool's Draw Betrayed'').
+  The biggest risk for white is that black does not play 2. \ldots Ng8
+  but rather a normal move like 2. \ldots g6 (``Fool's Draw Betrayed'').
   This can happen if black is not metagaming at all (for Nf6 is a
@@
 Otherwise, of course, the best chances of survival would trivially be
-with the two kings, since they always survive. Loss includes
-resignation, since most high-level games actually end once the
-defeated player agrees that loss is inevitable. We can think of this
-common case like king seppuku. Many games also end in time forfeit,
-which is like the king's poor diet and lifestyle choices leading to
-a death by natural causes.
+with the two kings, since they are never formally captured. Loss
+includes resignation, since most high-level games actually end once
+the defeated player agrees that loss is inevitable. We can think of
+this common case like king seppuku. Many games also end in time
+forfeit, which is like the king's poor diet and lifestyle choices
+leading to a death by natural causes.
 
-Neither side has a decisive advantage, and many games end in a draw,
-with both kings surviving. So the survival chances of a king are
-clearly greater than 50\%---pretty decent odds. Is it possible that
-any other piece has even better chances?
+Neither side is believed to have a decisive advantage, and many games
+end in a draw, with both kings surviving. So the survival chances of a
+king are likely greater than 50\%; pretty decent odds. Is it possible
+that any other piece has even better chances? Let's find out---our
+lives may depend on it!
 
 \footnotetext{
-  But is it? The lichess implementation (although known
-  to be buggy\cite{kingme}) does not permit a threefold repetition
-  claim in this situation, which got me thinking that maybe there is
-  some subtlety here. Is the starting position special somehow, not
+  But is it?
+
+  First of all, although either player is allowed to {\it claim} a
+  draw after three repetitions of the same position, it is not
+  automatic. However, FIDE rules do declare that the game simply ends
+  in a draw upon {\it five} repetitions. Of course it is easy to
+  extend the Fool's Draw to accommodate this.
+
+  Second: The lichess implementation (although known to be
+  buggy\cite{kingme}) does not permit a threefold repetition claim in
+  this situation, which got me thinking that maybe there is some
+  subtlety here. Is the starting position special somehow, not
   counting as having occurred? The relevant statute, from the FIDE
@@
 
-Okay, this is pretty much off the cuff, but let’s say, just for white,
+Okay, this is pretty much off the cuff, but let's say, just for white,
 from most to least likely to survive:
@@
 I figure the king has got to be somewhere near the middle of the pack
-since he dies in half of games featuring a winner — but with slightly
+since he dies in half of games featuring a winner---but with slightly
 higher-than-even odds of surviving, since some games end in a draw.
@@
 their fellows on the left. But maybe it should actually be worse,
-since if they die, it’s because they failed to protect the king. Plus,
+since if they die, it's because they failed to protect the king. Plus,
 having heard the tip about C Pawn loud and clear, I’m gonna assume
@@
 (Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
-pawn in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a
-good example of a piece that has very different survival rates in
-different opening preferences. Since the variance is otherwise very
-low, I did not attempt to
+pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
+and declining the pawn are both popular and sound responses. This may
+be a good example of a piece that has substantially different survival
+rates in different opening preferences. Since the variance is
+otherwise extremely low, I only report means for the remainder of the
+paper.
 
@@
 as a result, the survival chances for the kings are both close to
-50\%.
+50\%. Although the database contains games in many time formats and
+with all varieties of human skill (including over a thousand games by
+Magnus Carlsen, the world champion and highest rated player of all
+time\footnote{Although to be fair, his username ``DrDrunkenstein''
+  suggests he may not play at full strength.}), blitz ($\sim 5$ minutes
+per side) and bullet ($\sim 1$ minute) games are predominant. In
+Section~\ref{sec:comparative} I show some slices of the data for
+comparison. % XXX could summarize whether this is a real effect or
+% not, once I have the slices!
 
@@
 Other than the bishops---which have no legal way to reach half of the
-squares---every piece reaches every square in at least a thousand
+squares---every piece ends on every square in at least a thousand
 games. So we have enough samples to have reasonable confidence in our
@@
 
+\section{Guesses and slices}
+%  
+%  \begin{tabular}{|l|l|}
+%  \hline
+%    tom & actual \\
+%   \hline
+%   \Pawn f    &  \\
+%   \Pawn c    &  \\
+%   \Pawn g    &  \\
+%   \Pawn a    &  \\
+%   \Pawn h    &  \\
+%   \Pawn b    &  \\
+%   \Rook h    &  \\
+%   \Rook a    &  \\
+%   \King      &  \\
+%   \Queen     &  \\
+%   \Bishop f  &  \\
+%   \Bishop c  &  \\
+%   \Knight g  &  \\
+%   \Knight b  &  \\
+%   \Pawn e    &  \\
+%   \Pawn d    &  \\
+%  \hline
+%  \end{tabular}
+%
+
 % Battle chess

### 2019-02-17 trunk/chess/maketable.cc

@@ -52,6 +52,41 @@ static constexpr const char *const PIECE_NAME[32] = {
   "g1 knight",
   "h1 rook", };
 
+// For makeperms.cc
+static constexpr const char *const CONSTANT_NAME[32] = {
+  "BLACK_ROOK_A",
+  "BLACK_KNIGHT_B",
+  "BLACK_BISHOP_C",
+  "BLACK_QUEEN",
+  "BLACK_KING",
+  "BLACK_BISHOP_F",
+  "BLACK_KNIGHT_G",
+  "BLACK_ROOK_H",
+  "BLACK_PAWN_A",
+  "BLACK_PAWN_B",
+  "BLACK_PAWN_C",
+  "BLACK_PAWN_D",
+  "BLACK_PAWN_E",
+  "BLACK_PAWN_F",
+  "BLACK_PAWN_G",
+  "BLACK_PAWN_H",
+  "WHITE_PAWN_A",
+  "WHITE_PAWN_B",
+  "WHITE_PAWN_C",
+  "WHITE_PAWN_D",
+  "WHITE_PAWN_E",
+  "WHITE_PAWN_F",
+  "WHITE_PAWN_G",
+  "WHITE_PAWN_H",
+  "WHITE_ROOK_A",
+  "WHITE_KNIGHT_B",
+  "WHITE_BISHOP_C",
+  "WHITE_QUEEN",
+  "WHITE_KING",
+  "WHITE_BISHOP_F",
+  "WHITE_KNIGHT_G",
+  "WHITE_ROOK_H", };
+
 // Take a piece number 0-31 and return its Position-style piece.
 static uint8 PiecePiece(int p) {
   static const Position startpos;
@@ -72,7 +107,7 @@ void ReadStats(const string &filename, Stats *stat_buckets) {
   for (int b = 0; b < NUM_BUCKETS; b++) {
     Stats *stats = &stat_buckets[b];
     int64 num_games;
-    CHECK(s >> num_games);
+    CHECK(s >> num_games) << filename;
     stats->num_games += num_games;
     for (int p = 0; p < 32; p++) {
       int64 on;
@@ -540,6 +575,30 @@ void GenReport(Stats *stat_buckets) {
   // <span style="position:absolute; color:#333" class="white piece">&#9812;</span>
 
 
+  {
+    FILE *f = fopen("report-probs.cc", "wb");
+    
+    vector<pair<int, float>> ranks;
+    for (int p : pieces) {
+      ranks.emplace_back(p, survived[p].Mean());
+    }
[…]

### 2019-02-17 trunk/chess/rungames.cc
Simultaneously compute stats for multiple criteria, since reading and executing games is basically the entire cost

@@ -10,6 +10,7 @@
 #include <unistd.h>
 
 #include "base/logging.h"
+#include "base/stringprintf.h"
 #include "util.h"
 #include "city.h"
 
@@ -29,14 +30,28 @@ static constexpr int64 MAX_GAMES = 0LL;
 // #define SELF_CHECK true
 #undef SELF_CHECK
 
-enum class Criteria {
-  ALL_GAMES,
+enum Criteria {
+  ALL_GAMES = 0,
   TITLED_ONLY,
   BULLET_ONLY,
   BLITZ_ONLY,
   RAPID_ONLY,
   CLASSICAL_ONLY,
+
+  NUM_CRITERIA,
 };
+
+string CriteriaName(Criteria crit) {
+  switch (crit) {
+  case ALL_GAMES: return "all";
+  case TITLED_ONLY: return "titled";
+  case BULLET_ONLY: return "bullet";
+  case BLITZ_ONLY: return "blitz";
+  case RAPID_ONLY: return "rapid";
+  case CLASSICAL_ONLY: return "classical";
+  default: return "BAD_CRITERIA";
+  }
+}
 
 static constexpr const char *const PIECE_NAME[32] = {
   "a8 rook",
@@ -62,29 +77,77 @@ static constexpr const char *const PIECE_NAME[32] = {
   "h1 rook", };
 
 struct Processor {
-  Processor(Criteria crit) : crit(crit) {}
-  const Criteria crit;
-  
-  bool Accept(const PGN &pgn) const {
-    switch (crit) {
-    case Criteria::ALL_GAMES:
-      return true;
-    case Criteria::TITLED_ONLY:
-      return 
-	ContainsKey(pgn.meta, "WhiteTitle") ||
-	ContainsKey(pgn.meta, "BlackTitle");
-      break;
-    case Criteria::BULLET_ONLY:
-    case Criteria::BLITZ_ONLY:
-    case Criteria::RAPID_ONLY:
-    case Criteria::CLASSICAL_ONLY:
-      LOG(FATAL) << "Unimplemented :(";
-      return false;
+  Processor(uint32 want_crit_set) : want_crit_set(want_crit_set) {  }
 
-    default:
-      LOG(FATAL) << "Bad criteria";
-      return false;
+  // Get the subset of want_crit_set that the game actually
+  // possesses. This is used to filter the game, and to decide
+  // which stat buckets to accumulate results into.
+  uint32 GetCriteria(const PGN &pgn) const {
+    uint32 result = 0;
+
[…]

### 2019-02-17 chess/papers/survival.tex
Some work on survival paper, including code for making figures

@@
   difficult to agree on the approach to drawing, let alone establish
-  that this is both player's desire. Since the rules of
-  chess-to-the-death can't forbid us from colluding at this moment, I
-  hereby declare that the following is the correct approach:
+  that this is both players' desire. Since the rules of
+  chess-to-the-death can't forbid us from colluding right now as you
+  read this paper, I hereby declare that the following is the correct
+  approach:
 \begin{enumerate}[label=\arabic*.]
@@
 
-\section{Hypotheses}
+\section{Hypotheses} \label{sec:hypotheses}
 
@@
 
-% XXX: Move this stuff to the appendix.
-
 Here are my guesses.
@@
   is, the a1 and a8 rooks probably have about the same survival
-  chances (it's known that white has a slight statistical advantage
-  but it is probably only around 1\%). So these guesses will be
-  written about white's pieces.
+  chances (it's known that white has a slight statistical
+  advantage~\cite{chesstego} but it is probably only around 1\%). So
+  these guesses will be written about white's pieces.
 \item The d2 and e2 pawns are very active in common openings, and
@@
 
-This leaves my final ranking, from most surviving to most dead:
-
-\begin{enumerate}
-\item  f2
-\item  c2
-\item  g2
-\item  a2
-\item  h2
-\item  b2,
-\item  h1 rook,
-\item  a1 rook,
-\item  king,
-\item  queen,
-\item  f1 bishop,
-\item  c1 bishop,
-\item  g1 knight,
-\item  b1 knight,
-\item  e2,
-\item  d2,
-\end{enumerate}
+The final ranking that I committed to, from most surviving to most dead:
+\pawn f, \pawn c, \pawn g, \pawn a, \pawn h, \pawn b, \rook h, \rook a,
+\king, \queen, \bishop f, \knight g, \pawn e, \pawn d.
 
@@
 how to spell the difficult word aficionado without spell-check, he is
-not good at chess. A few drinking buddies with varying chessperience were
-also consulted for their wagers.
-
-\subsection{Ben}\footnote{Ben does not prefer to use the shift key.}
-
-edge pawns almost never played til endgame let alone traded off
-
-ph
-pa
-
-not quite sure where these should go (pb more likely to see play in
-queenside minority attacks in k-side castle games?)
-
-pf
-pg
-pb
-
-rook play more likely to be active on q side than on k side (also the
-classic Nxc7 fork in low rank play), but overall more likely to stay
-tucked away compared to q
-
-Rh
-Ra
-
-i think IQP positions are more likely than not saccing e in e4 openings
-but on the other hand d is often traded off in e4 openings while vice
-versa is not as true
-
-pe
-pd
-
-q probably involved in many checkmates (low ranked play) or resignations
-before traded off (high ranked play)
-
-Q
-
-just randomly guessing k dies in about 1/3 of games, times 1/2 for 2
-sides
-
-K
-
-this pawn is a super goner (sicilian, QGA, ...)
-
-pc
-
-most doomed seem to be the minor pieces as i'd guess at least half of
-them get traded off on c/f/3/6 or e/d/4/5 in near every game so
-
-Bc
-Bf
-Nb
-Ng
-
-\subsection{Jim}
-
-Most-to-least-survival hero tier list for chess (patch 1.0):
-
-1.: King
- -- If I estimate that about 2/3rds of all regular pieces are captured in an average game, and the probability of any non-king piece being captured is uniform, then the king is clearly the most likely to survive.
-(I'm going to break symmetry here and rank black king less likely to survive than white king.)
-
-2. Both Rooks
- -- Kept in reserve for castling purposes.
-
-3. The A, B, G, and H pawns
- -- maybe people will forget to move them because they are far from the center.
-
-4. Both Knights
- -- They are slippery, but they often get deep into enemy territory quickly.
-
-5. The C,D,E,F pawns
- -- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.
-
-6. Both Bishops
- -- https://youtu.be/gDnE-5lD7w8
-
-7. The Queen
- -- A high-value target, seems unlikely to survive.
-
-This list was assembled for chess.report by ix, who only barely
-understands the rules of chess and rarely plays. This is probably why
-the justifications get increasingly nonsensical.
-
-\subsection{David}
-
-Here's my ranking, along with the probabilities of survival that I guess for each piece:
-
-white g-pawn: 0.72
-black g-pawn: 0.72
-white b-pawn: 0.69
-black a-pawn: 0.66
-black b-pawn: 0.65
-white king: 0.65
-white a-pawn: 0.64
-black king: 0.55
-black f-pawn: 0.54
-white f-pawn: 0.54
-black h-pawn: 0.53
-white h-pawn: 0.53
-white kingside rook: 0.52
-black kingside rook: 0.52
-white queenside rook: 0.51
-black queenside rook: 0.51
-white c-pawn: 0.49
-black c-pawn: 0.44
-white light-squared bishop: 0.33
-black light-squared bishop: 0.33
-white dark-squared bishop: 0.32
-black dark-squared bishop: 0.32
-white queen: 0.31
-black queen: 0.30
-white kingside knight: 0.29
-black kingside knight: 0.29
-white queenside knight: 0.28
-black queenside knight: 0.28
-white e-pawn: 0.19
-black d-pawn: 0.18
-white d-pawn: 0.17
-black e-pawn: 0.16
-
-\subsection{William}
-
-Okay, this is pretty much off the cuff, but let’s say, just for white,
-from most to least likely to survive:
-
-Queenside Rook
-B Pawn
-A Pawn
-Kingside Bishop
-C Pawn
-Queen
-King
-H Pawn
-Queenside Bishop
-F Pawn
-G Pawn
-Queenside Knight
-Kingside Rook
-Kingside Knight
-D Pawn
-E Pawn
-
-Some motivated reasoning:
-
-I figure the king has got to be somewhere near the middle of the pack
-since he dies in half of games featuring a winner — but with slightly
-higher-than-even odds of surviving, since some games end in a draw.
-I’m probably mixing up means and medians here somehow..
-
-...
-King
-...
-
-I’m gonna assume castling happens more often on the King’s side, so
-let’s give Kingside Rook and F, G, and H Pawns a better shot than
-their fellows on the left. But maybe it should actually be worse,
-since if they die, it’s because they failed to protect the king. Plus,
-having heard the tip about C Pawn loud and clear, I’m gonna assume
-that bad boy most often becomes a new Queen, which means he gets more
-survival points than the real Queen herself.
-
-...
-C Pawn
-Queen
-A Pawn
-B Pawn
-Queenside Rook
-King
-F Pawn
-G Pawn
-H Pawn
-Kingside Rook
-...
-
-D and E Pawn are nothing but pawns, and they mostly sacrifice
-themselves to the cause.
-
-...
-D Pawn
-E Pawn
-
-Randomizing within these constraints gives us our starting point. Then
-the wildcard Bishops and Knights get randomly distributed through what
-remains to come up with this final answer shown above.
+not good at chess. A few drinking buddies with varying chessperience
+were also consulted for their wagers; these are compared in
+Section~\ref{sec:buddies}.
 
@@
 (Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
-pawn in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a
-good example of a piece that has very different survival rates in
-different opening preferences. Since the variance is otherwise
-extremely low, I only report means for the remainder of the paper.
+pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
+and declining the pawn are both popular and sound responses. This may
+be a good example of a piece that has substantially different survival
+rates in different opening preferences. Since the variance is
+otherwise extremely low, I only report means for the remainder of the
+paper.
 
@@
 as a result, the survival chances for the kings are both close to
-50\%.
+50\%. Although the database contains games in many time formats and
+with all varieties of human skill (including over a thousand games by
+Magnus Carlsen, the world champion and highest rated player of all
+time\footnote{Although to be fair, his username ``DrDrunkenstein''
+  suggests he may not play at full strength.}), blitz ($\sim 5$ minutes
+per side) and bullet ($\sim 1$ minute) games are predominant. In
+Section~\ref{sec:slices} I show some slices of the data for
+comparison. % XXX could summarize whether this is a real effect or
+% not, once I have the slices!
 
@@
 achievement for reaching this square this way!
+% 12 times for titled players, although there are a few things that
+% are even rarer, like dying on a8.
 % TODO: Perhaps could find a game where this happened?!
@@
 
+\section{Guesses and slices}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{hypotheses}
+  \caption{
+    Piece rankings (from most surviving to most dead); either a
+    human or hypothesis or the actual results across all games. 
+    The {\bf actual} column appears multiple times so that each human
+    gets a chance to be adjacent to it; this makes for the easiest
+    visual comparison. Lines connect the same piece in adjacent columns,
+    and are darker if the pairs have more different ranks.
+  } \label{fig:buddies}
+\end{figure}
+
+It is a good and humbling act to compare the actual results to the
+hypotheses gathered before the experiment. Figure~\ref{fig:buddies}
+compares the ranking across all games (slice {\bf Actual}) to the
+author author (slice {\bf Tom}) and his drinking buddies (slices {\bf
+  Ben}, {\bf Jim}, {\bf David}, {\bf William}). There are a number of
+different reasonable ways to measure the accuracy of this type of
+position; a very simple one is the sum of the absolute differences in
+rank for each piece (e.g. if in one ranking \king is \#3, and in the
+other \#5, then this contributes 2 to the total error). By this
+metric, Ben has the best prediction (98 error), followed by David
+(138) and Tom (148). Tom and David had the most similar predictions
+(116) and Jim and William the most different (312). The expected error
+between two completely random permutations is about 341, so all of
+these guesses are significantly better than chance. Note in the actual
+ranking, many pieces have very similar survival probabilities, and
+many guesses are ambivalent about groups of pieces. Weighting each
+rank difference equally is therefore an oversimplification. It would
+have been better to ask each participant to give probabilities, as
+David did; this would give us more sensitive error metrics and more
+opportunities to spend the afternoon making visualizations.
+
+Several drinking buddies gave rationales for their hypotheses (mine
+appear in Section~\ref{sec:hypotheses}).
+
+\medskip
+{\bf Ben} does not prefer to use the shift key, a typographic quirk I
+replicated faithfully here even though it burns my eyes:
+
+\begin{quote}
+edge pawns almost never played til endgame let alone traded off (\pawn
+h, \pawn a)
+
+not quite sure where these should go (pb more likely to see play in
+queenside minority attacks in k-side castle games?) (\pawn f, \pawn g, \pawn b)
+
+rook play more likely to be active on q side than on k side (also the
+classic Nxc7 fork in low rank play), but overall more likely to stay
+tucked away compared to q (\rook h, \rook a)
+
+i think IQP positions are more likely than not saccing e in e4 openings
+but on the other hand d is often traded off in e4 openings while vice
+versa is not as true (\pawn e, \pawn d)
+
+q probably involved in many checkmates (low ranked play) or resignations
+before traded off (high ranked play) (\queen)
+
+just randomly guessing k dies in about 1/3 of games, times 1/2 for 2
+sides (\king)
+
+this pawn is a super goner (sicilian, QGA, ...) (\pawn c)
+
+most doomed seem to be the minor pieces as i'd guess at least half of
+them get traded off on c/f/3/6 or e/d/4/5 in near every game so (\bishop c, \bishop f, \knight b, \knight g)
+\end{quote}
+
+{\bf Jim} ``barely understands the rules of chess'' and ``rarely plays.'' His justifications get ``increasingly nonsensical:''
+
+\begin{quote}
+Most-to-least-survival hero tier list for chess (patch 1.0):
+
+1.: King --- If I estimate that about 2/3rds of all regular pieces are
+captured in an average game, and the probability of any non-king piece
+being captured is uniform, then the king is clearly the most likely to
+survive. (I'm going to break symmetry here and rank black king less
+likely to survive than white king.)
+
+2. Both Rooks --- Kept in reserve for castling purposes.
+
+3. The A, B, G, and H pawns --- maybe people will forget to move them because they are far from the center.
+
+4. Both Knights --- They are slippery, but they often get deep into enemy territory quickly.
+
+5. The C,D,E,F pawns --- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.
+
+6. Both Bishops --- https://youtu.be/gDnE-5lD7w8
+
+7. The Queen --- A high-value target, seems unlikely to survive.
+\end{quote}
+
+{\bf David} simply provided a ranking, along with survival
+probabilities, in typical understated style.
+
+\medskip
+{\bf William} notes that his guesses are
+``pretty much off the cuff,'' but provides some motivated reasoning:
+
+\begin{quote}
+I figure the king has got to be somewhere near the middle of the pack
+since he dies in half of games featuring a winner---but with slightly
+higher-than-even odds of surviving, since some games end in a draw.
+I'm probably mixing up means and medians here somehow..
+
+I'm gonna assume castling happens more often on the King's side, so
+let's give Kingside Rook and F, G, and H Pawns a better shot than
+their fellows on the left. But maybe it should actually be worse,
+since if they die, it's because they failed to protect the king. Plus,
+having heard the tip about C Pawn\footnote{I believe the ``tip'' here
+  was that I described the rest of us (William was the last
+  respondant) as disagreeing most on \pawn c. I think William
+  misinterpreted this ``tip.''} loud and clear, I'm gonna assume that
+bad boy most often becomes a new Queen, which means he gets more
+survival points than the real Queen herself.
+
+D and E Pawn are nothing but pawns, and they mostly sacrifice
+themselves to the cause.
+
+Randomizing within these constraints gives us our starting point. Then
+the wildcard Bishops and Knights get randomly distributed through what
+remains to come up with this final answer shown above.
+\end{quote}
+
+\subsection{Slices} \label{sec:slices}
+The survival probabilities differ depending on the conditions of the
+game; Figure~\ref{fig:slices} compares some of those slices. Here the
+{\bf All} slice is the same as the {\bf Actual} slice in
+Figure~\ref{fig:buddies}, and consists of all acceptable games in the
+database.
+
+The {\bf titled} slice includes only games where at least one of the
+players has an official title (Grandmaster, International Master, FIDE
+Master, etc.\footnote{Lichess used to award the LM ``Lichess Master'' title
+  to notable players on the site; this title is excluded.}). These
+games have high-quality play, although there are not many of them
+(only XXX), so the variance is somewhat higher. % XXX give error bounds?
+
+The other slices, {\bf bullet}, {\bf blitz}, {\bf rapid} and {\bf
+  classical} are different time control formats. Games on Lichess are
+played with a starting clock (per side) and an increment added to the
+clock after each player's move. The total time is taken to be the
+starting time plus 40 times the increment (with the idea that an
+average game has 40 moves per side); this is the same formula that
+Lichess uses. A bullet game is when the total time per side is
+between 30~seconds and 3~minutes; blitz is between 3 and 8~minutes;
+rapid is between 8 and 25~minutes; classical is any more than this
+(including untimed games).
+
 % Battle chess
 % Feminist game
-
-\nocite{chesstego}
 

### 2019-02-18 chess/papers/survival.tex
Rename Player to StatelessPlayer, in preparation for allowing stateful algorithms

@@
     and simultaneously a signal that a draw is desired. The other
-    advantage is that very weak players may simply copy what white does.
-    % XXX potential citation to 'elo world' if players are stateful..
+    advantage is that very weak players\cite{eloworld} may simply copy what white does.
     In doing so, they will also play this move. 
@@
 
-  Since this is one of the shortest possible routes to a draw, I hereby
-  dub this line the ``Fool's Draw,'' by analogy with the Fool's Mate.
+  Since this is one of the shortest possible routes to a draw, I
+  hereby dub this line the ``Fool's Draw,'' by analogy with the Fool's
+  Mate. In this line all pieces survive, which is anyway humane and
+  also advantageous in the case that you are someone is simultaneously
+  being one of the chesspieces to the death!
 
@@
 not good at chess. A few drinking buddies with varying chessperience
-were also consulted for their wagers; these are compared in
-Section~\ref{sec:buddies}.
+were also consulted for their wagers; these are compared to each other
+and to the actual results in Section~\ref{sec:guesses}.
 
@@
 processing these took some care for efficiency and parallelism.
+Fortunately, I have a computer with just an obscene number of cores
+and truly excessive RAM, so you gotta use that for
+something.\footnote{ To torture your own desktop computer, source code
+  is available at
+  \url{sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/chess/}.}
 
-Other than that, I simply parsed and simulated each game. For each
+Other than that, I simply implemented the rules of chess, wrote a PGN
+parser, then parsed and simulated each game. For each
 of the 32 pieces in the starting position, I tracked its current
@@
 
-\begin{figure}[htb]
+\begin{figure}[t]
   \begin{center}
@@
 this was the most controversial among the drinking buddies
-(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
+(Section~\ref{sec:guesses}). Note that \pawn c is the sacrificed
 pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
@@
 time\footnote{Although to be fair, his username ``DrDrunkenstein''
-  suggests he may not play at full strength.}), blitz ($\sim 5$ minutes
-per side) and bullet ($\sim 1$ minute) games are predominant. In
-Section~\ref{sec:slices} I show some slices of the data for
-comparison. % XXX could summarize whether this is a real effect or
-% not, once I have the slices!
+  suggests he may not play at full strength.}), blitz ($\sim\! 5$
+minutes per side) and bullet ($\sim\! 1$ minute) games are
+predominant. Nonetheless, the results are fairly robust across
+different time formats and skill levels. In Section~\ref{sec:slices} I
+show some slices of the data for comparison.
 
@@
 such knowledge could help calibrate your anxiety.
+
+% XXX using this a way to play the game
 
@@
 achievement for reaching this square this way!
-% 12 times for titled players, although there are a few things that
-% are even rarer, like dying on a8.
+% 12 times for titled players (in new dataset too), although there are
+% a few things that are even rarer, like dying on a8.
 % TODO: Perhaps could find a game where this happened?!
@@
 
-\section{Guesses and slices}
+\section{Guesses and slices} \label{sec:guesses}
 
@@
 
-It is a good and humbling act to compare the actual results to the
-hypotheses gathered before the experiment. Figure~\ref{fig:buddies}
-compares the ranking across all games (slice {\bf Actual}) to the
-author author (slice {\bf Tom}) and his drinking buddies (slices {\bf
-  Ben}, {\bf Jim}, {\bf David}, {\bf William}). There are a number of
-different reasonable ways to measure the accuracy of this type of
-position; a very simple one is the sum of the absolute differences in
-rank for each piece (e.g. if in one ranking \king is \#3, and in the
-other \#5, then this contributes 2 to the total error). By this
-metric, Ben has the best prediction (98 error), followed by David
-(138) and Tom (148). Tom and David had the most similar predictions
-(116) and Jim and William the most different (312). The expected error
-between two completely random permutations is about 341, so all of
-these guesses are significantly better than chance. Note in the actual
-ranking, many pieces have very similar survival probabilities, and
-many guesses are ambivalent about groups of pieces. Weighting each
-rank difference equally is therefore an oversimplification. It would
-have been better to ask each participant to give probabilities, as
-David did; this would give us more sensitive error metrics and more
-opportunities to spend the afternoon making visualizations.
+Like all good scientific research, I explicitly compare the actual
+results to the hypotheses gathered before the experiment; this is a
+hygenic and humbling exercise. Figure~\ref{fig:buddies} compares the
+ranking across all games (slice {\bf Actual}) to the author author
+(slice {\bf Tom}) and his drinking buddies (slices {\bf Ben}, {\bf
+  Jim}, {\bf David}, {\bf William}). There are a number of different
+reasonable ways to measure the accuracy of this type of position; a
+very simple one is the sum of the absolute differences in rank for
+each piece (e.g. if in one ranking \king is \#3, and in the other \#5,
+then this contributes 2 to the total error). By this metric, Ben has
+the best prediction (98 error), followed by David (138) and Tom (148).
+Tom and David had the most similar predictions (116) and Jim and
+William the most different (312). The expected error between two
+completely random permutations is about 341, so all of these guesses
+are significantly better than chance. Note in the actual ranking, many
+pieces have very similar survival probabilities, and many guesses are
+ambivalent about groups of pieces. Weighting each rank difference
+equally is therefore an oversimplification. It would have been better
+to ask each participant to give probabilities, as David did; this
+would give us more sensitive error metrics and more opportunities to
+spend the afternoon making visualizations.
 
@@
 \subsection{Slices} \label{sec:slices}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{slices}
+  \caption{
+    Ranks and survival probabilities for different subsets of games.
+    The same piece in adjacent rows is connected to highlight
+    differences in the ranking, as before. The time formats all
+    exhibit similar ranking with only small perturbations. Games
+    including a titled player (rightmost column) are the most
+    different, although we have far fewer samples in this set,
+    so variance becomes significant.
+  } \label{fig:slices}
+\end{figure}
+
 The survival probabilities differ depending on the conditions of the
@@
 players has an official title (Grandmaster, International Master, FIDE
-Master, etc.\footnote{Lichess used to award the LM ``Lichess Master'' title
-  to notable players on the site; this title is excluded.}). These
-games have high-quality play, although there are not many of them
-(only XXX), so the variance is somewhat higher. % XXX give error bounds?
+Master, etc.\footnote{Lichess used to award the LM ``Lichess Master''
+  title to notable players on the site; this title is excluded from
+  the sample.}). These games have high-quality play, but far fewer
+samples (``only'' 3.4 million). This set exhibits significant
+variance; for example, the survival rate of \pawn c ranges from
+$38$--$46\%$. This is both because of the small sample size and the
+bucketization by player name; there are few enough titled players that
+an individual's preference in openings and style of play moves the
+values of their entire bucket. I caution against reading too much into
+this column. It seems we can at least conclude that these games tend
+to be much bloodier (Kings aside, survival rates are lower across the
+board); top players are less likely to fall for traps early in the
+game, perhaps more willing to sacrifice material, and more likely to
+play into endgames where almost every piece is exchanged. If being a
+chesspiece to the death, you do {\em not} want to have a Grandmaster
+playing the game!
 
-The other slices, {\bf bullet}, {\bf blitz}, {\bf rapid} and {\bf
-  classical} are different time control formats. Games on Lichess are
-played with a starting clock (per side) and an increment added to the
-clock after each player's move. The total time is taken to be the
-starting time plus 40 times the increment (with the idea that an
-average game has 40 moves per side); this is the same formula that
-Lichess uses. A bullet game is when the total time per side is
-between 30~seconds and 3~minutes; blitz is between 3 and 8~minutes;
-rapid is between 8 and 25~minutes; classical is any more than this
-(including untimed games).
+% XXX give % error % bounds?
+
+On the other hand, the other slices all have enough samples that the
+variance is minimal. These slices, {\bf bullet} (151,261,707 games),
+{\bf blitz} (236,050,938 games), {\bf rapid} (98,606,558) and {\bf
+  classical} (13,886,352) are different time control formats. Games on
+Lichess are played with a starting clock (per side) and an increment
+added to the clock after each player's move. The game is classified
+according to the estimated total time: The starting time plus $40 \times$
+the increment (with the idea that an average game has 40 moves per
+side); this is the same formula that Lichess uses. A bullet game is
+when the total time per side is between 30~seconds and 3~minutes;
+blitz is between 3 and 8~minutes; rapid is between 8 and 25~minutes;
+classical is any more than this (including untimed games). Each of
+these slices has enough samples that the variance is very low.
+Here we see that the ranking is rather stable across the range of
+time formats, which was not what I expected. This should increase
+our confidence that the results are really inherent to chess, not
+just the particulars of this data set.
 

### 2019-02-18 chess/makeperm.cc

@@ -322,7 +322,7 @@ static string PieceName(int x) {
 }
 
 // Actual results over all games (500m)
-vector<pair<int, float>> prob_actual = {
+vector<pair<int, float>> actual_probs = {
   {WHITE_PAWN_H, 0.715432286},
   {BLACK_PAWN_A, 0.705200613},
   {WHITE_PAWN_A, 0.704906046},
@@ -374,20 +374,207 @@ int RankError(const vector<int> &a,
   }
   return delta;
 }
-	      
+
+
+vector<pair<int, float>> titled_probs = {
+  {WHITE_PAWN_H, 0.669196844},
+  {BLACK_PAWN_A, 0.667333484},
+  {WHITE_PAWN_A, 0.662262022},
+  {BLACK_PAWN_H, 0.656060696},
+  {WHITE_PAWN_G, 0.624552369},
+  {BLACK_PAWN_G, 0.605713427},
+  {WHITE_PAWN_B, 0.568870246},
+  {BLACK_PAWN_B, 0.564766288},
+  {WHITE_KING, 0.538510501},
+  {WHITE_PAWN_F, 0.536402941},
+  {BLACK_PAWN_F, 0.528088152},
+  {BLACK_KING, 0.510553479},
+  {BLACK_ROOK_A, 0.49197644},
+  {WHITE_ROOK_A, 0.489860713},
+  {WHITE_ROOK_H, 0.476608098},
+  {BLACK_ROOK_H, 0.474159986},
+  {WHITE_PAWN_C, 0.423101097},
+  {WHITE_QUEEN, 0.395854831},
+  {BLACK_QUEEN, 0.388097733},
+  {BLACK_PAWN_C, 0.373647988},
+  {BLACK_PAWN_E, 0.369938105},
+  {BLACK_BISHOP_F, 0.312111437},
+  {WHITE_BISHOP_F, 0.308377415},
+  {WHITE_PAWN_E, 0.303896815},
+  {BLACK_BISHOP_C, 0.287457943},
+  {BLACK_PAWN_D, 0.287427694},
+  {WHITE_BISHOP_C, 0.277486771},
+  {WHITE_PAWN_D, 0.257768214},
+  {WHITE_KNIGHT_B, 0.231363401},
+  {BLACK_KNIGHT_B, 0.230935827},
+  {BLACK_KNIGHT_G, 0.200624391},
+  {WHITE_KNIGHT_G, 0.200406671},
+};
+
[…]

### 2019-02-18 chess/player.cc

@@ -40,7 +40,7 @@ static int TypeValue(uint8 t) {
   }
 }
 
-struct FirstMovePlayer : public Player {
+struct FirstMovePlayer : public StatelessPlayer {
 
   static int WhiteCode(const Move &m) {
     int src = (int)m.src_row * 8 + (int)m.src_col;
@@ -79,7 +79,7 @@ struct FirstMovePlayer : public Player {
   }
 };
 
-struct RandomPlayer : public Player {
+struct RandomPlayer : public StatelessPlayer {
   RandomPlayer() : rc(PlayerUtil::GetSeed()) {
     rc.Discard(800);
   }
@@ -100,7 +100,7 @@ struct RandomPlayer : public Player {
   ArcFour rc;
 };
 
-struct CCCPPlayer : public Player {
+struct CCCPPlayer : public StatelessPlayer {
 
   struct LabeledMove {
     Move m;
@@ -204,7 +204,7 @@ struct CCCPPlayer : public Player {
   }
 };
 
-struct AlphabeticalPlayer : public Player {
+struct AlphabeticalPlayer : public StatelessPlayer {
 
   struct LabeledMove {
     Move m;
@@ -694,7 +694,7 @@ struct Symmetry180Player : public SymmetryPlayer {
   }
 };
 
-struct SinglePlayerPlayer : public Player {
+struct SinglePlayerPlayer : public StatelessPlayer {
   explicit SinglePlayerPlayer(int max_depth) : 
     max_depth(max_depth), 
     rc(PlayerUtil::GetSeed()) {
@@ -797,74 +797,74 @@ struct SinglePlayerPlayer : public Player {
 
 }  // namespace
 
-Player *CreateFirstMove() {
+StatelessPlayer *CreateFirstMove() {
   return new FirstMovePlayer;
 }
 
-Player *CreateRandom() {
+StatelessPlayer *CreateRandom() {
   return new RandomPlayer;
 }
 
-Player *CreateAlphabetical() {
+StatelessPlayer *CreateAlphabetical() {
   return new AlphabeticalPlayer;
 }
 
-Player *CreateCCCP() {
+StatelessPlayer *CreateCCCP() {
   return new CCCPPlayer;
 }
 
-Player *CreatePacifist() {
+StatelessPlayer *CreatePacifist() {
   return new PacifistPlayer;
 }
 
-Player *CreateMinOpponentMoves() {
[…]

### 2019-02-18 trunk/chess/chessmaster.h
Rename Player to StatelessPlayer, in preparation for allowing stateful algorithms

@@ -46,8 +46,8 @@ private:
 };
 
 // Level 1. Makes random move if chessmaster fails for some reason.
-Player *CreateChessmaster1();
+StatelessPlayer *CreateChessmaster1();
 // Level 2.
-Player *CreateChessmaster2();
+StatelessPlayer *CreateChessmaster2();
 
 #endif

### 2019-02-20 chess/fate-player.h
Add fatalist, survivalist, equalizer. improve Player interface and ability to select games for tournament

@@ -9,4 +9,9 @@ Player *Dangerous();
 Player *Popular();
 Player *Rare();
 
+Player *Survivalist();
+Player *Fatalist();
+
+Player *Equalizer();
+
 #endif

### 2019-02-22 nand/paper.tex
Search for impl of nand on binary3

@@
 The real numbers are described by IEEE 754, most recently revised in
-2008\cite{ieee754}. Every real number has a sign, a mantissa, and an
+AD 2008\cite{ieee754}. Every real number has a sign, a mantissa, and an
 exponent. Actually, this understates the elegance of real numbers,
@@
 you count very high or very low. Excitingly, we also have both
-positive and negative versions of 0.
+positive and negative versions of 0. Some numbers have multiple
+representations, and most numbers cannot be represented at all.
 
@@
 % really lean into the joke that we choose the "number" NAN
-
-\nocite{fluint8}
 
@@
 but require less space and allow faster calculations. For our purposes
-in this paper, since we only need to represent the two values \nan\ and
+in this paper, since we only need to represent the two numbers \nan\ and
 \inf, we are interested in the smallest possible representation.
@@
 precision'' is 16, 4 bits is ``eighth precision.'' Given how nicely
-all this works out, shouldn't there be a {\tt eighth} base type in
+all this works out, shouldn't there be an {\tt eighth} base type in
 most modern programming languages and GPUs? Since there are so few
@@
 
+% hardware notes:
+% http://openocd.org/ looks like a good way to do in-device
+% programming like over SWD, because it runs on the pi. A board
+% could have jumpers or even software route to different
+% gpio pins. (You would need to run at 3v3 for this of course.)
+% I guess make sure the chip is compatible, but there are lots
+% explicitly listed.
+
+% Some supported devices:
+% AT91SAM7 - no fp in the M7
+% at91sam3u - no fp here either :/
+
+% LPC1700, LPC1800,
+% LPC2000, LPC4300, , STR7x, STR9x, LM3, STM32x and
+% EFM32
+
+% ok the STM32f3 is supported, has FPU, 51 IO pins,
+% here is an example breakout:
+% http://ebrombaugh.studionebula.com/embedded/stm32f303breakout/index.html
+% requires a crystal, voltage regulator, and some filter capacitors,
+% but looks manageable. (And the chip has some 5V tolerant TTL I/O
+% pins too, which could be nice for other projects)
+% also maybe a ferrite bead (for input to analog VDDA, so probably
+% not needed for this project)
+
+Idea: Make a FPU. How small could it be?
+I guess the simplest thing that could be described as faithful
+would compute the NAND function of two binary3 numbers, but on
+the entire domain. There are six input pins, three output pins,
+and a table size of 8^2 = 64.
+
+A nice simple nand function is
+f(x, y) = inf - max(x + y, -inf)
+
+What is a simple implementation of this function?
+It can only output inf or nan, so that's good. Of the three
+output pins, the first is always 0, the second always 1, so
+at worst we would just need a table of 64 bits for this last
+value.
+
+Further simplification:
+
+if isfinite(x) && isfinite(y)
+then
+  // inf - (x + y) = inf
+  0
+else
+// x + y is nan when x is nan, y is nan, or x and y are
+// infinites with different signs. If they are both -inf,
+// then we have max(-inf, -inf) anyway, which is the same
+// as max(nan, -inf). So we have
+  if x == inf && y == inf
+    // both positive infinities
+    // inf - inf = nan
+    1
+  else
+    // inf - -inf = inf
+    0
+
+So ultimately this function only returns 1 in the case
+that both inputs are exactly +inf, the pattern 0 1 0.
+
+If the inputs are a0 a1 a2, b0, b1, b2, and outputs
+are c0, c1, c2, then:
+c0 = 0
+c1 = 1
+c2 = !a0 && !a2 && !b0 && !b2 && a1 && b1
+
+// Could adapt find.sml to find solutions for this,
+// using only NAND. Only 2^6 = 64 possible values
+// for these variables
+
+let x = nand(a0, a2)   // !a0 || !a2
+let y = nand(b0, b2)   // !b0 || !b2
+
+let out =
+
+#  0 & 0 & 0 &   $+0$ \\
+#  0 & 0 & 1 &   +1
+#  0 & 1 & 0 &   $+\infty$ \\
+#  0 & 1 & 1 &    \nan \\
+#  1 & 0 & 0 &   $-0$ \\
+#  1 & 0 & 1 &   $-1$ \\
+#  1 & 1 & 0 &   $-\infty$ \\
+#  1 & 1 & 1 &    \nan \\
+
+% would be cool if the meta NAND gate were addressable
+% from the rasberry pi (well duh of course it is?) because
+% then we could also use it to simulate itself in software
+
 \nocite{ieee754}

### 2019-02-24 nand/paper.tex
Abrupt power failures are a good reminder to back up your work

@@
 % ok the STM32f3 is supported, has FPU, 51 IO pins,
+% specifically like the
+% https://www.mouser.com/ProductDetail/STMicroelectronics/STM32F303RDT7?qs=%2Fha2pyFaduikpRt0bOqlahEvgTujL2BrdYPen6RZRRk%3D
+
 % here is an example breakout:
@@
 
-let x = nand(a0, a2)   // !a0 || !a2
-let y = nand(b0, b2)   // !b0 || !b2
+let x = nand(!a0, !a2)
+let y = nand(!b0, !b2)
+let z = nand(a1, b1)
 
-let out =
+let w = Not(Nand(Not(x), Not(y)))
+let out = Not(Nand(w, Not(z))
+
+... this is pretty much a straightforward Nand construction.
+I guess it doesn't really simplify much across units.
+
+or with more gates,
+Not(Or(z, Or(y, x)))
+
+and I mean obviously it's just
+And(Not (Or (a0, a2, b0, b2)), a1, b1)
+if you have all the gates directly.
+
+we can certainly make chips implementing each of these gates,
+since we have floating point implementations of each, but
+then we sort of lose the final joke since we'd just build
+a floating point unit that computes the NAND instruction.
+
+so let's try the pure-nand approach.
+
+implementing the nots also with nands, this works out to 14
+gates, which is ``just'' three chips. very reasonable. I was
+actually hoping for it to be a little more ridiculous.
+(maybe there is a lower pin-count device with an easier
+package? programming them is still going to be a huge pain)
+
+if we used the meta-nand gates, we'd have 14 * 3 = 42 chips,
+which is OTOH probably too many.
+
+with this meta-fpu, we have 2 binary3 inputs and 1 binary3
+output, but each ``bit'' is represented as a binary3 number
+itself. So this makes (3 * 2 + 1) * 3 = 27 wires to interface
+with it. This is exactly as many GPIO the pi has, so that
+is good, although IIRC a few pi pins have special annoyances,
+and it doesn't leave us any space for a programmer.
+Maybe just best to make a separate programmer and connect it
+up to different test points manually. Then we could reuse
+the programmer pretty easily.
+
+Note that in a pinch we can ``cheat'' pretty easily here
+since some of the pins have constant values. Probably justifiable
+is that the meta-nand always outputs 0 (nan) for g and 1 (inf) for h,
+plus two of the bits of i are fixed because it will only represent
+nan or inf. But really would be ``prettiest'' to not make this cheat,
+I mean, it's a pretty silly excursion anyway.
+
+oh jk there is actually gpio0 too, so 28 pins on pi.
 

### 2019-02-28 trunk/chess/papers/survival.tex
Simultaneously compute stats for multiple criteria, since reading and executing games is basically the entire cost
Some light tweaks to survival paper

@@
 
-This leaves my final ranking, from most surviving to most dead:
-
-\begin{enumerate}
-\item  f2
-\item  c2
-\item  g2
-\item  a2
-\item  h2
-\item  b2,
-\item  h1 rook,
-\item  a1 rook,
-\item  king,
-\item  queen,
-\item  f1 bishop,
-\item  c1 bishop,
-\item  g1 knight,
-\item  b1 knight,
-\item  e2,
-\item  d2,
-\end{enumerate}
+The final ranking that I predict, from most surviving to most dead:
+\pawn f, \pawn c, \pawn g, \pawn a, \pawn h, \pawn b, \rook h, \rook a,
+\king, \queen, \bishop f, \knight g, \pawn e, \pawn d.
 
@@
 how to spell the difficult word aficionado without spell-check, he is
-not good at chess. A few drinking buddies with varying chessperience were
-also consulted for their wagers.
-
-\subsection{Ben}\footnote{Ben does not prefer to use the shift key.}
-
-edge pawns almost never played til endgame let alone traded off
-
-ph
-pa
-
-not quite sure where these should go (pb more likely to see play in
-queenside minority attacks in k-side castle games?)
-
-pf
-pg
-pb
-
-rook play more likely to be active on q side than on k side (also the
-classic Nxc7 fork in low rank play), but overall more likely to stay
-tucked away compared to q
-
-Rh
-Ra
-
-i think IQP positions are more likely than not saccing e in e4 openings
-but on the other hand d is often traded off in e4 openings while vice
-versa is not as true
-
-pe
-pd
-
-q probably involved in many checkmates (low ranked play) or resignations
-before traded off (high ranked play)
-
-Q
-
-just randomly guessing k dies in about 1/3 of games, times 1/2 for 2
-sides
-
-K
-
-this pawn is a super goner (sicilian, QGA, ...)
-
-pc
-
-most doomed seem to be the minor pieces as i'd guess at least half of
-them get traded off on c/f/3/6 or e/d/4/5 in near every game so
-
-Bc
-Bf
-Nb
-Ng
-
-\subsection{Jim}
-
-Most-to-least-survival hero tier list for chess (patch 1.0):
-
-1.: King
- -- If I estimate that about 2/3rds of all regular pieces are captured in an average game, and the probability of any non-king piece being captured is uniform, then the king is clearly the most likely to survive.
-(I'm going to break symmetry here and rank black king less likely to survive than white king.)
-
-2. Both Rooks
- -- Kept in reserve for castling purposes.
-
-3. The A, B, G, and H pawns
- -- maybe people will forget to move them because they are far from the center.
-
-4. Both Knights
- -- They are slippery, but they often get deep into enemy territory quickly.
-
-5. The C,D,E,F pawns
- -- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.
-
-6. Both Bishops
- -- https://youtu.be/gDnE-5lD7w8
-
-7. The Queen
- -- A high-value target, seems unlikely to survive.
-
-This list was assembled for chess.report by ix, who only barely
-understands the rules of chess and rarely plays. This is probably why
-the justifications get increasingly nonsensical.
-
-\subsection{David}
-
-Here's my ranking, along with the probabilities of survival that I guess for each piece:
-
-white g-pawn: 0.72
-black g-pawn: 0.72
-white b-pawn: 0.69
-black a-pawn: 0.66
-black b-pawn: 0.65
-white king: 0.65
-white a-pawn: 0.64
-black king: 0.55
-black f-pawn: 0.54
-white f-pawn: 0.54
-black h-pawn: 0.53
-white h-pawn: 0.53
-white kingside rook: 0.52
-black kingside rook: 0.52
-white queenside rook: 0.51
-black queenside rook: 0.51
-white c-pawn: 0.49
-black c-pawn: 0.44
-white light-squared bishop: 0.33
-black light-squared bishop: 0.33
-white dark-squared bishop: 0.32
-black dark-squared bishop: 0.32
-white queen: 0.31
-black queen: 0.30
-white kingside knight: 0.29
-black kingside knight: 0.29
-white queenside knight: 0.28
-black queenside knight: 0.28
-white e-pawn: 0.19
-black d-pawn: 0.18
-white d-pawn: 0.17
-black e-pawn: 0.16
-
-\subsection{William}
-
-Okay, this is pretty much off the cuff, but let's say, just for white,
-from most to least likely to survive:
-
-Queenside Rook
-B Pawn
-A Pawn
-Kingside Bishop
-C Pawn
-Queen
-King
-H Pawn
-Queenside Bishop
-F Pawn
-G Pawn
-Queenside Knight
-Kingside Rook
-Kingside Knight
-D Pawn
-E Pawn
-
-Some motivated reasoning:
-
-I figure the king has got to be somewhere near the middle of the pack
-since he dies in half of games featuring a winner---but with slightly
-higher-than-even odds of surviving, since some games end in a draw.
-I’m probably mixing up means and medians here somehow..
-
-...
-King
-...
-
-I’m gonna assume castling happens more often on the King’s side, so
-let’s give Kingside Rook and F, G, and H Pawns a better shot than
-their fellows on the left. But maybe it should actually be worse,
-since if they die, it's because they failed to protect the king. Plus,
-having heard the tip about C Pawn loud and clear, I’m gonna assume
-that bad boy most often becomes a new Queen, which means he gets more
-survival points than the real Queen herself.
-
-...
-C Pawn
-Queen
-A Pawn
-B Pawn
-Queenside Rook
-King
-F Pawn
-G Pawn
-H Pawn
-Kingside Rook
-...
-
-D and E Pawn are nothing but pawns, and they mostly sacrifice
-themselves to the cause.
-
-...
-D Pawn
-E Pawn
-
-Randomizing within these constraints gives us our starting point. Then
-the wildcard Bishops and Knights get randomly distributed through what
-remains to come up with this final answer shown above.
+not good at chess. A few drinking buddies with varying chessperience
+were also consulted for their wagers; these are compared to each other
+and to the actual results in Section~\ref{sec:guesses}.
 
@@
 this was the most controversial among the drinking buddies
-(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
+(Section~\ref{sec:guesses}). Note that \pawn c is the sacrificed
 pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
-and declining the pawn are both popular and sound responses. This may
+and declining the pawn are both common and sound responses. This may
 be a good example of a piece that has substantially different survival
@@
 time\footnote{Although to be fair, his username ``DrDrunkenstein''
-  suggests he may not play at full strength.}), blitz ($\sim 5$ minutes
-per side) and bullet ($\sim 1$ minute) games are predominant. In
-Section~\ref{sec:comparative} I show some slices of the data for
-comparison. % XXX could summarize whether this is a real effect or
-% not, once I have the slices!
+  suggests he may not play at full strength.}), blitz ($\sim\! 5$
+minutes per side) and bullet ($\sim\! 1$ minute) games are
+predominant. Nonetheless, the results are fairly robust across
+different time formats and skill levels. In Section~\ref{sec:slices} I
+show some slices of the data for comparison.
 
@@
 achievement for reaching this square this way!
+% 12 times for titled players (in new dataset too), although there are
+% a few things that are even rarer, like dying on a8.
 % TODO: Perhaps could find a game where this happened?!
 Aside from pawns, the weirdest fate is for \knight g to die on h1,
-which happened 47,307 times. Corners are of course fairly garbage for
-the knight, although it is twice as likely to survive on this square.
+which happened 47,307 times. Corners are of course garbage for the
+knight, although it is twice as likely to survive on this square.
 
-% There may even be hidden achievements for reaching these squares!
+There are characteristic patterns for each of the pieces, which make
+sense given their starting positions movement rules. You could
+probably guess the piece just by looking at one of the heat maps
+below, although---spoiler alert---the piece is just listed right there
+and they are in order. Two independent things are communicated in
+these graphics: The chance that the piece ends the game on some square
+(alive or dead), and its survival chances there. In each map, a darker
+background color indicates that the piece ends on that square more
+often. The shade is based on the rank (64/64 black is most common,
+63/64 black is next most common, etc.) rather than the absolute
+probability, since otherwise the graphic looks boring. The number on
+the square is the percentage survival of the piece when it is last
+seen (either being captured or surviving to the end of the game) on
+that square. The four squares with the highest survival rates are
+underlined for your convenience.
 
-That said, there are characteristic patterns for each of the pieces,
-which make sense given their starting positions and patterns of
-movement. You could probably guess the piece just by looking at one of
-the heat maps below, although---spoiler alert---the piece is just
-listed right there and they are in order. Two independent things are
-communicated in these graphics: The chance that the piece ends the
-game on some square (alive or dead), and its survival chances there.
-In each map, a darker background color indicates that the piece ends
-on that square more often. The shade is based on the rank (64/64 black
-is most common, 63/64 black is next most common, etc.) rather than the
-absolute probability, since otherwise the graphic looks boring. The
-number on the square is the percentage survival of the piece when it
-is last seen (either being captured or surviving to the end of the
-game) on that square. The four squares with the highest survival rates
-are underlined for your convenience.
+\clearpage
 
@@
 
-\section{Guesses and slices}
-%  
-%  \begin{tabular}{|l|l|}
-%  \hline
-%    tom & actual \\
-%   \hline
-%   \Pawn f    &  \\
-%   \Pawn c    &  \\
-%   \Pawn g    &  \\
-%   \Pawn a    &  \\
-%   \Pawn h    &  \\
-%   \Pawn b    &  \\
-%   \Rook h    &  \\
-%   \Rook a    &  \\
-%   \King      &  \\
-%   \Queen     &  \\
-%   \Bishop f  &  \\
-%   \Bishop c  &  \\
-%   \Knight g  &  \\
-%   \Knight b  &  \\
-%   \Pawn e    &  \\
-%   \Pawn d    &  \\
-%  \hline
-%  \end{tabular}
-%
+\clearpage
+
+Aside from admiring the groovy pictures, the data can also be used as
+the basis of an algorithm for playing chess. There are several things
+to try; we could move pieces towards squares where they are more
+likely to survive than die (keeping pieces alive is good), or just
+towards squares where they are more likely to be positioned at the end
+of the game (our moves are more likely to resemble real chess moves
+because they put pieces in their proper places). Alas, it turns out
+that these are bad approaches to chess, both because they are boring
+(most pieces are actually most comfortable in their starting
+positions) and because they perform badly against even weak
+opponents~\cite{eloworld}.
+
+\section{Guesses and slices} \label{sec:guesses}
+
+Like all good scientific research, I explicitly compare the actual
+results to the hypotheses gathered before the experiment; this is a
+hygenic and humbling exercise. Figure~\ref{fig:buddies} compares the
+ranking across all games (slice {\bf Actual}) to the author author
+(slice {\bf Tom}) and his drinking buddies (slices {\bf Ben}, {\bf
+  Jim}, {\bf David}, {\bf William}). There are a number of different
+reasonable ways to measure the accuracy of this type of position; a
+very simple one is the sum of the absolute differences in rank for
+each piece (e.g. if in one ranking \king is \#3, and in the other \#5,
+then this contributes 2 to the total error). By this metric, Ben has
+the best prediction (98 error), followed by David (138) and Tom (148).
+Tom and David had the most similar predictions (116) and Jim and
+William the most different (312). The expected error between two
+completely random permutations is about 341, so all of these guesses
+are significantly better than chance. Note in the actual ranking, many
+pieces have very similar survival probabilities, and many guesses are
+ambivalent about groups of pieces. Weighting each rank difference
+equally is therefore an oversimplification. It would have been better
+to ask each participant to give probabilities, as David did; this
+would give us more sensitive error metrics and more opportunities to
+spend the afternoon making visualizations.
+
+Several drinking buddies gave rationales for their hypotheses (mine
+appear in Section~\ref{sec:hypotheses}).
+
+\medskip
+{\bf Ben} does not prefer to use the shift key, a typographic quirk I
+replicated faithfully here even though it burns my eyes:
+
+\begin{quote}
+edge pawns almost never played til endgame let alone traded off (\pawn
+h, \pawn a)
+
+not quite sure where these should go (pb more likely to see play in
+queenside minority attacks in k-side castle games?) (\pawn f, \pawn g, \pawn b)
+
+rook play more likely to be active on q side than on k side (also the
+classic Nxc7 fork in low rank play), but overall more likely to stay
+tucked away compared to q (\rook h, \rook a)
+
+i think IQP positions are more likely than not saccing e in e4 openings
+but on the other hand d is often traded off in e4 openings while vice
+versa is not as true (\pawn e, \pawn d)
+
+q probably involved in many checkmates (low ranked play) or resignations
+before traded off (high ranked play) (\queen)
+
+just randomly guessing k dies in about 1/3 of games, times 1/2 for 2
+sides (\king)
+
+this pawn is a super goner (sicilian, QGA, ...) (\pawn c)
+
+most doomed seem to be the minor pieces as i'd guess at least half of
+them get traded off on c/f/3/6 or e/d/4/5 in near every game so (\bishop c, \bishop f, \knight b, \knight g)
+\end{quote}
+
+\begin{figure}[t]
+  \includegraphics[width=\linewidth]{hypotheses}
+  \caption{
+    Piece rankings (from most surviving to most dead); either a
+    human or hypothesis or the actual results across all games. 
+    The {\bf actual} column appears multiple times so that each human
+    gets a chance to be adjacent to it; this makes for the easiest
+    visual comparison. Lines connect the same piece in adjacent columns,
+    and are darker if the pairs have more different ranks.
+  } \label{fig:buddies}
+\end{figure}
+
+{\bf Jim} ``barely understands the rules of chess'' and ``rarely plays.'' His justifications get ``increasingly nonsensical:''
+
+\begin{quote}
+Most-to-least-survival hero tier list for chess (patch 1.0):
+
+1.: King --- If I estimate that about 2/3rds of all regular pieces are
+captured in an average game, and the probability of any non-king piece
+being captured is uniform, then the king is clearly the most likely to
+survive. (I'm going to break symmetry here and rank black king less
+likely to survive than white king.)
+
+2. Both Rooks --- Kept in reserve for castling purposes.
+
+3. The A, B, G, and H pawns --- maybe people will forget to move them because they are far from the center.
+
+4. Both Knights --- They are slippery, but they often get deep into enemy territory quickly.
+
+5. The C,D,E,F pawns --- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.
+
+6. Both Bishops --- https://youtu.be/gDnE-5lD7w8
+
+7. The Queen --- A high-value target, seems unlikely to survive.
+\end{quote}
+
+{\bf David} simply provided a ranking, along with survival
+probabilities, in typical understated style.
+
+\medskip
+{\bf William} notes that his guesses are
+``pretty much off the cuff,'' but provides some motivated reasoning:
+
+\begin{quote}
+I figure the king has got to be somewhere near the middle of the pack
+since he dies in half of games featuring a winner---but with slightly
+higher-than-even odds of surviving, since some games end in a draw.
+I'm probably mixing up means and medians here somehow..
+
+I'm gonna assume castling happens more often on the King's side, so
+let's give Kingside Rook and F, G, and H Pawns a better shot than
+their fellows on the left. But maybe it should actually be worse,
+since if they die, it's because they failed to protect the king. Plus,
+having heard the tip about C Pawn\footnote{I believe the ``tip'' here
+  was that I described the rest of us (William was the last
+  respondant) as disagreeing most on \pawn c. I think William
+  misinterpreted this ``tip.''} loud and clear, I'm gonna assume that
+bad boy most often becomes a new Queen, which means he gets more
+survival points than the real Queen herself.
+
+D and E Pawn are nothing but pawns, and they mostly sacrifice
+themselves to the cause.
+
+Randomizing within these constraints gives us our starting point. Then
+the wildcard Bishops and Knights get randomly distributed through what
+remains to come up with this final answer shown above.
+\end{quote}
+
+\subsection{Slices} \label{sec:slices}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{slices}
+  \caption{
+    Ranks and survival probabilities for different subsets of games.
+    The same piece in adjacent rows is connected to highlight
+    differences in the ranking, as before. The time formats all
+    exhibit similar ranking with only small perturbations. Games
+    including a titled player (rightmost column) are the most
+    different, although we have far fewer samples in this set,
+    so variance becomes significant.
+  } \label{fig:slices}
+\end{figure}
+
+The survival probabilities differ depending on the conditions of the
+game; Figure~\ref{fig:slices} compares some of those slices. Here the
+{\bf All} slice is the same as the {\bf Actual} column in
+Figure~\ref{fig:buddies}, and consists of all acceptable games in the
+database.
+
+The {\bf Titled} slice includes only games where at least one of the
+players has an official title (Grandmaster, International Master, FIDE
+Master, etc.\footnote{Lichess used to award the LM ``Lichess Master''
+  title to notable players on the site; this title is excluded from
+  the sample.}). These games have high-quality play, but far fewer
+samples (``only'' 3.4 million). This set exhibits significant
+variance; for example, the survival rate of \pawn c ranges from
+$38$--$46\%$. This is both because of the small sample size and the
+bucketization by player name; there are few enough titled players that
+an individual's preference in openings and style of play changes the
+values of their entire bucket. I caution against reading too much into
+this column. It seems we can at least conclude that these games tend
+to be much bloodier (Kings aside, survival rates are lower across the
+board); top players are less likely to fall for traps early in the
+game, perhaps more willing to sacrifice material, and more likely to
+play into endgames where almost every piece is exchanged. If being a
+chesspiece to the death, you do {\em not} want to have a Grandmaster
+playing the game!
+
+% XXX give % error % bounds?
+
+On the other hand, the other slices all have enough samples that the
+variance is minimal. These slices, {\bf Bullet} (151,261,707 games),
+{\bf Blitz} (236,050,938 games), {\bf Rapid} (98,606,558) and {\bf
+  Classical} (13,886,352) are different time control formats. Games on
+Lichess are played with a starting clock (per side) and an increment
+added to the clock after each player's move. The game is classified
+according to the estimated total time: The starting time plus $40 \times$
+the increment (with the idea that an average game has 40 moves per
+side); this is the same formula that Lichess uses. A bullet game is
+when the total time per side is between 30~seconds and 3~minutes;
+blitz is between 3 and 8~minutes; rapid is between 8 and 25~minutes;
+classical is any more than this (including untimed games). Each of
+these slices has enough samples that the variance is very low.
+Here we see that the ranking is rather stable across the range of
+time formats, which was not what I expected. This should increase
+our confidence that the results are really inherent to chess, not
+just the particulars of this data set.
 
@@
 % Feminist game
-
-\nocite{chesstego}
 

### 2019-03-02 chess/papers/survival.tex
Submitted version

@@
 
+\clearpage
+
   Since this is one of the shortest possible routes to a draw, I
@@
   Mate. In this line all pieces survive, which is anyway humane and
-  also advantageous in the case that you are someone is simultaneously
+  also advantageous in the case that you or someone is simultaneously
   being one of the chesspieces to the death!
@@
 \begin{itemize}
-\item Black and white are probably not significantly different. That
+\item Black and white are probably not substantially different. That
   is, the a1 and a8 rooks probably have about the same survival
@@
 
-The final ranking that I committed to, from most surviving to most dead:
+The final ranking that I predict, from most surviving to most dead:
 \pawn f, \pawn c, \pawn g, \pawn a, \pawn h, \pawn b, \rook h, \rook a,
@@
 pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
-and declining the pawn are both popular and sound responses. This may
+and declining the pawn are both common and sound responses. This may
 be a good example of a piece that has substantially different survival
@@
 
-% XXX using this a way to play the game
-
 Other than the bishops---which have no legal way to reach half of the
@@
 Aside from pawns, the weirdest fate is for \knight g to die on h1,
-which happened 47,307 times. Corners are of course fairly garbage for
-the knight, although it is twice as likely to survive on this square.
+which happened 47,307 times. Corners are of course garbage for the
+knight, although it is twice as likely to survive on this square.
 
-% There may even be hidden achievements for reaching these squares!
-
-That said, there are characteristic patterns for each of the pieces,
-which make sense given their starting positions and patterns of
-movement. You could probably guess the piece just by looking at one of
-the heat maps below, although---spoiler alert---the piece is just
-listed right there and they are in order. Two independent things are
-communicated in these graphics: The chance that the piece ends the
-game on some square (alive or dead), and its survival chances there.
-In each map, a darker background color indicates that the piece ends
-on that square more often. The shade is based on the rank (64/64 black
-is most common, 63/64 black is next most common, etc.) rather than the
-absolute probability, since otherwise the graphic looks boring. The
-number on the square is the percentage survival of the piece when it
-is last seen (either being captured or surviving to the end of the
-game) on that square. The four squares with the highest survival rates
-are underlined for your convenience.
+There are characteristic patterns for each of the pieces, which make
+sense given their starting positions movement rules. You could
+probably guess the piece just by looking at one of the heat maps
+below, although---spoiler alert---the piece is just listed right there
+and they are in order. Two independent things are communicated in
+these graphics: The chance that the piece ends the game on some square
+(alive or dead), and its survival chances there. In each map, a darker
+background color indicates that the piece ends on that square more
+often. The shade is based on the rank (64/64 black is most common,
+63/64 black is next most common, etc.) rather than the absolute
+probability, since otherwise the graphic looks boring. The number on
+the square is the percentage survival of the piece when it is last
+seen (either being captured or surviving to the end of the game) on
+that square. The four squares with the highest survival rates are
+underlined for your convenience.
 
@@
 
+Aside from admiring the groovy pictures, the data can also be used as
+the basis of an algorithm for playing chess. There are several things
+to try; we could move pieces towards squares where they are more
+likely to survive than die (keeping pieces alive is good), or just
+towards squares where they are more likely to be positioned at the end
+of the game (our moves are more likely to resemble real chess moves
+because they put pieces in their proper places). Alas, it turns out
+that these are bad approaches to chess, both because they are boring
+(most pieces are actually most comfortable in their starting
+positions) and because they perform badly against even weak
+opponents~\cite{eloworld}.
+
 \section{Guesses and slices} \label{sec:guesses}
 
-\begin{figure}
-  \includegraphics[width=\linewidth]{hypotheses}
-  \caption{
-    Piece rankings (from most surviving to most dead); either a
-    human or hypothesis or the actual results across all games. 
-    The {\bf actual} column appears multiple times so that each human
-    gets a chance to be adjacent to it; this makes for the easiest
-    visual comparison. Lines connect the same piece in adjacent columns,
-    and are darker if the pairs have more different ranks.
-  } \label{fig:buddies}
-\end{figure}
-
 Like all good scientific research, I explicitly compare the actual
-results to the hypotheses gathered before the experiment; this is a
+results to the hypotheses gathered before the experiment; this is an
 hygenic and humbling exercise. Figure~\ref{fig:buddies} compares the
@@
 \end{quote}
+
+\begin{figure}[t]
+  \includegraphics[width=\linewidth]{hypotheses}
+  \caption{
+    Piece rankings (from most surviving to most dead); either a
+    human or hypothesis or the actual results across all games. 
+    The {\bf actual} column appears multiple times so that each human
+    gets a chance to be adjacent to it; this makes for the easiest
+    visual comparison. Lines connect the same piece in adjacent columns,
+    and are darker if the pairs have more different ranks.
+  } \label{fig:buddies}
+\end{figure}
 
@@
 
-\begin{figure}
+\begin{figure}[t]
   \includegraphics[width=\linewidth]{slices}
@@
 game; Figure~\ref{fig:slices} compares some of those slices. Here the
-{\bf All} slice is the same as the {\bf Actual} slice in
+{\bf All} slice is the same as the {\bf Actual} column in
 Figure~\ref{fig:buddies}, and consists of all acceptable games in the
@@
 
-The {\bf titled} slice includes only games where at least one of the
+The {\bf Titled} slice includes only games where at least one of the
 players has an official title (Grandmaster, International Master, FIDE
@@
 bucketization by player name; there are few enough titled players that
-an individual's preference in openings and style of play moves the
+an individual's preference in openings and style of play changes the
 values of their entire bucket. I caution against reading too much into
@@
 On the other hand, the other slices all have enough samples that the
-variance is minimal. These slices, {\bf bullet} (151,261,707 games),
-{\bf blitz} (236,050,938 games), {\bf rapid} (98,606,558) and {\bf
-  classical} (13,886,352) are different time control formats. Games on
+variance is minimal. These slices, {\bf Bullet} (151,261,707 games),
+{\bf Blitz} (236,050,938 games), {\bf Rapid} (98,606,558) and {\bf
+  Classical} (13,886,352) are different time control formats. Games on
 Lichess are played with a starting clock (per side) and an increment

### 2019-03-02 trunk/nand/paper.tex
v1 board sent for manufacturing

@@
-flinf flops and NaN gates
-flip FLOPs and NaN gates?
+\documentclass[twocolumn]{article}
+\usepackage[top=0.5in, left=0.45in, right=0.45in, bottom=0.5in]{geometry}
+
+\usepackage{url}
+% \usepackage{code}
+% \usepackage{cite}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
+
+\usepackage{nopageno}
+
+\interfootnotelinepenalty=0
+
+% lets me explicitly set a. or 1. etc. as enum label
+\usepackage{enumitem}
+
+\pagestyle{empty}
+
+\usepackage{ulem}
+% go back to italics for emphasis, though
+\normalem
+
+\usepackage{natbib}
+
+\setlength{\footnotesep}{2em}
+
+\newcommand\comment[1]{}
+\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
+
+\newcommand\nan{\textsf{NaN}}
+\renewcommand\inf{\textsf{inf}}
+
+\newcommand\plusminus{\pm}
+
+\begin{document}
+
+% flInf flops?
+\title{NaN gates and flip FLOPS}
+\author{Dr.~Tom~Murphy~VII~Ph.D.\footnote{
+Copyright \copyright\ 2019 the Regents of the Wikiplia
+Foundation. Appears in SIGBOVIK 2019 with the
+XXX JOKE
+of the Association for Computational Heresy; 
+{\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
+\$-0.00 } }
+
+\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
+\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
+\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
+\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
+\newcommand\at{\ensuremath{\scriptstyle @}}
+
+\date{1 April 2019}
+
+\maketitle \thispagestyle{empty}
+
+\begin{abstract}
+what do I put here
+\end{abstract}
+
+\section*{Introduction}
+
+Mathematics is fundamental to computer science, and the foundation of
+mathematics is the real numbers; this is obvious from the name. One of
+computing's dirtiest secrets, however, is that computers themselves
+are not based on real numbers---rather, they are based on so-called
+``ones'' and ``zeroes'' combined with ``logic gates'' simulated with
+transistors. While this suffices for most practical purposes, it is
+unsatisfying from a theoretical perspective.
+
+Recently, some progress has been made by human geniuses on completely
+replacing integer calculations with calculations on real
+numbers\cite{fluint8}. While this removes many of the hacks present in
+modern software, there are still many components of the computer (e.g.
+RAM, registers, the {\it scroll lock} LED, a tiny USB-powered fan that
+can cool you on hot summer days or during particularly strenuous
+programming sessions) % XXX cite skymall etc?
+that are not integer-based.
+
+In this paper I give a new foundation for computing based solely on
+real numbers. This ...
+
+I begin with a brief reminder of the definition of real numbers,
+although the reader is expected to be familiar as these are pretty
+fundamental to everything. The approach of the paper is then to
+identify a pair of real numbers that have nice properties
+(Section~\ref{sec:distinguished}), and then to give mathematical
+operations on these numbers that parallel the logical operations
+typically used in the construction of computers
+(Section~\ref{sec:logical}). I then discuss how these operations
+can be implemented efficiently (Section~\ref{sec:representations}).
+I conclude with some wild speculation.
+
+\section{Real numbers}
+The real numbers are described by IEEE 754, most recently revised in
+AD 2008\cite{ieee754}. Every real number has a sign, a mantissa, and an
+exponent. Actually, this understates the elegance of real numbers,
+since there are a number of numbers, such as \nan\ (``not a number'')
+which are not of this form; \nan\ nas no sign nor mantissa nor
+exponent. We also have \inf\ and $-\inf$, which do have a sign, but no
+mantissa nor exponent. These are the infinite numbers that you get if
+you count very high or very low. Excitingly, we also have both
+positive and negative versions of 0. Some numbers have multiple
+representations, and most numbers cannot be represented at all.
+
+The real numbers have an equality operation {\tt ==}. This operation
+has some very exciting properties: It is not reflexive (\nan\ {\tt ==}
+\nan\ is false), and does not obey substitution (for $+0$ {\tt ==} $-0$
+is true, but $\sfrac{1}{+0}$ {\tt ==} $\sfrac{1}{-0}$ is false).
+
+Real numbers are an absolute joy to work with.
+
+Beautiful identities such as
+
+$0^{\nan * \inf} = 1$
+
+\subsection{Choosing some distinguished values} \label{sec:distinguished}
+
+use nan and inf because other real numbers can be ``easily confused''
+with one another. And do we really want ordering relationships on
+our fundamental particles, like $0 < 1$?
+
+% really lean into the joke that we choose the "number" NAN
+
+ha okay now there is just notes
 
@@
 [inf, inf, inf, inf] = Inf
+
+IEEE 754, 3.3
+minimal encoding size:
+b, radix, 2
+p, precision, ?
+emax, maximum exponent, ?
+emin = always 1 - emax
+
+I deduce:
+p = t + 1
+    (because you get one implied bit by normalization)
+t = 1
+% using the table
+emax = $2^{k-p-1} - 1$, which is $2^{4-2-1}-1 = 2^1-1 = 1$
+bias = emax
+emin = 1 - emax = 0
+
+I think there's still some ambiguity about whether we have
+emax = 1, emin = 0,
+or
+emax = 0, emin = -1,
+probably both work, but I think the former has more support
+in the docs.
+
+\comment{
+a normal value is
+(-1)^sign x b^{exp} x significand
+(I think significand = mantissa?)
+
+by standard, need:
+- sign bit
+- exponent could maybe be empty, though it's sort of implied
+that you just have emax=0, emin= -1, so one bit
+- m could probably be empty (no bits mantissa = 0)
+
+DID YOU KNOW?
+``every finite floating point numbe is an integral multiple
+of the smallest subnormal magnitude b^{emin} * b^{1 - p}''
+
+fewest exponent bits w?
+The range of the encoding's biased exponent E shall include:
+- every integer between 1 and 2^{w}−2, inclusive, to encode normal numbers
+- the reserved value 0 to encode +/-0 and subnormal numbers
+- the reserved value 2^{w}−1 to encode +/-inf and NaNs
+
+so we need at least 0,
+and a reserved value for inf & nan,
+so if w = 1, we'd have 0 and 2^1-1 = 1,
+but this seems to violate the spec because you need
+``every integer between 1 and 2^w-2 inclusive'', and here 2^w-2 is 0,
+so either this interval is inverted (arguable?) and that's why it
+doesn't include its endpoints, or it's in violation
+but with w = 2,
+
+2^2 - 1 = 0b11 = 3 (inf, nan)
+1 to 2^2 - 2 = 0b01, 0b10 = 1,2 (normals)
+0 = 0b00 = 0 (zero, subnormals)
+
+(Note this is the biased exponent E, not e)
+
+2 bits looks clearly legal.
+
+ s       w bits           t bits
+[sign][ exponent ][  trailing significand T   ]
+
+when e=0b11, t determines inf or nan.
+- nan is any nonzero sign is ignored.
+(recall, nans can carry info, so that is why they get
+the nonzero case)
+- inf is zero
+
+so we clearly need 1 t-bit to encode inf vs nan,
+but it seems that this is enough
+
+so probable case is four bits:
+
+ - sign
+ - two exponent bits
+ - one mantissa bit
+
+The sign is always 0 (positive infinity; doesn't matter for nan)
+and the exponent is always 11 (since we always encode nan or
+positive inf). Unless we do any intermediate calculations
+(could check what all the intermediate values are in the above
+expressions and make sure they fit in 4 bits), a hardware
+realization of this could just hard-wire 3 of 4 bits? Is that
+better or worse?
+
+so another version of this joke is that a NAND gate is a super
+fast FPU for calculating inf - max(x + y, -inf) on four-bit
+IEEE 754 where the top 3 MSBs are fixed (although be careful,
+nan = 1 and inf = 0 in this representation. Is that maybe a
+better choice for this presentation? it just dualizes some
+gates)
+
+}
+
+\section{The binary4 representation}
+
+IEEE 754 natively defines several bit widths for floating-point
+values, such as the 32-bit binary32 (aka ``single-precision float'')
+and 64-bit binary64 (aka ``double-precision float''). The
+specification is parameterized to allow other bit widths; for example,
+half-precision 16-bit floats are common in GPU code for machine
+learning applications (XXX cite). Smaller floats sacrifice precision,
+but require less space and allow faster calculations. For our purposes
+in this paper, since we only need to represent the two numbers \nan\ and
+\inf, we are interested in the smallest possible representation.
+
+% TODO: what if each bit in the float is itself represented as
+% inf or NaN?
+
+This section describes the binary4 representation, a four-bit floating
+point number that is clearly allowed by the IEEE 754 standard.
+
+The representation of any floating-point number has a single sign bit,
+some number $w$ of exponent bits, and some number $t$ of mantissa
+bits. For binary32, $w = 8$ and $t = 23$; and with the sign bit we
+have $23 + 8 + 1 = 32$ bits as expected. We need at least a sign bit,
+but what are the smallest permissible values of $w$ and $t$?
+
+The most stringent constraint on $w$ comes in IEEE-754-2008~3.4, which
+states
+\begin{quote}
+  The range of the encoding's biased exponent E shall include:
+  \begin{enumerate}[label=---]
+    \item every integer between 1 and $2^w - 2$, inclusive, to encode
+      normal numbers
+    \item the reserved value 0 to encode $\plusminus 0$ and subnormal
+      numbers
+    \item the reserved value $2^w - 1$ to encode $\plusminus \infty$
+      and NaNs.
+  \end{enumerate}
+\end{quote}
+
+$E$ is the binary number encoded by $w$. It must include at least the
+two special values consisting of all zeroes and all ones (second and
+third clause). A conservative reading of ``every integer between 1 and
+$2^w - 2$'' seems to require that $1 \leq 2^w - 2$ (otherwise how
+could the interval be inclusive of its endpoints?), which would imply
+that $w$ is at least 2. (However, see Section~\ref{sec:binary3} for
+the hypothesized case where $w=1$.)
+
+The representation of \nan\ and \inf\ are distinguished by the value
+of $t$ when $E$ is all ones. We certainly need to distinguish these,
+so $t = 1$ is the minimal size.
+
+We have one sign bit, two exponent bits, and one mantissa bit, for a
+total of four. Since ``single precision'' is 32 bits, ``half
+precision'' is 16, 4 bits is ``eighth precision.'' Given how nicely
+all this works out, shouldn't there be an {\tt eighth} base type in
+most modern programming languages and GPUs? Since there are so few
+values representable, it would be practical for all the standard
+operations to be done in constant time via table lookups. All 16
+possible values are given in Figure~\ref{fig:binary4}.
+
+\begin{figure*}[tp]
+\begin{tabular}{|l|c|c|c|c|c|c|}
+  \hline
+  parameter & {\bf binary4} & binary16 & binary32 & binary64 & binary128 & binary$_k$ \\
+  \hline
+  $k$, storage in bits & {\bf 4} & 16 & 32 & 64 & 128 & multiple of 32 \\
+  $p$, precision in bits & {\bf 2} & 11 & 24 & 53 & 113 & $k$ - round(4 * log$_2$(k)) + 13 \\
+  $emax$, maximum exponent e & {\bf 1} & 15 & 127 & 1023 & 16383 & $2^{k-p-1}$ - 1 \\
+  $bias$ = $E$ - $e$ & {\bf 1} & 15 & 127 & 1023 & 16383 & emax \\
+  $sign bits$ & {\bf 1} & 1 & 1 & 1 & 1 & 1 \\
+  $w$, exponent width & {\bf 2} & 5 & 8 & 11 & 15 & round(4 * log$_2$(k)) - 13 \\
+  $t$, trailing significand width & {\bf 1} & 10 & 23 & 52 & 112 & k - w - 1 \\
+  $k$, storage width & {\bf 4} & 16 & 32 & 64 & 128 & 1 + w + t \\
+\hline
+\end{tabular}
+\caption{Parameters for the newly-introduced {\bf binary4} encoding
+  for IEEE-754, compared to the standard widths (see Table~3.5 in the
+  standard\cite{ieee754}). } \label{fig:binary4}
+\end{figure*}
+
+% bias = 1
+
+\begin{figure}[h]
+\begin{tabular}{|l@{\,}c@{\,}l|p{2.5in}|}
+\hline
+  $s$ & $E$ & $T$ & value \\
+  \hline
+0 & 00 & 0 &   $+0$ \\
+0 & 00 & 1 &    subnormal: % $2^{emin} * 2^{1-p} * T$ =
+                $2^0 * 2^{1-2} * 1$ =
+                $1 * \sfrac{1}{2} * 1$ = 0.5 \\
+0 & 01 & 0 &    normal: % $2^{E - bias} * (1 + 2^{1-p} * T)$
+                $2^0 * (1 + \sfrac{1}{2} * 0)$ = 1 \\
+0 & 01 & 1 &    $2^0 * (1 + \sfrac{1}{2} * 1)$ = 1.5 \\
+0 & 10 & 0 &    $2^1 * (1 + \sfrac{1}{2} * 0)$ = 2 \\
+0 & 10 & 1 &    $2^1 * (1 + \sfrac{1}{2} * 1)$ = 3 \\
+0 & 11 & 0 &   $+\infty$ \\
+0 & 11 & 1 &    \nan \\
+1 & 00 & 0 &   $-0$ \\
+1 & 00 & 1 &   $-0.5$ \\
+1 & 01 & 0 &   $-1$ \\ 
+1 & 01 & 1 &   $-1.5$ \\
+1 & 10 & 0 &   $-2$ \\
+1 & 10 & 1 &   $-3$ \\ 
+1 & 11 & 0 &   $-\infty$ \\
+1 & 11 & 1 &    \nan \\
+\hline
+\end{tabular}
+\caption{All 16 values representable in binary4 floating-point.
+  The format works reasonably well even at this very low precision,
+  although note how many of the values are not finite.} \label{fig:allvalues4}
+\end{figure}
+
+\subsection{The hypothesized binary3 format} \label{sec:binary3}
+The IEEE-754 representation clearly requires a sign bit, and for this
+purpose we need at least one bit for the mantissa in order to
+distinguish \nan\ and \inf. It is perhaps a stretch of the wording,
+but arguably the spec permits a 1-bit exponent ($w = 1$). To
+rationalize this we need to interpret the phrase ``every integer
+between $1$ and $2^w-2$ inclusive'' (that is, between 1 and 0 inclusive)
+as denoting the empty set.
+
+With one bit for sign, exponent, and mantissa, we can represent just 8
+different values. Here $emax$ is $0$, and the standard clearly
+requires $emin = 1 - emax$, so $emin = 1$. Certainly fishy for $emin$
+to be larger than $emax$, but we can just not stress out about it; the
+representable values are all reasonable-looking
+(Figure~\ref{fig:allvalues3}).
+
+% here, precision is still 2 (just one mantissa bit)
+% so emax is 2^{k-p-1}-1, which is 2^{3-2-1}-1 = 0
+% bias = 0
+% emin is 1 - emax = 1 (!?)
+% this is kind of absurd, it may still "work"?
+
+\begin{figure}[h]
+\begin{tabular}{|l@{\,}c@{\,}l|p{2.5in}|}
+\hline
+  $s$ & $E$ & $T$ & value \\
+  \hline
+0 & 0 & 0 &   $+0$ \\
+0 & 0 & 1 &    subnormal: % $2^{emin} * 2^{1-p} * T$ =
+                $2^1 * 2^{1-2} * 1$ =
+                $2 * \sfrac{1}{2} * 1$ = 1 \\
+0 & 1 & 0 &   $+\infty$ \\
+0 & 1 & 1 &    \nan \\
+1 & 0 & 0 &   $-0$ \\
+1 & 0 & 1 &   $-1$ \\
+1 & 1 & 0 &   $-\infty$ \\
+1 & 1 & 1 &    \nan \\
+\hline
+\end{tabular}
+\caption{All 8 values of the hypothetical binary3 representation.
+  There are no normal values; the only finite values are the
+  positive and negative zero and a single subnormal which
+  denotes $1$ (or $-1$).
+} \label{fig:allvalues3}
+\end{figure}
+
+% TODO: We could do without a mantissa if we didn't care about
+% distinguishing NaN and inf. With an empty t, we should
+% consider it to have value 0 (true? maybe it is defined as
+% a product so the empty string is the multiplicative unit 1?)
+% which means that we encode +/- inf, and nan cannot be encoded.
+% Now the operations will not be closed since e.g. 0/0 needs to
+% result in nan.
+
+% TODO: What if we have no mantissa or exponent, just a sign bit?
+% When E=0 and T=0, we can only encode +/- zero. This definitely
+% involves some leaps (and is not closed as above), but could
+% we find expressions that work for just +/- 0?
+
+% hardware notes:
+% http://openocd.org/ looks like a good way to do in-device
+% programming like over SWD, because it runs on the pi. A board
+% could have jumpers or even software route to different
+% gpio pins. (You would need to run at 3v3 for this of course.)
+% I guess make sure the chip is compatible, but there are lots
+% explicitly listed.
+
+% Some supported devices:
+% AT91SAM7 - no fp in the M7
+% at91sam3u - no fp here either :/
+
+% LPC1700, LPC1800,
+% LPC2000, LPC4300, , STR7x, STR9x, LM3, STM32x and
+% EFM32
+
+% ok the STM32f3 is supported, has FPU, 51 IO pins,
+% specifically like the
+% https://www.mouser.com/ProductDetail/STMicroelectronics/STM32F303RDT7?qs=%2Fha2pyFaduikpRt0bOqlahEvgTujL2BrdYPen6RZRRk%3D
+
+% here is an example breakout:
+% http://ebrombaugh.studionebula.com/embedded/stm32f303breakout/index.html
+% requires a crystal, voltage regulator, and some filter capacitors,
+% but looks manageable. (And the chip has some 5V tolerant TTL I/O
+% pins too, which could be nice for other projects)
+% also maybe a ferrite bead (for input to analog VDDA, so probably
+% not needed for this project)
+
+Idea: Make a FPU. How small could it be?
+I guess the simplest thing that could be described as faithful
+would compute the NAND function of two binary3 numbers, but on
+the entire domain. There are six input pins, three output pins,
+and a table size of 8^2 = 64.
+
+inf is logic true, 010 binary3
+nan is logic false, 011 binary3
+
+A nice simple nand function is
+f(x, y) = inf - max(x + y, -inf)
+
+What is a simple implementation of this function?
+It can only output inf or nan, so that's good. Of the three
+output pins, the first is always 0, the second always 1, so
+at worst we would just need a table of 64 bits for this last
+value.
+
+Further simplification:
+
+if isfinite(x) && isfinite(y)
+then
+  // inf - (x + y) = inf
+  0
+else
+// x + y is nan when x is nan, y is nan, or x and y are
+// infinites with different signs. If they are both -inf,
+// then we have max(-inf, -inf) anyway, which is the same
+// as max(nan, -inf). So we have
+  if x == inf && y == inf
+    // both positive infinities
+    // inf - inf = nan
+    1
+  else
+    // inf - -inf = inf
+    0
+
+So ultimately this function only returns 1 in the case
+that both inputs are exactly +inf, the pattern 0 1 0.
+
+If the inputs are a0 a1 a2, b0, b1, b2, and outputs
+are c0, c1, c2, then:
+c0 = 0
+c1 = 1
+c2 = !a0 && !a2 && !b0 && !b2 && a1 && b1
+
+// Could adapt find.sml to find solutions for this,
+// using only NAND. Only 2^6 = 64 possible values
+// for these variables
+
+let x = nand(!a0, !a2)
+let y = nand(!b0, !b2)
+let z = nand(a1, b1)
+
+let w = Not(Nand(Not(x), Not(y)))
+let out = Not(Nand(w, Not(z))
+
+... this is pretty much a straightforward Nand construction.
+I guess it doesn't really simplify much across units.
+
+or with more gates,
+Not(Or(z, Or(y, x)))
+
+and I mean obviously it's just
+And(Not (Or (a0, a2, b0, b2)), a1, b1)
+if you have all the gates directly.
+
+we can certainly make chips implementing each of these gates,
+since we have floating point implementations of each, but
+then we sort of lose the final joke since we'd just build
+a floating point unit that computes the NAND instruction.
+
+so let's try the pure-nand approach.
+
+implementing the nots also with nands, this works out to 14
+gates, which is ``just'' three chips. very reasonable. I was
+actually hoping for it to be a little more ridiculous.
+(maybe there is a lower pin-count device with an easier
+package? programming them is still going to be a huge pain)
+
+if we used the meta-nand gates, we'd have 14 * 3 = 42 chips,
+which is OTOH probably too many.
+
+with this meta-fpu, we have 2 binary3 inputs and 1 binary3
+output, but each ``bit'' is represented as a binary3 number
+itself. So this makes (3 * 2 + 1) * 3 = 27 wires to interface
+with it. This is exactly as many GPIO the pi has, so that
+is good, although IIRC a few pi pins have special annoyances,
+and it doesn't leave us any space for a programmer.
+Maybe just best to make a separate programmer and connect it
+up to different test points manually. Then we could reuse
+the programmer pretty easily.
+
+Note that in a pinch we can ``cheat'' pretty easily here
+since some of the pins have constant values. Probably justifiable
+is that the meta-nand always outputs 0 (nan) for g and 1 (inf) for h,
+plus two of the bits of i are fixed because it will only represent
+nan or inf. But really would be ``prettiest'' to not make this cheat,
+I mean, it's a pretty silly excursion anyway.
+
+oh jk there is actually gpio0 too, so 28 pins on pi.
+
+#  0 & 0 & 0 &   $+0$ \\
+#  0 & 0 & 1 &   +1
+#  0 & 1 & 0 &   $+\infty$ \\
+#  0 & 1 & 1 &    \nan \\
+#  1 & 0 & 0 &   $-0$ \\
+#  1 & 0 & 1 &   $-1$ \\
+#  1 & 1 & 0 &   $-\infty$ \\
+#  1 & 1 & 1 &    \nan \\
+
+% would be cool if the meta NAND gate were addressable
+% from the rasberry pi (well duh of course it is?) because
+% then we could also use it to simulate itself in software
+
+\nocite{ieee754}
+\bibliography{nand}{}
+\bibliographystyle{plain}
+
+\end{document}

### 2019-03-04 trunk/chess/blind/eval-unblinder.cc
Blind player that always reports one king for each side

@@ -21,6 +21,8 @@ using namespace std;
 
 using int64 = int64_t;
 
+static constexpr bool USE_SINGLE_KING = true;
+
 // If false, we don't even look at the game.
 static bool Eligible(const PGN &pgn) {
   // Ignore games that don't finish.
@@ -139,7 +141,7 @@ int main(int argc, char **argv) {
       // XXX something is wrong! Error rate is almost as high (27.24 vs 27.61)
       // as just guessing a totally empty board, which we definitely do better
       // than. I guess maybe it's not being simulated correctly?
-      Position guess = unblinder->Unblind(bits);
+      Position guess = unblinder->Unblind(USE_SINGLE_KING, bits);
 
       int piecem = 0;
       for (int r = 0; r < 8; r++) {

### 2019-03-04 trunk/chess/blind/unblinder.h

@@ -16,8 +16,10 @@ struct Unblinder {
   static void Layer64(uint64_t pos, std::vector<float> *v);
   
   // Note that the resulting position may be invalid (e.g. multiple
-  // black kings).
-  virtual Position Unblind(uint64_t pos) const = 0;
+  // black kings). If single_king is true, then constructs positions
+  // such that there is exactly on king per side (but could still
+  // be invalid because of e.g. mutual check).
+  virtual Position Unblind(bool single_king, uint64_t pos) const = 0;
 
   // Sample n predicted positions that are valid. Valid means:
   //  - It is black's move iff blackmove is true

### 2019-03-04 chess/blind/paper.tex
Blind player that always reports one king for each side

 make -j eval-unblinder.exe && ./eval-unblinder.exe net.val
.!Evaluating model net.val...
Loaded 50000 positions in 122.59s
Loaded model in 0.10s
Ran eval in 2.04s
Reading [net.val]
net.val: 4 layers.
net.val: num nodes: 64 1024 12288 567 837
net.val: indices per node/fns: 49 LEAKY_RELU 36 LEAKY_RELU 127 LEAKY_RELU 78 LEAKY_RELU
Read from net.val.
Invert index:
Check it:
ModelInfo [339885 rounds, 223224128 examples]
Over 50000 positions:
  9584 exactly correct (19.17%)
  161166 piece mistakes (3.22/pos)
  1630 castling mistakes (0.03/pos)
  19014 move mistakes (0.38/pos)
  181810 total mistakes (3.64/pos)

$ make -j eval-unblinder.exe && ./eval-unblinder.exe net-before-vacuum.val
make: 'eval-unblinder.exe' is up to date.
Evaluating model net-before-vacuum.val...
Loaded 50000 positions in 118.80s
Loaded model in 3.99s
Ran eval in 332.10s
Reading [net-before-vacuum.val]
net-before-vacuum.val: 4 layers.
net-before-vacuum.val: num nodes: 64 1024 12288 567 837
net-before-vacuum.val: indices per node/fns: 64 LEAKY_RELU 1024 LEAKY_RELU 12288 LEAKY_RELU 567 LEAKY_RELU
Read from net-before-vacuum.val.
Invert index:
Check it:
ModelInfo [146999 rounds, 9407936 examples]
Over 50000 positions:
  10601 exactly correct (21.20%)
  156052 piece mistakes (3.12/pos)
  1542 castling mistakes (0.03/pos)
  18818 move mistakes (0.38/pos)
  176412 total mistakes (3.53/pos)

without concurrent processes:
net.val: 1.61s  (something like 1035/sec/core)
net-before-vacuum: 229.85s (something like 7/sec/core)

single_kings net.val:
Over 50000 positions:
  9642 exactly correct (19.28%)
  162804 piece mistakes (3.26/pos)
  1633 castling mistakes (0.03/pos)
  19014 move mistakes (0.38/pos)
  183451 total mistakes (3.67/pos)

It's actually kind of heartening that it performs worse with this approach,
though we are more likely to get the whole board correct.

(cite #kingme)

% all bits set to 1 (with single_kings)
% (without single_kings, only difference is that the bottom right becomes a
% rook; no white queen)
\includegraphics{allon}

### 2019-03-04 chess/blind-player.cc

@@ -19,8 +19,6 @@ using int64 = int64_t;
 using Move = Position::Move;
 using namespace std;
 
-// TODO: 
-
 static constexpr bool VERBOSE = false;
 
 namespace {
@@ -31,8 +29,11 @@ static Unblinder *GetUnblinder() {
   return the_unblinder;
 }
 
-struct BlindYoloPlayer : public StatelessPlayer {
-  BlindYoloPlayer() : rc(PlayerUtil::GetSeed()) {
+struct BlindPlayer : public StatelessPlayer {
+  BlindPlayer(const string &name,
+	      bool single_king) : name(name),
+				  single_king(single_king),
+				  rc(PlayerUtil::GetSeed()) {
     fish.reset(new Stockfish(20, 1'000'000));
     CHECK(fish.get());
     unblinder = GetUnblinder();
@@ -100,8 +101,8 @@ struct BlindYoloPlayer : public StatelessPlayer {
   }
   
   Move MakeMove(const Position &orig_pos) override {
-    uint64 blinded = Unblinder::Blind(orig_pos);
-    Position predicted = unblinder->Unblind(blinded);
+    const uint64 blinded = Unblinder::Blind(orig_pos);
+    Position predicted = unblinder->Unblind(single_king, blinded);
 
     // Force the current player.
     predicted.SetBlackMove(orig_pos.BlackMove());
@@ -123,13 +124,13 @@ struct BlindYoloPlayer : public StatelessPlayer {
 	return m;
 
       if (VERBOSE)
-	printf("Move %s not legal. Predicted:\n%sActual:\n%s",
+	printf("\nMove %s not legal. Predicted:\n%sActual:\n%s",
 	       predicted.ShortMoveString(m).c_str(),
 	       predicted.BoardString().c_str(),
 	       orig_pos.BoardString().c_str());
     } else {
       if (VERBOSE)
-	printf("Predicted board not OK:\n%sActual:\n%s",
+	printf("\nPredicted board not OK:\n%sActual:\n%s",
 	       predicted.BoardString().c_str(),
 	       orig_pos.BoardString().c_str());
     }
@@ -141,15 +142,18 @@ struct BlindYoloPlayer : public StatelessPlayer {
   }
 
   string Name() const override {
-    return "blind_yolo";
+    return name;
   }
 
   string Desc() const override {
-    return StringPrintf("Predict a board state. If valid (or can be made "
+    return StringPrintf("Predict a board state%s. If valid (or can be made "
[…]

### 2019-03-06 chess/papers/blind.tex
Start on colorblind paper

\documentclass[twocolumn]{amsart}
\usepackage[top=0.5in, left=0.45in, right=0.45in, bottom=0.5in]{geometry}

\usepackage{url}
% \usepackage{code}
% \usepackage{cite}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{chessboard}

% \usepackage{chessfs}
\usepackage{adjustbox}

% Define black versions of pieces for inline use. Gross, but it works.
\newcommand{\Pawn}[1][1.3ex]{%
\adjustbox{Trim=4.3pt 2.6pt 4.3pt 0pt,width=#1,margin=0.2ex 0ex 0.2ex 0ex}{\BlackPawnOnWhite}%
}%
\newcommand{\Rook}[1][1.58ex]{%
\adjustbox{Trim=3.2pt 2.2pt 3.2pt 0pt,width=#1,raise=0ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackRookOnWhite}%
}%
\newcommand{\Knight}[1][1.85ex]{%
\adjustbox{Trim=2.3pt 2.35pt 2.5pt 0pt,width=#1,raise=-0.03ex,margin=0.14ex 0ex 0.14ex 0ex}{\BlackKnightOnWhite}%
}%
\newcommand{\Bishop}[1][1.79ex]{%
\adjustbox{Trim=2.3pt 2pt 2.3pt 0pt,width=#1,raise=-0.12ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackBishopOnWhite}%
}%
\newcommand{\Queen}[1][2.05ex]{%
\adjustbox{Trim=1.2pt 2.2pt 1.2pt 0pt,width=#1,raise=-0.08ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackQueenOnWhite}%
}%
\newcommand{\King}[1][1.95ex]{%
\adjustbox{Trim=2pt 2pt 2pt 0pt,width=#1,raise=-0.06ex,margin=0.13ex 0ex 0.13ex 0ex}{\BlackKingOnWhite}%
}%

\interfootnotelinepenalty=0

% lets me explicitly set a. or 1. etc. as enum label
\usepackage{enumitem}

\pagestyle{empty}

\usepackage{ulem}
% go back to italics for emphasis, though
\normalem

\usepackage{natbib}

\setlength{\footnotesep}{2em}

% \newcommand\comment[1]{}
\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}

\begin{document}

\title{Color- and piece-blind chess}
\author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
Appears in The Journal Of LaTeX Class Files with the insufficient
material of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1 tempo}

\setchessboard{showmover=false}

\newcommand\checkmate{\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}

\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
\newcommand\at{\ensuremath{\scriptstyle @}}

\date{1 April 2019}

\maketitle \thispagestyle{empty}

% \begin{abstract}
% CHESSMATE.
% \end{abstract}

\section*{Impressing humans}

What better way for humans to impress each other with their brains,
especially in movies, than to play chess---and to shout dramatically
CHECKMATE! upon surprise-checkmating their opponent? Well, one way is
to play chess while disadvantaged somehow, for example, by punching
each other in the face repeatedly during the game to impair brain
function (see Chess Boxing~\cite{chessboxing}). Another common
distraction is to play a multitude of games against many opponents at
the same time, in a so-called ``simultaneous exhibition.'' The idea is
that this is more challenging because of the need to maintain mental
state for so many games at once, whereas your opponents only need to
maintain state for one game. In truth, simultaneous exhibitions easily
fall to a ``man-in-the-middle attack.'' If the purported genius simply
creates a bipartite matching of the games played with the white pieces
and the games played with black, he can mechanically forward moves
between these pairs of boards. This requires only constant state (see
next section) per pair of games, and guarantees an even score for the
exhibition. So that's not very impressive.

Another disadvantage that humans sometimes use to impress each other
is a blindfold; they only hear the opponent announce the moves and
must keep the position on the board in their mind's eye, both for
the sake of remembering it and while exploring potential moves.
% This is effective, since although the .
Disadvantages can be combined, such as in the final scene of the 1988
documentary {\it Bloodsport} where Jean Claude van Damme is blinded by
an illicit foreign substance during the final martial art
battle.\footnote{JCVD does not play chess on camera, but it is implied
  that he is also holding a simultaneous exhibition between rounds in
  a different room of the underground Hong Kong illegal karate
  complex.}

\section{Impressing computers}

In contrast, it is much more difficult to impress computers or impress
people with computers. For example, it is well known that computers
function {\em better} when struck, so chess boxing becomes trivial.
Playing multiple games simutaneously is an easy extension of playing a
single game, and when it comes to computers playing chess, largely,
the jig is up; it is now easy for chess programs, running on consumer
hardware, to defeat the strongest human players. Blindfold chess is
the natural interface for a chess computer; it is actually {\em much
  more difficult} to have the computer interpret the opponent's move
by visually studying a physical board!

Computers can still impress us and each other with computational chess
feats other than playing. For example, computers are largely concerned
with filling up their memories with efficiently encoded data.

compactly storing a position. constant state

I think a version I liked was 64 bits (this gives the positions of all
pieces), followed by 32 records giving the piece values (there are 6:
pawn, knight, bishop, rook, queen, king) and colors. To pick between
the 12 possibilities we need 4 bits, and there is some slack here of
course. anyway, that's 64 + 32*4 = 192 bits, plus 15 bits for other
state. Note we could use the c_rook trick to save 4 bits for castling
rights (14 possibilities), not to mention a en_passantable pawn (15)
and a king whose turn it is (even 16). Clear slack here, but that
would leave only the 50-move rule at 7 bits, for a total of 199.
Wikipedia says the huffman approach is ``maximum of 204 bits, and
often much less.'' but the article also contains many bugs, like
the misconception that there can only be four rooks.

% 
% % make -j eval-unblinder.exe && ./eval-unblinder.exe net.val
% .!Evaluating model net.val...
% Loaded 50000 positions in 122.59s
% Loaded model in 0.10s
% Ran eval in 2.04s
% Reading [net.val]
% net.val: 4 layers.
% net.val: num nodes: 64 1024 12288 567 837
% net.val: indices per node/fns: 49 LEAKY_RELU 36 LEAKY_RELU 127 LEAKY_RELU 78 LEAKY_RELU
% Read from net.val.
% Invert index:
% Check it:
% ModelInfo [339885 rounds, 223224128 examples]
% Over 50000 positions:
%   9584 exactly correct (19.17%)
%   161166 piece mistakes (3.22/pos)
%   1630 castling mistakes (0.03/pos)
%   19014 move mistakes (0.38/pos)
%   181810 total mistakes (3.64/pos)
% 
% % $ make -j eval-unblinder.exe && ./eval-unblinder.exe net-before-vacuum.val
% make: 'eval-unblinder.exe' is up to date.
% Evaluating model net-before-vacuum.val...
% Loaded 50000 positions in 118.80s
% Loaded model in 3.99s
% Ran eval in 332.10s
% Reading [net-before-vacuum.val]
% net-before-vacuum.val: 4 layers.
% net-before-vacuum.val: num nodes: 64 1024 12288 567 837
% net-before-vacuum.val: indices per node/fns: 64 LEAKY_RELU 1024 LEAKY_RELU 12288 LEAKY_RELU 567 LEAKY_RELU
% Read from net-before-vacuum.val.
% Invert index:
% Check it:
% ModelInfo [146999 rounds, 9407936 examples]
% Over 50000 positions:
%   10601 exactly correct (21.20%)
%   156052 piece mistakes (3.12/pos)
%   1542 castling mistakes (0.03/pos)
%   18818 move mistakes (0.38/pos)
%   176412 total mistakes (3.53/pos)
% 
% 
% 
% without concurrent processes:
% net.val: 1.61s  (something like 1035/sec/core)
% net-before-vacuum: 229.85s (something like 7/sec/core)
% 
% 
% single_kings net.val:
% Over 50000 positions:
%   9642 exactly correct (19.28%)
%   162804 piece mistakes (3.26/pos)
%   1633 castling mistakes (0.03/pos)
%   19014 move mistakes (0.38/pos)
%   183451 total mistakes (3.67/pos)
% 
% It's actually kind of heartening that it performs worse with this approach,
% though we are more likely to get the whole board correct.
% 
% (cite #kingme)
%

% all bits set to 1 (with single_kings)
% (without single_kings, only difference is that the bottom right becomes a
% rook; no white queen)
\includegraphics{blind-allon}

\bibliography{chess}{}
\bibliographystyle{plain}

\end{document}

### 2019-03-07 chess/papers/blind.tex
Some paper writing, got myself side-tracked

@@
 
-\section*{Impressing humans}
+\section{Impressing humans}
 
@@
 Another disadvantage that humans sometimes use to impress each other
-is a blindfold; they only hear the opponent announce the moves and
-must keep the position on the board in their mind's eye, both for
-the sake of remembering it and while exploring potential moves.
+is a blindfold (worn over the eyes). In this predicament they only
+hear the opponent announce moves and must keep the position on the
+board in their mind's eye, both for the sake of remembering it and
+while exploring potential moves.
 % This is effective, since although the .
@@
 In contrast, it is much more difficult to impress computers or impress
-people with computers. For example, it is well known that computers
-function {\em better} when struck, so chess boxing becomes trivial.
+people with computers. When it comes to computers playing chess,
+largely, the jig is up; it is now easy for chess programs, running on
+consumer hardware, to defeat the strongest human players. It is well
+known that striking a computer actually {\em fixes} it, so chess
+boxing becomes trivial. Blindfold chess is the natural interface for a
+chess computer; it is actually {\em much more difficult} to have the
+computer interpret the opponent's move by visually studying a physical
+board!
+
 Playing multiple games simutaneously is an easy extension of playing a
-single game, and when it comes to computers playing chess, largely,
-the jig is up; it is now easy for chess programs, running on consumer
-hardware, to defeat the strongest human players. Blindfold chess is
-the natural interface for a chess computer; it is actually {\em much
-  more difficult} to have the computer interpret the opponent's move
-by visually studying a physical board!
+single game, although in principle the scale of such a thing could
+still be impressive. With a modern chess engine, it is easy to scale
+to an arbitrary number of games, since the exhibitor can make progress
+by observing one of the boards, computing a strong move, and playing
+it; this requires $O(0)$ space because all of the state is stored
+externally in the exhibition itself. 
+% If playing blindfold, then the
+% main cost is storing the state of each game. 
+However, we run the risk of losing the tournament (other players may
+be yet stronger computers). The man-in-the-middle attack remains an
+efficient way way to minimize loss (ensuring an exactly even score).
+Suppose that there are $n$ games $G$ being played. We generate a
+bipartite matching that pairs games $\langle G_w, G_b \rangle$ (where
+we play as white against Bob and black against Walice, respectively).
+This is $n/2$ pairs, and we record which pair each game belongs to, in
+$n \log_2(n/2)$ bits. Now, to make progress, we look at a game $G_i$
+where it is our move, and consult our state for this pair. If the
+pair has the has-move bit (actually just store it as the move itself),
+
+ok, so probably the most efficient thing is to store instead the
+move played in each row. Moves are like 64*64*4 (promotion) = 6+6+2 bits.
+The only way to really motivate the board approach would be if it
+allows us to avoid the log(n/2) factor from storing the matching?
+
+so you have a set of black and white games, same number of each.
+goal is to get an even score, while using O(n) space.
+So we have an invariant that each of the games-as-white is paired
+with at least one compatible game-as-black (and vice versa). Of
+course at the beginning all pairs are compatible, and then white
+makes a move; the games are still compatible because any of these
+moves can be played on the black side. We just need to make sure
+that we don't play e.g. e4 in all the games, because we need enough
+black opponents to be playing against d4s and Nf3s and so on.
+So, after all the white players have made their moves, we have
+those boards, and need to make moves for black. For each, just pick
+a compatible board (i.e., something that represents a legal move)
+and play it. When we do so, we update the board to a state like
+``I made this move as white.'' (So at the very beginning, all boards
+are in the starting position with like ``I made this move as black,''
+of course not really, more like ``white opponent to move'').
+This is easy on the first move because all choices (that are in
+the state ``copied this move from white'').
+
+Now we wait for black players to make their moves. (Note waiting
+can be done in constant space. You just compare each board that
+has ``black opponent to move'' with the actual board state; if
+the move is compatible then you update it. But might you confuse
+the identities of two games? It would need to be the case that
+there exists board states $B_1$, $B_2$, $B$ where $B$ can
+legally follow $B_1$ and $B_2$ but $B_1$ and $B_2$ are different.
+Of course this is possible! $B$ could be like
+
+\chessboard[tinyboard, setfen=k7/8/8/8/8/8/8/K7 w - -, showmover]
+
+and then we have $B_1$ and $B_2$ as
+
+\chessboard[tinyboard, setfen=1k6/8/8/8/8/8/8/K7 b - -, showmover]
+\chessboard[tinyboard, setfen=Rk6/8/8/8/8/8/8/K7 b - -, showmover]
+
+And so if we misassign $B$ to $B_1$, and then later see $B'$:
+
+\chessboard[tinyboard, setfen=R7/1k6/8/8/8/8/8/K7 w - -, showmover]
+
+We are stuck because $B'$ cannot follow $B_2$.
+
+So if we're actually going to do it with this method, we would
+need to be able to compute the bipartite matching looking at all
+the constraints. There certainly exists one, and any one will do. 
+(Actually what we want is called a ``perfect matching.'')
+So the question is, can we compute a perfect matching in constant
+space? I mean, it seems like not. How could we even {\em store}
+the matching? Well, we don't actually need to store the matching
+because it's not *literally* what we need; we just want to have
+assigned the games according to a perfect match. So one approach
+is like, pick a random assignment. If we get stuck, just start
+over (so we would have to keep two board states, in order to
+be able to ``undo'', but that is still constant) and commit
+the whole thing if we succeed. We're actually quite likely to
+succeed since ambiguities are pretty rare. So this may work
+fine in practice. The main thing we achieve here is that we
+don't need to store the matching itself between rounds; the
+matching is ``expensive'' because it contains an index, which
+is logarithmic size in the number of games. So, with this
+method, we get $O(n)$ space, which is actually around $200 n$.
+
+For each pair, we store the state of the board $B$
+as last we saw it in $G_w$ or $G_b$.
+
+both Walice and Bob have made their moves (that is, it is white's turn
+in $G_w$ and
+
+$G_i$, we look up which pair $p$ it is, and look at $B_p$. This board
+
+Note that
 
@@
 course. anyway, that's 64 + 32*4 = 192 bits, plus 15 bits for other
-state. Note we could use the c_rook trick to save 4 bits for castling
-rights (14 possibilities), not to mention a en_passantable pawn (15)
+state. Note we could use the c\_rook trick to save 4 bits for castling
+rights (14 possibilities), not to mention a en\_passantable pawn (15)
 and a king whose turn it is (even 16). Clear slack here, but that
@@
 the misconception that there can only be four rooks.
+
+So now comes the main idea of the paper, which is also spoilered in the
+very clear paper title. What if we represented boards {\em only} as the
+64-bit mask telling us what squares are occupied? The encoding is very
+lossy, of course, but it often contains enough information to deduce
+the state of the board, or ...
 
@@
 % rook; no white queen)
-\includegraphics{blind-allon}
+\includegraphics[width=0.9 \linewidth]{blind-allon}
 

### 2019-03-08 chess/papers/blind.tex
More hacking on sidetrack, need to rewrite again

@@
 efficient way way to minimize loss (ensuring an exactly even score).
-Suppose that there are $n$ games $G$ being played. We generate a
-bipartite matching that pairs games $\langle G_w, G_b \rangle$ (where
-we play as white against Bob and black against Walice, respectively).
-This is $n/2$ pairs, and we record which pair each game belongs to, in
-$n \log_2(n/2)$ bits. Now, to make progress, we look at a game $G_i$
-where it is our move, and consult our state for this pair. If the
-pair has the has-move bit (actually just store it as the move itself),
+The simplest way to do this is to explicily generate a perfect
+bipartite matching over the $n$ games $G$ being played. This consists
+of $n/2$ pairs $\langle G_w, G_b \rangle$ (where we play as white
+against Bob and black against Walice, respectively). Since each game
+starts in the starting position, this is very easy; we can just assign
+the matches consecutively. Along with each pair we also record which
+of the following states we are in:
+\begin{enumerate}
+\item We are waiting for a move from Walice (our white opponent) \label{state:mim1}
+\item We have seen a move from Walice, which is \underline{\hspace{6em}}. \label{state:mim2}
+\item We are waiting for a move from Bob (our black opponent) \label{state:mim3}
+\item We have seen a move from Bob, which is \underline{\hspace{6em}}. \label{state:mim4}
+\end{enumerate}
+If in State~\ref{state:mim1}, we just watch $G_b$ until Walice makes a
+move, then record it and proceed to State~\ref{state:mim2}. We consume
+the move and move to State~\ref{state:mim3} by playing that move in
+$G_w$ against Bob (where it must be our turn). We can immediately seek
+out that game or wait until we naturally come upon it. However, we
+should only approach $G_w$ when the pair of games is in
+State~\ref{state:mim3}, etc., otherwise we will not have a move to play.
 
-ok, so probably the most efficient thing is to store instead the
-move played in each row. Moves are like 64*64*4 (promotion) = 6+6+2 bits.
-The only way to really motivate the board approach would be if it
-allows us to avoid the log(n/2) factor from storing the matching?
+There are $n/2$ pairs, with two bits for the state, no more\footnote{
+  There are only 1792 pairs of squares between which pieces can ever
+  move (Section~\ref{sec:legalmove}), so $11+2$ bits suffices, with
+  some added complexity.} than $\log_2(64 * 64 * 4) = 14$ bits for
+each move (source square, destination square, and 2 bits to
+distinguish promotion to queen, rook, bishop, or knight). However, we
+also need to store the matching of $G_w$ to $G_b$; this can be done
+with a pair of indices (or e.g. memory addresses) but unfortunately,
+this requires $\log_2(n)$ bits to represent. So overall this approach
+requires $O(n \log(n))$ space to play a simultaneous exhibition of $n$
+games.
+
+It appears to be possible to reduce the space usage per game to a
+constant. In order to perform a man-in-the-middle attack, we need
+a perfect matching between the white games and black games. It is
+not essential that the matching be stable over time; for example
+if we are forwarding moves between Walice and Bob, and between
+Waluigi and Bario, and these games happen to transpose to the same
+position, then it works just fine to switch to forwarding between
+Walice and Bario; Waluigi and Bob. So, rather than store the
+matching explicitly, we can reconstruct it from the stored state.
+We two vectors of $n/2$ rows, the ``source'' and ``destination'' rows,  (XXX I think we do not need destination)
+each containing a game {\em position} (not move).
+This position represents XXXX (like the last time we saw it?).
+The positions are initialized to the starting position. Like before
+we can be in one of four states:
+\begin{itemize}
+\item Waiting for all ``source'' players (who play as white) to complete their moves. \label{state:cmim1}
+\item Forwarding our moves from ``source'' to our black opponents. \label{state:cmim2}
+\item `` \ldots '' (who play as black) `` \ldots '' \label{state:cmim3}
+\item `` \ldots '' white opponents. \label{state:cmim3}
+\end{itemize}
+
+\newcommand\stepsto[1]{\stackrel{#1}{\rightarrow}}
+
+% (XXX So actually, simpler here is just: Loop until all white players have
+% made their moves. Then copy all their boards to our board vector.)
+While in State~\label{state:cmim1} (which we also start in), we
+compare our stored source board states to the boards where we play as
+black. Once the white opponent has made a move, resulting in $B'$,
+there will be some position $B$ in our ``source'' set that is a valid
+predecessor for $B'$ (that is, there exists some legal move $M$ that
+advances $B$ to $B'$). For example, at the beginning, $B$ may be the
+starting position (in fact it must, since all boards are initialized
+this way) and $B'$ may be the position resulting from $1. e4$. With
+the matching constructed (see below) we copy each $B'$ into our set of
+boards, forgetting $B$.\footnote{In fact, we do not even need to
+  construct the matching, although it may help us with bookkeeping
+  (e.g., which games have already been copied?), depending on how the
+  exhibition itself is represented.} Then we transition to
+State~\label{state:cmim2}. In this state, we figure out what moves to
+play in the pending boards. By invariant, each one $B$ will be
+compatible with at least one of our source boards $B'$. Compatible
+means that there exists a move $m$ where playing $m$ in board $B$
+results in $B'$. If $m$ exists, it is unique (and easy to compute), so
+this is the move that we will play. The only tricky thing is figuring
+out which boards go with which; this is the problem of reconstructing
+the perfect matching.
+
+Although if $B \stepsto{m} B'$ it is easy to deduce $m$, it is not
+possible to compute $B$ from $B'$, or even from $B'$ and $m$. This is
+because we may have both $B_a \stepsto{m_a} B'$ and $B_b \stepsto{m_b}
+B'$ with $B_a \neq B_b$. For example with $B'$
+
+\chessboard[tinyboard, setfen=k7/8/8/8/8/8/8/K7 w - -, showmover]
+
+we could have $B_a$ and $B_b$ be
+
+\chessboard[tinyboard, setfen=Nk6/8/8/8/8/8/8/K7 b - -, showmover]
+\chessboard[tinyboard, setfen=Rk6/8/8/8/8/8/8/K7 b - -, showmover]
+
+Both of which can precede $B'$ (the move is even the same: Kxa1).
+So it is not enough to greedily assign edges in our perfect match;
+if we choose the edge $B_b$ to go with $B'$, we might later find
+$B'_2$:
+
+\chessboard[tinyboard, setfen=R7/1k6/8/8/8/8/8/K7 w - -, showmover]
+
+\ldots and have no possible matching board, since it cannot follow
+$B_a$.
+
+Fortunately, we know the
+ - there exists a perfect matching (players playing legally)
+ - we can tell if we found one
 
@@
 I think a version I liked was 64 bits (this gives the positions of all
-pieces), followed by 32 records giving the piece values (there are 6:
-pawn, knight, bishop, rook, queen, king) and colors. To pick between
-the 12 possibilities we need 4 bits, and there is some slack here of
-course. anyway, that's 64 + 32*4 = 192 bits, plus 15 bits for other
-state. Note we could use the c\_rook trick to save 4 bits for castling
-rights (14 possibilities), not to mention a en\_passantable pawn (15)
-and a king whose turn it is (even 16). Clear slack here, but that
-would leave only the 50-move rule at 7 bits, for a total of 199.
-Wikipedia says the huffman approach is ``maximum of 204 bits, and
-often much less.'' but the article also contains many bugs, like
-the misconception that there can only be four rooks.
+pieces),\footnote{(We can also do a little better. Note that only up
+  to 32 of the 64 bits can be set, so you could represent
+  $\binom{60}{32}$ in $\sim 61$ bits and use e.g. a third king to
+  indicate that there are no more actually occupied squares if fewer
+  than 32 pieces are present. However, this gets vastly more
+  complicated for only 3 bits of savings.)} followed by 32 records
+giving the piece values (there are 6: pawn, knight, bishop, rook,
+queen, king) and colors. To pick between the 12 possibilities we need
+4 bits, and there is some slack here of course. anyway, that's 64 +
+32*4 = 192 bits, plus 15 bits for other state. Note we could use the
+c\_rook trick to save 4 bits for castling rights (14 possibilities),
+not to mention a en\_passantable pawn (15) and a king whose turn it is
+(even 16). Clear slack here, but that would leave only the 50-move
+rule at 7 bits, for a total of 199. Wikipedia says the huffman
+approach is ``maximum of 204 bits, and often much less.'' but the
+article also contains many bugs, like the misconception that there can
+only be four rooks.
 

### 2019-03-08 trunk/chess/papers/blind.tex
More actual progress here; need to rewrite the simul part now that I figured it out

--- 2019-03-08 chess/papers/blind.tex
@@
 
+% Computers can still impress us and each other with computational chess
+% feats other than playing. For example, computers are largely concerned
+% with filling up their memories with efficiently encoded data.
+
 Playing multiple games simutaneously is an easy extension of playing a
@@
 play in the pending boards. By invariant, each one $B$ will be
-compatible with at least one of our source boards $B'$. Compatible
+compatible with at least one of our source boards $B'$; we can iterate
+over source boards until we find one in constant space (and mark it
+as ``taken'' with . Compatible
 means that there exists a move $m$ where playing $m$ in board $B$
@@
 
-Fortunately, we know the
- - there exists a perfect matching (players playing legally)
- - we can tell if we found one
+Fortunately, we know that there exists a perfect matching (assuming
+all players are playing legal moves) and we can tell if we found
+one (i.e., we didn't get stuck). So, one strategy that works is
+to choose randomly when there is some ambiguity, and start over
+from the beginning if we ever get stuck. In practice this will
+be pretty efficient, since convergent board states are unusual.
+We only need a single pseudorandom pool for the entire process,
+so it can be O(n) bits; this seems easily enough to generate all
+possible permutations of $n/2$ items. Even $n^{2n}$ grows much
+faster than $n!$. If we don't like the random approach, I believe
+it is also possible to compute \verb+next_permutation+ in
+constant space; so we can just explicitly try all orderings for $B'$
+(this takes exponential time).
 
-so you have a set of black and white games, same number of each.
-goal is to get an even score, while using O(n) space.
-So we have an invariant that each of the games-as-white is paired
-with at least one compatible game-as-black (and vice versa). Of
-course at the beginning all pairs are compatible, and then white
-makes a move; the games are still compatible because any of these
-moves can be played on the black side. We just need to make sure
-that we don't play e.g. e4 in all the games, because we need enough
-black opponents to be playing against d4s and Nf3s and so on.
-So, after all the white players have made their moves, we have
-those boards, and need to make moves for black. For each, just pick
-a compatible board (i.e., something that represents a legal move)
-and play it. When we do so, we update the board to a state like
-``I made this move as white.'' (So at the very beginning, all boards
-are in the starting position with like ``I made this move as black,''
-of course not really, more like ``white opponent to move'').
-This is easy on the first move because all choices (that are in
-the state ``copied this move from white'').
+Once we have paired up each $B$ and $B'$, we simply compute the
+move (which we now know exists) and play it in that game. We
+then wait for the black opponents to play their moves, copy the
+resulting board states into our vector and repeat the above
+process (but flipping ``black'' and ``white'').
 
-Now we wait for black players to make their moves. (Note waiting
-can be done in constant space. You just compare each board that
-has ``black opponent to move'' with the actual board state; if
-the move is compatible then you update it. But might you confuse
-the identities of two games? It would need to be the case that
-there exists board states $B_1$, $B_2$, $B$ where $B$ can
-legally follow $B_1$ and $B_2$ but $B_1$ and $B_2$ are different.
-Of course this is possible! $B$ could be like
+Although this is more involved than the previous approach, and may
+take exponential time, it allows us to play against $n$ simultaneous
+opponents using $O(n)$ space!\footnote{ 
+%
+If the games in the exhibition were presented to us in some way that
+allowed us to visit them in the same order multiple times, we could
+perhaps get away with {\em only} the random pool? XXX }
+%
+The actual space used per game is primarily the representation of
+a chess position, plus a few bits for bookkeeping. So, representing
+boards compactly gives us a way to increase the number of simultaneous
+games we can play for a given storage size.
 
-\chessboard[tinyboard, setfen=k7/8/8/8/8/8/8/K7 w - -, showmover]
+Mainly, we need to store the pieces and their locations. There are a
+few other bits, like whose turn it is (1 bit), whether each player can
+still castle king- and queen-side (4 bits), and whether and where an
+en passant capture is possible (4 bits).\footnote{
+%
+Technically, we need to store a lot of additional information with the
+board in order to completely implement the rules of
+chess.\cite{fiderules} The trickiest of these involve the rules for
+draw by repetition, which make reference to the history of the game
+(See Footnote~1 in {\it Survival in chessland}\cite{survival} and seem
+to require storing all previous board states. Fortunately, if we are
+being this picky, then we also know that the length of a chess game is
+bounded by a constant: Rule 9.6.2 ends the game in a draw if both
+players have made 75 moves without a pawn move or
+capture,\footnotemark\ so it suffices to store the $75*2$ most recent
+(half-)moves. This sucks so most people don't do it (for example, FEN
+notation only gives the number of such moves, and so cannot implement
+the draw by repetition). On the other hand, if we insist, then this
+may give us a simpler route to a constant-space exhibition, since the
+$B \stepsto{m} B'$ relation is probably reversible with such
+information.}\footnotetext{These are two types of moves that make it impossible
+  to formally repeat a position that preceded them. Castling also has
+  this property, but doesn't count because it is a ``secret move.''}
 
-and then we have $B_1$ and $B_2$ as
+With 64 squares, six pieces for each of the two sides, plus the empty
+square, a straightforward representation of the board uses $64*4=256$
+bits. The Thursd'z Institute considered more compact
+representations\cite{representations}; one nice choice works as follows:
 
-\chessboard[tinyboard, setfen=1k6/8/8/8/8/8/8/K7 b - -, showmover]
-\chessboard[tinyboard, setfen=Rk6/8/8/8/8/8/8/K7 b - -, showmover]
+Start with a single 64-bit mask which indicates, for each square on
+the board, whether it contains any piece. Note that there can only be
+up to 32 pieces on the board. To tell what these pieces are, we then
+follow with 32 four-bit records; these indicate the piece's color and
+type.\footnote{(Since only 32 of the 64 bits can be set, you could do
+  slightly better by representing $\binom{60}{32}$ in $\sim 61$ bits.
+  When fewer than 32 squares are occupied, we can use a record
+  containing e.g. a third king (otherwise impossible) to indicate that
+  we should ignore the remaining bits. However, this gets vastly more
+  complicated for only 3 bits of savings.)} 
+%
+With the 9 bits of extra state above, this totals $64+32*4+9=201$
+bits. There is some slack in the representation, because there are
+only 12 actual possibilities for a piece but we can indicate 16 with 4
+bits. It is great to abuse this slack to save bits; for example, we
+can store a new type of rook that is still able to castle (it can only
+be in the corners and thus also indicates its own color), eliminating
+the 4 castling bits. We can similarly introduce an en-passantable
+pawn, saving the 4 bits of en passant state; this piece can only be in
+ranks 4 or 5, so it also indicates its color. We can also have a
+``king whose turn it is'' for each side, saving the side-to-move bit.
+This totals a nice round $64+32*4=192$ bits.\footnote{ Since this is
+  SIGBOVIK, I am freed from the burden of comparing related work. I
+  did however read the rather bad Wikipedia article on the
+  topic\cite{wikipediaboard} which describes a Huffman-based encoding
+  that uses a ``maximum of 204 bits, and often much less.'' This also
+  includes a 7-bit 50-move counter (but you really need to implement a
+  75-move counter; 50 moves only allow a player to {\em claim} a draw)
+  so should be compared as $197$ bits. But the article also contains
+  many bugs, like the misconception that there can only be four total
+  rooks (pawns are allowed to promote to rook). So the approach
+  described here is both {\em more efficient} and {\em more correct}.}
+%
+This would allow approximately an 11 billion-game simultaneous
+exhibition in RAM on my desktop computer.
 
-And so if we misassign $B$ to $B_1$, and then later see $B'$:
-
-\chessboard[tinyboard, setfen=R7/1k6/8/8/8/8/8/K7 w - -, showmover]
-
-We are stuck because $B'$ cannot follow $B_2$.
-
-So if we're actually going to do it with this method, we would
-need to be able to compute the bipartite matching looking at all
-the constraints. There certainly exists one, and any one will do. 
-(Actually what we want is called a ``perfect matching.'')
-So the question is, can we compute a perfect matching in constant
-space? I mean, it seems like not. How could we even {\em store}
-the matching? Well, we don't actually need to store the matching
-because it's not *literally* what we need; we just want to have
-assigned the games according to a perfect match. So one approach
-is like, pick a random assignment. If we get stuck, just start
-over (so we would have to keep two board states, in order to
-be able to ``undo'', but that is still constant) and commit
-the whole thing if we succeed. We're actually quite likely to
-succeed since ambiguities are pretty rare. So this may work
-fine in practice. The main thing we achieve here is that we
-don't need to store the matching itself between rounds; the
-matching is ``expensive'' because it contains an index, which
-is logarithmic size in the number of games. So, with this
-method, we get $O(n)$ space, which is actually around $200 n$.
-
-For each pair, we store the state of the board $B$
-as last we saw it in $G_w$ or $G_b$.
-
-both Walice and Bob have made their moves (that is, it is white's turn
-in $G_w$ and
-
-$G_i$, we look up which pair $p$ it is, and look at $B_p$. This board
-
-Note that
-
-Computers can still impress us and each other with computational chess
-feats other than playing. For example, computers are largely concerned
-with filling up their memories with efficiently encoded data.
-
-compactly storing a position. constant state
-
-I think a version I liked was 64 bits (this gives the positions of all
-pieces),\footnote{(We can also do a little better. Note that only up
-  to 32 of the 64 bits can be set, so you could represent
-  $\binom{60}{32}$ in $\sim 61$ bits and use e.g. a third king to
-  indicate that there are no more actually occupied squares if fewer
-  than 32 pieces are present. However, this gets vastly more
-  complicated for only 3 bits of savings.)} followed by 32 records
-giving the piece values (there are 6: pawn, knight, bishop, rook,
-queen, king) and colors. To pick between the 12 possibilities we need
-4 bits, and there is some slack here of course. anyway, that's 64 +
-32*4 = 192 bits, plus 15 bits for other state. Note we could use the
-c\_rook trick to save 4 bits for castling rights (14 possibilities),
-not to mention a en\_passantable pawn (15) and a king whose turn it is
-(even 16). Clear slack here, but that would leave only the 50-move
-rule at 7 bits, for a total of 199. Wikipedia says the huffman
-approach is ``maximum of 204 bits, and often much less.'' but the
-article also contains many bugs, like the misconception that there can
-only be four rooks.
-
+\medskip
 So now comes the main idea of the paper, which is also spoilered in the
@@
 lossy, of course, but it often contains enough information to deduce
-the state of the board, or ...
+the state of the board. For example, can you tell what position this is?
+
+\chessboard[setfen=8/8/8/8/8/8/8/8,
+  showmover=false, 
+  smallboard,  
+  color=gray,
+  strokeopacity=0,
+  markstyle={[fill]circle},
+  markfields={a1,b1,c1,d1,e1,f1,g1,h1,a2,b2,c2,d2,e4,f2,g2,h2,a7,b7,c7,d7,e7,f7,g7,h7,a8,b8,c8,d8,e8,f8,g8,h8
+}]
+
+Correct! It is
+
+\begin{tabular}{p{1.85in}p{1.4in}}
+\chessboard[setfen=nrbqkbrn/pppppppp/8/8/4P3/8/PPPP1PPP/NRBBNRQK b - - 34 19, showmover=false, smallboard] & {
+\vspace{-1.65in} \small after 1. Nf3 Nf6 2. e4 Ng4 3. Bc4 Ne5 4. O-O Ng6
+5. Kh1 Rg8 6. Nc3 Nh8 7. Qe2 Nc6 8. Rb1 Rb8
+9. Nb5 Nb4 10. Rd1 Nd5 11. Qf1 Nb6 12. Qg1 Na8
+13. Nbd4 Nb6 14. Rf1 Na8 15. Be2 Ng6 16. Bd1 Nh8
+17. Nb3 Ng6 18. Ne1 Nh8 19. Na1}
+\end{tabular} \vspace{1em}
+
+So this is kind of like blindfold chess ... XXX
 

### 2019-03-08 trunk/chess/papers/blind.tex
Good progress on the unblinder part

@@
 each other in the face repeatedly during the game to impair brain
-function (see Chess Boxing~\cite{chessboxing}). Another common
+function (see Chess Boxing~\cite{wikipediachessboxing}). Another common
 distraction is to play a multitude of games against many opponents at
@@
   complex.}
+% Note: blindfold simuls are banned in russia as a health hazard!?
 
@@
   move (Section~\ref{sec:legalmove}), so $11+2$ bits suffices, with
-  some added complexity.} than $\log_2(64 * 64 * 4) = 14$ bits for
+  some added complexity.} than $\log_2(64 \times 64 \times 4) = 14$ bits for
 each move (source square, destination square, and 2 bits to
@@
 perhaps get away with {\em only} the random pool? XXX }
-%
+
+\subsection{Board representations} \label{sec:representation}
+
 The actual space used per game is primarily the representation of
@@
 players have made 75 moves without a pawn move or
-capture,\footnotemark\ so it suffices to store the $75*2$ most recent
+capture,\footnotemark\ so it suffices to store the $75\times 2$ most recent
 (half-)moves. This sucks so most people don't do it (for example, FEN
@@
 With 64 squares, six pieces for each of the two sides, plus the empty
-square, a straightforward representation of the board uses $64*4=256$
+square, a straightforward representation of the board uses $64 \times 4=256$
 bits. The Thursd'z Institute considered more compact
@@
 %
-With the 9 bits of extra state above, this totals $64+32*4+9=201$
+With the 9 bits of extra state above, this totals $64+32 \ times 4+9=201$
 bits. There is some slack in the representation, because there are
@@
 ``king whose turn it is'' for each side, saving the side-to-move bit.
-This totals a nice round $64+32*4=192$ bits.\footnote{ Since this is
+This totals a nice round $64+32 \times 4=192$ bits.\footnote{ Since this is
   SIGBOVIK, I am freed from the burden of comparing related work. I
@@
 
-So this is kind of like blindfold chess ... XXX
+\section{Color- and piece-blind chess}
 
+So this is kind of like blindfold chess, but for computers! Instead of
+being blind to the board, and only relying on our memory (trivial for
+computers), we'll only be able to see where pieces are positioned on
+the board, but not what type or color they are. Of course, we also
+have to prohibit the computer from simply remembering the board, so
+the algorithm must be stateless. Specifically, we want a function
+\[
+  {\tt makemove} : {\tt uint64} \rightarrow {\tt move list}
+\]
+that makes a move from a single board position, represented just as 64
+bits. This is a single move; the move list represents our preference
+for the move to make in descending order, and we commit to the first
+move that is actually legal. It does not work well to insist that the
+function return a single move, as it will often be the case that the
+board is misinterpreted and a move is in fact illegal; forcing
+forfeit\footnote{FIDE rules state that the {\em second} attempt at an
+  illegal move results in forfeit (7.5.5).} would mean that almost all
+games end in forfeit, which is boring. On the other hand, allowing the
+function to try again upon learning a move is illegal would allow it
+to interactively ``interrogate'' the board state
+somewhat.\footnote{Similar to Kriegspiel~\cite{wikipediakriegspiel}, although
+  in that game at least one's own pieces are known!} This seems
+counter to the spirit of color- and piece-blind chess, so we instead
+require the function to rank all moves ahead of time.
+
+\section{Unblinding the board}
+
+I went about this by building a function that ``unblinds'' a board; it
+has type
+\[
+{\tt unblind} : {\tt uint64} \rightarrow {\tt position}
+\]
+
+This function is natural for machine learning. It is easy to generate
+copious training data from actual games by simply blinding positions
+into their corresponding 64-bit numbers; I just randomly sampled 100
+million positions from Feburary~2017 on lichess.org.
+
+I repurposed the custom neural network code from my seminal paper {\it
+  Red i removal with artificial retinal networks}~\cite{redi} after
+discovering that artificial retinal networks are actually {\em
+  isomorphic} to neural networks. The main advantage of this code is
+that it allows for sparse networks, but the real reason to use it is
+that I would rather spend dozens of hours debugging my own code, pay a
+larger electric bill, and get worse results in the end, than to spend
+a short while trying to get someone else's probably-very-good neural
+network training package to compile.
+
+The structure of the network is as follows. The input layer is 64
+nodes, one for each of the 64 bits, with each node set to either {\tt
+  1.0f} or {\tt 0.0f}. Three hidden layers of size 1024, 12288, and 567
+do the neural magic. The output layer is 837 nodes; the bulk of which
+is a ``one-hot'' representation of the predictions for the 64 squares,
+each with 13 possible contents (black or white, six pieces, or empty).
+This is $64 \times 13 = 832$ nodes. Then four nodes to predict the
+four castling bits, and one to predict the current side to move. This
+model does not predict the possibility for {\em en passant} capture,
+nor move counters or position repetition. This will not be its main
+source of disadvantage!
+
+I trained the network in two phases, first starting with a densely
+connected one (model size 160 MB), and then after I get fed up with
+how slow training was, a ``vacuumed'' version of the network where I
+removed edges with low absolute weights (model size 5 MB) to continue
+training. Removing edges based on an absolute weight threshold is
+very unprincipled (since a downstream edge can magnify a contribution
+arbitrarily) but I did it anyway.
+
+Everything was trained on a single GPU, though a fairly decent one in
+2018, the EVGA GeForce GTX 1080 FTW (really), using OpenCL. Biases
+were initialized to $0$ and weights with a Gaussian of mean $0$ and
+standard deviation $0.025$. In the first phase, there were 64 examples
+per round, and after vacuuming, 2048. The round learning rate
+$\alpha_r$ started at $0.1$ and descended linearly to $0.002$ over
+500,000 rounds; and the learning rate when updating weights (for each
+example) $\alpha$ is $\alpha_r / \verb+examples_per_round+$. In no way
+am I recommending these parameters. Fiddling with the parameters to
+make it do its black magic or alternately carom off to a sea of NaNs
+or zeroes is for sure the worst part of neural networks. In fact, I
+initially started with the classical sigmoid transfer function, but
+``upgraded'' to the ``leaky rectified linear'' function
 %
+$$((potential < 0.0f) ? potential * 0.01f : potential)$$
+%
+after getting fed up with sigmoid weights caroming off (see
+``vanishing gradient problem'' and/or ``exploding gradient problem'').
+The final model was trained over 339,885 rounds on 223 million
+examples. It did not appear to show signs of improving for
+several days before I terminated it.
+
+\subsection{Statistical evaluation}
+
+The unblinding component can be evaluated on its own, by running
+it on examples sampled independently of the training set. Over
+50,000 examples, these were the results:
+
+9,584 predicted boards were exactly correct (19.17\%). There were a
+total of 161,166 piece mistakes, which is an average of 3.22 per
+board. This is wherever we predict a square's contents incorrectly.
+There were only 1630 castling status mistakes, an average of 0.03 per
+board (there can be up to four mistakes per board). This is probably
+because when the king and rook are in their starting squares, castling
+is almost always still allowed. In 19,014 boards, we mispredicted
+whose move it is (38\%). This appears to be the most difficult thing
+to predict, which is not surprising.\footnote{ Prior to ``vacuuming'',
+  the 160 MB network actually performed slightly better than the final
+  5 MB network, with 21.20\% of boards exactly correct, and an average
+  of 3.12 mistakes per board. This suggests that the model may only be
+  doing a limited amount of generalization, instead mostly memorizing
+  board positions. Representing the 223 million examples seen exactly
+  (using our best board representation described in
+  Section~\ref{sec:representation}) would take 42.8 GB, so at 5 MB at
+  least the data is represented compactly, if also rather lossily.}
+
+\subsection{Subjective evaluation}
+
+The unblinder {\em must} make mistakes since the 64-bit representation
+is ambiguous. Subjectively, the unblinder makes reasonable mistakes.
+It is excellent at common openings, usually getting these exactly
+correct. On the other hand, it is fairly bad at sparse endgames, where
+it is difficult to tell a pawn from a rook from a king. It is terrible
+at unlikely positions that can be confused for likely ones. If you are
+playing against it and know how it works, it is easy to trick it by
+doing something like capturing one of its starting-position pawns with
+your queen; nobody does this in real games (because the queen can be
+immediately recaptured), so the square is predicted as a pawn and the
+queen ``disappears'' to the unblinder (Figure~\ref{fig:trick}). Having
+an invisible queen in your camp is of course very dangerous. Resolving
+ambiguities in favor of more likely positions is the right thing for
+the model to do, so this is just an inherent flaw with the
+decomposition of the problem. There are some ways we can account for
+this (Section~\ref{sec:using}). % XXX make sure I did this work
+
+\begin{figure}[ht]
+XXX actual example
+\caption{The white queen ``disappears'' after QxWhatever, because
+  unblinding predicts it to be one of black's own pawns, which is far
+  more likely in that square.} \label{fig:trick}
+\end{figure}
+
+A few things are distinctly disappointing about its performance. Even
+outside of ``likely'' positions, it usually predicts that pieces on
+black's side of the board are black, and vice versa
+(Figure~\ref{fig:allon}). This makes sense, but suggests serious
+limitation on using the prediction to play chess. Less forgivably, it
+sometimes predicts more than one king per side (or zero), which is
+always wrong. Actually, an early version had this problem in spades,
+usually frequently predicting two or three kings. Upon debugging, I
+had simply made the noob mistake of printing both King and Knight with
+the letter ``K.'' Ha! It often predicts the ``wrong'' number of
+bishops (etc.), or places them on the same color. This is technically
+possible through promotion, but usually a bad guess, since promotion
+is relatively rare, and moreso promotion to a piece other than a
+queen. An approach that might handle this better (but may have
+different downsides) would be instead to predict the ``fates'' of the
+32 initial pieces. The fate of a pawn can include promotion, but no
+other pieces can. This would {\em require} that the model only predict
+a single location for each king, among other things. However, this
+would require a much larger output layer (32 pieces can move to 64
+squares, plus promotion) and it is not always clear how to interpret
+its value into a position (for example, if two pieces are predicted
+strongly to be on the same square).
+
+\begin{figure}[ht]
+\includegraphics[width=0.9 \linewidth]{blind-allon}
+\caption{What the model predicts (in single-king mode;
+  Section~\ref{sec:using}) for {\tt 0xFFFFFFFFFFFFFFFF}, the board
+  with all bits set. This is an impossible position, but it gives some
+  idea of the model's biases for each square. Notably, most of the pieces
+  on each half of the board have a single color. This makes sense, but
+  also suggests substantial limitation. When single-king mode is off,
+  the bottom right king is predicted as a white rook.} \label{fig:allon}
+\end{figure}
+
 % % make -j eval-unblinder.exe && ./eval-unblinder.exe net.val
@@
 
-% all bits set to 1 (with single_kings)
-% (without single_kings, only difference is that the bottom right becomes a
-% rook; no white queen)
-\includegraphics[width=0.9 \linewidth]{blind-allon}
-
 \bibliography{chess}{}

### 2019-03-09 trunk/chess/papers/blind.tex
Wrote about using unblind to play

@@
 %
-$$((potential < 0.0f) ? potential * 0.01f : potential)$$
+$$(p < 0) \,?\, p \times 0.01 : p$$
 %
@@
 
-The unblinding component can be evaluated on its own, by running
-it on examples sampled independently of the training set. Over
-50,000 examples, these were the results:
+The unblinding component can be evaluated on its own, by running it on
+examples sampled independently of the training set. The model outputs
+a score for each possible contents of each square; we simply
+discretize to the highest-scoring one (same too for the predicted
+castling state and turn). Over 50,000 examples, these were the
+results:
 
@@
 different downsides) would be instead to predict the ``fates'' of the
-32 initial pieces. The fate of a pawn can include promotion, but no
-other pieces can. This would {\em require} that the model only predict
-a single location for each king, among other things. However, this
-would require a much larger output layer (32 pieces can move to 64
-squares, plus promotion) and it is not always clear how to interpret
-its value into a position (for example, if two pieces are predicted
-strongly to be on the same square).
+32 initial pieces~\cite{survival}. The fate of a pawn includes what if
+any piece it has promoted to, but no other pieces can. This would {\em
+  require} that the model only predict a single location for each
+king, among other things. However, this would require a much larger
+output layer (32 pieces can move to 64 squares, plus promotion) and it
+is not always clear how to interpret its value into a position (for
+example, if two pieces are predicted strongly to be on the same
+square).
 
@@
 \end{figure}
+
+\section{Playing blind}
+
+Once we have predicated a board state, we can play with it. The
+simplest way to do this is to use a strong chess engine to pick a move
+for the position. Here, I use Stockfish with a node budget of 1
+million nodes, which takes about 1 second of CPU time per move on my
+computer; henceforth {\sf stockfish1m}.
+
+There are some complications:
+\begin{itemize}
+\item Frequently, the unblinded board will not match the true
+  position, and Stockfish will choose a move that is illegal. So, as discussed
+  before, we actually return a prioritzed list of moves. For this first
+  experiment, we just return the move Stockfish recommends, followed by all
+  other moves ordered randomly.
+\item Stockfish is a very strong engine, and in my opinion the code is generally
+  good, but it is very sensitive to bad FEN (the notation used to give it a
+  position to evaluate) strings. Given a bad string, like one that says castling
+  is possible when the king isn't in its home square, often crashes the engine.
+  So we need to make sure to actually pass valid positions. I accomplish this
+  by making the following modifications to the predicted board:
+  \begin{itemize}
+  \item If a castling bit is set, but castling is not possible due to the position
+    of the king or rook, I clear the bit.
+  \item Set the side-to-move to be the correct actual value. This uses
+    the unblinded state, so is superficially cheating. But note that
+    if we get the side wrong, then Stockfish's move will always be
+    illegal: Moves are specified as a source and destination
+    square,\footnote{Plus promotion piece. Castling is represented as
+      a two-square move of the king.} and so the source square of
+    Stockfish's move would always be a piece of the wrong piece's
+    color. So this is equivalent to (but twice as efficient as)
+    running stockfish twice, one for each side, and prioritizing the
+    predicted side's move first.
+  \end{itemize}
+  This won't fix all positions, for example, if the white and black king are
+  adjacent to one another in mutual check. If an irreparable problem is detected,
+  then I just return a uniformly random move list.
+\end{itemize}
+
+It is easy to beat this chess engine, by tricking it as in
+Figure~\ref{fig:trick}, although this involves unnatural moves, so it
+may only apply if you know how it works.
+% XXX example game?
+Measuring how well it plays in an absolute sense is a subject of some
+interest, so I wrote a separate paper about that~\cite{eloworld}. This
+algorithm, called \verb+blind_yolo+, had an Elo World score of XXX.
+
+\subsection{We three kings}
+
+When evaluating the first version I found that it was producing a
+fairly high number of illegal positions, which was causing us to fall
+back on making random moves, which is mostly boring. The second version
+reduces the rate of illegal positions due to too many or too few kings.
+
+ - we get scores with each piece
+ etc.
+
+This version, called \verb+blind_kings+, performs consistently better
+than \verb+blind_yolo+ (XXX give actual record). It had an Elo World
+score of XXX.
+
+\cite{kingme}
+
+\section{Conclusion}
+
+- Other structures for the network
+- Stochastic approach
+- How do the solutions compare to other chess algorithms that
+  could be applied in the blind? Random works, also lexicographic,
+  any others? Most are inspecting the board state, obviously.
+- Could estimate the upper bound for accuracy by blinding every
+  game in the database, and doing a ``lookup''-based version of
+  the eval.
 

### 2019-03-09 chess/blind-player.cc
Spycheck, not yet implemented

@@ -31,7 +31,9 @@ static Unblinder *GetUnblinder() {
 
 struct BlindPlayer : public StatelessPlayer {
   BlindPlayer(const string &name,
+	      bool spycheck,
 	      bool single_king) : name(name),
+				  spycheck(spycheck),
 				  single_king(single_king),
 				  rc(PlayerUtil::GetSeed()) {
     fish.reset(new Stockfish(20, 1'000'000));
@@ -104,6 +106,21 @@ struct BlindPlayer : public StatelessPlayer {
     const uint64 blinded = Unblinder::Blind(orig_pos);
     Position predicted = unblinder->Unblind(single_king, blinded);
 
+    // HERE: Do spy check if enabled. Note that it would be
+    // wrong for us to do this using the correct side-to-move, because
+    // we don't actually have that information. (Below it can be seen
+    // as an optimization, equivalent to running stockfish for both
+    // possibilities, since the potentially legal move sets would be
+    // completely disjoint.)
+    if (spycheck) {
+      // ... Possible to do this a symmetric way, though? I guess we
+      // can just prioritize any legal move where the source and
+      // destination pieces are predicted to be the same color, but
+      // maybe put moves for the predicted side-to-move first. Also
+      // makes sense to break ties by capturing with a weaker
+      // (predicted) piece, probably.
+    }
+
     // Force the current player.
     predicted.SetBlackMove(orig_pos.BlackMove());
 
@@ -146,13 +163,16 @@ struct BlindPlayer : public StatelessPlayer {
   }
 
   string Desc() const override {
-    return StringPrintf("Predict a board state%s. If valid (or can be made "
+    return StringPrintf("Predict a board state%s.%s If valid (or can be made "
 			"valid trivially), use stockfish1m to make a move. "
 			"If move is invalid or other problem, random.",
+			(spycheck ? " First, spy-check pieces that we think are our "
+			 "own." : ""),
 			(single_king ? "" : " with exactly one king per side"));
   }
 
   const string name;
+  const bool spycheck = false;
   const bool single_king = false;
   ArcFour rc;
   std::unique_ptr<Stockfish> fish;
@@ -164,9 +184,13 @@ struct BlindPlayer : public StatelessPlayer {
 
 
 Player *BlindYolo() {
-  return new MakeStateless<BlindPlayer, string, bool>("blind_yolo", false);
[…]

### 2019-03-09 trunk/chess/papers/blind.tex
Spycheck, not yet implemented

@@
 
-When evaluating the first version I found that it was producing a
-fairly high number of illegal positions, which was causing us to fall
-back on making random moves, which is mostly boring. The second version
-reduces the rate of illegal positions due to too many or too few kings.
+When evaluating the first version I found that it was predicting a
+disappointingly high number of illegal positions in practice, which
+was causing us to fall back on making random moves, which is mostly
+boring. The second version reduces the rate of illegal positions due
+to too many or too few kings~\cite{kingme}.
 
- - we get scores with each piece
- etc.
+The model predicts a score for each type of piece in each square,
+and we do not have to necessarily interpret it by always taking
+the highest-scoring piece. This version first finds the two squares
+with the highest scores for the white king, and same for the black
+king. We take two in case the same square is predicted for both.
+Then this square gets one of the kings (whichever has higher score)
+and the other king goes in the highest-scoring unclaimed square. The
+rest of the squares just get the highest-scoring prediction as before,
+but we never predict kings.
 
 This version, called \verb+blind_kings+, performs consistently better
-than \verb+blind_yolo+ (XXX give actual record). It had an Elo World
-score of XXX.
+than \verb+blind_yolo+ (63 wins, 45 losses, 412 draws).
+% XXX could run a lot more games between these two.
+% as white: 31/24/207, as black 21/32/205
+It had an Elo World score of XXX.
 
-\cite{kingme}
+\subsection{Spy check}
+
+Say \verb+blind_kings+ is playing as black; it remains easy to fool it
+by moving black pieces into white's camp, since they are usually then
+predicted to be white pieces. We can defend against this somewhat.
+Since it is illegal to capture one's own piece, there is little risk
+in trying; if it is indeed our own piece then the move will be
+rejected, and if it is not our piece, then capturing is good for two
+reasons: One, we capture a piece, and two, we avoid having
+Stockfish make a move in this incorrectly predicted board. (Of course
+there are many reasons why eagerly capturing a piece can be a bad
+idea, but at this level of play, an edge in material is likely worth
+it.)
+
+There is one subtlety here, which is that we can't fairly use the
+information of the actual side-to-move, as above. So there is some
+additional chance that we end up making a dumb move because we
+both mispredicted the side-to-move and the identity of some pieces.
+
+(XXX evaluation here)
 

### 2019-03-09 trunk/chess/papers/blind.tex
Implemented spycheck (not tested)

@@
 
-There is one subtlety here, which is that we can't fairly use the
-information of the actual side-to-move, as above. So there is some
-additional chance that we end up making a dumb move because we
-both mispredicted the side-to-move and the identity of some pieces.
+There is one subtlety here. Above we argued that it was safe to use
+the actual side-to-move instead of the predicted one; but here it
+would not be equivalent to do so. Instead, we first prioritize all
+apparent spy-check moves where the predicted source piece matches the
+predicted side-to-move, then we try the opposite. (Ties are broken by
+preferring to capture with a lower-value predicted piece, and then
+randomly.) Due to this, there is some additional chance that we end up
+making an especially dumb move because we both mispredicted the
+side-to-move and the identity of some pieces.
 
 (XXX evaluation here)
+
+\subsection{Future work}
+
+The predicted board often expresses uncertainty about some squares,
+which could be thought of as probabilities. A principled improvement
+would be to try to find moves that are good in expectation, that is,
+integrated all possible boards in proportion to their predicted
+probability. A good approximation might be had by sampling a bunch
+of boards according to the predicted distribution, and then using
+Stockfish to score the top $k$ moves for each; we can then order
+moves by their expected score. Unfortunately, it is not easy to
+efficiently get Stockfish to generate scored moves for $k \neq 1$.
+Even with $k = 1$, this approach would be slow, taking about a
+second for each (distinct) sampled board. So I did not try it,
+at least not before submitting this paper.
 
@@
 - Other structures for the network
-- Stochastic approach
 - How do the solutions compare to other chess algorithms that

### 2019-03-09 trunk/chess/player-util.h

@@ -20,7 +20,9 @@ struct PlayerUtil {
 
   static bool ParseLongMove(const std::string &move_s,
 			    bool black, Position::Move *m);
-  
+
+  // Get the best item from the non-empty vector f, where
+  // f(a, b) returns true if a is better than b.
   template<class T, class F>
   static const T &GetBest(const std::vector<T> &v, F f) {
     CHECK(!v.empty());

### 2019-03-09 chess/papers/blind.tex
Some conclusions and things

@@
 board!
+
+% Computers can still impress us and each other with computational chess
+% feats other than playing. For example, computers are largely concerned
+% with filling up their memories with efficiently encoded data.
 
@@
 
-Fortunately, we know the
- - there exists a perfect matching (players playing legally)
- - we can tell if we found one
+Fortunately, we know that there exists a perfect matching (assuming
+all players are playing legal moves) and we can tell if we found
+one (i.e., we didn't get stuck). So, one strategy that works is
+to choose randomly when there is some ambiguity, and start over
+from the beginning if we ever get stuck. In practice this will
+be pretty efficient, since convergent board states are unusual.
+We only need a single pseudorandom pool for the entire process,
+so it can be O(n) bits; this seems easily enough to generate all
+possible permutations of $n/2$ items. Even $n^{2n}$ grows much
+faster than $n!$. If we don't like the random approach, I believe
+it is also possible to compute \verb+next_permutation+ in
+constant space; so we can just explicitly try all orderings for $B'$
+(this takes exponential time).
 
-so you have a set of black and white games, same number of each.
-goal is to get an even score, while using O(n) space.
-So we have an invariant that each of the games-as-white is paired
-with at least one compatible game-as-black (and vice versa). Of
-course at the beginning all pairs are compatible, and then white
-makes a move; the games are still compatible because any of these
-moves can be played on the black side. We just need to make sure
-that we don't play e.g. e4 in all the games, because we need enough
-black opponents to be playing against d4s and Nf3s and so on.
-So, after all the white players have made their moves, we have
-those boards, and need to make moves for black. For each, just pick
-a compatible board (i.e., something that represents a legal move)
-and play it. When we do so, we update the board to a state like
-``I made this move as white.'' (So at the very beginning, all boards
-are in the starting position with like ``I made this move as black,''
-of course not really, more like ``white opponent to move'').
-This is easy on the first move because all choices (that are in
-the state ``copied this move from white'').
+Once we have paired up each $B$ and $B'$, we simply compute the
+move (which we now know exists) and play it in that game. We
+then wait for the black opponents to play their moves, copy the
+resulting board states into our vector and repeat the above
+process (but flipping ``black'' and ``white'').
 
-Now we wait for black players to make their moves. (Note waiting
-can be done in constant space. You just compare each board that
-has ``black opponent to move'' with the actual board state; if
-the move is compatible then you update it. But might you confuse
-the identities of two games? It would need to be the case that
-there exists board states $B_1$, $B_2$, $B$ where $B$ can
-legally follow $B_1$ and $B_2$ but $B_1$ and $B_2$ are different.
-Of course this is possible! $B$ could be like
+Although this is more involved than the previous approach, and may
+take exponential time, it allows us to play against $n$ simultaneous
+opponents using $O(n)$ space!\footnote{ 
+%
+If the games in the exhibition were presented to us in some way that
+allowed us to visit them in the same order multiple times, we could
+perhaps get away with {\em only} the random pool? XXX }
 
-\chessboard[tinyboard, setfen=k7/8/8/8/8/8/8/K7 w - -, showmover]
+\subsection{Board representations} \label{sec:representation}
 
-and then we have $B_1$ and $B_2$ as
+The actual space used per game is primarily the representation of
+a chess position, plus a few bits for bookkeeping. So, representing
+boards compactly gives us a way to increase the number of simultaneous
+games we can play for a given storage size.
 
-\chessboard[tinyboard, setfen=1k6/8/8/8/8/8/8/K7 b - -, showmover]
-\chessboard[tinyboard, setfen=Rk6/8/8/8/8/8/8/K7 b - -, showmover]
+Mainly, we need to store the pieces and their locations. There are a
+few other bits, like whose turn it is (1 bit), whether each player can
+still castle king- and queen-side (4 bits), and whether and where an
+en passant capture is possible (4 bits).\footnote{
+%
+Technically, we need to store a lot of additional information with the
+board in order to completely implement the rules of
+chess.\cite{fiderules} The trickiest of these involve the rules for
+draw by repetition, which make reference to the history of the game
+(See Footnote~1 in {\it Survival in chessland}\cite{survival} and seem
+to require storing all previous board states. Fortunately, if we are
+being this picky, then we also know that the length of a chess game is
+bounded by a constant: Rule 9.6.2 ends the game in a draw if both
+players have made 75 moves without a pawn move or
+capture,\footnotemark\ so it suffices to store the $75\times 2$ most recent
+(half-)moves. This sucks so most people don't do it (for example, FEN
+notation only gives the number of such moves, and so cannot implement
+the draw by repetition). On the other hand, if we insist, then this
+may give us a simpler route to a constant-space exhibition, since the
+$B \stepsto{m} B'$ relation is probably reversible with such
+information.}\footnotetext{These are two types of moves that make it impossible
+  to formally repeat a position that preceded them. Castling also has
+  this property, but doesn't count because it is a ``secret move.''}
 
-And so if we misassign $B$ to $B_1$, and then later see $B'$:
+With 64 squares, six pieces for each of the two sides, plus the empty
+square, a straightforward representation of the board uses $64 \times 4=256$
+bits. The Thursd'z Institute considered more compact
+representations\cite{representations}; one nice choice works as follows:
 
-\chessboard[tinyboard, setfen=R7/1k6/8/8/8/8/8/K7 w - -, showmover]
+Start with a single 64-bit mask which indicates, for each square on
+the board, whether it contains any piece. Note that there can only be
+up to 32 pieces on the board. To tell what these pieces are, we then
+follow with 32 four-bit records; these indicate the piece's color and
+type.\footnote{Since only 32 of the 64 bits can be set, you could do
+  slightly better by representing $\binom{60}{32}$ in $\sim 61$ bits.
+  When fewer than 32 squares are occupied, we can use a record
+  containing e.g. a third king (otherwise impossible) to indicate that
+  we should ignore the remaining bits. However, this gets vastly more
+  complicated for only 3 bits of savings.} 
+%
+With the 9 bits of extra state above, this totals $64+32 \times 4+9=201$
+bits. There is some slack in the representation, because there are
+only 12 actual possibilities for a piece but we can indicate 16 with 4
+bits. It is great to abuse this slack to save bits; for example, we
+can store a new type of rook that is still able to castle (it can only
+be in the corners and thus also indicates its own color), eliminating
+the 4 castling bits. We can similarly introduce an en-passantable
+pawn, saving the 4 bits of en passant state; this piece can only be in
+ranks 4 or 5, so it also indicates its color. We can also have a
+``king whose turn it is'' for each side, saving the side-to-move bit.
+This totals a nice round $64+32 \times 4=192$ bits.\footnote{ Since this is
+  SIGBOVIK, I am freed from the burden of comparing related work. I
+  did however read the rather bad Wikipedia article on the
+  topic\cite{wikipediaboard} which describes a Huffman-based encoding
+  that uses a ``maximum of 204 bits, and often much less.'' This also
+  includes a 7-bit 50-move counter (but you really need to implement a
+  75-move counter; 50 moves only allow a player to {\em claim} a draw)
+  so should be compared as $197$ bits. But the article also contains
+  many bugs, like the misconception that there can only be four total
+  rooks (pawns are allowed to promote to rook). So the approach
+  described here is both {\em more efficient} and {\em more correct}.}
+%
+This would allow approximately an 11 billion-game simultaneous
+exhibition in RAM on my desktop computer.
 
-We are stuck because $B'$ cannot follow $B_2$.
-
-So if we're actually going to do it with this method, we would
-need to be able to compute the bipartite matching looking at all
-the constraints. There certainly exists one, and any one will do. 
-(Actually what we want is called a ``perfect matching.'')
-So the question is, can we compute a perfect matching in constant
-space? I mean, it seems like not. How could we even {\em store}
-the matching? Well, we don't actually need to store the matching
-because it's not *literally* what we need; we just want to have
-assigned the games according to a perfect match. So one approach
-is like, pick a random assignment. If we get stuck, just start
-over (so we would have to keep two board states, in order to
-be able to ``undo'', but that is still constant) and commit
-the whole thing if we succeed. We're actually quite likely to
-succeed since ambiguities are pretty rare. So this may work
-fine in practice. The main thing we achieve here is that we
-don't need to store the matching itself between rounds; the
-matching is ``expensive'' because it contains an index, which
-is logarithmic size in the number of games. So, with this
-method, we get $O(n)$ space, which is actually around $200 n$.
-
-For each pair, we store the state of the board $B$
-as last we saw it in $G_w$ or $G_b$.
-
-both Walice and Bob have made their moves (that is, it is white's turn
-in $G_w$ and
-
-$G_i$, we look up which pair $p$ it is, and look at $B_p$. This board
-
-Note that
-
-Computers can still impress us and each other with computational chess
-feats other than playing. For example, computers are largely concerned
-with filling up their memories with efficiently encoded data.
-
-compactly storing a position. constant state
-
-I think a version I liked was 64 bits (this gives the positions of all
-pieces),\footnote{(We can also do a little better. Note that only up
-  to 32 of the 64 bits can be set, so you could represent
-  $\binom{60}{32}$ in $\sim 61$ bits and use e.g. a third king to
-  indicate that there are no more actually occupied squares if fewer
-  than 32 pieces are present. However, this gets vastly more
-  complicated for only 3 bits of savings.)} followed by 32 records
-giving the piece values (there are 6: pawn, knight, bishop, rook,
-queen, king) and colors. To pick between the 12 possibilities we need
-4 bits, and there is some slack here of course. anyway, that's 64 +
-32*4 = 192 bits, plus 15 bits for other state. Note we could use the
-c\_rook trick to save 4 bits for castling rights (14 possibilities),
-not to mention a en\_passantable pawn (15) and a king whose turn it is
-(even 16). Clear slack here, but that would leave only the 50-move
-rule at 7 bits, for a total of 199. Wikipedia says the huffman
-approach is ``maximum of 204 bits, and often much less.'' but the
-article also contains many bugs, like the misconception that there can
-only be four rooks.
-
+\medskip
 So now comes the main idea of the paper, which is also spoilered in the
@@
 lossy, of course, but it often contains enough information to deduce
-the state of the board, or ...
+the state of the board. For example, can you tell what position this is?
 
+\chessboard[setfen=8/8/8/8/8/8/8/8,
+  showmover=false, 
+  smallboard,  
+  color=gray,
+  strokeopacity=0,
+  markstyle={[fill]circle},
+  markfields={a1,b1,c1,d1,e1,f1,g1,h1,a2,b2,c2,d2,e4,f2,g2,h2,a7,b7,c7,d7,e7,f7,g7,h7,a8,b8,c8,d8,e8,f8,g8,h8
+}]
+
+Correct! It is
+
+\begin{tabular}{p{1.85in}p{1.4in}}
+\chessboard[setfen=nrbqkbrn/pppppppp/8/8/4P3/8/PPPP1PPP/NRBBNRQK b - - 34 19, showmover=false, smallboard] & {
+\vspace{-1.65in} \small after 1. Nf3 Nf6 2. e4 Ng4 3. Bc4 Ne5 4. O-O Ng6
+5. Kh1 Rg8 6. Nc3 Nh8 7. Qe2 Nc6 8. Rb1 Rb8
+9. Nb5 Nb4 10. Rd1 Nd5 11. Qf1 Nb6 12. Qg1 Na8
+13. Nbd4 Nb6 14. Rf1 Na8 15. Be2 Ng6 16. Bd1 Nh8
+17. Nb3 Ng6 18. Ne1 Nh8 19. Na1}
+\end{tabular} \vspace{1em}
+
+\section{Color- and piece-blind chess}
+
+So this is kind of like blindfold chess, but for computers! Instead of
+being blind to the board, and only relying on our memory (trivial for
+computers), we'll only be able to see where pieces are positioned on
+the board, but not what type or color they are. Of course, we also
+have to prohibit the computer from simply remembering the board, so
+the algorithm must be stateless. Specifically, we want a function
+\[
+  {\tt makemove} \;:\; {\tt uint64} \rightarrow {\tt move\ list}
+\]
+that makes a move from a single board position, represented just as 64
+bits. This is a single move; the move list represents our preference
+for the move to make in descending order, and we commit to the first
+move that is actually legal. It does not work well to insist that the
+function return a single move, as it will often be the case that the
+board is misinterpreted and a move is in fact illegal; forcing
+forfeit\footnote{FIDE rules state that the {\em second} attempt at an
+  illegal move results in forfeit (7.5.5).} would mean that almost all
+games end in forfeit, which is boring. On the other hand, allowing the
+function to try again upon learning a move is illegal would allow it
+to interactively ``interrogate'' the board state
+somewhat.\footnote{Similar to Kriegspiel~\cite{wikipediakriegspiel}, although
+  in that game at least one's own pieces are known!} This seems
+counter to the spirit of color- and piece-blind chess, so we instead
+require the function to rank all moves ahead of time.
+
+\section{Unblinding the board}
+
+I went about this by building a function that ``unblinds'' a board; it
+has type
+\[
+{\tt unblind} \;:\; {\tt uint64} \rightarrow {\tt position}
+\]
+
+This function is natural for machine learning. It is easy to generate
+copious training data from actual games by simply blinding positions
+into their corresponding 64-bit numbers; I just randomly sampled 100
+million positions from Feburary~2017 on lichess.org.
+
+I repurposed the custom neural network code from my seminal paper {\it
+  Red i removal with artificial retinal networks}~\cite{redi} after
+discovering that artificial retinal networks are actually {\em
+  isomorphic} to neural networks. The main advantage of this code is
+that it allows for sparse networks, but the real reason to use it is
+that I would rather spend dozens of hours debugging my own code, pay a
+larger electric bill, and get worse results in the end, than to spend
+a short while trying to get someone else's probably-very-good neural
+network training package to compile.
+
+The structure of the network is as follows. The input layer is 64
+nodes, one for each of the 64 bits, with each node set to either {\tt
+  1.0f} or {\tt 0.0f}. Three hidden layers of size 1024, 12288, and 567
+do the neural magic. The output layer is 837 nodes; the bulk of which
+is a ``one-hot'' representation of the predictions for the 64 squares,
+each with 13 possible contents (black or white, six pieces, or empty).
+This is $64 \times 13 = 832$ nodes. Then four nodes to predict the
+four castling bits, and one to predict the current side to move. This
+model does not predict the possibility for {\em en passant} capture,
+nor move counters or position repetition. This will not be its main
+source of disadvantage!
+
+I trained the network in two phases, first starting with a densely
+connected one (model size 160 MB), and then after I get fed up with
+how slow training was, a ``vacuumed'' version of the network where I
+removed edges with low absolute weights (model size 5 MB) to continue
+training. Removing edges based on an absolute weight threshold is
+very unprincipled (since a downstream edge can magnify a contribution
+arbitrarily) but I did it anyway.
+
+Everything was trained on a single GPU, though a fairly decent one in
+2018, the EVGA GeForce GTX 1080 FTW (really), using OpenCL. Biases
+were initialized to $0$ and weights with a Gaussian of mean $0$ and
+standard deviation $0.025$. In the first phase, there were 64 examples
+per round, and after vacuuming, 2048. The round learning rate
+$\alpha_r$ started at $0.1$ and descended linearly to $0.002$ over
+500,000 rounds; and the learning rate when updating weights (for each
+example) $\alpha$ is $\alpha_r / \verb+examples_per_round+$. In no way
+am I recommending these parameters. Fiddling with the parameters to
+make it do its black magic or alternately carom off to a sea of NaNs
+or zeroes is for sure the worst part of neural networks. In fact, I
+initially started with the classical sigmoid transfer function, but
+``upgraded'' to the ``leaky rectified linear'' function
 %
+$$(p < 0) \,?\, p \times 0.01 : p$$
+%
+after getting fed up with sigmoid weights caroming off (see
+``vanishing gradient problem'' and/or ``exploding gradient problem'').
+The final model was trained over 339,885 rounds on 223 million
+examples. It did not appear to show signs of improving for
+several days before I terminated it.
+
+\subsection{Statistical evaluation}
+
+The unblinding component can be evaluated on its own, by running it on
+examples sampled independently of the training set. The model outputs
+a score for each possible contents of each square; we simply
+discretize to the highest-scoring one (same too for the predicted
+castling state and turn). Over 50,000 examples, these were the
+results:
+
+9,584 predicted boards were exactly correct (19.17\%). There were a
+total of 161,166 piece mistakes, which is an average of 3.22 per
+board. This is wherever we predict a square's contents incorrectly.
+There were only 1630 castling status mistakes, an average of 0.03 per
+board (there can be up to four mistakes per board). This is probably
+because when the king and rook are in their starting squares, castling
+is almost always still allowed. In 19,014 boards, we mispredicted
+whose move it is (38\%). This appears to be the most difficult thing
+to predict, which is not surprising.\footnote{ Prior to ``vacuuming'',
+  the 160 MB network actually performed slightly better than the final
+  5 MB network, with 21.20\% of boards exactly correct, and an average
+  of 3.12 mistakes per board. This suggests that the model may only be
+  doing a limited amount of generalization, instead mostly memorizing
+  board positions. Representing the 223 million examples seen exactly
+  (using our best board representation described in
+  Section~\ref{sec:representation}) would take 42.8 GB, so at 5 MB at
+  least the data is represented compactly, if also rather lossily.}
+
+\subsection{Subjective evaluation}
+
+The unblinder {\em must} make mistakes since the 64-bit representation
+is ambiguous. Subjectively, the unblinder makes reasonable mistakes.
+It is excellent at common openings, usually getting these exactly
+correct. On the other hand, it is fairly bad at sparse endgames, where
+it is difficult to tell a pawn from a rook from a king. It is terrible
+at unlikely positions that can be confused for likely ones. If you are
+playing against it and know how it works, it is easy to trick it by
+doing something like capturing one of its starting-position pawns with
+your queen; nobody does this in real games (because the queen can be
+immediately recaptured), so the square is predicted as a pawn and the
+queen ``disappears'' to the unblinder (Figure~\ref{fig:trick}). Having
+an invisible queen in your camp is of course very dangerous. Resolving
+ambiguities in favor of more likely positions is the right thing for
+the model to do, so this is just an inherent flaw with the
+decomposition of the problem. There are some ways we can account for
+this (Section~\ref{sec:spycheck}).
+
+\begin{figure}[ht]
+XXX actual example
+\caption{The white queen ``disappears'' after QxWhatever, because
+  unblinding predicts it to be one of black's own pawns, which is far
+  more likely in that square.} \label{fig:trick}
+\end{figure}
+
+A few things are distinctly disappointing about its performance. Even
+outside of ``likely'' positions, it usually predicts that pieces on
+black's side of the board are black, and vice versa
+(Figure~\ref{fig:allon}). This makes sense, but suggests serious
+limitation on using the prediction to play chess. Less forgivably, it
+sometimes predicts more than one king per side (or zero), which is
+always wrong. Actually, an early version had this problem in spades,
+usually frequently predicting two or three kings. Upon debugging, I
+had simply made the noob mistake of printing both King and Knight with
+the letter ``K.'' Ha! It often predicts the ``wrong'' number of
+bishops (etc.), or places them on the same color. This is technically
+possible through promotion, but usually a bad guess, since promotion
+is relatively rare, and moreso promotion to a piece other than a
+queen. An approach that might handle this better (but may have
+different downsides) would be instead to predict the ``fates'' of the
+32 initial pieces~\cite{survival}. The fate of a pawn includes what if
+any piece it has promoted to, but no other pieces can. This would {\em
+  require} that the model only predict a single location for each
+king, among other things. However, this would require a much larger
+output layer (32 pieces can move to 64 squares, plus promotion) and it
+is not always clear how to interpret its value into a position (for
+example, if two pieces are predicted strongly to be on the same
+square).
+
+\begin{figure}[ht]
+\includegraphics[width=0.9 \linewidth]{blind-allon}
+\caption{What the model predicts (in single-king mode;
+  Section~\ref{sec:using}) for {\tt 0xFFFFFFFFFFFFFFFF}, the board
+  with all bits set. This is an impossible position, but it gives some
+  idea of the model's biases for each square. Notably, most of the pieces
+  on each half of the board have a single color. This makes sense, but
+  also suggests substantial limitation. When single-king mode is off,
+  the bottom right king is predicted as a white rook.} \label{fig:allon}
+\end{figure}
+
+\section{Playing blind}
+
+Once we have predicated a board state, we can play with it. The
+simplest way to do this is to use a strong chess engine to pick a move
+for the position. Here, I use Stockfish with a node budget of 1
+million nodes, which takes about 1 second of CPU time per move on my
+computer; henceforth {\sf stockfish1m}.
+
+There are some complications:
+\begin{itemize}
+\item Frequently, the unblinded board will not match the true
+  position, and Stockfish will choose a move that is illegal. So, as discussed
+  before, we actually return a prioritzed list of moves. For this first
+  experiment, we just return the move Stockfish recommends, followed by all
+  other moves ordered randomly.
+\item Stockfish is a very strong engine, and in my opinion the code is generally
+  good, but it is very sensitive to bad FEN (the notation used to give it a
+  position to evaluate) strings. Given a bad string, like one that says castling
+  is possible when the king isn't in its home square, often crashes the engine.
+  So we need to make sure to actually pass valid positions. I accomplish this
+  by making the following modifications to the predicted board:
+  \begin{itemize}
+  \item If a castling bit is set, but castling is not possible due to the position
+    of the king or rook, I clear the bit.
+  \item Set the side-to-move to be the correct actual value. This uses
+    the unblinded state, so is superficially cheating. But note that
+    if we get the side wrong, then Stockfish's move will always be
+    illegal: Moves are specified as a source and destination
+    square,\footnote{Plus promotion piece. Castling is represented as
+      a two-square move of the king.} and so the source square of
+    Stockfish's move would always be a piece of the wrong piece's
+    color. So this is equivalent to (but twice as efficient as)
+    running stockfish twice, one for each side, and prioritizing the
+    predicted side's move first.
+  \end{itemize}
+  This won't fix all positions, for example, if the white and black king are
+  adjacent to one another in mutual check. If an irreparable problem is detected,
+  then I just return a uniformly random move list.
+\end{itemize}
+
+It is easy to beat this chess engine, by tricking it as in
+Figure~\ref{fig:trick}, although this involves unnatural moves, so it
+may only apply if you know how it works.
+% XXX example game?
+Measuring how well it plays in an absolute sense is a subject of some
+interest, so I wrote a separate paper about that~\cite{eloworld}. This
+algorithm, called \verb+blind_yolo+, had an Elo World score of XXX.
+
+\subsection{We three kings}
+
+When evaluating the first version I found that it was predicting a
+disappointingly high number of illegal positions in practice, which
+was causing us to fall back on making random moves, which is mostly
+boring. The second version reduces the rate of illegal positions due
+to too many or too few kings~\cite{kingme}.
+
+The model predicts a score for each type of piece in each square, and
+we do not have to necessarily interpret it by always taking the
+highest-scoring piece. This version first finds the two squares with
+the highest scores for the white king, and same for the black king. We
+take two in case the same square is predicted for both. Then this
+square gets one of the kings (whichever has higher score) and the
+other king goes in the highest-scoring unclaimed square. The rest of
+the squares get the highest-scoring prediction as before, but we never
+predict kings for them.
+
+This change just affects the unblinding process, so we can directly
+evaluate its accuracy. It gets 19.28\% of positions exactly correct
+(slightly better), with an average of 3.26 piece mistakes per position
+(slightly worse). This is expected; we exchange local mistakes (each
+was trained independently to minimize its local error) for global
+correctness (which is not taken into account at all during training).
+
+This version, called \verb+blind_kings+, performs consistently better
+than \verb+blind_yolo+ (63 wins, 45 losses, 412 draws).
+% XXX could run a lot more games between these two.
+% as white: 31/24/207, as black 21/32/205
+It had an Elo World score of XXX.
+
+\subsection{Spy check} \label{sec:spycheck}
+
+Say \verb+blind_kings+ is playing as black; it remains easy to fool it
+by moving black pieces into white's camp, since they are usually then
+predicted to be white pieces. We can defend against this somewhat.
+Since it is illegal to capture one's own piece, there is little risk
+in trying; if it is indeed our own piece then the move will be
+rejected, and if it is not our piece, then capturing is good for two
+reasons: One, we capture a piece, and two, we avoid having
+Stockfish make a move in this incorrectly predicted board. (Of course
+there are many reasons why eagerly capturing a piece can be a bad
+idea, but at this level of play, an edge in material is likely worth
+it.)
+
+There is one subtlety here. Above we argued that it was safe to use
+the actual side-to-move instead of the predicted one; but here it
+would not be equivalent to do so. Instead, we first prioritize all
+apparent spy-check moves where the predicted source piece matches the
+predicted side-to-move, then we try the opposite. (Ties are broken by
+preferring to capture with a lower-value predicted piece, and then
+randomly.) Due to this, there is some additional chance that we end up
+making an especially dumb move because we both mispredicted the
+side-to-move and the identity of some pieces.
+
+(XXX evaluation here)
+
+\subsection{Future work}
+
+The predicted board often expresses uncertainty about some squares,
+which could be thought of as probabilities. A principled improvement
+would be to try to find moves that are good in expectation, that is,
+integrated all possible boards in proportion to their predicted
+probability. A good approximation might be had by sampling a bunch
+of boards according to the predicted distribution, and then using
+Stockfish to score the top $k$ moves for each; we can then order
+moves by their expected score. Unfortunately, it is not easy to
+efficiently get Stockfish to generate scored moves for $k \neq 1$.
+Even with $k = 1$, this approach would be slow, taking about a
+second for each (distinct) sampled board. So I did not try it,
+at least not before submitting this paper.
+
+\medskip
+I initially considered trying to solve this whole problem with neural
+networks. The current best known engine in the world (AlphaZero) at
+least {\em uses} a neural network.
+% XXX is the whole thing just a network? I guessed it's probably
+% alpha-beta plus a NN evaluation function?
+The biggest advantage would be that it would naturally be able to
+consider multiple moves under uncertainty about the board state, as
+just discussed, without any particular extra logic. My plan was to
+make a multiple different components that coud be evaluated
+separately, starting with the unblinder described, followed by a unit
+that predicts legal moves, and then a unit that takes these two (and
+also the 64-bit blinded mask if it likes) and scores each move.
+Training a reasonable unblinder took longer than I expected, and the
+legal move predictor never really worked that well,
+% (XXX some evaluation of it)
+so I just settled for basing it off the single unblinder unit. Can you
+do better?
+
+\section{Conclusion}
+% 
+% - How do the solutions compare to other chess algorithms that
+%   could be applied in the blind? Random works, also lexicographic,
+%   any others? Most are inspecting the board state, obviously.
+% - Could estimate the upper bound for accuracy by blinding every
+%   game in the database, and doing a ``lookup''-based version of
+%   the eval.
+
+I would like to thank the little people (pawns) and the anonymous
+referrees.
+
 % % make -j eval-unblinder.exe && ./eval-unblinder.exe net.val
@@
 %   183451 total mistakes (3.67/pos)
-% 
-% It's actually kind of heartening that it performs worse with this approach,
-% though we are more likely to get the whole board correct.
-% 
-% (cite #kingme)
-%
-
-% all bits set to 1 (with single_kings)
-% (without single_kings, only difference is that the bottom right becomes a
-% rook; no white queen)
-\includegraphics[width=0.9 \linewidth]{blind-allon}
 

### 2019-03-10 trunk/chess/papers/blind.tex
OK, this is pretty much ready to submit except for the XXXs when I get back home

@@
 each other in the face repeatedly during the game to impair brain
-function (see Chess Boxing~\cite{wikipediachessboxing}). Another common
-distraction is to play a multitude of games against many opponents at
-the same time, in a so-called ``simultaneous exhibition.'' The idea is
-that this is more challenging because of the need to maintain mental
-state for so many games at once, whereas your opponents only need to
-maintain state for one game. In truth, simultaneous exhibitions easily
-fall to a ``man-in-the-middle attack.'' If the purported genius simply
-creates a bipartite matching of the games played with the white pieces
-and the games played with black, he can mechanically forward moves
-between these pairs of boards. This requires only constant state (see
-next section) per pair of games, and guarantees an even score for the
-exhibition. So that's not very impressive.
+function (see Chess Boxing~\cite{wikipediachessboxing}). Another
+common distraction is to play a multitude of games against many
+opponents at the same time, in a so-called ``simultaneous
+exhibition.'' The idea is that this is more challenging because of the
+need to maintain mental state for so many games at once, whereas your
+opponents only need to maintain state for one game. In truth,
+simultaneous exhibitions easily fall to a ``man-in-the-middle
+attack.'' If the purported genius simply creates a perfect bipartite
+matching of the games played with the white pieces and the games
+played with black, he can mechanically forward moves between these
+pairs of boards. This requires only constant state (see next section)
+per pair of games, and guarantees an even score for the exhibition. So
+that's not very impressive.
 
@@
 is a blindfold (worn over the eyes). In this predicament they only
-hear the opponent announce moves and must keep the position on the
+hear the opponent announce moves and must imagine the position on the
 board in their mind's eye, both for the sake of remembering it and
@@
 consumer hardware, to defeat the strongest human players. It is well
-known that striking a computer actually {\em fixes} it, so chess
-boxing becomes trivial. Blindfold chess is the natural interface for a
+known that striking a computer actually {\em fixes} it, so Chess
+Boxing becomes trivial. Blindfold chess is the natural interface for a
 chess computer; it is actually {\em much more difficult} to have the
@@
 single game, although in principle the scale of such a thing could
-still be impressive. With a modern chess engine, it is easy to scale
+still be impressive. This is also impressive to other computers, who
+are largely concerned with filling up their memories with efficiently
+coded data. With a modern chess engine, it is easy to scale
 to an arbitrary number of games, since the exhibitor can make progress
 by observing one of the boards, computing a strong move, and playing
-it; this requires $O(0)$ space because all of the state is stored
+it; this requires $O(1)$ space because all of the state is stored
 externally in the exhibition itself. 
@@
 be yet stronger computers). The man-in-the-middle attack remains an
-efficient way way to minimize loss (ensuring an exactly even score).
+efficient way to minimize loss (ensuring an exactly even score).
 The simplest way to do this is to explicily generate a perfect
@@
 \item We are waiting for a move from Walice (our white opponent) \label{state:mim1}
-\item We have seen a move from Walice, which is \underline{\hspace{6em}}. \label{state:mim2}
+\item We have seen a move from Walice, which is \underline{\hspace{3em}}. \label{state:mim2}
 \item We are waiting for a move from Bob (our black opponent) \label{state:mim3}
@@
 Walice and Bario; Waluigi and Bob. So, rather than store the
-matching explicitly, we can reconstruct it from the stored state.
-We two vectors of $n/2$ rows, the ``source'' and ``destination'' rows,  (XXX I think we do not need destination)
-each containing a game {\em position} (not move).
-This position represents XXXX (like the last time we saw it?).
-The positions are initialized to the starting position. Like before
-we can be in one of four states:
-\begin{itemize}
-\item Waiting for all ``source'' players (who play as white) to complete their moves. \label{state:cmim1}
-\item Forwarding our moves from ``source'' to our black opponents. \label{state:cmim2}
-\item `` \ldots '' (who play as black) `` \ldots '' \label{state:cmim3}
-\item `` \ldots '' white opponents. \label{state:cmim3}
-\end{itemize}
+matching explicitly, we can reconstruct it from the stored state
+at each step.
+
+Let's think about the process of forwarding the moves from our white
+opponents to our black opponents; the reverse is of course symmetric.
+The first step will be to wait for the white opponents to make their
+moves, and then copy these $n/2$ positions (not moves) into a vector
+in memory.
 
 \newcommand\stepsto[1]{\stackrel{#1}{\rightarrow}}
+The black opponents are waiting for a move from us. Next, we'll copy
+their $n/2$ positions into memory, aligned with the $n/2$ positions
+already present. Let's say that the relation that defines a legal
+move in chess is
+%
+$$
+B \stepsto{m} B'
+$$ 
+%
+where $B$ is the position before the move $m$, and $B'$ is the
+resulting position. Our goal is to align the games such that $B'$ (a
+position copied from our white opponent) is aligned with $B$ (a
+position pending a move for our black opponent) in memory; this will
+represent the perfect matching. Computing $m$ from $B$ and $B'$ when
+$B \stepsto{m} B'$ is easy (and unique), so this allows us to read off
+and play the move for each row.
 
-% (XXX So actually, simpler here is just: Loop until all white players have
-% made their moves. Then copy all their boards to our board vector.)
-While in State~\label{state:cmim1} (which we also start in), we
-compare our stored source board states to the boards where we play as
-black. Once the white opponent has made a move, resulting in $B'$,
-there will be some position $B$ in our ``source'' set that is a valid
-predecessor for $B'$ (that is, there exists some legal move $M$ that
-advances $B$ to $B'$). For example, at the beginning, $B$ may be the
-starting position (in fact it must, since all boards are initialized
-this way) and $B'$ may be the position resulting from $1. e4$. With
-the matching constructed (see below) we copy each $B'$ into our set of
-boards, forgetting $B$.\footnote{In fact, we do not even need to
-  construct the matching, although it may help us with bookkeeping
-  (e.g., which games have already been copied?), depending on how the
-  exhibition itself is represented.} Then we transition to
-State~\label{state:cmim2}. In this state, we figure out what moves to
-play in the pending boards. By invariant, each one $B$ will be
-compatible with at least one of our source boards $B'$; we can iterate
-over source boards until we find one in constant space (and mark it
-as ``taken'' with . Compatible
-means that there exists a move $m$ where playing $m$ in board $B$
-results in $B'$. If $m$ exists, it is unique (and easy to compute), so
-this is the move that we will play. The only tricky thing is figuring
-out which boards go with which; this is the problem of reconstructing
-the perfect matching.
+By invariant, it will be possible to produce such an alignment. For
+example, the first time we do this, each $B$ will be the starting
+position, and $B'$ will be a legal move made by white from the
+starting position. Any alignment will work. Let's say that just one of
+the white opponents played 1. d4, resulting in $B_0$; then one black
+opponent will make a legal response to this (say 1. \ldots Nf6, giving
+$B_1$). Then $B_1$ can be $B'$ for the next round, which we can align
+with $B=B_0$, and so on.
 
+The only tricky thing is figuring out which boards go with which.
 Although if $B \stepsto{m} B'$ it is easy to deduce $m$, it is not
@@
 
-we could have $B_a$ and $B_b$ be
+\noindent \ldots we could have $B_a$ and $B_b$ be
 
@@
 
-\ldots and have no possible matching board, since it cannot follow
-$B_a$.
+\noindent \ldots and have no possible matching board, since it cannot
+legally follow $B_a$.
 
@@
 We only need a single pseudorandom pool for the entire process,
-so it can be O(n) bits; this seems easily enough to generate all
-possible permutations of $n/2$ items. Even $n^{2n}$ grows much
+so it can be $O(n)$ bits; this seems easily enough to generate all
+possible permutations of $n/2$ items. Even $2^{2n}$ grows much
 faster than $n!$. If we don't like the random approach, I believe
@@
 take exponential time, it allows us to play against $n$ simultaneous
-opponents using $O(n)$ space!\footnote{ 
-%
-If the games in the exhibition were presented to us in some way that
-allowed us to visit them in the same order multiple times, we could
-perhaps get away with {\em only} the random pool? XXX }
+opponents using $O(n)$ space!
+% If the games in the exhibition were presented to us in some way that
+% allowed us to visit them in the same order multiple times, we could
+% perhaps get away with {\em only} the random pool? XXX
 
@@
 draw by repetition, which make reference to the history of the game
-(See Footnote~1 in {\it Survival in chessland}\cite{survival} and seem
+(See Footnote~1 in {\it Survival in chessland}~\cite{survival}) and seem
 to require storing all previous board states. Fortunately, if we are
@@
 bits. The Thursd'z Institute considered more compact
-representations\cite{representations}; one nice choice works as follows:
+representations~\cite{representations}; one nice choice works as follows:
 
@@
 follow with 32 four-bit records; these indicate the piece's color and
-type.\footnote{(Since only 32 of the 64 bits can be set, you could do
-  slightly better by representing $\binom{60}{32}$ in $\sim 61$ bits.
+type.\footnote{Since only 32 of the 64 bits can be set, you could do
+  slightly better by representing $\binom{64}{32}$ in $\sim 61$ bits.
   When fewer than 32 squares are occupied, we can use a record
@@
   we should ignore the remaining bits. However, this gets vastly more
-  complicated for only 3 bits of savings.)} 
+  complicated for only 3 bits of savings.} 
 %
-With the 9 bits of extra state above, this totals $64+32 \ times 4+9=201$
+With the 9 bits of extra state above, this totals $64+32 \times 4+9=201$
 bits. There is some slack in the representation, because there are
@@
   did however read the rather bad Wikipedia article on the
-  topic\cite{wikipediaboard} which describes a Huffman-based encoding
+  topic~\cite{wikipediaboard} which describes a Huffman-based encoding
   that uses a ``maximum of 204 bits, and often much less.'' This also
@@
 \[
-  {\tt makemove} : {\tt uint64} \rightarrow {\tt move list}
+  {\tt makemove} \;:\; {\tt uint64} \rightarrow {\tt move\ list}
 \]
@@
 \[
-{\tt unblind} : {\tt uint64} \rightarrow {\tt position}
+{\tt unblind} \;:\; {\tt uint64} \rightarrow {\tt position}
 \]
@@
 make it do its black magic or alternately carom off to a sea of NaNs
-or zeroes is for sure the worst part of neural networks. In fact, I
+or zeroes is for sure the worst part of neural networks. Indeed, I
 initially started with the classical sigmoid transfer function, but
@@
 decomposition of the problem. There are some ways we can account for
-this (Section~\ref{sec:using}). % XXX make sure I did this work
+this (Section~\ref{sec:spycheck}).
 
@@
 always wrong. Actually, an early version had this problem in spades,
-usually frequently predicting two or three kings. Upon debugging, I
-had simply made the noob mistake of printing both King and Knight with
-the letter ``K.'' Ha! It often predicts the ``wrong'' number of
-bishops (etc.), or places them on the same color. This is technically
-possible through promotion, but usually a bad guess, since promotion
-is relatively rare, and moreso promotion to a piece other than a
-queen. An approach that might handle this better (but may have
-different downsides) would be instead to predict the ``fates'' of the
-32 initial pieces~\cite{survival}. The fate of a pawn includes what if
-any piece it has promoted to, but no other pieces can. This would {\em
-  require} that the model only predict a single location for each
-king, among other things. However, this would require a much larger
-output layer (32 pieces can move to 64 squares, plus promotion) and it
-is not always clear how to interpret its value into a position (for
-example, if two pieces are predicted strongly to be on the same
+frequently predicting two or three kings. Upon debugging, I had simply
+made the noob mistake of printing both King and Knight with the letter
+``K.'' Ha! It often predicts the ``wrong'' number of bishops (etc.),
+or places them on the same color. This is technically possible through
+promotion, but usually a bad guess, since promotion is relatively
+rare, and moreso promotion to a piece other than a queen. An approach
+that might handle this better (but may have different downsides) would
+be instead to predict the ``fates'' of the 32 initial
+pieces~\cite{survival}. The fate of a pawn includes what if any piece
+it has promoted to, but this is not necessary for the other pieces.
+This would {\em require} that the model only predict a single location
+for each king, among other things. However, this would require a much
+larger output layer (32 pieces can move to 64 squares, plus promotion)
+and it is not always clear how to interpret its value into a position
+(for example, if two pieces are predicted strongly to be on the same
 square).
@@
 \caption{What the model predicts (in single-king mode;
-  Section~\ref{sec:using}) for {\tt 0xFFFFFFFFFFFFFFFF}, the board
+  Section~\ref{sec:singleking}) for {\tt 0xFFFFFFFFFFFFFFFF}, the board
   with all bits set. This is an impossible position, but it gives some
@@
 million nodes, which takes about 1 second of CPU time per move on my
-computer; henceforth {\sf stockfish1m}.
-
-There are some complications:
+computer. There are some complications:
 \begin{itemize}
@@
   position, and Stockfish will choose a move that is illegal. So, as discussed
-  before, we actually return a prioritzed list of moves. For this first
+  before, we actually return a prioritized list of moves. For this first
   experiment, we just return the move Stockfish recommends, followed by all
@@
   \begin{itemize}
-  \item If a castling bit is set, but castling is not possible due to the position
-    of the king or rook, I clear the bit.
+  \item If a castling bit is set, but castling is not possible due to
+    the position of the king or rook, clear the bit.
   \item Set the side-to-move to be the correct actual value. This uses
@@
 interest, so I wrote a separate paper about that~\cite{eloworld}. This
-algorithm, called \verb+blind_yolo+, had an Elo World score of XXX.
+algorithm, called \verb+blind_yolo+, had an Elo World score of XXX. It
+beats a purely random player with a score of 101 wins, 27 losses, and
+389 draws. Making moves purely at random is one of the few fair
+comparisons, since the random strategy also works with color- and
+piece-blind chess.
 
-\subsection{We three kings}
+\subsection{We three kings} \label{sec:singleking}
 
@@
 
-The model predicts a score for each type of piece in each square,
-and we do not have to necessarily interpret it by always taking
-the highest-scoring piece. This version first finds the two squares
-with the highest scores for the white king, and same for the black
-king. We take two in case the same square is predicted for both.
-Then this square gets one of the kings (whichever has higher score)
-and the other king goes in the highest-scoring unclaimed square. The
-rest of the squares just get the highest-scoring prediction as before,
-but we never predict kings.
+The model predicts a score for each type of piece in each square, and
+we do not have to necessarily interpret it by always taking the
+highest-scoring piece. This version first finds the two squares with
+the highest scores for the white king, and same for the black king. We
+take two in case the same square is predicted for both. Then this
+square gets one of the kings (whichever has higher score) and the
+other king goes in the highest-scoring unclaimed square. The rest of
+the squares get the highest-scoring prediction as before, but we never
+predict kings for them.
+
+This change just affects the unblinding process, so we can directly
+evaluate its accuracy. It gets 19.28\% of positions exactly correct
+(slightly better), with an average of 3.26 piece mistakes per position
+(slightly worse). This is expected; we exchange local mistakes (each
+was trained independently to minimize its local error) for global
+correctness (which is not taken into account at all during training).
 
@@
 
-\subsection{Spy check}
+\subsection{Spy check} \label{sec:spycheck}
 
@@
 
+\subsubsection{No, {\em u r} a lnetwork} \label{sec:legalmove}
+I initially considered trying to solve this whole problem with neural
+networks. The current best known engine in the world (AlphaZero) at
+least {\em uses} a neural network.
+% XXX is the whole thing just a network? I guessed it's probably
+% alpha-beta plus a NN evaluation function?
+The biggest advantage would be that it would naturally be able to
+consider multiple moves under uncertainty about the board state, as
+just discussed, without any particular extra logic. My plan was to
+make a multiple different components that could be evaluated
+separately, starting with the unblinder described, followed by a unit
+that predicts legal moves, and then a unit that takes these two (and
+also the 64-bit blinded mask if it likes) and scores each move.
+Predicting a legal move is also a natural function for machine
+learning; a move can be given just as a source and destination
+square.\footnote{ There are also four choices for promotion when
+  moving a pawn into the last rank. It is always the case that if any
+  promotion is legal, all choices are legal, so this does not need to
+  be encoded in this phase. Also, at this level of play, always
+  promoting to queen is a very safe simplification.} Many pairs of
+squares are always impossible (e.g. no piece can ever move from A1 to
+B8); so there are only 1792 potential moves to predict. However,
+training a reasonable unblinder took longer than I expected, and the
+legal move predictor never really worked that well (it has a harder
+job),
+% (XXX some evaluation of it)
+so I just settled for basing it off the single unblinder unit. Can you
+do better?
+
 \section{Conclusion}
+% 
+% - How do the solutions compare to other chess algorithms that
+%   could be applied in the blind? Random works, also lexicographic,
+%   any others? Most are inspecting the board state, obviously.
+% - Could estimate the upper bound for accuracy by blinding every
+%   game in the database, and doing a ``lookup''-based version of
+%   the eval.
 
-- Other structures for the network
-- How do the solutions compare to other chess algorithms that
-  could be applied in the blind? Random works, also lexicographic,
-  any others? Most are inspecting the board state, obviously.
-- Could estimate the upper bound for accuracy by blinding every
-  game in the database, and doing a ``lookup''-based version of
-  the eval.
+I would like to thank the little people (pawns) and the anonymous
+referrees.
 
@@
 %   183451 total mistakes (3.67/pos)
-% 
-% It's actually kind of heartening that it performs worse with this approach,
-% though we are more likely to get the whole board correct.
-% 
-% (cite #kingme)
-%
 

### 2019-03-11 trunk/chess/papers/weak.tex
Experiment to predict legal moves. Packed representation of moves. Mode for tournament where we only compute one row/col. Add stockfish node option and Stockfish1M player
start writing 'weak'

@@
+% [computermodern], 
+\documentclass[numbers]{sigplanconf}
+
+\usepackage{url}
+% \usepackage{code}
+% \usepackage{cite}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{chessboard}
+
+% \usepackage{chessfs}
+\usepackage{adjustbox}
+
+% Define black versions of pieces for inline use. Gross, but it works.
+\newcommand{\Pawn}[1][1.3ex]{%
+\adjustbox{Trim=4.3pt 2.6pt 4.3pt 0pt,width=#1,margin=0.2ex 0ex 0.2ex 0ex}{\BlackPawnOnWhite}%
+}%
+\newcommand{\Rook}[1][1.58ex]{%
+\adjustbox{Trim=3.2pt 2.2pt 3.2pt 0pt,width=#1,raise=0ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackRookOnWhite}%
+}%
+\newcommand{\Knight}[1][1.85ex]{%
+\adjustbox{Trim=2.3pt 2.35pt 2.5pt 0pt,width=#1,raise=-0.03ex,margin=0.14ex 0ex 0.14ex 0ex}{\BlackKnightOnWhite}%
+}%
+\newcommand{\Bishop}[1][1.79ex]{%
+\adjustbox{Trim=2.3pt 2pt 2.3pt 0pt,width=#1,raise=-0.12ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackBishopOnWhite}%
+}%
+\newcommand{\Queen}[1][2.05ex]{%
+\adjustbox{Trim=1.2pt 2.2pt 1.2pt 0pt,width=#1,raise=-0.08ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackQueenOnWhite}%
+}%
+\newcommand{\King}[1][1.95ex]{%
+\adjustbox{Trim=2pt 2pt 2pt 0pt,width=#1,raise=-0.06ex,margin=0.13ex 0ex 0.13ex 0ex}{\BlackKingOnWhite}%
+}%
+
+\interfootnotelinepenalty=0
+
+% lets me explicitly set a. or 1. etc. as enum label
+\usepackage{enumitem}
+
+\pagestyle{empty}
+
+\usepackage{ulem}
+% go back to italics for emphasis, though
+\normalem
+
+\usepackage{natbib}
+
+\setlength{\footnotesep}{2em}
+
+% \newcommand\comment[1]{}
+\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
+
+\begin{document}
+
+\conferenceinfo{SIGBOVIK~2019}{April 1, 2019, Pittsburgh, Pennsylvania, USA}
+\copyrightyear{2019}
+\title{Elo World,}
+\subtitle{a framework for benchmarking weak chess engines}
+\authorinfo{Dr. Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
+\toappear{
+Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
+Appears in SIGBOVIK 19119 with the
+%
+insufficient material 
+%
+of the Association for Computational Heresy; {\em IEEEEEE!}
+press, Verlag-Verlag volume no.~0x40-2A. 1600 rating points}
+
+\ccsdesc[500]{Evaluation methodologies~Tournaments}
+\ccsdesc[500]{Chess~Being bad at it}
+% \ccsdesc[100]{Software and its engineering~Control structures}
+% \ccsdesc[300]{Theory of computation~Control primitives}
+
+\terms
+pawn, horse, bishop, castle, queen, king
+\keywords
+chess, bad
+
+\setchessboard{showmover=false}
+
+\newcommand\checkmate{\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}
+
+\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
+\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
+\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
+\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
+\newcommand\at{\ensuremath{\scriptstyle @}}
+
+\date{1 April 2019}
+
+\maketitle \thispagestyle{empty}
+
+Fiddly bits aside, it is a solved problem to maintain a numeric rating
+of players for some game (for example chess, but also sports,
+e-sports, probably also z-sports if that's a thing). Though it has
+some competition (suggesting the need for a meta-rating system to
+compare them), the Elo Rating System~\cite{elo1978rating} is a simple
+and effective way to do it.
+
+The gist of this system is to track a score for each player. The
+scores are defined such that the expected outcomes of games can
+be computed from the scores (for example, a player with a rating
+of 2400 should win 9/10 of her games against a player with a
+rating of 2000). If the true outcome doesn't match the
+
+.. so we can just leave it at that
+
+% Names:
+% "Stalemate"
+% "Weaksauce"
+% "back rank"
+% "blunder"
+% something with ..elo..   "elo world"
+% inFIDElity
+% blunderbuss
+%
+
+Interpolation, like Scoville scale.
+
 % Dramatic failure of the "suicide king" strategy against
@@
 1. d4 g5 2. Bxg5 Nc6 3. Bxe7 Kxe7 4. d5 Kd6
-5. dxc6+ Kc5 6. Qd6+ Kc4 7. e4#
+5. dxc6+ Kc5 6. Qd6+ Kc4 7. e4++
 
@@
 
+% worstfish (black) somehow beating suicide king
+% white forces the mate. Note that worstfish
+% does assume its opponent will pick good moves...
+1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
+5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
+9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
+13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
+17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
+21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
+25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
+29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
+33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
+37. Qxd7+ Bxd7++
+
+% Fate-based algorithms.
+% The deterministic version is fairly boring, since the
+% highest probability states are similar to the starting
+% position (this might not be true for "survivalist"
+% or "fatalist"? check?)
+
+% weighted random is almost the same as random, because
+% the weights tend to be very similar when treated
+% additively. They perform slightly different from
+% random, as so (101.5 million games):
+
+%  player	elo	w/l/d
+%  dangerous	986.29	1374577/1501526/15683897
+%  popular	997.34	1360053/1502995/15696952
+%  random_move	1000.79	1540798/1337668/15681534
+%  rare	1005.85	1554600/1326707/15678693
+%  safe	1009.72	1341030/1502162/15716808
+
+% then with normalization:
+%  player	elo	w/l/d
+%  popular	990.93	25756/43561/346683
+%  safe	993.49	28082/40769/347149
+%  dangerous	996.60	30063/37451/348486
+%  rare	1005.82	50656/27390/337954
+%  random_move	1013.15	41747/27133/347120
+
 // Good site for some computer chess programs with ELO:
@@
 
+\nocite{elo1978rating}
+\nocite{topple}
 // citation for knights distance:
-// Counting the Number of Squares Reachable in k
-// Knight's Moves
-// Amanda M. Miller, David L. Farnsworth
-// Open Journal of Discrete Mathematics, 2013, 3, 151-154
-// http://dx.doi.org/10.4236/ojdm.2013.33027
+\nocite{miller2013counting}
+
+\bibliography{chess}{}
+\bibliographystyle{plain}
+
+\end{document}

### 2019-03-11 chess/papers/weak.tex
Better progress updates in Tournament. About to do incremental rewrite
Tournament now accumulates results to database. Separated elo/table code out into its own program.
Add fatalist, survivalist, equalizer. improve Player interface and ability to select games for tournament
Did some writing on weak.tex

--- 2019-02-09 trunk/chess/papers/weak.tex
@@
+% [computermodern], 
+\documentclass[numbers]{sigplanconf}
+
+\usepackage{url}
+% \usepackage{code}
+% \usepackage{cite}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{chessboard}
+
+% \usepackage{chessfs}
+\usepackage{adjustbox}
+
+% Define black versions of pieces for inline use. Gross, but it works.
+\newcommand{\Pawn}[1][1.3ex]{%
+\adjustbox{Trim=4.3pt 2.6pt 4.3pt 0pt,width=#1,margin=0.2ex 0ex 0.2ex 0ex}{\BlackPawnOnWhite}%
+}%
+\newcommand{\Rook}[1][1.58ex]{%
+\adjustbox{Trim=3.2pt 2.2pt 3.2pt 0pt,width=#1,raise=0ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackRookOnWhite}%
+}%
+\newcommand{\Knight}[1][1.85ex]{%
+\adjustbox{Trim=2.3pt 2.35pt 2.5pt 0pt,width=#1,raise=-0.03ex,margin=0.14ex 0ex 0.14ex 0ex}{\BlackKnightOnWhite}%
+}%
+\newcommand{\Bishop}[1][1.79ex]{%
+\adjustbox{Trim=2.3pt 2pt 2.3pt 0pt,width=#1,raise=-0.12ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackBishopOnWhite}%
+}%
+\newcommand{\Queen}[1][2.05ex]{%
+\adjustbox{Trim=1.2pt 2.2pt 1.2pt 0pt,width=#1,raise=-0.08ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackQueenOnWhite}%
+}%
+\newcommand{\King}[1][1.95ex]{%
+\adjustbox{Trim=2pt 2pt 2pt 0pt,width=#1,raise=-0.06ex,margin=0.13ex 0ex 0.13ex 0ex}{\BlackKingOnWhite}%
+}%
+
+\interfootnotelinepenalty=0
+
+% lets me explicitly set a. or 1. etc. as enum label
+\usepackage{enumitem}
+
+\pagestyle{empty}
+
+\usepackage{ulem}
+% go back to italics for emphasis, though
+\normalem
+
+\usepackage{natbib}
+
+\setlength{\footnotesep}{2em}
+
+% \newcommand\comment[1]{}
+\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
+
+\begin{document}
+
+\conferenceinfo{SIGBOVIK~2019}{April 1, 2019, Pittsburgh, Pennsylvania, USA}
+\copyrightyear{2019}
+\title{Elo World,}
+\subtitle{a framework for benchmarking weak chess engines}
+\authorinfo{Dr. Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
+\toappear{
+Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
+Appears in SIGBOVIK 19119 with the
+%
+title inflation
+%
+of the Association for Computational Heresy; {\em IEEEEEE!}
+press, Verlag-Verlag volume no.~0x40-2A. 1600 rating points}
+
+\ccsdesc[500]{Evaluation methodologies~Tournaments}
+\ccsdesc[500]{Chess~Being bad at it}
+% \ccsdesc[100]{Software and its engineering~Control structures}
+% \ccsdesc[300]{Theory of computation~Control primitives}
+
+\terms
+pawn, horse, bishop, castle, queen, king
+\keywords
+chess, bad
+
+\setchessboard{showmover=false}
+
+\newcommand\checkmate{\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}
+
+\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
+\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
+\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
+\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
+\newcommand\at{\ensuremath{\scriptstyle @}}
+
+\date{1 April 2019}
+
+\maketitle \thispagestyle{empty}
+
+\section{Introduction}
+
+Fiddly bits aside, it is a solved problem to maintain a numeric rating
+of players for some game (for example chess, but also sports,
+e-sports, probably also z-sports if that's a thing). Though it has
+some competition (suggesting the need for a meta-rating system to
+compare them), the Elo Rating System~\cite{elo1978rating} is a simple
+and effective way to do it. This paper is concerned with Elo in chess,
+its original purpose.
+
+The gist of this system is to track a score for each player. The
+scores are defined such that the expected outcomes of games can
+be computed from the scores (for example, a player with a rating
+of 2400 should win 9/10 of her games against a player with a
+rating of 2000). If the true outcome (of e.g.~a tournament) doesn't
+match the expected outcome, then both player's scores are adjusted
+towards values that would have produced the expected result. Over
+time, scores thus become a more accurate reflection of player's skill,
+while also allowing for players to get stronger or weaker. This
+system is carefully described elsewhere, so we can just leave it
+at that.
+
+The players need not be human, and in fact this can facilitate running
+many games and thereby getting accurate ratings.
+
+The problem this paper is concerned with is that basically all chess
+tournaments (whether with humans or computers or both) are between
+players who know how to play chess, are interested in winning their
+games, and have some reasonable level of skill. This makes it hard to
+give a rating to weak players: They just lose every single game and so
+tend towards a rating of $-\infty$\footnote{Some organizations don't
+  even let ratings fall beneath a certain level, for example, the
+  lowest possible USCF rating is 100.} Even if other comparatively
+weak players existed to participate in the tournament and occasionally
+lose to the player under study, it may still be difficult to
+understand how this cohort performs in any absolute sense. (In
+contrast we have ``the highest ever human rating was 2882,'' and
+``there are 5,323 players with ratings in 2200--2299'' and ``Players
+whose ratings drop below 1000 are listed on the next list as
+'delisted'.''\cite{fideratings}.) It may also be the case that all
+weak players always lose to all strong players, making it unclear just
+how wide the performance gap between the two sets is. The goal of
+this paper is to expand the dynamic range of chess ratings to span
+all the way from extremely weak players to extremely strong ones, while
+providing canonical reference points along the way.
+
+\section{Elo world}
+
+Elo World is a tournament with dozens of computer players. The
+computer players include some traditional chess engines, but also many
+algorithms chosen for their simplicity, as well as some designed to
+be competitively bad.
+
+The fun part is the various algorithms, so let's get into that. First,
+some ground rules and guidelines:
+\begin{itemize}
+\item No resigning (forfeiting) or claiming a draw. The player will
+  only be asked to make a move when there exists one, and it must
+  choose a move in finite time. (In practice, most of the players are
+  extremely fast, with the slowest ones using about one second of
+  CPU time per move.)
+\item The player can retain state for the game, and executes moves
+  sequentially (for either the white or black pieces), but cannot have
+  state meaningfully span games. For example, it is not permitted to
+  do man-in-the-middle attacks~\cite{blind} or learn opponent's moves
+  from previous rounds, or to get better at chess. The majority of
+  players are actually completely stateless, just a partial function
+  of type ${\tt position} \rightarrow {\tt move}$.
+\item black and white behave the same
+\item 
+\end{itemize}
+
+Ideally, such a system would be robust against adding new players,
+but this is probably not the case.
+
+% Names:
+% "Stalemate"
+% "Weaksauce"
+% "back rank"
+% "blunder"
+% something with ..elo..   "elo world"
+% inFIDElity
+% blunderbuss
+%
+
+Interpolation, like Scoville scale.
+
 % Dramatic failure of the "suicide king" strategy against

### 2019-03-13 chess/papers/weak.tex
Switch to two column, and start digging in

@@
 % [computermodern], 
-\documentclass[numbers]{sigplanconf}
+% \documentclass[numbers]{sigplanconf}
+\documentclass[10pt,preprint,twocolumn]{acmart}
+
+\usepackage[T1]{fontenc}
+\usepackage{libertine}%% Only as example for the romans/sans fonts
+\usepackage[scaled=0.85]{beramono}
 
@@
 
+% skip acm copyright shits
+\makeatletter
+\def\@copyrightspace{\relax}
+\makeatother
+
+\setcopyright{none}
+
 \begin{document}
 
-\conferenceinfo{SIGBOVIK~2019}{April 1, 2019, Pittsburgh, Pennsylvania, USA}
-\copyrightyear{2019}
-\title{Elo World,}
-\subtitle{a framework for benchmarking weak chess engines}
-\authorinfo{Dr. Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
-\toappear{
-Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
+% \copyrightyear{2019}
+
+\acmConference{SIGBOVIK~2019}{April 1, 2019, Pittsburgh, Pennsylvania, USA}
+
+\title[Elo World]{Elo World, a framework for \\ benchmarking weak chess \\ engines}
+% \authorinfo{Dr. Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
+\author{Dr. Tom Murphy VII Ph.D.}\email{tom7@tom7.org}
+\thanks{Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
 Appears in SIGBOVIK 19119 with the
@@
 
-\terms
-pawn, horse, bishop, castle, queen, king
-\keywords
-chess, bad
+\keywords{pawn, horse, bishop, castle, queen, king}
+% \keywords{chess, bad}
 
@@
 
-Fiddly bits aside, it is a solved problem to maintain a numeric rating
-of players for some game (for example chess, but also sports,
+Fiddly bits aside, it is a solved problem to maintain a numeric skill
+rating of players for some game (for example chess, but also sports,
 e-sports, probably also z-sports if that's a thing). Though it has
@@
 
-The gist of this system is to track a score for each player. The
-scores are defined such that the expected outcomes of games can
-be computed from the scores (for example, a player with a rating
-of 2400 should win 9/10 of her games against a player with a
-rating of 2000). If the true outcome (of e.g.~a tournament) doesn't
-match the expected outcome, then both player's scores are adjusted
-towards values that would have produced the expected result. Over
-time, scores thus become a more accurate reflection of player's skill,
-while also allowing for players to get stronger or weaker. This
-system is carefully described elsewhere, so we can just leave it
-at that.
+The gist of this system is to track a single score for each player.
+The scores are defined such that the expected outcomes of games can be
+computed from the scores (for example, a player with a rating of 2400
+should win 9/10 of her games against a player with a rating of 2000).
+If the true outcome (of e.g.~a tournament) doesn't match the expected
+outcome, then both player's scores are adjusted towards values that
+would have produced the expected result. Over time, scores thus become
+a more accurate reflection of players' skill, while also allowing for
+players to change skill level. This system is carefully described
+elsewhere, so we can just leave it at that.
 
 The players need not be human, and in fact this can facilitate running
-many games and thereby getting accurate ratings.
+many games and thereby getting arbitrarily accurate ratings.
 
@@
 whose ratings drop below 1000 are listed on the next list as
-'delisted'.''\cite{fideratings}.) It may also be the case that all
+'delisted'.''~\cite{fideratings}) It may also be the case that all
 weak players always lose to all strong players, making it unclear just
@@
   from previous rounds, or to get better at chess. The majority of
-  players are actually completely stateless, just a partial function
-  of type ${\tt position} \rightarrow {\tt move}$.
+  players are actually completely stateless, just a function of type
+  ${\tt position} \rightarrow {\tt move}$.
 \item black and white behave the same
-\item 
+\item Avoid game-tree search. Minimax, alpha--beta, etc. are the
+  correct way to play chess programmatically. They are well-studied
+  (i.e., boring) and effective, and so not well suited to our problem.
+  A less obvious issue is that they are endlessly parameterizable, for
+  example the search ply; this leaves us with a million things to
+  fiddle with. In any case, several traditional chess engines are
+  included for comparison.
 \end{itemize}
+
+% dispense with alpha-beta, etc.
+
+\subsection{The players}
+
+\newcommand\describeplayer[1]{\smallskip\noindent {\texttt{\textbf{#1}}}.\quad}
+\newcommand\player[1]{\texttt{#1}}
+\newcommand\deterministic{\colorbox{black}{\color{white} \textsf{det}}}
+\newcommand\traditional{\colorbox{black}{\color{white} \textsf{trad}}}
+% funny to use that v symbol made of leaves
+\newcommand\vegetarian{\colorbox{black}{\color{white} \textsf{v}}}
+\newcommand\canonical{\colorbox{black}{\color{white} \textsf{canon}}}
+\newcommand\stateful{\colorbox{black}{\color{white} \textsf{state}}}
+\newcommand\asymmetric{\colorbox{black}{\color{white} \textsf{asym}}}
+
+\describeplayer{random} We must begin with the most canonical of all
+strategies: Choosing a legal move uniformly at random. This is a
+lovely choice for Elo World, for several reasons: It is simple to
+describe. It is clearly canonical, in that anyone undertaking a
+similar project would come up with the same thing. It is capable of
+producing any sequence of moves, and thus completely spans the gamut
+from the worst possible player to the best. If we run the tournament
+long enough, it will eventually draw games even against a hypothetical
+perfect opponent, a sort of Boltzmann Brilliancy. Note that this
+strategy actually does keep state (the pseudorandom pool), despite the
+admonition above. We can see this as not really state but a simulation
+of an external source of ``true'' randomness. Most other players
+fall back on making random moves to break ties or when their primary
+strategy does not apply. \canonical
+
+\describeplayer{same\_color} When playing as white, put pieces on
+white squares. Vice versa for black. This is accomplished by counting
+the number of white pieces on white squares {\em after} each possible
+move, and then playing one of the moves that maximizes this number.
+Ties are broken randomly. Like many algorithms described this way, it
+tends to reach a plateau where the metric cannot be increased in a
+single move, and then plays randomly along this local maximum
+(Figure~\ref{fig:samecolor}). This particular strategy moves one
+knight at most once (because they always change color when moving)
+unless forced; on the other hand both bishops can be safely moved
+anywhere when the metric is maxed out.
+
+\begin{figure}[ht]
+\chessboard[setfen=1n1r3q/1Q2r1bp/1p1p1p1n/p1pPp1p1/P1PkR3/1P1P1P1N/6BP/1N1R3K b - - 85 73,showmover=false]
+\caption{\player{same\_color} (playing as white) checkmates
+  \player{same\_color} (playing as black) on move 73. Note that since
+  white opened with Nh3, the h2 pawn is stuck on a black square.
+  Moving the knight out of the way would require that move to be
+  forced. } \label{fig:samecolor}
+\end{figure}
+
+\describeplayer{opposite\_color} Same idea, opposite parity.
+
+% XXX more clear/efficient description here?
+\describeplayer{pacifist} Avoid moves that mate the opponent, and
+failing that, avoid moves that check, and failing that, avoid moves
+that capture pieces, and failing that, capture lower value pieces.
+Break ties randomly. This is one of the worst strategies, drawing
+against players that are not ambitious about normal chess pursuits,
+and easily losing to simple strategies. On the other hand, it does
+rarely get forced into mating its opponent by aggressive but weak
+players. \vegetarian
+
+\describeplayer{first\_move} Make the lexicographically first legal
+move. The moves are ordered as $\langle$ \verb+src_row+,
+\verb+src_col+, \verb+dst_row+, \verb+dst_col+,
+\verb+promote_to+ $\rangle$ for white (rank $1$ is row $0$) and the rows
+are reversed for black to make the strategy symmetric. Tends to
+produce draws (by repetition), because knights and rooks can often
+move back and forth on the first few files. \deterministic
+
+\describeplayer{alphabetical} Make the alphabetically first move,
+using standard PGN short notation. White and black both try to move
+towards A1. \asymmetric \deterministic
+
+\describeplayer{worstfish}
+\describeplayer{generous}
+\describeplayer{safe}
+\describeplayer{popular}
+\describeplayer{dangerous}
+\describeplayer{rare}
+\describeplayer{no\_i\_insist}
+\describeplayer{sym\_mirror\_x}
+\describeplayer{huddle}
+\describeplayer{survivalist}
+\describeplayer{sym\_180}
+\describeplayer{sym\_mirror\_y}
+\describeplayer{suicide\_king}
+\describeplayer{reverse\_starting}
+\describeplayer{fatalist}
+\describeplayer{blind\_yolo}
+\describeplayer{cccp}
+\describeplayer{equalizer}
+\describeplayer{blind\_kings}
+\describeplayer{swarm}
+\describeplayer{min\_oppt\_moves}
+\describeplayer{chessmaster.nes\_lv1}
+\describeplayer{stockfish1m\_r32768}
+\describeplayer{chessmaster.nes\_lv2}
+\describeplayer{stockfish1m\_r16384}
+\describeplayer{stockfish0}
+\describeplayer{stockfish5}
+\describeplayer{stockfish1m\_r8192}
+\describeplayer{topple10k}
+\describeplayer{stockfish15}
+\describeplayer{stockfish10}
+\describeplayer{stockfish1m\_r4096}
+\describeplayer{stockfish20}
+\describeplayer{topple1m}
+\describeplayer{stockfish1m\_r2048}
+\describeplayer{stockfish1m\_r1024}
+\describeplayer{stockfish1m\_r512}
+\describeplayer{stockfish1m\_r256}
+\describeplayer{stockfish1m\_r64}
+\describeplayer{stockfish1m}
+\describeplayer{stockfish1m\_r128}
 
@@
 
-\bibliography{chess}{}
-\bibliographystyle{plain}
+% \balance
+% \bibliographystyle{plain}
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{chess}
 

### 2019-03-13 chess/player.cc

@@ -784,7 +784,6 @@ struct SinglePlayerPlayer : public StatelessPlayer {
 };
 
 // TODO:
-//  - Stockfish, with and without opening book / endgame tablebases
 //  - Maximize net control of squares
 //  - Try to move your pieces to certain squares:
 //       - try to get all pieces close to the center
@@ -795,7 +794,13 @@ struct SinglePlayerPlayer : public StatelessPlayer {
 //  - prefer moves that can be mirrored by the opponent
 //
 //  - select eligible squares with langton's ant, game of life
-
+//
+//  mimic the move your opponent just played (though I guess flipped),
+//  if possible. rather similar to sym_mirror_y
+//
+//  checkmate: make a move that minimizes the number of moves I would
+//  have to make in order to checkmate the opponent, assuming I am
+//  the only one making moves.
 
 }  // namespace
 

### 2019-03-13 trunk/chess/papers/weak.tex
Wrote about some more players in the hotel

@@
 
+% skip acm copyright shits
+\makeatletter
+\def\@copyrightspace{\relax}
+\makeatother
+
+\setcopyright{none}
+
 \begin{document}
 
-\conferenceinfo{SIGBOVIK~2019}{April 1, 2019, Pittsburgh, Pennsylvania, USA}
-\copyrightyear{2019}
-\title{Elo World,}
-\subtitle{a framework for benchmarking weak chess engines}
-\authorinfo{Dr. Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
-\toappear{
-Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
+% \copyrightyear{2019}
+
+\acmConference{SIGBOVIK~2019}{April 1, 2019, Pittsburgh, Pennsylvania, USA}
+
+\title[Elo World]{Elo World, a framework for \\ benchmarking weak chess \\ engines}
+% \authorinfo{Dr.~Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
+\author{Dr.~Tom Murphy VII Ph.D.}\email{tom7@tom7.org}
+\thanks{Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
 Appears in SIGBOVIK 19119 with the
 %
-insufficient material 
+title inflation
 %
@@
 
-Fiddly bits aside, it is a solved problem to maintain a numeric rating
-of players for some game (for example chess, but also sports,
+\section{Introduction}
+
+Fiddly bits aside, it is a solved problem to maintain a numeric skill
+rating of players for some game (for example chess, but also sports,
 e-sports, probably also z-sports if that's a thing). Though it has
@@
 compare them), the Elo Rating System~\cite{elo1978rating} is a simple
-and effective way to do it.
+and effective way to do it. This paper is concerned with Elo in chess,
+its original purpose.
 
-The gist of this system is to track a score for each player. The
-scores are defined such that the expected outcomes of games can
-be computed from the scores (for example, a player with a rating
-of 2400 should win 9/10 of her games against a player with a
-rating of 2000). If the true outcome doesn't match the
+The gist of this system is to track a single score for each player.
+The scores are defined such that the expected outcomes of games can be
+computed from the scores (for example, a player with a rating of 2400
+should win 9/10 of her games against a player with a rating of 2000).
+If the true outcome (of e.g.~a tournament) doesn't match the expected
+outcome, then both player's scores are adjusted towards values that
+would have produced the expected result. Over time, scores thus become
+a more accurate reflection of players' skill, while also allowing for
+players to change skill level. This system is carefully described
+elsewhere, so we can just leave it at that.
 
-.. so we can just leave it at that
+The players need not be human, and in fact this can facilitate running
+many games and thereby getting arbitrarily accurate ratings.
+
+The problem this paper is concerned with is that basically all chess
+tournaments (whether with humans or computers or both) are between
+players who know how to play chess, are interested in winning their
+games, and have some reasonable level of skill. This makes it hard to
+give a rating to weak players: They just lose every single game and so
+tend towards a rating of $-\infty$\footnote{Some organizations don't
+  even let ratings fall beneath a certain level, for example, the
+  lowest possible USCF rating is 100.} Even if other comparatively
+weak players existed to participate in the tournament and occasionally
+lose to the player under study, it may still be difficult to
+understand how this cohort performs in any absolute sense. (In
+contrast we have ``the highest ever human rating was 2882,'' and
+``there are 5,323 players with ratings in 2200--2299'' and ``Players
+whose ratings drop below 1000 are listed on the next list as
+'delisted'.''~\cite{fideratings}) It may also be the case that all
+weak players always lose to all strong players, making it unclear just
+how wide the performance gap between the two sets is. The goal of
+this paper is to expand the dynamic range of chess ratings to span
+all the way from extremely weak players to extremely strong ones, while
+providing canonical reference points along the way.
+
+\section{Elo world}
+
+Elo World is a tournament with dozens of computer players. The
+computer players include some traditional chess engines, but also many
+algorithms chosen for their simplicity, as well as some designed to
+be competitively bad.
+
+The fun part is the various algorithms, so let's get into that. First,
+some ground rules and guidelines:
+\begin{itemize}
+\item No resigning (forfeiting) or claiming a draw. The player will
+  only be asked to make a move when there exists one, and it must
+  choose a move in finite time. (In practice, most of the players are
+  extremely fast, with the slowest ones using about one second of
+  CPU time per move.)
+\item The player can retain state for the game, and executes moves
+  sequentially (for either the white or black pieces), but cannot have
+  state meaningfully span games. For example, it is not permitted to
+  do man-in-the-middle attacks~\cite{blind} or learn opponent's moves
+  from previous rounds, or to get better at chess. The majority of
+  players are actually completely stateless, just a function of type
+  ${\tt position} \rightarrow {\tt move}$.
+\item black and white behave the same
+\item Avoid game-tree search. Minimax, alpha--beta, etc. are the
+  correct way to play chess programmatically. They are well-studied
+  (i.e., boring) and effective, and so not well suited to our problem.
+  A less obvious issue is that they are endlessly parameterizable, for
+  example the search ply; this leaves us with a million things to
+  fiddle with. In any case, several traditional chess engines are
+  included for comparison.
+\end{itemize}
+
+% dispense with alpha-beta, etc.
+
+\subsection{Simple players}
+
+\newcommand\describeplayer[1]{\smallskip\noindent {\texttt{\textbf{#1}}}.\quad}
+\newcommand\player[1]{\texttt{#1}}
+\newcommand\deterministic{\colorbox{black}{\color{white} \textsf{det}}}
+\newcommand\traditional{\colorbox{black}{\color{white} \textsf{trad}}}
+% funny to use that v symbol made of leaves
+\newcommand\vegetarian{\colorbox{black}{\color{white} \textsf{v}}}
+\newcommand\canonical{\colorbox{black}{\color{white} \textsf{canon}}}
+\newcommand\stateful{\colorbox{black}{\color{white} \textsf{state}}}
+\newcommand\asymmetric{\colorbox{black}{\color{white} \textsf{asym}}}
+
+\describeplayer{random} We must begin with the most canonical of all
+strategies: Choosing a legal move uniformly at random. This is a
+lovely choice for Elo World, for several reasons: It is simple to
+describe. It is clearly canonical, in that anyone undertaking a
+similar project would come up with the same thing. It is capable of
+producing any sequence of moves, and thus completely spans the gamut
+from the worst possible player to the best. If we run the tournament
+long enough, it will eventually draw games even against a hypothetical
+perfect opponent, a sort of Boltzmann Brilliancy. Note that this
+strategy actually does keep state (the pseudorandom pool), despite the
+admonition above. We can see this as not really state but a simulation
+of an external source of ``true'' randomness. Most other players
+fall back on making random moves to break ties or when their primary
+strategy does not apply. \canonical
+
+\describeplayer{same\_color} When playing as white, put pieces on
+white squares. Vice versa for black. This is accomplished by counting
+the number of white pieces on white squares {\em after} each possible
+move, and then playing one of the moves that maximizes this number.
+Ties are broken randomly. Like many algorithms described this way, it
+tends to reach a plateau where the metric cannot be increased in a
+single move, and then plays randomly along this local maximum
+(Figure~\ref{fig:samecolor}). This particular strategy moves one
+knight at most once (because they always change color when moving)
+unless forced; on the other hand both bishops can be safely moved
+anywhere when the metric is maxed out.
+
+\begin{figure}[ht]
+\chessboard[setfen=1n1r3q/1Q2r1bp/1p1p1p1n/p1pPp1p1/P1PkR3/1P1P1P1N/6BP/1N1R3K b - - 85 73,showmover=false]
+\caption{\player{same\_color} (playing as white) checkmates
+  \player{same\_color} (playing as black) on move 73. Note that since
+  white opened with Nh3, the h2 pawn is stuck on a black square.
+  Moving the knight out of the way would require that move to be
+  forced. } \label{fig:samecolor}
+\end{figure}
+
+\describeplayer{opposite\_color} Same idea, opposite parity.
+
+% XXX more clear/efficient description here?
+\describeplayer{pacifist} Avoid moves that mate the opponent, and
+failing that, avoid moves that check, and failing that, avoid moves
+that capture pieces, and failing that, capture lower value pieces.
+Break ties randomly. This is one of the worst strategies, drawing
+against players that are not ambitious about normal chess pursuits,
+and easily losing to simple strategies. On the other hand, it does
+rarely get forced into mating its opponent by aggressive but weak
+players. \vegetarian
+
+\describeplayer{first\_move} Make the lexicographically first legal
+move. The moves are ordered as $\langle\!\!$ \verb+src_row+,
+\verb+src_col+, \verb+dst_row+, \verb+dst_col+,
+\verb+promote_to+ $\!\!\rangle$ for white (rank $1$ is row $0$) and the rows
+are reversed for black to make the strategy symmetric. Tends to
+produce draws (by repetition), because knights and rooks can often
+move back and forth on the first few files. \deterministic
+
+\describeplayer{alphabetical} Make the alphabetically first move,
+using standard PGN short notation. White and black both try to move
+towards A1. \asymmetric \deterministic
+
+\describeplayer{huddle} As white, make moves that minimize the total
+distance between white pieces and the white king. Distance is the
+Chebyshev distance, which is the number of moves a king takes to move
+between squares. This forms a defensive wall around the king
+(Figure~\ref{fig:huddle}).
+
+\begin{figure}[ht]
+\chessboard[setfen=2r4r/5n2/1p4p1/nP5p/1Ppk3P/2NPRNP1/1bRPPP2/2BQKB2 b - - 80 158,showmover=false]
+\caption{\player{huddle} (white) checkmates \player{pacifist} on move
+  158 with substantial help. Note that white's distal pawns have
+  advanced; these are actually the same distance from the King as they
+  are in their home row, since the distance metric includes diagonal
+  moves. } \label{fig:huddle}
+\end{figure}
+
+\describeplayer{swarm} Like \player{huddle}, but instead move pieces
+such that they are close to opponent's king. This is essentially an
+all-out attack with no planning, and manages to be one of the
+better-performing ``simple'' strategies. From the bloodbaths it
+creates, it even manages a few victories against strong opponents
+(Figure~\ref{fig:swarm}).
+
+% XXX illustrate the moves here
+\begin{figure}[ht]
+\chessboard[setfen=rnbqkb1r/pp1ppp1p/7p/1Bp5/3PnP2/8/PPP3PP/RN1QK1NR b KQkq - 1 5,showmover=false]
+\caption{\player{swarm} (white) vs.~\player{stockfish1m\_r4096}
+  with black to move after 1.~d4 Nf6 2.~Bh6 gxh6 3.~f4 c5 4.~e4 Nxe4
+  5.~Bb5. Black blunders 5\ldots f6??, which must have been a random
+  move as \player{swarm} immediately wins with 6.~Qh5++.} \label{fig:swarm}
+\end{figure}
+
+\describeplayer{generous} Move so as to maximize the number of
+opponent moves that capture our pieces, weighting by the value of the
+offered piece ($\pawn = 1$, $\bishop = \knight = 3$, $\rook = 5$,
+$\queen = 9$). A piece that can be catpured multiple ways is counted
+multiple times.
+
+\describeplayer{no\_i\_insist} Like \player{generous}, but be overwhelmingly
+polite by trying to {\em force} the opponent to accept the gift of material.
+There are three tiers: Avoid at all costs mating the opponent (and moreso
+checkmating). Stalemate is polite, but it is more canonical for two
+polite players to form a draw by repetition, from continually offering up
+material to one another and declining it. Next, avoid situations where
+the opponent can refuse our gift; among these, prefer the move where the
+opponent must capture the highest value piece. Finally, prefer moves
+where the expected value of the offered material (i.e.~against random play)
+is highest. (This means that if there are three moves, and one captures a
+rook but the others capture nothing, the value is $5/3$.) This strategy
+almost never wins, but is not among the worst players, since it often forces
+a draw by exchanging all its pieces.
+
+\describeplayer{reverse\_starting} This player thinks that the board is
+upside-down, and as white, tries to put its pieces where black pieces
+start. This often causes some conflict since the black pieces are already
+there, but can also end peacefully (Figure~\ref{fig:reversestarting}).
+
+\begin{figure}[ht]
+\chessboard[setfen=1NB1K2R/8/8/8/8/8/6p1/rnbqkbn1 b - - 19 64,showmover=false]
+\caption{\player{reverse\_starting} draws against \player{reverse\_starting}
+by repetition, both players having happily moved their surviving pieces into the
+correct spots.} \label{fig:reversestarting}
+\end{figure}
+
+\subsection{Fate-based players}
+
+\subsection{Engine-based players}
+
+\subsection{Blind players}
+
+(I guess we can also segue by discussing handicaps)
+(end with this so that we can segue into dilution methods)
+
+\subsection{Diultion methods}
+% Stockfish1m_32768 loses to reverse_starting in 3 moves!
+1. e4 g5 2. Ba6 f5 3. Qh5
+
+\describeplayer{worstfish}
+\describeplayer{safe}
+\describeplayer{popular}
+\describeplayer{dangerous}
+\describeplayer{rare}
+\describeplayer{survivalist}
+\describeplayer{sym\_mirror\_x}
+\describeplayer{sym\_180}
+\describeplayer{sym\_mirror\_y}
+\describeplayer{suicide\_king}
+\describeplayer{fatalist}
+\describeplayer{blind\_yolo}
+\describeplayer{cccp}
+\describeplayer{equalizer}
+\describeplayer{blind\_kings}
+\describeplayer{min\_oppt\_moves}
+\describeplayer{chessmaster.nes\_lv1}
+\describeplayer{stockfish1m\_r32768}
+\describeplayer{chessmaster.nes\_lv2}
+\describeplayer{stockfish1m\_r16384}
+\describeplayer{stockfish0}
+\describeplayer{stockfish5}
+\describeplayer{stockfish1m\_r8192}
+\describeplayer{topple10k}
+\describeplayer{stockfish15}
+\describeplayer{stockfish10}
+\describeplayer{stockfish1m\_r4096}
+\describeplayer{stockfish20}
+\describeplayer{topple1m}
+\describeplayer{stockfish1m\_r2048}
+\describeplayer{stockfish1m\_r1024}
+\describeplayer{stockfish1m\_r512}
+\describeplayer{stockfish1m\_r256}
+\describeplayer{stockfish1m\_r64}
+\describeplayer{stockfish1m}
+\describeplayer{stockfish1m\_r128}
+
+Ideally, such a system would be robust against adding new players,
+but this is probably not the case.
 

### 2019-03-13 chess/papers/weak.tex
Writing on the shuttle and waiting for the gate, describing some more players

@@
 \title[Elo World]{Elo World, a framework for \\ benchmarking weak chess \\ engines}
-% \authorinfo{Dr. Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
-\author{Dr. Tom Murphy VII Ph.D.}\email{tom7@tom7.org}
+% \authorinfo{Dr.~Tom Murphy VII Ph.D.}{tom7.org Foundation}{tom7@tom7.org}
+\author{Dr.~Tom Murphy VII Ph.D.}\email{tom7@tom7.org}
 \thanks{Copyright \copyright\ 2019 the Regents of the Wikiplia Foundation.
@@
 
-\subsection{The players}
+\subsection{Simple players}
 
@@
 \describeplayer{first\_move} Make the lexicographically first legal
-move. The moves are ordered as $\langle$ \verb+src_row+,
+move. The moves are ordered as $\langle\!\!$ \verb+src_row+,
 \verb+src_col+, \verb+dst_row+, \verb+dst_col+,
-\verb+promote_to+ $\rangle$ for white (rank $1$ is row $0$) and the rows
+\verb+promote_to+ $\!\!\rangle$ for white (rank $1$ is row $0$) and the rows
 are reversed for black to make the strategy symmetric. Tends to
@@
 
-\describeplayer{worstfish}
-\describeplayer{generous}
-\describeplayer{safe}
-\describeplayer{popular}
-\describeplayer{dangerous}
-\describeplayer{rare}
-\describeplayer{no\_i\_insist}
-\describeplayer{sym\_mirror\_x}
-\describeplayer{huddle}
-\describeplayer{survivalist}
-\describeplayer{sym\_180}
-\describeplayer{sym\_mirror\_y}
-\describeplayer{suicide\_king}
-\describeplayer{reverse\_starting}
-\describeplayer{fatalist}
-\describeplayer{blind\_yolo}
-\describeplayer{cccp}
-\describeplayer{equalizer}
-\describeplayer{blind\_kings}
-\describeplayer{swarm}
-\describeplayer{min\_oppt\_moves}
-\describeplayer{chessmaster.nes\_lv1}
-\describeplayer{stockfish1m\_r32768}
-\describeplayer{chessmaster.nes\_lv2}
-\describeplayer{stockfish1m\_r16384}
+\describeplayer{huddle} As white, make moves that minimize the total
+distance between white pieces and the white king. Distance is the
+Chebyshev distance, which is the number of moves a king takes to move
+between squares. This forms a defensive wall around the king
+(Figure~\ref{fig:huddle}).
+
+\begin{figure}[ht]
+\chessboard[setfen=2r4r/5n2/1p4p1/nP5p/1Ppk3P/2NPRNP1/1bRPPP2/2BQKB2 b - - 80 158,showmover=false]
+\caption{\player{huddle} (white) checkmates \player{pacifist} on move
+  158 with substantial help. Note that white's distal pawns have
+  advanced; these are actually the same distance from the King as they
+  are in their home row, since the distance metric includes diagonal
+  moves. } \label{fig:huddle}
+\end{figure}
+
+\describeplayer{swarm} Like \player{huddle}, but instead move pieces
+such that they are close to opponent's king. This is essentially an
+all-out attack with no planning, and manages to be one of the
+better-performing ``simple'' strategies. From the bloodbaths it
+creates, it even manages a few victories against strong opponents
+(Figure~\ref{fig:swarm}).
+
+% XXX illustrate the moves here
+\begin{figure}[ht]
+\chessboard[setfen=rnbqkb1r/pp1ppp1p/7p/1Bp5/3PnP2/8/PPP3PP/RN1QK1NR b KQkq - 1 5,showmover=false]
+\caption{\player{swarm} (white) vs.~\player{stockfish1m\_r4096}
+  with black to move after 1.~d4 Nf6 2.~Bh6 gxh6 3.~f4 c5 4.~e4 Nxe4
+  5.~Bb5. Black blunders 5\ldots f6??, which must have been a random
+  move as \player{swarm} immediately wins with 6.~Qh5++.} \label{fig:swarm}
+\end{figure}
+
+\describeplayer{generous} Move so as to maximize the number of
+opponent moves that capture our pieces, weighting by the value of the
+offered piece ($\pawn = 1$, $\bishop = \knight = 3$, $\rook = 5$,
+$\queen = 9$). A piece that can be catpured multiple ways is counted
+multiple times.
+
+\describeplayer{no\_i\_insist} Like \player{generous}, but be overwhelmingly
+polite by trying to {\em force} the opponent to accept the gift of material.
+There are three tiers: Avoid at all costs mating the opponent (and moreso
+checkmating). Stalemate is polite, but it is more canonical for two
+polite players to form a draw by repetition, from continually offering up
+material to one another and declining it. Next, avoid situations where
+the opponent can refuse our gift; among these, prefer the move where the
+opponent must capture the highest value piece. Finally, prefer moves
+where the expected value of the offered material (i.e.~against random play)
+is highest. (This means that if there are three moves, and one captures a
+rook but the others capture nothing, the value is $5/3$.) This strategy
+almost never wins, but is not among the worst players, since it often forces
+a draw by exchanging all its pieces.
+
+\describeplayer{reverse\_starting} This player thinks that the board is
+upside-down, and as white, tries to put its pieces where black pieces
+start. This often causes some conflict since the black pieces are already
+there, but can also end peacefully (Figure~\ref{fig:reversestarting}).
+
+\begin{figure}[ht]
+\chessboard[setfen=1NB1K2R/8/8/8/8/8/6p1/rnbqkbn1 b - - 19 64,showmover=false]
+\caption{\player{reverse\_starting} draws against \player{reverse\_starting}
+by repetition, both players having happily moved their surviving pieces into the
+correct spots.} \label{fig:reversestarting}
+\end{figure}
+
+% TODO: Cyrillic tag or toast at the end
+\describeplayer{cccp} Prioritize moves that Checkmate, Check, Capture
+or Push, in that order. Push means to move pieces as deep as possible
+into enemy territory (without any regard for their safety). Ties are
+broken deterministically by the source/destination squares, so this
+one is technically asymmetric. \deterministic \asymmetric
+
+\describeplayer{suicide\_king} Take a random move that minimizes the
+distance between the two kings. Putting one's king out in the open is
+very unprincipled (Figure~\ref{fig:cccpsuicide}), but it does produce
+enough pressure to win against unambitious opponents.
+
+\begin{figure}[ht]
+  \chessboard[setfen=r1bq1bnr/pppp1p1p/2PQ4/8/2k6/8/PPP1PPPP/RN2KBNR b KQ - 0 7,showmover=false]
+\caption{\player{suicide\_king} (black) dramatically failing against
+  \player{cccp}'s slightly more principled play, after 1.~d4
+  g5 2.~Bxg5 Nc6 3.~Bxe7 Kxe7 4.~d5 Kd6 5.~dxc6+ Kc5 6.~Qd6+ Kc4.
+  White delivers a discovered mate with 7.~e4++.} \label{fig:cccpsuicide}
+\end{figure}
+
+% XXX include examples of symmetric players, as they are aesthetic.
+\describeplayer{sym\_mirror\_y} As white, try to maximize symmetry
+when flipping vertically. Zero penalty for opposing e.g.~a \King with
+a \king, but a small penalty when the pieces are not the same type,
+and a larger penalty when they are not opposite coors. The starting
+position is already symmetric this way, so this usually has the effect
+of copying the opponent's moves when possible. As ``copy your
+opponent's moves'' is a commonly known (but underspecified) strategy
+discovered by many children, this player is close to being canonical.
+However, it admits a bit too much arbitrary choice in the penalties
+assigned.
+
+\describeplayer{sym\_mirror\_x} As \player{sym\_mirror\_y}, but
+maximize symmetry when flipping horizontally. This does not make much
+chess sense, but can produce aesthetic arrangements
+(Figure~\ref{fig:XXX}).
+
+% XXX show self play perhaps?
+\describeplayer{sym\_180} As \player{sym\_mirror\_y}, but maximize
+symmetry under $180\deg$ rotation of the board. An emergent priority is
+to ``castle'' with the king and queen to ``fix'' them.
+
+\describeplayer{min\_oppt\_moves} Take a move that minimizes the
+number of resulting legal moves for the opponent, breaking ties
+randomly. This is a beautifully simple approach that generalizes many
+chess principles: Occupying space reduces the destination squares
+available to the opponent; capturing their pieces reduces the number
+of their pieces that they can move; pinning pieces or checking the
+king further reduces the legal moves; and mating the opponent is the
+best possible move.\footnote{However note that it does not distinguish
+  checkmate and stalemate, despite these having very different
+  results.} Among the players in the paper, this one is Pareto
+efficient in terms of its simplicity and effectiveness. \canonical
+
+% XXX cite blog comment?
+\describeplayer{equalizer} Prefer to move a piece that has been
+moved the fewest times, and then prefer moving to a square that
+has been visited the fewest times. Castling counts as moving
+both the king and rook, and visiting both destination squares.
+This is the first strategy described that keeps meaningful state.
+\canonical \stateful
+
+\subsection{Fate-based players}
+
+If we keep state, then we can track the location of each piece as it
+moves around the board (allowing us to distinguish the two rooks, or
+follow a pawn as it promotes). We can then use statistics on the
+average fates of each piece over hundreds of millions of games to
+guide the moves. These statistics give us, for each piece (e.g.~the c2
+pawn) and square, how likely it is to end the game on that square, and
+how likely it is to be alive or dead there when the game
+ends~\cite{survival}.
+
+\describeplayer{safe} This strategy moves pieces towards squares where
+they are likely to end the game alive. For this strategy and several
+others, simply moving to maximize this score (e.g.~its sum or product
+over all pieces) is very boring, since the score is almost always
+maximized in the starting position. So this strategy actually makes
+each move randomly, weighted by the total score of the resulting
+positions. The scores are normalized (with the lowest-scoring move
+receiving weight 0.0 and the highest 1.0) and then sampled according
+to these weights. Without normalization, the play is almost identical
+to uniformly random, since the weights of the resulting positions tend
+to be very similar (dominated by the many pieces that don't move). But
+it's pretty arbitrary.
+
+\describeplayer{popular} Like \player{safe}, but the score for a
+piece/square pair is the probability that the piece ends on that
+square, whether it lives or dies. This player likes to follow the
+crowd!
+
+\describeplayer{dangerous} The dual of \player{safe}; the score is
+the probability that the piece dies on that square. Note that a
+king is said to die if it is checkmated. This player likes to
+live life on the edge!
+
+\describeplayer{rare} The dual of \player{dangerous}; the score
+is one minus the probability of ending the game on that square. This
+player has a thirst for adventure!
+
+\describeplayer{survivalist} Like the above, but the score is the
+ratio of the survival and death probabilities on the square. In the
+data set, every piece ends alive or dead in every square (except
+for the bishops, which can only legally occupy squares of their
+color) at least 1000 times, so each ratio is defined. Here, the
+sums of ratios have plenty of variability, and the highest
+ratios are not so often on the starting squares. So with this
+strategy, we simply do a weighted sample from the moves according
+to their (non-normalized) scores.
+
+\describeplayer{fatalist} The dual of \player{fatalist}; the score is
+the ratio of the death and survival probability on the square. This
+player knows that even if you win, you just have to keep playing over
+and over again, so you might as well get it over with!
+
+\subsection{Engine-based players}
+
+...
+
 \describeplayer{stockfish0}
 \describeplayer{stockfish5}
-\describeplayer{stockfish1m\_r8192}
-\describeplayer{topple10k}
 \describeplayer{stockfish15}
 \describeplayer{stockfish10}
+\describeplayer{stockfish20}
+
+\describeplayer{worstfish} (compare ``losing chess'')
+uses stockfish20; we could use 1m (but slow), but it
+probably doesn't benefit from this depth of analysis,
+since the engine assumes the opponent is playing well.
+To do even worse, we would want to play moves that
+trap the opponent into beating us, as well as ones that 
+avoid repetition draws
+
+\describeplayer{chessmaster.nes\_lv1}
+\describeplayer{chessmaster.nes\_lv2} \deterministic
+\describeplayer{stockfish1m\_r32768}
+\describeplayer{stockfish1m\_r16384}
+\describeplayer{stockfish1m\_r8192}
+\describeplayer{topple1m}
+\describeplayer{topple10k}
+\describeplayer{stockfish1m}
+
+\subsection{Blind players}
+
+\describeplayer{blind\_yolo}
+\describeplayer{blind\_kings}
+% \describeplayer{blind\_spycheck} (described in paper, not yet run)
+
+(I guess we can also segue by discussing handicaps)
+(end with this so that we can segue into dilution methods)
+
+\subsection{Diultion methods}
+% Stockfish1m_32768 loses to reverse_starting in 3 moves!
+1. e4 g5 2. Ba6 f5 3. Qh5
+
 \describeplayer{stockfish1m\_r4096}
-\describeplayer{stockfish20}
-\describeplayer{topple1m}
 \describeplayer{stockfish1m\_r2048}
@@
 \describeplayer{stockfish1m\_r64}
-\describeplayer{stockfish1m}
 \describeplayer{stockfish1m\_r128}
@@
 Interpolation, like Scoville scale.
-
-% Dramatic failure of the "suicide king" strategy against
-% CCCP.
-1. d4 g5 2. Bxg5 Nc6 3. Bxe7 Kxe7 4. d5 Kd6
-5. dxc6+ Kc5 6. Qd6+ Kc4 7. e4++
 
@@
 
+\section{Conclusion}
+
+The author would like to thank the anonymous referees
+of the Special Interest Group on
+B
+O
+V
+In
+Chess journal for their careful reviews.
+
 \nocite{elo1978rating}

### 2019-03-14 trunk/chess/papers/weak.tex
Writing on the plane, describing some more players

@@
 from the worst possible player to the best. If we run the tournament
-long enough, it will eventually draw games even against a hypothetical
-perfect opponent, a sort of Boltzmann Brilliancy. Note that this
-strategy actually does keep state (the pseudorandom pool), despite the
-admonition above. We can see this as not really state but a simulation
-of an external source of ``true'' randomness. Most other players
-fall back on making random moves to break ties or when their primary
-strategy does not apply. \canonical
+long enough, it will eventually at least draw games even against a
+hypothetical perfect opponent, a sort of Boltzmann Brilliancy. Note
+that this strategy actually does keep state (the pseudorandom pool),
+despite the admonition above. We can see this as not really state but
+a simulation of an external source of ``true'' randomness. Most other
+players fall back on making random moves to break ties or when their
+primary strategy does not apply. \canonical
 
@@
 \describeplayer{first\_move} Make the lexicographically first legal
-move. The moves are ordered as $\langle\!\!$ \verb+src_row+,
+move. The moves are ordered as $\langle\!$~\verb+src_row+,
 \verb+src_col+, \verb+dst_row+, \verb+dst_col+,
-\verb+promote_to+ $\!\!\rangle$ for white (rank $1$ is row $0$) and the rows
+\verb+promote_to+~$\!\rangle$ for white (rank $1$ is row $0$) and the rows
 are reversed for black to make the strategy symmetric. Tends to
@@
 
+% TODO: Cyrillic tag or toast at the end
+\describeplayer{cccp} Prioritize moves that Checkmate, Check, Capture
+or Push, in that order. Push means to move pieces as deep as possible
+into enemy territory (without any regard for their safety). Ties are
+broken deterministically by the source/destination squares, so this
+one is technically asymmetric. \deterministic \asymmetric
+
+\describeplayer{suicide\_king} Take a random move that minimizes the
+distance between the two kings. Putting one's king out in the open is
+very unprincipled (Figure~\ref{fig:cccpsuicide}), but it does produce
+enough pressure to win against unambitious opponents.
+
+\begin{figure}[ht]
+  \chessboard[setfen=r1bq1bnr/pppp1p1p/2PQ4/8/2k6/8/PPP1PPPP/RN2KBNR b KQ - 0 7,showmover=false]
+\caption{\player{suicide\_king} (black) dramatically failing against
+  \player{cccp}'s slightly more principled play, after 1.~d4
+  g5 2.~Bxg5 Nc6 3.~Bxe7 Kxe7 4.~d5 Kd6 5.~dxc6+ Kc5 6.~Qd6+ Kc4.
+  White delivers a discovered mate with 7.~e4++.} \label{fig:cccpsuicide}
+\end{figure}
+
+% XXX include examples of symmetric players, as they are aesthetic.
+\describeplayer{sym\_mirror\_y} As white, try to maximize symmetry
+when flipping vertically. Zero penalty for opposing e.g.~a \King\, with
+a \king, but a small penalty when the pieces are not the same type,
+and a larger penalty when they are not opposite colors. The starting
+position is already symmetric this way, so this usually has the effect
+of copying the opponent's moves when possible. As ``copy your
+opponent's moves'' is a common (but underspecified) strategy
+discovered by many children, this player is close to being canonical.
+However, it admits a bit too much arbitrary choice in the penalties
+assigned.
+
+\describeplayer{sym\_mirror\_x} As \player{sym\_mirror\_y}, but
+maximize symmetry when flipping horizontally. This does not make much
+chess sense, but can produce aesthetic arrangements
+(Figure~\ref{fig:XXX}).
+
+% XXX show self play perhaps?
+\describeplayer{sym\_180} As \player{sym\_mirror\_y}, but maximize
+symmetry under 180 degree
+% XXX latex symbol for ``degrees''
+rotation of the board. An emergent priority is
+to ``castle'' with the king and queen to ``fix'' them.
+
+\describeplayer{min\_oppt\_moves} Take a move that minimizes the
+number of resulting legal moves for the opponent, breaking ties
+randomly. This is a beautifully simple approach that generalizes many
+chess principles: Occupying space reduces the destination squares
+available to the opponent; capturing their pieces reduces the number
+of their pieces that they can move; pinning pieces or checking the
+king further reduces the legal moves; and mating the opponent is the
+best possible move.\footnote{However note that it does not distinguish
+  checkmate and stalemate, despite these having very different
+  results.} Among the players in the paper, this one is Pareto
+efficient in terms of its simplicity and effectiveness. \canonical
+
+% XXX cite blog comment?
+\describeplayer{equalizer} Prefer to move a piece that has been
+moved the fewest times, and then prefer moving to a square that
+has been visited the fewest times. Castling counts as moving
+both the king and rook, and visiting both destination squares.
+This is the first strategy described that keeps meaningful state.
+\canonical \stateful
+
 \subsection{Fate-based players}
+
+If we keep state, then we can track the location of each piece as it
+moves around the board (allowing us to distinguish the two rooks, or
+follow a pawn as it promotes). We can then use statistics on the
+average fates of each piece over hundreds of millions of games to
+guide the moves. These statistics give us, for each piece (e.g.~the c2
+pawn) and square, how likely it is to end the game on that square, and
+how likely it is to be alive or dead there when the game
+ends~\cite{survival}.
+
+\describeplayer{safe} This strategy moves pieces towards squares where
+they are likely to end the game alive. For this strategy and several
+others, simply moving to maximize this score (e.g.~its sum or product
+over all pieces) is very boring, since the score is almost always
+maximized in the starting position. So this strategy actually makes
+each move randomly, weighted by the total score of the resulting
+positions. The scores are normalized (with the lowest-scoring move
+receiving weight 0.0 and the highest 1.0) and then sampled according
+to these weights. Without normalization, the play is almost identical
+to uniformly random, since the weights of the resulting positions tend
+to be very similar (dominated by the many pieces that don't move). But
+it's pretty arbitrary.
+
+\describeplayer{popular} Like \player{safe}, but the score for a
+piece/square pair is the probability that the piece ends on that
+square, whether it lives or dies. This player likes to follow the
+crowd!
+
+\describeplayer{dangerous} The dual of \player{safe}; the score is the
+probability that the piece dies on that square. Note that a king is
+said to die if it is checkmated or his side resigns. This player likes
+to live life on the edge!
+
+\describeplayer{rare} The dual of \player{dangerous}; the score
+is one minus the probability of ending the game on that square. This
+player has a thirst for adventure!
+
+\describeplayer{survivalist} Like the above, but the score is the
+ratio of the survival and death probabilities on the square. In the
+data set, every piece ends alive or dead in every square (except
+for the bishops, which can only legally occupy squares of their
+color) at least 1000 times, so each ratio is defined. Here, the
+sums of ratios have plenty of variability, and the highest
+ratios are not so often on the starting squares. So with this
+strategy, we simply do a weighted sample from the moves according
+to their (non-normalized) scores.
+
+\describeplayer{fatalist} The dual of \player{survivalist}; the score is
+the ratio of the death and survival probability on the square. This
+player knows that even if you win, you just have to keep playing over
+and over again, so you might as well get it over with!
 
@@
 
+Of course, people have been making more serious attempts at automating
+chess since before computers, and there are thousands of chess engines
+out there. We include a few in here to represent the high end of
+skill, and to make sure that weaker players are evaluated somewhat in
+terms of their ability to play chess proper, not just beat other weak
+players.
+
+% XXX citations for stockfish
+\describeplayer{stockfish0} Stockfish is probably the strongest
+open-source chess engine (or even publicly available engine); at full
+strength its play is estimated to be around 3500 on the FIDE scale.
+Aside from being quite machine dependent (it can search more moves in
+a given amount of time when it has a fast CPU with many cores), there
+are many options to fiddle with. Stockfish can use both opening books
+and endgame tables; neither is used here. It also has a ``difficulty''
+setting, which is set here to 0. Stockfish is run in a separate
+process, and the board is reset before each move, but I am not
+extremely hygienic about flushing e.g.~internal hash tables between
+moves. One consequence of this attempt at statelessness is that
+Stockfish sometimes walks into a draw by repetition in positions
+where it would be able to win, because it doesn't know that it is
+repeating positions.
+
+% TODO: Does lichess publish the difficulty settings ? Nice to compare
+% to my subjective skill level
+\describeplayer{stockfish5} Stockfish as above, but set to difficulty
+5.
+\describeplayer{stockfish10} Same, at difficulty 10.
+\describeplayer{stockfish15} Same, at difficulty 15.
+\describeplayer{stockfish20} And difficulty 20, the maximum. Even
+at this difficulty, Stockfish produces moves basically instantaneously.
+
+\describeplayer{stockfish1m} As expected, the engine's performance
+increases steadily as the difficulty setting increases (without
+apparently affecting the time to make moves). I don't know what it's
+doing with these settings. The true way to unleash chess engines is to
+give them a lot of CPU and memory to search. Since the tournament is
+run simultaneously across about 60 threads\footnote{ The computer is
+  the completely egregious AMD 2990WX ``Threadripper~2'', which has 32
+  cores (for 64 hardware threads) and 250 Watts of power dissipation
+  at load. The torture of this CPU was part of the impetus for the
+  paper.} using dozens of gigabytes of memory, and sometimes I would
+play {\it Hollow Knight} while it ran, I wanted to avoid having the
+performance be dependent on the scheduling environment. So here,
+Stockfish is given a ``node'' budget of 1~million, hence {\sf 1m}. It
+takes about one second per move when running alone, and is easily the
+strongest player evaluated.
+
+\begin{figure}[ht]
+% worstfish (black) somehow beating suicide king
+% white forces the mate. Note that worstfish
+% does assume its opponent will pick good moves...
+1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
+5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
+9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
+13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
+17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
+21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
+25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
+29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
+33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
+37. Qxd7+ Bxd7++
+\caption{TODO} \label{fig:worstfish}
+\end{figure}
+
+\describeplayer{worstfish} On the other hand, a strong chess engine
+can also be used to play badly. When playing as white, for each legal
+move, I ask Stockfish (configured as \player{stockfish0}) to evaluate
+the resulting position from black's perspective.\footnote{By asking it
+  to make a move, which also returns the evaluation of its move. The
+  UCI protocol does not seem to offer a way to evaluate a position
+  directly.} I then choose the move that produces the best position
+for black. This is easily the worst player evaluated, but it is not
+hard to imagine ways it could be worse. Indeed, playing for a loss is
+a common twist on chess ({\it Losing
+  Chess}~\cite{losingchesswikipedia}), although the rules of this
+variant require that you capture a piece if you are able to (without
+this constraint it is fairly boring). So, engines that support this
+variant cannot be directly applied. We could use stronger settings of
+Stockfish, but since it already invokes Stockfish for each legal move,
+it is also one of the slowest players. Like Stockfish, it occasionally
+blunders an otherwise losing position into a draw by repetition. But
+most importantly, its search strategy when evaluating positions is not
+applying the correct logic ($\alpha--\beta$ pruning); it assumes that
+its opponent will choose strong moves, and that it will itself play
+strong moves in future plies. As a result, it sometimes allows the
+opponent to create a situation where \player{worstfish} is forced to
+checkmate its opponent (Figure~\ref{fig:worstfish}).
+
+\describeplayer{topple10k} Topple~\cite{topple} is another strong,
+open source engine, provided to keep Stockfish on its toes. Here,
+its node budget is 10,000. \traditional
+
+\describeplayer{topple1m} Topple with a node budget of 1 million,
+as \player{stockfish1m}. Stockfish almost always wins, though it
+is not clear whether the budgets are actually comparable. \traditional
+
+\describeplayer{chessmaster.nes\_lv1}
+\describeplayer{chessmaster.nes\_lv2} \deterministic
+
 \subsection{Blind players}
+
+\describeplayer{blind\_yolo}
+\describeplayer{blind\_kings}
+% \describeplayer{blind\_spycheck} (described in paper, not yet run)
 
@@
 
-\describeplayer{worstfish}
-\describeplayer{safe}
-\describeplayer{popular}
-\describeplayer{dangerous}
-\describeplayer{rare}
-\describeplayer{survivalist}
-\describeplayer{sym\_mirror\_x}
-\describeplayer{sym\_180}
-\describeplayer{sym\_mirror\_y}
-\describeplayer{suicide\_king}
-\describeplayer{fatalist}
-\describeplayer{blind\_yolo}
-\describeplayer{cccp}
-\describeplayer{equalizer}
-\describeplayer{blind\_kings}
-\describeplayer{min\_oppt\_moves}
-\describeplayer{chessmaster.nes\_lv1}
 \describeplayer{stockfish1m\_r32768}
-\describeplayer{chessmaster.nes\_lv2}
 \describeplayer{stockfish1m\_r16384}
-\describeplayer{stockfish0}
-\describeplayer{stockfish5}
 \describeplayer{stockfish1m\_r8192}
-\describeplayer{topple10k}
-\describeplayer{stockfish15}
-\describeplayer{stockfish10}
 \describeplayer{stockfish1m\_r4096}
-\describeplayer{stockfish20}
-\describeplayer{topple1m}
 \describeplayer{stockfish1m\_r2048}
@@
 57. Qh7+
-
-% worstfish (black) somehow beating suicide king
-% white forces the mate. Note that worstfish
-% does assume its opponent will pick good moves...
-1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
-5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
-9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
-13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
-17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
-21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
-25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
-29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
-33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
-37. Qxd7+ Bxd7++
 

### 2019-03-14 chess/papers/blind.tex
Actually add Spycheck player, some writing, go to bed

@@
 type.\footnote{Since only 32 of the 64 bits can be set, you could do
-  slightly better by representing $\binom{60}{32}$ in $\sim 61$ bits.
+  slightly better by representing $\binom{64}{32}$ in $\sim 61$ bits.
   When fewer than 32 squares are occupied, we can use a record
@@
 Everything was trained on a single GPU, though a fairly decent one in
-2018, the EVGA GeForce GTX 1080 FTW (really), using OpenCL. Biases
+2018, the EVGA GeForce GTX 1080 ``FTW'' (really), using OpenCL. Biases
 were initialized to $0$ and weights with a Gaussian of mean $0$ and
@@
 \begin{figure}[ht]
-XXX actual example
-\caption{The white queen ``disappears'' after QxWhatever, because
-  unblinding predicts it to be one of black's own pawns, which is far
+\chessboard[tinyboard,setfen=rnbqkbnr/ppp2pQp/8/3pp3/4P3/8/PPPP1PPP/RNB1KBNR b KQkq - 0 3,showmover=false] (a)
+  \includegraphics[width=0.75 \linewidth]{queendisappears} (b)
+  \caption{(a) Position after 1.~e4 e5 2.~Qg4 d5 3.~Qxg7; note the white
+    queen strangely on g7.
+    (b) The bitmask for this position and the unblinder's prediction.
+    The queen ``disappears'' after Qxg7, because
+    unblinding predicts it to be one of black's own pawns---far
     more likely in that square.} \label{fig:trick}
@@
 
-Once we have predicated a board state, we can play with it. The
+Once we have predicted a board state, we can play with it. The
 simplest way to do this is to use a strong chess engine to pick a move
@@
     illegal: Moves are specified as a source and destination
-    square,\footnote{Plus promotion piece. Castling is represented as
+    square,\footnote{\,Plus promotion piece. Castling is represented as
       a two-square move of the king.} and so the source square of
@@
 would be to try to find moves that are good in expectation, that is,
-integrated all possible boards in proportion to their predicted
-probability. A good approximation might be had by sampling a bunch
-of boards according to the predicted distribution, and then using
-Stockfish to score the top $k$ moves for each; we can then order
-moves by their expected score. Unfortunately, it is not easy to
-efficiently get Stockfish to generate scored moves for $k \neq 1$.
-Even with $k = 1$, this approach would be slow, taking about a
-second for each (distinct) sampled board. So I did not try it,
-at least not before submitting this paper.
+integrated over all possible boards in proportion to their predicted
+probability. A good approximation might be had by sampling a bunch of
+boards according to the predicted distribution, and then using
+Stockfish to score the top $k$ moves for each; we can then order moves
+by their expected score. Unfortunately, it is not easy to efficiently
+get Stockfish to generate scored moves for $k \neq 1$. Even with $k =
+1$, this approach would be slow, taking about a second for each
+(distinct) sampled board. So I did not try it, at least not before
+submitting this paper.
 
-\medskip
+\subsubsection{No, {\em u r} a lnetwork} \label{sec:legalmove}
 I initially considered trying to solve this whole problem with neural
@@
 just discussed, without any particular extra logic. My plan was to
-make a multiple different components that coud be evaluated
+make multiple different components that could be evaluated
 separately, starting with the unblinder described, followed by a unit
@@
 also the 64-bit blinded mask if it likes) and scores each move.
-Training a reasonable unblinder took longer than I expected, and the
-legal move predictor never really worked that well,
+Predicting a legal move is also a natural function for machine
+learning; a move can be given just as a source and destination
+square.\footnote{ There are also four choices for promotion when
+  moving a pawn into the last rank. It is always the case that if any
+  promotion is legal, all choices are legal, so this does not need to
+  be encoded in this phase. Also, at this level of play, always
+  promoting to queen is a very safe simplification.} Many pairs of
+squares are always impossible (e.g. no piece can ever move from A1 to
+B8); so there are only 1792 potential moves to predict. However,
+training a reasonable unblinder took longer than I expected, and the
+legal move predictor never really worked that well (it has a harder
+job),
 % (XXX some evaluation of it)
@@
 
-I would like to thank the little people (pawns) and the anonymous
-referrees.
+I would like to thank the little people (pawns) and the the author-
+and content-blind anonymous referrees.
 

### 2019-03-14 chess/papers/weak.tex
Actually add Spycheck player, some writing, go to bed

@@
 
-\usepackage[T1]{fontenc}
+\usepackage[T2A,T1]{fontenc}
 \usepackage{libertine}%% Only as example for the romans/sans fonts
 \usepackage[scaled=0.85]{beramono}
+
+\usepackage[utf8]{inputenc}
+\usepackage[russian,english]{babel}
 
@@
 \usepackage{chessboard}
+\usepackage{gensymb}
 
@@
 
-\newcommand\describeplayer[1]{\smallskip\noindent {\texttt{\textbf{#1}}}.\quad}
+\newcommand\describeplayer[1]{\smallskip\noindent {\texttt{\textbf{#1}}}.\ \ }
 \newcommand\player[1]{\texttt{#1}}
@@
 \chessboard[setfen=1NB1K2R/8/8/8/8/8/6p1/rnbqkbn1 b - - 19 64,showmover=false]
-\caption{\player{reverse\_starting} draws against \player{reverse\_starting}
+\caption{\player{reverse\_starting} draws against itself
 by repetition, both players having happily moved their surviving pieces into the
@@
 broken deterministically by the source/destination squares, so this
-one is technically asymmetric. \deterministic \asymmetric
+one is technically asymmetric.
+% \begin{otherlanguage*}{russian}
+%   Ни шагу назад!
+% \end{otherlanguage*}
+\deterministic \asymmetric
 
@@
 \describeplayer{sym\_mirror\_y} As white, try to maximize symmetry
-when flipping vertically. Zero penalty for opposing e.g.~a \King with
+when flipping vertically. Zero penalty for opposing e.g.~a \King\, with
 a \king, but a small penalty when the pieces are not the same type,
-and a larger penalty when they are not opposite coors. The starting
+and a larger penalty when they are not opposite colors. The starting
 position is already symmetric this way, so this usually has the effect
 of copying the opponent's moves when possible. As ``copy your
-opponent's moves'' is a commonly known (but underspecified) strategy
+opponent's moves'' is a common (but underspecified) strategy
 discovered by many children, this player is close to being canonical.
@@
 \describeplayer{sym\_180} As \player{sym\_mirror\_y}, but maximize
-symmetry under $180\deg$ rotation of the board. An emergent priority is
+symmetry under 180\degree\ rotation
+% XXX latex symbol for ``degrees''
+of the board. An emergent priority is
 to ``castle'' with the king and queen to ``fix'' them.
@@
 
-\describeplayer{dangerous} The dual of \player{safe}; the score is
-the probability that the piece dies on that square. Note that a
-king is said to die if it is checkmated. This player likes to
-live life on the edge!
+\describeplayer{dangerous} The dual of \player{safe}; the score is the
+probability that the piece dies on that square. Note that a king is
+said to die if it is checkmated or his side resigns. This player likes
+to live life on the edge!
 
@@
 
-\describeplayer{fatalist} The dual of \player{fatalist}; the score is
+\describeplayer{fatalist} The dual of \player{survivalist}; the score is
 the ratio of the death and survival probability on the square. This
@@
 
-...
+Of course, people have been making more serious attempts at automating
+chess since before computers, and there are thousands of chess engines
+out there. We include a few in here to represent the high end of
+skill, and to make sure that weaker players are evaluated somewhat in
+terms of their ability to play chess proper, not just beat other weak
+players.
 
-\describeplayer{stockfish0}
-\describeplayer{stockfish5}
-\describeplayer{stockfish15}
-\describeplayer{stockfish10}
-\describeplayer{stockfish20}
+% XXX citations for stockfish
+\describeplayer{stockfish0} Stockfish is probably the strongest
+open-source chess engine (or even publicly available engine); at full
+strength its play is estimated to be around 3500 on the FIDE scale.
+Aside from being quite machine-dependent (it can search more moves in
+a given amount of time when it has a fast CPU with many cores), there
+are many options to fiddle with. Stockfish can use both opening books
+and endgame tables; neither is used here. It also has a ``difficulty''
+setting, which is set here to 0. Stockfish is run in a separate
+process, and the board is reset before each move, but I am not
+extremely hygienic about flushing e.g.~internal hash tables between
+moves. One consequence of this attempt at statelessness is that
+Stockfish sometimes walks into a draw by repetition in positions
+where it would be able to win, because it doesn't know that it is
+repeating positions.
 
-\describeplayer{worstfish} (compare ``losing chess'')
-uses stockfish20; we could use 1m (but slow), but it
-probably doesn't benefit from this depth of analysis,
-since the engine assumes the opponent is playing well.
-To do even worse, we would want to play moves that
-trap the opponent into beating us, as well as ones that 
-avoid repetition draws
+% TODO: Does lichess publish the difficulty settings ? Nice to compare
+% to my subjective skill level
+\describeplayer{stockfish5} Stockfish as above, but at difficulty 5.
 
-\describeplayer{chessmaster.nes\_lv1}
-\describeplayer{chessmaster.nes\_lv2} \deterministic
-\describeplayer{stockfish1m\_r32768}
-\describeplayer{stockfish1m\_r16384}
-\describeplayer{stockfish1m\_r8192}
-\describeplayer{topple1m}
-\describeplayer{topple10k}
-\describeplayer{stockfish1m}
+\describeplayer{stockfish10} Same, at difficulty 10.
+
+\describeplayer{stockfish15} Same, at difficulty 15.
+
+\describeplayer{stockfish20} And difficulty 20, the maximum. Even
+at this difficulty, Stockfish produces moves basically instantaneously.
+
+\describeplayer{stockfish1m} As expected, the engine's performance
+increases steadily as the difficulty setting increases (without
+apparently affecting the time to make moves). I don't know what it's
+doing with these settings. The true way to unleash chess engines is to
+give them a lot of CPU and memory to search. Since the tournament is
+run simultaneously across about 60 threads\footnote{ The computer is
+  the completely egregious AMD 2990WX ``Threadripper~2,'' which has 32
+  cores (for 64 hardware threads) and 250 Watts of power dissipation
+  at load. The torture of this CPU was part of the impetus for the
+  paper.} using dozens of gigabytes of memory, and sometimes I would
+play {\it Hollow Knight} while it ran, I wanted to avoid having the
+performance be dependent on the scheduling environment. So here,
+Stockfish is given a ``node'' budget of 1~million, hence {\sf 1m}. It
+takes about one second per move when running alone, and is easily the
+strongest player (type) evaluated.
+
+\begin{figure}[ht]
+  \chessboard[setfen=r1bK1b2/1p1pn3/p2k3N/1np1pQ2/2P2P1P/N2P4/R5B1/4B2R w - - 1 37,showmover=false]
+  \caption{\player{suicide\_king} (white) to move against
+    \player{worstfish} after 36 moves. Since the kings
+    are already at their minimum distance, white will make
+    a move at random. 37.~Qxe5++ wins for white immediately,
+    but \player{suicide\_king} plays 37.~Qxd7??. The only
+    legal move is 37\ldots Bxd7++, so \player{worstfish}
+    must play it, and thus wins one of its only victories.} \label{fig:worstfish}
+% full game:
+% 1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
+% 5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
+% 9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
+% 13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
+% 17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
+% 21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
+% 25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
+% 29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
+% 33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
+% 37. Qxd7+ Bxd7++
+\end{figure}
+
+\describeplayer{worstfish} On the other hand, a strong chess engine
+can also be used to play badly. When playing as white, for each legal
+move, I ask Stockfish (configured as \player{stockfish0}) to evaluate
+the resulting position from black's perspective.\footnote{By asking it
+  to make a move, which also returns the evaluation of its move. The
+  UCI protocol does not seem to offer a way to evaluate a position
+  directly.} I then choose the move that produces the best position
+for black. This is easily the worst player evaluated, but it is not
+hard to imagine ways it could be worse. Indeed, playing for a loss is
+a common twist on chess ({\it Losing
+  Chess}~\cite{wikipedialosingchess}), although the rules of this
+variant require that you capture a piece if you are able to (without
+this constraint it is fairly boring). So, engines that support this
+variant cannot be directly applied. We could use stronger settings of
+Stockfish, but since it already invokes Stockfish for each legal move,
+it is also one of the slowest players. Like Stockfish, it occasionally
+blunders an otherwise losing position into a draw by repetition. But
+most importantly, its search strategy when evaluating positions is not
+applying the correct logic ($\alpha$--$\beta$ pruning); it assumes that
+its opponent will choose strong moves, and that it will itself play
+strong moves in future plies. As a result, it sometimes allows the
+opponent to create a situation where \player{worstfish} is forced to
+checkmate its opponent (Figure~\ref{fig:worstfish}).
+
+\describeplayer{topple10k} Topple~\cite{topple} is another strong,
+open source engine, provided to keep Stockfish on its toes. Here,
+its node budget is 10,000. \traditional
+
+\describeplayer{topple1m} Topple with a node budget of 1 million,
+which like \player{stockfish1m} takes about one second per move.
+Stockfish almost always wins, though it is not clear whether the
+budgets are actually comparable. \traditional
+
+\describeplayer{chessmaster.nes\_lv1} \deterministic \traditional
+
+\describeplayer{chessmaster.nes\_lv2} \deterministic \traditional
 
@@
 \describeplayer{blind\_yolo}
+
 \describeplayer{blind\_kings}
@@
 
+\describeplayer{stockfish1m\_r32768}
+\describeplayer{stockfish1m\_r16384}
+\describeplayer{stockfish1m\_r8192}
 \describeplayer{stockfish1m\_r4096}
@@
 
-% worstfish (black) somehow beating suicide king
-% white forces the mate. Note that worstfish
-% does assume its opponent will pick good moves...
-1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
-5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
-9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
-13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
-17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
-21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
-25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
-29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
-33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
-37. Qxd7+ Bxd7++
-
 % Fate-based algorithms.
@@
 
+Shout out to the Thursd'z Institute and anonymous commenters
+on my blog for discussion of players and suggestions. Several
+ideas were suggested by multiple people, increasing my confidence
+that they are somehow canonical.
+
 The author would like to thank the anonymous referees

### 2019-03-15 chess/tournament.cc
Added icons for the players and wrote about chessmaster, etc.

@@ -62,6 +62,13 @@ static constexpr int SELFPLAY_TARGET = 5;
 
 typedef Player *(*Entrant)();
 
+
+static Player *Stockfish1M_57344() {
+  return new BlendRandom<57344>(Stockfish1M());
+}
+static Player *Stockfish1M_49152() {
+  return new BlendRandom<49152>(Stockfish1M());
+}
 static Player *Stockfish1M_32768() {
   return new BlendRandom<32768>(Stockfish1M());
 }
@@ -142,6 +149,8 @@ const vector<Entrant> &GetEntrants() {
 			
 			Stockfish1M,
 
+			Stockfish1M_57344,
+			Stockfish1M_49152,
 			Stockfish1M_32768,
 			Stockfish1M_16384,
 			Stockfish1M_8192,

### 2019-03-15 trunk/chess/papers/weak.tex
Added icons for the players and wrote about chessmaster, etc.

@@
 
+% \usepackage[utf8]{inputenc}
+% \usepackage[russian,english]{babel}
+
 \usepackage{url}
@@
 
-\newcommand\describeplayer[1]{\smallskip\noindent {\texttt{\textbf{#1}}}.\quad}
+\newcommand\describeplayer[1]{\smallskip\noindent {\texttt{\textbf{#1}}}.\ \ }
 \newcommand\player[1]{\texttt{#1}}
-\newcommand\deterministic{\colorbox{black}{\color{white} \textsf{det}}}
-\newcommand\traditional{\colorbox{black}{\color{white} \textsf{trad}}}
-% funny to use that v symbol made of leaves
-\newcommand\vegetarian{\colorbox{black}{\color{white} \textsf{v}}}
-\newcommand\canonical{\colorbox{black}{\color{white} \textsf{canon}}}
-\newcommand\stateful{\colorbox{black}{\color{white} \textsf{state}}}
-\newcommand\asymmetric{\colorbox{black}{\color{white} \textsf{asym}}}
+\newcommand\russian{
+  \includegraphics[height=0.9em]{zazdarovye}
+}
+\newcommand\deterministic{
+  \includegraphics[height=0.9em]{deterministic}
+}
+\newcommand\traditional{
+  \includegraphics[height=0.9em]{traditional}
+}
+\newcommand\vegetarian{
+  \includegraphics[height=0.9em]{vegetarian}
+}
+\newcommand\canonical{
+  \includegraphics[height=0.9em]{canonical}
+}
+\newcommand\stateful{
+  \includegraphics[height=0.9em]{stateful}
+}
+\newcommand\asymmetric{
+  \includegraphics[height=0.9em]{asymmetric}
+}
 
-\describeplayer{random} We must begin with the most canonical of all
+\describeplayer{random\_move} We must begin with the most canonical of all
 strategies: Choosing a legal move uniformly at random. This is a
@@
   advanced; these are actually the same distance from the King as they
-  are in their home row, since the distance metric includes diagonal
+  would be in their home row, since the distance metric includes diagonal
   moves. } \label{fig:huddle}
@@
 broken deterministically by the source/destination squares, so this
-one is technically asymmetric. \deterministic \asymmetric
+one is technically asymmetric.
+% \begin{otherlanguage*}{russian}
+% За здоровье!
+% \end{otherlanguage*}
+\deterministic \asymmetric \russian
 
@@
 
-% XXX include examples of symmetric players, as they are aesthetic.
 \describeplayer{sym\_mirror\_y} As white, try to maximize symmetry
@@
 maximize symmetry when flipping horizontally. This does not make much
-chess sense, but can produce aesthetic arrangements
-(Figure~\ref{fig:XXX}).
+chess sense, but can produce aesthetic arrangements.
 
-% XXX show self play perhaps?
 \describeplayer{sym\_180} As \player{sym\_mirror\_y}, but maximize
-symmetry under 180 degree
-% XXX latex symbol for ``degrees''
-rotation of the board. An emergent priority is
+symmetry under 180\degree\ rotation
+of the board (Figure~\ref{fig:sym180}. An emergent priority is
 to ``castle'' with the king and queen to ``fix'' them.
+
+\begin{figure}
+  \chessboard[setfen=2bkr3/8/n1r1pppp/1PppQ2P/N2BPPP1/1PPP1R1N/8/3RKB2 b - - 8 66,showmover=false]
+  \caption{\player{sym\_180} (white) vs. \player{pacifist} after 66 moves. Note that the position is not quite rotationally symmetric, but is close given the material.} \label{fig:sym180}
+\end{figure}
 
@@
 
-% XXX cite blog comment?
 \describeplayer{equalizer} Prefer to move a piece that has been
@@
 to be very similar (dominated by the many pieces that don't move). But
-it's pretty arbitrary.
+it's pretty arbitrary. \stateful
 
@@
 square, whether it lives or dies. This player likes to follow the
-crowd!
+crowd! \stateful
 
@@
 said to die if it is checkmated or his side resigns. This player likes
-to live life on the edge!
+to live life on the edge! \stateful
 
@@
 is one minus the probability of ending the game on that square. This
-player has a thirst for adventure!
+player has a thirst for adventure! \stateful
 
@@
 strategy, we simply do a weighted sample from the moves according
-to their (non-normalized) scores.
+to their (non-normalized) scores. \stateful
 
@@
 player knows that even if you win, you just have to keep playing over
-and over again, so you might as well get it over with!
+and over again, so you might as well get it over with! \stateful
 
@@
 
-% XXX citations for stockfish
-\describeplayer{stockfish0} Stockfish is probably the strongest
-open-source chess engine (or even publicly available engine); at full
-strength its play is estimated to be around 3500 on the FIDE scale.
-Aside from being quite machine dependent (it can search more moves in
-a given amount of time when it has a fast CPU with many cores), there
-are many options to fiddle with. Stockfish can use both opening books
-and endgame tables; neither is used here. It also has a ``difficulty''
-setting, which is set here to 0. Stockfish is run in a separate
-process, and the board is reset before each move, but I am not
-extremely hygienic about flushing e.g.~internal hash tables between
-moves. One consequence of this attempt at statelessness is that
-Stockfish sometimes walks into a draw by repetition in positions
-where it would be able to win, because it doesn't know that it is
-repeating positions.
+\describeplayer{stockfish0} Stockfish~\cite{stockfish} is probably the
+strongest open-source chess engine (or even publicly available
+engine); at full strength its play is estimated to be around 3500 Elo
+on the FIDE scale. Aside from being quite machine-dependent (it can
+search more moves in a given amount of time when it has a fast CPU
+with many cores), there are many options to fiddle with. Stockfish can
+use both opening books and endgame tables; neither is used here. It
+also has a ``difficulty'' setting, which is set here to 0. Stockfish
+is run in a separate process, and the board is reset before each move,
+but I am not extremely hygienic about flushing e.g.~internal hash
+tables between moves. One consequence of this attempt at statelessness
+is that Stockfish sometimes walks into a draw by repetition in
+positions where it would be able to win, because it doesn't know that
+it is repeating positions. \traditional
 
@@
 % to my subjective skill level
-\describeplayer{stockfish5} Stockfish as above, but set to difficulty
-5.
+\describeplayer{stockfish5} Stockfish as above, but at difficulty 5.
+\traditional
+
 \describeplayer{stockfish10} Same, at difficulty 10.
+\traditional
+
 \describeplayer{stockfish15} Same, at difficulty 15.
+\traditional
+
 \describeplayer{stockfish20} And difficulty 20, the maximum. Even
 at this difficulty, Stockfish produces moves basically instantaneously.
+\traditional
 
@@
 run simultaneously across about 60 threads\footnote{ The computer is
-  the completely egregious AMD 2990WX ``Threadripper~2'', which has 32
+  the completely egregious AMD 2990WX ``Threadripper~2,'' which has 32
   cores (for 64 hardware threads) and 250 Watts of power dissipation
@@
 takes about one second per move when running alone, and is easily the
-strongest player evaluated.
+strongest player (type) evaluated.
+\traditional
 
 \begin{figure}[ht]
-% worstfish (black) somehow beating suicide king
-% white forces the mate. Note that worstfish
-% does assume its opponent will pick good moves...
-1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
-5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
-9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
-13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
-17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
-21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
-25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
-29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
-33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
-37. Qxd7+ Bxd7++
-\caption{TODO} \label{fig:worstfish}
+  \chessboard[setfen=r1bK1b2/1p1pn3/p2k3N/1np1pQ2/2P2P1P/N2P4/R5B1/4B2R w - - 1 37,showmover=false]
+  \caption{\player{suicide\_king} (white) to move against
+    \player{worstfish} after 36 moves. Since the kings
+    are already at their minimum distance, white will make
+    a move at random. 37.~Qxe5++ wins for white immediately,
+    but \player{suicide\_king} plays 37.~Qxd7??. The only
+    legal move is 37\ldots Bxd7++, so \player{worstfish}
+    must play it, and thus wins one of its only victories.}
+  \label{fig:worstfish}
+% full game:
+% 1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
+% 5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
+% 9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
+% 13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
+% 17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
+% 21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
+% 25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
+% 29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
+% 33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
+% 37. Qxd7+ Bxd7++
 \end{figure}
@@
 for black. This is easily the worst player evaluated, but it is not
-hard to imagine ways it could be worse. Indeed, playing for a loss is
-a common twist on chess ({\it Losing
-  Chess}~\cite{losingchesswikipedia}), although the rules of this
-variant require that you capture a piece if you are able to (without
-this constraint it is fairly boring). So, engines that support this
-variant cannot be directly applied. We could use stronger settings of
-Stockfish, but since it already invokes Stockfish for each legal move,
-it is also one of the slowest players. Like Stockfish, it occasionally
-blunders an otherwise losing position into a draw by repetition. But
-most importantly, its search strategy when evaluating positions is not
-applying the correct logic ($\alpha--\beta$ pruning); it assumes that
-its opponent will choose strong moves, and that it will itself play
-strong moves in future plies. As a result, it sometimes allows the
-opponent to create a situation where \player{worstfish} is forced to
-checkmate its opponent (Figure~\ref{fig:worstfish}).
+hard to imagine ways it could be worse. Indeed, a common twist on
+chess is to play for a loss, called {\it Antichess} or {\it Losing
+  Chess}~\cite{wikipedialosingchess}. Recently it was even proved that
+white can always win (i.e.~lose)~\cite{watkins2017losing} in this
+variant! However, the variant requires that you capture a piece if
+you are able to, so strategies and engines that support this variant
+cannot be directly applied.
+We could use stronger settings of Stockfish, but since it already
+invokes Stockfish for each legal move, it is also one of the slowest
+players. Like Stockfish, it occasionally blunders an otherwise losing
+position into a draw by repetition. But most importantly, its search
+strategy when evaluating positions is not applying the correct logic
+($\alpha$--$\beta$ pruning); it assumes that its opponent will choose
+strong moves, and that it will itself play strong moves in future
+plies. As a result, it sometimes allows the opponent to create a
+situation where \player{worstfish} is forced to checkmate its opponent
+(Figure~\ref{fig:worstfish}).
 
-\describeplayer{topple10k} Topple~\cite{topple} is another strong,
+\describeplayer{topple10k} Topple~\cite{topple} is another strong
 open source engine, provided to keep Stockfish on its toes. Here,
@@
 \describeplayer{topple1m} Topple with a node budget of 1 million,
-as \player{stockfish1m}. Stockfish almost always wins, though it
-is not clear whether the budgets are actually comparable. \traditional
+which like \player{stockfish1m} takes about one second per move.
+Stockfish almost always wins, though it is not clear whether the
+budgets are actually comparable. \traditional
 
-\describeplayer{chessmaster.nes\_lv1}
-\describeplayer{chessmaster.nes\_lv2} \deterministic
+\describeplayer{chessmaster.nes\_lv1} This is {\it Chessmaster} for
+the Nintendo Entertainment System, published in AD 1989. Moves are
+extracted from the game via emulation. This proved to be somewhat
+delicate, because the in-memory representation of the board is not
+that simple (it appears to be represented in two parallel arrays,
+perhaps using the ``0x88 method'') and the game understandably goes
+wild if you mess it up. To play a move, I restore a saved machine
+state in the ``board editor'' mode, and then modify memory to contain
+the desired board. I then emulate button presses to operate the game
+menu and return to ``playing'' mode. Chessmaster repairs its internal
+data structures from this board, and makes a move. During this time I
+mash controller buttons to skip its dramatic tunes and modal messages
+like \verb+CHECK+. Once some memory locations reach certain values,
+the move has been played; I can diff the before and after boards to
+uniquely determine the move. Since this approach uses emulation, it
+would normally be completely deterministic, but I deliberately stall
+for a random number of frames so that it can advance its internal
+pseudorandom state. A nice thing about this engine is that it is an
+earnest attempt at writing a good engine, but limited to the
+techniques of the 1980s, and running on hardware that was underpowered
+even for its day. It finishes well behind the modern engines, but
+ahead of all the non-traditional strategies. \traditional
+
+\describeplayer{chessmaster.nes\_lv2} Same as
+\player{chessmaster.nes\_lv1}, but increasing the ``Level Of Play'' to
+``Newcomer 2.'' Chessmaster has stronger levels still, but in these
+modes it will think for several NES minutes, which takes multiple
+seconds to emulate---too slow for our purposes. Still much
+weaker than modern engines (Figure~\ref{fig:chessmaster}).
+\traditional
+
+\begin{figure}[ht]
+  \includegraphics[width=0.75 \linewidth]{chessmaster}
+  \caption{\player{stockfish1m} (2019; running on a modern machine) beats
+    \player{chessmaster.nes\_lv2} (1989; running on an emulated NES)
+    thousands of times without a loss or draw.} \label{fig:chessmaster}
+\end{figure}
 
@@
 
-\describeplayer{blind\_yolo}
-\describeplayer{blind\_kings}
-% \describeplayer{blind\_spycheck} (described in paper, not yet run)
+\describeplayer{blind\_yolo} This player is only allowed to see the
+64-bit mask of where pieces are placed on the board, but not their
+colors or types. It is described in {\it Color- and Piece-blind
+  Chess}~\cite{blind}.
 
-(I guess we can also segue by discussing handicaps)
-(end with this so that we can segue into dilution methods)
+\describeplayer{blind\_kings} As \player{blind\_yolo}, but forcing
+the prediction to produce exactly one king for each side, which
+improves its accuracy a little.
+
+\describeplayer{blind\_spycheck} As \player{blind\_kings}, but
+performing a ``spy check'' before each move. Here, the player tries
+capturing each piece of a given (predicted) color with other pieces
+of that same (predicted) color. If such a move is legal, then one
+of the two pieces was mispredicted, so we prefer the capture over
+sending an incorrect position to the engine.
+
+Each of these uses a neural network to predict the configuration of
+the board, and then the equivalent of \player{stockfish1m} to make a
+move for that predicted board. (If that move is illegal because the
+board was mispredicted, then it plays a random legal move.) These
+players can therefore be seen as handicaps on \player{stockfish1m}.
 
 \subsection{Diultion methods}
+
+Even at its weakest setting, \player{stockfish0} crushes all of the
+nontraditional strategies. This is not surprising, but it creates
+a problem for quantifying their difference in skill (do they win
+one in a million games? One in 10^{100}?). Having intermediate
+players allows some Elo points to flow between the two tiers
+transitively. One nice way to construct intermediate players is
+by interpolation. We can interpolate between any two players,\footnote{
+  Even if they are stateful! The interface separates ``please
+  give me a move for the current state'' and ``the following
+  move was played; please update your state,'' allowing them
+  to disagree. So we can just keep two independent states for
+  the two players.} but interpolating with the most canonical
+player, \player{random\_move}.
+
 % Stockfish1m_32768 loses to reverse_starting in 3 moves!
@@
 \nocite{miller2013counting}
+\nocite{watkins2017losing}
 

### 2019-03-16 chess/papers/blind.tex
Multi-round elo, quartiles, generate heatmap png for figure. Submitted versions of papers

@@
 interest, so I wrote a separate paper about that~\cite{eloworld}. This
-algorithm, called \verb+blind_yolo+, had an Elo World score of XXX. It
+algorithm, called \verb+blind_yolo+, had an Elo World score of $489 \pm 2$. It
 beats a purely random player with a score of 101 wins, 27 losses, and
@@
 
-This version, called \verb+blind_kings+, performs consistently better
+This version, called \verb+blind_kings+, performs a small amount better
 than \verb+blind_yolo+ (63 wins, 45 losses, 412 draws).
@@
 % as white: 31/24/207, as black 21/32/205
-It had an Elo World score of XXX.
+It had an Elo World score of $502 \pm 3$.
 
@@
 
-(XXX evaluation here)
+This version, \verb+blind_spycheck+, works significantly better than
+\verb+blind_kings+. It has an Elo World score of $547 \pm 1$, somewhere
+between a 93.75--96.875\% dilution of \verb+stockfish1m+ (the third
+best non-engine player).
 

### 2019-03-16 chess/headless-graphics.cc

@@ -21,6 +21,17 @@ void SetPixel(int w, int h, int x, int y,
   (*argb)[i + 3] = b;
 }
 
+void FillRect(int w, int h,
+	      int rectx, int recty, int rectw, int recth,
+	      uint8 a, uint8 r, uint8 g, uint8 b,
+	      vector<uint8> *argb) {
+  for (int yy = 0; yy < recth; yy++) {
+    for (int xx = 0; xx < rectw; xx++) {
+      SetPixel(w, h, rectx + xx, recty + yy, a, r, g, b, argb);
+    }
+  }
+}
+
 void SaveARGB(const vector<uint8> &argb, int width, int height,
 	      const string &filename) {
   CHECK(argb.size() == width * height * 4);
@@ -30,10 +41,10 @@ void SaveARGB(const vector<uint8> &argb, int width, int height,
     // This is all messed up, maybe because I got ARGB wrong in emu
     // (TODO: fix that) and also stb_ wants little-endian words. But
     // anyway, it works like this.
-    uint8 b = argb[i + 0];
-    uint8 g = argb[i + 1];
-    uint8 r = argb[i + 2];
-    uint8 a = argb[i + 3];
+    uint8 a = argb[i + 0];
+    uint8 r = argb[i + 1];
+    uint8 g = argb[i + 2];
+    uint8 b = argb[i + 3];
     
     rgba[i + 0] = r;
     rgba[i + 1] = g;

### 2019-03-16 chess/tournament.cc

@@ -62,7 +62,15 @@ static constexpr int SELFPLAY_TARGET = 5;
 
 typedef Player *(*Entrant)();
 
-
+static Player *Stockfish1M_64512() {
+  return new BlendRandom<64512>(Stockfish1M());
+}
+static Player *Stockfish1M_63488() {
+  return new BlendRandom<63488>(Stockfish1M());
+}
+static Player *Stockfish1M_61440() {
+  return new BlendRandom<61440>(Stockfish1M());
+}
 static Player *Stockfish1M_57344() {
   return new BlendRandom<57344>(Stockfish1M());
 }
@@ -149,6 +157,9 @@ const vector<Entrant> &GetEntrants() {
 			
 			Stockfish1M,
 
+			Stockfish1M_64512,
+			Stockfish1M_63488,
+			Stockfish1M_61440,
 			Stockfish1M_57344,
 			Stockfish1M_49152,
 			Stockfish1M_32768,

### 2019-03-16 trunk/chess/papers/weak.tex
Multi-round elo, quartiles, generate heatmap png for figure. Submitted versions of papers

@@
 \usepackage{gensymb}
+\usepackage{supertabular}
 
@@
 
-The problem this paper is concerned with is that basically all chess
+The problem this paper addresses is that basically all chess
 tournaments (whether with humans or computers or both) are between
@@
 give a rating to weak players: They just lose every single game and so
-tend towards a rating of $-\infty$\footnote{Some organizations don't
+tend towards a rating of $-\infty$.\footnote{Some organizations don't
   even let ratings fall beneath a certain level, for example, the
@@
   ${\tt position} \rightarrow {\tt move}$.
-\item black and white behave the same
+\item The player should try to ``behave the same'' when playing with
+  the white or black pieces. Of course this can't be literally true,
+  and in fact some algorithms can't be symmetric, so it's, like, a
+  suggestion.
 \item Avoid game-tree search. Minimax, alpha--beta, etc. are the
@@
 }
+
+\begin{figure}[t]
+\begin{tabular}{cl}
+  \deterministic & The player is deterministic \\
+  \traditional & Traditional approach to chess (e.g. engine) \\
+  \vegetarian & Vegetarian or vegetarian option available \\
+  \canonical & A canonical algorithm! \\
+  \stateful & Stateful (not including pseudorandom pool) \\
+  \asymmetric & Asymmetric \\
+\end{tabular}
+\caption{Key} \label{fig:key}
+\end{figure}
 
@@
 \describeplayer{alphabetical} Make the alphabetically first move,
-using standard PGN short notation. White and black both try to move
+using standard PGN short notation (e.g. ``a3'' $<$ ``O-O'' $<$ ``Qxg7''). White and black both try to move
 towards A1. \asymmetric \deterministic
@@
 
+% Looks like this is the only one with *no* wins!
 \describeplayer{no\_i\_insist} Like \player{generous}, but be overwhelmingly
@@
 upside-down, and as white, tries to put its pieces where black pieces
-start. This often causes some conflict since the black pieces are already
-there, but can also end peacefully (Figure~\ref{fig:reversestarting}).
+start. Since here we have a specific configuration in mind, we can produce
+a distance metric by computing the total distance from each piece to
+its desired destination. Each piece uses a different distance metric;
+Chebyshev distance for the King, Manhattan distance for the Rook, and
+a pretty weird function for the Knight~\cite{miller2013counting}.
+Trying to move the pieces into the reversed starting position often
+causes some conflict since the black pieces are already there, but can
+also end peacefully (Figure~\ref{fig:reversestarting}).
 
@@
 
-% TODO: Cyrillic tag or toast at the end
 \describeplayer{cccp} Prioritize moves that Checkmate, Check, Capture
@@
 one is technically asymmetric.
-% \begin{otherlanguage*}{russian}
-% За здоровье!
-% \end{otherlanguage*}
 \deterministic \asymmetric \russian
@@
 symmetry under 180\degree\ rotation
-of the board (Figure~\ref{fig:sym180}. An emergent priority is
+of the board (Figure~\ref{fig:sym180}). An emergent priority is
 to ``castle'' with the king and queen to ``fix'' them.
@@
 ends~\cite{survival}.
+
+% (could work this into a footnote)
+% weighted random is almost the same as random, because
+% the weights tend to be very similar when treated
+% additively. They perform slightly different from
+% random, as so (101.5 million games):
+
+%  player	elo	w/l/d
+%  dangerous	986.29	1374577/1501526/15683897
+%  popular	997.34	1360053/1502995/15696952
+%  random_move	1000.79	1540798/1337668/15681534
+%  rare	1005.85	1554600/1326707/15678693
+%  safe	1009.72	1341030/1502162/15716808
+
+% then with normalization:
+%  player	elo	w/l/d
+%  popular	990.93	25756/43561/346683
+%  safe	993.49	28082/40769/347149
+%  dangerous	996.60	30063/37451/348486
+%  rare	1005.82	50656/27390/337954
+%  random_move	1013.15	41747/27133/347120
 
@@
   paper.} using dozens of gigabytes of memory, and sometimes I would
-play {\it Hollow Knight} while it ran, I wanted to avoid having the
-performance be dependent on the scheduling environment. So here,
+play {\it Hollow Knight} ({\it aka} \knight) while it ran, I wanted to avoid having the
+chess skill be dependent on the scheduling environment. So here,
 Stockfish is given a ``node'' budget of 1~million, hence {\sf 1m}. It
@@
 
-\describeplayer{chessmaster.nes\_lv2} Same as
-\player{chessmaster.nes\_lv1}, but increasing the ``Level Of Play'' to
-``Newcomer 2.'' Chessmaster has stronger levels still, but in these
-modes it will think for several NES minutes, which takes multiple
-seconds to emulate---too slow for our purposes. Still much
-weaker than modern engines (Figure~\ref{fig:chessmaster}).
+\describeplayer{chessmaster.nes\_lv2} As above, but increasing the
+``Level Of Play'' to ``Newcomer 2.'' Chessmaster has stronger levels
+still, but in these modes it will think for several NES minutes, which
+takes multiple seconds to emulate---too slow for our purposes. Still
+much weaker than modern engines (Figure~\ref{fig:chessmaster}).
 \traditional
@@
 
-\subsection{Diultion methods}
+\subsection{Interpolation methods}
 
 Even at its weakest setting, \player{stockfish0} crushes all of the
-nontraditional strategies. This is not surprising, but it creates
-a problem for quantifying their difference in skill (do they win
-one in a million games? One in 10^{100}?). Having intermediate
-players allows some Elo points to flow between the two tiers
-transitively. One nice way to construct intermediate players is
-by interpolation. We can interpolate between any two players,\footnote{
-  Even if they are stateful! The interface separates ``please
-  give me a move for the current state'' and ``the following
-  move was played; please update your state,'' allowing them
-  to disagree. So we can just keep two independent states for
-  the two players.} but interpolating with the most canonical
-player, \player{random\_move}.
+nontraditional strategies. This is not surprising, but it creates a
+problem for quantifying their difference in skill (do they win one in
+a million games? One in $10^{100}$?). Having intermediate players
+allows some Elo points to flow between the two tiers transitively. One
+nice way to construct intermediate players is by interpolation. We can
+interpolate between any two players,\footnote{ Even if they are
+  stateful! The interface separates ``please give me a move for the
+  current state'' and ``the following move was played; please update
+  your state,'' allowing them to disagree. So we can just keep two
+  independent states for the two players.} but it is most natural to
+interpolate with the most canonical player, \player{random\_move}.
 
-% Stockfish1m_32768 loses to reverse_starting in 3 moves!
-1. e4 g5 2. Ba6 f5 3. Qh5
+\describeplayer{stockfish1m\_r{\it nnn}} There are 15 players in this
+group, each characterized by a number $nnn$. Before each move, we
+generate a random 16-bit number; if the number is less than $nnn$ then
+we play a random move and otherwise, a move as \player{stockfish1m}.
+The values of $nnn$ are chosen so that we mix a power-of-two
+fraction of noise: \player{stockfish1m\_r32768} blends half random
+with half strong moves, but we also have $1/4$ random, $1/8$ random,
+$1/16$, $1/32$, \ldots, $1/1024$. These give us a nice smooth
+gradation at high levels of play (Figure~\ref{fig:heatmap}). At the
+extremes, many games have no random moves, and the performance is
+difficult to distinguish from \player{stockfish1m}.
+% in fact, highest IS r64?
+Even at half random moves, \player{stockfish1m\_r32768} consistently
+beats the non-traditional players. So we also dilute stockfish by
+mixing with a majority of random moves: \player{stockfish1m\_r49152}
+is $3/4$ random, and we also have $7/8$, $15/16$, $31/32$, and
+$63/64$. At this point it becomes hard to distinguish from
+\player{random\_move}.
 
-\describeplayer{stockfish1m\_r32768}
-\describeplayer{stockfish1m\_r16384}
-\describeplayer{stockfish1m\_r8192}
-\describeplayer{stockfish1m\_r4096}
-\describeplayer{stockfish1m\_r2048}
-\describeplayer{stockfish1m\_r1024}
-\describeplayer{stockfish1m\_r512}
-\describeplayer{stockfish1m\_r256}
-\describeplayer{stockfish1m\_r64}
-\describeplayer{stockfish1m\_r128}
+Note that playing $1/64$ excellent moves and the rest randomly doesn't
+do much to help one's performance. On the other hand, playing one
+random move for $63/64$ strong ones does create a significant
+handicap! Against strong opponents, it's easy to see how a mistake can
+end the game. Even against weak ones, a blunder can be fatal
+(Figure~\ref{fig:s32768}).
+% TODO: graph elo with random fraction
 
-Ideally, such a system would be robust against adding new players,
-but this is probably not the case.
+\begin{figure}[ht]
+\chessboard[setfen=rnbqkbnr/ppppp2p/B7/5ppQ/4P3/8/PPPP1PPP/RNB1K1NR b KQkq - 1 3,showmover=false]
+\caption{A demonstration of how quickly random play can ruin a game.
+  \player{stockfish1m\_r32768} (black; plays half random moves
+  and half strong moves) loses to \player{reverse\_starting} in one of
+  the shortest possible sequences: 1.~e4 g5 2.~Ba6 f5 3.~Qh5++. Both
+  of black's moves are bad (so they must be random) but white's 2.~Ba6
+  is a blunder as well. White is just pushing pieces as far as possible;
+  the mate is purely accidental.} \label{fig:s32768}
+\end{figure}
+
+%  \describeplayer{stockfish1m\_r64512} 63/64
+%  \describeplayer{stockfish1m\_r63488} 31/32
+%  \describeplayer{stockfish1m\_r61440} 15/16
+%  \describeplayer{stockfish1m\_r57344} 7/8
+%  \describeplayer{stockfish1m\_r49152} 3/4
+%  \describeplayer{stockfish1m\_r32768} 1/2
+%  \describeplayer{stockfish1m\_r16384} 1/4
+%  \describeplayer{stockfish1m\_r8192} 1/8
+%  \describeplayer{stockfish1m\_r4096} 1/16
+%  \describeplayer{stockfish1m\_r2048} 1/32
+%  \describeplayer{stockfish1m\_r1024} 1/64
+%  \describeplayer{stockfish1m\_r512} 1/128
+%  \describeplayer{stockfish1m\_r256} 1/256
+%  \describeplayer{stockfish1m\_r128} 1/512
+%  \describeplayer{stockfish1m\_r64} 1/1024
+
+\section{Running the tournament}
+
+Given the players, it's a simple matter to simulate a bunch of games.
+Since the tournament runs for tens of thousands of CPU hours, there is
+of course some code to allow it to checkpoint its work, to make
+efficient use of CPU and RAM, and to provide attractive ANSI graphics
+showing off its activity---but this is all straightforward.\footnote{
+  The source code can be found at
+  \url{sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/chess/}
+} The
+tournament accumulates win/loss/draw counts for each pair of players
+(playing as both white and black), as well as example games used to
+debug or create the figures in this paper.
+
+After the tournament is done, or from any checkpoint, I can run a
+separate program to produce Elo (and other) rankings. Elo depends
+on the order in which games are played, because it is designed to take
+into account changes in player skill over time. However, it is easy
+to just make up a random order in which the games were played, because
+these players do not change over time.
+
+Despite its strong reputation, I found Elo to be rather finnicky. It
+is sensitive to the starting scores for the players, and the factor
+$k$ which governs how strong of a correction to make after a game.
+With players of such vastly different skill, it is also very sensitive
+to the order in which the games are played.\footnote{For example,
+  suppose the weak player \player{random\_move} has one win against
+  \player{stockfish1m}. If this win happens early, when the two
+  players have their initial Elo scores, then it doesn't affect
+  anything. If it happens last, when \player{stockfish1m} has
+  thousands of Elo points over \player{random\_move}, then it produces
+  a very strong correction.} For similar reasons, it is also very
+sensitive to imbalance in the number of games played between a
+particular pair of players; if some player mostly has games against
+weak players, this can artificially inflate its score.
+
+To control for these effects, I compute the maximum number $n$ for
+which we have $n$ games played between every pair of players. I then
+seed each player with an initial Elo score of 1000, and sample exactly
+$n$ games to play from each cell without replacement. This is so that
+we play a balanced number of games. I also randomize the global order
+of all games. Then I do this over again, with a smaller update factor
+$k$, for 20 passes. I don't know whether it's the fact that I reduce
+$k$ over time, or that I end up with effectively 20 times the number
+of games played,\footnote{To be clear, running 20 passes means that
+  games can be reused, which is not ideal.} but doing this
+significantly reduces variance. I run the whole thing 19 times and the
+result is the median Elo score. I also computed the 25\th\ and
+75\th\ percentiles, which were within 1--2\% of the median, which is
+good. The Elo scores appear in Section~\ref{sec:results}; $n > 400$
+for this run.
+
+Ideally, such a system would be robust against adding new players, but
+this is probably not the case; it is easy to see from the results
+(Figure~\ref{fig:heatmap}) that there are certain matchups that favor
+one of the players despite its average skill. During development, I
+often noticed significant swings in Elo scores as players were added,
+especially before there were middle-tier players in the mix. One way
+to deal with this would be to run the Elo simulations multiple times
+while randomly ablating players from the tournament; we could then at
+least estimate the variance in the Elo scores under the removal of
+players, which is like adding players in reverse. I did not implement
+such a thing because of oppressive SIGBOVIK deadlines.
+
+\section{Results} \label{sec:results}
+
+% maybe whole-page?
+\begin{figure}[ht]
+  \includegraphics[width=\linewidth]{elo-heatmap}
+  \caption{ The matrix of outcomes for all players. There is too much
+    data to print, so we just have a bunch of colors, mostly because
+    it looks rad. If you don't have a color printer or TV it will
+    probably not look rad. Rows indicate the player playing as white,
+    from worst (\player{worstfish}, top) to best
+    (\player{stockfish1m}, bottom). Columns for the player with the
+    black pieces, in the same order from left to right. Green
+    indicates a cell that is predominately wins (for white), red is
+    losses, and blue is draws. The right and bottom third of the
+    graphic are all different levels/dilutions of strong engines
+    (Stockfish and Topple). This creates a nice smooth gradient where
+    better engines regularly beat even slightly weaker ones, but start
+    to produce draws at the highest levels. They consistently destroy
+    weak engines; a cell with an $\times$ indicates that it only
+    contained wins or only contained losses (not even draws).
+    \\
+    At the low end, there is a large splat of matchups that mostly
+    produce draws against one another, but can consistently beat
+    the weakest tier. In the top-left corner, a square of blue
+    draws from low-ambition play; these aggressively bad players
+    almost never win games, even when playing each other.
+    \\
+    Microtexture outside of these broad trends comes from matchups
+    where the player is unusually suited or weak against that
+    particular opponent. For example, the bright red cell on the
+    diagonal near the center is \player{cccp} vs. \player{cccp}; this
+    determinisic strategy alwasy wins in self-play as black.
+  } \label{fig:heatmap}
+\end{figure}
+
+The main use of the Elo~World tournament is to benchmark some new
+engine or algorithm for playing chess, particularly if it is not that
+good~\cite{blind}. There are a few ways that we can interpret the
+results:
+
+The {\bf Elo score}, as described.
+
+We can find a {\bf comparable interpolated player}. This is like the
+  Scoville scale for measuring how spicy something is: Some
+  golden-tongued blind tasters are given the pure chili oil and asked
+  to distinguish it from a cup of sugar water, which they of course
+  can do. They then dilute the chili oil 1:1 with sugar water, and try
+  again. The number of dilutions before the testers cannot reliably
+  tell the drink from sugar water yields the score on the Scoville
+  scale. Here, for example, we can say that \player{blind\_spycheck}
+  performs between \player{stockfish1m\_r63488} and
+  \player{\_r61440}, so it is approximately a
+  93.75--96.875\% diluted Stockfish.
+
+We can compute a {\bf Markov probability}. The tournament table can be
+  easily thought of as a Markov transition matrix.
+  Imagine that there is a
+  Champion trophy, held by the player corresponding to some row. Each
+  cell in that row contains the probability that in a game betweeen
+  those two players, the trophy would be passed to that player. We
+  treat draws as choosing one of the two sides by a coin flip (or,
+  equivalently, splitting the trophy in half); and for the games that
+  the row player wins, the trophy is retained (probability mass assigned
+  to a self-transition). It is easy to compute this matrix from the
+  win/loss/draw counts in the tournament table, and it is not sensitive
+  to imbalance like the Elo calculation is.
+  Presented this way, we can compute the stationary distribution (if
+  it exists, which it typically will), which basically tells us that
+  after an extremely large number of games, what is the probability
+  that a given player holds the Champion trophy? This tends to
+  agree with the Elo score, but there are a few places where it does
+  not (e.g. \player{same\_color} has a much higher p(Champion) score
+  than its Elo seems to warrant).
+
+And so finally, a table with numbers:
+\smallskip
+
+% \begin{footnotesize}
+\tablefirsthead{%
+  \hline
+  Player & Elo & p(Champion) \\
+  \hline
+  }
+\tablehead{%
+  \hline
+  \ldots\ Player & Elo & p(Champion) \\
+  \hline
+}
+\tabletail{%
+  \multicolumn{3}{|c|}{\ldots} \\
+  \hline
+}
+\tablelasttail{%
+  \hline
+}
+\begin{supertabular}{|r|l|l|}
+\player{worstfish} & 188.54 & 0.00000071 \\
+\player{pacifist} & 286.90 & 0.00000509 \\
+\player{alphabetical} & 320.91 & 0.00000276 \\
+\player{generous} & 342.71 & 0.00000306 \\
+\player{popular} & 349.48 & 0.00000374 \\
+\player{dangerous} & 349.60 & 0.00000291 \\
+\player{safe} & 350.64 & 0.00000454 \\
+\player{first\_move} & 357.71 & 0.00000434 \\
+\player{rare} & 361.39 & 0.00000323 \\
+\player{no\_i\_insist} & 378.23 & 0.00000244 \\
+\player{huddle} & 399.55 & 0.00000933 \\
+\player{sym\_180} & 429.94 & 0.00000331 \\
+\player{same\_color} & 430.90 & 0.00002590 \\
+\player{opposite\_color} & 432.11 & 0.00000293 \\
+\player{sym\_mirror\_x} & 438.87 & 0.00000305 \\
+\player{survivalist} & 439.03 & 0.00000681 \\
+\player{random\_move} & 439.21 & 0.00000462 \\
+\player{fatalist} & 439.89 & 0.00000314 \\
+\player{sym\_mirror\_y} & 442.50 & 0.00000716 \\
+\player{reverse\_starting} & 445.56 & 0.00000301 \\
+\player{suicide\_king} & 447.02 & 0.00000493 \\
+\player{stockfish1m\_r64512} & 462.82 & 0.00000297 \\
+\player{stockfish1m\_r63488} & 482.44 & 0.00000356 \\
+\player{blind\_yolo} & 488.97 & 0.00000488 \\
+\player{blind\_kings} & 501.98 & 0.00000633 \\
+\player{equalizer} & 504.21 & 0.00000653 \\
+\player{stockfish1m\_r61440} & 521.45 & 0.00001173 \\
+\player{swarm} & 534.03 & 0.00000623 \\
+\player{blind\_spycheck} & 546.91 & 0.00002554 \\
+\player{cccp} & 553.54 & 0.00000544 \\
+\player{min\_oppt\_moves} & 597.04 & 0.00001076 \\
+\player{stockfish1m\_r57344} & 600.54 & 0.00001112 \\
+\player{stockfish1m\_r49152} & 752.22 & 0.00000737 \\
+\player{chessmaster.nes\_lv1} & 776.15 & 0.00002055 \\
+\player{stockfish1m\_r32768} & 976.20 & 0.00003025 \\
+\player{chessmaster.nes\_lv2} & 989.21 & 0.00012094 \\
+\player{stockfish1m\_r16384} & 1277.73 & 0.00012509 \\
+\player{stockfish0} & 1335.69 & 0.00008494 \\
+\player{stockfish5} & 1644.63 & 0.00070207 \\
+\player{stockfish1m\_r8192} & 1690.15 & 0.00152052 \\
+\player{topple10k} & 1771.65 & 0.00179298 \\
+\player{stockfish10} & 1915.23 & 0.00257648 \\
+\player{stockfish15} & 1952.93 & 0.00310436 \\
+\player{stockfish1m\_r4096} & 2020.25 & 0.00886928 \\
+\player{stockfish20} & 2139.47 & 0.00721173 \\
+\player{topple1m} & 2218.71 & 0.01091286 \\
+\player{stockfish1m\_r2048} & 2261.50 & 0.03132272 \\
+\player{stockfish1m\_r1024} & 2425.72 & 0.06759788 \\
+\player{stockfish1m\_r512} & 2521.78 & 0.11122236 \\
+\player{stockfish1m\_r256} & 2581.97 & 0.15364889 \\
+\player{stockfish1m\_r128} & 2609.45 & 0.17861429 \\
+\player{stockfish1m\_r64} & 2637.78 & 0.20508090 \\
+\player{stockfish1m} & 2644.10 & 0.21523142 \\
+\end{supertabular}
+% \end{footnotesize}
+
+%  % miraculous win for "generous" strategy as black:
+%  % playing against "suicide king". White forces black
+%  % to mate him:
+%  1. a4 b5 2. d3 Nh6 3. Kd2 Ng8 4. Kc3 Nh6
+%  5. Kd4 c6 6. Ke5 f5 7. a5 Qb6 8. Bg5 Qe3+
+%  9. Bxe3 Na6 10. f3 Nc5 11. Ra2 Ne4 12. f4 Nd2
+%  13. Bf2 Nf3+ 14. exf3 Ng4+ 15. Kxf5 g5 16. Nh3 e5
+%  17. Bh4 Ba3 18. b4 e4 19. Nf2 c5 20. c4 Ba6
+%  21. g3 Rc8 22. Be2 Kf7 23. fxe4 d5 24. Qc1 dxe4
+%  25. Rg1 bxc4 26. Rd1 cxd3 27. Qxc5 Rhf8 28. Nd2 Bc4
+%  29. Kxg5 Rc7 30. Bf3 Ke8 31. Qe5+ Kf7 32. Qe6+ Kg7
+%  33. Qb6 Rxf4 34. Nb3 Bb5 35. Qb7 d2 36. Nh3 Bc1
+%  37. Qxa7 Be2 38. Nd4 h5 39. Ra1 Nf2 40. Raxc1 Nh1
+%  41. Rf1 d1=Q 42. Qxc7+ Kh8 43. Kg6 Bc4 44. Bxd1 e3
+%  45. Qf7 Bb3 46. Rf3 Nf2 47. Rc6 e2 48. Rc7 Rf5
+%  49. Rc4 Rf4 50. Nb5 e1=Q 51. Rc2 Qc3 52. g4 Rf5
+%  53. Ng1 Qc7 54. Rcc3 Nd3 55. Rc5 Nf2 56. Nc3 Ba2
+%  57. Qh7+
+%
+
+% Good site for some computer chess programs with ELO:
+% http://www.computerchess.org.uk/ccrl/404/
+
+\section{Conclusion}
+
+Shout out to the Thursd'z Institute and anonymous commenters
+on my blog for discussion of players and suggestions. Several
+ideas were suggested by multiple people, increasing my confidence
+that they are somehow canonical.
+
+The author would also like to thank the anonymous referees
+of the Special Interest Group on
+Bafflingly
+Overdone
+Ventures
+In
+Chess journal for their careful reviews.
 
@@
 
-Interpolation, like Scoville scale.
-
-% miraculous win for "generous" strategy as black:
-% playing against "suicide king". White forces black
-% to mate him:
-1. a4 b5 2. d3 Nh6 3. Kd2 Ng8 4. Kc3 Nh6
-5. Kd4 c6 6. Ke5 f5 7. a5 Qb6 8. Bg5 Qe3+
-9. Bxe3 Na6 10. f3 Nc5 11. Ra2 Ne4 12. f4 Nd2
-13. Bf2 Nf3+ 14. exf3 Ng4+ 15. Kxf5 g5 16. Nh3 e5
-17. Bh4 Ba3 18. b4 e4 19. Nf2 c5 20. c4 Ba6
-21. g3 Rc8 22. Be2 Kf7 23. fxe4 d5 24. Qc1 dxe4
-25. Rg1 bxc4 26. Rd1 cxd3 27. Qxc5 Rhf8 28. Nd2 Bc4
-29. Kxg5 Rc7 30. Bf3 Ke8 31. Qe5+ Kf7 32. Qe6+ Kg7
-33. Qb6 Rxf4 34. Nb3 Bb5 35. Qb7 d2 36. Nh3 Bc1
-37. Qxa7 Be2 38. Nd4 h5 39. Ra1 Nf2 40. Raxc1 Nh1
-41. Rf1 d1=Q 42. Qxc7+ Kh8 43. Kg6 Bc4 44. Bxd1 e3
-45. Qf7 Bb3 46. Rf3 Nf2 47. Rc6 e2 48. Rc7 Rf5
-49. Rc4 Rf4 50. Nb5 e1=Q 51. Rc2 Qc3 52. g4 Rf5
-53. Ng1 Qc7 54. Rcc3 Nd3 55. Rc5 Nf2 56. Nc3 Ba2
-57. Qh7+
-
-% Fate-based algorithms.
-% The deterministic version is fairly boring, since the
-% highest probability states are similar to the starting
-% position (this might not be true for "survivalist"
-% or "fatalist"? check?)
-
-% weighted random is almost the same as random, because
-% the weights tend to be very similar when treated
-% additively. They perform slightly different from
-% random, as so (101.5 million games):
-
-%  player	elo	w/l/d
-%  dangerous	986.29	1374577/1501526/15683897
-%  popular	997.34	1360053/1502995/15696952
-%  random_move	1000.79	1540798/1337668/15681534
-%  rare	1005.85	1554600/1326707/15678693
-%  safe	1009.72	1341030/1502162/15716808
-
-% then with normalization:
-%  player	elo	w/l/d
-%  popular	990.93	25756/43561/346683
-%  safe	993.49	28082/40769/347149
-%  dangerous	996.60	30063/37451/348486
-%  rare	1005.82	50656/27390/337954
-%  random_move	1013.15	41747/27133/347120
-
-// Good site for some computer chess programs with ELO:
-// http://www.computerchess.org.uk/ccrl/404/
-
-\section{Conclusion}
-
-Shout out to the Thursd'z Institute and anonymous commenters
-on my blog for discussion of players and suggestions. Several
-ideas were suggested by multiple people, increasing my confidence
-that they are somehow canonical.
-
-The author would like to thank the anonymous referees
-of the Special Interest Group on
-B
-O
-V
-In
-Chess journal for their careful reviews.
-
-\nocite{elo1978rating}
-\nocite{topple}
-// citation for knights distance:
-\nocite{miller2013counting}
-\nocite{watkins2017losing}
-
 % \balance

### 2019-03-21 trunk/nand/board/rom/main.c
Successfully programmed the board

@@ -1,0 +1,78 @@
+#include "stm32f30x.h"
+
+/* zyp's cycle count delay routines */
+volatile uint32_t* demcr = (uint32_t*)0xE000EDFC;
+volatile uint32_t* dwt_ctrl = (uint32_t*)0xe0001000;
+volatile uint32_t* dwt_cyccnt = (uint32_t*)0xe0001004;
+
+void cyccnt_enable() {
+  *demcr |= (1<<24);
+  *dwt_ctrl |= 1;
+}
+
+void cyclesleep(uint32_t cycles) {
+  uint32_t start = *dwt_cyccnt;
+
+  while (*dwt_cyccnt - start < cycles) {}
+}
+
+// see _gpio.c for light docs
+static void InitPinOutB(uint16_t pins) {
+  GPIO_InitTypeDef gpio;
+
+  gpio.GPIO_Pin = pins;
+  gpio.GPIO_Mode = GPIO_Mode_OUT;
+  // For this purpose, extreme speeds are not necessary and
+  // just increase noise.
+  gpio.GPIO_Speed = GPIO_Speed_2MHz;
+  // gpio.GPIO_Speed = GPIO_Speed_50MHz;
+  // We want push-pull since we connect these directly to
+  // inputs.
+  gpio.GPIO_OType = GPIO_OType_PP;
+  // I think pull-up/pull-down doesn't make sense when using
+  // push/pull.
+  gpio.GPIO_PuPd = GPIO_PuPd_NOPULL;
+  GPIO_Init(GPIOB, &gpio);
+}
+
+int main(void) {
+  // Here clock literally means the oscillator that drives
+  // this part of the chip, in other words, "turn on the GPIO B
+  // peripheral so that we can use it."
+  RCC_AHBPeriphClockCmd(RCC_AHBPeriph_GPIOB, ENABLE);
+
+  /* Enable PB13,14,15 for output */
+  InitPinOutB(GPIO_Pin_13 | GPIO_Pin_14 | GPIO_Pin_15);
+
+  GPIOB->BSRR = (1<<13) | (1<<14) | (1<<15);
+
+  /* start cycle counter */
+  cyccnt_enable();
+
+  /* loop forever */
+  for (;;) {
+
+    GPIOB->BRR = 1<<13;
+    GPIOB->BSRR = 1<<14;
+
+    cyclesleep(0x084aa20);
+
+    GPIOB->BRR = 1<<14;
[…]

### 2019-03-21 nand/paper.tex
Successfully programmed the board

--- 2019-01-27 nand/paper.tex
@@
-\documentclass[twocolumn]{article}
+\documentclass[twocolumn,cm]{article}
 \usepackage[top=0.5in, left=0.45in, right=0.45in, bottom=0.5in]{geometry}
@@
 \usepackage{url}
-% \usepackage{code}
+\usepackage{code}
 % \usepackage{cite}
@@
 
+\newcommand\ieee[2]{[{\it IEEE 754-2008}, #1, p#2]}
+
 % flInf flops?
@@
 Foundation. Appears in SIGBOVIK 2019 with the
-XXX JOKE
+half precision
 of the Association for Computational Heresy; 
@@
 \begin{abstract}
-what do I put here
+Yes, this paper contains many layers of abstraction.
 \end{abstract}
@@
 programming sessions) % XXX cite skymall etc?
-that are not integer-based.
+that are not integer-based, and thus cannot be replaced with real numbers
+via this techique.
 
 In this paper I give a new foundation for computing based solely on
-real numbers. This ...
-
-I begin with a brief reminder of the definition of real numbers,
-although the reader is expected to be familiar as these are pretty
-fundamental to everything. The approach of the paper is then to
+real numbers. I begin with a brief reminder of the definition of real
+numbers, although the reader is expected to be familiar as these are
+pretty fundamental to everything. The approach of the paper is then to
 identify a pair of real numbers that have nice properties
@@
 typically used in the construction of computers
-(Section~\ref{sec:logical}). I then discuss how these operations
-can be implemented efficiently (Section~\ref{sec:representations}).
-I conclude with some wild speculation.
+(Section~\ref{sec:logical}). I then discuss how these operations can
+be implemented efficiently (Section~\ref{sec:hardware}). I
+conclude with some wild speculation.
 
@@
 you count very high or very low. Excitingly, we also have both
-positive and negative versions of 0.
+positive and negative versions of 0. Some numbers have multiple
+representations, and almost all numbers cannot be represented at all.
 
 The real numbers have an equality operation {\tt ==}. This operation
-has some very exciting properties: It is not reflexive (\nan\ {\tt ==}
-\nan\ is false), and does not obey substitution (for $+0$ {\tt ==} $-0$
-is true, but $\sfrac{1}{+0}$ {\tt ==} $\sfrac{1}{-0}$ is false).
+has some very exciting properties which are unusual for an equivalence
+relation: It is not reflexive (\nan\ {\tt ==} \nan\ is false), and
+does not obey substitution (for $+0$ {\tt ==} $-0$ is true, but
+$\sfrac{1}{+0}$ {\tt ==} $\sfrac{1}{-0}$ is false).
 
-Real numbers are an absolute joy to work with.
+As a result, real numbers are an absolute joy to work with.
 
@@
 
-use nan and inf because other real numbers can be ``easily confused''
-with one another. And do we really want ordering relationships on
-our fundamental particles, like $0 < 1$?
+Computing will need at least two different values. We could choose
+$0.0$ and $1.0$ as in ``binary,'' but these numbers are extremely
+arbitrary; why not $1.0$ and $2.0$? $e$ and $\sfrac{\pi}{2}$? These
+numbers are easily confused with one another. It seems better to use
+distinguished values, making the resulting mathematics more
+distinguished. One of the most distinguished numbers is
+\nan\ (Figure~\ref{fig:gentlenan}). One nice thing about using the
+number \nan\ is that it is not comparable to other numbers, e.g. both
+$\nan < 0.0$ and $0.0 < \nan$ are false. Does it really make sense for
+our fundamental particles to be ordered (e.g. $0 < 1$)? The lack of
+symmetry is abhorrent.
+
+\begin{figure}[ht]
+  \begin{center}
+  \includegraphics[width=0.55 \linewidth]{gentlenan}
+  \caption{A distinguished gentleNaN.} \label{fig:gentlenan}
+  \end{center}
+\end{figure}
+
+The two numbers we choose need to be different; alas they cannot both
+be \nan, since although \nan\ is different from \nan\ (\nan\ {\tt !=}
+\nan), it is not possible to tell them apart (except that
+\nan\ actually has multiple binary
+representations---see Section~\ref{sec:representation}). Another great choice is
+$+\inf$ or $-\inf$. We choose to use $+\inf$ in order to break
+symmetry, and because it will make our scientific contribution more
+positive.
+% So the two values will be \nan\ and \inf; it may
+% be helpful to think of these as false and true, (or 0 and 1), respectively.
 
@@
 
-\nocite{fluint8}
+\subsection{IEEEuler's Identity}
 
-ha okay now there is just notes
+Moreover, \nan\ and \inf\ are part of the pantheon of special values, exhibiting
+exquisite properties, such as IEEEuler's identity:
 
-Some useful primitives:
+\begin{huge}
+  \[
+  e^{i \pi} + 1^{\nan \times \inf} = 0
+  \]
+\end{huge}
 
-        These are called minNum and maxNum in IEE 754-2008 (5.3.1, p19)
+because $1^n$ is $1$\footnote{This paper uses both exponents and footnotes
+  extensively; please be careful of the difference.}, even for $n=\nan$.\footnote{\ieee{9.2.1}{44}} Another nice
+pair of properties ties these fundamental constants together a
+different way:
 
-        "If exactly one argument is NaN, they return the other.
-        If both are NaN they return NaN.
+\begin{huge}
+  \[
+  {(e^{i \pi})}^{-\inf} = {\tt compound}(\nan, 0)
+  \]
+\end{huge}
 
-        NAND  0 1     MAX  nan inf
-             +---         +-------
-           0 |1 1     nan |nan inf
-           1 |1 0     inf |inf inf
+${\tt compound}(x, n)$ is the ``compound interest'' function
+$(1 + x)^n$, defined in the IEEE 754 standard, but only available
+in C via floating point extensions~\cite{fpextensions}. This
+function is $1$ for $n=0$ and $x=\nan$.\footnote{\ieee{9.2.1}{44}}
+More excitingly, we have $e^{i \pi} = -1$ (Euler) and ${-1}^{-\inf} =
+1$ ``because all large positive floating-point values are even
+integers.''~\cite{crationale510}
 
-                    HYPOT  nan inf
-                          +-------
-                      nan |nan inf
-                      inf |inf inf
+\subsection{NaN's Not GNU} \label{sec:logical}
 
-        So if 0 = nan and 1 = inf, then MAX is OR.
+People who work with real numbers are often taught that the number
+\nan\ is propagated through all expressions that use it (e.g. $\nan -
+1 = \nan$), like some kind of GNU Public Licensed number. This is a
+misconception. We already saw in the beautiful identities above that
+some expressions involving \nan\ do not result in \nan, like $1^{\nan}
+= 1$ and ${\tt compound}(\nan, 0) = 0$. But it is also the case that
+$1^{\inf} = 1$ and ${\tt compound}(\inf, 0) = 0$. Are there mathematical
+functions that distinguish between \nan\ and \inf?
 
-        We also have AND, since only inf*inf (aka 1 AND 1)
-        returns inf in that truth table.
+It turns out that there are! For example, the functions {\tt minNum}
+and {\tt maxNum} (\ieee{5.3.1}{19}) take two arguments and return the
+min and max, respectively. They have the special, distinguished property
+that ``if exactly one argument is \nan, they return the other. If
+both are \nan\ they return \nan.''
 
-        Is AND and OR together complete? (I don't think so?)
-        Do we have NOT?
+\begin{figure}[ht]
+\renewcommand*{\arraystretch}{1.4}
+  \begin{tabular}{cc}
+  \begin{tabular}{r|cc}
+    \multicolumn{1}{r}{${\tt maxNum}(a,b)$} & \multicolumn{1}{c}{\nan} & \inf \\[0.2em]\cline{2-3}
+    \nan & \nan & \inf \\
+    \inf & \inf & \inf \\
+  \end{tabular} &
+  \begin{tabular}{r|cc}
+    \multicolumn{1}{r}{$a * b$} & \multicolumn{1}{c}{\nan} & \inf \\[0.2em]\cline{2-3}
+    \nan & \nan & \nan \\
+    \inf & \nan & \inf \\
+  \end{tabular}
+  \end{tabular}
+  \caption{
+    The behavior of some mathematical functions on our distinguished values \nan\ and \inf.
+    {\tt maxNum} returns \inf\ if either of its arguments is \inf\ (some other functions have
+    this property, like {\tt hypot}). $a * b$ is \inf\ only
+    if both of its arguments are \inf\ (there are many other examples, like $a + b$).
+  } \label{fig:truthtables}
+\end{figure}
 
-YES!
+With functions such as these, we can begin constructing the building blocks
+of more interesting functions (Figure~\ref{fig:truthtables}). Unfortunately,
+${\tt maxNum}(a,b)$ and $a*b$ are not complete on their own; we additionally
+need at least a function $f(x)$ where $f(\nan) = \inf$ and $f(\inf) = \nan$.
+Does such a function exist? Yes! Several can be built from IEEE 754 primitives:
+\[
+\begin{array}{rcl}
+  f(x) & = & {\tt minNum}(-x, -1.0) + \inf \\
+  f(x) & = & {\tt hypot}(\nan, {\tt maxNum}(1/x, -\inf)) \\
+  f(x) & = & \inf - {\tt maxNum}(x, 1.0) \\
+  f(x) & = & {\tt sqrt}({\tt copysign}(\inf, -x)) \\
+\end{array}
+\]
+You can try these out in your favorite programming language, and if they don't
+work, your implementation is not IEEE 754 compliant. Why do these work? Let's
+take the first one, and compare \nan\ and \inf\ for $x$:
+\[
+\begin{array}{l|r}
+  \hline
+  x = \nan                             & x = \inf \\
+  \hline
+  {\tt minNum}(-x, -1.0) + \inf    & {\tt minNum}(-x, -1.0) + \inf \\
+  {\tt minNum}(-\nan, -1.0) + \inf & {\tt minNum}(-\inf, -1.0) + \inf \\
+      -1.0 + \inf                      & -\inf + \inf \\
+      \inf                             & \nan \\
+\end{array}
+\]
 
-Here are some:
-Min(-x, -1.0) + inf
-hypot(nan, max(1/x, -inf))
-inf - max(x, 1)
-sqrt(copysign(inf, -x))
+Thinking of \nan\ as {\tt false} and \inf\ as {\tt true}, we now have {\tt AND} ({\tt maxNum}),
+{\tt OR} ($*$), and {\tt NOT} (${\tt minNum}(-x, -1.0)$). With these we can create arbitrary
+functions $f(a_1, a_2, \ldots, a_n)$ that return our choice of $\nan$ or $\inf$ for the $2^n$
+different combinations of arguments. It is also possible to find more direct expressions that
+compute simple functions (Figure~\ref{fig:truthtables2}).
 
-Now direct search for some gates:
-NAND: inf - max(x + y, -inf)
-AND: x + y
-OR: hypot(x, y)
-XOR: Abs(max(y, -inf) - max(x, -inf))
-     Abs(min(y, -x) + max(y, -inf))
-     (Wouldn't be surprised if there's a simpler one,
-     but my search does not find it before it gets
-     out of hand...)
+\begin{figure}[ht]
+\renewcommand*{\arraystretch}{1.4}
 
-NOR: -inf / max(y, max(x, -1))
+  \begin{tabular}{r|cc}
+    \multicolumn{1}{r}{$\inf - {\tt maxNum}(a + b, -\inf)$} & \multicolumn{1}{c}{\nan} & \inf \\[0.2em]\cline{2-3}
+    \nan & \inf & \inf \\
+    \inf & \inf & \nan \\
+    \multicolumn{3}{c}{}\\[0.1em]
+    \multicolumn{1}{r}{${\tt abs}({\tt minNum}(b, -a) + {\tt maxNum}(b, -\inf))$} & \multicolumn{1}{c}{\nan} & \inf \\[0.2em]\cline{2-3}
+    \nan & \nan & \inf \\
+    \inf & \inf & \nan \\
+    \multicolumn{3}{c}{}\\[0.1em]
+    \multicolumn{1}{r}{$-\inf / {\tt maxNum}(b, {\tt maxNum}(a, -1))$} & \multicolumn{1}{c}{\nan} & \inf \\[0.2em]\cline{2-3}
+    \nan & \inf & \nan \\
+    \inf & \nan & \nan \\
+  \end{tabular}
 
-And how about if we leave out min, max? They are certainly
-reasonable since they are defined in IEEE 754, but perhaps
-not the best joke since they are not as math-y as e.g. hypot
+  \caption{
+    Some interesting functions of two variables. They are isomorphic to the boolean functions {\tt NAND},
+    {\tt XOR} and {\tt NOR} respectively, but more beautiful.
+  } \label{fig:truthtables2}
+\end{figure}
 
-NOR: sqrt(copysign(-inf, -hypot(y, x)))
-(Here, hypot is just computing OR.)
+\begin{figure*}[tp]
+\begin{tabular}{|l|c|c|c|c|c|c|}
+  \hline
+  parameter & {\bf binary4} & binary16 & binary32 & binary64 & binary128 & binary$_k$ \\
+  \hline
+  $k$, storage in bits & {\bf 4} & 16 & 32 & 64 & 128 & multiple of 32 \\
+  $p$, precision in bits & {\bf 2} & 11 & 24 & 53 & 113 & $k$ - round(4 * log$_2$(k)) + 13 \\
+  $emax$, maximum exponent e & {\bf 1} & 15 & 127 & 1023 & 16383 & $2^{k-p-1}$ - 1 \\
+  $bias$ = $E$ - $e$ & {\bf 1} & 15 & 127 & 1023 & 16383 & emax \\
+  $sign bits$ & {\bf 1} & 1 & 1 & 1 & 1 & 1 \\
+  $w$, exponent width & {\bf 2} & 5 & 8 & 11 & 15 & round(4 * log$_2$(k)) - 13 \\
+  $t$, trailing significand width & {\bf 1} & 10 & 23 & 52 & 112 & k - w - 1 \\
+  $k$, storage width & {\bf 4} & 16 & 32 & 64 & 128 & 1 + w + t \\
+\hline
+\end{tabular}
+\caption{Parameters for the newly-introduced {\bf binary4} encoding
+  for IEEE 754, compared to the standard widths (see Table~3.5 in the
+  standard\cite{ieee754}). } \label{fig:binary4}
+\end{figure*}
 
-So another NOT is:
-sqrt(copysign(-inf, -x))  (actually can be inf)
+I found these functions through computer search,\footnote{
+  Source code is available at \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/nand/}
+  } using a technique
+like bottom-up logic programming~\cite{pfenning2006logic}. I start
+with a small set of constants, including arguments {\sf a} and {\sf
+  b}, and then compute all of the expressions that can be made by
+applying a single mathematical function (e.g. ${\tt abs}(x)$, $-x$) or
+binary mathematical function ($x + y$, $x / y$, ${\tt maxNum}(x, y)$)
+to existing expressions. The expression is actually a collection of
+values taken on for each possible substitution (in $\{\nan, \inf\}$)
+to arguments {\sf a} and {\sf b} (i.e., it represents a function). If
+the expression has the correct values for each possible assignment to
+the arguments, then we are done. We only need to keep one (the
+smallest) expression that represents a distinct function, but note
+that we have to consider intermediate expressions that compute values
+other than \nan\ and \inf. Also note that we need one of {\tt minNum},
+{\tt maxNum} or {\tt copySign} in order to compute the {\tt NOT}
+function; we could think of these functions as therefore essential
+to mathematical completeness.
+% TODO: Could show all 36 functions without these..
 
-XOR: abs(copysign(-inf, y) - copysign(-inf, x))
-OR: hypot(y, x)
-AND: x + y
-NAND: inf - copysign(-inf, x + y)
+Particularly nice is $\inf - {\tt maxNum}(a + b, -\inf)$, which
+returns $\inf$ if either of its arguments is $\nan$. We will call this
+the ``{\tt NAN} gate'', for ``Not \nan''. The {\tt NAN} gate is
+exciting because it can be used on its own to construct all other
+boolean functions! We can use \nan, \inf, and this function to
+construct any computer and any computable function. Beautiful!
 
-ok cool, but copysign kinda has the same problems as min,max.
-maybe without that?
+To program with numbers on computers, the real numbers are
+represented as strings of bits. Next we'll talk about efficient
+representations that allow us to manipulate \nan\ and \inf\ 
+with {\tt NAN} gates.
 
-the database saturates with 36 entries if we remove
-copysign and min, max. They are these:
+\section{The binary4 representation} \label{sec:representation}
 
-[nan, nan, nan, nan] = NaN
-[nan, nan, nan, ~inf] = Minus(Neg(y), x)
-[nan, nan, nan, ~2.0] = Plus(Tanh(Minus(Neg(y), x)), -1)
-[nan, nan, nan, ~1.0] = Tanh(Minus(Neg(y), x))
-[nan, nan, nan, 0.0] = Pow(y, Neg(x))
-[nan, nan, nan, 1.0] = Exp(Pow(y, Neg(x)))
-[nan, nan, nan, 2.0] = Plus(Exp(Pow(y, Neg(x))), 1)
-[nan, nan, nan, inf] = Plus(y, x)
-[nan, nan, ~inf, ~inf] = Neg(y)
-[nan, nan, ~2.0, ~2.0] = Plus(Tanh(Neg(y)), -1)
-[nan, nan, ~1.0, ~1.0] = Tanh(Neg(y))
-[nan, nan, ~0.0, ~0.0] = Div(-1, y)
-[nan, nan, 1.0, 1.0] = Exp(Div(-1, y))
-[nan, nan, 2.0, 2.0] = Minus(1, Tanh(Neg(y)))
-[nan, nan, inf, inf] = y
-[nan, ~inf, nan, ~inf] = Neg(x)
-[nan, ~inf, ~inf, ~inf] = Neg(Hypot(y, x))
-[nan, ~2.0, nan, ~2.0] = Plus(Tanh(Neg(x)), -1)
-[nan, ~2.0, ~2.0, ~2.0] = Plus(Tanh(Neg(Hypot(y, x))), -1)
-[nan, ~1.0, nan, ~1.0] = Tanh(Neg(x))
-[nan, ~1.0, ~1.0, ~1.0] = Tanh(Neg(Hypot(y, x)))
-[nan, ~0.0, nan, ~0.0] = Div(-1, x)
-[nan, ~0.0, ~0.0, ~0.0] = Div(-1, Hypot(y, x))
-[nan, 1.0, nan, 1.0] = Exp(Div(-1, x))
-[nan, 1.0, 1.0, 1.0] = Exp(Div(-1, Hypot(y, x)))
-[nan, 2.0, nan, 2.0] = Minus(1, Tanh(Neg(x)))
-[nan, 2.0, 2.0, 2.0] = Minus(1, Tanh(Neg(Hypot(y, x))))
-[nan, inf, nan, inf] = x
-[nan, inf, inf, inf] = Hypot(y, x)
-[~inf, ~inf, ~inf, ~inf] = NegInf
-[~2.0, ~2.0, ~2.0, ~2.0] = Plus(-1, -1)
-[~1.0, ~1.0, ~1.0, ~1.0] = -1
-[0.0, 0.0, 0.0, 0.0] = 0
-[1.0, 1.0, 1.0, 1.0] = 1
-[2.0, 2.0, 2.0, 2.0] = Plus(1, 1)
-[inf, inf, inf, inf] = Inf
+% bias = 1
 
-IEEE 754, 3.3
-minimal encoding size:
-b, radix, 2
-p, precision, ?
-emax, maximum exponent, ?
-emin = always 1 - emax
-
-I deduce:
-p = t + 1
-    (because you get one implied bit by normalization)
-t = 1
-% using the table
-emax = $2^{k-p-1} - 1$, which is $2^{4-2-1}-1 = 2^1-1 = 1$
-bias = emax
-emin = 1 - emax = 0
-
-I think there's still some ambiguity about whether we have
-emax = 1, emin = 0,
-or
-emax = 0, emin = -1,
-probably both work, but I think the former has more support
-in the docs.
-
-\comment{
-a normal value is
-(-1)^sign x b^{exp} x significand
-(I think significand = mantissa?)
-
-by standard, need:
-- sign bit
-- exponent could maybe be empty, though it's sort of implied
-that you just have emax=0, emin= -1, so one bit
-- m could probably be empty (no bits mantissa = 0)
-
-DID YOU KNOW?
-``every finite floating point numbe is an integral multiple
-of the smallest subnormal magnitude b^{emin} * b^{1 - p}''
-
-fewest exponent bits w?
-The range of the encoding's biased exponent E shall include:
-- every integer between 1 and 2^{w}−2, inclusive, to encode normal numbers
-- the reserved value 0 to encode +/-0 and subnormal numbers
-- the reserved value 2^{w}−1 to encode +/-inf and NaNs
-
-so we need at least 0,
-and a reserved value for inf & nan,
-so if w = 1, we'd have 0 and 2^1-1 = 1,
-but this seems to violate the spec because you need
-``every integer between 1 and 2^w-2 inclusive'', and here 2^w-2 is 0,
-so either this interval is inverted (arguable?) and that's why it
-doesn't include its endpoints, or it's in violation
-but with w = 2,
-
-2^2 - 1 = 0b11 = 3 (inf, nan)
-1 to 2^2 - 2 = 0b01, 0b10 = 1,2 (normals)
-0 = 0b00 = 0 (zero, subnormals)
-
-(Note this is the biased exponent E, not e)
-
-2 bits looks clearly legal.
-
- s       w bits           t bits
-[sign][ exponent ][  trailing significand T   ]
-
-when e=0b11, t determines inf or nan.
-- nan is any nonzero sign is ignored.
-(recall, nans can carry info, so that is why they get
-the nonzero case)
-- inf is zero
-
-so we clearly need 1 t-bit to encode inf vs nan,
-but it seems that this is enough
-
-so probable case is four bits:
-
- - sign
- - two exponent bits
- - one mantissa bit
-
-The sign is always 0 (positive infinity; doesn't matter for nan)
-and the exponent is always 11 (since we always encode nan or
-positive inf). Unless we do any intermediate calculations
-(could check what all the intermediate values are in the above
-expressions and make sure they fit in 4 bits), a hardware
-realization of this could just hard-wire 3 of 4 bits? Is that
-better or worse?
-
-so another version of this joke is that a NAND gate is a super
-fast FPU for calculating inf - max(x + y, -inf) on four-bit
-IEEE 754 where the top 3 MSBs are fixed (although be careful,
-nan = 1 and inf = 0 in this representation. Is that maybe a
-better choice for this presentation? it just dualizes some
-gates)
-
-}
-
-\section{The binary4 representation}
+\begin{figure}[ht]
+\begin{center}
+  \begin{tabular}{|l@{\,}c@{\,}l|p{2.5in}|}
+\hline
+  $s$ & $E$ & $T$ & value \\
+  \hline
+0 & 00 & 0 &   $+0$ \\
+0 & 00 & 1 &    subnormal: % $2^{emin} * 2^{1-p} * T$ =
+                $2^0 * 2^{1-2} * 1$ =
+                $1 * \sfrac{1}{2} * 1$ = 0.5 \\
+0 & 01 & 0 &    normal: % $2^{E - bias} * (1 + 2^{1-p} * T)$
+                $2^0 * (1 + \sfrac{1}{2} * 0)$ = 1 \\
+0 & 01 & 1 &    $2^0 * (1 + \sfrac{1}{2} * 1)$ = 1.5 \\
+0 & 10 & 0 &    $2^1 * (1 + \sfrac{1}{2} * 0)$ = 2 \\
+0 & 10 & 1 &    $2^1 * (1 + \sfrac{1}{2} * 1)$ = 3 \\
+0 & 11 & 0 &   $+\inf$ \\
+0 & 11 & 1 &    \nan \\
+1 & 00 & 0 &   $-0$ \\
+1 & 00 & 1 &   $-0.5$ \\
+1 & 01 & 0 &   $-1$ \\ 
+1 & 01 & 1 &   $-1.5$ \\
+1 & 10 & 0 &   $-2$ \\
+1 & 10 & 1 &   $-3$ \\ 
+1 & 11 & 0 &   $-\inf$ \\
+1 & 11 & 1 &    \nan \\
+\hline
+\end{tabular}
+\end{center}
+  \caption{All 16 values representable in binary4 floating-point.
+  The format works reasonably well even at this very low precision,
+  although note how many of the values are not finite.} \label{fig:allvalues4}
+\end{figure}
 
@@
 half-precision 16-bit floats are common in GPU code for machine
-learning applications (XXX cite). Smaller floats sacrifice precision,
+learning applications~\cite{wikipedia16}. Smaller floats sacrifice precision,
 but require less space and allow faster calculations. For our purposes
-in this paper, since we only need to represent the two values \nan\ and
+in this paper, since we only need to represent the two numbers \nan\ and
 \inf, we are interested in the smallest possible representation.
-
-% TODO: what if each bit in the float is itself represented as
-% inf or NaN?
 
@@
 
-The most stringent constraint on $w$ comes in IEEE-754-2008~3.4, which
+The most stringent constraint on $w$ comes in \ieee{3.4}{9}, which
 states
 \begin{quote}
-  The range of the encoding's biased exponent E shall include:
+  The range of the encoding's biased exponent $E$ shall include:
   \begin{enumerate}[label=---]
@@
 
-\begin{figure*}[tp]
-\begin{tabular}{|l|c|c|c|c|c|c|}
-  \hline
-  parameter & {\bf binary4} & binary16 & binary32 & binary64 & binary128 & binary$_k$ \\
-  \hline
-  $k$, storage in bits & {\bf 4} & 16 & 32 & 64 & 128 & multiple of 32 \\
-  $p$, precision in bits & {\bf 2} & 11 & 24 & 53 & 113 & $k$ - round(4 * log$_2$(k)) + 13 \\
-  $emax$, maximum exponent e & {\bf 1} & 15 & 127 & 1023 & 16383 & $2^{k-p-1}$ - 1 \\
-  $bias$ = $E$ - $e$ & {\bf 1} & 15 & 127 & 1023 & 16383 & emax \\
-  $sign bits$ & {\bf 1} & 1 & 1 & 1 & 1 & 1 \\
-  $w$, exponent width & {\bf 2} & 5 & 8 & 11 & 15 & round(4 * log$_2$(k)) - 13 \\
-  $t$, trailing significand width & {\bf 1} & 10 & 23 & 52 & 112 & k - w - 1 \\
-  $k$, storage width & {\bf 4} & 16 & 32 & 64 & 128 & 1 + w + t \\
-\hline
-\end{tabular}
-\caption{Parameters for the newly-introduced {\bf binary4} encoding
-  for IEEE-754, compared to the standard widths (see Table~3.5 in the
-  standard\cite{ieee754}). } \label{fig:binary4}
-\end{figure*}
+Four bits is not many, but is it possible to represent these two
+values more efficiently?
 
-% bias = 1
+\subsection{The hypothesized binary3 format} \label{sec:binary3}
 
-\begin{figure}[h]
+\begin{figure}[ht]
+\begin{center}
 \begin{tabular}{|l@{\,}c@{\,}l|p{2.5in}|}
@@
   \hline
-0 & 00 & 0 &   $+0$ \\
-0 & 00 & 1 &    subnormal: % $2^{emin} * 2^{1-p} * T$ =
-                $2^0 * 2^{1-2} * 1$ =
-                $1 * \sfrac{1}{2} * 1$ = 0.5 \\
-0 & 01 & 0 &    normal: % $2^{E - bias} * (1 + 2^{1-p} * T)$
-                $2^0 * (1 + \sfrac{1}{2} * 0)$ = 1 \\
-0 & 01 & 1 &    $2^0 * (1 + \sfrac{1}{2} * 1)$ = 1.5 \\
-0 & 10 & 0 &    $2^1 * (1 + \sfrac{1}{2} * 0)$ = 2 \\
-0 & 10 & 1 &    $2^1 * (1 + \sfrac{1}{2} * 1)$ = 3 \\
-0 & 11 & 0 &   $+\infty$ \\
-0 & 11 & 1 &    \nan \\
-1 & 00 & 0 &   $-0$ \\
-1 & 00 & 1 &   $-0.5$ \\
-1 & 01 & 0 &   $-1$ \\ 
-1 & 01 & 1 &   $-1.5$ \\
-1 & 10 & 0 &   $-2$ \\
-1 & 10 & 1 &   $-3$ \\ 
-1 & 11 & 0 &   $-\infty$ \\
-1 & 11 & 1 &    \nan \\
+0 & 0 & 0 &   $+0$ \\
+0 & 0 & 1 &    subnormal: % $2^{emin} * 2^{1-p} * T$ =
+                $2^1 * 2^{1-2} * 1$ =
+                $2 * \sfrac{1}{2} * 1$ = 1 \\
+0 & 1 & 0 &   $+\inf$ \\
+0 & 1 & 1 &    \nan \\
+1 & 0 & 0 &   $-0$ \\
+1 & 0 & 1 &   $-1$ \\
+1 & 1 & 0 &   $-\inf$ \\
+1 & 1 & 1 &    \nan \\
 \hline
 \end{tabular}
-\caption{All 16 values representable in binary4 floating-point.
-  The format works reasonably well even at this very low precision,
-  although note how many of the values are not finite.} \label{fig:allvalues4}
+\caption{All 8 values of the hypothetical binary3 representation.
+  There are no normal values; the only finite values are the
+  positive and negative zero and a single subnormal which
+  denotes $1$ (or $-1$).
+} \label{fig:allvalues3}
+\end{center}
 \end{figure}
 
-\subsection{The hypothesized binary3 format} \label{sec:binary3}
-The IEEE-754 representation clearly requires a sign bit, and for this
+The IEEE 754 representation clearly requires a sign bit, and for this
 purpose we need at least one bit for the mantissa in order to
@@
 between $1$ and $2^w-2$ inclusive'' (that is, between 1 and 0 inclusive)
-as denoting the empty set.
+as denoting the empty set. This seems reasonable.
 
@@
 
-% here, precision is still 2 (just one mantissa bit)
-% so emax is 2^{k-p-1}-1, which is 2^{3-2-1}-1 = 0
-% bias = 0
-% emin is 1 - emax = 1 (!?)
-% this is kind of absurd, it may still "work"?
+\section{A hardware math accelerator} \label{sec:hardware}
 
-\begin{figure}[h]
-\begin{tabular}{|l@{\,}c@{\,}l|p{2.5in}|}
-\hline
-  $s$ & $E$ & $T$ & value \\
-  \hline
-0 & 0 & 0 &   $+0$ \\
-0 & 0 & 1 &    subnormal: % $2^{emin} * 2^{1-p} * T$ =
-                $2^1 * 2^{1-2} * 1$ =
-                $2 * \sfrac{1}{2} * 1$ = 1 \\
-0 & 1 & 0 &   $+\infty$ \\
-0 & 1 & 1 &    \nan \\
-1 & 0 & 0 &   $-0$ \\
-1 & 0 & 1 &   $-1$ \\
-1 & 1 & 0 &   $-\infty$ \\
-1 & 1 & 1 &    \nan \\
-\hline
-\end{tabular}
-\caption{All 8 values of the hypothetical binary3 representation.
-  There are no normal values; the only finite values are the
-  positive and negative zero and a single subnormal which
-  denotes $1$ (or $-1$).
-} \label{fig:allvalues3}
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[width=0.95 \linewidth]{chip}
+  \caption{The STM32F303RDT6 wired up as 5 {\tt NAN} gates, shown here
+    {\it in situ}. This is a portion of a larger schematic.
+    Also shown is some support hardware needed for each microprocessor:
+    A programming header, 5 filter capacitors, a crystal oscillator circuit,
+    and a reset switch with external pull-up.
+  }
+  \label{fig:schematic}
+  \end{center}
 \end{figure}
 
-% TODO: We could do without a mantissa if we didn't care about
-% distinguishing NaN and inf. With an empty t, we should
-% consider it to have value 0 (true? maybe it is defined as
-% a product so the empty string is the multiplicative unit 1?)
-% which means that we encode +/- inf, and nan cannot be encoded.
-% Now the operations will not be closed since e.g. 0/0 needs to
-% result in nan.
+So now we know that we can build arbitrary computers with the {\tt
+  NAN} gate, representing the interconnects between the gates efficiently
+with binary3-coded real numbers. All that remains is an efficient
+implementation of the {\tt NAN} gate itself. We could emulate such
+a thing in software, but software is much slower than hardware;
+we would also like to maximize the number of times that we can flip
+between states of the gate (the flip FLOPS) per second.
 
-% TODO: What if we have no mantissa or exponent, just a sign bit?
-% When E=0 and T=0, we can only encode +/- zero. This definitely
-% involves some leaps (and is not closed as above), but could
-% we find expressions that work for just +/- 0?
+Fortunately, there are several pieces of hardware that implement
+IEEE 754 real numbers. I found a moderately-priced micrprocessor
+(\$6.48/ea.), the STM32F303RDT6. This is a 32-bit ARM Cortex M4F
+processor with hardware floating-point running at 72MHz~\cite{stm32}.
+In the rather-difficult-to-solder 10mm surface-mount LQFP64 package,
+it has 64 pins, 51 of which can be used for general-purpose IO.
+Since a {\tt NAN} gate using the binary3 representation needs 9
+pins ($3 \times 2$ for the inputs, $3$ for the outputs), it is
+feasible to implement five {\tt NAN} gates on the same chip with
+a few pins left over for jiggery pokery (Figure~\ref{fig:schematic}).
 
-\nocite{ieee754}
+\begin{figure*}[t!]
+  \begin{center}
+    \includegraphics[width=0.55 \linewidth]{board}
+  \end{center}\vspace{-0.1in}
+  \caption{A beautiful hand-routed circuit board implementing a
+    universal math accelerator, using only the universal {\tt NAN} gate
+    implemented with native floating point hardware.}
+  \label{fig:board}
+\end{figure*}
+
+The hardware math accelerator itself can be thought of as a floating
+point unit (FPU), but one that is streamlined to run only a single
+instruction, the universal function $\inf - {\tt maxNum}(a + b, -\inf)$.
+This is a function taking two binary3 real numbers and
+outputting a single binary3 number. Since there are only $2^6 = 64$
+possible inputs, it can be straightforwardly implemented with table
+lookup, but this would require dozens of microprocessors, which
+might exceed our power budget. In fact there is significant structure
+to the function; for one thing, it can only return $\nan$ or $\inf$ (even
+if arguments like -1.0 or 0.0 are given), and the
+binary3 representation of these only differ in one bit. Equivalent
+logic to determine that bit is as follows:
+
+\begin{small}
+\begin{code}
+if isfinite(a) && isfinite(b)
+then
+  // inf - (a + b) = inf
+  0
+else
+  // a + b is nan when a is nan, b is nan, or a and b are
+  // infinites with different signs. If they are both
+  // -inf, then we have max(-inf, -inf) anyway, which is
+  // the same as max(nan, -inf). So we have:
+  if a == inf && b == inf
+    // both positive infinities
+    // inf - inf = nan
+    1
+  else
+    // inf - -inf = inf
+    0
+\end{code}
+\end{small}
+
+So ultimately this function only returns 1 in the case that both
+inputs are exactly $+\inf$, the pattern {\tt 0 1 0}.
+
+If the inputs are a0 a1 a2, b0, b1, b2, and outputs
+are c0, c1, c2, then:
+\begin{small}
+\begin{code}
+c0 = 0
+c1 = 1
+c2 = !a0 && !a2 && !b0 && !b2 && a1 && b1
+\end{code}
+\end{small}
+
+So we can hardwire the outputs c0 and c1, and use the microprocessor-based
+{\tt NAN} gates to compute c2 as a small boolean function.
+
+Of course, each $0$ or $1$ above is actually itself a binary3-coded
+\nan\ or \inf. Thus on the physical circuit board, this math
+accelerator has $2 \times 3 \times 3$ input pins and $1 \times 3
+\times 3$ output pins. This is just shy of the total number of IO pins
+on the Raspberry~Pi, so we use such a computer to drive the math
+accelerator. Given the large number of traces and small footprint of
+the microprocessors, routing the board gets somewhat involved
+(Figure~\ref{fig:board}).
+
+As of the SIGBOVIK 2019 deadline, such a circuit board has been
+manufactured in China and is in possession of the author (actually the
+minimum order quantity of 10), but he is somewhat nervous about his
+ability to hand-solder these 0.1mm surface-mount leads, so we'll see
+how it goes! Please see \url{http://tom7.org/nan} for project updates
+or an embarrassing {\tt 404 Not Found} if I fail to reboot computing
+using the beautiful foundation of real numbers.
+
 \bibliography{nand}{}

### 2019-04-01 trunk/nand/paper.tex
Oops, fix rom to actually call float version

--- 2019-01-26 trunk/nand/paper.tex
@@
-flinf flops and NaN gates
-flip FLOPs and NaN gates?
+\documentclass[twocolumn,cm]{article}
+\usepackage[top=0.5in, left=0.45in, right=0.45in, bottom=0.5in]{geometry}
 
-Some useful primitives:
+\usepackage{url}
+\usepackage{code}
+% \usepackage{cite}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
 
-        These are called minNum and maxNum in IEE 754-2008 (5.3.1, p19)
+\usepackage{nopageno}
 
-        "If exactly one argument is NaN, they return the other.
-        If both are NaN they return NaN.
+\interfootnotelinepenalty=0
 
-        NAND  0 1     MAX  nan inf
-             +---         +-------
-           0 |1 1     nan |nan inf
-           1 |1 0     inf |inf inf
+% lets me explicitly set a. or 1. etc. as enum label
+\usepackage{enumitem}
 
-                    HYPOT  nan inf
-                          +-------
-                      nan |nan inf
-                      inf |inf inf
+\pagestyle{empty}
 
-        So if 0 = nan and 1 = inf, then MAX is OR.
+\usepackage{ulem}
+% go back to italics for emphasis, though
+\normalem
 
-        We also have AND, since only inf*inf (aka 1 AND 1)
-        returns inf in that truth table.
+\usepackage{natbib}
 
-        Is AND and OR together complete? (I don't think so?)
-        Do we have NOT?
+\setlength{\footnotesep}{2em}
 
-YES!
+\newcommand\comment[1]{}
+\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
 
-Here are some:
-Min(-x, -1.0) + inf
-hypot(nan, max(1/x, -inf))
-inf - max(x, 1)
-sqrt(copysign(inf, -x))
+\newcommand\nan{\textsf{NaN}}
+\renewcommand\inf{\textsf{inf}}
 
-Now direct search for some gates:
-NAND: inf - max(x + y, -inf)
-AND: x + y
-OR: hypot(x, y)
-XOR: Abs(max(y, -inf) - max(x, -inf))
-     Abs(min(y, -x) + max(y, -inf))
-     (Wouldn't be surprised if there's a simpler one,
-     but my search does not find it before it gets
-     out of hand...)
+\newcommand\plusminus{\pm}
 
-NOR: -inf / max(y, max(x, -1))
+\begin{document}
 
-And how about if we leave out min, max? They are certainly
-reasonable since they are defined in IEEE 754, but perhaps
-not the best joke since they are not as math-y as e.g. hypot
+\newcommand\ieee[2]{[{\it IEEE 754-2008}, #1, p#2]}
 
-NOR: sqrt(copysign(-inf, -hypot(y, x)))
-(Here, hypot is just computing OR.)
+% flInf flops?
+\title{NaN gates and flip FLOPS}
+\author{Dr.~Tom~Murphy~VII~Ph.D.\footnote{
+Copyright \copyright\ 2019 the Regents of the Wikiplia
+Foundation. Appears in SIGBOVIK 2019 with the
+half precision
+of the Association for Computational Heresy; 
+{\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
+\$-0.00 } }
 
-So another NOT is:
-sqrt(copysign(-inf, -x))  (actually can be inf)
+\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
+\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
+\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
+\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
+\newcommand\at{\ensuremath{\scriptstyle @}}
 
-XOR: abs(copysign(-inf, y) - copysign(-inf, x))
-OR: hypot(y, x)
-AND: x + y
-NAND: inf - copysign(-inf, x + y)
+\date{1 April 2019}
 
-ok cool, but copysign kinda has the same problems as min,max.
-maybe without that?
+\maketitle \thispagestyle{empty}
 
-the database saturates with 36 entries if we remove
-copysign and min, max. They are these:
+\begin{abstract}
+Yes, this paper contains many layers of abstraction.
+\end{abstract}
 
-[nan, nan, nan, nan] = NaN
-[nan, nan, nan, ~inf] = Minus(Neg(y), x)
-[nan, nan, nan, ~2.0] = Plus(Tanh(Minus(Neg(y), x)), -1)
-[nan, nan, nan, ~1.0] = Tanh(Minus(Neg(y), x))
-[nan, nan, nan, 0.0] = Pow(y, Neg(x))
-[nan, nan, nan, 1.0] = Exp(Pow(y, Neg(x)))
-[nan, nan, nan, 2.0] = Plus(Exp(Pow(y, Neg(x))), 1)
-[nan, nan, nan, inf] = Plus(y, x)
-[nan, nan, ~inf, ~inf] = Neg(y)
-[nan, nan, ~2.0, ~2.0] = Plus(Tanh(Neg(y)), -1)
-[nan, nan, ~1.0, ~1.0] = Tanh(Neg(y))
-[nan, nan, ~0.0, ~0.0] = Div(-1, y)
-[nan, nan, 1.0, 1.0] = Exp(Div(-1, y))
-[nan, nan, 2.0, 2.0] = Minus(1, Tanh(Neg(y)))
-[nan, nan, inf, inf] = y
-[nan, ~inf, nan, ~inf] = Neg(x)
-[nan, ~inf, ~inf, ~inf] = Neg(Hypot(y, x))
-[nan, ~2.0, nan, ~2.0] = Plus(Tanh(Neg(x)), -1)
-[nan, ~2.0, ~2.0, ~2.0] = Plus(Tanh(Neg(Hypot(y, x))), -1)
-[nan, ~1.0, nan, ~1.0] = Tanh(Neg(x))
-[nan, ~1.0, ~1.0, ~1.0] = Tanh(Neg(Hypot(y, x)))
-[nan, ~0.0, nan, ~0.0] = Div(-1, x)
-[nan, ~0.0, ~0.0, ~0.0] = Div(-1, Hypot(y, x))
-[nan, 1.0, nan, 1.0] = Exp(Div(-1, x))
-[nan, 1.0, 1.0, 1.0] = Exp(Div(-1, Hypot(y, x)))
-[nan, 2.0, nan, 2.0] = Minus(1, Tanh(Neg(x)))
-[nan, 2.0, 2.0, 2.0] = Minus(1, Tanh(Neg(Hypot(y, x))))
-[nan, inf, nan, inf] = x
-[nan, inf, inf, inf] = Hypot(y, x)
-[~inf, ~inf, ~inf, ~inf] = NegInf
-[~2.0, ~2.0, ~2.0, ~2.0] = Plus(-1, -1)
-[~1.0, ~1.0, ~1.0, ~1.0] = -1
-[0.0, 0.0, 0.0, 0.0] = 0
-[1.0, 1.0, 1.0, 1.0] = 1
-[2.0, 2.0, 2.0, 2.0] = Plus(1, 1)
-[inf, inf, inf, inf] = Inf
+\section*{Introduction}
+
+Mathematics is fundamental to computer science, and the foundation of
+mathematics is the real numbers; this is obvious from the name. One of
+computing's dirtiest secrets, however, is that computers themselves
+are not based on real numbers---rather, they are based on so-called
+``ones'' and ``zeroes'' combined with ``logic gates'' simulated with
+transistors. While this suffices for most practical purposes, it is
+unsatisfying from a theoretical perspective.
+
+Recently, some progress has been made by human geniuses on completely
+replacing integer calculations with calculations on real
+numbers\cite{fluint8}. While this removes many of the hacks present in
+modern software, there are still many components of the computer (e.g.
+RAM, registers, the {\it scroll lock} LED, a tiny USB-powered fan that
+can cool you on hot summer days or during particularly strenuous
+programming sessions) % XXX cite skymall etc?
+that are not integer-based, and thus cannot be replaced with real numbers
+via this techique.
+
+In this paper I give a new foundation for computing based solely on
+real numbers. I begin with a brief reminder of the definition of real
+numbers, although the reader is expected to be familiar as these are
+pretty fundamental to everything. The approach of the paper is then to
+identify a pair of real numbers that have nice properties
+(Section~\ref{sec:distinguished}), and then to give mathematical
+operations on these numbers that parallel the logical operations
+typically used in the construction of computers
+(Section~\ref{sec:logical}). I then discuss how these operations can
+be implemented efficiently (Section~\ref{sec:hardware}). I
+conclude with some wild speculation.
+
+\section{Real numbers}
+The real numbers are described by IEEE 754, most recently revised in
+AD 2008\cite{ieee754}. Every real number has a sign, a mantissa, and an
+exponent. Actually, this understates the elegance of real numbers,
+since there are a number of numbers, such as \nan\ (``not a number'')
+which are not of this form; \nan\ nas no sign nor mantissa nor
+exponent. We also have \inf\ and $-\inf$, which do have a sign, but no
+mantissa nor exponent. These are the infinite numbers that you get if
+you count very high or very low. Excitingly, we also have both
+positive and negative versions of 0. Some numbers have multiple
+representations, and almost all numbers cannot be represented at all.
+
+The real numbers have an equality operation {\tt ==}. This operation
+has some very exciting properties which are unusual for an equivalence
+relation: It is not reflexive (\nan\ {\tt ==} \nan\ is false), and
+does not obey substitution (for $+0$ {\tt ==} $-0$ is true, but
+$\sfrac{1}{+0}$ {\tt ==} $\sfrac{1}{-0}$ is false).
+
+As a result, real numbers are an absolute joy to work with.
+
+\subsection{Choosing some distinguished values} \label{sec:distinguished}
+
+Computing will need at least two different values. We could choose
+$0.0$ and $1.0$ as in ``binary,'' but these numbers are extremely
+arbitrary; why not $1.0$ and $2.0$? $e$ and $\sfrac{\pi}{2}$? These
+numbers are easily confused with one another. It seems better to use
+distinguished values, making the resulting mathematics more
+distinguished. One of the most distinguished numbers is
+\nan\ (Figure~\ref{fig:gentlenan}). One nice thing about using the
+number \nan\ is that it is not comparable to other numbers, e.g. both
+$\nan < 0.0$ and $0.0 < \nan$ are false. Does it really make sense for
+our fundamental particles to be ordered (e.g. $0 < 1$)? The lack of
+symmetry is abhorrent.
+
+\begin{figure}[ht]
+  \begin{center}
+  \includegraphics[width=0.55 \linewidth]{gentlenan}
+  \caption{A distinguished gentleNaN.} \label{fig:gentlenan}
+  \end{center}
+\end{figure}
+
+The two numbers we choose need to be different; alas they cannot both
+be \nan, since although \nan\ is different from \nan\ (\nan\ {\tt !=}
+\nan), it is not possible to tell them apart (except that
+\nan\ actually has multiple binary
+representations---see Section~\ref{sec:representation}). Another great choice is
+$+\inf$ or $-\inf$. We choose to use $+\inf$ in order to break
+symmetry, and because it will make our scientific contribution more
+positive.
+% So the two values will be \nan\ and \inf; it may
+% be helpful to think of these as false and true, (or 0 and 1), respectively.
+
+% really lean into the joke that we choose the "number" NAN
+
+\subsection{IEEEuler's Identity}
+
+Moreover, \nan\ and \inf\ are part of the pantheon of special values, exhibiting
+exquisite properties, such as IEEEuler's identity:
+
+\begin{huge}
+  \[
+  e^{i \pi} + 1^{\nan \times \inf} = 0
+  \]
+\end{huge}
+
+because $1^n$ is $1$\footnote{This paper uses both exponents and footnotes
+  extensively; please be careful of the difference.}, even for $n=\nan$.\footnote{\ieee{9.2.1}{44}} Another nice
+pair of properties ties these fundamental constants together a
+different way:
+
+\begin{huge}
+  \[
+  {(e^{i \pi})}^{-\inf} = {\tt compound}(\nan, 0)
+  \]
+\end{huge}
+
+${\tt compound}(x, n)$ is the ``compound interest'' function
+$(1 + x)^n$, defined in the IEEE 754 standard, but only available
+in C via floating point extensions~\cite{fpextensions}. This
+function is $1$ for $n=0$ and $x=\nan$.\footnote{\ieee{9.2.1}{44}}
+More excitingly, we have $e^{i \pi} = -1$ (Euler) and ${-1}^{-\inf} =
+1$ ``because all large positive floating-point values are even
+integers.''~\cite{crationale510}
+
+\subsection{NaN's Not GNU} \label{sec:logical}
+
+People who work with real numbers are often taught that the number
+\nan\ is propagated through all expressions that use it (e.g. $\nan -
+1 = \nan$), like some kind of GNU Public Licensed number. This is a
+misconception. We already saw in the beautiful identities above that
+some expressions involving \nan\ do not result in \nan, like $1^{\nan}
+= 1$ and ${\tt compound}(\nan, 0) = 0$. But it is also the case that
+$1^{\inf} = 1$ and ${\tt compound}(\inf, 0) = 0$. Are there mathematical
+functions that distinguish between \nan\ and \inf?
+
+It turns out that there are! For example, the functions {\tt minNum}
+and {\tt maxNum} (\ieee{5.3.1}{19}) take two arguments and return the
+min and max, respectively. They have the special, distinguished property
+that ``if exactly one argument is \nan, they return the other. If
+both are \nan\ they return \nan.''
+
+\begin{figure}[ht]
+\renewcommand*{\arraystretch}{1.4}
+  \begin{tabular}{cc}
+  \begin{tabular}{r|cc}
+    \multicolumn{1}{r}{${\tt maxNum}(a,b)$} & \multicolumn{1}{c}{\nan} & \inf \\[0.2em]\cline{2-3}
+    \nan & \nan & \inf \\
+    \inf & \inf & \inf \\
+  \end{tabular} &
+  \begin{tabular}{r|cc}
+    \multicolumn{1}{r}{$a * b$} & \multicolumn{1}{c}{\nan} & \inf \\[0.2em]\cline{2-3}
+    \nan & \nan & \nan \\
+    \inf & \nan & \inf \\
+  \end{tabular}
+  \end{tabular}
+  \caption{
+    The behavior of some mathematical functions on our distinguished values \nan\ and \inf.
+    {\tt maxNum} returns \inf\ if either of its arguments is \inf\ (some other functions have
+    this property, like {\tt hypot}). $a * b$ is \inf\ only
+    if both of its arguments are \inf\ (there are many other examples, like $a + b$).
+  } \label{fig:truthtables}
+\end{figure}
+
+With functions such as these, we can begin constructing the building blocks
+of more interesting functions (Figure~\ref{fig:truthtables}). Unfortunately,
+${\tt maxNum}(a,b)$ and $a*b$ are not complete on their own; we additionally
+need at least a function $f(x)$ where $f(\nan) = \inf$ and $f(\inf) = \nan$.
+Does such a function exist? Yes! Several can be built from IEEE 754 primitives:
+\[
+\begin{array}{rcl}
+  f(x) & = & {\tt minNum}(-x, -1.0) + \inf \\
+  f(x) & = & {\tt hypot}(\nan, {\tt maxNum}(1/x, -\inf)) \\
+  f(x) & = & \inf - {\tt maxNum}(x, 1.0) \\
+  f(x) & = & {\tt sqrt}({\tt copysign}(\inf, -x)) \\
+\end{array}
+\]
+You can try these out in your favorite programming language, and if they don't
+work, your implementation is not IEEE 754 compliant. Why do these work? Let's
+take the first one, and compare \nan\ and \inf\ for $x$:
+\[
+\begin{array}{l|r}
+  \hline
+  x = \nan                             & x = \inf \\
+  \hline
+  {\tt minNum}(-x, -1.0) + \inf    & {\tt minNum}(-x, -1.0) + \inf \\
+  {\tt minNum}(-\nan, -1.0) + \inf & {\tt minNum}(-\inf, -1.0) + \inf \\
+      -1.0 + \inf                      & -\inf + \inf \\
+      \inf                             & \nan \\
+\end{array}
+\]
+
+Thinking of \nan\ as {\tt false} and \inf\ as {\tt true}, we now have {\tt AND} ({\tt maxNum}),
+{\tt OR} ($*$), and {\tt NOT} (${\tt minNum}(-x, -1.0)$). With these we can create arbitrary
+functions $f(a_1, a_2, \ldots, a_n)$ that return our choice of $\nan$ or $\inf$ for the $2^n$
+different combinations of arguments. It is also possible to find more direct expressions that
+compute simple functions (Figure~\ref{fig:truthtables2}).
+
+\begin{figure}[ht]
+\renewcommand*{\arraystretch}{1.4}
+
+  \begin{tabular}{r|cc}
+    \multicolumn{1}{r}{$\inf - {\tt maxNum}(a + b, -\inf)$} & \multicolumn{1}{c}{\nan} & \inf \\[0.2em]\cline{2-3}
+    \nan & \inf & \inf \\
+    \inf & \inf & \nan \\
+    \multicolumn{3}{c}{}\\[0.1em]
+    \multicolumn{1}{r}{${\tt abs}({\tt minNum}(b, -a) + {\tt maxNum}(b, -\inf))$} & \multicolumn{1}{c}{\nan} & \inf \\[0.2em]\cline{2-3}
+    \nan & \nan & \inf \\
+    \inf & \inf & \nan \\
+    \multicolumn{3}{c}{}\\[0.1em]
+    \multicolumn{1}{r}{$-\inf / {\tt maxNum}(b, {\tt maxNum}(a, -1))$} & \multicolumn{1}{c}{\nan} & \inf \\[0.2em]\cline{2-3}
+    \nan & \inf & \nan \\
+    \inf & \nan & \nan \\
+  \end{tabular}
+
+  \caption{
+    Some interesting functions of two variables. They are isomorphic to the boolean functions {\tt NAND},
+    {\tt XOR} and {\tt NOR} respectively, but more beautiful.
+  } \label{fig:truthtables2}
+\end{figure}
+
+\begin{figure*}[tp]
+\begin{tabular}{|l|c|c|c|c|c|c|}
+  \hline
+  parameter & {\bf binary4} & binary16 & binary32 & binary64 & binary128 & binary$_k$ \\
+  \hline
+  $k$, storage in bits & {\bf 4} & 16 & 32 & 64 & 128 & multiple of 32 \\
+  $p$, precision in bits & {\bf 2} & 11 & 24 & 53 & 113 & $k$ - round(4 * log$_2$(k)) + 13 \\
+  $emax$, maximum exponent e & {\bf 1} & 15 & 127 & 1023 & 16383 & $2^{k-p-1}$ - 1 \\
+  $bias$ = $E$ - $e$ & {\bf 1} & 15 & 127 & 1023 & 16383 & emax \\
+  $sign bits$ & {\bf 1} & 1 & 1 & 1 & 1 & 1 \\
+  $w$, exponent width & {\bf 2} & 5 & 8 & 11 & 15 & round(4 * log$_2$(k)) - 13 \\
+  $t$, trailing significand width & {\bf 1} & 10 & 23 & 52 & 112 & k - w - 1 \\
+  $k$, storage width & {\bf 4} & 16 & 32 & 64 & 128 & 1 + w + t \\
+\hline
+\end{tabular}
+\caption{Parameters for the newly-introduced {\bf binary4} encoding
+  for IEEE 754, compared to the standard widths (see Table~3.5 in the
+  standard\cite{ieee754}). } \label{fig:binary4}
+\end{figure*}
+
+I found these functions through computer search,\footnote{
+  Source code is available at \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/nand/}
+  } using a technique
+like bottom-up logic programming~\cite{pfenning2006logic}. I start
+with a small set of constants, including arguments {\sf a} and {\sf
+  b}, and then compute all of the expressions that can be made by
+applying a single mathematical function (e.g. ${\tt abs}(x)$, $-x$) or
+binary mathematical function ($x + y$, $x / y$, ${\tt maxNum}(x, y)$)
+to existing expressions. The expression is actually a collection of
+values taken on for each possible substitution (in $\{\nan, \inf\}$)
+to arguments {\sf a} and {\sf b} (i.e., it represents a function). If
+the expression has the correct values for each possible assignment to
+the arguments, then we are done. We only need to keep one (the
+smallest) expression that represents a distinct function, but note
+that we have to consider intermediate expressions that compute values
+other than \nan\ and \inf. Also note that we need one of {\tt minNum},
+{\tt maxNum} or {\tt copySign} in order to compute the {\tt NOT}
+function; we could think of these functions as therefore essential
+to mathematical completeness.
+% TODO: Could show all 36 functions without these..
+
+Particularly nice is $\inf - {\tt maxNum}(a + b, -\inf)$, which
+returns $\inf$ if either of its arguments is $\nan$. We will call this
+the ``{\tt NAN} gate'', for ``Not \nan''. The {\tt NAN} gate is
+exciting because it can be used on its own to construct all other
+boolean functions! We can use \nan, \inf, and this function to
+construct any computer and any computable function. Beautiful!
+
+To program with numbers on computers, the real numbers are
+represented as strings of bits. Next we'll talk about efficient
+representations that allow us to manipulate \nan\ and \inf\ 
+with {\tt NAN} gates.
+
+\section{The binary4 representation} \label{sec:representation}
+
+% bias = 1
+
+\begin{figure}[ht]
+\begin{center}
+  \begin{tabular}{|l@{\,}c@{\,}l|p{2.5in}|}
+\hline
+  $s$ & $E$ & $T$ & value \\
+  \hline
+0 & 00 & 0 &   $+0$ \\
+0 & 00 & 1 &    subnormal: % $2^{emin} * 2^{1-p} * T$ =
+                $2^0 * 2^{1-2} * 1$ =
+                $1 * \sfrac{1}{2} * 1$ = 0.5 \\
+0 & 01 & 0 &    normal: % $2^{E - bias} * (1 + 2^{1-p} * T)$
+                $2^0 * (1 + \sfrac{1}{2} * 0)$ = 1 \\
+0 & 01 & 1 &    $2^0 * (1 + \sfrac{1}{2} * 1)$ = 1.5 \\
+0 & 10 & 0 &    $2^1 * (1 + \sfrac{1}{2} * 0)$ = 2 \\
+0 & 10 & 1 &    $2^1 * (1 + \sfrac{1}{2} * 1)$ = 3 \\
+0 & 11 & 0 &   $+\inf$ \\
+0 & 11 & 1 &    \nan \\
+1 & 00 & 0 &   $-0$ \\
+1 & 00 & 1 &   $-0.5$ \\
+1 & 01 & 0 &   $-1$ \\ 
+1 & 01 & 1 &   $-1.5$ \\
+1 & 10 & 0 &   $-2$ \\
+1 & 10 & 1 &   $-3$ \\ 
+1 & 11 & 0 &   $-\inf$ \\
+1 & 11 & 1 &    \nan \\
+\hline
+\end{tabular}
+\end{center}
+  \caption{All 16 values representable in binary4 floating-point.
+  The format works reasonably well even at this very low precision,
+  although note how many of the values are not finite.} \label{fig:allvalues4}
+\end{figure}
+
+IEEE 754 natively defines several bit widths for floating-point
+values, such as the 32-bit binary32 (aka ``single-precision float'')
+and 64-bit binary64 (aka ``double-precision float''). The
+specification is parameterized to allow other bit widths; for example,
+half-precision 16-bit floats are common in GPU code for machine
+learning applications~\cite{wikipedia16}. Smaller floats sacrifice precision,
+but require less space and allow faster calculations. For our purposes
+in this paper, since we only need to represent the two numbers \nan\ and
+\inf, we are interested in the smallest possible representation.
+
+This section describes the binary4 representation, a four-bit floating
+point number that is clearly allowed by the IEEE 754 standard.
+
+The representation of any floating-point number has a single sign bit,
+some number $w$ of exponent bits, and some number $t$ of mantissa
+bits. For binary32, $w = 8$ and $t = 23$; and with the sign bit we
+have $23 + 8 + 1 = 32$ bits as expected. We need at least a sign bit,
+but what are the smallest permissible values of $w$ and $t$?
+
+The most stringent constraint on $w$ comes in \ieee{3.4}{9}, which
+states
+\begin{quote}
+  The range of the encoding's biased exponent $E$ shall include:
+  \begin{enumerate}[label=---]
+    \item every integer between 1 and $2^w - 2$, inclusive, to encode
+      normal numbers
+    \item the reserved value 0 to encode $\plusminus 0$ and subnormal
+      numbers
+    \item the reserved value $2^w - 1$ to encode $\plusminus \infty$
+      and NaNs.
+  \end{enumerate}
+\end{quote}
+
+$E$ is the binary number encoded by $w$. It must include at least the
+two special values consisting of all zeroes and all ones (second and
+third clause). A conservative reading of ``every integer between 1 and
+$2^w - 2$'' seems to require that $1 \leq 2^w - 2$ (otherwise how
+could the interval be inclusive of its endpoints?), which would imply
+that $w$ is at least 2. (However, see Section~\ref{sec:binary3} for
+the hypothesized case where $w=1$.)
+
+The representation of \nan\ and \inf\ are distinguished by the value
+of $t$ when $E$ is all ones. We certainly need to distinguish these,
+so $t = 1$ is the minimal size.
+
+We have one sign bit, two exponent bits, and one mantissa bit, for a
+total of four. Since ``single precision'' is 32 bits, ``half
+precision'' is 16, 4 bits is ``eighth precision.'' Given how nicely
+all this works out, shouldn't there be an {\tt eighth} base type in
+most modern programming languages and GPUs? Since there are so few
+values representable, it would be practical for all the standard
+operations to be done in constant time via table lookups. All 16
+possible values are given in Figure~\ref{fig:binary4}.
+
+Four bits is not many, but is it possible to represent these two
+values more efficiently?
+
+\subsection{The hypothesized binary3 format} \label{sec:binary3}
+
+\begin{figure}[ht]
+\begin{center}
+\begin{tabular}{|l@{\,}c@{\,}l|p{2.5in}|}
+\hline
+  $s$ & $E$ & $T$ & value \\
+  \hline
+0 & 0 & 0 &   $+0$ \\
+0 & 0 & 1 &    subnormal: % $2^{emin} * 2^{1-p} * T$ =
+                $2^1 * 2^{1-2} * 1$ =
+                $2 * \sfrac{1}{2} * 1$ = 1 \\
+0 & 1 & 0 &   $+\inf$ \\
+0 & 1 & 1 &    \nan \\
+1 & 0 & 0 &   $-0$ \\
+1 & 0 & 1 &   $-1$ \\
+1 & 1 & 0 &   $-\inf$ \\
+1 & 1 & 1 &    \nan \\
+\hline
+\end{tabular}
+\caption{All 8 values of the hypothetical binary3 representation.
+  There are no normal values; the only finite values are the
+  positive and negative zero and a single subnormal which
+  denotes $1$ (or $-1$).
+} \label{fig:allvalues3}
+\end{center}
+\end{figure}
+
+The IEEE 754 representation clearly requires a sign bit, and for this
+purpose we need at least one bit for the mantissa in order to
+distinguish \nan\ and \inf. It is perhaps a stretch of the wording,
+but arguably the spec permits a 1-bit exponent ($w = 1$). To
+rationalize this we need to interpret the phrase ``every integer
+between $1$ and $2^w-2$ inclusive'' (that is, between 1 and 0 inclusive)
+as denoting the empty set. This seems reasonable.
+
+With one bit for sign, exponent, and mantissa, we can represent just 8
+different values. Here $emax$ is $0$, and the standard clearly
+requires $emin = 1 - emax$, so $emin = 1$. Certainly fishy for $emin$
+to be larger than $emax$, but we can just not stress out about it; the
+representable values are all reasonable-looking
+(Figure~\ref{fig:allvalues3}).
+
+\section{A hardware math accelerator} \label{sec:hardware}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[width=0.95 \linewidth]{chip}
+  \caption{The STM32F303RDT6 wired up as 5 {\tt NAN} gates, shown here
+    {\it in situ}. This is a portion of a larger schematic.
+    Also shown is some support hardware needed for each microprocessor:
+    A programming header, 5 filter capacitors, a crystal oscillator circuit,
+    and a reset switch with external pull-up.
+  }
+  \label{fig:schematic}
+  \end{center}
+\end{figure}
+
+So now we know that we can build arbitrary computers with the {\tt
+  NAN} gate, representing the interconnects between the gates efficiently
+with binary3-coded real numbers. All that remains is an efficient
+implementation of the {\tt NAN} gate itself. We could emulate such
+a thing in software, but software is much slower than hardware;
+we would also like to maximize the number of times that we can flip
+between states of the gate (the flip FLOPS) per second.
+
+Fortunately, there are several pieces of hardware that implement
+IEEE 754 real numbers. I found a moderately-priced microprocessor
+(\$6.48/ea.), the STM32F303RDT6. This is a 32-bit ARM Cortex M4F
+processor with hardware floating-point running at 72MHz~\cite{stm32}.
+In the rather-difficult-to-solder 10mm surface-mount LQFP64 package,
+it has 64 pins, 51 of which can be used for general-purpose IO.
+Since a {\tt NAN} gate using the binary3 representation needs 9
+pins ($3 \times 2$ for the inputs, $3$ for the outputs), it is
+feasible to implement five {\tt NAN} gates on the same chip with
+a few pins left over for jiggery pokery (Figure~\ref{fig:schematic}).
+
+\begin{figure*}[t!]
+  \begin{center}
+    \includegraphics[width=0.55 \linewidth]{board}
+  \end{center}\vspace{-0.1in}
+  \caption{A beautiful hand-routed circuit board implementing a
+    universal math accelerator, using only the universal {\tt NAN} gate
+    implemented with native floating point hardware.}
+  \label{fig:board}
+\end{figure*}
+
+The hardware math accelerator itself can be thought of as a floating
+point unit (FPU), but one that is streamlined to run only a single
+instruction, the universal function $\inf - {\tt maxNum}(a + b, -\inf)$.
+This is a function taking two binary3 real numbers and
+outputting a single binary3 number. Since there are only $2^6 = 64$
+possible inputs, it can be straightforwardly implemented with table
+lookup, but this would require dozens of microprocessors, which
+might exceed our power budget. In fact there is significant structure
+to the function; for one thing, it can only return $\nan$ or $\inf$ (even
+if arguments like -1.0 or 0.0 are given), and the
+binary3 representation of these only differ in one bit. Equivalent
+logic to determine that bit is as follows:
+
+\begin{small}
+\begin{code}
+if isfinite(a) && isfinite(b)
+then
+  // inf - (a + b) = inf
+  0
+else
+  // a + b is nan when a is nan, b is nan, or a and b are
+  // infinites with different signs. If they are both
+  // -inf, then we have max(-inf, -inf) anyway, which is
+  // the same as max(nan, -inf). So we have:
+  if a == inf && b == inf
+    // both positive infinities
+    // inf - inf = nan
+    1
+  else
+    // inf - -inf = inf
+    0
+\end{code}
+\end{small}
+
+So ultimately this function only returns 1 in the case that both
+inputs are exactly $+\inf$, the pattern {\tt 0 1 0}.
+
+If the inputs are a0 a1 a2, b0, b1, b2, and outputs
+are c0, c1, c2, then:
+\begin{small}
+\begin{code}
+c0 = 0
+c1 = 1
+c2 = !a0 && !a2 && !b0 && !b2 && a1 && b1
+\end{code}
+\end{small}
+
+So we can hardwire the outputs c0 and c1, and use the microprocessor-based
+{\tt NAN} gates to compute c2 as a small boolean function.
+
+Of course, each $0$ or $1$ above is actually itself a binary3-coded
+\nan\ or \inf. Thus on the physical circuit board, this math
+accelerator has $2 \times 3 \times 3$ input pins and $1 \times 3
+\times 3$ output pins. This is just shy of the total number of IO pins
+on the Raspberry~Pi, so we use such a computer to drive the math
+accelerator. Given the large number of traces and small footprint of
+the microprocessors, routing the board gets somewhat involved
+(Figure~\ref{fig:board}).
+
+As of the SIGBOVIK 2019 deadline, such a circuit board has been
+manufactured in China and is in possession of the author (actually the
+minimum order quantity of 10), but he is somewhat nervous about his
+ability to hand-solder these 0.1mm surface-mount leads, so we'll see
+how it goes! Please see \url{http://tom7.org/nan} for project updates
+or an embarrassing {\tt 404 Not Found} if I fail to reboot computing
+using the beautiful foundation of real numbers.
+
+\bibliography{nand}{}
+\bibliographystyle{plain}
+
+\end{document}

### 2019-04-03 chess/papers/blind.tex
Some misc improvements while wrapping up NaN gate talk

@@
 
-Say \verb+blind_kings+ is playing as black; it remains easy to fool it
+Say \verb+blind_kings+ is playing as white; it remains easy to fool it
 by moving black pieces into white's camp, since they are usually then
@@
 
-I would like to thank the little people (pawns) and the the author-
+I would like to thank the little people (pawns) and the author-
 and content-blind anonymous referrees.

### 2019-04-03 cc-lib/image.h

@@ -32,6 +32,11 @@ struct ImageRGBA {
 
   void SetPixel(int x, int y, uint8 r, uint8 g, uint8 b, uint8 a);
   void SetPixel32(int x, int y, uint32 rgba);
+
+  // Blend pixel with existing data.
+  // Note: Currently assumes existing alpha is 0xFF.
+  void BlendPixel(int x, int y, uint8 r, uint8 g, uint8 b, uint8 a);
+  // void BlendPixel32(int x, int y, uint32 rgba);
   
   const int width, height;
   // Size width * height * 4.

### 2019-04-03 trunk/nand/board/iface/nandy.h
Some misc improvements while wrapping up NaN gate talk

@@ -17,6 +17,8 @@
 #include <tuple>
 #include <deque>
 #include <functional>
+#include <unistd.h>
+#include <ctime>
 
 #include "base/logging.h"
 #include "base/stringprintf.h"

### 2019-04-03 trunk/nand/board/iface/test-nandy.cc

@@ -20,6 +20,36 @@ using uint64 = uint64_t;
 
 #define SAVE_IMAGE 0
 
+static void BenchNandy() {
+  ArcFour rc{"nandy"};
+  auto Rand3 = [&rc]() { return Binary3(rc.Byte() & 0b111); };
+  Nandy nandy;
+  for (int i = 0; i < Nandy::MEM_SIZE; i++) {
+    nandy.MEM[i] = Rand3();
+  }
+
+  int64 start = time(nullptr);
+  static constexpr int NUM_STEPS = 100000000;
+  for (int y = 0; y < NUM_STEPS; y++) {
+    nandy.Step();
+  }
+
+  int64 elapsed = time(nullptr) - start;
+  printf("%d steps in %lld = %.2f/s\n",
+	 NUM_STEPS, elapsed,
+	 (double)NUM_STEPS / (double)elapsed);
+
+  // Make sure result is used.
+  vector<Binary3> row = nandy.GetState();
+  string trace;
+  for (int x = 0; x < row.size(); x++) {
+    uint8 rgb = row[x].Bits();
+    trace.push_back('0' + rgb);
+  }
+  printf("Final: %s\n",
+	 MD5::Ascii(MD5::Hash(trace)).c_str());
+}
+
 static void TestNandy() {
   ArcFour rc{"nandy"};
   auto Rand3 = [&rc]() { return Binary3(rc.Byte() & 0b111); };
@@ -99,6 +129,9 @@ void Test2Nandy() {
 
 
 int main(int argc, char **argv) {
+  BenchNandy();
+  return 0;
+    
   TestNandy();
   
   (void)TestNandy;

### 2020-03-01 chess/papers/longest.tex
Start on this turd/gem

\documentclass[twocolumn]{article}
\usepackage[top=0.25in, left=0.25in, right=0.25in, bottom=0.25in]{geometry}

\usepackage{chessboard}
\usepackage{setspace}
\usepackage{seqsplit}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{code}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{textcomp}

\pagestyle{empty}

\renewcommand\comment[1]{}

\newcommand\st{$^{\mathrm{st}}$}
\newcommand\nd{$^{\mathrm{nd}}$}
\newcommand\rd{$^{\mathrm{rd}}$}
\renewcommand\th{$^{\mathrm{th}}$}
\newcommand\tm{$^{\mbox{\tiny \textsc{tm}}}$}

\newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}

\newcommand\citef[1]{\addtocounter{footnote}{1}\footnotetext{\cite{#1}}\ensuremath{^{\mbox{\footnotesize [\thefootnote]}}}}

\setchessboard{showmover=false}

\begin{document}

\title{Is this the longest Chess game?}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2020 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2015 with the {\tt ShortMoveString} of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
\textcurrency 17696
}
}

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{1 April 2020}

\maketitle

\section{Introduction}

In my experience, most chess games end in a few moves. Players who
seek longevity typically do so by playing many games of chess, rather
than a protracted

Although many people ``know how to play'' chess, almost nobody fully
understands the rules of chess, most authoritatively given by
FIDE~\cite{fiderules}. (See for example Figure~\ref{fig:castle960} for
a minor chess scandal that erupted in 2019 over an obscure corner case
in the rules.) Several of these ``deep cuts'' have to do with game-ending
conditions that were introduced to keep the game from going on forever.

Many chess moves are reversible (e.g. moving the knight forward and
back to its starting position~\cite{survival}), so informal games of
chess could last forever with the players repeating a short cycle. In
AD~1561, Ruy L\'opez added the ``fifty-move rule'' to prevent infinite
games.\footnote{This rule only applied to games started after its
  introduction, so it is possible that some pre-1561 games are still
  in progress and may never end.} This rule (detailed below) ensures
that irreversible moves are regularly played, and so the game always
makes progress towards an end state. Another rule, ``threefold
repetition'' also guarantees termination as a sort of backup plan
(either of these rules would suffice on its own).

% This is good for the insurers and reinsurers of chess tournaments
% (games are guaranteed to end, rather than simply ending with
% probability 1).
So, chess is formally a finite game. This good for computer
scientists, since it means that chess has a trivial O(1) optimal
solution. This allows us to move onto other important questions, like:
What is the longest chess game? In this paper I show how to compute
such a game, and then gratuitously present all of its
% XXX CHECK
17,696 
%
half-moves. Even if you are a chess expert (``chexpert''), I bet you
will be surprised at some of the corner cases in the rules that are
involved.

Speaking of rules, let's first detail the two main rules that limit the
length of the game.

\begin{figure}
  \begin{center}
    \chessboard[setfen=2nr1r1k/pb1n1pqp/1p2p1p1/1N1pN3/5P2/1P4P1/P2PPQ1P/2R1K1RB,showmover=false]
  \end{center}
  \caption{ ({\it Nepomniachtchi -- So, 2019.}) White to move during
    a speed Chess960 (aka ``Fischer Random'') tournament. In this
    variant, the pieces start in different positions, but castling
    rules are such that the king and rook end up on the same squares
    that they would in normal chess. As a result, it is possible for
    the king or rook to not move during castling, or for the
    destination square for the king to already be occupied by the
    rook. Attempting to castle in the position depicted,
    (Figure~\ref{fig:castle960}), grandmaster Ian Nepomniachtchi
    first touched the rook to move it out of the way. However,
    piece-touching rules require that when castling, the player
    first moves the king; but how? The rook is occupying g1! One
    commenter suggested tossing the king into the air, then sliding
    the rook to f1 while the king is airborne, and then watching the
    king land dead center on its target. The arbiter required Nepo
    to make a rook move instead, but this was later appealed, and
    the game replayed. } \label{fig:castle960}
\end{figure}

(Players are collaborating, perhaps playing chess to the death)

\subsection{The seventy-five move rule}

The ``fifty-move rule''

Although usually known by the name ``fifty-move rule'' ()

\subsection{Fivefold repetition}

Although only

% \begin{figure}[htp]
% \input{table-two}
% \caption{Minimal joining strings for every letter (rows) to every other letter (columns).} \label{fig:table}
% \end{figure}

\begin{sloppypar}
\tiny
\setstretch{0.1}
\hyphenpenalty=1
\pretolerance=2
\tolerance=2

\noindent \input{slow}
0--0

\end{sloppypar}

\nocite{chesstego}

\bibliography{chess}{}
\bibliographystyle{plain}

\end{document}

### 2020-03-09 trunk/chess/papers/longest.tex
Some work on longest paper/code. Shaved one slack move off, so feeling pretty confident in the argument that this is the longest...

\documentclass[twocolumn]{article}
\usepackage[top=0.25in, left=0.25in, right=0.25in, bottom=0.25in]{geometry}

\usepackage{chessboard}
\usepackage{setspace}
\usepackage{seqsplit}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{code}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{textcomp}

% XXX search for common typo "movies"

% lets me explicitly set a. or 1. etc. as enum label
\usepackage{enumitem}

\usepackage[pdftex,
            pdfauthor={Dr. Tom Murphy VII Ph.D.},
            pdftitle={Is this the longest chess game?},
            pdfkeywords={SIGBOVIK, chess}]{hyperref}

\pagestyle{empty}

\renewcommand\comment[1]{}

\newcommand\st{$^{\mathrm{st}}$}
\newcommand\nd{$^{\mathrm{nd}}$}
\newcommand\rd{$^{\mathrm{rd}}$}
\renewcommand\th{$^{\mathrm{th}}$}
\newcommand\tm{$^{\mbox{\tiny \textsc{tm}}}$}

\newcommand\sfrac[2]{{}\,$^{#1}$\!/{}\!$_{#2}$}

\newcommand\citef[1]{\addtocounter{footnote}{1}\footnotetext{\cite{#1}}\ensuremath{^{\mbox{\footnotesize [\thefootnote]}}}}

\setchessboard{showmover=false}

% Define black versions of pieces for inline use. Gross, but it works.
\newcommand{\Pawn}[1][1.3ex]{%
\adjustbox{Trim=4.3pt 2.6pt 4.3pt 0pt,width=#1,margin=0.2ex 0ex 0.2ex 0ex}{\BlackPawnOnWhite}%
}%
\newcommand{\Rook}[1][1.58ex]{%
\adjustbox{Trim=3.2pt 2.2pt 3.2pt 0pt,width=#1,raise=0ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackRookOnWhite}%
}%
\newcommand{\Knight}[1][1.85ex]{%
\adjustbox{Trim=2.3pt 2.35pt 2.5pt 0pt,width=#1,raise=-0.03ex,margin=0.14ex 0ex 0.14ex 0ex}{\BlackKnightOnWhite}%
}%
\newcommand{\Bishop}[1][1.79ex]{%
\adjustbox{Trim=2.3pt 2pt 2.3pt 0pt,width=#1,raise=-0.12ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackBishopOnWhite}%
}%
\newcommand{\Queen}[1][2.05ex]{%
\adjustbox{Trim=1.2pt 2.2pt 1.2pt 0pt,width=#1,raise=-0.08ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackQueenOnWhite}%
}%
\newcommand{\King}[1][1.95ex]{%
\adjustbox{Trim=2pt 2pt 2pt 0pt,width=#1,raise=-0.06ex,margin=0.13ex 0ex 0.13ex 0ex}{\BlackKingOnWhite}%
}%

\begin{document}

\title{Is this the longest Chess game?}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2020 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2015 with the {\tt ShortMoveString} of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
% XXX should match the length of the game
\textcurrency 17697
}
}

\renewcommand\>{$>$}
\newcommand\<{$<$}

\date{1 April 2020}

\maketitle

\section{Introduction}

In my experience, most chess games end in a few moves. Players who
seek longevity typically do so by playing many games of chess, rather
than a protracted XXX

Although many people ``know how to play'' chess, almost nobody fully
understands the rules of chess, most authoritatively given by
FIDE~\cite{fiderules}. (See for example Figure~\ref{fig:castle960} for
a minor chess scandal that erupted in 2019 over an obscure corner case
in the rules.) Several of these ``deep cuts'' have to do with game-ending
conditions that were introduced to keep the game from going on forever.

Many chess moves are reversible (e.g. moving the knight forward and
back to its starting position~\cite{survival}), so informal games of
chess could last forever with the players repeating a short cycle. In
AD~1561, Ruy L\'opez added the ``fifty-move rule'' to prevent infinite
games.\footnote{This rule only applied to games started after its
  introduction, so it is possible that some pre-1561 games are still
  in progress and may never end.} This rule (detailed below) ensures
that irreversible moves are regularly played, and so the game always
makes progress towards an end state. Another rule, ``threefold
repetition'' also guarantees termination as a sort of backup plan
(either of these rules would suffice on its own).

% This is good for the insurers and reinsurers of chess tournaments
% (games are guaranteed to end, rather than simply ending with
% probability 1).
So, chess is formally a finite game. This good for computer
scientists, since it means that chess has a trivial O(1) optimal
solution. This allows us to move onto other important questions, like:
What is the longest chess game? In this paper I show how to compute
such a game, and then gratuitously present all of its
% XXX check that this is the final one
17,697 
%
half-moves. Even if you are a chess expert (``chexpert''), I bet you
will be surprised at some of the corner cases in the rules that are
involved.

Speaking of rules, let's first detail the two main rules that limit the
length of the game.

\begin{figure}
  \begin{center}
    \chessboard[
      % fillarea=a3-h1,printarea=a3-h1,
      setfen=2nr1r1k/pb1n1pqp/1p2p1p1/1N1pN3/5P2/1P4P1/P2PPQ1P/2R1K1RB,
      pgfstyle=curvemove,markmoves={e1-g1},
      pgfstyle=straightmove,markmoves={g1-f1},
      showmover=false]
  \end{center}
  \caption{ ({\it Nepomniachtchi -- So, 2019.}) White to move during a
    speed Chess960 (aka ``Fischer Random'') tournament. In this
    variant, the pieces start in different positions, but castling
    rules are such that the king and rook end up on the same squares
    that they would in normal chess. As a result, it is possible for
    the king or rook to not move during castling, or for the
    destination square for the king to already be occupied by the
    rook. Attempting to castle in the position depicted, grandmaster
    Ian Nepomniachtchi first touched the rook to move it out of the
    way. However, piece-touching rules require that when castling, the
    player first moves the king (and ``Each move must be played with
    one hand only''); but how? The rook is occupying g1! One commenter
    suggested tossing the king into the air, then sliding the rook to
    f1 while the king is airborne, and then watching the king land
    dead center on its target. The arbiter required Nepo to make a
    rook move instead, but this was later appealed, and the game
    replayed. } \label{fig:castle960}
\end{figure}

% TODO: (Players are collaborating, perhaps playing chess to the death)

\subsection{The seventy-five move rule} \label{sec:75}

The ``fifty-move rule,''~\cite{wikipedia50move} as it is usually
known, requires that an irreversible move is played at a minimum pace.
For the sake of this rule, irreversible moves are considered captures
and pawn moves,%
\footnote{ Other irreversible moves include: Castling, moving a piece
  so as to lose castling rights, and declining to capture {\it en
    passant} (such capture must be made immediately, so if the option
  is not taken, it cannot be regained). The fifty-move rule could be
  soundly expanded to include these, but, that's just like, not the
  rule, man.}
%
including promotion and {\it en passant} (which is also a capturing
move) but not the movement of a previously promoted piece.
Specifically~\cite{fiderules}:

\begin{quote}
{\bf 9.3}.\quad The game is drawn, upon a correct claim by the player having
the move, if:
  \begin{enumerate}[label=\alph*.]
    \item he writes his move on his scoresheet and declares to the
      arbiter his intention to make this move, which shall result in
      the last 50 moves having been made by each player without the
      movement of any pawn and without any capture, or
    \item the last 50 consecutive moves have been made by each player
      without the movement of any pawn and without any capture.
  \end{enumerate}
\end{quote}

Note that what is provided here is the option for a player to {\em
  claim} a draw (and the two provisions essentially allow either player
to claim the draw at the moment of the 50\th\ move). If neither player
is interested in a draw, either because they think their position is
still winning, or are just trying to create the longest ever chess game,
the game legally continues. That's why what is actually relevant for this
paper is provision 9.6, which defines a draw:

\begin{quote}
  {\bf 9.6}.\quad If one or both of the following occur(s) then the
  game is drawn:

  \ldots 9.6.2. any series of at least 75 moves have been made by each
  player without the movement of any pawn and without any capture. If
  the last move resulted in checkmate, that shall take precedence.
\end{quote}

\begin{figure}[htp]
  \begin{center}
    \chessboard[setfen=8/r6n/8/8/5k2/3K4/7N/3b3Q b - - 0 1]
  \end{center}
  \caption{Black to move and mate in 545 moves (!). The position was
    found (by Zakharov and Makhnichev~\cite{tablebase7}) while
    building an endgame tablebase of all possible 7-piece positions.
    Of course, the game ends prematurely in a draw because of the
    75-move rule. } \label{fig:matein545}
\end{figure}

Draws after 75 moves (per player, so really 150 half-moves) are
compulsory.

Interestingly (at least as interesting as anything in this dubious
affair) it is known that some otherwise winning endgame positions
require more than 50 moves to execute (Figure~\ref{fig:matein545}).
The rules of chess have at various times allowed for longer timers in
such known situations, but were later simplified to the fixed 50 (and
75) move limit.

\subsection{Fivefold repetition} \label{sec:fivefold}

The 75-move rule is rarely applied in practice, but its
counterpart, ``threefold repetition''~\cite{wikipediathreefold} is often
the cause of draws in modern chess. This rule states that if the same
position appears three times, the players can claim a draw:

\begin{quote}
\begin{itemize}
  \item[9.2.2.] Positions are considered the same if and only if the
same player has the move, pieces of the same kind and colour occupy
the same squares and the possible moves of all the pieces of both
players are the same. Thus positions are not the same if:
\begin{itemize}
\item[9.2.2.1.]	at the start of the sequence a pawn could have been captured en passant
\item[9.2.2.2.]	a king had castling rights with a rook that has not been moved, but forfeited these after moving. The castling rights are lost only after the king or rook is moved.
\end{itemize}
\end{itemize}
\end{quote}

Like the 75-move rule, this rule has an optional version (upon
three repetitions) and a mandatory one in 9.6:

\begin{quote}
  [The game is a draw if \ldots] \\
  9.6.1. the same position has appeared, as in 9.2.2, at least five times.
\end{quote}

% (I seem to remember that this rule used to be about moves, not positions,
% but someone discovered infinite sequences of moves that don't repeat, right?
% I can't find this citation. Maybe can recreat those sequences.)
%
% Fascinatingly (at least as fascinating as anything in this
% questionable undertaking), this rule

\medskip
Many implementations of chess ignore these rules or treat them
incorrectly. Implementation of the seventy-five move rule simply
requires a count of how many moves have transpired since a pawn move
or capture, but programs typically do not force a draw after 75 moves.
Repetition requires more work, since the program must keep track of
each of the states reached since the last irreversible move. There are
also some corner cases, such as ambiguity as to whether the starting
position has ``appeared'' before the first move~\cite{survival}. The
ubiquitous FEN notation for describing chess positions does not even
include any information about states previously reached.

\subsection{Dead position} \label{sec:insufficient}

The informal version of this rule (``insufficient material'') states
that if neither side has enough pieces to mate the opponent (for
example, a king and bishop can never mate a bare king) then the game
is draw. Again, the formal rule is more subtle:

\begin{quote}
5.2.2. The game is drawn when a position has arisen in which neither
player can checkmate the opponent's king with any series of legal
moves. The game is said to end in a `dead position'. This immediately
ends the game \ldots
\end{quote}

This clearly includes the well-known material-based cases like king
and knight vs.~king, but it also surprisingly includes many other
specific positions, especially those with forced captures
(Figure~\ref{fig:insufficient}).

\begin{figure}[htp]
  \begin{center}
    \chessboard[setfen=8/8/5K2/8/8/8/Q7/k7 b - - 0 1]
  \end{center}
  \caption{Black to move and draw in 0 (!). Most players and even
    chexperts believe that the only legal move is Kxa2, and
    that the game then ends in a draw with ``insufficient material.''
    However, this game is {\em already over}. Since neither black
    nor white can win via any series of legal moves, by rule 5.2.2
    the game immediately ends in a `dead position'. (Although see
    Section~\ref{sec:ambiguity} for possible ambiguity in this
    rule.)} \label{fig:insufficient}
\end{figure}

The insufficient material rule is curious in that it requires
nontrivial computation to implement. In order to know whether the
position is a draw, an implementation needs to be able to decide
whether or not a series of moves that results in checkmate exists.
Note that this is not nearly as bad as normal game tree search because
the two sides can collaborate to produce the mate (it is not a
$\exists \forall \exists \forall \ldots$ but rather $\exists \exists
\exists \exists \ldots$). Still, such ``helpmates'' can still be
quite deep (dozens of moves) and are interesting enough to be a
common source of chess puzzles. Proving the non-existence of a
helpmate can be very difficult indeed (Figure~\ref{fig:nohelp}).

\begin{figure}[htp]
  \begin{center}
    \chessboard[setfen=k1bB4/8/2p1p1p1/1pPpPpP1/1P1P1P2/8/8/K1Bb4 w - - 0 1]
  \end{center}
  \caption{Thinking of implementing the rules of chess? To be correct,
    you'll need your program to be able to deduce that no helpmates are
    possible in this position and thus the game is over. Stockfish rates
    this as +0.4 for white, even searching to depth 92.
    (Position is due to user {\tt supercat} on Chess StackExchange.)
  } \label{fig:nohelp}
\end{figure}

% XXX might be a good place to talk about the self-referential "maximal"
% and "minimal" interpretations of this rule?

\medskip
For completeness, note that there are other routes to a draw
(stalemate, draw offers) which we can ignore; it is easy to avoid
these situations when generating the longest game.

\section{Generating the longest game}

It is generally not hard to avoid repeating positions, so the main
obstacle we'll face is the 75-move rule. Let's call an irreversible
move that resets the 75-move counter a {\em critical move}; this is
a pawn move or capturing move (or both). The structure of the game
will be a series of critical moves (I will call these ``critical''
moves) with a maximal sequence of pointless reversible moves in
between them. If we execute the maximum number of critical moves and
make 149 half-moves (just shy of triggering the compulsory version of
the 75-move rule) between them, then this will be fairly easily
seen as a maximal game.

In fact, for most positions, it is easy to waste 149 moves and return
to the exact same position. So, the strategy for generating the
longest game can mostly be broken into two phases: Make a game with a
maximum number of critical moves (it can also contain other moves) and
then pad that game out to maximum-length excursions in between its
critical moves.

\subsection{Maximal critical moves}

Critical moves are pawn moves and captures. There are 16 pawns, each
of which has 6 squares to move into before promoting, so this is
$16\times 6$ critical moves. There are 14 capturable pieces, plus the
16 pawns (they can be captured after promoting) so this nets an
additional $16+14$ critical moves for a total of $126$. Each critical
move can be made after a maximum of $75+74$ reversible moves, giving
$(75+74+1) \times 126 = 18900$ moves. The final move would capture the
last piece, yielding a draw due to the remaining kings being insufficient
material (``dead position''). This is the upper bound.

We will not quite be able to use the entire critical move budget. If
pawns only move forward in single steps, they will eventually get
blocked by the opposing pawn on the same file. Pawns can move
diagonally off their starting file only by capturing. We have plenty
of capturing to do anyway, so this in no problem. With four captures
per side, the pawns can be doubled, with a clear route to promotion,
like so:

\begin{center}
\chessboard[setfen=2bqkb2/1p1p1p1p/1p1p1p1p/8/8/P1P1P1P1/P1P1P1P1/2BQKB2 w - - 0 1]
\end{center}

Each of these captures is {\em both} a pawn move {\em and} a capturing
move, however. This means that we lose $4+4$ critical moves off our
total budget. $150 \times 118 = 17700$ is the new upper bound.

\paragraph{Parity.} If a critical move (such as a pawn move) is made by
white, then the first pointless excursion is made by black. The
players alternate these pointless moves until black has made 75 and
white 74. Now it is white's move, and white must make a critical move
or the game ends due to the 75-move rule. If instead we wanted black
to make the next critical move, this would happen after black has made
74 and white 74 non-critical moves. Any time this happens we lose one
move against the upper bound. So, we want to minimize the number of
times we switch which player is making the critical moves. Obviously
we must switch at least once, because both black and white must
make critical moves (upper bound $17699$).

\paragraph{Starting condition.} The first critical move should be made
by black. The starting position (with white to move) is analogous to
the situation just described, as if black has just made a critical move,
and it's white's turn. White will play 75 reversible moves, black plays
74, and the 150th move is black making the first critical move.

Note that white is quite constrained during this beginning phase,
unable to make any pawn moves. Only the white knights can escape the
back rank. When we try to insert $149$ pointless moves, we'll only be
able to move the knights and rooks, and doing this 75 times must leave
e.g. one of the knights on an opposite-colored square.\footnote{Knight
  moves always change the color of the knight's square, and same for
  rook moves constrained to the a1/b1 or g1/h1 squares.} So we have to
be a little careful about the position in which we make black's first
critical move.

Since white can only free their two knights, it will not be possible
for black to double four sets of their pawns (each requires capturing
a white piece). Executing as many critical moves as possible get us
to a position like

\begin{center}
\chessboard[setfen=r1bqkb2/8/5r2/p1pn4/5p1p/1p1p1p1p/PPPPPPPP/n3K3 w q - 0 47]
\end{center}

\ldots at which point we need to switch to making critical moves by white ({\bf Switch 1}).
White can promote their unblocked pawns,\footnote{White could promote all their pawns, but this would require.. actually maybe this is possible? Try...
  % In this game you get stuck at:
  % 1N1NN1RQ/4k3/7N/p1p2b1Q/5p1p/1p1p1p1p/1P6/3K4 b - - 0 95

  )
}
making progress to a position
like

\begin{center}
  \chessboard[setfen=2krQ3/8/7N/p1p2b2/PRPQ1p1p/PpPp1p1p/8/5K2 b - - 2 86]
\end{center}

\ldots and then we switch back to black again ({\bf Switch 2}). Black
promotes the remaining pawns, giving something like

\begin{center}
  \chessboard[setfen=2krbK2/8/7r/8/P1P5/PnP5/1n1n1n2/1n1n1n2 w - - 0 118]
\end{center}

\ldots with nothing left to do. We need to promote the white pawns
before capturing them. So we switch back to white ({\bf Switch 3}) and
promote them all. White can capture all of black's remaining pieces
during this phase of critical moves, resulting in a board like

\begin{center}
  \chessboard[setfen=N2Q1K2/N7/N7/8/8/7k/8/8 b - - 0 149]
\end{center}

\comment{
  position after above: N2Q1K2/N7/N7/8/8/7k/8/8 b - - 0 149

   proof game:
  1. Nf3 h6 2. Ng1 h5 3. Nf3 f5 4. Nc3 d5
5. Ne4 b5 6. Ne5 Rh6 7. Ng4 Rc6 8. Nh6 gxh6
9. Nf6+ exf6 10. Rb1 Na6 11. Ra1 Nb4 12. Rb1 Ne7
13. Ra1 Ng6 14. Rb1 Nf4 15. Ra1 Nh3 16. Rg1 Nxg1
17. Rb1 Nh3 18. Ra1 Nf4 19. Rb1 d4 20. Ra1 Nbd5
21. Rb1 Ne3 22. Ra1 Nxf1 23. Rb1 Ne3 24. Ra1 Nxd1
25. Rb1 Ne3 26. Ra1 Ned5 27. Rb1 Nd3+ 28. Kf1 Nxc1
29. Ra1 h4 30. Rb1 h3 31. Ra1 h5 32. Rb1 h4
33. Ra1 f4 34. Rb1 f5 35. Ra1 f3 36. Rb1 f4
37. Ra1 d3 38. Rb1 b4 39. Ra1 Nb3 40. Rb1 a6
41. Ra1 a5 42. Rb1 Rf6 43. Ra1 c6 44. Rb1 c5
45. Ra1 Nxa1 46. Ke1 b3 47. g3 Re6 48. g4 Rf6
49. g5 Re6 50. g6 Rf6 51. g7 Re6 52. g8=Q Rf6
53. Qg5 Re6 54. Qe5 Bg7 55. Qd4 Qg5 56. Kd1 Qg3
57. hxg3 Bh6 58. g4 Bg7 59. g5 Bh6 60. g6 Bg5
61. g7 Bh6 62. g8=N Bd7 63. e3 O-O-O 64. e4 Rd6
65. e5 Rc6 66. e6 Rb6 67. e7 Rc6 68. e8=Q Rg6
69. c3 Rg3 70. fxg3 Bc6 71. Nxh6 Bd7 72. g4 Bc6
73. g5 Bd7 74. g6 Bc6 75. g7 Bd7 76. g8=R Nf6
77. a3 Nd5 78. a4 Nf6 79. c4 Nd5 80. Rg6 Nc3+
81. Ke1 Nc2+ 82. Kf1 Na3 83. dxc3 Bf5 84. Rb6 Bd7
85. bxa3 Bf5 86. Rb4 axb4 87. Qf8 cxd4 88. Qe8 b2
89. Qf8 b1=N 90. Qe8 d2 91. Qf8 d1=N 92. Qe8 b3
93. Qf8 b2 94. Qe8 Nd2+ 95. Kg1 b1=N 96. Qf8 Nb2
97. Qe8 Nb3 98. Qf8 d3 99. Qe8 d2 100. Qf8 d1=N
101. Kf1 f2 102. Ke2 f1=N 103. Kf3 h2 104. Qe8 Bd7
105. Ke4 f3 106. Qf7 Nfd2+ 107. Kf4 f2 108. Kg5 f1=N
109. Kg6 h1=N 110. Kf6 h3 111. Kg6 h2 112. Kf6 Nhf2
113. Kg6 h1=R 114. Kg7 Be6 115. Kh7 Bxf7 116. Kg7 Be8
117. Kf8 Rxh6 118. a5 Rh5 119. a6 Rh4 120. a7 Rh3
121. a8=N Rh2 122. a4 Rh3 123. a5 Rh4 124. a6 Rh5
125. a7 Rh4 126. Nc7 Rh3 127. a8=N Kd7 128. Na6 Ke6
129. c5 Kf6 130. c6 Kg6 131. c7 Kh5 132. c8=N Rh2
133. c4 Rh3 134. c5 Rh2 135. c6 Rh1 136. c7 Rh2
137. Na7 Rh1 138. c8=Q Nc5 139. Qxc5+ Kh4 140. Qxf2+ Kh5
141. Qxf1 Kh6 142. Qxd1 Kh7 143. Qxd2 Kg6 144. Qxb2 Kg5
145. Qxb1 Kh4 146. Qxh1+ Kg5 147. Qc6 Kh4 148. Qxe8 Kh3
149. Qxd8
  }

But !

// The way we'll find this game is to manually generate a slow game,
// that is, one where we make make 16*6 pawn moves (but 8 are
captures) // and capture all the pieces except for a single queen, and
then // capture it, too. Then all that remains is to artificially slow
// down this game even more, by inserting 75+74 half-moves that get us
back // to the position we're already in (without repeating 5 times).
The // hypothesis is that such sequences are numerous, so we can find
// them easily (perhaps by just randomly moving a piece around, or //
perhaps by lazily generating some kind of graph structure).

\subsection{Better approach}

OK, need to rewrite the above for this superior game:

1. Nf3 b6 2. Nc3 g6 3. Ne5 g5 4. Nd5 b5
5. Nc4 bxc4 6. Nf4 gxf4 7. Rg1 c6 8. Rh1 c5
9. Rg1 f6 10. Rh1 f5 11. b3 Nc6 12. b4 Nd4
13. b5 Nb3 14. cxb3 Bb7 15. a3 Rc8 16. a4 Rc6
17. a5 Rd6 18. a6 Re6 19. axb7 Rd6 20. b8=N Re6
21. b6 Nh6 22. b7 Ng4 23. Nc6 Bh6 24. b8=Q Kf8
25. b4 Ke8 26. b5 Kf8 27. Qe5 Ke8 28. b6 Kf8
29. b7 Ke8 30. b8=N Bg5 31. h3 Bh4 32. hxg4 Bg3
33. fxg3 Qc8 34. g5 Qa6 35. Nb4 Kd8 36. Qe3 Kc7
37. g6 Kb6 38. g7 Rf6 39. g8=Q Re6 40. g4 Rf6
41. g5 Re6 42. g6 Rf6 43. g7 Re6 44. Qf8 Rf6
45. g8=N Rd6 46. Nf6 Qb5 47. Nh5 Rg8 48. g3 Rgg6
49. g4 Rge6 50. g5 Rf6 51. g6 Rc6 52. g7 Qa6
53. g8=N Qa5 54. d3 Qa6 55. d4 Qa5 56. d5 Qa6
57. dxc6 Qa5 58. c7 Qa6 59. c8=R Qa4 60. Qh3 Qa6
61. e3 Qa5 62. e4 Qa6 63. e5 Qa5 64. exf6 Qa6
65. Qg7 Qa5 66. f7 Qa6 67. Re8 Qa5 68. f8=R Qa6
69. Qdg4 Qa5 70. Nc6 Qa6 71. Ne5 Qa5 72. Nf3 Qa6
73. Nd3 Qa5+ 74. Ke2 Qa6 75. Nf2 Qa5 76. Qh8 Qa3
77. Nd1 Qa2+ 78. Nd2 Qa3 79. Kf2 Qa2 80. Kg1 Qa3
81. Rxa3 f3 82. Kh2 f2 83. Bd3 f1=R 84. Qf4 Rxf4
85. Ng7 h6 86. Be4 Rxe4 87. Ne6 f4 88. Nf6 f3
89. Ng8 f2 90. Nf6 f1=B 91. Rd3 Bxd3 92. Bb2 a6
93. Bc3 a5 94. Bb2 a4 95. Bc3 a3 96. Bb2 a2
97. Bc3 a1=R 98. Ba5+ Rxa5 99. Nf4 c3 100. Ne2 c2
101. Nf4 c1=N 102. Ne2 c4 103. Ng4 Nxe2 104. Nge3 c3
105. Nf5 c2 106. Nfe3 c1=R 107. Re1 h5 108. Qf3 h4
109. Kg2 h3+ 110. Kf2 h2 111. Qfh5 h1=R 112. Ng4 Rxg4
113. Nc4+ Bxc4 114. Nb2 d6 115. Rg1 d5 116. Rg2 d4
117. Kf3 d3 118. Kf2 d2 119. Qh2 d1=R 120. Kf3 Nc3
121. Kf2 Rg8 122. Kf3 e6 123. Kf4 e5+ 124. Kf5 e4+
125. Kf6 e3 126. Rg3 e2 127. Rg2 e1=Q 128. Nd3 Rxd3
129. Rg1 Rgxg1 130. Q8h3 Rxh3 131. Ra8 Rxa8 132. Rb8+ Rxb8
133. Qxb8+ Ka5 134. Qg3 Bf7 135. Qxe1 Rg5 136. Qxc1 Ka6
137. Qxc3 Kb6 138. Qxh3 Kc5 139. Kxf7 Kc6 140. Qxh1+ Kd7
141. Kf6 Ke8 142. Qg2 Rg8 143. Ke6 Rg7 144. Qxg7 Kd8
145. Qd7\#

Black's first phase results in this:
\chessboard[setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11]

We made two captures, and each of white's pawns can be freed with
at most one capture, and two are already clear. So this totals 8 pawn
captures, the (conjectured) minimum. (Switch 1) Now white's phase:

White's first phase then ends here:

\chessboard[setfen=4RRNQ/p2pp2p/1k6/2p2p1N/2p2pQ1/R6Q/3N4/2BN1BKR b - - 0 81]

We promoted all of white's pieces, and captured all of black's. (Switch 2).
Black again. Black promotes all his pawns:

\chessboard[setfen=4RRrQ/8/1k3K2/r7/2b5/2n5/1N4RQ/2rrq2r w - - 0 128]

... and then proceeds to capture all of white's remaining pieces:

\chessboard[setfen=1r6/8/1k3K2/8/2b5/7r/4n3/2r1q1rr w - - 0 134]

Now we switch back to white making critical moves (Switch 3). White mops up
all black's pieces with the king, which is the only piece left.

... this ends up not working; when black mates the white queen is
black's 75th move, but white has only made 74. So it should really be
white that mates black! (this is kinda like the asymmetry in the starting
condition.)

So instead at this moment...

\chessboard[setfen=1r6/8/1k3K2/8/2b5/2n4r/7Q/2r1q1rr w - - 0 133]

.. we leave the white queen alive, and capture all black's pieces for real!
then mate with the white queen

\chessboard[setfen=3k4/3Q4/4K3/8/8/8/8/8 b - - 2 145]

%8/8/2k5/4K3/7q/8/8/8 w - - 1 1
% 8846. Ke6 [72 72] Kc7 [72 73] 8847. Kf7 [73 73] Kd6 [73 74] 8848. Ke8 [74 74] Q
% e7 [74 75]Slack 0: 114 time(s)

\subsection{Inserting excursions} \label{sec:excursions}

\section{A longest game}

\begin{sloppypar}
\tiny
\setstretch{0.1}
\hyphenpenalty=1
\pretolerance=2
\tolerance=2

(a long game goes here)

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad
minim veniam, quis nostrud exercitation ullamco laboris nisi ut
aliquip ex ea commodo consequat. Duis aute irure dolor in
reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla
pariatur. Excepteur sint occaecat cupidatat non proident, sunt in
culpa qui officia deserunt mollit anim id est laborum.

 \noindent \input{slow}
\quad 1-0

\end{sloppypar}

\nocite{chesstego}

\bibliography{chess}{}
\bibliographystyle{plain}

\end{document}

### 2020-03-09 trunk/chess/papers/survival.tex
Some work on longest paper/code. Shaved one slack move off, so feeling pretty confident in the argument that this is the longest...

@@
 
-\clearpage
-
 \newcommand{\chessmap}[2]{\begin{center} \includegraphics[width=0.65 \linewidth]{#1} {#2} \end{center}\vspace{-0.1in}}
@@
 
-\clearpage
-
 Aside from admiring the groovy pictures, the data can also be used as
@@
 Like all good scientific research, I explicitly compare the actual
-results to the hypotheses gathered before the experiment; this is a
+results to the hypotheses gathered before the experiment; this is an
 hygenic and humbling exercise. Figure~\ref{fig:buddies} compares the

### 2020-03-10 trunk/chess/papers/longest.tex
Clean up the critical moves section

@@
 % XXX should match the length of the game
-\textcurrency 17697
+\textcurrency 17,697
 }
@@
 than a protracted XXX
+
+(shortest game is well known)
 
@@
 % and "minimal" interpretations of this rule?
+% Need to have a  \label{sec:ambiguity} somewhere.
 
@@
 per side, the pawns can be doubled, with a clear route to promotion,
-like so:
+like in Figure~\ref{fig:doubled}.
 
+\begin{figure}[h]
 \begin{center}
   \chessboard[setfen=2bqkb2/1p1p1p1p/1p1p1p1p/8/8/P1P1P1P1/P1P1P1P1/2BQKB2 w - - 0 1]
+  \caption{One way to clear the promotion routes for all pawns with
+    eight captures. } \label{fig:doubled}
 \end{center}
+\end{figure}
 
@@
 move, however. This means that we lose $4+4$ critical moves off our
-total budget. $150 \times 118 = 17700$ is the new upper bound.
+total budget. $150 \times (126 - 8) = 17700$ is the new upper bound.
 
 \paragraph{Parity.} If a critical move (such as a pawn move) is made by
-white, then the first pointless excursion is made by black. The
-players alternate these pointless moves until black has made 75 and
-white 74. Now it is white's move, and white must make a critical move
+white, then the first non-critical move is made by black. The
+players alternate these pointless moves until black has made $75$ and
+white $74$. Now it is white's move, and white must make a critical move
 or the game ends due to the 75-move rule. If instead we wanted black
 to make the next critical move, this would happen after black has made
-74 and white 74 non-critical moves. Any time this happens we lose one
+$74$ and white $74$ non-critical moves. Any time this happens we lose one
 move against the upper bound. So, we want to minimize the number of
@@
 we must switch at least once, because both black and white must
-make critical moves (upper bound $17699$).
+make critical moves. This reduces the upper bound to $17699$.
 
@@
 the situation just described, as if black has just made a critical move,
-and it's white's turn. White will play 75 reversible moves, black plays
-74, and the 150th move is black making the first critical move.
+and it's white's turn. White will play $75$ reversible moves, black plays
+$74$, and the 150\th\ move is black making the first critical move.
 
-Note that white is quite constrained during this beginning phase,
-unable to make any pawn moves. Only the white knights can escape the
-back rank. When we try to insert $149$ pointless moves, we'll only be
-able to move the knights and rooks, and doing this 75 times must leave
-e.g. one of the knights on an opposite-colored square.\footnote{Knight
-  moves always change the color of the knight's square, and same for
-  rook moves constrained to the a1/b1 or g1/h1 squares.} So we have to
-be a little careful about the position in which we make black's first
-critical move.
+Note that white is quite constrained during this beginning phase, as
+pawn moves are critical moves and must be avoided. Only the white
+knights can escape the back rank. When we try to insert $149$
+pointless moves, we'll only be able to move the knights and rooks, and
+doing this $75$ times must leave e.g. one of the knights on an
+opposite-colored square.\footnote{Knight moves always change the color
+  of the knight's square, and same for rook moves constrained to the
+  a1/b1 or g1/h1 squares.} So we have to be a little careful about the
+position in which we make black's first critical move.
 
-Since white can only free their two knights, it will not be possible
-for black to double four sets of their pawns (each requires capturing
-a white piece). Executing as many critical moves as possible get us
-to a position like
+Since white can only free their two knights, these are the only pieces
+that can be captured by black pawns. So it will not be possible for
+black to double four sets of their pawns as in
+Figure~\ref{fig:doubled}. This would require white to have a phase of
+critical moves to free moves to capture, then black again to finish
+doubling pawns, then white again to promote its freed pawns. Each switch
+costs move off the theoretical max. We can be more efficient with an
+asymmetric approach.
 
+Black's first phase of critical moves instead results in this:
+
+\comment{
+1. Nf3 b6 2. Nc3 g6 3. Ne5 g5 4. Nd5 b5
+5. Nc4 bxc4 6. Nf4 gxf4 7. Rg1 c6 8. Rh1 c5
+9. Rg1 f6 10. Rh1 f5 
+}
 \begin{center}
-\chessboard[setfen=r1bqkb2/8/5r2/p1pn4/5p1p/1p1p1p1p/PPPPPPPP/n3K3 w q - 0 47]
+\chessboard[setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11]
 \end{center}
 
-\ldots at which point we need to switch to making critical moves by white ({\bf Switch 1}).
-White can promote their unblocked pawns,\footnote{White could promote all their pawns, but this would require.. actually maybe this is possible? Try...
-  % In this game you get stuck at:
-  % 1N1NN1RQ/4k3/7N/p1p2b1Q/5p1p/1p1p1p1p/1P6/3K4 b - - 0 95
-
-  )
-}
-making progress to a position
-like
-
-\begin{center}
-  \chessboard[setfen=2krQ3/8/7N/p1p2b2/PRPQ1p1p/PpPp1p1p/8/5K2 b - - 2 86]
-\end{center}
-
-\ldots and then we switch back to black again ({\bf Switch 2}). Black
-promotes the remaining pawns, giving something like
-
-\begin{center}
-  \chessboard[setfen=2krbK2/8/7r/8/P1P5/PnP5/1n1n1n2/1n1n1n2 w - - 0 118]
-\end{center}
-
-\ldots with nothing left to do. We need to promote the white pawns
-before capturing them. So we switch back to white ({\bf Switch 3}) and
-promote them all. White can capture all of black's remaining pieces
-during this phase of critical moves, resulting in a board like
-
-\begin{center}
-  \chessboard[setfen=N2Q1K2/N7/N7/8/8/7k/8/8 b - - 0 149]
-\end{center}
+The white b and g pawns have a clear route to promotion. White can
+free each of the remaining 6 pawns with a single capture. Between this
+and black's two pawn captures, this is the minimum of 8 pawn moves
+that are also captures. Since black has plenty of freed pieces, white
+can promote all of their pawns during their own phase of critical
+moves, resulting in:
 
 \comment{
-  position after above: N2Q1K2/N7/N7/8/8/7k/8/8 b - - 0 149
-
-   proof game:
-  1. Nf3 h6 2. Ng1 h5 3. Nf3 f5 4. Nc3 d5
-5. Ne4 b5 6. Ne5 Rh6 7. Ng4 Rc6 8. Nh6 gxh6
-9. Nf6+ exf6 10. Rb1 Na6 11. Ra1 Nb4 12. Rb1 Ne7
-13. Ra1 Ng6 14. Rb1 Nf4 15. Ra1 Nh3 16. Rg1 Nxg1
-17. Rb1 Nh3 18. Ra1 Nf4 19. Rb1 d4 20. Ra1 Nbd5
-21. Rb1 Ne3 22. Ra1 Nxf1 23. Rb1 Ne3 24. Ra1 Nxd1
-25. Rb1 Ne3 26. Ra1 Ned5 27. Rb1 Nd3+ 28. Kf1 Nxc1
-29. Ra1 h4 30. Rb1 h3 31. Ra1 h5 32. Rb1 h4
-33. Ra1 f4 34. Rb1 f5 35. Ra1 f3 36. Rb1 f4
-37. Ra1 d3 38. Rb1 b4 39. Ra1 Nb3 40. Rb1 a6
-41. Ra1 a5 42. Rb1 Rf6 43. Ra1 c6 44. Rb1 c5
-45. Ra1 Nxa1 46. Ke1 b3 47. g3 Re6 48. g4 Rf6
-49. g5 Re6 50. g6 Rf6 51. g7 Re6 52. g8=Q Rf6
-53. Qg5 Re6 54. Qe5 Bg7 55. Qd4 Qg5 56. Kd1 Qg3
-57. hxg3 Bh6 58. g4 Bg7 59. g5 Bh6 60. g6 Bg5
-61. g7 Bh6 62. g8=N Bd7 63. e3 O-O-O 64. e4 Rd6
-65. e5 Rc6 66. e6 Rb6 67. e7 Rc6 68. e8=Q Rg6
-69. c3 Rg3 70. fxg3 Bc6 71. Nxh6 Bd7 72. g4 Bc6
-73. g5 Bd7 74. g6 Bc6 75. g7 Bd7 76. g8=R Nf6
-77. a3 Nd5 78. a4 Nf6 79. c4 Nd5 80. Rg6 Nc3+
-81. Ke1 Nc2+ 82. Kf1 Na3 83. dxc3 Bf5 84. Rb6 Bd7
-85. bxa3 Bf5 86. Rb4 axb4 87. Qf8 cxd4 88. Qe8 b2
-89. Qf8 b1=N 90. Qe8 d2 91. Qf8 d1=N 92. Qe8 b3
-93. Qf8 b2 94. Qe8 Nd2+ 95. Kg1 b1=N 96. Qf8 Nb2
-97. Qe8 Nb3 98. Qf8 d3 99. Qe8 d2 100. Qf8 d1=N
-101. Kf1 f2 102. Ke2 f1=N 103. Kf3 h2 104. Qe8 Bd7
-105. Ke4 f3 106. Qf7 Nfd2+ 107. Kf4 f2 108. Kg5 f1=N
-109. Kg6 h1=N 110. Kf6 h3 111. Kg6 h2 112. Kf6 Nhf2
-113. Kg6 h1=R 114. Kg7 Be6 115. Kh7 Bxf7 116. Kg7 Be8
-117. Kf8 Rxh6 118. a5 Rh5 119. a6 Rh4 120. a7 Rh3
-121. a8=N Rh2 122. a4 Rh3 123. a5 Rh4 124. a6 Rh5
-125. a7 Rh4 126. Nc7 Rh3 127. a8=N Kd7 128. Na6 Ke6
-129. c5 Kf6 130. c6 Kg6 131. c7 Kh5 132. c8=N Rh2
-133. c4 Rh3 134. c5 Rh2 135. c6 Rh1 136. c7 Rh2
-137. Na7 Rh1 138. c8=Q Nc5 139. Qxc5+ Kh4 140. Qxf2+ Kh5
-141. Qxf1 Kh6 142. Qxd1 Kh7 143. Qxd2 Kg6 144. Qxb2 Kg5
-145. Qxb1 Kh4 146. Qxh1+ Kg5 147. Qc6 Kh4 148. Qxe8 Kh3
-149. Qxd8
-  }
-
-But !
-
-// The way we'll find this game is to manually generate a slow game,
-// that is, one where we make make 16*6 pawn moves (but 8 are
-captures) // and capture all the pieces except for a single queen, and
-then // capture it, too. Then all that remains is to artificially slow
-// down this game even more, by inserting 75+74 half-moves that get us
-back // to the position we're already in (without repeating 5 times).
-The // hypothesis is that such sequences are numerous, so we can find
-// them easily (perhaps by just randomly moving a piece around, or //
-perhaps by lazily generating some kind of graph structure).
-
-\subsection{Better approach}
-
-OK, need to rewrite the above for this superior game:
-
-1. Nf3 b6 2. Nc3 g6 3. Ne5 g5 4. Nd5 b5
-5. Nc4 bxc4 6. Nf4 gxf4 7. Rg1 c6 8. Rh1 c5
-9. Rg1 f6 10. Rh1 f5 11. b3 Nc6 12. b4 Nd4
+11. b3 Nc6 12. b4 Nd4
 13. b5 Nb3 14. cxb3 Bb7 15. a3 Rc8 16. a4 Rc6
@@
 77. Nd1 Qa2+ 78. Nd2 Qa3 79. Kf2 Qa2 80. Kg1 Qa3
-81. Rxa3 f3 82. Kh2 f2 83. Bd3 f1=R 84. Qf4 Rxf4
+81. Rxa3
+}
+\begin{center}
+  \chessboard[setfen=4RRNQ/p2pp2p/1k6/2p2p1N/2p2pQ1/R6Q/3N4/2BN1BKR b - - 0 81]
+\end{center}
+
+Now black can promote all of its pawns and capture white's pieces.
+We actually leave the white queen; this turns out to be essential:
+
+\comment{
+\ldots f3 82. Kh2 f2 83. Bd3 f1=R 84. Qf4 Rxf4
 85. Ng7 h6 86. Be4 Rxe4 87. Ne6 f4 88. Nf6 f3
@@
 129. Rg1 Rgxg1 130. Q8h3 Rxh3 131. Ra8 Rxa8 132. Rb8+ Rxb8
+}
+
+\begin{center}
+  \chessboard[setfen=1r6/8/1k3K2/8/2b5/2n4r/7Q/2r1q1rr w - - 0 133]
+\end{center}
+
+Finally, white captures all of black's pieces, and mates the black king.
+
+\comment{
 133. Qxb8+ Ka5 134. Qg3 Bf7 135. Qxe1 Rg5 136. Qxc1 Ka6
@@
 145. Qd7\#
+}
 
-Black's first phase results in this:
-\chessboard[setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11]
+\begin{center}
+  \chessboard[setfen=3k4/3Q4/4K3/8/8/8/8/8 b - - 2 145]
+\end{center}
 
-We made two captures, and each of white's pawns can be freed with
-at most one capture, and two are already clear. So this totals 8 pawn
-captures, the (conjectured) minimum. (Switch 1) Now white's phase:
+(this is three switches)
 
-White's first phase then ends here:
+\paragraph{Ending condition.} The way this game ends is subtle for
+several reasons. First, note that we left the white queen on the board
+and used it for mate. It is required that white be the one mating for
+parity reasons, similar to the reason black must make the first critical
+move. Since white makes critical moves in the last phase, black leads
+on non-critical moves; at the moment white makes the checkmating move,
+black has made 75 non-critical moves, and white 74. White's 75th move
+mates.
 
-\chessboard[setfen=4RRNQ/p2pp2p/1k6/2p2p1N/2p2pQ1/R6Q/3N4/2BN1BKR b - - 0 81]
+But doesn't this trigger the 75-move rule? No, this rule (9.6.2;
+Section~\ref{sec:75}) has a special exception for checkmate: ``If the
+last move resulted in checkmate, that shall take precedence.''
+Essentially, we can treat checkmate as a type of critical move.
 
-We promoted all of white's pieces, and captured all of black's. (Switch 2).
-Black again. Black promotes all his pawns:
+Why not capture the white queen too? This would be a critical move,
+but once two kings remain, the game ends immediately in a draw. So
+there is no length advantage here over checkmate. In fact we could do
+worse: Consider the position in Figure~\ref{fig:insufficient} where
+black is {\em forced} to capture the queen. This game is over prior to
+the capture, so this is shorter than the checkmating sequence! Even if
+the king had an escape square, it is arguable
+(Section~\ref{sec:ambiguity}) that the game is over in any case: There
+is no legal sequence of moves that leads to mate. Either the king
+captures the queen (and then clearly no mate is possible with just the
+two kings) or the king escapes, but in doing so plays the
+75\th\ non-critical move, and triggers a draw by the 75-move rule.
 
-\chessboard[setfen=4RRrQ/8/1k3K2/r7/2b5/2n5/1N4RQ/2rrq2r w - - 0 128]
-
-... and then proceeds to capture all of white's remaining pieces:
-
-\chessboard[setfen=1r6/8/1k3K2/8/2b5/7r/4n3/2r1q1rr w - - 0 134]
-
-Now we switch back to white making critical moves (Switch 3). White mops up
-all black's pieces with the king, which is the only piece left.
-
-... this ends up not working; when black mates the white queen is
-black's 75th move, but white has only made 74. So it should really be
-white that mates black! (this is kinda like the asymmetry in the starting
-condition.)
-
-So instead at this moment...
-
-\chessboard[setfen=1r6/8/1k3K2/8/2b5/2n4r/7Q/2r1q1rr w - - 0 133]
-
-.. we leave the white queen alive, and capture all black's pieces for real!
-then mate with the white queen
-
-\chessboard[setfen=3k4/3Q4/4K3/8/8/8/8/8 b - - 2 145]
-
-%8/8/2k5/4K3/7q/8/8/8 w - - 1 1
-% 8846. Ke6 [72 72] Kc7 [72 73] 8847. Kf7 [73 73] Kd6 [73 74] 8848. Ke8 [74 74] Q
-% e7 [74 75]Slack 0: 114 time(s)
+The most foolproof way to ensure that mate is always possible is for the
+game itself to end in mate. This sidesteps any ambiguity about the way
+the 75-move condition should be interpreted, as well.
 
 \subsection{Inserting excursions} \label{sec:excursions}
+
+TODO
+
+\section{Future work}
+
+Some puzzles?
 

### 2020-03-15 trunk/chess/papers/longest.tex
Almost done with this boring paper

@@
 \usepackage{textcomp}
+
+\usepackage{adjustbox}
 
@@
 Verlag-Verlag volume no.~0x40-2A.
-% XXX should match the length of the game
 \textcurrency 17,697
@@
 
-Many chess moves are reversible (e.g. moving the knight forward and
+Many chess moves are reversible (e.g.~moving the knight forward and
 back to its starting position~\cite{survival}), so informal games of
@@
 What is the longest chess game? In this paper I show how to compute
-such a game, and then gratuitously present all of its
-% XXX check that this is the final one
-17,697 
-%
-half-moves. Even if you are a chess expert (``chexpert''), I bet you
+such a game, and then gratuitously present all of its 17,697 
+moves. Even if you are a chess expert (``chexpert''), I bet you
 will be surprised at some of the corner cases in the rules that are
@@
 
-Speaking of rules, let's first detail the two main rules that limit the
-length of the game.
+Speaking of rules, let's first detail the three main rules that limit
+the length of the game. These rules cause the game to end in a draw
+(tie) when certain conditions are met.
 
@@
 
-Draws after 75 moves (per player, so really 150 half-moves) are
-compulsory.
+Draws after 75 moves (per player, so really 150 moves) are compulsory.
 
@@
 % but someone discovered infinite sequences of moves that don't repeat, right?
-% I can't find this citation. Maybe can recreat those sequences.)
+% I can't find this citation. Maybe can recreate those sequences.)
 %
@@
 example, a king and bishop can never mate a bare king) then the game
-is draw. Again, the formal rule is more subtle:
+is drawn. Again, the formal rule is more subtle:
 
@@
 moves. The game is said to end in a `dead position'. This immediately
-ends the game \ldots
+ends the game\ldots
 \end{quote}
@@
 $\exists \forall \exists \forall \ldots$ but rather $\exists \exists
-\exists \exists \ldots$). Still, such ``helpmates'' can still be
-quite deep (dozens of moves) and are interesting enough to be a
-common source of chess puzzles. Proving the non-existence of a
-helpmate can be very difficult indeed (Figure~\ref{fig:nohelp}).
+\exists \exists \ldots$). Still, such ``helpmates'' can still be quite
+deep (dozens of moves) and are interesting enough to be a common
+source of chess puzzles. Proving the non-existence of a helpmate can
+be very difficult indeed (Figure~\ref{fig:nohelp}).\footnote{Perhaps
+  an enterprising reader can prove that for generalized chess, it is
+  NP-hard to decide whether the game has ended due to this rule?}
 
@@
 
-% XXX might be a good place to talk about the self-referential "maximal"
-% and "minimal" interpretations of this rule?
-% Need to have a  \label{sec:ambiguity} somewhere.
+\subsubsection{Ambiguity} \label{sec:ambiguity}
+
+Moreover, this rule contains some ambiguity. The phrase ``any series
+of legal moves'' is usually taken to mean something like, ``the
+players alternate legal moves and follow most of the normal rules of
+chess.'' In my opinion it is hard to justify an interpretation like
+this.
+
+First, the rules specifically define ``legal move'' in Article~3, and
+3.10.1 says ``A move is {\em legal} when all the relevant requirements
+of Articles 3.1 -- 3.9 have been fulfilled.'' These requirements
+describe the movement of each piece as you are familiar (e.g. 3.3
+``The rook may move to any square along the file or the rank on which
+it stands.''). They also disallow capturing one's own pieces, or
+moving when in check. However, they allow as legal some moves that
+would otherwise be prohibited, like capturing the opponent's king
+(this is excluded by 1.4.1, outside the definition of ``legal'').
+Capturing the opponent's king is generally not useful for
+demonstrating that checkmate is possible
+(Figure~\ref{fig:kingcapture}), so this is mostly a curiosity.
+
+\begin{figure}
+  \begin{center}
+    \begin{tabular}{cc}
+      \chessboard[smallboard,marginleft=false,marginright=false,setfen=BBKBRRnk/PPPPPPnn/3P1BNR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] &
+      \chessboard[smallboard,marginright=false,setfen=BBKBRR1B/PPPPnP1n/3P2NR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] \\
+      (a) & (b) \\
+  \end{tabular}
+  \end{center}
+
+  \caption{In this contrived and impossible position (a), white has many
+    easy paths to mate.  All of black's pieces are pinned.
+    There is also a ``series of legal moves''
+    where black mates white: \bishop xg7++  \bishop xh8?? \Knight xe7++.
+    Capturing the king (by moving twice in a row) is a ``legal move''
+    (despite not being allowed by other rules), and doing so unpins
+    black's knight to deliver a smothered mate (b). Of course, abuse of
+    this dubious technical possibility doesn't change the status of
+    the position, since we already have a mating sequence by white.
+  } \label{fig:kingcapture}
+\end{figure}
+
+Second, what is a ``series'' of legal moves? It seems completely
+consistent to allow the white player to make several legal moves in
+a row, for example. The rules about alternating moves are
+again outside the definition of ``legal move'' and ``series'' is
+never defined.
+
+We could instead interpret ``any series of legal moves'' as ``taking
+the entirety of the rules of chess, any continuation of the game that
+ends in checkmate for either player.'' I like this better, although it
+creates its own subtle issues. For example, should the position be
+considered dead if there is a checkmating sequence, but it requires
+entering a fivefold repetition or exhaustion of the 75-move rule? If
+so, this would end the game prematurely, and so it has implications
+for the longest possible game (Section~\ref{sec:ending}). Even more
+esoterically, this interpretation causes the rule to be
+self-referential: A sequence must also be allowed by the rule being
+defined. A normal person would take the ``least fixed point'' (in the
+Kripke sense~\cite{kripke1975outline}) of this self-referential
+definition (fewest positions are drawn). But it is also consistent to
+interpret it {\em maximally}---in which case the longest chess game is
+zero moves!
 
@@
 between them. If we execute the maximum number of critical moves and
-make 149 half-moves (just shy of triggering the compulsory version of
+make 149 moves (just shy of triggering the compulsory version of
 the 75-move rule) between them, then this will be fairly easily
@@
 to the exact same position. So, the strategy for generating the
-longest game can mostly be broken into two phases: Make a game with a
+longest game can mostly be broken into two tasks: Make a game with a
 maximum number of critical moves (it can also contain other moves) and
 then pad that game out to maximum-length excursions in between its
-critical moves.
+critical moves.\footnote{The first phase was constructed by hand.
+  Software for inserting excursions and checking the result is
+  available at:
+  \url{sf.net/p/tom7misc/svn/HEAD/tree/trunk/chess/longest.cc} }
 
-\subsection{Maximal critical moves}
+\subsection{Maximal critical moves} \label{sec:maxcritical}
 
@@
 $16\times 6$ critical moves. There are 14 capturable pieces, plus the
-16 pawns (they can be captured after promoting) so this nets an
+16 pawns (they can be captured after promoting); capturing them nets an
 additional $16+14$ critical moves for a total of $126$. Each critical
@@
 last piece, yielding a draw due to the remaining kings being insufficient
-material (``dead position''). This is the upper bound.
+material (``dead position''). This is our starting upper bound.
 
@@
 
-Each of these captures is {\em both} a pawn move {\em and} a capturing
-move, however. This means that we lose $4+4$ critical moves off our
+However, each of these captures is {\em both} a pawn move {\em and} a capturing
+move. This means that we lose $4+4$ critical moves off our
 total budget. $150 \times (126 - 8) = 17700$ is the new upper bound.
@@
 pointless moves, we'll only be able to move the knights and rooks, and
-doing this $75$ times must leave e.g. one of the knights on an
+doing this $75$ times must leave e.g.~one of the knights on an
 opposite-colored square.\footnote{Knight moves always change the color
@@
 Figure~\ref{fig:doubled}. This would require white to have a phase of
-critical moves to free moves to capture, then black again to finish
+critical moves to free pieces to capture, then black again to finish
 doubling pawns, then white again to promote its freed pawns. Each switch
-costs move off the theoretical max. We can be more efficient with an
+costs one move off the na\"ive max. We can be more efficient with an
 asymmetric approach.
@@
 \begin{center}
-\chessboard[setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11]
+  \chessboard[setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11,
+    pgfstyle=straightmove,markmoves={a6-b7,c3-b4,d6-c7,e6-f7,f3-g4,h6-g7},
+  ]
 \end{center}
@@
 free each of the remaining 6 pawns with a single capture. Between this
-and black's two pawn captures, this is the minimum of 8 pawn moves
+and black's two pawn captures, this is the optimal 8 pawn moves
 that are also captures. Since black has plenty of freed pieces, white
@@
 
-(this is three switches)
+As we switch which color is making critical moves a total of three
+times, we must come shy of the na\"ive maximum of 17,700 moves by
+three. This gives us an upper bound of 17,697 for this approach,
+which we will be able to achieve.
 
-\paragraph{Ending condition.} The way this game ends is subtle for
-several reasons. First, note that we left the white queen on the board
-and used it for mate. It is required that white be the one mating for
-parity reasons, similar to the reason black must make the first critical
-move. Since white makes critical moves in the last phase, black leads
-on non-critical moves; at the moment white makes the checkmating move,
-black has made 75 non-critical moves, and white 74. White's 75th move
-mates.
+\paragraph{Ending condition.} \label{sec:ending} The way this game ends
+is subtle for several reasons. First, note that we left the white
+queen on the board and used it for mate. It is required that white be
+the one mating for parity reasons, similar to the reason black must
+make the first critical move. Since white makes critical moves in the
+last phase, black leads on non-critical moves; at the moment white
+makes the checkmating move, black has made 75 non-critical moves, and
+white 74. White's 75th move mates.
 
@@
 (Section~\ref{sec:ambiguity}) that the game is over in any case: There
-is no legal sequence of moves that leads to mate. Either the king
+is no series of legal moves that leads to mate. Either the king
 captures the queen (and then clearly no mate is possible with just the
@@
 
-TODO
+The game described in Section~\ref{sec:maxcritical} has 118 critical
+moves in 289 total moves. There is some inefficiency between the
+critical moves, but this doesn't matter since we are trying to generate
+a long game anyway. In fact, the next step will be to add {\em as much
+  inefficiency as possible} in between the critical moves.
 
-\section{Future work}
+% Diagram here or earlier would probably help!
 
-Some puzzles?
+Each ``critical section,'' which is the series of moves ending in a
+critical move, can be treated independently. If black ends the section
+with a critical move, then we want $75+74$ non-critical moves to be
+played, and then black's critical move. For white, of course, $74+75$.
+There are many ways we could try to make these $149$ moves; we don't
+even have to use the moves that are already there as long as we end
+up in the right position to make the critical move. But a simple
+approach suffices.
 
-\section{A longest game}
+For each critical section, we loop over all of the positions
+encountered, and attempt inserting excursions that return us to the
+same position but waste moves. There are two types of excursions we
+try: Even excursions (each player makes two moves) and odd (each
+player makes three moves).
+
+\paragraph{Even excursions.} This four move sequence moves two pieces
+of $X_1$ and $X_2$ of opposite colors. $X_1$ moves from $s_1$ to $d_1$
+then $X_2$ from $s_2$ to $d_2$, then $X_1$ moves from $d_1$ back to
+$s_1$, and $X_2$ from $d_2$ back to $s_2$. Easy. Any piece can perform
+this maneuver other than pawns (which would be critical moves anyway)
+as long as there are legal squares (considering check, etc.). All of
+$s_1$, $d_1$, $s_2$, $d_2$ must be distinct. No shorter excursions are
+possible.
+
+\paragraph{Odd excursions.} This is the straightforward extension to
+three squares ($s_i \rightarrow m_i \rightarrow d_i \rightarrow s_i$),
+for a six-move sequence. The squares for each piece must be distinct
+but it is possible for e.g.~$m_2$ to equal $s_1$. Knights cannot
+perform this trick; each move changes the color of the square the
+knight sits on, which causes a contradiction with a cycle of length
+three. All other pieces can do it with sufficient room. The king, for
+example, can move horizontally, then diagonally, then vertically back
+to its starting square.
+
+Note that odd excursions are not possible early in the game (prior to
+white moving any pawns), because even when the knights are free, the
+rooks only have two squares (and thus the same color parity argument
+applies as knights). Some opportunity can be created by having a black
+knight capture one of white's bishops. Fortunately, we do not need any
+odd excursions at this point in the game.
+
+We find excursions by just looping over possible moves that satisfy
+the criteria, prioritizing odd excursions if the target (divided by
+two) is odd. In order to avoid triggering the fivefold repetition
+rule, we also keep track of all of the positions encountered, and
+never enter a position more than two times. (Here we avoid even
+threefold repetition.) It is not necessary to look beyond the critical
+section, because critical moves make it impossible to return to a
+prior position.
+
+This process is not at all guaranteed to work; it may fail to fill the
+critical section. Indeed, as discussed in
+Section~\ref{sec:maxcritical}, we must have at least one move of slack
+whenever we switch from a critical move by one player to the other.
+In practice this approach succeeds readily, and manages to waste 149
+moves in each critical section of the input game, save for the three
+times that parity requires one move of slack. The full game is uselessly
+included in a very tiny font in Section~\ref{sec:longest}.
+
+\section{Reader's guide}
+
+The paper demonstrates a game with three ``switches'' of which side is
+making critical moves; each costs a move against the na\"ive maximum
+due to parity. We clearly need at least one switch (both sides must
+make critical moves), but is it possible to do it with only two? If
+not, can this lower bound be proved?
+
+\smallskip
+The game given is believed to be maximal, as measured in the number of
+moves. But, other metrics exist. For example, the letter g is slightly
+wider than f, so moving \Queen g3 is {\em typographically} longer than
+\Bishop f1. PGN format itself can be stretched by making moves that
+need to be disambiguated (\bishop ff7 means ``move the \bishop\ on the
+f file to f7'' (not ``the giant sword from Final Fantasy 7'', as many
+believe)) or checking the opponent's king for a bonus +. Some moves
+are longer in terms of distance traveled; moving the queen or bishop
+between opposite corners is $\sqrt{2} \times 7$ squares! What is the
+longest game according to these or other metrics?
+
+% XXX
+% Are there any practical implications of the
+
+\smallskip
+In {\it Chess}, it is impossible to capture your own pieces. How does
+this limitation apply to your own life?
+
+\smallskip
+In many games of {\it Chess}, the black and white pieces are found to
+disagree. What does this say about society?
+
+\smallskip
+How do you feel about the ending of the game? Is it disappointing that
+the result is not symmetric (i.e.~a draw), or elegant that it
+demonstrates its own avoidance of the ``dead position'' rule? Was it
+what you expected?
+
+\smallskip
+{\it Is this the longest Chess game?} is the author's sixth paper
+about chess for SIGBOVIK, but it is widely believed that nobody wants
+to read this kind of thing. What is wrong with him?
+
+\section{A longest game} \label{sec:longest}
+
+There are jillions of possible games that satisfy the description
+above and reach 17,697 moves; here is one of them.\footnote{
+  It can also be downloaded at \url{tom7.org/chess/longest.pgn}.
+  Many chess programs fail to load the whole game, but this is
+  because they decided not to implement the full glory of chess.}
+%
+Critical moves are marked with {\large \bf [bold]}. Note that in the
+standard PGN format for listing games, move numbers are ``full moves''
+consisting of a move by white and then black, whereas we use ``move''
+in this paper to mean ``half-move''; an individual move by one of the
+players. Thus the game ends during full move 8,849, one half of
+17,697.
+
+\medskip
 
@@
 \tolerance=2
-
-(a long game goes here)
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad
-minim veniam, quis nostrud exercitation ullamco laboris nisi ut
-aliquip ex ea commodo consequat. Duis aute irure dolor in
-reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla
-pariatur. Excepteur sint occaecat cupidatat non proident, sunt in
-culpa qui officia deserunt mollit anim id est laborum.
 

### 2020-03-15 trunk/chess/chess.cc

@@ -1386,6 +1386,14 @@ string Position::ShortMoveString(Move m) {
   }
 }
 
+string Position::PGNMoveSuffix(Move m) const {
+  Position p = *this;
+  p.ApplyMove(m);
+  if (p.IsMated()) return "++";
+  else if (p.IsInCheck()) return "+";
+  else return "";
+}
+
 string Position::DebugMoveString(Move m) {
   string s = "xx->xx";
   s[0] = m.src_col + 'a';

### 2020-03-15 chess/papers/longest.tex
Submitted version

@@
 \usepackage{textcomp}
+
+\usepackage{adjustbox}
+
+% XXX search for common typo "movies"
+
+% lets me explicitly set a. or 1. etc. as enum label
+\usepackage{enumitem}
+
+\usepackage[pdftex,
+            pdfauthor={Dr. Tom Murphy VII Ph.D.},
+            pdftitle={Is this the longest chess game?},
+            pdfkeywords={SIGBOVIK, chess}]{hyperref}
 
@@
 
+% Define black versions of pieces for inline use. Gross, but it works.
+\newcommand{\Pawn}[1][1.3ex]{%
+\adjustbox{Trim=4.3pt 2.6pt 4.3pt 0pt,width=#1,margin=0.2ex 0ex 0.2ex 0ex}{\BlackPawnOnWhite}%
+}%
+\newcommand{\Rook}[1][1.58ex]{%
+\adjustbox{Trim=3.2pt 2.2pt 3.2pt 0pt,width=#1,raise=0ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackRookOnWhite}%
+}%
+\newcommand{\Knight}[1][1.85ex]{%
+\adjustbox{Trim=2.3pt 2.35pt 2.5pt 0pt,width=#1,raise=-0.03ex,margin=0.14ex 0ex 0.14ex 0ex}{\BlackKnightOnWhite}%
+}%
+\newcommand{\Bishop}[1][1.79ex]{%
+\adjustbox{Trim=2.3pt 2pt 2.3pt 0pt,width=#1,raise=-0.12ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackBishopOnWhite}%
+}%
+\newcommand{\Queen}[1][2.05ex]{%
+\adjustbox{Trim=1.2pt 2.2pt 1.2pt 0pt,width=#1,raise=-0.08ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackQueenOnWhite}%
+}%
+\newcommand{\King}[1][1.95ex]{%
+\adjustbox{Trim=2pt 2pt 2pt 0pt,width=#1,raise=-0.06ex,margin=0.13ex 0ex 0.13ex 0ex}{\BlackKingOnWhite}%
+}%
+
 \begin{document}
@@
 Verlag-Verlag volume no.~0x40-2A.
-\textcurrency 17696
+\textcurrency 17,697
 }
@@
 
-In my experience, most chess games end in a few moves. Players who
-seek longevity typically do so by playing many games of chess, rather
-than a protracted
+In my experience, most chess games end in a few moves. If you want to
+play a lot of chess moves, you just play a lot of chess games. Still,
+there are games that seem to go on foreverrrrrrrrrr. Perhaps the
+players are trying to {\em lull each other into a false sense of
+security} while {\em waiting for the moment to strike}, or perhaps they
+are stalling in a game of {\em Chess to the Death}.
 
@@
 a minor chess scandal that erupted in 2019 over an obscure corner case
-in the rules.) Several of these ``deep cuts'' have to do with game-ending
-conditions that were introduced to keep the game from going on forever.
+in the rules.) Several of these ``deep cuts'' have to do with
+game-ending conditions that were introduced to avoid interminable
+games.
 
-Many chess moves are reversible (e.g. moving the knight forward and
+Many chess moves are reversible (e.g.~moving the knight forward and
 back to its starting position~\cite{survival}), so informal games of
@@
 % probability 1).
-So, chess is formally a finite game. This good for computer
+So, chess is formally a finite game. This is good for computer
 scientists, since it means that chess has a trivial O(1) optimal
@@
 What is the longest chess game? In this paper I show how to compute
-such a game, and then gratuitously present all of its
-% XXX CHECK
-17,696 
-%
-half-moves. Even if you are a chess expert (``chexpert''), I bet you
+such a game, and then gratuitously present all of its 17,697 
+moves. Even if you are a chess expert (``chexpert''), I bet you
 will be surprised at some of the corner cases in the rules that are
@@
 
-Speaking of rules, let's first detail the two main rules that limit the
-length of the game.
+Speaking of rules, let's first detail the three main rules that limit
+the length of the game. These rules cause the game to end in a draw
+(tie) when certain conditions are met.
 
-\begin{figure}
+\begin{figure}[b!]
   \begin{center}
-    \chessboard[setfen=2nr1r1k/pb1n1pqp/1p2p1p1/1N1pN3/5P2/1P4P1/P2PPQ1P/2R1K1RB,showmover=false]
+    \chessboard[
+      % fillarea=a3-h1,printarea=a3-h1,
+      marginbottom=false,margintop=false,
+      setfen=2nr1r1k/pb1n1pqp/1p2p1p1/1N1pN3/5P2/1P4P1/P2PPQ1P/2R1K1RB,
+      pgfstyle=curvemove,markmoves={e1-g1},
+      pgfstyle=straightmove,markmoves={g1-f1},
+      showmover=false]
   \end{center}
-  \caption{ ({\it Nepomniachtchi -- So, 2019.}) White to move during
-    a speed Chess960 (aka ``Fischer Random'') tournament. In this
+  \caption{ ({\it Nepomniachtchi -- So, 2019.}) White to move during a
+    speed Chess960 (aka ``Fischer Random'') tournament. In this
     variant, the pieces start in different positions, but castling
@@
     destination square for the king to already be occupied by the
-    rook. Attempting to castle in the position depicted,
-    (Figure~\ref{fig:castle960}), grandmaster Ian Nepomniachtchi
-    first touched the rook to move it out of the way. However,
-    piece-touching rules require that when castling, the player
-    first moves the king; but how? The rook is occupying g1! One
-    commenter suggested tossing the king into the air, then sliding
-    the rook to f1 while the king is airborne, and then watching the
-    king land dead center on its target. The arbiter required Nepo
-    to make a rook move instead, but this was later appealed, and
-    the game replayed. } \label{fig:castle960}
+    rook. Attempting to castle in the position depicted, grandmaster
+    Ian Nepomniachtchi first touched the rook to move it out of the
+    way. However, piece-touching rules require that when castling, the
+    player first moves the king (and ``Each move must be played with
+    one hand only''); but how? The rook is occupying g1! One commenter
+    suggested tossing the king into the air, then sliding the rook to
+    f1 while the king is airborne, and then watching the king land
+    dead center on its target. The arbiter required Nepo to make a
+    rook move instead, but this was later appealed, and the game
+    replayed. } \label{fig:castle960}
 \end{figure}
 
-(Players are collaborating, perhaps playing chess to the death)
+% TODO: (Players are collaborating, perhaps playing chess to the death)
 
-\subsection{The seventy-five move rule}
+\subsection{The seventy-five move rule} \label{sec:75}
 
-The ``fifty-move rule''
+The ``fifty-move rule,''~\cite{wikipedia50move} as it is usually
+known, requires that an irreversible move is played at a minimum pace.
+For the sake of this rule, irreversible moves are considered captures
+and pawn moves,%
+\footnote{ Other irreversible moves include: Castling, moving a piece
+  so as to lose castling rights, and declining to capture {\it en
+    passant} (such capture must be made immediately, so if the option
+  is not taken, it cannot be regained). The fifty-move rule could be
+  soundly expanded to include these, but, that's just like, not the
+  rule, man.}
+%
+including promotion and {\it en passant} (which is also a capturing
+move) but not the movement of a previously promoted piece.
+Specifically~\cite{fiderules}:
 
-Although usually known by the name ``fifty-move rule'' ()
+\begin{quote}
+{\bf 9.3}.\quad The game is drawn, upon a correct claim by the player having
+the move, if:
+  \begin{enumerate}[label=\alph*.]
+    \item he writes his move on his scoresheet and declares to the
+      arbiter his intention to make this move, which shall result in
+      the last 50 moves having been made by each player without the
+      movement of any pawn and without any capture, or
+    \item the last 50 consecutive moves have been made by each player
+      without the movement of any pawn and without any capture.
+  \end{enumerate}
+\end{quote}
 
-\subsection{Fivefold repetition}
+Note that what is provided here is the option for a player to {\em
+  claim} a draw (and the two provisions essentially allow either player
+to claim the draw at the moment of the 50\th\ move). If neither player
+is interested in a draw, either because they think their position is
+still winning, or are just trying to create the longest ever chess game,
+the game legally continues. That's why what is actually relevant for this
+paper is provision 9.6, which defines a draw:
 
-Although only
+\begin{quote}
+  {\bf 9.6}.\quad If one or both of the following occur(s) then the
+  game is drawn:
 
-% \begin{figure}[htp]
-% \input{table-two}
-% \caption{Minimal joining strings for every letter (rows) to every other letter (columns).} \label{fig:table}
-% \end{figure}
+  \ldots 9.6.2. any series of at least 75 moves have been made by each
+  player without the movement of any pawn and without any capture. If
+  the last move resulted in checkmate, that shall take precedence.
+\end{quote}
+
+Draws after 75 moves (per player, so really 150 moves) are compulsory.
+
+Interestingly (at least as interesting as anything in this dubious
+affair) it is known that some otherwise winning endgame positions
+require more than 50 moves to execute (Figure~\ref{fig:matein545}).
+The rules of chess have at various times allowed for longer timers in
+such known situations, but were later simplified to the fixed 50 (and
+75) move limit.
+
+\subsection{Fivefold repetition} \label{sec:fivefold}
+
+The 75-move rule is rarely applied in practice, but its
+counterpart, ``threefold repetition''~\cite{wikipediathreefold} is often
+the cause of draws in modern chess. This rule states that if the same
+position appears three times, the players can claim a draw:
+
+\begin{quote}
+\begin{itemize}
+  \item[9.2.2.] Positions are considered the same if and only if the
+same player has the move, pieces of the same kind and colour occupy
+the same squares and the possible moves of all the pieces of both
+players are the same. [...]
+\comment{
+  Thus positions are not the same if:
+\begin{itemize}
+\item[9.2.2.1.]	at the start of the sequence a pawn could have been captured en passant
+\item[9.2.2.2.]	a king had castling rights with a rook that has not been moved, but forfeited these after moving. The castling rights are lost only after the king or rook is moved.
+\end{itemize}
+}
+\end{itemize}
+\end{quote}
+
+Like the 75-move rule, this rule has an optional version (upon
+three repetitions) and a mandatory one in 9.6:
+
+\begin{quote}
+  [The game is a draw if \ldots] \\
+  9.6.1. the same position has appeared, as in 9.2.2, at least five times.
+\end{quote}
+
+Fascinatingly (at least as fascinating as anything in this
+questionable undertaking), this rule used to require {\em consecutive}
+repetition of moves. However, there exist infinite sequences of moves
+with no consecutive $n$-fold repetition. For example, in the starting
+position, white and black can move either of their knights out and
+back. Let {\bf 0} be \knight c3 \Knight c6 \knight b1 \Knight b8
+(queenside knights move, returning to the starting position) and {\bf
+  1} be \knight f3 \Knight f6 \knight g1 \Knight g8 (kingside). Now
+the Prouhet--Thue--Morse sequence~\cite{thue1906uber}
+0110100110010110\ldots\footnote{
+  Let $s_0$ be the string {\tt "0"}, then
+  $s_i$ is $s_{i-1} \overline{s_{i-1}}$
+  where $\overline{\tt 0} = {\tt 1}$ and $\overline{\tt 1} = {\tt 0}$.
+}
+%
+can be executed. This infinite
+sequence is cube-free (does not contain $SSS$ for any non-empty string
+$S$)~\cite{oeisa010060}, and therefore never violates the consecutive
+threefold repetition rule~\cite{euwe1929mengentheoretische}.
+
+\medskip
+Many implementations of chess ignore these rules or treat them
+incorrectly. Implementation of the seventy-five move rule simply
+requires a count of how many moves have transpired since a pawn move
+or capture, but programs typically do not force a draw after 75 moves.
+Repetition requires more work, since the program must keep track of
+each of the states reached since the last irreversible move. There are
+also some corner cases, such as ambiguity as to whether the starting
+position has ``appeared'' before the first move~\cite{survival}. The
+ubiquitous FEN notation for describing chess positions does not even
+include any information about states previously reached.
+
+\begin{figure}[htp]
+  \begin{center}
+    \chessboard[setfen=8/r6n/8/8/5k2/3K4/7N/3b3Q b - - 0 1]
+  \end{center}
+  \caption{Black to move and mate in 545 moves (!). The position was
+    found (by Zakharov and Makhnichev~\cite{tablebase7}) while
+    building an endgame tablebase of all possible 7-piece positions.
+    Of course, the game ends prematurely in a draw because of the
+    75-move rule. } \label{fig:matein545}
+\end{figure}
+
+\subsection{Dead position} \label{sec:insufficient}
+
+The informal version of this rule (``insufficient material'') states
+that if neither side has enough pieces to mate the opponent (for
+example, a king and bishop can never mate a bare king) then the game
+is drawn. Again, the formal rule is more subtle:
+
+\begin{quote}
+5.2.2. The game is drawn when a position has arisen in which neither
+player can checkmate the opponent's king with any series of legal
+moves. The game is said to end in a `dead position'. This immediately
+ends the game\ldots
+\end{quote}
+
+This clearly includes the well-known material-based cases like king
+and knight vs.~king, but it also surprisingly includes many other
+specific positions, especially those with forced captures
+(Figure~\ref{fig:insufficient}).
+
+\begin{figure}[htp]
+  \begin{center}
+    \chessboard[margintop=false,marginbottom=false,setfen=8/8/5K2/8/8/8/Q7/k7 b - - 0 1]
+  \end{center}
+  \caption{Black to move and draw in 0 (!). Most players and even
+    chexperts believe that the only legal move is Kxa2, and
+    that the game then ends in a draw with ``insufficient material.''
+    However, this game is {\em already over}. Since neither black
+    nor white can win via any series of legal moves, by rule 5.2.2
+    the game immediately ends in a `dead position'. (Although see
+    Section~\ref{sec:ambiguity} for possible ambiguity in this
+    rule.)} \label{fig:insufficient}
+\end{figure}
+
+The insufficient material rule is curious in that it requires
+nontrivial computation to implement. In order to know whether the
+position is a draw, an implementation needs to be able to decide
+whether or not a series of moves that results in checkmate exists.
+Note that this is not nearly as bad as normal game tree search because
+the two sides can collaborate to produce the mate (it is not a
+$\exists \forall \exists \forall \ldots$ but rather $\exists \exists
+\exists \exists \ldots$). Still, such ``helpmates'' can still be quite
+deep (dozens of moves) and are interesting enough to be a common
+source of chess puzzles. Proving the non-existence of a helpmate can
+be very difficult indeed (Figure~\ref{fig:nohelp}).\footnote{Perhaps
+  an enterprising reader can prove that for generalized chess, it is
+  NP-hard to decide whether the game has ended due to this rule?}
+
+\begin{figure}[htp]
+  \begin{center}
+    \chessboard[margintop=false,marginbottom=false,setfen=k1bB4/8/2p1p1p1/1pPpPpP1/1P1P1P2/8/8/K1Bb4 w - - 0 1]
+  \end{center}
+  \caption{Thinking of implementing the rules of chess? To be correct,
+    you'll need your program to be able to deduce that no helpmates are
+    possible in this position and thus the game is over. Stockfish rates
+    this as +0.4 for white, even searching to depth 92.
+    (Position is due to user {\tt supercat} on Chess StackExchange.)
+  } \label{fig:nohelp}
+\end{figure}
+
+\subsubsection{Ambiguity} \label{sec:ambiguity}
+
+Moreover, this rule contains some ambiguity. The phrase ``any series
+of legal moves'' is usually taken to mean something like, ``the
+players alternate legal moves and follow most of the normal rules of
+chess.'' In my opinion it is hard to justify an interpretation like
+this.
+
+First, the rules specifically define ``legal move'', with 3.10.1
+saying ``A move is {\em legal} when all the relevant requirements of
+Articles 3.1 -- 3.9 have been fulfilled.'' These requirements describe
+the movement of each piece as you are familiar (e.g. 3.3 ``The rook
+may move to any square along the file or the rank on which it
+stands.''). They also disallow capturing one's own pieces, or moving
+when in check. However, they allow as legal some moves that would
+otherwise be prohibited, like capturing the opponent's king (this is
+excluded by 1.4.1, outside the definition of ``legal''). Capturing the
+opponent's king is generally not useful for demonstrating that
+checkmate is possible (Figure~\ref{fig:kingcapture}), so this is
+mostly a curiosity.
+
+\begin{figure}
+  \begin{center}
+    \begin{tabular}{cc}
+      \chessboard[smallboard,marginleft=false,marginright=false,setfen=BBKBRRnk/PPPPPPnn/3P1BRR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] &
+      \chessboard[smallboard,marginright=false,setfen=BBKBRR1B/PPPPnP1n/3P2RR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] \\
+      (a) & (b) \\
+  \end{tabular}
+  \end{center}
+
+  \caption{In this contrived and impossible position (a), white has many
+    easy paths to mate.  All of black's pieces are pinned.
+    There is also a ``series of legal moves''
+    where black mates white: \bishop xg7++  \bishop xh8?? \Knight xe7++.
+    Capturing the king (by moving twice in a row) is a ``legal move''
+    (despite not being allowed by other rules), and doing so unpins
+    black's knight to deliver a smothered mate (b). Of course, abuse of
+    this dubious technical possibility doesn't change the status of
+    the position, since we already have a mating sequence by white.
+  } \label{fig:kingcapture}
+\end{figure}
+
+Second, what is a ``series'' of legal moves? It seems completely
+consistent to allow the white player to make several legal moves in
+a row, for example. The rules about alternating moves are
+again outside the definition of ``legal move'' and ``series'' is
+never defined.
+
+We could instead interpret ``any series of legal moves'' as ``taking
+the entirety of the rules of chess, any continuation of the game that
+ends in checkmate for either player.'' I like this better, although it
+creates its own subtle issues. For example, should the position be
+considered dead if there is a checkmating sequence, but it requires
+entering a fivefold repetition or exhaustion of the 75-move rule? If
+so, this would end the game prematurely, and so it has implications
+for the longest possible game (Section~\ref{sec:ending}). Even more
+esoterically, this interpretation causes the rule to be
+self-referential: A sequence must also be allowed by the rule being
+defined. A normal person would take the ``least fixed point'' (in the
+Kripke sense~\cite{kripke1975outline}) of this self-referential
+definition (fewest positions are drawn). But it is also consistent to
+interpret it {\em maximally}---in which case the longest chess game is
+zero moves!
+
+\medskip
+For completeness, note that there are other routes to a draw
+(stalemate, draw offers) which we can ignore; it is easy to avoid
+these situations when generating the longest game.
+
+\section{Generating the longest game}
+
+It is generally not hard to avoid repeating positions, so the main
+obstacle we'll face is the 75-move rule. Let's call an irreversible
+move that resets the 75-move counter a {\em critical move}; this is
+a pawn move or capturing move (or both). The structure of the game
+will be a series of critical moves (I will call these ``critical''
+moves) with a maximal sequence of pointless reversible moves in
+between them. If we execute the maximum number of critical moves and
+make 149 moves (just shy of triggering the compulsory version of
+the 75-move rule) between them, then this will be fairly easily
+seen as a maximal game.
+
+In fact, for most positions, it is easy to waste 149 moves and return
+to the exact same position. So, the strategy for generating the
+longest game can mostly be broken into two tasks: Make a game with a
+maximum number of critical moves (it can also contain other moves) and
+then pad that game out to maximum-length excursions in between its
+critical moves.\footnote{The first phase was constructed by hand.
+  Software for inserting excursions and checking the result is
+  available at:
+  \url{sf.net/p/tom7misc/svn/HEAD/tree/trunk/chess/longest.cc} }
+
+\subsection{Maximal critical moves} \label{sec:maxcritical}
+
+Critical moves are pawn moves and captures. There are 16 pawns, each
+of which has 6 squares to move into before promoting, so this is
+$16\times 6$ critical moves. There are 14 capturable pieces, plus the
+16 pawns (they can be captured after promoting); capturing them nets an
+additional $16+14$ critical moves for a total of $126$. Each critical
+move can be made after a maximum of $75+74$ reversible moves, giving
+$(75+74+1) \times 126 = 18900$ moves. The final move would capture the
+last piece, yielding a draw due to the remaining kings being insufficient
+material (``dead position''). This is our starting upper bound.
+
+We will not quite be able to use the entire critical move budget. If
+pawns only move forward in single steps, they will eventually get
+blocked by the opposing pawn on the same file. Pawns can move
+diagonally off their starting file only by capturing. We have plenty
+of capturing to do anyway, so this in no problem. With four captures
+per side, the pawns can be doubled, with a clear route to promotion,
+like in Figure~\ref{fig:doubled}.
+
+\begin{figure}[h]
+\begin{center}
+  \chessboard[setfen=2bqkb2/1p1p1p1p/1p1p1p1p/8/8/P1P1P1P1/P1P1P1P1/2BQKB2 w - - 0 1]
+  \caption{One way to clear the promotion routes for all pawns with
+    eight captures. } \label{fig:doubled}
+\end{center}
+\end{figure}
+
+However, each of these captures is {\em both} a pawn move {\em and} a capturing
+move. This means that we lose $4+4$ critical moves off our
+total budget. $150 \times (126 - 8) = 17700$ is the new upper bound.
+
+\paragraph{Parity.} If a critical move (such as a pawn move) is made by
+white, then the first non-critical move is made by black. The
+players alternate these pointless moves until black has made $75$ and
+white $74$. Now it is white's move, and white must make a critical move
+or the game ends due to the 75-move rule. If instead we wanted black
+to make the next critical move, this would happen after black has made
+$74$ and white $74$ non-critical moves. Any time this happens we lose one
+move against the upper bound. So, we want to minimize the number of
+times we switch which player is making the critical moves. Obviously
+we must switch at least once, because both black and white must
+make critical moves. This reduces the upper bound to $17699$.
+
+\paragraph{Starting condition.} The first critical move should be made
+by black. The starting position (with white to move) is analogous to
+the situation just described, as if black has just made a critical move,
+and it's white's turn. White will play $75$ reversible moves, black plays
+$74$, and the 150\th\ move is black making the first critical move.
+
+Note that white is quite constrained during this beginning phase, as
+pawn moves are critical moves and must be avoided. Only the white
+knights can escape the back rank. When we try to insert $149$
+pointless moves, we'll only be able to move the knights and rooks, and
+doing this $75$ times must leave e.g.~one of the knights on an
+opposite-colored square.\footnote{Knight moves always change the color
+  of the knight's square, and same for rook moves constrained to the
+  a1/b1 or g1/h1 squares.} So we have to be a little careful about the
+position in which we make black's first critical move.
+
+Since white can only free their two knights, these are the only pieces
+that can be captured by black pawns. So it will not be possible for
+black to double four sets of their pawns as in
+Figure~\ref{fig:doubled}. This would require white to have a phase of
+critical moves to free pieces to capture, then black again to finish
+doubling pawns, then white again to promote its freed pawns. Each switch
+costs one move off the na\"ive max. We can be more efficient with an
+asymmetric approach.
+
+Black's first phase of critical moves instead results in this:
+
+\comment{
+1. Nf3 b6 2. Nc3 g6 3. Ne5 g5 4. Nd5 b5
+5. Nc4 bxc4 6. Nf4 gxf4 7. Rg1 c6 8. Rh1 c5
+9. Rg1 f6 10. Rh1 f5 
+}
+\begin{center}
+  \chessboard[
+    margintop=false,marginbottom=false,
+    setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11,
+    pgfstyle=straightmove,markmoves={a6-b7,c3-b4,d6-c7,e6-f7,f3-g4,h6-g7},
+  ]
+\end{center}
+
+\smallskip
+\noindent The white b and g pawns have a clear route to promotion. White can
+free each of the remaining 6 pawns with a single capture. Between this
+and black's two pawn captures, this is the optimal 8 pawn moves
+that are also captures. Since black has plenty of freed pieces, white
+can promote all of their pawns during their own phase of critical
+moves, resulting in:
+
+\comment{
+11. b3 Nc6 12. b4 Nd4
+13. b5 Nb3 14. cxb3 Bb7 15. a3 Rc8 16. a4 Rc6
+17. a5 Rd6 18. a6 Re6 19. axb7 Rd6 20. b8=N Re6
+21. b6 Nh6 22. b7 Ng4 23. Nc6 Bh6 24. b8=Q Kf8
+25. b4 Ke8 26. b5 Kf8 27. Qe5 Ke8 28. b6 Kf8
+29. b7 Ke8 30. b8=N Bg5 31. h3 Bh4 32. hxg4 Bg3
+33. fxg3 Qc8 34. g5 Qa6 35. Nb4 Kd8 36. Qe3 Kc7
+37. g6 Kb6 38. g7 Rf6 39. g8=Q Re6 40. g4 Rf6
+41. g5 Re6 42. g6 Rf6 43. g7 Re6 44. Qf8 Rf6
+45. g8=N Rd6 46. Nf6 Qb5 47. Nh5 Rg8 48. g3 Rgg6
+49. g4 Rge6 50. g5 Rf6 51. g6 Rc6 52. g7 Qa6
+53. g8=N Qa5 54. d3 Qa6 55. d4 Qa5 56. d5 Qa6
+57. dxc6 Qa5 58. c7 Qa6 59. c8=R Qa4 60. Qh3 Qa6
+61. e3 Qa5 62. e4 Qa6 63. e5 Qa5 64. exf6 Qa6
+65. Qg7 Qa5 66. f7 Qa6 67. Re8 Qa5 68. f8=R Qa6
+69. Qdg4 Qa5 70. Nc6 Qa6 71. Ne5 Qa5 72. Nf3 Qa6
+73. Nd3 Qa5+ 74. Ke2 Qa6 75. Nf2 Qa5 76. Qh8 Qa3
+77. Nd1 Qa2+ 78. Nd2 Qa3 79. Kf2 Qa2 80. Kg1 Qa3
+81. Rxa3
+}
+\begin{center}
+  \chessboard[
+    margintop=false,marginbottom=false,
+    setfen=4RRNQ/p2pp2p/1k6/2p2p1N/2p2pQ1/R6Q/3N4/2BN1BKR b - - 0 81]
+\end{center}
+
+\smallskip
+\noindent Now black can promote all of its pawns and capture white's pieces.
+We actually leave the white queen; this turns out to be essential:
+
+\comment{
+\ldots f3 82. Kh2 f2 83. Bd3 f1=R 84. Qf4 Rxf4
+85. Ng7 h6 86. Be4 Rxe4 87. Ne6 f4 88. Nf6 f3
+89. Ng8 f2 90. Nf6 f1=B 91. Rd3 Bxd3 92. Bb2 a6
+93. Bc3 a5 94. Bb2 a4 95. Bc3 a3 96. Bb2 a2
+97. Bc3 a1=R 98. Ba5+ Rxa5 99. Nf4 c3 100. Ne2 c2
+101. Nf4 c1=N 102. Ne2 c4 103. Ng4 Nxe2 104. Nge3 c3
+105. Nf5 c2 106. Nfe3 c1=R 107. Re1 h5 108. Qf3 h4
+109. Kg2 h3+ 110. Kf2 h2 111. Qfh5 h1=R 112. Ng4 Rxg4
+113. Nc4+ Bxc4 114. Nb2 d6 115. Rg1 d5 116. Rg2 d4
+117. Kf3 d3 118. Kf2 d2 119. Qh2 d1=R 120. Kf3 Nc3
+121. Kf2 Rg8 122. Kf3 e6 123. Kf4 e5+ 124. Kf5 e4+
+125. Kf6 e3 126. Rg3 e2 127. Rg2 e1=Q 128. Nd3 Rxd3
+129. Rg1 Rgxg1 130. Q8h3 Rxh3 131. Ra8 Rxa8 132. Rb8+ Rxb8
+}
+
+\begin{center}
+  \chessboard[
+    margintop=false,marginbottom=false,
+    setfen=1r6/8/1k3K2/8/2b5/2n4r/7Q/2r1q1rr w - - 0 133]
+\end{center}
+
+\smallskip
+\noindent Finally, white captures all of black's pieces, and mates the black king:
+
+\comment{
+133. Qxb8+ Ka5 134. Qg3 Bf7 135. Qxe1 Rg5 136. Qxc1 Ka6
+137. Qxc3 Kb6 138. Qxh3 Kc5 139. Kxf7 Kc6 140. Qxh1+ Kd7
+141. Kf6 Ke8 142. Qg2 Rg8 143. Ke6 Rg7 144. Qxg7 Kd8
+145. Qd7\#
+}
+
+\begin{center}
+  \chessboard[
+    margintop=false,marginbottom=false,
+    setfen=3k4/3Q4/4K3/8/8/8/8/8 b - - 2 145]
+\end{center}
+
+\smallskip
+\noindent Since we switch which color is making critical moves a total of three
+times, we must come shy of the na\"ive maximum of 17,700 moves by
+three. This gives us an upper bound of 17,697 for this approach,
+which we will be able to achieve.
+
+\paragraph{Ending condition.} \label{sec:ending} The way this game ends
+is subtle for several reasons. First, note that we left the white
+queen on the board and used it for mate. It is required that white be
+the one mating for parity reasons, similar to the reason black must
+make the first critical move. Since white makes critical moves in the
+last phase, black leads on non-critical moves; at the moment white
+makes the checkmating move, black has made 75 non-critical moves, and
+white 74. White's 75\th\ move mates.
+
+But doesn't this trigger the 75-move rule? No, this rule (9.6.2;
+Section~\ref{sec:75}) has a special exception for checkmate: ``If the
+last move resulted in checkmate, that shall take precedence.''
+Essentially, we can treat checkmate as a type of critical move.
+
+Why not capture the white queen too? This would be a critical move,
+but once two kings remain, the game ends immediately in a draw. So
+there is no length advantage here over checkmate. In fact, attempting
+to capture may {\em shorten} the game: Consider the position in
+Figure~\ref{fig:insufficient} where black is {\em forced} to capture
+the queen. This game is over prior to the capture, so this is shorter
+than the checkmating sequence! Even if the king had an escape square,
+it is arguable (Section~\ref{sec:ambiguity}) that the game is over in
+any case: There is no ``series of legal moves'' that leads to mate.
+Either the king captures the queen (and then clearly no mate is
+possible with just the two kings) or the king escapes, but in doing so
+plays the 75\th\ non-critical move, and triggers a draw by the 75-move
+rule.
+
+The most foolproof way to ensure that mate is always possible is for the
+game itself to end in mate. This sidesteps any ambiguity about the way
+the 75-move condition should be interpreted, as well.
+
+\subsection{Inserting excursions} \label{sec:excursions}
+
+The game described in Section~\ref{sec:maxcritical} has 118 critical
+moves in 289 total moves. There is some inefficiency between the
+critical moves, but this doesn't matter since we are trying to generate
+a long game anyway. In fact, the next step will be to add {\em as much
+  inefficiency as possible} in between the critical moves.
+
+% Diagram here or earlier would probably help!
+
+Each ``critical section,'' which is the series of moves ending in a
+critical move, can be treated independently. If black ends the section
+with a critical move, then we want $75+74$ non-critical moves to be
+played, and then black's critical move. For white, of course, $74+75$.
+There are many ways we could try to make these $149$ moves; we don't
+even have to use the moves that are already there as long as we end
+up in the right position to make the critical move. But a simple
+approach suffices.
+
+For each critical section, we loop over all of the positions
+encountered, and attempt inserting excursions that return us to the
+same position but waste moves. There are two types of excursions we
+try: Even excursions (each player makes two moves) and odd (each
+player makes three moves).
+
+\paragraph{Even excursions.} This four move sequence moves two pieces
+of $X_1$ and $X_2$ of opposite colors. $X_1$ moves from $s_1$ to $d_1$
+then $X_2$ from $s_2$ to $d_2$, then $X_1$ moves from $d_1$ back to
+$s_1$, and $X_2$ from $d_2$ back to $s_2$. Easy. Any piece can perform
+this maneuver other than pawns (which would be critical moves anyway)
+as long as there are legal squares (considering check, etc.). All of
+$s_1$, $d_1$, $s_2$, $d_2$ must be distinct. No shorter excursions are
+possible.
+
+\paragraph{Odd excursions.} This is the straightforward extension to
+three squares ($s_i \rightarrow m_i \rightarrow d_i \rightarrow s_i$),
+for a six-move sequence. The squares for each piece must be distinct
+but it is possible for e.g.~$m_2$ to equal $s_1$. Knights cannot
+perform this trick; each move changes the color of the square the
+knight sits on, which causes a contradiction with a cycle of length
+three. All other pieces can do it with sufficient room. The king, for
+example, can move horizontally, then diagonally, then vertically back
+to its starting square.
+
+Note that odd excursions are not possible early in the game (prior to
+white moving any pawns), because even when the knights are free, the
+rooks only have two squares (and thus the same color parity argument
+applies as knights). Some opportunity can be created by having a black
+knight capture one of white's bishops. Fortunately, we do not need any
+odd excursions at this point in the game.
+
+We find excursions by just looping over possible moves that satisfy
+the criteria, prioritizing odd excursions if the target (divided by
+two) is odd. In order to avoid triggering the fivefold repetition
+rule, we also keep track of all of the positions encountered, and
+never enter a position more than two times. (Here we avoid even
+threefold repetition.) It is not necessary to look beyond the critical
+section, because critical moves make it impossible to return to a
+prior position.
+
+This process is not at all guaranteed to work; it may fail to fill the
+critical section. Indeed, as discussed in
+Section~\ref{sec:maxcritical}, we must have at least one move of slack
+whenever we switch from a critical move by one player to the other.
+In practice this approach succeeds readily, and manages to waste 149
+moves in each critical section of the input game, save for the three
+times that parity requires one move of slack. The full game is uselessly
+included in a very tiny font in Section~\ref{sec:longest}.
+
+\section{Reader's guide}
+
+The paper demonstrates a game with three ``switches'' of which side is
+making critical moves; each costs a move against the na\"ive maximum
+due to parity. We clearly need at least one switch (both sides must
+make critical moves), but is it possible to do it with only two? If
+not, can this lower bound be proved?
+
+\smallskip
+The game given is believed to be maximal, as measured in the number of
+moves. But, other metrics exist. For example, the letter g is slightly
+wider than f, so moving \Queen g3 is {\em typographically} longer than
+\Bishop f1. PGN format itself can be stretched by making moves that
+need to be disambiguated (\bishop ff7 means ``move the \bishop\ on the
+f file to f7'' (not ``the giant sword from Final Fantasy 7'', as many
+believe)) or checking the opponent's king for a bonus +. Some moves
+are longer in terms of distance traveled; moving the queen or bishop
+between opposite corners is $\sqrt{2} \times 7$ squares! What is the
+longest game according to these or other metrics?
+
+\smallskip
+In {\it Chess}, it is impossible to capture your own pieces. How does
+this limitation apply to your own life?
+
+\smallskip
+In many games of {\it Chess}, the black and white pieces are found to
+disagree. What does this say about society?
+
+\smallskip
+How do you feel about the ending of the game? Is it disappointing that
+the result is not symmetric (i.e.~a draw), or elegant that it
+demonstrates its own avoidance of the ``dead position'' rule? Was it
+what you expected?
+
+\smallskip
+The character called \queen\ often plays an important role in the story.
+How would you describe her personality? How does she develop over the
+course of the game?
+
+\smallskip
+{\it Is this the longest Chess game?} is the author's sixth paper
+about chess for SIGBOVIK, but it is widely believed that nobody wants
+to read this kind of thing. What is wrong with him?
+
+\section{A longest game} \label{sec:longest}
+
+There are jillions of possible games that satisfy the description
+above and reach 17,697 moves; here is one of them.\footnote{
+  It can also be downloaded at \url{tom7.org/chess/longest.pgn}.
+  Many chess programs fail to load the whole game, but this is
+  because they decided not to implement the full glory of chess.}
+%
+Critical moves are marked with {\large \bf [bold]}. Note that in the
+standard PGN format for listing games, move numbers are ``full moves''
+consisting of a move by white and then black, whereas we use ``move''
+in this paper to mean ``half-move''; an individual move by one of the
+players. Thus the game ends during full move 8,849, one half of
+17,697.
+
+\medskip
 
@@
 \noindent \input{slow}
-0--0
+\quad 1-0
 

### 2020-03-23 trunk/chess/papers/longest.tex
This year, not 5 years ago

@@
 \usepackage{adjustbox}
-
-% XXX search for common typo "movies"
 
@@
 Copyright \copyright\ 2020 the Regents of the Wikiplia
-Foundation. Appears in SIGBOVIK 2015 with the {\tt ShortMoveString} of the
+Foundation. Appears in SIGBOVIK 2020 with the {\tt ShortMoveString} of the
 Association for Computational Heresy; {\em IEEEEEE!} press,
@@
 
-In my experience, most chess games end in a few moves. Players who
-seek longevity typically do so by playing many games of chess, rather
-than a protracted XXX
-
-(shortest game is well known)
+In my experience, most chess games end in a few moves. If you want to
+play a lot of chess moves, you just play a lot of chess games. Still,
+there are games that seem to go on foreverrrrrrrrrr. Perhaps the
+players are trying to {\em lull each other into a false sense of
+security} while {\em waiting for the moment to strike}, or perhaps they
+are stalling in a game of {\em Chess to the Death}.
 
@@
 a minor chess scandal that erupted in 2019 over an obscure corner case
-in the rules.) Several of these ``deep cuts'' have to do with game-ending
-conditions that were introduced to keep the game from going on forever.
+in the rules.) Several of these ``deep cuts'' have to do with
+game-ending conditions that were introduced to avoid interminable
+games.
 
@@
 
-\begin{figure}
+\begin{figure}[b!]
   \begin{center}
@@
       % fillarea=a3-h1,printarea=a3-h1,
+      marginbottom=false,margintop=false,
       setfen=2nr1r1k/pb1n1pqp/1p2p1p1/1N1pN3/5P2/1P4P1/P2PPQ1P/2R1K1RB,
@@
 
-\begin{figure}[htp]
-  \begin{center}
-    \chessboard[setfen=8/r6n/8/8/5k2/3K4/7N/3b3Q b - - 0 1]
-  \end{center}
-  \caption{Black to move and mate in 545 moves (!). The position was
-    found (by Zakharov and Makhnichev~\cite{tablebase7}) while
-    building an endgame tablebase of all possible 7-piece positions.
-    Of course, the game ends prematurely in a draw because of the
-    75-move rule. } \label{fig:matein545}
-\end{figure}
-
 Draws after 75 moves (per player, so really 150 moves) are compulsory.
@@
 the same squares and the possible moves of all the pieces of both
-players are the same. Thus positions are not the same if:
+players are the same. [...]
+\comment{
+  Thus positions are not the same if:
 \begin{itemize}
@@
 \end{itemize}
+}
 \end{itemize}
@@
 
-% (I seem to remember that this rule used to be about moves, not positions,
-% but someone discovered infinite sequences of moves that don't repeat, right?
-% I can't find this citation. Maybe can recreate those sequences.)
+Fascinatingly (at least as fascinating as anything in this
+questionable undertaking), this rule used to require {\em consecutive}
+repetition of moves. However, there exist infinite sequences of moves
+with no consecutive $n$-fold repetition. For example, in the starting
+position, white and black can move either of their knights out and
+back. Let {\bf 0} be \knight c3 \Knight c6 \knight b1 \Knight b8
+(queenside knights move, returning to the starting position) and {\bf
+  1} be \knight f3 \Knight f6 \knight g1 \Knight g8 (kingside). Now
+the Prouhet--Thue--Morse sequence~\cite{thue1906uber}
+0110100110010110\ldots\footnote{
+  Let $s_0$ be the string {\tt "0"}, then
+  $s_i$ is $s_{i-1} \overline{s_{i-1}}$
+  where $\overline{\tt 0} = {\tt 1}$ and $\overline{\tt 1} = {\tt 0}$.
+}
 %
-% Fascinatingly (at least as fascinating as anything in this
-% questionable undertaking), this rule
+can be executed. This infinite
+sequence is cube-free (does not contain $SSS$ for any non-empty string
+$S$)~\cite{oeisa010060}, and therefore never violates the consecutive
+threefold repetition rule~\cite{euwe1929mengentheoretische}.
 
@@
 include any information about states previously reached.
+
+\begin{figure}[htp]
+  \begin{center}
+    \chessboard[setfen=8/r6n/8/8/5k2/3K4/7N/3b3Q b - - 0 1]
+  \end{center}
+  \caption{Black to move and mate in 545 moves (!). The position was
+    found (by Zakharov and Makhnichev~\cite{tablebase7}) while
+    building an endgame tablebase of all possible 7-piece positions.
+    Of course, the game ends prematurely in a draw because of the
+    75-move rule. } \label{fig:matein545}
+\end{figure}
 
@@
   \begin{center}
-    \chessboard[setfen=8/8/5K2/8/8/8/Q7/k7 b - - 0 1]
+    \chessboard[margintop=false,marginbottom=false,setfen=8/8/5K2/8/8/8/Q7/k7 b - - 0 1]
   \end{center}
@@
   \begin{center}
-    \chessboard[setfen=k1bB4/8/2p1p1p1/1pPpPpP1/1P1P1P2/8/8/K1Bb4 w - - 0 1]
+    \chessboard[margintop=false,marginbottom=false,setfen=k1bB4/8/2p1p1p1/1pPpPpP1/1P1P1P2/8/8/K1Bb4 w - - 0 1]
   \end{center}
@@
 
-First, the rules specifically define ``legal move'' in Article~3, and
-3.10.1 says ``A move is {\em legal} when all the relevant requirements
-of Articles 3.1 -- 3.9 have been fulfilled.'' These requirements
-describe the movement of each piece as you are familiar (e.g. 3.3
-``The rook may move to any square along the file or the rank on which
-it stands.''). They also disallow capturing one's own pieces, or
-moving when in check. However, they allow as legal some moves that
-would otherwise be prohibited, like capturing the opponent's king
-(this is excluded by 1.4.1, outside the definition of ``legal'').
-Capturing the opponent's king is generally not useful for
-demonstrating that checkmate is possible
-(Figure~\ref{fig:kingcapture}), so this is mostly a curiosity.
+First, the rules specifically define ``legal move'', with 3.10.1
+saying ``A move is {\em legal} when all the relevant requirements of
+Articles 3.1 -- 3.9 have been fulfilled.'' These requirements describe
+the movement of each piece as you are familiar (e.g. 3.3 ``The rook
+may move to any square along the file or the rank on which it
+stands.''). They also disallow capturing one's own pieces, or moving
+when in check. However, they allow as legal some moves that would
+otherwise be prohibited, like capturing the opponent's king (this is
+excluded by 1.4.1, outside the definition of ``legal''). Capturing the
+opponent's king is generally not useful for demonstrating that
+checkmate is possible (Figure~\ref{fig:kingcapture}), so this is
+mostly a curiosity.
 
@@
     \begin{tabular}{cc}
-      \chessboard[smallboard,marginleft=false,marginright=false,setfen=BBKBRRnk/PPPPPPnn/3P1BNR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] &
-      \chessboard[smallboard,marginright=false,setfen=BBKBRR1B/PPPPnP1n/3P2NR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] \\
+      \chessboard[smallboard,marginleft=false,marginright=false,setfen=BBKBRRnk/PPPPPPnn/3P1BRR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] &
+      \chessboard[smallboard,marginright=false,setfen=BBKBRR1B/PPPPnP1n/3P2RR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] \\
       (a) & (b) \\
@@
 \begin{center}
-  \chessboard[setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11,
+  \chessboard[
+    margintop=false,marginbottom=false,
+    setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11,
     pgfstyle=straightmove,markmoves={a6-b7,c3-b4,d6-c7,e6-f7,f3-g4,h6-g7},
@@
 
-The white b and g pawns have a clear route to promotion. White can
+\smallskip
+\noindent The white b and g pawns have a clear route to promotion. White can
 free each of the remaining 6 pawns with a single capture. Between this
@@
 \begin{center}
-  \chessboard[setfen=4RRNQ/p2pp2p/1k6/2p2p1N/2p2pQ1/R6Q/3N4/2BN1BKR b - - 0 81]
+  \chessboard[
+    margintop=false,marginbottom=false,
+    setfen=4RRNQ/p2pp2p/1k6/2p2p1N/2p2pQ1/R6Q/3N4/2BN1BKR b - - 0 81]
 \end{center}
 
-Now black can promote all of its pawns and capture white's pieces.
+\smallskip
+\noindent Now black can promote all of its pawns and capture white's pieces.
 We actually leave the white queen; this turns out to be essential:
@@
 \begin{center}
-  \chessboard[setfen=1r6/8/1k3K2/8/2b5/2n4r/7Q/2r1q1rr w - - 0 133]
+  \chessboard[
+    margintop=false,marginbottom=false,
+    setfen=1r6/8/1k3K2/8/2b5/2n4r/7Q/2r1q1rr w - - 0 133]
 \end{center}
 
-Finally, white captures all of black's pieces, and mates the black king.
+\smallskip
+\noindent Finally, white captures all of black's pieces, and mates the black king:
 
@@
 \begin{center}
-  \chessboard[setfen=3k4/3Q4/4K3/8/8/8/8/8 b - - 2 145]
+  \chessboard[
+    margintop=false,marginbottom=false,
+    setfen=3k4/3Q4/4K3/8/8/8/8/8 b - - 2 145]
 \end{center}
 
-As we switch which color is making critical moves a total of three
+\smallskip
+\noindent Since we switch which color is making critical moves a total of three
 times, we must come shy of the na\"ive maximum of 17,700 moves by
@@
 makes the checkmating move, black has made 75 non-critical moves, and
-white 74. White's 75th move mates.
+white 74. White's 75\th\ move mates.
 
@@
 but once two kings remain, the game ends immediately in a draw. So
-there is no length advantage here over checkmate. In fact we could do
-worse: Consider the position in Figure~\ref{fig:insufficient} where
-black is {\em forced} to capture the queen. This game is over prior to
-the capture, so this is shorter than the checkmating sequence! Even if
-the king had an escape square, it is arguable
-(Section~\ref{sec:ambiguity}) that the game is over in any case: There
-is no series of legal moves that leads to mate. Either the king
-captures the queen (and then clearly no mate is possible with just the
-two kings) or the king escapes, but in doing so plays the
-75\th\ non-critical move, and triggers a draw by the 75-move rule.
+there is no length advantage here over checkmate. In fact, attempting
+to capture may {\em shorten} the game: Consider the position in
+Figure~\ref{fig:insufficient} where black is {\em forced} to capture
+the queen. This game is over prior to the capture, so this is shorter
+than the checkmating sequence! Even if the king had an escape square,
+it is arguable (Section~\ref{sec:ambiguity}) that the game is over in
+any case: There is no ``series of legal moves'' that leads to mate.
+Either the king captures the queen (and then clearly no mate is
+possible with just the two kings) or the king escapes, but in doing so
+plays the 75\th\ non-critical move, and triggers a draw by the 75-move
+rule.
 
@@
 
-% XXX
-% Are there any practical implications of the
-
 \smallskip
@@
 what you expected?
+
+\smallskip
+The character called \queen\ often plays an important role in the story.
+How would you describe her personality? How does she develop over the
+course of the game?
 

### 2021-03-08 lowercase/paper/lowercase.tex
Multi font generation and tweaks. Add and start paper

%

% TODO: SDF vs SFD typo factory!

\documentclass[twocolumn]{amsart}
\usepackage[top=0.75in, left=0.65in, right=0.65in, bottom=0.6in]{geometry}

\usepackage{url}
% \usepackage{code}
% \usepackage{cite}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{chessboard}
% IPA symbols. safe turns off overrides for like \! which I still want
\usepackage[safe]{tipa}

% \usepackage{chessfs}
\usepackage{adjustbox}

\usepackage[most]{tcolorbox}

\interfootnotelinepenalty=0

% lets me explicitly set a. or 1. etc. as enum label
\usepackage{enumitem}

\pagestyle{empty}

\usepackage{ulem}
% go back to italics for emphasis, though
\normalem

\usepackage{natbib}

\setlength{\footnotesep}{2em}

% \newcommand\comment[1]{}
\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}

\begin{document}

\title{Lowestcase and Uppestcase letters: Further adventures in Derp Learning}
\author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
Appears in SIGBOVIK~2021 with the
OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1 em}

\setchessboard{showmover=false}

\newcommand\makelowercase{{\sf make\_lowercase}}
\newcommand\makeuppercase{{\sf make\_uppercase}}

\renewcommand\th{\ensuremath{{}^{\textrm{th}}}}
\newcommand\st{\ensuremath{{}^{\textrm{st}}}}
\newcommand\rd{\ensuremath{{}^{\textrm{rd}}}}
\newcommand\nd{\ensuremath{{}^{\textrm{nd}}}}
\newcommand\at{\ensuremath{\scriptstyle @}}

\date{1 April 2021}

\maketitle \thispagestyle{empty}

\sloppypar

\section{Introduction}

Have you ever been writing something on the internet and wanted to convey
that you ARE FEELING ANGRY? Conversely, have you ever fired back a super
quick dm and u wanted to make it clear that it was like super ca\textipa{Z}
and so u didnt use ne capitals or punctuation dots except 4 that one place
where u needed to use the international phonetic alphabet because u dont
no how to write ca\textipa{Z} as in short for casual without it lol

If so, you made use of the fact that all letters have UPPERCASE
VERSIONS (e.g.~signifying ANGER) and lowercase versions
(e.g.~signifying u dont care lol). These dimensions have other uses,
for example, it is polite to start a person's name with a capital
letter to show that you took the time to regard their humanity (as it
takes extra work to press the caps lock key, press the first letter of
their name, and then press ther caps lock key again to turn it off).
In German, Nouns start with uppercase Letters, signifying Superiority
over other grammatical Categories like Verbs and Adjectives. Lowercase
letters can be used to conserve printer ink. Actually, I'm not sure that
lowercase letters have any other uses, but let's just roll with it.

The thing is: What if I'm even MORE ANGRY THAN I WAS BEFORE? There are
some standard sorts of typographic emphasis, like I can be {\bf BOLD
  ANGRY} or \textbf{\textit{\large BIG BOLD ITALIC UNDERLINE ANGRY}}
or { \Large \textbf{\textit{\uuline{COMBINE A LOT OF THESE ANGERS}}}},
each with its own nuances, depending on the cascading style sheet or
LaTeX class file. To be even more casual than lowercase, u can learn 2
write like this, and {\scriptsize shrink away} and also \sout{cross
  out ur words in shame in advance of them even being read}, but there
are few other options for de-emphasis. Plus, when I'm FEELING PRETTY
ANGRY, TOM, how do I capitalize that already-capitalized T in order to
show the proper reverence for your humanity?

This paper is about unshackling this dimension of human expression by
introducing letterforms further along the uppercase and lowercase
dimensions. Basically, we want to know what the upper{\it er}case
version of uppercase T is, and a lower{\it er}case version of
lowercase t is.

\subsection{Induction}

Today we're just concerned with English letters, of which there are
only 26. To create an upperercase and lowerercase alphabet by hand is
O(52 pick up), which for a guy who likes drawing letters anyway and
who alphabetized Star~Wars for fun, is not much to ask. In fact I
drew such alphabets in Figure~\ref{fig:manual} just now.

\begin{figure}[ht]
% \includegraphics[width=0.9 \linewidth]{manual}
\caption{TODO} \label{fig:manual}
\end{figure}

But, why do easy fun things by hand when you can build a complicated
automatic solution which produces much worse results? Well, there is
no good reason. I could claim that this allows us to automatically
upperercase any font, which is true, but the results are at best
moderately letter-like lumps. In principle there are several other
interesting things we can do, like apply the function over and over to
approach the uppestcase and lowestcase letters. This sounds fun, but
the results themselves are not going to impress. But the story of
getting there may be interesting, and even as it turns out to be
``derp learning,'' there will be opportunities for more good puns. So
let's just roll with it!

\section{Capital A Artificial Intelligence}

% XXX want some macro for referring to a letter's shape, maybe
% with a box around it?

\newcommand\letterform[1]{\tcbox[
    nobeforeafter,
    tcbox raise base,
    top=0pt,bottom=0pt,left=-1pt,right=-1pt,
    left skip=1pt,
    right skip=1pt,
    arc=2pt,outer arc=2pt,
    boxrule=0.15mm,
    colback=white,
    colframe=white!50!black
    ]{\textrm{#1}}}

%     leftrule=0.2pt,rightrule=0.2pt,toprule=0.2pt,bottomrule=0.2pt,
%     enhanced jigsaw,
%    borderline horizontal={0.2pt}{0pt}{dashed},
%    borderline vertical={0.2pt}{0pt}{dashed}

We want to machine-learn two functions, \makelowercase\ and
\makeuppercase. Each takes a letterform and returns a letterform (we
can choose how these are represented) and does the needful, e.g.
\makelowercase(\letterform{A}) should return \letterform{a}. In order to learn this function, we'll at least need a lot of
examples to use as training data. A training example for
\makelowercase\ is a letterform and its expected corresponding
lowercase one. We can ``easily'' find a large amount of examples by
using existing fonts, and pairing their \letterform{A} with their
\letterform{a}, and so on for all 26 letters, and symmetrically for
\makeuppercase.

However, if we only give uppercase letters to \makelowercase, it
may very well learn how to generate the corresponding lowercase letter
but be unable to do anything interesting for other letterforms. This
is a problem because we want to use this function to see
what e.g.~\makelowercase(\letterform{a}) is.

This is not (only) the problem of overfitting. An overfit model could
work well on the letter \letterform{A} from one font (because it has
seen that font before) but fail on \letterform{A} from a new font. The
property that we want is that the learned function can also produce an
interesting result on a shape it's never seen before, like
\letterform{\textipa{Z}}\,. That is, it has generalized the idea of
``how to make a shape lowercase,'' not simply ``how to make a capital
A shape lowercase.''

The problem with this is that we don't have any training data other
than existing fonts to tell us what the lowercase of some arbitrary
shape should look like. Without examples of this form, the problem is
unconstrained. \makelowercase\ could learn to generate empty output
for anything it doesn't recognize as a capital letter, and still have
perfect performance on the training and test set. It is hard to
generate training data of this form (even by hand) as we don't have
much idea {\em a priori} of what a lowerercase \letterform{a} should
look like (except for e.g.~One Artist's Impression from
Figure~\ref{fig:manual}).

\newcommand\trainingexample[2]{$\langle$\letterform{#1}$,\,$\letterform{#2}$\rangle$}

\newcommand\weirdcharlo{\includegraphics[width=0.8em]{weirdchar-lo}}
\newcommand\weirdcharup{\includegraphics[width=0.8em]{weirdchar-up}}

This brings us to the one decent idea in this paper (which by the way
only sort of works, but let's just roll with it). We can at least
express one characteristic property of the \makelowercase\ function
that ought to be true even for letterforms we don't have examples of:
It ought to be the inverse of \makeuppercase. So, we train these two
models in tandem. \makelowercase\ is fed training examples from the
sample fonts like \trainingexample{Q}{q} etc.~and \makeuppercase\ gets
\trainingexample{e}{E} etc.~as expected. We also run the current
version of \makeuppercase\ on some letter-like shapes, which produces
some other shape. For example, say that
\makeuppercase(\letterform{\weirdcharlo}) outputs
\letterform{\weirdcharup}. We have no idea if this is good or not, so
we don't update the model. However, we {\em do} provide the training
example to \trainingexample{\weirdcharup}{\weirdcharlo} to the
\makelowercase\ training queue and penalize {\em it} if it did not
predict \letterform{\weirdcharup}. In this way, whatever \makeuppercase\ is
doing, we ask \makelowercase\ to learn the inverse. We of course also
simultaneously do the symmetric thing, using the output of
\makelowercase\ to create training examples for
\makeuppercase\ (Figure~\ref{fig:cotraining}).

\begin{figure}[ht]
% \includegraphics[width=0.9 \linewidth]{manual}
\caption{TODO cotraining} \label{fig:cotraining}
\end{figure}

Because \makelowercase\ is getting training examples of
uppercase/lowercase pairs from real fonts, it remains grounded on real
letters. It is also free to generate new shapes in the open domain
(outside \letterform{A}--\letterform{Z}). However, it is penalized if
its behavior is not the inverse of whatever \makeuppercase\ is
currently doing. And since we do the symmetric thing for
\makeuppercase\, there is a (slow) feedback loop between the two
models that keeps them from straying too far from the grounded
examples. The idea is that this allows them to do some creative
generalization outside their native domains, but in a way that
still has some constraint.

In practice, we don't feed arbitrary shapes to the models. We
just need something letter-like, and in fact we have a large
collection of letter-like shapes among our existing fonts. So
we pass already-lowercase shapes to \makelowercase, in order
to generate inversion examples for training \makeuppercase.
These shapes are clearly letter-like (they {\em are} letters) and
are also of interest to us anyway, since we want to try to
generate lowerercase and upperercase letters with these models.

% $\langle\,$\letterform{Q}, \letterform{q}$\,\rangle$
% $\langle\,$\letterform{e}, \letterform{E}$\,\rangle$

\section{1000001 Free Fonts}

% no copyright intended

% (lowercase i intelligence: redi)

% The care and feeding of sparse matrices

% Perfect letters, hallucinated

% autotracing

% CAPS LOCK, LOCK, CAPTAIN'S LOCK

% when two letters coincide, a shift-reduce conflict

% cool s

% lowercase uppercase letters etc.

\nocite{murphy2019blind}
\nocite{murphy2019eloworld}

\bibliography{lowercase}{}
\bibliographystyle{plain}

\end{document}

### 2021-03-08 lowercase/font-problem.cc

@@ -2496,15 +2496,21 @@ float FontProblem::GuessRightEdge(
   // The inner box is the nominal character box, and we are trying to
   // guess its right edge (like how pad_left defines a left edge).
   // Consider only pixels >pad_top and <pad_bot. Sum up the
-  // total SDF power in each column:
+  // total SDF power in each column, and the number of pixels above
+  // the onedge value.
 
   vector<float> power;
+  vector<int> content;
   for (int x = 0; x < config.sdf_size; x++) {
     float col_power = 0.0f;
+    int col_content = 0;
     for (int y = config.pad_top; y < config.sdf_size - config.pad_bot; y++) {
-      col_power += sdf.GetPixel(x, y);
+      const float f = sdf.GetPixel(x, y);
+      col_power += f;
+      if (f > config.onedge_value / 255.0f) col_content++;
     }
     power.push_back(col_power);
+    content.push_back(col_content);
   }
 
   // Now find the last column that has power less than the left edge's
@@ -2514,6 +2520,11 @@ float FontProblem::GuessRightEdge(
   for (int x = config.sdf_size - 1; x >= 0; x--) {
     if (power[x] >= target) {
       // TODO: interpolate for better quality!
+      return x + 1;
+    }
+    // Don't allow any of the actual shape to protrude, either.
+    // (Could also interpolate here.)
+    if (content[x] > 0) {
       return x + 1;
     }
   }
@@ -2528,6 +2539,21 @@ float FontProblem::TTFBaseline(const SDFConfig &config) {
   return 1.0f - (config.pad_bot / (float)(config.sdf_size - config.pad_top));
 }
 
+// XXX probably useless?
+void FontProblem::ScaleChar(float scale, TTF::Char *ch) {
+  for (TTF::Contour &contour : ch->contours) {
+    for (TTF::Path &p : contour.paths) {
+      p.x *= scale;
+      p.y *= scale;
+      if (p.type == TTF::PathType::BEZIER) {
+        p.cx *= scale;
+        p.cy *= scale;
+      }
+    }
+  }
+
+  ch->width *= scale;
+}
 
 TTF::Char FontProblem::ToChar(
     const SDFConfig &config,

### 2021-03-08 lowercase/ttf.h

@@ -177,6 +177,10 @@ struct TTF {
     std::map<char, Char> chars;
     float baseline = 0.75f;
     float linegap = 0.0f;
+    // Additional scale factor with baseline origin. Typical use is to
+    // remove unwanted space above the character. Affects widths but
+    // not baseline, ascent, or descent.
+    float extra_scale = 1.0f;
 
     // Serialize and deserialize.
     std::string ToString() const;

### 2021-03-08 trunk/lowercase/font-problem.h
Multi font generation and tweaks. Add and start paper

@@ -304,6 +304,10 @@ struct FontProblem {
       const std::vector<TTF::Contour> &contours,
       float right_edge);
 
+  // Scale the glyph using an origin of 0,0. This changes its location
+  // relative to the baseline so it might not be that useful.
+  static void ScaleChar(float scale, TTF::Char *ch);
+
   // Derive the TTF::Font baseline from the SDF config (the bottom padding).
   static float TTFBaseline(const SDFConfig &config);
 

### 2021-03-10 trunk/lowercase/paper/lowercase.tex
Some work on paper

--- 2021-03-08 lowercase/paper/lowercase.tex
@@
 
-% XXX want some macro for referring to a letter's shape, maybe
-% with a box around it?
+% XXX introduce the meaning of the \letterform syntax somewhere?
 
@@
 
-\newcommand\weirdcharlo{\includegraphics[width=0.8em]{weirdchar-lo}}
-\newcommand\weirdcharup{\includegraphics[width=0.8em]{weirdchar-up}}
+\newcommand\weirdcharlo{\includegraphics[width=0.75em]{weirdchar-lo}}
+\newcommand\weirdcharup{\includegraphics[width=0.75em]{weirdchar-up}}
+\newcommand\lowerlowera{\includegraphics[width=0.6em]{lowerlowera}}
 
@@
 \begin{figure}[ht]
-% \includegraphics[width=0.9 \linewidth]{manual}
-\caption{TODO cotraining} \label{fig:cotraining}
+\includegraphics[width=0.9 \linewidth]{training}
+\caption{ Simultaneously training the two models. This example
+  illustrates how a pair of letterforms \letterform{A} and
+  \letterform{a} from the same font becomes four training examples.
+  The pair straightforwardly generates an example
+  \trainingexample{A}{a} for the \makelowercase\ queue, and an example
+  \trainingexample{a}{A} for the \makeuppercase\ queue. Separately, we
+  supply \letterform{a} to the \makelowercase\ model, simply to get
+  the current output \letterform{\lowerlowera} (no model updates are
+  performed). But this pair reversed becomes a training example
+  \trainingexample{\lowerlowera}{a} for the \makeuppercase\ queue.
+} \label{fig:cotraining}
 \end{figure}
@@
 uppercase/lowercase pairs from real fonts, it remains grounded on real
-letters. It is also free to generate new shapes in the open domain
+letters. It is also free to generate new shapes for the open domain
 (outside \letterform{A}--\letterform{Z}). However, it is penalized if
@@
 
-In practice, we don't feed arbitrary shapes to the models. We
-just need something letter-like, and in fact we have a large
-collection of letter-like shapes among our existing fonts. So
-we pass already-lowercase shapes to \makelowercase, in order
-to generate inversion examples for training \makeuppercase.
-These shapes are clearly letter-like (they {\em are} letters) and
-are also of interest to us anyway, since we want to try to
-generate lowerercase and upperercase letters with these models.
-
-% $\langle\,$\letterform{Q}, \letterform{q}$\,\rangle$
-% $\langle\,$\letterform{e}, \letterform{E}$\,\rangle$
+In practice, we don't feed arbitrary shapes to the models. We just
+need something letter-like, and in fact we have a large collection of
+letter-like shapes among our existing fonts! We pass already-lowercase
+shapes to \makelowercase, in order to generate inversion examples for
+training \makeuppercase. These shapes are clearly letter-like (they
+{\em are} letters) and are also of interest to us anyway, since we
+want to try to generate lowerercase and upperercase letters from
+the trained models.
 
@@
 
+Sprechen of fonts, I downloaded every font I could find on the whole
+internet. This was overkill. The resulting directory tree contained
+over 100,000 files, many of which were duplicates. Exact duplicates
+are easy to find, but since many of these files were the result of 30
+years of community transmission, they had acquired various mutations.
+One of the first things I did was write software to automatically
+remove files that were essentially duplicates even if they weren't
+exactly the same bytes.
+
+Next, my lord, do people have bad taste! And I say this as someone who
+made dozens of amateurish fonts\cite{dbzfonts} as a high school and
+college student and who is contributing several new questionable
+fonts\label{sec:newfonts} as a result of this paper. The database is
+just filled with garbage that is unusable for this project: Fonts that
+are completely illegible, fonts that are missing most of their
+characters, fonts with millions of control points, Comic Sans MS,
+fonts where every glyph is a drawing of a train, fonts where
+everything is fine except that just the lowercase r has a width of
+{\tt MAX\_INT}, and so on. So I built a UI
+(Figure~\ref{fig:sortition}) for efficiently and mind-numbingly
+cleaning up the database by marking fonts as broken or suitable (and
+also categorizing them as serif, sans-serif, decorative, techno, etc.,
+which I never used). In doing this I noticed another extremely common
+problem, which was that many fonts had the same letter shapes for
+uppercase and lowercase letters. This would not do for the current
+application!
+
+\begin{figure*}[ht]
+\includegraphics[width=0.9 \linewidth]{sortition}
+\caption{ The interactive font data-cleaning UI. A seemingly endless
+  series of fonts presents, with single keypresses putting the fonts
+  into common categories such as (b)roken.
+  % XXX maybe write more?
+} \label{fig:sortition}
+\end{figure*}
+
+But why manually mark fonts with nearly the same upper- and
+lowercase letters, when you could build a complicated automatic
+solution? The first pass identified fonts whose letters were
+exactly the same, but this was only a small fraction of the
+problematic fonts. A common issue was that the lowercase characters
+were very slightly modified versions of the uppercase ones, often
+scaled and translated and then perhaps ``optimized'' during the
+font export.
+
+So, for a given font, I want to reject it if for most pairs of cased
+letters \letterform{A},\letterform{a}, \letterform{a} is close to a
+linear transformation of \letterform{A}. This problem can probably be
+solved with math, but it didn't sound that fun. Instead I tried out
+a new tool, and it worked well enough that I've now added it to the
+permanent rotation: Black-box function optimizers.
+
+\paragraph{Black-box optimization.} If you have a function and want
+to find arguments that minimize its output, the most efficient
+techniques are generally those like gradient descent. (In fact, the
+backpropagation algorithm we use to training the neural network in
+Section~\ref{sec:neural} is gradient descent on the function that
+takes the model weights and produces an error value for each output
+node.) The problem with this is that you need to do some math to
+compute the gradient function, and anyway you need to deal with fiddly
+bits (Section~\ref{sec:fiddly}) unless the function is convex and
+smooth, which it will not be. If you don't want to deal with that,
+and have a fast computer (and who doesn't?), black-box optimization
+algorithms are worth considering. Here, the interface\footnote{
+  Here a simplfied wrapper around BiteOpt~\cite{biteopt} in my {\tt cc-lib}
+  library. See \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/cc-lib/opt/}.}
+  is just something like (C++):
+
+\begin{verbatim}
+  double Minimize1D(
+      const std::function<double(double)> &f,
+      double lower_bound,
+      double upper_bound,
+      int iters);
+\end{verbatim}
+
+which takes a function \verb+f+ of type \verb+double+ $\rightarrow$
+\verb+double+, finite bounds on the argument's value, the maximum
+number of times to call that function, and returns the argument it
+found that produced the minimal value. Not as fast as gradient
+descent, but in practice ``if the function is kinda smooth'' these
+optimizers produce excellent results! The chief selling point for me
+is that I don't need to think about anything except for writing the
+function that I want minimized, which I just express in normal code.
+
+\begin{figure}[ht]
+\includegraphics[width=0.9 \linewidth]{casealignment}
+\caption{ Example alignment to reject the font
+  {\tt DrippingGooExtended}.
+  At left, \letterform{A} (red) and
+  \letterform{a} (green) rendered with the identity transform, and
+  their alignment ($35\%$ difference) below. At right, the transform
+  found by the black-box optimizer and the resulting alignment with
+  $1.7\%$ difference. Note that the shapes are still not an exact
+  match (the export process has to round the data to integers and
+  might apply other non-linear transformations like curve
+  simplification), but these are clearly not a useful pair for
+  the current problem.} \label{fig:casealignment}
+\end{figure}
+
+In this case, I render the letterform \letterform{A} and then optimize
+a four argument function taking {\tt xoff}, {\tt yoff}, {\tt xscale},
+{\tt yscale}. This function renders \letterform{a} with those
+parameters, then just computes the difference in the two rendered
+bitmaps. This finds the best alignment of the two letterforms (under
+the linear transformation) in a few hundred milliseconds
+(Figure~\ref{fig:casealignment}). If the disagreement is low as a
+function of the total pixels, then we say that the letters have the
+same case. If enough of them have the same case, we reject the font. I
+set the thresholds by looking at the P/R curve computed on random
+hand-labeled examples (Figure~\ref{fig:samecasepr}).
+
+\begin{figure}[ht]
+\includegraphics[width=0.9 \linewidth]{samecasepr}
+\caption{ Precision--recall curve for automatically detecting fonts
+  that have basically the same upper- and lowercase shapes. It's good!
+  This is how you want 'em to look!
+} \label{fig:samecasepr}
+\end{figure}
+
+\medskip
+
+I labeled fonts using the UI until I had 10,000 that were clean enough
+for a training set and passed the same-case heuristic.
+
+\section{The simplest thing that might work}
+
+Before getting fancy (which we we will) it's good engineering hygiene
+to try the simpliest thing that might just work (it doesn't). Fonts
+are represented as vector data (lines and quadratic B\'ezier curves).
+Can we just train a network that takes these lines and curves as input
+and predicts the lower- or uppercased letter in the same format? (No.)
+
+We'll at least put the data in a somewhat normalized form. The neural
+network will take a fixed number of inputs to a fixed number of
+outputs, so a simple approach is to decide on some maximum number
+control points per letter, and only try training on fonts whose
+letterforms fit within that budget. Letterforms can be made of
+multiple contours (e.g. a stacked \letterform{g} typically has two
+holes in it, and \letterform{j} has two disjoint parts). I found that
+most clean fonts had three or fewer contours, and when sorting them by
+descending length, most fit within 38, 14, and 10 endpoints for the
+three. So, I only train on fonts where all of the letters fit within
+this budget.\footnote{It would not be a good idea to reject only the letters
+that don't fit, because it might result in the network being trained
+on more \letterform{l}s (tends to be simple) than \letterform{g}s
+(tends to be complex).}
+
+% TODO: Illustrate a letterform and its contours?
+
+Rather than try to work with both lines and B\'ezier curves, I
+normalize each contour to only contain B\'eziers, by turning a line
+segment into an equivalent B\'ezier with its control point at the
+midpoint. This frees us from having to distinguish the two types in
+the data. We also need each of the three contours to not be {\em too
+  short}, so I fill out the fixed-size buffers by repeating the last
+point. This is not great but does have the correct meaning (bunch of
+useless zero-length edges). It has the property that any predicted
+data can be rendered and has a straightforward point-by-point error
+function (which might not be the case if we were predicting a dynamic
+number of points). Finally, I output the contour's points starting
+from the point closest to the origin in order to reduce one needless
+degree of freedom.
+
+% XXX total number of parameters?
+The network I trained has an input layer size of $154 = (38 + 14 + 10)
+\times 4 + 3 \times 2$ (one control point and one end point per
+B\'ezier curve), plus a starting point for each of the three contours.
+The output layer is the same size, plus 26 (see below). There are
+three hidden layers of size 308, 308, 360. The first two layers are
+dense and the third sparse (about half of the weights are zero). All
+layers are leaky rectified linear units. If you're taking notes,
+don't, as again this does not work well, and I don't know how people
+figure out what the right network size is anyway. I just made it up.
+You can give {\em me} your notes.
+
+\newcommand\nan{\textsf{NaN}}
+\renewcommand\inf{\textsf{inf}}
+
+I trained the network using a home-grown (why??) GPU-based package
+that I wrote for {\em Red i removal with artificial retina
+  networks}~\cite{murphy2015redi}---an example of ``lowercase i
+artificial intelligence''---and have improved as I repurposed it for
+other projects, such as {\em Color- and piece-blind
+  chess}~\cite{murphy2019blind}. It is ``tried and true'' in the
+sense that ``every time I tried using it, I truly wanted to throw
+my computer out the window, and retire to a hermitage in the glade
+whenceforth I shall nevermore be haunted by a model which has
+overnight become a sea of \inf{}s and \nan{}s.''
+
+% TODO: Screenshot of this thing's output.
+% TODO: Font file generated from this.
+% TODO: Fitting the prediction to the actual in order to coax
+% better results, failing.
+
+% the sdf problem
+
+\section{The care and feeding of sparse matrices} \label{sec:neural}
+
+\label{sec:fiddly}
+
 % no copyright intended
 
-% (lowercase i intelligence: redi)
+% stretch the loss function near the onedge value
 
-% The care and feeding of sparse matrices
+%
 

### 2021-03-11 trunk/lowercase/network.cc
Generate font from vector version, write about this

@@ -97,6 +97,14 @@ int64 Network::Bytes() const {
   return ret;
 }
 
+int64 Network::TotalParameters() const {
+  int64 params = 0LL;
+  for (int i = 0; i < num_layers; i++) {
+    params += layers[i].weights.size() + layers[i].biases.size();
+  }
+  return params;
+}
+
 void Network::RunForward(Stimulation *stim) const {
   for (int src = 0; src < num_layers; src++) {
     RunForwardLayer(stim, src);
@@ -249,7 +257,7 @@ void Network::StructuralCheck() const {
   CHECK(renderstyle.size() == num_layers + 1);
 
   CHECK(inverted_indices.size() == num_layers);
-  
+
   for (int i = 0; i < num_layers; i++) {
     const Layer &layer = layers[i];
     const int num_prev_nodes = num_nodes[i];

### 2021-03-11 lowercase/paper/lowercase.tex
Generate font from vector version, write about this

@@
 
+Sprechen of fonts, I downloaded every font I could find on the whole
+internet. This was overkill. The resulting directory tree contained
+over 100,000 files, many of which were duplicates. Exact duplicates
+are easy to find, but since many of these files were the result of 30
+years of community transmission, they had acquired various mutations.
+One of the first things I did was write software to automatically
+remove files that were essentially duplicates even if they weren't
+exactly the same bytes.
+
+Next, my lord, do people have bad taste! And I say this as someone who
+made dozens of amateurish fonts\cite{dbzfonts} as a high school and
+college student and who is contributing several new questionable
+fonts\label{sec:newfonts} as a result of this paper. The database is
+just filled with garbage that is unusable for this project: Fonts that
+are completely illegible, fonts that are missing most of their
+characters, fonts with millions of control points, Comic Sans MS,
+fonts where every glyph is a drawing of a train, fonts where
+everything is fine except that just the lowercase r has a width of
+{\tt MAX\_INT}, and so on. So I built a UI
+(Figure~\ref{fig:sortition}) for efficiently and mind-numbingly
+cleaning up the database by marking fonts as broken or suitable (and
+also categorizing them as serif, sans-serif, decorative, techno, etc.,
+which I never used). In doing this I noticed another extremely common
+problem, which was that many fonts had the same letter shapes for
+uppercase and lowercase letters. This would not do for the current
+application!
+
+\begin{figure*}[ht]
+\includegraphics[width=0.9 \linewidth]{sortition}
+\caption{ The interactive font data-cleaning UI. A seemingly endless
+  series of fonts presents, with single keypresses putting the fonts
+  into common categories such as (b)roken.
+  % XXX maybe write more?
+} \label{fig:sortition}
+\end{figure*}
+
+But why manually mark fonts with nearly the same upper- and
+lowercase letters, when you could build a complicated automatic
+solution? The first pass identified fonts whose letters were
+exactly the same, but this was only a small fraction of the
+problematic fonts. A common issue was that the lowercase characters
+were very slightly modified versions of the uppercase ones, often
+scaled and translated and then perhaps ``optimized'' during the
+font export.
+
+So, for a given font, I want to reject it if for most pairs of cased
+letters \letterform{A},\letterform{a}, \letterform{a} is close to a
+linear transformation of \letterform{A}. This problem can probably be
+solved with math, but it didn't sound that fun. Instead I tried out
+a new tool, and it worked well enough that I've now added it to the
+permanent rotation: Black-box function optimizers.
+
+\paragraph{Black-box optimization.} If you have a function and want
+to find arguments that minimize its output, the most efficient
+techniques are generally those like gradient descent. (In fact, the
+backpropagation algorithm we use to training the neural network in
+Section~\ref{sec:neural} is gradient descent on the function that
+takes the model weights and produces an error value for each output
+node.) The problem with this is that you need to do some math to
+compute the gradient function, and anyway you need to deal with fiddly
+bits (Section~\ref{sec:fiddly}) unless the function is convex and
+smooth, which it will not be. If you don't want to deal with that,
+and have a fast computer (and who doesn't?), black-box optimization
+algorithms are worth considering. Here, the interface\footnote{
+  Here a simplfied wrapper around BiteOpt~\cite{biteopt} in my {\tt cc-lib}
+  library. See \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/cc-lib/opt/}.}
+  is just something like (C++):
+
+\begin{verbatim}
+  double Minimize1D(
+      const std::function<double(double)> &f,
+      double lower_bound,
+      double upper_bound,
+      int iters);
+\end{verbatim}
+
+which takes a function \verb+f+ of type \verb+double+ $\rightarrow$
+\verb+double+, finite bounds on the argument's value, the maximum
+number of times to call that function, and returns the argument it
+found that produced the minimal value. Not as fast as gradient
+descent, but in practice ``if the function is kinda smooth'' these
+optimizers produce excellent results! The chief selling point for me
+is that I don't need to think about anything except for writing the
+function that I want minimized, which I just express in normal code.
+
+\begin{figure}[ht]
+\includegraphics[width=0.9 \linewidth]{casealignment}
+\caption{ Example alignment to reject the font
+  {\tt DrippingGooExtended}.
+  At left, \letterform{A} (red) and
+  \letterform{a} (green) rendered with the identity transform, and
+  their alignment ($35\%$ difference) below. At right, the transform
+  found by the black-box optimizer and the resulting alignment with
+  $1.7\%$ difference. Note that the shapes are still not an exact
+  match (the export process has to round the data to integers and
+  might apply other non-linear transformations like curve
+  simplification), but these are clearly not a useful pair for
+  the current problem.} \label{fig:casealignment}
+\end{figure}
+
+In this case, I render the letterform \letterform{A} and then optimize
+a four argument function taking {\tt xoff}, {\tt yoff}, {\tt xscale},
+{\tt yscale}. This function renders \letterform{a} with those
+parameters, then just computes the difference in the two rendered
+bitmaps. This finds the best alignment of the two letterforms (under
+the linear transformation) in a few hundred milliseconds
+(Figure~\ref{fig:casealignment}). If the disagreement is low as a
+function of the total pixels, then we say that the letters have the
+same case. If enough of them have the same case, we reject the font. I
+set the thresholds by looking at the P/R curve computed on random
+hand-labeled examples (Figure~\ref{fig:samecasepr}).
+
+\begin{figure}[ht]
+\includegraphics[width=0.9 \linewidth]{samecasepr}
+\caption{ Precision--recall curve for automatically detecting fonts
+  that have basically the same upper- and lowercase shapes. It's good!
+  This is how you want 'em to look!
+} \label{fig:samecasepr}
+\end{figure}
+
+\medskip
+
+I labeled fonts using the UI until I had 10,000 that were clean enough
+for a training set and passed the same-case heuristic.
+
+\section{The simplest thing that might work} \label{sec:vectorversion}
+
+Before getting fancy (which we we will) it's good engineering hygiene
+to try the simpliest thing that might just work (it doesn't). Fonts
+are represented as vector data (lines and quadratic B\'ezier curves).
+Can we just train a network that takes these lines and curves as input
+and predicts the lower- or uppercased letter in the same format? (No.)
+
+We'll at least put the data in a somewhat normalized form. The neural
+network will take a fixed number of inputs to a fixed number of
+outputs, so a simple approach is to decide on some maximum number
+control points per letter, and only try training on fonts whose
+letterforms fit within that budget. Letterforms can be made of
+multiple contours (e.g. a stacked \letterform{g} typically has two
+holes in it, and \letterform{j} has two disjoint parts). I found that
+most clean fonts had three or fewer contours, and when sorting them by
+descending length, most fit within 38, 14, and 10 endpoints for the
+three. So, I only train on fonts where all of the letters fit within
+this budget.\footnote{It would not be a good idea to reject only the letters
+that don't fit, because it might result in the network being trained
+on more \letterform{l}s (tends to be simple) than \letterform{g}s
+(tends to be complex).}
+
+% TODO: Illustrate a letterform and its contours?
+
+Rather than try to work with both lines and B\'ezier curves, I
+normalize each contour to only contain B\'eziers, by turning a line
+segment into an equivalent B\'ezier with its control point at the
+midpoint. This frees us from having to distinguish the two types in
+the data. We also need each of the three contours to not be {\em too
+  short}, so I fill out the fixed-size buffers by repeating the last
+point. This is not great but does have the correct meaning (bunch of
+useless zero-length edges). It has the property that any predicted
+data can be rendered and has a straightforward point-by-point error
+function (which might not be the case if we were predicting a dynamic
+number of points). Finally, I output the contour's points starting
+from the point closest to the origin in order to reduce one needless
+degree of freedom.
+
+The network I trained has an input layer size of $154 = (38 + 14 + 10)
+\times 4 + 3 \times 2$ (one control point and one end point per
+B\'ezier curve), plus a starting point for each of the three contours.
+The output layer is the same size, plus 26 (see below). There are
+three hidden layers of size 308, 308, 360. The first layer is dense
+and the remainder are sparse (two thirds to a half of the weights are
+zero), for just about 1 million total parameters. All layers are leaky
+rectified linear units. If you're taking notes, don't, as again this
+does not work well, and I don't know how people figure out what the
+right network size is anyway. I just made it up. You can give {\em me}
+your notes.
+
+\paragraph{Bonus outputs}. The output includes the predicted shape,
+and also 26 individual predictors for each of the 26 letters. So a
+training example is actually like \letterform{C} $\rightarrow$
+\letterform{c} $[0, 0, 1, 0, 0, \ldots, 0]$, with the 1 in the third
+place because C is the third letter. We don't need these outputs for
+the problem itself (e.g. to lowercase new letter shapes), but there
+are several ideas behind this. First, the lowercasing function we're
+trying to learn does take into account the letter of the alphabet
+being lowercased. By asking the network to learn this as well (it is
+penalized when it gets the prediction wrong), it must learn features
+that allow it to distinguish different letters, and those features are
+available for use by outputs we {\em do} care about. This is an easy
+way to coax it to learn features that I know are meaningful without
+having to actually engineer feature extractors by hand (or first train
+separate models, etc.). Perhaps the most useful thing is that it's
+very clear what the right answer is, so it gives me an easy way to see
+if the network is learning anything {\em at all}. (It does.) Finally,
+we can do some silly stuff with these; see
+Section~\ref{sec:hallucination}.
+
+\newcommand\nan{\textsf{NaN}}
+\renewcommand\inf{\textsf{inf}}
+
+I trained the network using a home-grown (why??) GPU-based package
+that I wrote for {\em Red i removal with artificial retina
+  networks}~\cite{murphy2015redi}---an example of ``lowercase i
+artificial intelligence''---and have improved as I repurposed it for
+other projects, such as {\em Color- and piece-blind
+  chess}~\cite{murphy2019blind}. It is ``tried and true'' in the
+sense that ``every time I tried using it, I truly wanted to throw
+my computer out the window, and retire to a hermitage in the glade
+whenceforth I shall nevermore be haunted by a model which has
+overnight become a sea of \inf{}s and \nan{}s.''
+
+\begin{figure}[ht]
+\includegraphics[width=0.9 \linewidth]{trainingvector}
+\caption{ Screenshot of {\tt train.exe} running on the vector-based
+  version of the problem. Shown is the \makelowercase\ model's output
+  (bottom) on four shapes from four fonts (top). Some dust in between
+  is the activation of the network's layers. At the very bottom, the
+  26 predictions for ``what letter is this?''. The output for
+  \letterform{j} is not too bad; you can see the distinct dot (a
+  separate contour) and it sort of looks like a \letterform{j}. The
+  \letterform{e} also has two pieces as expected but is otherwise
+  garbage. The model is unsure whether the second input is an H
+  or a K, and has predicted a shape sort of ambiguously between those
+  two. The \lowercase{z} is also an embarrassment.
+} \label{fig:trainingvector}
+\end{figure}
+
+I was so {\tiny confident} that this wouldn't work that I only trained
+a \makelowercase\ model and didn't even worry about the more
+complicated simultaneous training setup yet. I ran this for about
+22,000 rounds, some 90 million training examples. Indeed it does not
+work (Figure~\ref{fig:trainingvector}). It is not a total catastrophe,
+though. The model can clearly recognize letters (though perhaps just by
+memorization) and some of the shapes it generates are along the right
+lines. ({\em Along the right lines}, get it??)
+
+It's a bit hard on the
+
+% TODO: Font file generated from this.
+% TODO: Fitting the prediction to the actual in order to coax
+% better results, failing.
+
+% XXX cite futura
+% XXX add downloadable font too
+\begin{figure*}[tp]
+\includegraphics[width=0.9 \linewidth]{futurda}
+\caption{ Type specimen for the generated font {\bf Futurda}. This is the
+  classic font Futura, run through the vector based lowercasing model
+  (Section~\ref{sec:vectorversion}). At top are Futura's letterforms
+  \letterform{A}--\letterform{Z} run through the model and so ``made
+  lowercase'' (it's obviously garbage). Next are the \letterform{a}--\letterform{z},
+  made even more lowercase. Also rubbish.
+  At the bottom are the pangrams ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
+  and ``Sphinx of black quartz, judge my vow!'' Although the text is
+  illegible and kind of uninteresting, it does have a certain dynamic
+  Rorschach aesthetic, like a collection of delicate moths pinned
+  to paperboard, that one could consider framing or publishing in
+  the proceedings of SIGBOVIK 2021.
+} \label{fig:futurda}
+\end{figure*}
+
+% lowerercase Franklin Gothic. Franklin mint... not guaranteed
+% to go up in value
+
+% the sdf problem
+
+\section{The care and feeding of sparse matrices} \label{sec:neural}
+
+\label{sec:fiddly}
+
 % no copyright intended
 
-% (lowercase i intelligence: redi)
+% stretch the loss function near the onedge value
 
-% The care and feeding of sparse matrices
+%
 
-% Perfect letters, hallucinated
+\section{Perfect letters, hallucinated}
+\label{sec:hallucination}
 

### 2021-03-11 lowercase/get-modelinfo.cc

@@ -15,21 +15,21 @@ using namespace std;
 
 int main(int argc, char **argv) {
   const string modelfile = argc > 1 ? (string)argv[1] : "net0.val";
-  
+
   // Try loading from disk; null on failure.
   printf("Load network from %s...\n", modelfile.c_str());
   std::unique_ptr<Network> net(Network::ReadNetworkBinary(modelfile));
-  
+
   CHECK(net.get() != nullptr) << modelfile;
 
   // Get histogram of weights per layer.
   // Only layers with inputs (i.e. not the input layer) have weights.
   const int num_histos = net->num_layers;
-    
+
   static constexpr int HISTOW = 800;
   static constexpr int HISTOH = 220;
   static constexpr int MARGIN = 4;
-  
+
   const int WIDTH = HISTOW * 2 + MARGIN;
   const int HEIGHT = HISTOH * num_histos;
 
@@ -46,14 +46,15 @@ int main(int argc, char **argv) {
   // XXX even though TZ is set and 'date +"%H:%M"' works as expected
   // in mingw, this reports UTC time?
   tt -= 3600 * 5;
-  
+
   strftime(dates, 127, "%d %b %Y  %H:%M", localtime(&tt));
   vector<string> lines = {
     StringPrintf("%s  round %lld   examples %lld   bytes %lld   real layers %d",
                  dates,
                  net->rounds, net->examples, net->Bytes(), net->num_layers),
-    StringPrintf("  Input: %dx%dx%d = %d",
-                 net->width[0], net->height[0], net->channels[0], net->num_nodes[0]),
+    StringPrintf("  Input: %dx%dx%d = %d                %lld total params",
+                 net->width[0], net->height[0], net->channels[0], net->num_nodes[0],
+                 net->TotalParameters()),
   };
 
   for (int layer_idx = 0; layer_idx < net->num_layers; layer_idx++) {
@@ -78,10 +79,10 @@ int main(int argc, char **argv) {
   for (int i = 0; i < lines.size(); i++) {
     img.BlendText2x32(0, i * 20, 0xCCCCCCFF, lines[i]);
   }
-  
+
   const string outfile = argc > 2 ? (string)argv[2] : "modelinfo.png";
   img.Save(outfile);
-  
+
   printf("Wrote %s.\n", outfile.c_str());
   return 0;
 }

### 2021-03-11 lowercase/vector-makefont.cc

@@ -1,0 +1,163 @@
+
+#include <vector>
+#include <string>
+#include <algorithm>
+#include <cstdint>
+#include <memory>
+
+#include "timer.h"
+#include "font-problem.h"
+
+#include "image.h"
+#include "lines.h"
+#include "base/stringprintf.h"
+
+#include "network.h"
+#include "threadutil.h"
+
+using namespace std;
+
+using uint8 = uint8_t;
+using uint32 = uint32_t;
+using int64 = int64_t;
+
+
+// XXX somehow needs to be shared?
+static constexpr int ROW0_MAX_PTS = 38;
+static constexpr int ROW1_MAX_PTS = 14;
+static constexpr int ROW2_MAX_PTS = 10;
+constexpr std::initializer_list<int> row_max_points = {
+  ROW0_MAX_PTS,
+  ROW1_MAX_PTS,
+  ROW2_MAX_PTS,
+};
+
+
+struct Op {
+  char input_char;
+  char output_char;
+  Op(char i, char o) : input_char(i), output_char(o) {}
+};
+
+struct Config {
+  string input_font;
+
+  string font_name = "Heavenica";
+  string filename = "heavenica.sfd";
+  float extra_scale = 1.0f;
+  float linegap = 0.0f;
+  float blank_width = 1.0f;
+
+  vector<Op> letters;
+};
+
+static bool IsLetter(char c) {
+  return (c >= 'a' && c <= 'z') || (c >= 'A' && c <= 'Z');
+}
+
+// Lowerercase Franklin Gothic
+Config Franklin() {
+  Config cfg;
+  cfg.input_font = "franklin.ttf";
+  cfg.font_name = "Franklin TODO CLEVER NAME";
+  cfg.filename = "franklow.sfd";
+  /*
+  cfg.extra_scale = 1.2f;
+  cfg.linegap = -0.1f;
+  */
+  cfg.blank_width = 0.75f;
+
+  for (int i = 32; i < 127; i++) {
+    cfg.letters.emplace_back(i, i);
+  }
+  return cfg;
+}
+
+// Lowerercase Franklin Gothic
+Config Futurda() {
+  Config cfg;
+  cfg.input_font = "futura.ttf";
+  cfg.font_name = "Futurda";
+  cfg.filename = "futurda.sfd";
+  /*
+  cfg.extra_scale = 1.2f;
+  cfg.linegap = -0.1f;
+  */
+  cfg.blank_width = 0.75f;
+
[…]

### 2021-03-12 lowercase/hallucinate.cc
Hallucinate uses cache (huge speedup) and outputs 64 bit mask in bitmap mode. Generate (traced) bitmap SFDs. Little bit in the paper

@@ -8,6 +8,7 @@
 #include <algorithm>
 #include <cstdint>
 #include <memory>
+#include <unordered_map>
 
 #include "timer.h"
 #include "font-problem.h"
@@ -25,6 +26,8 @@ using namespace std;
 using uint8 = uint8_t;
 using uint32 = uint32_t;
 using int64 = int64_t;
+
+using Image8x8 = FontProblem::Image8x8;
 
 // If true, find the input that maximizes the prediction
 // for the target letter, and minimizes others.
@@ -623,8 +626,8 @@ static ImageA DrawSDFShapes(const Network &net,
 }
 
 // Generate random 8x8 bitmap images.
-static ImageA Random8x8(const Network &prednet,
-                        int target_letter) {
+static std::pair<FontProblem::Image8x8, ImageA>
+Random8x8(const Network &prednet, int target_letter) {
   constexpr int N = 8 * 8;
 
   auto Make8x8 = [](const std::vector<double> &inputs) {
@@ -640,13 +643,27 @@ static ImageA Random8x8(const Network &prednet,
   int num_calls = 0;
   double sdf_time = 0.0;
   double run_time = 0.0;
+
+  int64 cache_hits = 0;
+  std::unordered_map<uint64, double> cache;
+  // TODO: Cache result, since optimizer doesn't know that these
+  // are rounded to bools?
   std::function<double(const std::vector<double> &inputs)> IsLetter =
     [&prednet, target_letter, &Make8x8, &num_calls,
-     &sdf_time, &run_time](
+     &sdf_time, &run_time, &cache_hits, &cache](
         const std::vector<double> &inputs) -> double {
-
+      num_calls++;
+      
       Timer sdf_timer;
       FontProblem::Image8x8 img8x8 = Make8x8(inputs);
+      {
+        auto it = cache.find(img8x8.To64());
+        if (it != cache.end()) {
+          cache_hits++;
+          return it->second;
+        }
+      }
+
       ImageA sdf = FontProblem::SDF36From8x8(img8x8);
       sdf_time += sdf_timer.MS();
       Timer run_timer;
@@ -671,11 +688,12 @@ static ImageA Random8x8(const Network &prednet,
       int edges = img8x8.Edges();
       penalty += edges / 100.0;
 
-      num_calls++;
       if (num_calls % 1000 == 0)
-        printf("[%c] %d calls, %d pixels, %d edges, penalty: %.4f\n",
[…]

### 2021-03-12 trunk/lowercase/paper/lowercase.tex
Hallucinate uses cache (huge speedup) and outputs 64 bit mask in bitmap mode. Generate (traced) bitmap SFDs. Little bit in the paper

@@
 \newcommand\at{\ensuremath{\scriptstyle @}}
+
+\renewcommand\paragraph[1]{\smallskip{\bf #1}\enspace}
 
@@
 
-We want to machine-learn two functions, \makelowercase\ and
+We want to machine-learn~\cite{neuralnetwork} two functions, \makelowercase\ and
 \makeuppercase. Each takes a letterform and returns a letterform (we
@@
 
-Sprechen of fonts, I downloaded every font I could find on the whole
+Sprechen of Fonts, I downloaded every font I could find on the whole
 internet. This was overkill. The resulting directory tree contained
@@
 \includegraphics[width=0.9 \linewidth]{casealignment}
-\caption{ Example alignment to reject the font
-  {\tt DrippingGooExtended}.
-  At left, \letterform{A} (red) and
+\caption{ Example alignment to reject the font {\tt
+    DrippingGooExtended}. At left, \letterform{A} (red) and
   \letterform{a} (green) rendered with the identity transform, and
@@
   $1.7\%$ difference. Note that the shapes are still not an exact
-  match (the export process has to round the data to integers and
-  might apply other non-linear transformations like curve
-  simplification), but these are clearly not a useful pair for
-  the current problem.} \label{fig:casealignment}
+  match (probably errors introduced in the font export process, which
+  has to round the data to integers and might apply other non-linear
+  transformations like curve simplification), but these are clearly
+  not a useful pair for the current
+  problem.} \label{fig:casealignment}
 \end{figure}
@@
 
-\section{The simplest thing that might work}
+\section{The simplest thing that might work} \label{sec:vectorversion}
 
 Before getting fancy (which we we will) it's good engineering hygiene
-to try the simpliest thing that might just work (it doesn't). Fonts
+to try the simplest thing that might just work (it doesn't). Fonts
 are represented as vector data (lines and quadratic B\'ezier curves).
@@
 
-% XXX total number of parameters?
 The network I trained has an input layer size of $154 = (38 + 14 + 10)
@@
 The output layer is the same size, plus 26 (see below). There are
-three hidden layers of size 308, 308, 360. The first two layers are
-dense and the third sparse (about half of the weights are zero). All
-layers are leaky rectified linear units. If you're taking notes,
-don't, as again this does not work well, and I don't know how people
-figure out what the right network size is anyway. I just made it up.
-You can give {\em me} your notes.
+three hidden layers of size 308, 308, 360. The first layer is dense
+and the remainder are sparse,
+%(two thirds to a half of the weights are zero),
+for just about 1 million total parameters. All layers are leaky
+rectified linear units. If you're taking notes, don't, as again this
+does not work well, and I don't know how people figure out what the
+right network size is anyway. I just made it up. You can give {\em me}
+your notes.
+
+\paragraph{Bonus outputs.} The output includes the predicted shape,
+and also 26 individual predictors for each of the 26 letters. So a
+training example is actually like \letterform{C} $\rightarrow$
+\letterform{c} $[0, 0, 1, 0, 0, \ldots, 0]$, with the 1 in the third
+place because C is the third letter. We don't need these outputs for
+the problem itself (e.g. to lowercase new letter shapes), but there
+are several ideas behind this. First, the lowercasing function we're
+trying to learn does take into account the letter of the alphabet
+being lowercased. By asking the network to learn this as well (it is
+penalized when it gets the prediction wrong), it must learn features
+that allow it to distinguish different letters, and those features are
+available for use by outputs we {\em do} care about. This is an easy
+way to coax it to learn features that I know are meaningful without
+having to actually engineer feature extractors by hand (or first train
+separate models, etc.). Perhaps the most useful thing is that it's
+very clear what the right answer is, so it gives me an easy way to see
+if the network is learning anything {\em at all}. (It does.) Finally,
+we can do some silly stuff with these; see
+Section~\ref{sec:hallucination}.
 
@@
 
-% TODO: Screenshot of this thing's output.
-% TODO: Font file generated from this.
-% TODO: Fitting the prediction to the actual in order to coax
-% better results, failing.
+\begin{figure}[ht]
+\includegraphics[width=0.9 \linewidth]{trainingvector}
+\caption{ Screenshot of {\tt train.exe} running on the vector-based
+  version of the problem. Shown is the \makelowercase\ model's output
+  (bottom) on four shapes from four fonts (top). Some dust in between
+  is the activation of the network's layers. At the very bottom, the
+  26 predictions for ``what letter is this?''. The output for
+  \letterform{j} is not too bad; you can see the distinct dot (a
+  separate contour) and it sort of looks like a \letterform{j}. The
+  \letterform{e} also has two pieces as expected but is otherwise
+  garbage. The model is unsure whether the second input is an H
+  or a K, and has predicted a shape sort of ambiguously between those
+  two. The \lowercase{z} is also an embarrassment.
+} \label{fig:trainingvector}
+\end{figure}
+
+I was so {\tiny confident} that this wouldn't work that I only trained
+a \makelowercase\ model and didn't even worry about the more
+complicated simultaneous training setup yet. I ran this for about
+22,000 rounds, some 90 million training examples. Indeed it does not
+work (Figure~\ref{fig:trainingvector}). It is not a total catastrophe,
+though. We can tell from the 26 bonus outputs that the model can
+clearly recognize letters (though perhaps just by memorization). Some
+of the shapes it generates are along the right lines. ({\em Along the
+  right lines}, get it??) I did not feel ANGRY at these results because
+I expected it to not really work.
+
+\paragraph{Just try making it more complicated!} This
+problem of predicting the vector shape directly is a lot to ask of a
+neural network, at least set up this way. One thing that did not sit
+well with me is that the network could generate a perfect-looking
+result, but because it didn't have the points in the expected order,
+it would be penalized. This makes it harder to learn. True, but
+my reflex to make things more complicated XXX
+
+Ideally we would compute the difference between the predicted shape
+and the expected shape, regardless of how they're drawn. Aside from
+being a bit computationally challenging, this won't really work
+because we need to attribute error directly to each output in order to
+actually update the model in training. So I spent an afternoon writing
+a routine to compute the best alignment of points between the actual
+and expected outputs (Figure~\ref{fig:rotatej}). Aside from being
+harder than it looked, my alignment code ended up being pretty slow
+relative to the rest of training, even worse since it ran on the CPU
+instead of GPU, and the training speed became UPSETTINGLY SLOW. So, I
+discarded this excursion and threw this whole endeavour in the trash!
+But I must have confused the Recycle~Bin icon with the fairly
+complicated export-to-Truetype Font process that I built, because I
+ran the model on the venerable Futura~\cite{futura} font and generated
+{\bf Futurda} (Figure~\ref{fig:futurda}), which can be downloaded from my
+website and installed for your desktop publishing needs.\footnote{Font
+  downloads are available at \url{http://tom7.org/lowercase/}.}
+
+% TODO: actually upload the fonts and make sure the URL is what you want
+
+\begin{figure}[ht]
+\includegraphics[width=0.5 \linewidth]{rotatej}
+\caption{ Example alignment (from debug UI) of predicted shape (blue)
+  to expected shape (green). We require each point to be mapped (red)
+  to a point from the expected contour in a monotonic order (but
+  several can be mapped to the same one), so that we can attribute
+  error to each point. I hope it was fun to write this code because
+  it was too slow to do pretty much anything but generate this figure!
+} \label{fig:rotatej}
+\end{figure}
+
+% XXX add downloadable font too
+\begin{figure*}[tp]
+\includegraphics[width=0.9 \linewidth]{futurda}
+\caption{ Type specimen for the generated font {\bf Futurda}. This is the
+  classic font Futura, run through the vector based lowercasing model
+  (Section~\ref{sec:vectorversion}). At top are Futura's letterforms
+  \letterform{A}--\letterform{Z} run through the model and so ``made
+  lowercase'' (it's obviously garbage). Next are
+  \letterform{a}--\letterform{z},
+  made even more lowercase. Also rubbish.
+  At the bottom are the pangrams ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
+  and ``Sphinx of black quartz, judge my vow!'' Although the text is
+  illegible and kind of uninteresting, it does have a certain dynamic
+  Rorschach aesthetic, like a collection of delicate moths pinned
+  to paperboard, that one could consider framing or publishing in
+  the proceedings of SIGBOVIK 2021.
+} \label{fig:futurda}
+\end{figure*}
+
+% lowerercase Franklin Gothic. Franklin mint... not guaranteed
+% to go up in value
+
+% obviously, need to express different cases of anger at various
+% problems or successes?
 
 % the sdf problem
+
+\section{SDFs}
+
+Don't give up! The fixed-size input/output of neural networks is
+better suited to something like an array of pixels, and fonts can of
+course be represented this way as well. To stay in the realm of what
+my desktop computer with a single GeForce 1080 can do, I wanted to
+keep the number of inputs and outputs pretty small. There's already an
+excellent technique for representing font data as compact bitmaps,
+which comes from computer graphics, called Signed Distance Fields
+(SDFs)~\cite{green2007improved}. In a standard rasterization of a
+font, each pixel of a bitmap contains 1 or 0, or perhaps an
+anti-aliased value in-between. In an SDF, the bitmap instead contains
+the distance to the nearest edge, and is signed (e.g. values inside
+the shape are $> 0$, value outside are $< 0$). Actually in
+practice we offset and saturate the values so that they are all in
+$[0,1]$ (or bytes in $[0,255]$), with some nonzero ``on-edge value''
+(say, 0.5) standing for ``distance 0''. In order to display the font
+at the size of your choice, you then resize the SDF image with bilinear
+interpolation, and then simply threshold the image. This works
+surprisingly well (Figure~\ref{fig:sdf}).
+
+SDFs seem well-suited for machine learning. They contain more
+information per pixel than a plain bitmap, so we can use a smaller
+input and output size. On the input
+side, extremal pixels that would almost never be set in a bitmap still
+have significant information (distance to the character). The error
+function is just pixel-by-pixel difference. The rendering of the
+output is inherently tolerant of some noise because of the sampling
+and thresholding. So, this seemed like it might work well! (It doesn't
+work that well.)
+
+I got some stats on the font database, and determined the following
+parameters for the SDFs.
+ (here, $36\times36$ pixel images)
+% Figure: bitmap vs SDF
 
@@
 
-% Perfect letters, hallucinated
+\section{Perfect letters, hallucinated}
+\label{sec:hallucination}
 

### 2021-03-13 lowercase/font-problem.h
Just realized that I was writing about an older and much worse version of the vector net, ugh

@@ -170,14 +170,16 @@ struct FontProblem {
   };
 
   // Specific to the 36x36 bitmap problem I mostly worked with;
-  // tuned for speed. The 8x8 bitmap is placed at 4,6, which
-  // empirically seems to be the "right" place for it, for uppercase
-  // letterforms at least.
+  // tuned for speed.
+  // Uppercase letters place the 8x8 bitmap at 4,6, which
+  // empirically seems to be the "right" place for it.
+  // Lowercase letters go at 4,9.
   //
   // Note that these SDFs agree with SDFFromBitmap but are not
   // really right (they are too "low res") because that routine
   // assumes the bitmap is at least as big as the output sdf.
-  static ImageA SDF36From8x8(Image8x8 bits);
+  static ImageA SDF36From8x8Uppercase(Image8x8 bits);
+  static ImageA SDF36From8x8Lowercase(Image8x8 bits);  
 
   // Deprecated!
   static ImageA SDFThresholdAA(uint8_t onedge_value,

### 2021-03-13 lowercase/sdf-figure.cc

@@ -1,0 +1,132 @@
+
+#include <vector>
+#include <string>
+#include <algorithm>
+#include <cstdint>
+#include <memory>
+
+#include "timer.h"
+#include "font-problem.h"
+#include "network.h"
+
+#include "image.h"
+
+using namespace std;
+
+using uint8 = uint8_t;
+using uint32 = uint32_t;
+using int64 = int64_t;
+
+#if 0
+static void Gen(const FontProblem::SDFConfig &config,
+                const Network &make_lowercase,
+                const Network &make_uppercase,
+                const ImageA &sdf,
+                const string &filename) {
+  const int SCALE = 5;
+  const int QUALITY = 4;
+
+  ImageRGBA islands;
+  (void)FontProblem::VectorizeSDF(config, sdf, &islands);
+  islands = islands.ScaleBy(4);
+
+  FontProblem::Gen5Result gen5result =
+    FontProblem::Gen5(config, make_lowercase, make_uppercase, sdf);
+
+  FontProblem::Gen5ImagesResult result =
+    FontProblem::Gen5Images(gen5result, SCALE, QUALITY);
+
+  // All should be the same size.
+  const int TILEW = result.input.Width();
+  const int IMGH = result.input.Height();
+  // Room for predictors at bottom.
+  const int TILEH = IMGH + 24;
+
+  ImageRGBA out(TILEW * 3, 3 * TILEH);
+  out.Clear32(0x000000FF);
+
+  out.BlendImage(TILEW, TILEH, result.input);
+  out.BlendImage(0, TILEH, result.low);
+  out.BlendImage(0, 0, result.low_up);
+  out.BlendImage(TILEW * 2, TILEH, result.up);
+  out.BlendImage(TILEW * 2, TILEH * 2, result.up_low);
+
+  out.BlendImage(TILEW + 4, 4, islands);
+
+  auto DrawPreds = [&out](int startx, int starty,
+                          const std::array<float, 26> &preds) {
+      for (int i = 0; i < 26; i++) {
+        const uint8 v = FontProblem::FloatByte(preds[i]);
+        string s = "A";
+        s[0] += i;
+        out.BlendText(startx + 9 * (i % 13),
+                      starty + (i / 13) * 10,
+                      v, v, 0xFF, 0xFF,
[…]

### 2021-03-13 trunk/lowercase/paper/lowercase.tex
Just realized that I was writing about an older and much worse version of the vector net, ugh

@@
 
-\documentclass[twocolumn]{amsart}
+\documentclass[twocolumn]{article} % {amsart}
 \usepackage[top=0.75in, left=0.65in, right=0.65in, bottom=0.6in]{geometry}
 
+\usepackage{float}
 \usepackage{url}
@@
 
-%     leftrule=0.2pt,rightrule=0.2pt,toprule=0.2pt,bottomrule=0.2pt,
-%     enhanced jigsaw,
-%    borderline horizontal={0.2pt}{0pt}{dashed},
-%    borderline vertical={0.2pt}{0pt}{dashed}
-
-We want to machine-learn~\cite{neuralnetwork} two functions, \makelowercase\ and
-\makeuppercase. Each takes a letterform and returns a letterform (we
-can choose how these are represented) and does the needful, e.g.
-\makelowercase(\letterform{A}) should return \letterform{a}. In order to learn this function, we'll at least need a lot of
-examples to use as training data. A training example for
+We want to machine-learn~\cite{neuralnetwork} two functions,
+\makelowercase\ and \makeuppercase. Each takes a letterform and
+returns a letterform (we can choose how these are represented) and
+does the needful, e.g. \makelowercase(\letterform{A}) should return
+\letterform{a}. In order to learn this function, we'll at least need a
+lot of examples to use as training data. A training example for
 \makelowercase\ is a letterform and its expected corresponding
@@
 \begin{figure}[ht]
+\begin{centering}
   \includegraphics[width=0.5 \linewidth]{rotatej}
+\end{centering}
 \caption{ Example alignment (from debug UI) of predicted shape (blue)
@@
 
+% FIXME: Wait I had much better screenshots, see video/lowercase
+
 % XXX add downloadable font too
 \begin{figure*}[tp]
+\begin{centering}
   \includegraphics[width=0.9 \linewidth]{futurda}
+\end{centering}
 \caption{ Type specimen for the generated font {\bf Futurda}. This is the
@@
 
+\begin{figure}[ht]
+  \includegraphics[width=0.95 \linewidth]{sdf-figure}
+  \caption{
+    The signed distance function representation of a letterform.
+    At the very left, a $36\times 36$ pixel rasterization of
+    the character without anti-aliasing, for comparison. Any
+    scaling of this will have chunky pixel artifacts. Next,
+    a $36\times 36$ pixel SDF of same. Third, simply scaling
+    that $36\times 36$ image to $180 \times 180$ pixels with
+    bilinear sampling. Finally, that image thresholded to
+    produce a $180 \times 180$ pixel rasterization, which is
+    far superior despite being derived from a $36 \times 36$ pixel
+    bitmap. Typically this process is performed at an even higher
+    scale and then downsampled to produce an anti-aliased image.
+    } \label{fig:sdf}
+\end{figure}
+
 SDFs seem well-suited for machine learning. They contain more
 information per pixel than a plain bitmap, so we can use a smaller
-input and output size. On the input
-side, extremal pixels that would almost never be set in a bitmap still
-have significant information (distance to the character). The error
-function is just pixel-by-pixel difference. The rendering of the
-output is inherently tolerant of some noise because of the sampling
-and thresholding. So, this seemed like it might work well! (It doesn't
-work that well.)
+input and output size. On the input side, extremal pixels that would
+almost never be set in a bitmap still have significant information
+(distance to the character). The error function is just pixel-by-pixel
+difference. The rendering of the output is inherently tolerant of some
+noise because of the sampling and thresholding. So, this seemed like
+it might work really well! (It doesn't work that well.)
 
-I got some stats on the font database, and determined the following
-parameters for the SDFs.
- (here, $36\times36$ pixel images)
-% Figure: bitmap vs SDF
+I computed some stats on the font database, and determined the
+following parameters for the fixed-size SDFs we train on. The images
+are $36 \times 36$ pixels. The character box is placed such that there
+are $2$ pixels of top padding, and $9$ pixels of left and bottom
+padding. The character box is only ``nominal'' in the sense that the
+font's contours can exceed its bounds, and this is completely normal
+for a letter like \letterform{j} (which goes below the baseline and
+often hangs to the left of the origin as well). I used an ``on-edge
+value'' of 220 (because much more of the SDF is outside the letter
+than inside) and the distance is scaled as 15 units per pixel (so that
+pixels on the outer edge often have non-zero values). Compared to the
+first version, I was somewhat more permissive in what fonts I trained
+on, since there was no inherent limit to the number of contours their
+complexity. I did exclude fonts whose rasterizations exceeded the
+bounds of the SDF, which is possible (very wide \letterform{W} or
+low-descending \letterform{j} perhaps) but rare.
 
@@
 
+Having committed to the representation, again it is ``just'' a matter
+of heating up the GPU to apply some linear and non-linear transforms.
+The initial network had an input size of $36 \times 36 = 1296$ for the
+SDF, and the output the same plus $26$ bonus outputs (one for each
+letter, as before). I started with three hidden layers of $1296$, $1296$,
+and $2916$ nodes, each sparse ($80\%$ of the weights are zero). Again,
+don't take notes. This one works a bit better than before, but still
+not impressive. The node references are assigned spatially (something
+like the $20\%$ of the nodes on the previous layer that are closest to
+the next layer's node) but due to a bug the spatial locality is
+actually pretty strange.
+% XXX if I show layer weights, forward-reference them here:
+% It's responsible for the XXX
+Every layer's transfer function is ``leaky relu'' ({\tt x > 0 \? x : 0.1 * x})
+which is fast to compute and better than sigmoids in the output since
+correct values will not just be 0 and 1. It would definitely make sense
+to use convolutional layers for this problem, as features like serifs,
+lines, curves, and so on could appear throughout the input and output.
+I just haven't built support for that in my weird home-grown software,
+yet.
+
+I also adapted my weird home-grown software to train the
+\makeuppercase\ and \makelowercase\ models simultanously. Two models
+fit comfortably in GPU memory, with plenty of space for a stream of
+training data (one training instance is only about 10kb). The only
+challenging thing is arranging for them to feed each other generated
+``inversion'' examples (Figure~\ref{fig:cotraining}), but this is just
+a matter of programming, thank God. I should remember to do projects
+that are mostly a matter of programming. Each round, $25\%$ of the
+batch consists of inverted examples from the symmetric model's output
+from a recent round. I make sure that one model is not allowed to
+get too far ahead (more than 2 rounds) of the other in training,
+because I want this feedback loop to be somewhat tight.
+
+So I did that and let it run for a month. Actually I had to start over
+several times with different parameters and initialization weights
+because it would get stuck
+
+% you propagate my back, I'll propagate yours
+
 \label{sec:fiddly}
+
+% TODO loss over time
+
+% some other things that are worth a mention:
+%   autoparallel
+%   efficiency improvements (restrict, compile-time constants)
 
@@
 
-\section{Perfect letters, hallucinated}
-\label{sec:hallucination}
+\section{Perfect letters, hallucinated} \label{sec:hallucination}
+
+lorem ipsum (Figure~\ref{fig:perfect})
+
+\begin{figure*}[tp]
+\begin{centering}
+  \includegraphics[width=0.9 \linewidth]{perfecthallucination}
+\end{centering}
+\caption{
+%  XXX Why doesn't this caption come out the way I'm expecting?
+%  (in amsart style... can't have two figure*s on the same page?)
+   Type specimen for the generated font {\bf Perfect Hallucination}.
+  TODO write a description!
+} \label{fig:perfect}
+\end{figure*}
+
+% LSDF
+
+dolor sit amet
+
+\section{Chess-playing}
 
@@
 
+\section{Acknowledgements}
+
+For this project I used \verb+stb_truetype.h+ from the excellent stb
+collection of single-file libraries.~\cite{stb} I did push its limits
+somewhat and ``automatically discovered'' assertion failures and other
+crashes (e.g. during blackbox optimization on all 100k fonts), but it
+saved a lot of time. This library only helps with reading the fonts
+and generating SDFs. To generate TTFs, I had to write my own pipeline,
+which generates FontForge's .SFD (a typo factory right there) files,
+and then did the final export with FontForge~\cite{fontforge}. Thanks
+Jason Reed for this suggestion.
+
 \nocite{murphy2019blind}
 \nocite{murphy2019eloworld}
+\nocite{stb}
 

### 2021-03-13 trunk/lowercase/font-problem.cc

@@ -1121,8 +1121,8 @@ ImageA FontProblem::SDFFromBitmap(const SDFConfig &config,
   return sdf;
 }
 
-
-ImageA FontProblem::SDF36From8x8(Image8x8 img) {
+template<int TOP>
+static ImageA SDF36From8x8(FontProblem::Image8x8 img) {
   constexpr int sdf_size = 36;
   // nominal size of image.
   constexpr int size = 18;
@@ -1132,7 +1132,6 @@ ImageA FontProblem::SDF36From8x8(Image8x8 img) {
   constexpr float scale = size / (float)sdf_size;
 
   constexpr int LEFT = 4;
-  constexpr int TOP = 6;
   constexpr int WIDTH = 8;
   constexpr int HEIGHT = 8;
   if (false)
@@ -1269,7 +1268,13 @@ ImageA FontProblem::SDF36From8x8(Image8x8 img) {
   return sdf;
 }
 
+ImageA FontProblem::SDF36From8x8Uppercase(Image8x8 img) {
+  return SDF36From8x8<6>(img);
+}
 
+ImageA FontProblem::SDF36From8x8Lowercase(Image8x8 img) {
+  return SDF36From8x8<7>(img);
+}
 
 std::pair<ImageA, std::array<float, 26>>
 FontProblem::RunSDFModel(const Network &net,

### 2021-03-13 trunk/lowercase/hallucinate.cc

@@ -35,6 +35,9 @@ using Image8x8 = FontProblem::Image8x8;
 // to the vector [0, 0, ... 0, 1, 0, ... 0].
 static constexpr bool FIND_MAX = false;
 
+// Generating perfect uppercase letters, or lowercase?
+static constexpr bool FIND_UPPERCASE = false;
+
 
 static constexpr FontProblem::SDFConfig SDF_CONFIG = {};
 
@@ -645,9 +648,10 @@ Random8x8(const Network &prednet, int target_letter) {
   double run_time = 0.0;
 
   int64 cache_hits = 0;
+  // Cache results, since optimizer doesn't know that these doubles
+  // are discretized to bools. Helps a lot!
   std::unordered_map<uint64, double> cache;
-  // TODO: Cache result, since optimizer doesn't know that these
-  // are rounded to bools?
+  
   std::function<double(const std::vector<double> &inputs)> IsLetter =
     [&prednet, target_letter, &Make8x8, &num_calls,
      &sdf_time, &run_time, &cache_hits, &cache](
@@ -664,7 +668,10 @@ Random8x8(const Network &prednet, int target_letter) {
         }
       }
 
-      ImageA sdf = FontProblem::SDF36From8x8(img8x8);
+      ImageA sdf =
+        FIND_UPPERCASE ? FontProblem::SDF36From8x8Uppercase(img8x8) :
+        FontProblem::SDF36From8x8Lowercase(img8x8);
+      
       sdf_time += sdf_timer.MS();
       Timer run_timer;
       const auto pred =
@@ -677,11 +684,11 @@ Random8x8(const Network &prednet, int target_letter) {
                         return pred[i];
                       });
 
-      // Target about 50% of pixels set.
-      // (My handmade A had 33. Maybe should be fewer
-      // for lowercase?)
+      // Target about 50% of pixels set for uppercase (my handmade A had 33)
+      // and 37.5% for lowercase.
       int pixels = img8x8.PixelsOn();
-      int pixel_cost = abs(pixels - 32);
+      constexpr int target_pixels = FIND_UPPERCASE ? 32 : 24;
+      int pixel_cost = abs(pixels - target_pixels);
       penalty += pixel_cost / 50.0;
 
       // Provide a small penalty for noisy images.
@@ -716,7 +723,10 @@ Random8x8(const Network &prednet, int target_letter) {
          sdf_time / num_calls, run_time / num_calls);
   CHECK(best.size() == N);
[…]

### 2021-03-13 lowercase/paper/lowercase.tex
Fixed various bugs in my first attempt to use the old vector models. Generate new, much better fonts from the second version (duhhh). Update paper to tell this story

@@
 
-\documentclass[twocolumn]{amsart}
+\documentclass[twocolumn]{article} % {amsart}
 \usepackage[top=0.75in, left=0.65in, right=0.65in, bottom=0.6in]{geometry}
 
-\usepackage{url}
+\usepackage{float}
+
+% \usepackage{url}
+% \usepackage{xurl}
+\usepackage[pdftex]{hyperref}
+
 % \usepackage{code}
@@
 
-%     leftrule=0.2pt,rightrule=0.2pt,toprule=0.2pt,bottomrule=0.2pt,
-%     enhanced jigsaw,
-%    borderline horizontal={0.2pt}{0pt}{dashed},
-%    borderline vertical={0.2pt}{0pt}{dashed}
-
-We want to machine-learn two functions, \makelowercase\ and
-\makeuppercase. Each takes a letterform and returns a letterform (we
-can choose how these are represented) and does the needful, e.g.
-\makelowercase(\letterform{A}) should return \letterform{a}. In order to learn this function, we'll at least need a lot of
-examples to use as training data. A training example for
+We want to machine-learn~\cite{neuralnetwork} two functions,
+\makelowercase\ and \makeuppercase. Each takes a letterform and
+returns a letterform (we can choose how these are represented) and
+does the needful, e.g. \makelowercase(\letterform{A}) should return
+\letterform{a}. In order to learn this function, we'll at least need a
+lot of examples to use as training data. A training example for
 \makelowercase\ is a letterform and its expected corresponding
@@
 \begin{figure*}[ht]
+\centering
   \includegraphics[width=0.9 \linewidth]{sortition}
@@
 \begin{figure}[ht]
+\centering
   \includegraphics[width=0.9 \linewidth]{casealignment}
-\caption{ Example alignment to reject the font
-  {\tt DrippingGooExtended}.
-  At left, \letterform{A} (red) and
+\caption{ Example alignment to reject the font {\tt
+    DrippingGooExtended}. At left, \letterform{A} (red) and
   \letterform{a} (green) rendered with the identity transform, and
@@
 \begin{figure}[ht]
+\centering
   \includegraphics[width=0.9 \linewidth]{samecasepr}
@@
 Before getting fancy (which we we will) it's good engineering hygiene
-to try the simpliest thing that might just work (it doesn't). Fonts
+to try the simplest thing that might just work (it doesn't). Fonts
 are represented as vector data (lines and quadratic B\'ezier curves).
@@
 most clean fonts had three or fewer contours, and when sorting them by
-descending length, most fit within 38, 14, and 10 endpoints for the
-three. So, I only train on fonts where all of the letters fit within
-this budget.\footnote{It would not be a good idea to reject only the letters
-that don't fit, because it might result in the network being trained
-on more \letterform{l}s (tends to be simple) than \letterform{g}s
-(tends to be complex).}
+descending length, basically all of them fit within 100, 25, and 16
+endpoints for the three. So, I only train on fonts where all of the
+letters fit within this budget.\footnote{It would not be a good idea
+  to reject only the letters that don't fit, because it might result
+  in the network being trained on more \letterform{l}s (tends to be
+  simple) than \letterform{g}s (tends to be complex).}
 
@@
 function (which might not be the case if we were predicting a dynamic
-number of points). Finally, I output the contour's points starting
-from the point closest to the origin in order to reduce one needless
-degree of freedom.
+number of points).
 
-The network I trained has an input layer size of $154 = (38 + 14 + 10)
+The network I trained has an input layer size of $570 = (100 + 25 + 16)
 \times 4 + 3 \times 2$ (one control point and one end point per
@@
 three hidden layers of size 308, 308, 360. The first layer is dense
-and the remainder are sparse (two thirds to a half of the weights are
-zero), for just about 1 million total parameters. All layers are leaky
-rectified linear units. If you're taking notes, don't, as again this
+and the remainder are sparse,
+%(two thirds to a half of the weights are zero),
+for just about 1 million total parameters. All layers are leaky
+rectified linear units ({\tt x > 0 ? x : 0.1 * x}), which is fast to
+compute and better than sigmoids in the output since correct values
+will not just be 0 and 1. If you're taking notes, don't, as again this
 does not work well, and I don't know how people figure out what the
@@
 
-\paragraph{Bonus outputs}. The output includes the predicted shape,
+\paragraph{Bonus outputs.} The output includes the predicted shape,
 and also 26 individual predictors for each of the 26 letters. So a
@@
 are several ideas behind this. First, the lowercasing function we're
-trying to learn does take into account the letter of the alphabet
-being lowercased. By asking the network to learn this as well (it is
-penalized when it gets the prediction wrong), it must learn features
-that allow it to distinguish different letters, and those features are
-available for use by outputs we {\em do} care about. This is an easy
-way to coax it to learn features that I know are meaningful without
-having to actually engineer feature extractors by hand (or first train
-separate models, etc.). Perhaps the most useful thing is that it's
-very clear what the right answer is, so it gives me an easy way to see
-if the network is learning anything {\em at all}. (It does.) Finally,
-we can do some silly stuff with these; see
-Section~\ref{sec:hallucination}.
+trying to learn does depend on the letter of the alphabet being
+lowercased (in an extreme case, consider the lowercase-L
+\letterform{l} and the uppercase-i \letterform{I}, which look the same
+in many fonts but have different lowercase letterforms). By asking the
+network to learn this as well (it is penalized when it gets the
+prediction wrong), it must learn features that allow it to distinguish
+different letters, and those features are available for use by outputs
+we {\em do} care about. This is an easy way to coax it to learn
+features that I know are meaningful without having to actually
+engineer feature extractors by hand (or first train separate models,
+etc.). Similarly, I could have asked it to predict whether the font
+is italic, serif, the character's width, or anything else I have on
+hand. Perhaps the most useful thing is that it's very clear what the
+right answer is, so it gives me an easy way to see if the network is
+learning anything {\em at all}. (It does.) Finally, we can do some
+silly stuff with these; see Section~\ref{sec:hallucination}.
 
@@
 \begin{figure}[ht]
+\centering
   \includegraphics[width=0.9 \linewidth]{trainingvector}
-\caption{ Screenshot of {\tt train.exe} running on the vector-based
+\caption{ Screenshot of {\tt train.exe} running on the first vector-based
   version of the problem. Shown is the \makelowercase\ model's output
@@
 work (Figure~\ref{fig:trainingvector}). It is not a total catastrophe,
-though. The model can clearly recognize letters (though perhaps just by
-memorization) and some of the shapes it generates are along the right
-lines. ({\em Along the right lines}, get it??)
+though. We can tell from the 26 bonus outputs that the model can
+clearly recognize letters (though perhaps just by memorization). Some
+of the shapes it generates are along the right lines. ({\em Along the
+  right lines}, get it??) I did not feel ANGRY at these results
+because I expected it to not really work. Still, it ``has output'' and
+so it can be used to generate a font. I made every glyph in Comic Sans
+MS~\cite{comicsans} lowercase using the model (with the exception of the
+\letterform{\%} character, which has too many contours---{\it five}!).
+Mostly this model produces small, non-confident scrawls, like little
+grains of sand, so this font is called {\bf Comic Sands}
+(Figure~\ref{fig:comicsands}). The TrueType version can be downloaded
+from my website and installed for your desktop publishing
+needs.\footnote{Font downloads are available at
+  \url{http://tom7.org/lowercase/}.}
 
-It's a bit hard on the
-
-% TODO: Font file generated from this.
-% TODO: Fitting the prediction to the actual in order to coax
-% better results, failing.
-
-% XXX cite futura
-% XXX add downloadable font too
 \begin{figure*}[tp]
-\includegraphics[width=0.9 \linewidth]{futurda}
-\caption{ Type specimen for the generated font {\bf Futurda}. This is the
-  classic font Futura, run through the vector based lowercasing model
-  (Section~\ref{sec:vectorversion}). At top are Futura's letterforms
+\centering
+  \includegraphics[width=0.95 \linewidth]{comicsands}
+\caption{
+  Type specimen for the generated font {\bf Comic Sands}. This is
+  the hateful Comic Sans MS run through an early vector-based lowercasing
+  model (Section~\ref{sec:vectorversion}). At top are Comic Sans's letterforms
   \letterform{A}--\letterform{Z} run through the model and so ``made
-  lowercase'' (it's obviously garbage). Next are the \letterform{a}--\letterform{z},
+  lowercase'' (it's obviously garbage). Next are
+  \letterform{a}--\letterform{z},
   made even more lowercase. Also rubbish.
-  At the bottom are the pangrams ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
-  and ``Sphinx of black quartz, judge my vow!'' Although the text is
-  illegible and kind of uninteresting, it does have a certain dynamic
+  At the bottom are the illegible pangrams
+  ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
+  and ``Sphinx of black quartz, judge my vow!'' Although the output
+  barely resembles letters, it does have a certain dynamic
   Rorschach aesthetic, like a collection of delicate moths pinned
   to paperboard, that one could consider framing or publishing in
-  the proceedings of SIGBOVIK 2021.
+  the proceedings of SIGBOVIK 2021. It is certainly an improvement on
+  the original font.
+} \label{fig:comicsands}
+\end{figure*}
+
+\subsection{Just try making it more complicated!} \label{sec:complicated}
+
+This problem of predicting the vector shape directly is a lot to ask
+of a neural network, at least set up this way. One thing that did not
+sit well with me is that the network could in principle generate a
+perfect-looking result, but because it didn't have the points in the
+expected order, it would be penalized. This makes it harder to learn,
+and more prone to overfitting.\footnote{For example, imagine if the database
+contains two versions of Helvetica that just have their points in a
+different order---which is very likely the case btw---the model will
+have to learn how to distinguish between these, but using information
+we just don't care about.} This was one case where my questionable
+reflex to make things more complicated did pay off!
+
+First, I reduced the number of points in the input and output.
+Reducing the dimension of the function being learned generally makes
+learning a lot faster. This had the side-effect of reducing the number
+of eligible fonts (by about half), and by nature these fonts are
+simpler shapes. These effects alone could be responsible for the
+improved performance of this second try.
+
+I also output each contour's points in a normalized order, starting
+from the point closest to the origin. This removes one needless
+degree of freedom.\footnote{We can see how this manifests in
+  the biases on the output layer, which are a proxy for the
+  ``average prediction''. In the first model, because of the
+  unstructured order, these are mostly near 0.5 (center of the
+  character) or 0.0 (degenerate, unused contours). In this new model,
+  the distribution of biases is much more flat; it can learn that
+  ``the first point tends to be near 0.25,0.25'' and ``the seventh
+  point tends to be near 0.64,0.3.''}
+
+Aside from the changes in the input (now 254 nodes) and output (280),
+this second version has three sparse hidden layers of size 508, 508,
+and 560 nodes; the first two are dense and the latter sparse. The
+final model after some pruning had 609k parameters.
+
+As this was training, I worked on another improvement. Ideally we
+would compute the difference between the predicted shape and the
+expected shape, regardless of how they're drawn. Aside from being a
+bit computationally challenging, this won't really work because we
+need to attribute error directly to each output in order to actually
+update the model in training. I spent a valuable vacation day writing
+a routine to compute the best alignment of points between the actual
+and expected outputs (Figure~\ref{fig:trainingrotate}). Aside from being
+harder than it looked, my alignment code ended up being pretty slow
+relative to the rest of training, even worse since it ran on the CPU
+instead of GPU, which reduced the training speed by $50\%$. I let it
+run for 80,000 rounds, some 331 million training examples, but
+eventually got bored of waiting on this approach that was slow to
+train and seemed like a complicated version of an bad, oversimplified
+approach. So, I control-C'd that thing and threw this whole endeavour
+in the trash! But I must have confused the Recycle~Bin icon with the
+fairly complicated export-to-TrueType Font process that I built,
+because I ran the model on the venerable Futura~\cite{futura} font and
+generated {\bf Futurda} (Figure~\ref{fig:futurda}).
+% 
+% \begin{figure}[ht]
+% \centering
+%   \includegraphics[width=0.5 \linewidth]{rotatej}
+% \caption{ Example alignment (from debug UI) of predicted shape (blue)
+%   to expected shape (green). We require each point to be mapped (red)
+%   to a point from the expected contour in a monotonic order (but
+%   several can be mapped to the same one), so that we can attribute
+%   error to each point. I hope it was fun to write this code because
+%   it was too slow to do pretty much anything but generate this figure!
+% } \label{fig:rotatej}
+% \end{figure}
+%
+
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.95 \linewidth]{trainingrotate}
+\caption{ Screenshot (somewhat compacted) of training from near the
+  final round of the vector model's training, illustrating the
+  permissive loss function that finds the best alignment. At the bottom are
+  the predicted lowercase shapes (blue), also shown with their
+  expected shape (green). We require each point to be mapped (red)
+  to a point from the expected contour in a monotonic order (but
+  several can be mapped to the same one), so that we can attribute
+  error to each point.
+} \label{fig:trainingrotate}
+\end{figure}
+
+% XXX downloadable
+
+\begin{figure*}[tp]
+\centering
+  \includegraphics[width=0.95 \linewidth]{futurda}
+\caption{ Type specimen for the generated font {\bf Futurda}. This is
+  the classic font Futura, run through the final, improved
+  vector-based model (Section~\ref{sec:complicated}) to make each
+  letter lowercase. The letterforms \letterform{A}--\letterform{Z}
+  (top) become quite readable lowercase versions. The extra-lowercase
+  \letterform{a}--\letterform{z} are also almost legible, but are
+  mostly just scaled-down and screwed up versions of the lowercase
+  letterforms. Could definitely imagine this appear in the
+  ``distressed fonts'' category of a 10001 Free TrueType Fonts CD-ROM
+  in the 1990s, though.
 } \label{fig:futurda}
@@
 
+% obviously, need to express different cases of anger at various
+% problems or successes?
+
 % the sdf problem
+
+\section{SDFs}
+
+Don't give up! The fixed-size input/output of neural networks is
+better suited to something like an array of pixels, and fonts can of
+course be represented this way as well. To stay in the realm of what
+my desktop computer with a single GeForce 1080 can do, I wanted to
+keep the number of inputs and outputs pretty small. There's already an
+excellent technique for representing font data as compact bitmaps,
+which comes from computer graphics, called Signed Distance Fields
+(SDFs)~\cite{green2007improved}. In a standard rasterization of a
+font, each pixel of a bitmap contains 1 or 0, or perhaps an
+anti-aliased value in-between. In an SDF, the bitmap instead contains
+the distance to the nearest edge, and is signed (e.g. values inside
+the shape are $> 0$, value outside are $< 0$). Actually in
+practice we offset and saturate the values so that they are all in
+$[0,1]$ (or bytes in $[0,255]$), with some nonzero ``on-edge value''
+(say, 0.5) standing for ``distance 0''. In order to display the font
+at the size of your choice, you then resize the SDF image with bilinear
+interpolation, and then simply threshold the image. This works
+surprisingly well (Figure~\ref{fig:sdf}).
+
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.95 \linewidth]{sdf-figure}
+\caption{
+  The signed distance function representation of a letterform.
+  At the very left, a $36\times 36$ pixel rasterization of
+  the character without anti-aliasing, for comparison. Any
+  scaling of this will have chunky pixel artifacts. Next,
+  a $36\times 36$ pixel SDF of same. Third, simply scaling
+  that $36\times 36$ image to $180 \times 180$ pixels with
+  bilinear sampling. Finally, that image thresholded to
+  produce a $180 \times 180$ pixel rasterization, which is
+  far superior despite being derived from a $36 \times 36$ pixel
+  bitmap. Typically this process is performed at an even higher
+  scale and then downsampled to produce an anti-aliased image.
+} \label{fig:sdf}
+\end{figure}
+
+SDFs seem well-suited for machine learning. They contain more
+information per pixel than a plain bitmap, so we can use a smaller
+input and output size. On the input side, extremal pixels that would
+almost never be set in a bitmap still have significant information
+(distance to the character). The error function is just pixel-by-pixel
+difference. The rendering of the output is inherently tolerant of some
+noise because of the sampling and thresholding. So, this seemed like
+it might work really well! (It doesn't work that well.)
+
+I computed some stats on the font database, and determined the
+following parameters for the fixed-size SDFs we train on. The images
+are $36 \times 36$ pixels. The character box is placed such that there
+are $2$ pixels of top padding, and $9$ pixels of left and bottom
+padding. The character box is only ``nominal'' in the sense that the
+font's contours can exceed its bounds, and this is completely normal
+for a letter like \letterform{j} (which goes below the baseline and
+often hangs to the left of the origin as well). I used an ``on-edge
+value'' of 220 (because much more of the SDF is outside the letter
+than inside) and the distance is scaled as 15 units per pixel (so that
+pixels on the outer edge often have non-zero values). Compared to the
+first version, I was somewhat more permissive in what fonts I trained
+on, since there was no inherent limit to the number of contours or their
+complexity. I did exclude fonts whose rasterizations exceeded the
+bounds of the SDF, which is possible (very wide \letterform{W} or
+low-descending \letterform{j} perhaps) but rare.
 
@@
 
+Having committed to the representation, again it is ``just'' a matter
+of heating up the GPU to apply some linear and non-linear transforms.
+The initial network had an input size of $36 \times 36 = 1296$ for the
+SDF, and the output the same plus $26$ bonus outputs (one for each
+letter, as before). I started with three hidden layers of $1296$, $1296$,
+and $2916$ nodes, each sparse ($80\%$ of the weights are zero). Again,
+don't take notes. This one works a bit better than before, but still
+not impressive. The node references are assigned spatially (something
+like the $20\%$ of the nodes on the previous layer that are closest to
+the next layer's node) but due to a bug the spatial locality is
+actually pretty strange.
+% XXX if I show layer weights, forward-reference them here:
+% It's responsible for the XXX
+Every layer's transfer function is ``leaky relu'' again. It would
+definitely make sense to use convolutional layers for this problem, as
+features like serifs, lines, curves, and so on could appear throughout
+the input and output. I just haven't built support for that in my
+weird home-grown software, yet.
+
+I also adapted my weird home-grown software to train the
+\makeuppercase\ and \makelowercase\ models simultanously. Two models
+fit easily in GPU memory, with plenty of space for a stream of
+training data (one training instance is only about 10kb). The only
+challenging thing is arranging for them to feed each other generated
+``inversion'' examples (Figure~\ref{fig:cotraining}), but this is just
+a matter of programming, thank God. I should remember to do projects
+that are mostly a matter of programming. Each round, $25\%$ of the
+batch consists of inverted examples from the symmetric model's output
+from a recent round. Training happens asynchronously, but I make sure
+that one model is not allowed to get more than 2 rounds ahead of the
+other, because I want this feedback loop to be somewhat tight.
+
+So I did that and let it run for a month. Actually I had to start over
+several times with different parameters and initialization weights
+because it would get stuck (Figure~\ref{fig:nans}) right away. I
+prayed to the dark wizard of hyperparameter tuning until he smiled
+upon my initial conditions, knowing that somewhere he was adding
+another tick-mark next to my name in a tidy but ultimately scary
+Moleskine notebook that he bought on a whim in the Norman Y.~Mineta
+San Jose International Airport on a business trip, and still feels
+was overpriced for what it is.
+
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.9 \linewidth]{nans}
+\caption{
+  Divergent training after only 29 rounds. We have \nan{} total error
+  (is that good?). The example in column one is an inversion example
+  generated by the \makeuppercase\ model, which is why it also looks
+  like the Tunguska event, just of the opposite sign. The other two
+  are regular inputs, whose predicted outputs are black holes. Start
+  over!
+} \label{fig:nans}
+\end{figure}
+
+The training error over time appears in
+Figure~\ref{fig:sdfmergederror}. It looks like the ones I have seen in
+machine learning papers, although I don't like to read other people's
+papers because it just seems like spoilers.
+
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.99 \linewidth]{sdfmergederror}
+\caption{ The training error for the SDF models. The red curve is the
+  \makeuppercase\ model, which generally has a higher error rate
+  (perhaps simply because uppercase letters usually have more pixels
+  set) and blue is \makelowercase. The first few rounds have error
+  that's off the charts, well above 100. The most dramatic event is
+  around round 200,000, where I reduced the weight decay factor to
+  $0.999995$ (from $0.9995$). I guess you just need more nines to be
+  more reliable. There are some other visible peaks, which occur when
+  I do things like remove nodes with very low weights or which are
+  almost never activated (Section~\ref{sec:fiddly}). These momentarily
+  increase error but it is easily fine-tuned away (e.g. by learning
+  new biases). The peak at around 1.4M rounds is when I added a new
+  layer to the end of the model, which does seem to create a new
+  training regime (clear downward slope now); but this also
+  significantly increases the training cost per round. Even after
+  2,000,000 rounds, the network is still apparently improving, but at
+  a speed of about 1 pixel per several weeks. Eventually the extremely
+  strict SIGBOVIK deadlines mean you just have to call it done.
+} \label{fig:sdfmergederror}
+\end{figure}
+
+% you propagate my back, I'll propagate yours
+
 \label{sec:fiddly}
+
+% TODO loss over time
+
+% some other things that are worth a mention:
+%   autoparallel
+%   efficiency improvements (restrict, compile-time constants)
 
@@
 
-\section{Perfect letters, hallucinated}
-\label{sec:hallucination}
+\section{Perfect letters, hallucinated} \label{sec:hallucination}
+
+lorem ipsum (Figure~\ref{fig:perfect})
+
+\begin{figure*}[tp]
+\centering
+  \includegraphics[width=0.9 \linewidth]{perfecthallucination}
+\caption{
+%  XXX Why doesn't this caption come out the way I'm expecting?
+%  (in amsart style... can't have two figure*s on the same page?)
+   Type specimen for the generated font {\bf Perfect Hallucination}.
+  TODO write a description!
+} \label{fig:perfect}
+\end{figure*}
+
+% LSDF
+
+dolor sit amet
+
+\section{Chess-playing}
 

### 2021-03-13 lowercase/ttf.cc

@@ -522,7 +522,8 @@ static const std::unordered_map<char, string> &CharNames() {
 // the order of all contours in order to preserve their winding order.
 //
 // SFD format here: https://fontforge.org/docs/techref/sfdformat.html
-string TTF::Font::ToSFD(const string &name) const {
+string TTF::Font::ToSFD(const string &name,
+                        const string &copyright) const {
   // TTF uses integral coordinates. This is the size of the box
   // we discretize to. Larger numbers here mean less lost precision,
   // but since we are using 16 bit coordinates and it is normal for
@@ -548,6 +549,12 @@ string TTF::Font::ToSFD(const string &name) const {
 
   CHECK(descent >= 0) << "negative linegap removed entire descent?";
 
+  string name_no_space;
+  for (int i = 0; i < name.size(); i++) {
+    char c = name[i];
+    if (c != ' ') name_no_space += c;
+  }
+  
   // FYI the values in the Layer: command are what tell it that
   // we are using quadratic beziers.
   const int numchars = chars.size();
@@ -557,7 +564,7 @@ FontName: %s
 FullName: %s
 FamilyName: %s
 Weight: Regular
-Copyright: Copyright (c) 2021, Tom 7
+Copyright: %s
 UComments: "Generated by ttf.cc"
 Version: 001.000
 ItalicAngle: 0
@@ -597,7 +604,8 @@ AntiAlias: 1
 FitToEm: 0
 WinInfo: 54 18 11
 BeginChars: 256 %d
-)!", name.c_str(), name.c_str(), name.c_str(),
+)!", name_no_space.c_str(), name.c_str(), name.c_str(),
+     copyright.c_str(),
      ascent, descent,
      native_linegap, native_linegap, native_linegap,
      numchars);

### 2021-03-14 lowercase/paper/lowercase.tex
Fix bug in remapping that meant it didn't affect most of the image, ugh. Some additions to paper

@@
 
-\title{Lowestcase and Uppestcase letters: Further adventures in Derp Learning}
+\title{Lowestcase and Uppestcase letters: Advances in Derp Learning}
 \author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
@@
 
-\renewcommand\paragraph[1]{\smallskip{\bf #1}\enspace}
+\renewcommand\paragraph[1]{\smallskip \noindent{\bf #1}\enspace}
 
@@
 often hangs to the left of the origin as well). I used an ``on-edge
-value'' of 220 (because much more of the SDF is outside the letter
-than inside) and the distance is scaled as 15 units per pixel (so that
-pixels on the outer edge often have non-zero values). Compared to the
-first version, I was somewhat more permissive in what fonts I trained
-on, since there was no inherent limit to the number of contours or their
-complexity. I did exclude fonts whose rasterizations exceeded the
-bounds of the SDF, which is possible (very wide \letterform{W} or
-low-descending \letterform{j} perhaps) but rare.
+value'' of 0.862 (because much more of the SDF is outside the letter
+than inside) and the distance is scaled as 0.059 units per pixel (so
+that pixels on the outer edge often have non-zero values). Compared to
+the first version, I was somewhat more permissive in what fonts I
+trained on, since there was no inherent limit to the number of
+contours or their complexity. I did exclude fonts whose rasterizations
+exceeded the bounds of the SDF, which is possible (very wide
+\letterform{W} or low-descending \letterform{j} perhaps) but rare.
 
@@
 several times with different parameters and initialization weights
-because it would get stuck (Figure~\ref{fig:nans}) right away. I
-prayed to the dark wizard of hyperparameter tuning until he smiled
-upon my initial conditions, knowing that somewhere he was adding
-another tick-mark next to my name in a tidy but ultimately scary
-Moleskine notebook that he bought on a whim in the Norman Y.~Mineta
-San Jose International Airport on a business trip, and still feels
-was overpriced for what it is.
+because it would get stuck (Figure~\ref{fig:nans}) right away or as
+soon as I looked away from the computer. I prayed to the dark wizard
+of hyperparameter tuning until he smiled upon my initial conditions,
+knowing that somewhere he was adding another tick-mark next to my name
+in a tidy but ultimately terrifying Moleskine notebook that he bought
+on a whim in the Norman Y.~Mineta San Jose International Airport on a
+business trip, and still feels was overpriced for what it is.
 
@@
 
+\subsection{Fiddly bits} \label{sec:fiddly}
 The training error over time appears in
@@
 machine learning papers, although I don't like to read other people's
-papers because it just seems like spoilers.
+papers because it just seems like spoilers. There are several
+noticeable events in the curve, which came from me fiddling with the
+parameters or network as it ran. Here are some of the things I did:
+
+\paragraph{Vacuuming and culling.} Sometimes a node will just be
+dead (basically never activates) or an edge weight will be nearly
+zero. In these cases an equivalent, tidier network can be made by
+dropping the node or edge. Periodically I would perform these
+processes, sometimes feeling particularly choppy and removing like
+10\% of the parameters at a time. If these parameters are truly
+useless with no hope of recovery, we simply get faster training
+because there's less work to do.
+
+\paragraph{Widening.} The opposite thing is to introduce new nodes.
+Adding nodes to hidden layers is pretty easy. The thing that worked
+best for me is to increase the size of the layer by 10--15\%, where
+each new node has random incoming weights and bias 0. Then for each
+node on the next layer, I add edges to some subset of these new nodes
+(again generally 10\% of them) with weight 0. Since this weight is
+zero, the network computes the same function, but has new gradients
+to explore (in practice, it then experiences some shock after a
+few training rounds, but then quickly fine-tunes this away). More
+parameters means slower training, but also more potential to learn
+interesting functions, or overfit!
+
+\paragraph{Deepening.} It's also possible to add layers once the
+network is trained. This can be done anywhere, but I liked doing it on
+the output layer because this gets the most direct feedback from the
+training examples, and so it updates quickly and changes there are
+easy to understand. If you append the identity matrix (new layer is
+the same size as the previous; each node has weight 1.0 to its
+corresponding node and 0.0 elsewhere) then this network computes
+the same function but has new gradients to explore. Adding a layer
+did seem to help unlock a new training regime
+(Figure~\ref{fig:sdfmergederror}); subjectively it also reduced
+some weird artifacts in the SDFs that the model used to predict (makes
+sense; this most natural thing for this layer to do is learn how to
+predict a ``correction'' from the old prediction, for example
+by smoothing/sharpening it). This seems to be borne out by the
+weights, which are also fun to look at (Figure~\ref{fig:lastlayer}).
+
+\paragraph{Generating features.} On the other side, randomly sampling
+pixels from the input SDF does work, but I superstitiously believed
+that it might be better to have more spatially meaningful features. I
+wrote a program to generate a bunch of random simple features (one
+line/blob with positive weights, one line/blob with negative weights).
+It then chooses a set of them that are both {\em good} (maximum
+standard deviation on a sample of training data) and {\em different
+  from one another} (redundant or even partially redundant features
+are less valuable). It was nice to satisfy my supersition, and the
+dark wizard of superstitious fiddling with neural networks in the
+hope that they do the thing was pleased as well. The features are
+at least handsome (Figure~\ref{fig:makefeatures}).
+
+I assume these are all standard things that neural people do, but they
+do better and smarter versions of them because they are willing to
+read other people's papers instead of trying to figure things out from
+scratch all the time. But you gotta occupy yourself somehow while it
+crunches for a month.
+
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.99 \linewidth]{makefeatures}
+  \caption{ Some randomly-generated features with the selected ones
+    outlined in magenta, mostly shown here for aesthetic reasons.
+    Savvy Twitter user {\tt @iotathisworld} sees this as ``the classic
+    question: machine vision classification or 90s roller rink carpet
+    pattern?'' to which I deflect: ``Sorry, it's actually modern day
+    Port Authority bus upholstery or Gram stain of same!'' (But
+    actually machine vision classification was basically correct.)
+  } \label{fig:makefeatures}
+\end{figure}
 
@@
 
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.99 \linewidth]{lastlayer}
+  \caption{ The bottom-right corner of the weight matrix for the final
+    layer of the network. This layer was added to the network after
+    1.36 million rounds, initially as the identity matrix, and so can
+    be thought of partly as a correction of the network's output prior
+    to that round (though that part of the network continues to
+    evolve). The x-axis is the output nodes, and the y-axis is the
+    nodes of the previous layer. Note for example that the last 26
+    columns look pretty different; these are the predictors for the 26
+    letters, which occur in the output after the SDF pixels. Green
+    means positive and red means negative, so if you are looking at
+    this in a black-and-white printout, that may explain your current
+    confusion. The exact diagonal is a strong green, close to 1.0,
+    although over time these weights do diverge from the identity
+    somewhat. In the bottom-right corner of size $26^2$, we are
+    looking at how the 26 letter predictors are derived from the
+    previous layer's predictions. We see that most letters are
+    negatively correlated (makes sense; only one will ever be 1.0)
+    although there are some oddities (probably because it found some
+    other, better correlates). Nodes on this new layer have dense
+    references to these 26 predictions on the previous; this means
+    that the bottom row kind of represents biases for each of the 26
+    letters (what does an average 'e' look like?). I also included a
+    dense region above that, but this appears to have simply evolved
+    the same way as other $36^2$ chunks have (the rest are sparse).
+    These chunks have a large amount of spatial similarity (suggesting
+    that the sparse sampling would be adequate), with a meaning like
+    ``if this area of the image is bright, then this pixel should
+    be less bright.'' It is interesting that the pixels immediately
+    next to the diagonal are almost always strongly negative
+    (sharpening operation). ``Thank you for attending my TED talk.''
+    ---\,Figure~\ref{fig:lastlayer}}
+  \label{fig:lastlayer}
+\end{figure}
+
+For completeness, some other innovations that I feel are worth
+mentioning:
+
+Making the GPU code faster has really high value (could save weeks of
+waiting). Since I am using OpenCL (whoa, yeah, stop me right there, I
+know) I found a good trick was to generate different OpenCL code with
+constants baked for each layer (for example their size and
+\verb+indices_per_node+); this allows some of the inner loops to use
+faster tricks for e.g.~multiplication by a compile-time constant
+instead of depending on an argument or data. I have different routines
+for sparse and dense layers. It might even make sense to recompile the
+kernels for other parameters that change over the lifetime of
+training, like the learning rate. The {\tt fma} instruction (so named
+for the physical law $F=MA$) is a bit faster than
+\verb|potential += w * v|, and I guess the compiler can't do this
+itself because of IEEE horrors. But like, who cares? In my opinion you
+should be able to put it in ``fast machine learning mode'' where it
+readily makes precision errors, with a command-line option like
+\verb+--fml+. With all the tweaking, the easiest win was to use the
+{\tt restrict} keyword on arguments to tell the compiler that the
+input cannot alias the output, for example; this presumably helps
+it schedule instructions better.
+
+Various things in training run in parallel threads (e.g.~processing
+fonts, but also moving data to the GPU, backpropagation for each
+example, etc.). For a long time I had just been explicitly setting
+parallelism using superstitious constants. For this project I finally
+just wrote something that would automatically and empirically
+determine the number of threads that yielded the highest throughput,
+and persisted that information across program starts. This was a
+good idea and enters the permanent rotation.
+
+The actual error on the predicted SDFs is pretty low; for the
+\makelowercase\ model it is around 31.3, which is like if 2.4\% of the
+pixels were (completely) wrong, but the rest is exactly correct. In
+reality, of course, the error is distributed throughout the pixels,
+and some errors are a lot more important than others. Particularly,
+near the threshold value, a pixel goes from from being considered ``in
+the letter'' to ``outside'' with tiny changes in its value. Changes to
+a pixel with a value near 0.0 or 1.0 usually doesn't affect the output
+shape at all, in contrast. So one thing I did was map the loss
+function (comparing expected pixel value to actual) to ``stretch out''
+the region near the threshold, increasing the penalty (basically, the
+derivative) in that region and decreasing it elsewhere. Looking at the
+code again right now, I realize that I only applied this to the first
+row of the SDF (\verb+idx < SDF_SIZE+ instead of
+\verb+SDF_SIZE * SDF_SIZE+), so that was dumb. I will say in my
+defense that at least I felt disappointed that it didn't seem to make
+a difference!
+
 % you propagate my back, I'll propagate yours
-
-\label{sec:fiddly}
-
-% TODO loss over time
-
-% some other things that are worth a mention:
-%   autoparallel
-%   efficiency improvements (restrict, compile-time constants)
 
@@
 
-% stretch the loss function near the onedge value
+\subsection{Upperercase and Lowerercase fonts} \label{sec:fonts}
 
-%
+Let's make some fonts!
+
+% autotracing in here, I think?
 
@@
 % lowercase uppercase letters etc.
+% iterate this a lot of times?
+
+% run it to infinity to see the uppestcase, etc.
+% if these vanish, then we have a fixed point.
+% if not, or perhaps by doing some auto-gamma, then
+% it's like a fractal zoomer, and we can set the
+% record on the "most uppercase letter" by running
+% a jazillion times.
 

### 2021-03-14 lowercase/makefeatures.cc

@@ -3,8 +3,9 @@
 // (and more principledly) by the training process on random features,
 // but it seems good to start with better quality features?
 
-// This code generates the requested number of features to a file,
-// which can then be added to a model with widen.exe (not yet implemented!)
+// This code generates the requested number of features to a file
+// (a fake single-layer network) which can then be added to a model with
+// widen.exe.
 
 #include <string.h>
 #include <stdio.h>
@@ -151,7 +152,8 @@ struct TrainingExample {
   char letter;
 };
 
-// Generate num_output_features "good" features that are different from one another.
+// Generate num_output_features "good" features that are different
+// from one another.
 static void Generate(const std::string &model_filename,
                      int num_output_features, bool lowercasing) {
   std::unique_ptr<Network> net(Network::ReadNetworkBinary(model_filename));
@@ -571,7 +573,8 @@ static void Generate(const std::string &model_filename,
         // Compute initial distance.
         double dist = GetDist();
 
-        // Save the permutation if it is the global best, using double-checked lock.
+        // Save the permutation if it is the global best, using
+        // double-checked lock.
         auto SaveBest = [&best_m, &best_out, &best_dist, nof, &dist, &perm]() {
             if (dist > ReadWithLock(&best_m, &best_dist)) {
               WriteMutexLock ml(&best_m);
@@ -670,8 +673,9 @@ static void Generate(const std::string &model_filename,
   
   Timer make_timer;
   {
-    // Now save 'em. We make a fake one-layer network since the features
-    // have the same serialization needs as a hidden layer (no coincidence here!)
+    // Now save 'em. We make a fake one-layer network since the
+    // features have the same serialization needs as a hidden layer
+    // (no coincidence here!)
     vector<int> nn = {width * height, num_output_features};
     vector<int> ii = {feature_size};
     vector<TransferFunction> tfs = {LEAKY_RELU};

### 2021-03-14 trunk/lowercase/train.cc
Fix bug in remapping that meant it didn't affect most of the image, ugh. Some additions to paper

@@ -337,7 +337,7 @@ static string GetRemap() {
 
   [[maybe_unused]]
   auto Remap = [](int i, float x) {
-      if (i < SDF_SIZE) {
+      if (i < SDF_SIZE * SDF_SIZE) {
         // PERF could probably simplify this!
         if (x < ax) return x * s;
         if (x < bx) return ay + (x - ax) * interesting_slope;
@@ -365,7 +365,7 @@ static string GetRemap() {
       "%.9ff + (x - %.9ff) * %.9ff) : "
       // else return x
       "x)",
-      SDF_SIZE,
+      SDF_SIZE * SDF_SIZE,
       ax, s,
       bx, ay, ax, interesting_slope,
       by, bx, s);
@@ -2349,7 +2349,7 @@ struct Training {
     backward_comp = std::make_unique<AutoParallelComp>(
         32, 50, false,
         StringPrintf("autoparallel.%s.bwd.txt", experiment.c_str()));
-    
+
     decay_comp = std::make_unique<AutoParallelComp>(
         8, 50, false,
         StringPrintf("autoparallel.%s.dec.txt", experiment.c_str()));

### 2021-03-18 lowercase/paper/lowercase.tex
Find loops and render them, works

@@
 \section{1000001 Free Fonts}
+
+% no copyright intended
 
@@
 row of the SDF (\verb+idx < SDF_SIZE+ instead of
-\verb+SDF_SIZE * SDF_SIZE+), so that was dumb. I will say in my
-defense that at least I felt disappointed that it didn't seem to make
-a difference!
+\verb+SDF_SIZE * SDF_SIZE+), so that was dumb AND MAKES ME {\bf
+  ANGRY}. I will say in my defense that at least I felt disappointed
+at the time that it didn't seem to make a difference!
 
@@
 
-% no copyright intended
-
 \subsection{Upperercase and Lowerercase fonts} \label{sec:fonts}
 
-Let's make some fonts!
+Now that we have these expensive models, we can use them to make
+arbitrary letterforms uppercase or lowercase. The output is readily
+rasterized (using the standard thresholding approach for SDFs) but
+we'd actually like to have vector representations so that we can
+use them as TrueType fonts for our desktop publishing needs.
+
+\subsection{Tracing}
+Automatically tracing bitmaps into vector form is no doubt a solved
+problem, but I chose not to look at spoilers. Since we actually have a
+signed distance field, we can build a tracing routine directly off of
+that. The approach I took consists of three steps. First, I generate a
+bitmap of the SDF (at its native size) using the threshold. I separate
+the image into a nested tree of connected components in this pixel
+space; each component knows its single parent and whether it is
+``land'' (inside the shape) or ``sea''. Characters like \letterform{e}
+need internal cutouts, which are represented by a different winding
+order (clockwise or counter-clockwise) for the contour. Some of the
+tricky cases in computing this tree structure are given in f
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.9 \linewidth]{connected}
+  \caption{
+
+  } \label{fig:connected}
+  \end{figure}
 
@@
 
+% screenshot UI where I can draw letters and get live
+% predictions? Just for the video perhaps? I like the "cool s" joke?
+
 \section{Perfect letters, hallucinated} \label{sec:hallucination}
+
+Wow, this project is pretty involved, huh? Let's just add another
+dimension to it.
 
@@
 
+\section{To infinity, but let's stop there}
+
 % lowercase uppercase letters etc.
 % iterate this a lot of times?
+% "to infinity, but let's stop there"
 
@@
 % a jazillion times.
+
+% lowestcase is just dust, not surprising
+% uppestcase is actually a fixed point (not FixedSys)
+% ... so it's like a 3D manifold
 

### 2021-03-18 lowercase/renderloop.cc

@@ -1,0 +1,244 @@
+
+#include <vector>
+#include <string>
+#include <algorithm>
+#include <cstdint>
+#include <memory>
+
+#include "timer.h"
+#include "font-problem.h"
+
+#include "image.h"
+#include "lines.h"
+#include "base/stringprintf.h"
+#include "geom/marching.h"
+
+#include "network.h"
+#include "threadutil.h"
+
+using namespace std;
+
+static constexpr FontProblem::SDFConfig SDF_CONFIG = {};
+static constexpr int SDF_SIZE = SDF_CONFIG.sdf_size;
+
+using uint8 = uint8_t;
+using uint32 = uint32_t;
+using int64 = int64_t;
+
+vector<bool> Threshold(const ImageF &a) {
+  vector<bool> ret;
+  ret.reserve(a.Width() * a.Height());
+  for (int y = 0; y < a.Height(); y++) {
+    for (int x = 0; x < a.Width(); x++) {
+      ret.push_back(a.GetPixel(x, y) > (SDF_CONFIG.onedge_value / 255.0f));
+    }
+  }
+  return ret;
+}
+
+constexpr float SCALE = 2.0;
+constexpr int CYCLE_START = 245;
+constexpr int CYCLE_END = 377;
+constexpr bool LOWER = false;
+
+/*
+constexpr float SCALE = 1.5;
+constexpr int CYCLE_START = 202;
+constexpr int CYCLE_END = 389;
+constexpr bool LOWER = false;
+*/
+
+/*
+constexpr float SCALE = 3.0;
+constexpr int CYCLE_START = 2291;
+constexpr int CYCLE_END = 2957;
+constexpr bool LOWER = false;
+*/
+
+int main(int argc, char **argv) {
+  TTF helvetica("helvetica.ttf");
+
+  std::unique_ptr<Network> make_lowercase, make_uppercase;
+  make_lowercase.reset(Network::ReadNetworkBinary("net0.val"));
+  make_uppercase.reset(Network::ReadNetworkBinary("net1.val"));
+
+  CHECK(make_lowercase.get() != nullptr);
+  CHECK(make_uppercase.get() != nullptr);
+
+  vector<ImageF> cycle;
+
+  const Network *net =
+    LOWER ? make_lowercase.get() : make_uppercase.get();
+
+  Timer timer;  
+  {
+    std::optional<ImageA> sdfo =
+      helvetica.GetSDF('q', SDF_CONFIG.sdf_size,
[…]

### 2021-03-18 trunk/lowercase/makefont.cc
Find loops and render them, works

@@ -41,6 +41,8 @@ struct Config {
 
   string font_name = "Heavenica";
   string filename = "heavenica.sfd";
+  string copyright;
+
   float extra_scale = 1.0f;
   float linegap = 0.0f;
   float max_gamma = 1.0f;
@@ -60,6 +62,7 @@ Config Heavenica() {
   cfg.input_font = "helvetica.ttf";
   cfg.font_name = "Heavenica";
   cfg.filename = "heavenica.sfd";
+  cfg.copyright = "http://tom7.org/lowercase - Generated by SDF-based neural network. Makes capital letters (punctuation, etc.) from Helvetica 'even more capital', and lowercase are the predicted lowercase of those. Distribute freely";
   cfg.extra_scale = 1.2f;
   cfg.linegap = -0.1f;
   cfg.max_gamma = 0.95f;
@@ -86,6 +89,8 @@ Config SpezialHellvetica() {
   cfg.input_font = "helvetica.ttf";
   cfg.font_name = "Spezial Hellvetica";
   cfg.filename = "spezial-hellvetica.sfd";
+  cfg.copyright = "http://tom7.org/lowercase - Generated by SDF-based neural network. Makes lowercase letters (punctuation, etc.) from Helvetica 'even more lowercase', and capitals are the predicted uppercase of those. Distribute freely";
+
   // XXX tweak
   cfg.extra_scale = 1.3f;
   cfg.linegap = -0.05f;
@@ -113,6 +118,7 @@ Config ExponentialNewRoman() {
   cfg.input_font = "times.ttf";
   cfg.font_name = "Exponential New Roman";
   cfg.filename = "expnewroman.sfd";
+  cfg.copyright = "http://tom7.org/lowercase - Generated by SDF-based neural network. Makes capital letters (punctuation, etc.) from Times New Roman 'even more capital', and lowercase are the predicted lowercase of those. Distribute freely";
   cfg.extra_scale = 1.2f;
   cfg.linegap = -0.1f;
   cfg.max_gamma = 0.95f;
@@ -139,6 +145,7 @@ Config PlusNewRoman() {
   cfg.input_font = "times.ttf";
   cfg.font_name = "Plus New Roman";
   cfg.filename = "plusnewroman.sfd";
[…]

### 2021-03-18 trunk/lowercase/ui.cc

@@ -63,6 +63,7 @@ enum class Mode {
   SCALETEST,
   LOOPTEST,
   BEZDIST,
+  ZOOM,
   DRAW,
 };
 
@@ -136,8 +137,14 @@ private:
   float value = 1.0;
 };
 
+struct HashImageA {
+  std::size_t operator()(const ImageA &img) const {
+    return img.Hash();
+  }
+};
+
 struct UI {
-  Mode mode = Mode::BEZDIST;
+  Mode mode = Mode::ZOOM;
   DrawMode draw_mode = DrawMode::DRAWING;
 
   std::shared_mutex dirty_m;
@@ -178,7 +185,8 @@ struct UI {
   void DrawSDF();
   void DrawDrawing();
   void DrawBez();
-
+  void DrawZoom();
+  
   void ClearDrawing();
 
   // Draw line into drawing. Coordinates are relative to drawing, but may
@@ -225,14 +233,22 @@ struct UI {
   float current_xscale = 1.0, current_yscale = 1.0;
   float current_xoff = 0.0, current_yoff = 0.0;
   TTF times{"times.ttf"};
-
+  TTF helvetica{"helvetica.ttf"};
+  
   int bez_startx = 800, bez_starty = 128;
   int bez_cx = SCREENW - 32, bez_cy = SCREENH / 2;
   int bez_endx = 10, bez_endy = 188;
   int bez_px = 100, bez_py = 100;
   // double bez_time = 0.0;
   // int bez_cpx = 0, bez_cpy = 0;
-
+  
+  ImageF zoom_sdf{SDF_CONFIG.sdf_size, SDF_CONFIG.sdf_size};
+  int64 zoom_iters = 0;
+  bool zoom_lower = true;
+  bool zoom_loop = false;
+  float zoom_gamma = 1.0f;
+  std::unordered_map<ImageA, int, HashImageA> zoom_seen;
+  
   vector<FontProblem::Point> looptest_expected;
   vector<FontProblem::Point> looptest_actual;
   std::optional<FontProblem::LoopAssignment> looptest_assignment;
@@ -786,44 +802,6 @@ void UI::FloodFill(int x, int y, Uint32 color) {
   result_cv.notify_all();
 }
 
-// Returns (x, y, dist).
-// PERF! There are definitely analytical solutions to this, and also
-// much faster ways to optimize (root of a third-degree polynomial).
-static std::tuple<float, float, float>
-DistanceFromPointToBezierSearch(
-    // The point to test
-    float px, float py,
-    // Bezier start point
[…]

### 2021-03-20 lowercase/paper/lowercase.tex
More work on paper, getting there...

@@
-%
-
-% TODO: SDF vs SFD typo factory!
-
-\documentclass[twocolumn]{article} % {amsart}
+\documentclass[twocolumn]{article}
+% {amsart}
 \usepackage[top=0.75in, left=0.65in, right=0.65in, bottom=0.6in]{geometry}
@@
 \usepackage{graphicx}
-\usepackage{chessboard}
 % IPA symbols. safe turns off overrides for like \! which I still want
 \usepackage[safe]{tipa}
-
+\usepackage{chessboard}
 % \usepackage{chessfs}
-\usepackage{adjustbox}
+% \usepackage{adjustbox}
 
@@
 
-\interfootnotelinepenalty=0
-
 % lets me explicitly set a. or 1. etc. as enum label
-\usepackage{enumitem}
+% \usepackage{enumitem}
 
@@
 
+\interfootnotelinepenalty=0
 \setlength{\footnotesep}{2em}
 
-% \newcommand\comment[1]{}
 \newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
+
+% XXX the collection of crap above causes a first page
+% with just "1" on it (probably conflicting packages). This
+% magic just drops the first page, which is of course not
+% the right solution, but...
+\usepackage{atbegshi}% http://ctan.org/pkg/atbegshi
+\AtBeginDocument{\AtBeginShipoutNext{\AtBeginShipoutDiscard}}
 
@@
 lowercase t is.
+
+% shift-reduce conflict
 
@@
   mostly just scaled-down and screwed up versions of the lowercase
-  letterforms. Could definitely imagine this appear in the
+  letterforms. Could definitely imagine this appearing in the
   ``distressed fonts'' category of a 10001 Free TrueType Fonts CD-ROM
@@
   \includegraphics[width=0.9 \linewidth]{nans}
-\caption{
-  Divergent training after only 29 rounds. We have \nan{} total error
-  (is that good?). The example in column one is an inversion example
-  generated by the \makeuppercase\ model, which is why it also looks
-  like the Tunguska event, just of the opposite sign. The other two
-  are regular inputs, whose predicted outputs are black holes. Start
-  over!
+\caption{ Divergent training after only 29 rounds. We have \nan{}
+  total error (hard to say if that's good or bad?). The example in
+  column one is an inversion example generated by the
+  \makeuppercase\ model, which is why it also looks like the Tunguska
+  event, just of the opposite sign. The other two are regular inputs,
+  whose predicted outputs are black holes. Start over!
 } \label{fig:nans}
@@
 machine learning papers, although I don't like to read other people's
-papers because it just seems like spoilers. There are several
-noticeable events in the curve, which came from me fiddling with the
-parameters or network as it ran. Here are some of the things I did:
+papers because it just seems like spoilers, and reading is the
+opposite of writing! There are several noticeable events in the curve,
+which came from me fiddling with the parameters or network as it ran.
+Here are some of the things I did:
 
@@
 useless with no hope of recovery, we simply get faster training
-because there's less work to do.
+because there's less work to do. Speed is exhilarating!
 
@@
 parameters means slower training, but also more potential to learn
-interesting functions, or overfit!
+interesting functions, or overfit! Danger is exciting!
 
@@
 weights, which are also fun to look at (Figure~\ref{fig:lastlayer}).
+All problems in computer science can be solved by an additional
+layer of indirection!
 
@@
 hope that they do the thing was pleased as well. The features are
-at least handsome (Figure~\ref{fig:makefeatures}).
-
-I assume these are all standard things that neural people do, but they
+at least handsome (Figure~\ref{fig:makefeatures}). Creativity is
+enriching!
+
+I presume these are all standard things that neural people do, but they
 do better and smarter versions of them because they are willing to
@@
   2,000,000 rounds, the network is still apparently improving, but at
-  a speed of about 1 pixel per several weeks. Eventually the extremely
-  strict SIGBOVIK deadlines mean you just have to call it done.
+  a speed of about 1 pixel loss per several weeks. Eventually the
+  extremely strict SIGBOVIK deadlines mean you just have to call it done.
 } \label{fig:sdfmergederror}
@@
     be thought of partly as a correction of the network's output prior
-    to that round (though that part of the network continues to
+    to that round (though the remainder of the network continues to
     evolve). The x-axis is the output nodes, and the y-axis is the
@@
 constants baked for each layer (for example their size and
-\verb+indices_per_node+); this allows some of the inner loops to use
-faster tricks for e.g.~multiplication by a compile-time constant
+\verb+indices_per_node+); this allows the cmopiler to use faster
+tricks in inner loops for e.g.~multiplication by a compile-time constant
 instead of depending on an argument or data. I have different routines
@@
 order (clockwise or counter-clockwise) for the contour. Some of the
-tricky cases in computing this tree structure are given in f
+tricky cases in computing this tree structure are given in
+Figure~\ref{fig:connected}. Once I have this tree structure, I trace
+each pixel mass recursively (Figure~\ref{fig:trace}). I find a pixel
+on the edge, and then walk around that edge clockwise (simple case
+analysis on the three pixels ahead of me). As I walk the edge, I look
+at the normal (orthographic as we are doing orthography) of the edge
+and see where it reaches the edge value on the SDF; this point (a
+float) is output into the contour. I'm guaranteed to get back to where
+I started. I recurse by negating the SDF (outside becomes inside) and
+bitmap (land becomes sea), and reverse the winding order of the result
+of recursion.
+
+This gives me a perfectly fine line-based outline of the SDF's shape.
+Since I output points at every pixel, sometimes these points are
+inefficient (i.e. a series of colinear points on a straight line),
+and sometimes they reflect sharp corners that are not aesthetic. So
+I then take a second pass at each contour, and try fitting B\'ezier
+curves to sequences of points while the error remains low. Again I
+did fitting with a black-box optimizer, which is nice. However,
+the function being minimized also needs to be able to find the
+closest point on a B\'ezier curve to another point, and although
+this can also be done easily with the black-box optimizer, nesting
+an optimizer invocation inside another one proved to be way too slow.
+I found an old algorithm in a book I owned and was stymied by as a
+child~\cite{glassner1990graphics}.
 
@@
   \centering
-  \includegraphics[width=0.9 \linewidth]{connected}
+\includegraphics[width=0.6 \linewidth]{connected}
 \caption{
-
+  Some tricky cases to think about when generating the nested
+  connected components, as the first step of tracing SDFs. Area
+  {\bf 0} is the outside of the entire letterform, but note that
+  we should include the top-left corner even though it is not
+  reachable without leaving the bounds of the image. Area {\bf 1}'s
+  parent is 0; it has two holes within it, Areas {\bf 2} and {\bf 3}.
+  At the top left, the four pixel chunks making up area {\bf 4} are
+  not actually connected, but they separate the hole which is
+  child are {\bf 5}. This hole must have one parent, so it means
+  that all four pixel chunks are part of the same area {\bf 4}.
 } \label{fig:connected}
@@
 
-% autotracing in here, I think?
-
-% screenshot UI where I can draw letters and get live
-% predictions? Just for the video perhaps? I like the "cool s" joke?
+\begin{figure}[ht]
+  \centering
+\includegraphics[width=0.95 \linewidth]{trace}
+\caption{ Tracing the SDF from Figure~\ref{fig:sdf} into vector
+  format. Left image shows the nested connected components. Middle
+  image is the initial straight-line trace, and the right image
+  shows the simplified contours using quadratic B\'eziers.
+} \label{fig:trace}
+\end{figure}
 
 \section{Perfect letters, hallucinated} \label{sec:hallucination}
+
+All right, but why stop there??
+
+... the fonts in the previous section were created by modifying the
+case of existing letterforms, with mixed success. We can also do this
+to any letter-like shape. I built a UI for drawing letters and seeing
+them uppercased and lowercased (and then re-lowercased and
+re-uppercased) live, but it's impossible to demonstrate in paper
+form. It's pretty much what you'd expect.
+
+The UI also tells you how much your input resembles the various
+existing letters \letterform{A}--\letterform{Z} and
+\letterform{a}--\letterform{z} using the 26 bonus outputs that each
+model predicts. For example, I learned that the ``Cool
+S''~\cite{wikipediacools}:
+
+\begin{center}
+\includegraphics[width=0.1 \linewidth]{cools}
+\end{center}
+
+\noindent does not much resemble an S.
+% XXX ideally, numbers matching a scene in the video?
+
+This begs us to ask the question: What shapes {\em do} look like
+letters? Since the models will tell us, I can just search over shapes
+and ask them. The first thing I tried was to just generate random
+shapes and optimize their parameters to produce the highest possible
+prediction for the target letter, and the lowest possible prediction
+for the rest. This preduced results that are fully bonkers
+(Figure~\ref{fig:bonkersf}).
+
+We can improve the results by searching for inputs with a ``perfect''
+prediction (1.0) rather than making it as high as possible. These
+results may not have been fully bonkers, but were at least downright
+wacky. Since there appear to be a large variety of inputs that the
+model judges as ``perfect'', the most appealing results from this
+excursion involved scoring some additional properties of the
+hallucinated inputs to discourage them from being so barmy. I
+generated $8 \times 8$ bitmaps, plenty of pixels to make readable
+letters on classic computers. Rather than allowing them to be
+arbitrarily noisy, I also weakly optimized for (1) the number of
+pixels set being close to half (2) minimal transitions between
+on and off along each row and column. This produced shapes that
+are basically letter-like, but weird (Figure~\ref{fig:perfect}).
+
+\begin{figure}[tp]
+\centering
+  \includegraphics[width=0.9 \linewidth]{bonkersf}
+\caption{ A randomly generated SDF (left) and its rasterization
+  (right) which maximized the predicted score for \letterform{F}
+  (1.893 out of a nominal 1.0) while 
+  scoring all other letters low. Parts are recognizable as an
+  \letterform{F}, but other parts are fully bonkers. This is actually
+  one of the least weird ones.
+} \label{fig:bonkersf}
+\end{figure}
+
+\begin{figure*}[tp]
+\centering
+  \includegraphics[width=0.9 \linewidth]{perfecthallucination}
+\caption{
+   Type specimen for the generated font {\bf Perfect Hallucination}.
+   Each letter is an 8x8 bitmap that looks as close to ``perfect''
+   as possible to the model. Perfect here means that for \letterform{C},
+   the \makelowercase\ model outputs as close to the vector
+   $[0, 0, 1, 0, 0, \ldots, 0]$ (in its 26 letter predictors; the
+   actual lowercasing is ignored) as the optimizer could find. (Of
+   course I didn't search all $2^{64}$ inputs, but errors are
+   on the order of one part per thousand). The models are completely
+   successful at recognizing normal-looking letters as well, but
+   it likes these even better.
+} \label{fig:perfect}
+\end{figure*}
+
+% LSDF
+
+\section{Chess-playing}
+
+\newcommand\deterministic{
+  \includegraphics[height=0.8em]{deterministic}
+}
+\newcommand\asymmetric{
+  \includegraphics[height=0.8em]{asymmetric}
+}
+
+One obvious thing to do with a program that takes an $8 \times 8$
+bitmap and produces some kind of score for it is to use that
+program to play chess~\cite{murphy2019blind}. Here I entered
+26 such programs in the Elo World tournament~\cite{murphy2019eloworld},
+which allows us to see how they perform against each other
+and benchmark algorithms. (Badly.)
+
+At each turn, the algorithm takes the board state that would
+result from each legal move, and interprets it as an $8 \times 8$
+bitmap. It renders that bitmap as an SDF and then runs the
+\makelowercase\ model on it, and chooses the move that minimizes
+the difference between its letter predictions and the
+$[0, 0, \ldots 1, \ldots 0]$ vector corresponding to the letter
+we are ``playing as.'' \deterministic \asymmetric
+
+The players all perform fairly badly. Letters \letterform{E} and
+\letterform{S} are the worst (agreed on \letterform{S} being the worst,
+thank you very much!), even worse than the ``No, I insist!'' strategy
+that tries to force its opponent to capture its pieces. Letters
+\letterform{K} and \letterform{T} (eponymous! yeah!) perform the best,
+but still worse than random. The numeric players like $\pi$ and $e$
+are categorically better than the alphabetical ones, but this is not
+surprising because chess is more of a mathematical game than a
+linguistic one (Figure~\ref{fig:chess}).
+
+\begin{figure}[tp]
+  \centering
+  \chessboard[setfen=B2k4/8/1b2p2n/1b2p1pr/6Pp/4B2P/4q3/4K3 w - - 10 60,showmover=false]
+  \caption{
+    The letter \letterform{P} (white) loses to the numeric constant
+    $\pi$ (black) after 59 nonsensical moves. The \letterform{P}
+    player tries to move pieces such that the board looks like a
+    letter \letterform{P} (and no other) to the neural network.
+    The $\pi$ player uses
+    $3 - \pi$ to arithmetically decode the sequence of legal moves
+    (sorted alphabetically by PGN). Neither player is concerned with
+    chess, really, but the letter-based players are generally bad
+    because they are more likely to get stuck in local minima once
+    they are basically happy with the shape of the pieces.
+    %  1.
+Na3 g5 2.
+b3 e6 3.
+f4 Nc6 4.
+e4 Be7 5.
+e5 a6 6.
+c3 f6 7.
+h3 Bb4 8.
+g4 h5 9.
+Nf3 h4 10.
+Nd4 Rb8 11.
+Nf5 Na5 12.
+Nc2 Rh7 13.
+Nce3 d6 14.
+Qf3 Rh5 15.
+Nh6 Bd7 16.
+Bc4 c6 17.
+Rf1 Qc7 18.
+fxg5 Bxc3 19.
+b4 Bxb4 20.
+a3 Bc8 21.
+axb4 Qe7 22.
+Kd1 c5 23.
+d3 Nxh6 24.
+Ke2 Ra8 25.
+Rd1 dxe5 26.
+Bb5 Bd7 27.
+Ke1 c4 28.
+Nd5 Qg7 29.
+Nc3 Nb3 30.
+Bc6 Nd4 31.
+Be3 Nc2 32.
+Kf1 cxd3 33.
+Rdc1 d2 34.
+Qf4 Qh7 35.
+Qd4 Na3 36.
+Rxa3 a5 37.
+Qe4 b6 38.
+bxa5 dxc1=B 39.
+Qd5 Kd8 40.
+Qc5 Rb8 41.
+Ke1 Qe7 42.
+Bf3 fxg5 43.
+Bb7 Bxa3 44.
+Ba6 Bxc5 45.
+Bb5 Qd6 46.
+Ba6 Ba4 47.
+axb6 Bc6 48.
+Bc8 Qe7 49.
+Ba6 Rb7 50.
+Bxb7 Bxb6 51.
+Na2 Bb5 52.
+Nc1 Qa3 53.
+Nd3 Kd7 54.
+Nb2 Qxb2 55.
+Bc8 Ke8 56.
+Ba6 Kd7 57.
+Bc8 Kd8 58.
+Bb7 Qg2 59.
+Ba8 Qe2
+    } \label{fig:chess}
+\end{figure}
+
+% figure? drop it?
+{
+  \tiny
+\begin{tabular}{lrrrr}
+name & elo & win & loss & draw \\
+no i insist & 605.47 & 0 & 20405 & 42595 \\
+letter f & 605.60 & 1341 & 21653 & 40006 \\
+letter y & 606.00 & 1347 & 21688 & 39965 \\
+letter b & 607.39 & 1705 & 21870 & 39425 \\
+letter p & 607.61 & 1790 & 21963 & 39247 \\
+letter l & 608.75 & 1370 & 21306 & 40324 \\
+letter j & 610.86 & 1737 & 21525 & 39738 \\
+letter u & 611.41 & 1363 & 20947 & 40690 \\
+letter c & 612.38 & 1264 & 20787 & 40949 \\
+huddle & 612.60 & 1172 & 20494 & 41334 \\
+letter n & 612.81 & 1395 & 20802 & 40803 \\
+letter w & 613.64 & 1342 & 20723 & 40935 \\
+letter g & 613.72 & 1339 & 20660 & 41001 \\
+letter v & 614.18 & 1390 & 20606 & 41004 \\
+letter h & 615.04 & 1328 & 20452 & 41220 \\
+letter r & 615.22 & 1826 & 20932 & 40242 \\
+letter x & 615.70 & 1414 & 20457 & 41129 \\
+letter m & 616.31 & 1774 & 20809 & 40417 \\
+letter q & 618.85 & 1378 & 19993 & 41629 \\
+letter o & 619.08 & 1502 & 20078 & 41420 \\
+letter z & 620.09 & 1745 & 20275 & 40980 \\
+letter d & 620.10 & 1730 & 20369 & 40901 \\
+letter a & 620.57 & 1819 & 20212 & 40969 \\
+letter i & 622.50 & 2169 & 20424 & 40407 \\
+letter k & 622.90 & 1755 & 19878 & 41367 \\
+letter t & 623.30 & 2237 & 20271 & 40492 \\
+\ldots \\
+random move & 655.90 & 7267 & 20983 & 34750 \\
+\ldots \\
+chessmaster.nes lv1 & 1014.74 & 37302 & 13992 & 11706 \\
+\ldots
+stockfish1m & 2831.66 & 60167 & 493 & 2340 \\
+\end{tabular}
+}
+
+% CAPS LOCK, LOCK, CAPTAIN'S LOCK
+
+% when two letters coincide, a shift-reduce conflict
+
+\section{To infinity, but let's stop there}
 
@@
 dimension to it.
-
-lorem ipsum (Figure~\ref{fig:perfect})
-
-\begin{figure*}[tp]
-\centering
-  \includegraphics[width=0.9 \linewidth]{perfecthallucination}
-\caption{
-%  XXX Why doesn't this caption come out the way I'm expecting?
-%  (in amsart style... can't have two figure*s on the same page?)
-   Type specimen for the generated font {\bf Perfect Hallucination}.
-  TODO write a description!
-} \label{fig:perfect}
-\end{figure*}
-
-% LSDF
-
-dolor sit amet
-
-\section{Chess-playing}
-
-% autotracing
-
-% CAPS LOCK, LOCK, CAPTAIN'S LOCK
-
-% when two letters coincide, a shift-reduce conflict
-
-% cool s
-
-\section{To infinity, but let's stop there}
 
@@
 % ... so it's like a 3D manifold
+
+\begin{figure}[t]
+\centering
+  \includegraphics[width=0.7 \linewidth]{uppestcase3d}
+\caption{ 3D manifold showing a section of the repeating loop as the
+  \makeuppercase\ model iteratively uppercases a letter. (Shown here
+  is the input \lowercase{q} from iteration 245--377, but they all
+  converge to this same periodic shape.) Slices through this shape give a
+  letterform's outline (or usually, a linear interpolation between two
+  of them). The bottom of the shape is its ``beginning'' but it
+  appears to repeat like this forever.
+} \label{fig:uppestcase3d}
+\end{figure}
+
+% bristol stool chart
 

### 2021-03-20 lowercase/trace-sdf.cc

@@ -217,7 +217,7 @@ int main(int argc, char **argv) {
     }
   }
   Timer sdf_timer;
-  ImageA bitmap_sdf = FontProblem::SDFFromBitmap(SDF_CONFIG, input);
+  ImageF bitmap_sdf(FontProblem::SDFFromBitmap(SDF_CONFIG, input));
 
   FontProblem::Image8x8 pix;
   for (int i = 0; i < 64; i++) {
@@ -238,7 +238,7 @@ int main(int argc, char **argv) {
          "Num edges: %d\n",
          pix.PixelsOn(),
          pix.Edges());
-  ImageA pixel_sdf = FontProblem::SDF36From8x8(pix);
+  ImageF pixel_sdf(FontProblem::SDF36From8x8Uppercase(pix));
 
   printf("Took %.3f sec\n", sdf_timer.MS() / 1000.0);
 
@@ -259,7 +259,7 @@ int main(int argc, char **argv) {
   FontProblem::Gen5Result gen5result =
     FontProblem::Gen5(SDF_CONFIG, *make_lowercase, *make_uppercase, vector_sdf);
 
-  const ImageF &trace_sdf = gen5result.low;
+  const ImageF &trace_sdf = bitmap_sdf; // gen5result.low;
 
   ImageRGBA islands;
   Timer vectorize_timer;
@@ -282,7 +282,7 @@ int main(int argc, char **argv) {
   font.baseline = FontProblem::TTFBaseline(SDF_CONFIG);
   font.chars = {{'e', ttf_char}};
 
-  Util::WriteFile("trace.sfd", font.ToSFD("Traced"));
+  Util::WriteFile("trace.sfd", font.ToSFD("Traced", "Fake"));
 #else
 
   TTF::Char ttf_char;

### 2021-03-21 trunk/lowercase/paper/lowercase.tex
Wrote a bunch more...

@@
 \usepackage{float}
-\usepackage{url}
+
+% \usepackage{url}
+% \usepackage{xurl}
+\usepackage[pdftex]{hyperref}
+
 % \usepackage{code}
@@
 
+\interfootnotelinepenalty=0
 \setlength{\footnotesep}{2em}
 
-% \newcommand\comment[1]{}
 \newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}
 
+% XXX the collection of crap above causes a first page
+% with just "1" on it (probably conflicting packages). This
+% magic just drops the first page, which is of course not
+% the right solution, but...
+\usepackage{atbegshi}% http://ctan.org/pkg/atbegshi
+\AtBeginDocument{\AtBeginShipoutNext{\AtBeginShipoutDiscard}}
+
+\newcommand\upsigbovik{
+  \includegraphics[height=1em]{upsigbovik}
+}
+
 \begin{document}
 
-\title{Lowestcase and Uppestcase letters: Further adventures in Derp Learning}
+\title{Lowestcase and Uppestcase letters: Advances in Derp Learning}
 \author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
 Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
-Appears in SIGBOVIK~2021 with the
+Appears in \upsigbovik~2021 with the
 OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
@@
 takes extra work to press the caps lock key, press the first letter of
-their name, and then press ther caps lock key again to turn it off).
+their name, and then press the caps lock key again to turn it off).
 In German, Nouns start with uppercase Letters, signifying Superiority
@@
 
+% shift-reduce conflict
+% Shift is often used to increase conflict, not reduce it
+
 \subsection{Induction}
@@
 returns a letterform (we can choose how these are represented) and
-does the needful, e.g. \makelowercase(\letterform{A}) should return
+does the needful, e.g.~\makelowercase(\letterform{A}) should return
 \letterform{a}. In order to learn this function, we'll at least need a
@@
 
+% no copyright intended
+
 Sprechen of Fonts, I downloaded every font I could find on the whole
@@
 Next, my lord, do people have bad taste! And I say this as someone who
-made dozens of amateurish fonts\cite{dbzfonts} as a high school and
+made dozens of amateurish fonts~\cite{dbzfonts} as a high school and
 college student and who is contributing several new questionable
@@
 also categorizing them as serif, sans-serif, decorative, techno, etc.,
-which I never used). In doing this I noticed another extremely common
-problem, which was that many fonts had the same letter shapes for
-uppercase and lowercase letters. This would not do for the current
-application!
-
-\begin{figure*}[ht]
-\includegraphics[width=0.9 \linewidth]{sortition}
+which classifications I never used). In doing this I noticed another
+extremely common problem, which was that many fonts had the same
+letter shapes for uppercase and lowercase letters. This would not do
+for the current application!
+
+\begin{figure}[tp]
+\centering
+  \includegraphics[width=0.98 \linewidth]{sortition}
 \caption{ The interactive font data-cleaning UI. A seemingly endless
@@
 } \label{fig:sortition}
-\end{figure*}
+\end{figure}
 
@@
 techniques are generally those like gradient descent. (In fact, the
-backpropagation algorithm we use to training the neural network in
+backpropagation algorithm we use to train the neural network in
 Section~\ref{sec:neural} is gradient descent on the function that
@@
 node.) The problem with this is that you need to do some math to
-compute the gradient function, and anyway you need to deal with fiddly
-bits (Section~\ref{sec:fiddly}) unless the function is convex and
-smooth, which it will not be. If you don't want to deal with that,
-and have a fast computer (and who doesn't?), black-box optimization
-algorithms are worth considering. Here, the interface\footnote{
-  Here a simplfied wrapper around BiteOpt~\cite{biteopt} in my {\tt cc-lib}
-  library. See \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/cc-lib/opt/}.}
+compute the derivative of the function, and anyway you need to deal
+with fiddly bits (Section~\ref{sec:fiddly}) unless the function is
+convex and smooth, which it will not be. If you don't want to deal
+with that, and have a fast computer (and who doesn't?), black-box
+optimization algorithms are worth considering. Here, the
+interface\footnote{ Here a simplfied wrapper around
+  BiteOpt~\cite{biteopt} in my {\tt cc-lib} library. See
+  \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/cc-lib/opt/}.}
 is just something like (C++):
@@
 
-\begin{figure}[ht]
+\begin{figure}[tp]
+\centering
   \includegraphics[width=0.9 \linewidth]{casealignment}
@@
   $1.7\%$ difference. Note that the shapes are still not an exact
-  match (probably errors introduced in the font export process, which
+  match (probably noise introduced in the font export process, which
   has to round the data to integers and might apply other non-linear
@@
 letterforms fit within that budget. Letterforms can be made of
-multiple contours (e.g. a stacked \letterform{g} typically has two
+multiple contours (e.g.~a stacked \letterform{g} typically has two
 holes in it, and \letterform{j} has two disjoint parts). I found that
 most clean fonts had three or fewer contours, and when sorting them by
-descending length, most fit within 38, 14, and 10 endpoints for the
-three. So, I only train on fonts where all of the letters fit within
-this budget.\footnote{It would not be a good idea to reject only the letters
-that don't fit, because it might result in the network being trained
-on more \letterform{l}s (tends to be simple) than \letterform{g}s
-(tends to be complex).}
+descending length, basically all of them fit within 100, 25, and 16
+endpoints for the three. So, I only train on fonts where all of the
+letters fit within this budget.\footnote{It would not be a good idea
+  to reject only the letters that don't fit, because it might result
+  in the network being trained on more \letterform{l}s (tends to be
+  simple) than \letterform{g}s (tends to be complex).}
 
@@
 function (which might not be the case if we were predicting a dynamic
-number of points). Finally, I output the contour's points starting
-from the point closest to the origin in order to reduce one needless
-degree of freedom.
-
-The network I trained has an input layer size of $154 = (38 + 14 + 10)
+number of points).
+
+The network I trained has an input layer size of $570 = (100 + 25 + 16)
 \times 4 + 3 \times 2$ (one control point and one end point per
@@
 for just about 1 million total parameters. All layers are leaky
-rectified linear units. If you're taking notes, don't, as again this
+rectified linear units ({\tt x > 0 ? x : 0.1 * x}), which is fast to
+compute and better than sigmoids in the output since correct values
+will not just be 0 and 1. If you're taking notes, don't, as again this
 does not work well, and I don't know how people figure out what the
@@
 place because C is the third letter. We don't need these outputs for
-the problem itself (e.g. to lowercase new letter shapes), but there
+the problem itself (e.g.~to lowercase new letter shapes), but there
 are several ideas behind this. First, the lowercasing function we're
-trying to learn does take into account the letter of the alphabet
-being lowercased. By asking the network to learn this as well (it is
-penalized when it gets the prediction wrong), it must learn features
-that allow it to distinguish different letters, and those features are
-available for use by outputs we {\em do} care about. This is an easy
-way to coax it to learn features that I know are meaningful without
-having to actually engineer feature extractors by hand (or first train
-separate models, etc.). Perhaps the most useful thing is that it's
-very clear what the right answer is, so it gives me an easy way to see
-if the network is learning anything {\em at all}. (It does.) Finally,
-we can do some silly stuff with these; see
-Section~\ref{sec:hallucination}.
+trying to learn does depend on the letter of the alphabet being
+lowercased (in an extreme case, consider the lowercase-L
+\letterform{l} and the uppercase-i \letterform{I}, which look the same
+in many fonts but have different lowercase letterforms). By asking the
+network to learn this as well (it is penalized when it gets the
+prediction wrong), it must learn features that allow it to distinguish
+different letters, and those features are available for use by outputs
+we {\em do} care about. This is an easy way to coax it to learn
+features that I know are meaningful without having to actually
+engineer feature extractors by hand (or first train separate models,
+etc.). Similarly, I could have asked it to predict whether the font
+is italic, serif, the character's width, or anything else I have on
+hand. Perhaps the most useful thing is that it's very clear what the
+right answer is, so it gives me an easy way to see if the network is
+learning anything {\em at all}. (It does.) Finally, we can do some
+silly stuff with these; see Section~\ref{sec:hallucination}.
 
@@
   or a K, and has predicted a shape sort of ambiguously between those
-  two. The \lowercase{z} is also an embarrassment.
+  two. The \letterform{z} is also an embarrassment.
 } \label{fig:trainingvector}
@@
 of the shapes it generates are along the right lines. ({\em Along the
-  right lines}, get it??) I did not feel ANGRY at these results because
-I expected it to not really work.
-
-\paragraph{Just try making it more complicated!} This
-problem of predicting the vector shape directly is a lot to ask of a
-neural network, at least set up this way. One thing that did not sit
-well with me is that the network could generate a perfect-looking
-result, but because it didn't have the points in the expected order,
-it would be penalized. This makes it harder to learn. True, but
-my reflex to make things more complicated XXX
-
-Ideally we would compute the difference between the predicted shape
-and the expected shape, regardless of how they're drawn. Aside from
-being a bit computationally challenging, this won't really work
-because we need to attribute error directly to each output in order to
-actually update the model in training. So I spent an afternoon writing
-a routine to compute the best alignment of points between the actual
-and expected outputs (Figure~\ref{fig:rotatej}). Aside from being
-harder than it looked, my alignment code ended up being pretty slow
-relative to the rest of training, even worse since it ran on the CPU
-instead of GPU, and the training speed became UPSETTINGLY SLOW. So, I
-discarded this excursion and threw this whole endeavour in the trash!
-But I must have confused the Recycle~Bin icon with the fairly
-complicated export-to-Truetype Font process that I built, because I
-ran the model on the venerable Futura~\cite{futura} font and generated
-{\bf Futurda} (Figure~\ref{fig:futurda}), which can be downloaded from my
-website and installed for your desktop publishing needs.\footnote{Font
-  downloads are available at \url{http://tom7.org/lowercase/}.}
-
-% TODO: actually upload the fonts and make sure the URL is what you want
-
-\begin{figure}[ht]
-\begin{centering}
-  \includegraphics[width=0.5 \linewidth]{rotatej}
-\end{centering}
-\caption{ Example alignment (from debug UI) of predicted shape (blue)
-  to expected shape (green). We require each point to be mapped (red)
-  to a point from the expected contour in a monotonic order (but
-  several can be mapped to the same one), so that we can attribute
-  error to each point. I hope it was fun to write this code because
-  it was too slow to do pretty much anything but generate this figure!
-} \label{fig:rotatej}
-\end{figure}
-
-% FIXME: Wait I had much better screenshots, see video/lowercase
-
-% XXX add downloadable font too
+  right lines}, get it??) I did not feel ANGRY at these results
+because I expected it to not really work. Still, it ``has output'' and
+so it can be used to generate a font. I made every glyph in Comic Sans
+MS~\cite{comicsans} lowercase using the model (with the exception of the
+\letterform{{\sf \%}} character, which has too many contours---{\it five}!).
+Mostly this model produces small, non-confident scrawls, like little
+grains of sand, so this font is called {\bf Comic Sands}
+(Figure~\ref{fig:comicsands}). The TrueType version can be downloaded
+from my website and installed for your desktop publishing
+needs.\footnote{Font downloads are available at
+  \url{http://tom7.org/lowercase/}.}
+
 \begin{figure*}[tp]
-\begin{centering}
-  \includegraphics[width=0.9 \linewidth]{futurda}
-\end{centering}
-\caption{ Type specimen for the generated font {\bf Futurda}. This is the
-  classic font Futura, run through the vector based lowercasing model
-  (Section~\ref{sec:vectorversion}). At top are Futura's letterforms
+\centering
+  \includegraphics[width=0.95 \linewidth]{comicsands}
+\caption{
+  Type specimen for the generated font {\bf Comic Sands}. This is
+  the hateful Comic Sans MS run through an early vector-based lowercasing
+  model (Section~\ref{sec:vectorversion}). At top are Comic Sans's letterforms
   \letterform{A}--\letterform{Z} run through the model and so ``made
@@
   made even more lowercase. Also rubbish.
-  At the bottom are the pangrams ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
-  and ``Sphinx of black quartz, judge my vow!'' Although the text is
-  illegible and kind of uninteresting, it does have a certain dynamic
+  At the bottom are the illegible pangrams
+  ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
+  and ``Sphinx of black quartz, judge my vow!'' Although the output
+  barely resembles letters, it does have a certain wispy
   Rorschach aesthetic, like a collection of delicate moths pinned
   to paperboard, that one could consider framing or publishing in
-  the proceedings of SIGBOVIK 2021.
+  the proceedings of SIGBOVIK 2021. It is certainly an improvement on
+  the original font.
+} \label{fig:comicsands}
+\end{figure*}
+
+\subsection{Just try making it more complicated!} \label{sec:complicated}
+
+This problem of predicting the vector shape directly is a lot to ask
+of a neural network, at least set up this way. One thing that did not
+sit well with me is that the network could in principle generate a
+perfect-looking result, but because it didn't have the points in the
+expected order, it would be penalized. This makes it harder to learn,
+and more prone to overfitting.\footnote{For example, imagine if the database
+contains two versions of Helvetica that just have their points in a
+different order---which is very likely the case btw---the model will
+have to learn how to distinguish between these, but using information
+we just don't care about.} This was one case where my questionable
+reflex to make things more complicated did pay off!
+
+First, I reduced the number of points in the input and output.
+Reducing the dimension of the function being learned generally makes
+learning a lot faster. This had the side-effect of reducing the number
+of eligible fonts (by about half), and by nature these fonts are
+simpler shapes. These effects alone could be responsible for the
+improved performance of this second try.
+
+I also output each contour's points in a normalized order, starting
+from the point closest to the origin. This removes one needless degree
+of freedom.\footnote{We can see (well, it's not pictured since I have
+  far exceeded a reasonable number of figures in this paper, but {\em
+    I} can see) how this manifests in the biases on the output layer,
+  which are a proxy for the ``average prediction''. In the first
+  model, because of the unstructured order, these are mostly near 0.5
+  (center of the character) or 0.0 (degenerate, unused contours). In
+  this new model, the distribution of biases is much more flat; it can
+  learn that ``the first point tends to be near 0.25,0.25'' and ``the
+  seventh point tends to be near 0.64,0.3.''}
+
+Aside from the changes in the input (now 254 nodes) and output (280),
+this second version has three sparse hidden layers of size 508, 508,
+and 560 nodes; the first two are dense and the latter sparse. The
+final model after some pruning had 609k parameters.
+
+As this was training, I worked on another improvement. Ideally we
+would compute the difference between the predicted shape and the
+expected shape, regardless of how they're drawn. Aside from being a
+bit computationally challenging, this won't really work because we
+need to attribute error directly to each output in order to actually
+update the model in training. I spent a valuable vacation day writing
+a routine to compute the best alignment of points between the actual
+and expected outputs (Figure~\ref{fig:trainingrotate}). Aside from being
+harder than it looked, my alignment code ended up being pretty slow
+relative to the rest of training, even worse since it ran on the CPU
+instead of GPU, which reduced the training speed by $50\%$. I let it
+run for 80,000 rounds, some 331 million training examples, but
+eventually got bored of waiting on this approach that was slow to
+train and seemed like a complicated version of an bad, oversimplified
+approach. So, I control-C'd that thing and threw this whole endeavour
+in the trash! But I must have confused the Recycle~Bin icon with the
+fairly complicated export-to-TrueType Font process that I built,
+because I ran the model on the venerable Futura~\cite{futura} font and
+generated {\bf Futurda} (Figure~\ref{fig:futurda}).
+%
+% \begin{figure}[ht]
+% \centering
+%   \includegraphics[width=0.5 \linewidth]{rotatej}
+% \caption{ Example alignment (from debug UI) of predicted shape (blue)
+%   to expected shape (green). We require each point to be mapped (red)
+%   to a point from the expected contour in a monotonic order (but
+%   several can be mapped to the same one), so that we can attribute
+%   error to each point. I hope it was fun to write this code because
+%   it was too slow to do pretty much anything but generate this figure!
+% } \label{fig:rotatej}
+% \end{figure}
+%
+
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.95 \linewidth]{trainingrotate}
+\caption{ Screenshot (somewhat compacted) of training from near the
+  final round of the vector model's training, illustrating the
+  permissive loss function that finds the best alignment. At the bottom are
+  the predicted lowercase shapes (blue), also shown with their
+  expected shape (green). We require each point to be mapped (red)
+  to a point from the expected contour in a monotonic order (but
+  several can be mapped to the same one), so that we can attribute
+  error to each point.
+} \label{fig:trainingrotate}
+\end{figure}
+
+% XXX downloadable
+
+\begin{figure*}[tp]
+\centering
+  \includegraphics[width=0.95 \linewidth]{futurda}
+\caption{ Type specimen for the generated font {\bf Futurda}. This is
+  the classic font Futura, run through the final, improved
+  vector-based model (Section~\ref{sec:complicated}) to make each
+  letter lowercase. The letterforms \letterform{A}--\letterform{Z}
+  (top) become quite readable lowercase versions. The extra-lowercase
+  \letterform{a}--\letterform{z} are also almost legible, but are
+  mostly just scaled-down and screwed up versions of the lowercase
+  letterforms. Could definitely imagine this appearing in the
+  ``distressed fonts'' category of a 10001 Free TrueType Fonts CD-ROM
+  in the 1990s, though.
 } \label{fig:futurda}
@@
 anti-aliased value in-between. In an SDF, the bitmap instead contains
-the distance to the nearest edge, and is signed (e.g. values inside
+the distance to the nearest edge, and is signed (e.g.~values inside
 the shape are $> 0$, value outside are $< 0$). Actually in
@@
 \begin{figure}[ht]
+\centering
   \includegraphics[width=0.95 \linewidth]{sdf-figure}
@@
   far superior despite being derived from a $36 \times 36$ pixel
-    bitmap. Typically this process is performed at an even higher
+  image. Typically this process is performed at an even higher
   scale and then downsampled to produce an anti-aliased image.
@@
 often hangs to the left of the origin as well). I used an ``on-edge
-value'' of 220 (because much more of the SDF is outside the letter
-than inside) and the distance is scaled as 15 units per pixel (so that
-pixels on the outer edge often have non-zero values). Compared to the
-first version, I was somewhat more permissive in what fonts I trained
-on, since there was no inherent limit to the number of contours their
-complexity. I did exclude fonts whose rasterizations exceeded the
-bounds of the SDF, which is possible (very wide \letterform{W} or
-low-descending \letterform{j} perhaps) but rare.
+value'' of 0.862 (because much more of the SDF is outside the letter
+than inside) and the distance is scaled as 0.059 units per pixel
+(chosen so that pixels on the outer edge often have non-zero values).
+Compared to the first version, I was somewhat more permissive in what
+fonts I trained on, since there was no inherent limit to the number of
+contours or their complexity. I did exclude fonts whose rasterizations
+exceeded the bounds of the SDF, which is possible (very wide
+\letterform{W} or low-descending \letterform{j} perhaps) but rare.
 
@@
 % It's responsible for the XXX
-Every layer's transfer function is ``leaky relu'' ({\tt x > 0 \? x : 0.1 * x})
-which is fast to compute and better than sigmoids in the output since
-correct values will not just be 0 and 1. It would definitely make sense
-to use convolutional layers for this problem, as features like serifs,
-lines, curves, and so on could appear throughout the input and output.
-I just haven't built support for that in my weird home-grown software,
-yet.
+Every layer's transfer function is ``leaky relu'' again. It would
+definitely make sense to use convolutional layers for this problem, as
+features like serifs, lines, curves, and so on could appear throughout
+the input and output. I just haven't built support for that in my
+weird home-grown software, yet.
 
@@
 \makeuppercase\ and \makelowercase\ models simultanously. Two models
-fit comfortably in GPU memory, with plenty of space for a stream of
+fit easily in GPU memory, with plenty of space for a stream of
 training data (one training instance is only about 10kb). The only
@@
 ``inversion'' examples (Figure~\ref{fig:cotraining}), but this is just
-a matter of programming, thank God. I should remember to do projects
+a matter of programming, thank god. I should remember to do projects
 that are mostly a matter of programming. Each round, $25\%$ of the
 batch consists of inverted examples from the symmetric model's output
-from a recent round. I make sure that one model is not allowed to
-get too far ahead (more than 2 rounds) of the other in training,
-because I want this feedback loop to be somewhat tight.
+from a recent round. Training happens asynchronously, but I make sure
+that one model is not allowed to get more than 2 rounds ahead of the
+other, because I want this feedback loop to be somewhat tight.
 
@@
 several times with different parameters and initialization weights
-because it would get stuck
+because it would get stuck (Figure~\ref{fig:nans}) right away or as
+soon as I looked away from the computer. I prayed to the dark wizard
+of hyperparameter tuning until he smiled upon my initial conditions,
+knowing that somewhere he was adding another tick-mark next to my name
+in a tidy but ultimately terrifying Moleskine notebook that he bought
+on a whim in the Norman Y.~Mineta San Jose International Airport on a
+business trip, and still feels was overpriced for what it is.
+
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.9 \linewidth]{nans}
+\caption{ Divergent training after only 29 rounds. We have \nan{}
+  total error (hard to say if that's good or bad?). The example in
+  column one is an inversion example generated by the
+  \makeuppercase\ model, which is why it also looks like the Tunguska
+  event, just of the opposite sign. The other two are regular inputs,
+  whose predicted outputs are black holes. Start over!
+} \label{fig:nans}
+\end{figure}
+
+\subsection{Fiddly bits} \label{sec:fiddly}
+The training error over time appears in
+Figure~\ref{fig:sdfmergederror}. It looks like the ones I have seen in
+machine learning papers, although I don't like to read other people's
+papers because it just seems like spoilers, and reading is the
+opposite of writing! There are several noticeable events in the curve,
+which came from me fiddling with the parameters or network as it ran.
+Here are some of the things I did:
+
+\paragraph{Vacuuming and culling.} Sometimes a node will just be
+dead (basically never activates) or an edge weight will be nearly
+zero. In these cases an equivalent, tidier network can be made by
+dropping the node or edge. Periodically I would perform these
+processes, sometimes feeling particularly choppy and removing like
+10\% of the parameters at a time. If these parameters are truly
+useless with no hope of recovery, we simply get faster training
+because there's less work to do. Speed is exhilarating!
+
+\paragraph{Widening.} The opposite thing is to introduce new nodes.
+Adding nodes to hidden layers is pretty easy. The thing that worked
+best for me is to increase the size of the layer by 10--15\%, where
+each new node has random incoming weights and bias 0. Then for each
+node on the next layer, I add edges to some subset of these new nodes
+(again generally 10\% of them) with weight 0. Since this weight is
+zero, the network computes the same function, but has new gradients
+to explore (in practice, it then experiences some shock after a
+few training rounds, but then quickly fine-tunes this away). More
+parameters means slower training, but also more potential to learn
+interesting functions, or overfit! Danger is exciting!
+
+\paragraph{Deepening.} It's also possible to add layers once the
+network is trained. This can be done anywhere, but I liked doing it on
+the output layer because this gets the most direct feedback from the
+training examples, and so it updates quickly and changes there are
+easy to understand. If you append the identity matrix (new layer is
+the same size as the previous; each node has weight 1.0 to its
+corresponding node and 0.0 elsewhere) then this network computes
+the same function but has new gradients to explore. Adding a layer
+did seem to help unlock a new training regime
+(Figure~\ref{fig:sdfmergederror}); subjectively it also reduced
+some weird artifacts in the SDFs that the model used to predict (makes
+sense; this most natural thing for this layer to do is learn how to
+predict a ``correction'' from the old prediction, for example
+by smoothing/sharpening it). This seems to be borne out by the
+weights, which are also fun to look at (Figure~\ref{fig:lastlayer}).
+All problems in computer science can be solved by an additional
+layer of indirection!
+
+\paragraph{Generating features.} On the other side, randomly sampling
+pixels from the input SDF does work, but I superstitiously believed
+that it might be better to have more spatially meaningful features. I
+wrote a program to generate a bunch of random simple features (one
+line/blob with positive weights, one line/blob with negative weights).
+It then chooses a set of them that are both {\em good} (maximum
+standard deviation on a sample of training data) and {\em different
+  from one another} (redundant or even partially redundant features
+are less valuable). It was nice to satisfy my supersition, and the
+dark wizard of superstitious fiddling with neural networks in the
+hope that they do the thing was pleased as well. The features are
+at least handsome (Figure~\ref{fig:makefeatures}). Creativity is
+enriching!
+
+I presume these are all standard things that neural people do, but they
+do better and smarter versions of them because they are willing to
+read other people's papers instead of trying to figure things out from
+scratch all the time. But you gotta occupy yourself somehow while it
+crunches for a month.
+
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.99 \linewidth]{makefeatures}
+  \caption{ Some randomly-generated features with the selected ones
+    outlined in magenta, mostly shown here for aesthetic reasons.
+    Savvy Twitter user {\tt @iotathisworld} sees this as ``the classic
+    question: machine vision classification or 90s roller rink carpet
+    pattern?'' to which I deflect: ``Sorry, it's actually modern day
+    Port Authority bus upholstery or Gram stain of same!'' (But
+    actually machine vision classification was basically correct.)
+  } \label{fig:makefeatures}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.99 \linewidth]{sdfmergederror}
+\caption{ The training error for the SDF models. The red curve is the
+  \makeuppercase\ model, which generally has a higher error rate
+  (perhaps simply because uppercase letters usually have more pixels
+  set) and blue is \makelowercase. The first few rounds have error
+  that's off the charts, well above 100. The most dramatic event is
+  around round 200,000, where I reduced the weight decay factor to
+  $0.999995$ (from $0.9995$). I guess you just need more nines to be
+  more reliable. There are some other visible peaks, which occur when
+  I do things like remove nodes with very low weights or which are
+  almost never activated (Section~\ref{sec:fiddly}). These momentarily
+  increase error but it is easily fine-tuned away (e.g.~by learning
+  new biases). The peak at around 1.4M rounds is when I added a new
+  layer to the end of the model, which does seem to create a new
+  training regime (clear downward slope now); but this also
+  significantly increases the training cost per round. Even after
+  2,000,000 rounds, the network is still apparently improving, but at
+  a speed of about 1 pixel loss per several weeks. Eventually the
+  extremely strict SIGBOVIK deadlines mean you just have to call it done.
+} \label{fig:sdfmergederror}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+  \includegraphics[width=0.99 \linewidth]{lastlayer}
+  \caption{ The bottom-right corner of the weight matrix for the final
+    layer of the network. This layer was added to the network after
+    1.36 million rounds, initially as the identity matrix, and so can
+    be thought of partly as a correction of the network's output prior
+    to that round (though the remainder of the network continues to
+    evolve). The x-axis is the output nodes, and the y-axis is the
+    nodes of the previous layer. Note for example that the last 26
+    columns look pretty different; these are the predictors for the 26
+    letters, which occur in the output after the SDF pixels. Green
+    means positive and red means negative, so if you are looking at
+    this in a black-and-white printout, that may explain your current
+    confusion. The exact diagonal is a strong green, close to 1.0,
+    although over time these weights do diverge from the identity
+    somewhat. In the bottom-right corner of size $26^2$, we are
+    looking at how the 26 letter predictors are derived from the
+    previous layer's predictions. We see that most letters are
+    negatively correlated (makes sense; only one will ever be 1.0)
+    although there are some oddities (probably because it found some
+    other, better correlates). Nodes on this new layer have dense
+    references to these 26 predictions on the previous; this means
+    that the bottom row kind of represents biases for each of the 26
+    letters (what does an average 'e' look like?). I also included a
+    dense region above that, but this appears to have simply evolved
+    the same way as other $36^2$ chunks have (the rest are sparse).
+    These chunks have a large amount of spatial similarity (suggesting
+    that the sparse sampling would be adequate), with a meaning like
+    ``if this area of the image is bright, then this pixel should
+    be less bright.'' It is interesting that the pixels immediately
+    next to the diagonal are almost always strongly negative
+    (sharpening operation). ``Thank you for attending my TED talk.''
+    ---\,Figure~\ref{fig:lastlayer}}
+  \label{fig:lastlayer}
+\end{figure}
+
+For completeness, some other innovations that I feel are worth
+mentioning:
+
+Making the GPU code faster has really high value (could save weeks of
+waiting). Since I am using OpenCL (whoa, yeah, stop me right there, I
+know) I found a good technique was to generate different OpenCL code with
+constants baked for each layer (for example their size and
+\verb+indices_per_node+); this allows the compiler to use faster
+tricks in inner loops for e.g.~multiplication by a compile-time constant
+instead of depending on an argument or data. I have different routines
+for sparse and dense layers. It might even make sense to recompile the
+kernels for other parameters that change over the lifetime of
+training, like the learning rate. The {\tt fma} instruction (so named
+for the physical law $F=MA$) is a bit faster than
+\verb|potential += w * v|, and I guess the compiler can't do this
+itself because of IEEE horrors. But like, who cares? In my opinion you
+should be able to put it in ``fast machine learning mode'' where it
+readily makes precision errors, with a command-line option like
+\verb+--fml+. With all the tweaking, the easiest win was to use the
+{\tt restrict} keyword on arguments to tell the compiler that the
+input cannot alias the output, for example; this presumably helps
+it schedule instructions better.
+
+Various things in training run in parallel threads (e.g.~processing
+fonts, but also moving data to the GPU, backpropagation for each
+example, etc.). For a long time I had just been explicitly setting
+parallelism using superstitious constants. For this project I finally
+just wrote something that would automatically and empirically
+determine the number of threads that yielded the highest throughput,
+and persisted that information across program starts. This was a
+good idea and enters the permanent rotation.
+
+The actual error on the predicted SDFs is pretty low; for the
+\makelowercase\ model it is around 31.3, which is like if 2.4\% of the
+pixels were (completely) wrong, but the rest is exactly correct. In
+reality, of course, the error is distributed throughout the pixels,
+and some errors are a lot more important than others. Particularly,
+near the threshold value, a pixel goes from from being considered ``in
+the letter'' to ``outside'' with tiny changes in its value. Changes to
+a pixel with a value near 0.0 or 1.0 usually doesn't affect the output
+shape at all, in contrast. So one thing I did was map the loss
+function (comparing expected pixel value to actual) to ``stretch out''
+the region near the threshold, increasing the penalty (basically, the
+derivative) in that region and decreasing it elsewhere. Looking at the
+code again right now, I realize that I only applied this to the first
+row of the SDF (\verb+idx < SDF_SIZE+ instead of
+\verb+SDF_SIZE * SDF_SIZE+), so that was dumb AND MAKES ME {\bf
+  ANGRY}. I will say in my defense that at least I felt disappointed
+at the time that it didn't seem to make a difference! (The dark
+wizard of superstitious fiddling nods sagely.
 
@@
 
-\label{sec:fiddly}
-
-% TODO loss over time
-
-% some other things that are worth a mention:
-%   autoparallel
-%   efficiency improvements (restrict, compile-time constants)
-
-% no copyright intended
-
-% stretch the loss function near the onedge value
-
+Ultimately, each of the two models was trained for over 2 million
+rounds, which corresponds to 510 million training examples. Each
+model is about 24 megabytes.
+
+\subsection{Upperercase and Lowerercase fonts} \label{sec:fonts}
+
+Now that we have these expensive models, we can use them to make
+arbitrary letterforms uppercase or lowercase. The output is readily
+rasterized (using the standard thresholding approach for SDFs) but
+we'd actually like to have vector representations so that we can
+use them as TrueType fonts for our desktop publishing needs.
+
+\subsection{Tracing}
+Automatically tracing bitmaps into vector form is no doubt a solved
+problem, but I chose not to look at spoilers. Since we actually have a
+signed distance field, we can build a tracing routine directly off of
+that. The approach I took consists of three steps. First, I generate a
+bitmap of the SDF (at its native size) using the threshold. I separate
+the image into a nested tree of connected components in this pixel
+space; each component knows its single parent and whether it is
+``land'' (inside the shape) or ``sea''. Characters like \letterform{e}
+need internal cutouts, which are represented by a different winding
+order (clockwise or counter-clockwise) for the contour. Some of the
+tricky cases in computing this tree structure are given in
+Figure~\ref{fig:connected}. Once I have this tree structure, I trace
+each pixel mass recursively (Figure~\ref{fig:trace}). I find a pixel
+on the edge, and then walk around that edge clockwise (simple case
+analysis on the three pixels ahead of me). As I walk the edge, I look
+at the normal (orthographic as we are doing orthography) of the edge
+and see where it reaches the edge value on the SDF; this point (a
+float) is output into the contour. The process is guaranteed to return
+to where we started. I recurse by negating the SDF (outside
+becomes inside) and bitmap (land becomes sea), and reverse the winding
+order of the result of recursion.
+
+This gives me a perfectly fine line-based outline of the SDF's shape.
+Since I output points at every pixel, sometimes these points are
+inefficient (e.g.~a series of colinear points on a straight line),
+and sometimes they reflect sharp corners that are not aesthetic. So
+I then take a second pass at each contour, and try fitting B\'ezier
+curves to sequences of points while the error remains low. Again I
+did fitting with a black-box optimizer, which is nice. However,
+the function being minimized also needs to be able to find the
+closest point on a B\'ezier curve to another point, and although
+this can also be done easily with the black-box optimizer, nesting
+an optimizer invocation inside another one proved to be way too slow.
+I found an old algorithm in a book I owned and was stymied by as a
+child~\cite{glassner1990graphics}.
+
+\begin{figure}[ht]
+  \centering
+\includegraphics[width=0.6 \linewidth]{connected}
+\caption{
+  Some tricky cases to think about when generating the nested
+  connected components, as the first step of tracing SDFs. Area
+  {\bf 0} is the outside of the entire letterform, but note that
+  we should include the top-left corner even though it is not
+  reachable without leaving the bounds of the image. Area {\bf 1}'s
+  parent is 0; it has two holes within it, Areas {\bf 2} and {\bf 3}.
+  At the top left, the four pixel chunks making up area {\bf 4} are
+  not actually connected, but they separate the hole which is
+  child are {\bf 5}. This hole must have one parent, so it means
+  that all four pixel chunks are part of the same area {\bf 4}.
+} \label{fig:connected}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+\includegraphics[width=0.95 \linewidth]{trace}
+\caption{ Tracing the SDF from Figure~\ref{fig:sdf} into vector
+  format. Left image shows the nested connected components. Middle
+  image is the initial straight-line trace, and the right image
+  shows the simplified contours using quadratic B\'eziers.
+} \label{fig:trace}
+\end{figure}
+
+\begin{figure*}[tp]
+\centering
+\includegraphics[width=0.9 \linewidth]{heavenica} \\[0.5in]
+\includegraphics[width=0.9 \linewidth]{spezialhellvetica} \\
+\caption{
+  Type specimens for the generated font {\bf Heavenica} (top)
+  and {\bf Spezial Hellvetica} (bottom). The uppercase letters
+  in Heavenica are \makeuppercase\ applied to uppercase letters
+  from Helvetica, and the lowercase are \makelowercase\ applied
+  to those. These lower-upperercase letters resemble regular
+  uppercase letters, as they should; this gives you some idea
+  of the quality of the model. Spezial Hellvetica is the
+  symmetric thing (its lowercase letters are \makelowercase\ 
+  of Helvetica's lowercase). The sample text in this latter
+  case is ``Quartz jock vends BMW glyph fix. Twelve ziggurats
+  quickly jumped a finch box.''
+} \label{fig:heavenica}
+\end{figure*}
+
+\begin{figure*}[tp]
+\centering
+\includegraphics[width=0.9 \linewidth]{expnewroman} \\[0.5in]
+\includegraphics[width=0.9 \linewidth]{plusnewroman} \\
+\caption{
+  Type specimens for the generated font {\bf Exponential New Roman} (top)
+  and {\bf Plus New Roman} (bottom). These were produced with
+  the same procedure as in Figure~\ref{fig:heavenica}, but starting
+  with Times New Roman. The letterforms are clearly different, so
+  it's not as though the models are (just) memorizing a shape for
+  each letter. Notably, smudgy serifs reappear when the upperercase letters
+  are re-lowercased, as desired.
+  Sample text here is ``Amazingly few discotheques provide jukeboxes.
+  Those that don't MAKE ME QUITE ANGRY.'' and 
+  ``By Jove, my quirky study of lexicography won a prize! (the prize
+  was a crappy font)''
+} \label{fig:roman}
+\end{figure*}
+
+\medskip
+Now we're all set up to take an input shape (e.g.~from an existing
+font), run the \makeuppercase\ or \makelowercase\ model(s) on it,
+maybe multiple times, and trace the resulting SDF into a vector form
+that can be used in a font. I did this on the canonical sans serif
+font Helvetica~\cite{helvetica} and serif font Times New
+Roman~\cite{timesnewroman}.
+
+Helvetica means ``of Hell'', so making the font more uppercase give us
+{\bf Heavenica} (Figure~\ref{fig:heavenica}), since Heaven is ``up''
+from Hell. What's lower than Hell? Spezial Hell,\footnote{
+  I first learned about Spezial Hell from a Rugen Br\"au beer that I
+  drank in the Alps in Grindelwald, Switzerland ({\it la
+    Conf\'ed\'eration H\'elvetique}).
+  % Conf\oe{}deratio Helvetica
+}
 %
+as in ``There's a Spezial Hell for the scalpers and cryptocurrency
+environmental terrorists stockpiling GeForce 3000 series cards so that
+I can {\em not} just get {\em one} darn card at a reasonable price for
+my silly SIGBOVIK experiments.'' So the extra-lowercase version of
+Helvetica is {\bf Spezial Hellvetica} (also Figure~\ref{fig:heavenica}).
+
+Times New Roman refers to the multiplication operator in algebra,
+which has a natural uppercase in exponentiation. Thus the uppercase
+version of Times New Roman is {\bf Exponential New Roman}. Computing
+Tetration New Roman or $\uparrow\uparrow\uparrow\uparrow$ New
+Roman~\cite{knuth1976coping} is straightforward, but extremely
+punctilious SIGBOVIK page limits preclude showing them here. Of course
+the lowercase version is {\bf Plus New Roman} (and similarly implies
+Successor New Roman). Both fonts are shown in Figure~\ref{fig:roman}.
+
+The vector-based {\bf Futurda} font (Figure~\ref{fig:futurda}) is in
+some ways more readable than these, but for the sake of comparison,
+note that these fonts are actually doing something more interesting,
+as they are built with both the \makeuppercase\ and
+\makelowercase\ models. Futurda's \letterform{A}--\letterform{Z} are
+just the lowercase of existing uppercase letters, which already has a
+correct solution and which a ML model can simply learn through
+memorization. In contrast, none of the letterforms in Heavenica can
+come through memorization of a training example (at worst, memorizing
+an inversion example generated by the other model after {\em it}
+memorized something). Subjectively, the fonts are not very readable
+and only slightly interesting, but the two models did demonstrate a
+reasonable ability to invert one another's behavior.
 
@@
 
-lorem ipsum (Figure~\ref{fig:perfect})
+Oh, you don't like letters that look {\em bad}? Instead you want
+letters that look {\em good}? How about {\em best}?
+
+The fonts in the previous section were created by modifying the case
+of existing letterforms, with mixed success. We can also do this to
+any letter-like shape. I built a UI for drawing letters and seeing
+them uppercased and lowercased (and then re-lowercased and
+re-uppercased) live, but it's impossible to demonstrate in paper form.
+It's pretty much what you'd expect.
+
+The UI also tells you how much your input resembles the various
+existing letters \letterform{A}--\letterform{Z} and
+\letterform{a}--\letterform{z} using the 26 bonus outputs that each
+model predicts. For example, I learned that the ``Cool
+S''~\cite{wikipediacools}:
+
+\begin{center}
+\includegraphics[width=0.1 \linewidth]{cools}
+\end{center}
+
+\noindent does not much resemble an S.
+% XXX ideally, numbers matching a scene in the video?
+
+This begs us to ask the question: What shapes {\em do} look like
+letters? Since the models will tell us, I can just search over shapes
+and ask them. The first thing I tried was to just generate random
+shapes and optimize their parameters to produce the highest possible
+prediction for the target letter, and the lowest possible prediction
+for the rest. This preduced results that are fully bonkers
+(Figure~\ref{fig:bonkersf}).
+
+We can improve the results by searching for inputs with a ``perfect''
+prediction (1.0) rather than making it as high as possible. These
+results may not have been fully bonkers, but were at least downright
+wacky. Since there appear to be a large variety of inputs that the
+model judges as ``perfect'', the most appealing results from this
+excursion involved scoring some additional properties of the
+hallucinated inputs to discourage them from being so barmy. I
+generated $8 \times 8$ bitmaps, plenty of pixels to make readable
+letters on classic computers. Rather than allowing them to be
+arbitrarily noisy, I also weakly optimized for (1) the number of
+pixels set being close to half (2) minimal transitions between
+on and off along each row and column. This produced shapes that
+are basically letter-like, but weird (Figure~\ref{fig:perfect}).
+
+\begin{figure}[tp]
+\centering
+  \includegraphics[width=0.9 \linewidth]{bonkersf}
+\caption{ A randomly generated SDF (left) and its rasterization
+  (right) which maximized the predicted score for \letterform{F}
+  (1.893 out of a nominal 1.0) while 
+  scoring all other letters low. Parts are recognizable as an
+  \letterform{F}, but other parts are fully bonkers. This is actually
+  one of the least weird ones.
+} \label{fig:bonkersf}
+\end{figure}
 
 \begin{figure*}[tp]
-\begin{centering}
+\centering
   \includegraphics[width=0.9 \linewidth]{perfecthallucination}
-\end{centering}
 \caption{
-%  XXX Why doesn't this caption come out the way I'm expecting?
-%  (in amsart style... can't have two figure*s on the same page?)
    Type specimen for the generated font {\bf Perfect Hallucination}.
-  TODO write a description!
+   Each letter is an 8x8 bitmap that looks as close to ``perfect''
+   as possible to the model. Perfect here means that for \letterform{C},
+   the \makelowercase\ model outputs as close to the vector
+   $[0, 0, 1, 0, 0, \ldots, 0]$ (in its 26 letter predictors; the
+   actual lowercasing is ignored) as the optimizer could find. (Of
+   course I didn't search all $2^{64}$ inputs, but errors are
+   on the order of one part per thousand). The models are completely
+   successful at recognizing normal-looking letters as well, but
+   it likes these even better.
 } \label{fig:perfect}
@@
 
-dolor sit amet
-
 \section{Chess-playing}
 
-% autotracing
+\newcommand\deterministic{
+  \includegraphics[height=0.8em]{deterministic}
+}
+\newcommand\asymmetric{
+  \includegraphics[height=0.8em]{asymmetric}
+}
+
+One obvious thing to do with a program that takes an $8 \times 8$
+bitmap and produces some kind of score for it is to use that
+program to play chess~\cite{murphy2019blind}. Here I entered
+26 such programs in the Elo World tournament~\cite{murphy2019eloworld},
+which allows us to see how they perform against each other
+and benchmark algorithms. (Badly.)
+
+At each turn, the algorithm takes the board state that would
+result from each legal move, and interprets it as an $8 \times 8$
+bitmap. It renders that bitmap as an SDF and then runs the
+\makelowercase\ model on it, and chooses the move that minimizes
+the difference between its letter predictions and the
+$[0, 0, \ldots 1, \ldots 0]$ vector corresponding to the letter
+we are ``playing as.'' \deterministic \asymmetric
+
+After tens of thousands of games each, it is clear that the letter-based
+players are all bad at chess. Letters \letterform{E} and
+\letterform{S} perform the worst (agreed on \letterform{S} being the
+worst, thank you very much!), even worse than the ``No, I insist!''
+strategy that tries to force its opponent to capture its pieces.
+Letters \letterform{K} and \letterform{T} (eponymous! yeah!) perform
+the best, but still worse than random. The numeric players like $\pi$
+and $e$ are categorically better than the alphabetical ones, but this
+is not surprising because chess is more of a mathematical game than a
+linguistic one (Figure~\ref{fig:chess}).
+
+\begin{figure}[tp]
+  \centering
+  \chessboard[setfen=B2k4/8/1b2p2n/1b2p1pr/6Pp/4B2P/4q3/4K3 w - - 10 60,showmover=false]
+  \caption{
+    The letter \letterform{P} (white) loses to the numeric constant
+    $\pi$ (black) after 59 nonsensical moves. The \letterform{P}
+    player tries to move pieces such that the board looks like a
+    letter \letterform{P} (and no other) to the neural network.
+    The $\pi$ player uses
+    $3 - \pi$ to arithmetically decode the sequence of legal moves
+    (sorted alphabetically by PGN). Neither player is concerned with
+    chess, really, but the letter-based players are generally bad
+    because they are more likely to get stuck in local minima once
+    they are basically happy with the shape of the pieces.
+    %  1.
+Na3 g5 2.
+b3 e6 3.
+f4 Nc6 4.
+e4 Be7 5.
+e5 a6 6.
+c3 f6 7.
+h3 Bb4 8.
+g4 h5 9.
+Nf3 h4 10.
+Nd4 Rb8 11.
+Nf5 Na5 12.
+Nc2 Rh7 13.
+Nce3 d6 14.
+Qf3 Rh5 15.
+Nh6 Bd7 16.
+Bc4 c6 17.
+Rf1 Qc7 18.
+fxg5 Bxc3 19.
+b4 Bxb4 20.
+a3 Bc8 21.
+axb4 Qe7 22.
+Kd1 c5 23.
+d3 Nxh6 24.
+Ke2 Ra8 25.
+Rd1 dxe5 26.
+Bb5 Bd7 27.
+Ke1 c4 28.
+Nd5 Qg7 29.
+Nc3 Nb3 30.
+Bc6 Nd4 31.
+Be3 Nc2 32.
+Kf1 cxd3 33.
+Rdc1 d2 34.
+Qf4 Qh7 35.
+Qd4 Na3 36.
+Rxa3 a5 37.
+Qe4 b6 38.
+bxa5 dxc1=B 39.
+Qd5 Kd8 40.
+Qc5 Rb8 41.
+Ke1 Qe7 42.
+Bf3 fxg5 43.
+Bb7 Bxa3 44.
+Ba6 Bxc5 45.
+Bb5 Qd6 46.
+Ba6 Ba4 47.
+axb6 Bc6 48.
+Bc8 Qe7 49.
+Ba6 Rb7 50.
+Bxb7 Bxb6 51.
+Na2 Bb5 52.
+Nc1 Qa3 53.
+Nd3 Kd7 54.
+Nb2 Qxb2 55.
+Bc8 Ke8 56.
+Ba6 Kd7 57.
+Bc8 Kd8 58.
+Bb7 Qg2 59.
+Ba8 Qe2
+    } \label{fig:chess}
+\end{figure}
+
+% figure? drop it?
+{
+  \tiny
+\begin{tabular}{lrrrr}
+name & elo & win & loss & draw \\
+\hline  
+worstfish & 395.95 & 10 & 44406 & 18584 \\
+no i insist & 605.47 & 0 & 20405 & 42595 \\
+letter f & 605.60 & 1341 & 21653 & 40006 \\
+letter y & 606.00 & 1347 & 21688 & 39965 \\
+letter b & 607.39 & 1705 & 21870 & 39425 \\
+letter p & 607.61 & 1790 & 21963 & 39247 \\
+letter l & 608.75 & 1370 & 21306 & 40324 \\
+letter j & 610.86 & 1737 & 21525 & 39738 \\
+letter u & 611.41 & 1363 & 20947 & 40690 \\
+letter c & 612.38 & 1264 & 20787 & 40949 \\
+huddle & 612.60 & 1172 & 20494 & 41334 \\
+letter n & 612.81 & 1395 & 20802 & 40803 \\
+letter w & 613.64 & 1342 & 20723 & 40935 \\
+letter g & 613.72 & 1339 & 20660 & 41001 \\
+letter v & 614.18 & 1390 & 20606 & 41004 \\
+letter h & 615.04 & 1328 & 20452 & 41220 \\
+letter r & 615.22 & 1826 & 20932 & 40242 \\
+letter x & 615.70 & 1414 & 20457 & 41129 \\
+letter m & 616.31 & 1774 & 20809 & 40417 \\
+letter q & 618.85 & 1378 & 19993 & 41629 \\
+letter o & 619.08 & 1502 & 20078 & 41420 \\
+letter z & 620.09 & 1745 & 20275 & 40980 \\
+letter d & 620.10 & 1730 & 20369 & 40901 \\
+letter a & 620.57 & 1819 & 20212 & 40969 \\
+letter i & 622.50 & 2169 & 20424 & 40407 \\
+letter k & 622.90 & 1755 & 19878 & 41367 \\
+letter t & 623.30 & 2237 & 20271 & 40492 \\
+\ldots \\
+random move & 655.90 & 7267 & 20983 & 34750 \\
+\ldots \\
+chessmaster.nes lv1 & 1014.74 & 37302 & 13992 & 11706 \\
+\ldots \\
+stockfish1m & 2831.66 & 60167 & 493 & 2340 \\
+\end{tabular}
+}
 
@@
 
-% cool s
+\section{Other Applications}
+
+People often stop me on the street to ask, Tom, Why did you spend so
+much time and energy on this useless SIGBOVIK project? To which I say,
+Ha! At least I am not wasting my time {\it reading} SIGBOVIK papers or
+talking to strangers on the street! and run off. Although the main
+purpose of SIGBOVIK is to confound bibliometrics with ambiguously
+good-faith and high-quality research published in a clearly satirical
+but superfically decorous venue of a traditionally esteemed university,
+it is also possible for such work to have practical applications in
+arts and spycraft. You simply need to give it some thought.
+
+For example, it is well known to {\sf internet troIIs} and {\sf DOMAlN
+  NAME PHlSHERS} that the uppercase \letterform{i} and loweracse
+\letterform{L} are indistinguishable in many fonts, allowing for
+various Tomfoolery.\footnote{I actually wrote ``troiis'' and ``domaln
+  name phlshers'', hehe!} This can also be used for
+steganography---hiding messages inside text---by replacing letters
+with their alternates. Each replacement only encodes one bit of
+information, however. With the generic ability to uppercase and
+lowercase letterforms, we can exploit this ambiguity to generate
+a large variety of letterforms that can be used like this.
+
+For example, the following sequence of distinct letters are hard to
+distinguish from one another and could all be used in place of a
+lowercase \letterform{L} or uppercase \letterform{i}:
+
+\begin{center}
+\includegraphics[width=0.25 \textwidth]{stegtaller}
+\end{center}
+
+The letterforms are generated by repeated application of the
+\makeuppercase\ ($\uparrow$) and \makelowercase\ ($\downarrow$)
+networks to the lowercase \letterform{l} from Helvetica. They
+are, from left to right: $\letterform{l}$,
+$\uparrow \downarrow \letterform{l}$,
+$\downarrow\uparrow \uparrow \downarrow \letterform{l}$,
+$\uparrow \downarrow\uparrow \downarrow \letterform{l}$,
+$\uparrow \uparrow \downarrow\downarrow \letterform{l}$,
+$\downarrow\downarrow\uparrow \uparrow \downarrow \letterform{l}$,
+$\downarrow\uparrow \uparrow \downarrow\downarrow \letterform{l}$,
+$\uparrow \uparrow \uparrow \downarrow\downarrow \letterform{l}$,
+$\downarrow\uparrow \downarrow\uparrow \uparrow \downarrow \letterform{l}$,
+$\downarrow\uparrow \uparrow \downarrow\uparrow \downarrow \letterform{l}$,
+$\downarrow\uparrow \uparrow \uparrow \downarrow\downarrow \letterform{l}$,
+$\uparrow \downarrow\downarrow\uparrow \uparrow \downarrow \letterform{l}$.
+
+In this way we can encode much longer bit strings in a single
+character, even thousands of iterations deep if there is a reasonable
+balance of uppercasing and lowercasing operations. (Too many in a row
+will get us stuck; see Section~\ref{sec:lowestcase}.) Not all
+sequences lead to a shape like this (commonly they resemble
+\letterform{i} or \letterform{L} when starting from \letterform{l}),
+but we can easily create a codebook of ones that do. Maybe I have even
+hidden an intricate message in this paper for you to discover? (I
+didn't.)
+
+% taller:
+% adlnotwERTUZ
+% 
+% shorter:
+% ceikmqrsyHQSWXY
+% 
+% Other applications: Steganography with llIIllIl
+
+% a:
+% b: l
+% c: ll
+% d: ul
+% e: lll
+% f: lul
+% g: ull
+% h: uul
+% i: llll
+% j: llul
+% k: lull
+% l: luul
+% m: ulll
+% n: ulul
+% o: uull
+% p: uuul
+% q: lllll
+% r: lllul
+% s: llull
+% t: lluul
+% u: lulll
+% v: lulul
+% w: luull
+% x: luuul
+% y: ullll
+% z: ullul
+% A: ulull
+% B: uluul
+% C: uulll
+% D: uulul
+% E: uuull
+% F: uuuul
+% G: llllll
+% H: llllul
+% I: lllull
+% J: llluul
+% K: llulll
+% L: llulul
+% M: lluull
+% N: lluuul
+% O: lullll
+% P: lullul
+% Q: lulull
+% R: luluul
+% S: luulll
+% T: luulul
+% U: luuull
+% V: luuuul
+% W: ulllll
+% X: ulllul
+% Y: ullull
+% Z: ulluul
+
+\section{To infinity, but let's stop there} \label{sec:lowestcase}
+
+Wow, this project is pretty involved, huh? Let's just add another
+dimension to it!
+
+We looked at what we get when we run \makelowercase\ on an existing
+lowercase letter, making it lowerercase. Of course, we can run the
+model again, and get an even lowererercase letter. The process can be
+repeated indefinitely. As it turns out, lowercasing tends to make
+letters smaller and smaller (makes sense) and they eventually just
+turn to dust (Figure~\ref{fig:lowestcaseo}) and stay that way (makes
+sense; ``dust to dust'').
+
+It's possible to make the results a bit more interesting by injecting
+additional energy after each iteration, either by adjusting the gamma
+or ``zooming'' into the active area. This seems pretty arbitrary,
+though. I think it is right to conclude that the lowestcase versions
+of letters are canonically specks of dust.
+
+On the other hand, repeatedly uppercasing produces more interesting
+results. Uppercasing usually increases the scale of a letter, but
+this effect is limited by the finiteness of the SDF and the fact
+that XXX
+
+\begin{figure}[t]
+\centering
+  \includegraphics[width=0.80 \linewidth]{lowestcaseo}
+  \caption{ The first 27 iterations of the \makelowercase\ model on
+    the letter \letterform{o}. The lowercase model generally makes
+    letters get smaller and smaller until they disappear. There is
+    still energy in the SDF (left column) but no pixels exceed the
+    threshold so the rasterization is empty (right column) for about
+    16 iterations. Finally it reaches a stable state, a tiny piece of
+    colon-shaped dust, easily mistaken for a printing error. All
+    letters (from the test font Helvetica) converge to this shape,
+    except mysteriously the letter \letterform{v}. Is \letterform{v} a
+    different alien species, masquerading as a normal letter for
+    millions of years? } \label{fig:lowestcaseo}
+\end{figure}
 
 % lowercase uppercase letters etc.
+% iterate this a lot of times?
+% "to infinity, but let's stop there"
+
+% run it to infinity to see the uppestcase, etc.
+% if these vanish, then we have a fixed point.
+% if not, or perhaps by doing some auto-gamma, then
+% it's like a fractal zoomer, and we can set the
+% record on the "most uppercase letter" by running
+% a jazillion times.
+
+% lowestcase is just dust, not surprising
+% uppestcase is actually a fixed point (not FixedSys)
+% ... so it's like a 3D manifold
+
+\begin{figure}[t]
+\centering
+  \includegraphics[width=0.7 \linewidth]{uppestcase3d}
+\caption{ 3D manifold showing a section of the repeating loop as the
+  \makeuppercase\ model iteratively uppercases a letter. (Shown here
+  is the input \letterform{q} from iteration 245--377, but they all
+  converge to this same periodic shape.) Slices through this shape give a
+  letterform's outline (or usually, a linear interpolation between two
+  of them). The bottom of the shape is its ``beginning'' but it
+  appears to repeat like this forever.
+} \label{fig:uppestcase3d}
+\end{figure}
+
+% bristol stool chart
 
@@
 For this project I used \verb+stb_truetype.h+ from the excellent stb
-collection of single-file libraries.~\cite{stb} I did push its limits
+collection of single-file libraries~\cite{stb}. I did push its limits
 somewhat and ``automatically discovered'' assertion failures and other
-crashes (e.g. during blackbox optimization on all 100k fonts), but it
+crashes (e.g.~during blackbox optimization on all 100k fonts), but it
 saved a lot of time. This library only helps with reading the fonts
@@
 and then did the final export with FontForge~\cite{fontforge}. Thanks
-Jason Reed for this suggestion.
-
-\nocite{murphy2019blind}
-\nocite{murphy2019eloworld}
-\nocite{stb}
+Jason Reed for this suggestion, without which I would probably still
+be developing my own TTF file writer and custom hinting engine. Finally,
+I would like to thank the pseudonymous SIGBOVIK program committee for
+overseeing the proceedings, and the \upsigbovik\ program committee
+committee for overseeing {\em them}.
 

### 2021-03-21 lowercase/makefont.cc
Wrote a bunch more...

@@ -256,7 +256,7 @@ static void GenerateOne(const Network &make_lowercase,
               FontProblem::TTFBaseline(SDF_CONFIG);
             return make_pair(op.output_char, blank_char);
           }
-       }, 13);
+       }, 16);
 
   TTF::Font font;
   font.baseline = FontProblem::TTFBaseline(SDF_CONFIG);

### 2021-03-21 trunk/lowercase/paper/lowercase.tex
Add code for tracing hand-made bitmaps into fonts, add hand-made font and write about it, tidy up paper a little

@@
 % \usepackage{adjustbox}
+\usepackage{supertabular}
 
@@
 
-The thing is: What if I'm even MORE ANGRY THAN I WAS BEFORE? There are
-some standard sorts of typographic emphasis, like I can be {\bf BOLD
-  ANGRY} or \textbf{\textit{\large BIG BOLD ITALIC UNDERLINE ANGRY}}
-or { \Large \textbf{\textit{\uuline{COMBINE A LOT OF THESE ANGERS}}}},
-each with its own nuances, depending on the cascading style sheet or
-LaTeX class file. To be even more casual than lowercase, u can learn 2
-write like this, and {\scriptsize shrink away} and also \sout{cross
-  out ur words in shame in advance of them even being read}, but there
-are few other options for de-emphasis. Plus, when I'm FEELING PRETTY
-ANGRY, TOM, how do I capitalize that already-capitalized T in order to
-show the proper reverence for your humanity?
+There's nothing wrong with this (despite the classical advice to use
+shift to {\em reduce} conflict~\cite{aho1986compilers}). But the thing is:
+What if I'm even MORE ANGRY THAN I WAS BEFORE? There are some standard
+sorts of typographic emphasis, like I can be {\bf BOLD ANGRY} or
+\textbf{\textit{\large BIG BOLD ITALIC UNDERLINE ANGRY}} or { \Large
+  \textbf{\textit{\uuline{COMBINE A LOT OF THESE ANGERS}}}}, each with
+its own nuances, depending on the cascading style sheet or LaTeX class
+file. To be even more casual than lowercase, u can learn 2 write like
+this, and {\scriptsize shrink away} and also \sout{cross out ur words
+  in shame in advance of them even being read}, but there are few
+other options for de-emphasis. Plus, when I'm FEELING PRETTY ANGRY,
+TOM, how do I capitalize that already-capitalized T in order to show
+the proper reverence for your humanity?
 
@@
 
-% shift-reduce conflict
-% Shift is often used to increase conflict, not reduce it
-
 \subsection{Induction}
@@
 
-\begin{figure}[ht]
-% \includegraphics[width=0.9 \linewidth]{manual}
-\caption{TODO} \label{fig:manual}
+\begin{figure}[hb]
+\centering
+\includegraphics[width=0.9 \linewidth]{extracase}
+\caption{%
+  Probably someone already had this idea and did it before I was
+  even born, thus taking the fun out of it for the rest of us, but
+  here's a hand-made alphabet with ``upperercase'' and ``lowerercase''
+  letters. You can download this TrueType font from \url{tom7.org/lowercase}.
+  %
+} \label{fig:manual}
 \end{figure}
@@
 \begin{figure}[ht]
+\centering
 \includegraphics[width=0.9 \linewidth]{training}
@@
 
-\begin{figure}[ht]
+\begin{figure}[ht!b]
 \centering
@@
 at the time that it didn't seem to make a difference! (The dark
-wizard of superstitious fiddling nods sagely.
+wizard of superstitious fiddling nods sagely.)
 
@@
 font Helvetica~\cite{helvetica} and serif font Times New
-Roman~\cite{timesnewroman}.
+Roman~\cite{timesnewroman}. Each font is produced by a symmetric
+process; for example, to make a font ``more lowercase,'' I take the
+input font's lowercase alphabet and run \makelowercase\ on it (this
+becomes the output font's lowercase letters), and then run
+\makeuppercase\ on {\em those} to produce the output font's uppercase
+letters. These letters are usually recognizable as the ``normal''
+lowercase letters even though they've been through both neural
+networks. Before tracing, I do some automatic gamma adjustment of the
+SDFs (e.g.~at least 5\% of the pixels should be above the threshold),
+as the unadjusted letters seemed a bit too light. These fonts can
+also be downloaded from \url{http://tom7.org/lowercase/} for your
+corporate PowerPoint needs.
 
@@
 as in ``There's a Spezial Hell for the scalpers and cryptocurrency
-environmental terrorists stockpiling GeForce 3000 series cards so that
+environmental terrorists stockpiling GeForce 3000 series GPUs so that
 I can {\em not} just get {\em one} darn card at a reasonable price for
-my silly SIGBOVIK experiments.'' So the extra-lowercase version of
+my important SIGBOVIK experiments.'' So the extra-lowercase version of
 Helvetica is {\bf Spezial Hellvetica} (also Figure~\ref{fig:heavenica}).
@@
 \letterform{a}--\letterform{z} using the 26 bonus outputs that each
-model predicts. For example, I learned that the ``Cool
-S''~\cite{wikipediacools}:
+model predicts. For example, I learned that the
+``Cool~S''~\cite{wikipediacools}:
 
@@
 arbitrarily noisy, I also weakly optimized for (1) the number of
-pixels set being close to half (2) minimal transitions between
-on and off along each row and column. This produced shapes that
-are basically letter-like, but weird (Figure~\ref{fig:perfect}).
+pixels set being close to half and (2) minimal transitions between on
+and off along each row and column. This produced shapes that are
+basically letter-like, but weird (Figure~\ref{fig:perfect}).
 
@@
   (right) which maximized the predicted score for \letterform{F}
-  (1.893 out of a nominal 1.0) while 
-  scoring all other letters low. Parts are recognizable as an
-  \letterform{F}, but other parts are fully bonkers. This is actually
-  one of the least weird ones.
+  (1.893 out of a nominal 1.0) while scoring all other letters low.
+  Parts are recognizable as an \letterform{F}, but other parts are
+  fully bonkers. This is actually one of the least weird ones.
 } \label{fig:bonkersf}
@@
 the difference between its letter predictions and the
-$[0, 0, \ldots 1, \ldots 0]$ vector corresponding to the letter
+$[0, 0, \ldots 1, \ldots 0]$ vector selecting the letter
 we are ``playing as.'' \deterministic \asymmetric
 
-After tens of thousands of games each, it is clear that the letter-based
-players are all bad at chess. Letters \letterform{E} and
+After tens of thousands of games each, it is clear that the
+letter-based players are all bad at chess. Letters \letterform{E} and
 \letterform{S} perform the worst (agreed on \letterform{S} being the
 worst, thank you very much!), even worse than the ``No, I insist!''
-strategy that tries to force its opponent to capture its pieces.
-Letters \letterform{K} and \letterform{T} (eponymous! yeah!) perform
-the best, but still worse than random. The numeric players like $\pi$
-and $e$ are categorically better than the alphabetical ones, but this
-is not surprising because chess is more of a mathematical game than a
+strategy that tries to force its opponent to capture its pieces. The
+letter \letterform{T} (eponymous! yeah!) performs best, but still
+worse than random. The numeric players like $\pi$ and $e$ are
+categorically better than the alphabetical ones, but this is not
+surprising because chess is more of a mathematical game than a
 linguistic one (Figure~\ref{fig:chess}).
@@
 % figure? drop it?
+
+The abbreviated tournament results:
 {
-  \tiny
-\begin{tabular}{lrrrr}
-name & elo & win & loss & draw \\
+\scriptsize
+\begin{center}
+
+\tablefirsthead{%
+  {\bf name} & {\bf elo} & {\bf wins} & {\bf losses} & {\bf draws}   \\  
   \hline
+  }
+\tablehead{%
+  \ldots & {\bf elo} & {\bf wins} & {\bf losses} & {\bf draws}   \\    
+  \hline
+}
+\tabletail{%
+  \multicolumn{5}{c}{\ldots} \\
+  % \hline
+}
+\tablelasttail{%
+  % nothing
+}
+% \begin{tabular}{lrrrr}
+\begin{supertabular}{lrrrr}
+% name & elo & wins & losses & draws \\
+% \hline  
 worstfish & 395.95 & 10 & 44406 & 18584 \\
+\ldots \\
+letter e &  602.27 &  1717 & 22566 & 38717 \\
+letter s &  602.92 &  1310 & 22020 & 39670 \\
 no i insist & 605.47 & 0 & 20405 & 42595 \\
@@
 stockfish1m & 2831.66 & 60167 & 493 & 2340 \\
-\end{tabular}
+\end{supertabular}
+% \end{tabular}
+\end{center}
 }
@@
 good-faith and high-quality research published in a clearly satirical
-but superfically decorous venue of a traditionally esteemed university,
-it is also possible for such work to have practical applications in
-arts and spycraft. You simply need to give it some thought.
+but superfically decorous venue associated with a traditionally
+esteemed university, it is also possible for such work to have
+practical applications in arts and spycraft. You simply need to give
+it some thought.
 
@@
   name phlshers'', hehe!} This can also be used for
-steganography---hiding messages inside text---by replacing letters
-with their alternates. Each replacement only encodes one bit of
-information, however. With the generic ability to uppercase and
-lowercase letterforms, we can exploit this ambiguity to generate
-a large variety of letterforms that can be used like this.
+steganography---hiding messages inside text without the use of em
+dashes---by selectively replacing letters with their alternates. Each
+replacement only encodes one bit of information, however. With the
+generic ability to uppercase and lowercase letterforms, we can exploit
+this ambiguity to generate a large variety of letterforms that can be
+used like this.
 
@@
 
-In this way we can encode much longer bit strings in a single
-character, even thousands of iterations deep if there is a reasonable
-balance of uppercasing and lowercasing operations. (Too many in a row
-will get us stuck; see Section~\ref{sec:lowestcase}.) Not all
-sequences lead to a shape like this (commonly they resemble
-\letterform{i} or \letterform{L} when starting from \letterform{l}),
-but we can easily create a codebook of ones that do. Maybe I have even
-hidden an intricate message in this paper for you to discover? (I
-didn't.)
-
-% taller:
+% letters in figure:
 % adlnotwERTUZ
-% 
-% shorter:
-% ceikmqrsyHQSWXY
-% 
-% Other applications: Steganography with llIIllIl
 
@@
 
+In this way we can encode much longer bit strings in a single
+character, even thousands of iterations deep if there is a reasonable
+balance of uppercasing and lowercasing operations. (Too many in a row
+will get us stuck; see Section~\ref{sec:lowestcase}.) Not all
+sequences lead to a shape like this (commonly they resemble
+\letterform{i} or \letterform{L} when starting from \letterform{l}),
+but we can easily create a codebook of ones that do. Maybe I have even
+hidden an intricate message in this paper for you to discover? (I
+didn't.)
+
 \section{To infinity, but let's stop there} \label{sec:lowestcase}
@@
 turn to dust (Figure~\ref{fig:lowestcaseo}) and stay that way (makes
-sense; ``dust to dust'').
+sense; ``dust to dust''~\cite{commonprayerburial}).
 
@@
 On the other hand, repeatedly uppercasing produces more interesting
-results. Uppercasing usually increases the scale of a letter, but
-this effect is limited by the finiteness of the SDF and the fact
-that XXX
+results. Uppercasing usually increases the scale of a letter, but this
+effect is limited by the finiteness of the SDF and the fact that the
+outer edges are very unlikely to have high values. Moreover, although
+uppercase letters are large, they also have a lot of internal space.
+So iterations do not simply grow in size or fill the space, but
+repeatedly grow and deteriorate like an organism in the Game of
+Life~\cite{games1970fantastic}. Animating the SDF under iterations of
+the model is reminiscent of a flickering candle, recalling another
+well-known approach to increasing the emphasis of text,
+\url{flamingtext.com}. Is this eternal flame itself the uppestcase
+letter? Alas, such an effect is not possible in print (paper is too
+flammable).
+
+Iterating the \makeuppercase\ model with 32-bit floats does not form
+any cycles in 25,000,000 iterations, nor does it if the intermediates
+are quantized to 8-bit ints. This is curious because under visual
+observation the sequence does appear periodic, and in fact seems to be
+the same characteristic loop reached by all letters. Presumably there
+are some oscillations with long, relatively prime periods or even some
+pixels that are monotonically growing or shrinking, but very slowly.
+However, if I render the SDF as a 1-bit bitmap at $2\times$ scale, I
+get a perfect cycle of 132 frames, which appears to be two passes
+through the main characteristic loop, but slightly different each
+time. This is not a strange property for a letter to have; for example
+a typical \letterform{B} has two copies of the same basic idea in it.
+
+.... XXX more here ...
 
@@
 
-\section{Acknowledgements}
-
+\section{Conclusion}
+... XXX ...
+This fixed point
+suggests that the lowestcase letter is already on your keyboard,
+but it is not a point, it is just \letterform{:}.
+
+\paragraph{Futura Work}
+I think I pretty much beat this one to death, honestly.
+
+\paragraph{Acknowledgements}
 For this project I used \verb+stb_truetype.h+ from the excellent stb

### 2021-03-21 trunk/lowercase/template-makefont.cc

@@ -1,0 +1,166 @@
+
+#include <vector>
+#include <string>
+#include <algorithm>
+#include <cstdint>
+#include <memory>
+
+#include "timer.h"
+#include "font-problem.h"
+
+#include "image.h"
+#include "lines.h"
+#include "base/stringprintf.h"
+
+#include "threadutil.h"
+
+using namespace std;
+
+using uint8 = uint8_t;
+using uint32 = uint32_t;
+using int64 = int64_t;
+
+
+// XXX somehow needs to be shared?
+static constexpr int ROW0_MAX_PTS = 38;
+static constexpr int ROW1_MAX_PTS = 14;
+static constexpr int ROW2_MAX_PTS = 10;
+constexpr std::initializer_list<int> row_max_points = {
+  ROW0_MAX_PTS,
+  ROW1_MAX_PTS,
+  ROW2_MAX_PTS,
+};
+
+
+struct Op {
+  char input_char;
+  char output_char;
+  Op(char i, char o) : input_char(i), output_char(o) {}
+};
+
+struct Config {
+  string font_name = "Upperercase and Lowerercase";
+  string filename = "upperercase.sfd";
+  string copyright = "http://tom7.org/lowercase - "
+    "Hand-drawn font with 'extra uppercase' and 'extra lowercase' letters, "
+    "plus matching (but normal) numbers and punctuation to round out "
+    "the en-US keyboard characters. Copyright (c) 2021 Tom Murphy 7. "
+    "Distribute freely";
+  float baseline = 0.75f;
+  float extra_scale = 1.0f;
+  float linegap = 0.0f;
+  float blank_width = 0.66f;
+
+  int BSIZE = 800;
+  
+  vector<pair<string, string>> templates = {
+    {"upperercase.png",
+     "ABCDEFGHIJKLMNOPQRSTUVWXYZ"},
+    {"lowerercase.png",
+     "abcdefghijklmnopqrstuvwxyz"},
+    {"numbers.png",
+     "1234567890!@#$%^&*()"},
+    {"punctuation.png",
+     "-=_+[]{}\\|;:'\"<>?,./`~"},
+  };
+  // TODO punctuation...
+};
+
+// static constexpr FontProblem::SDFConfig SDF_CONFIG = {};
+// Use 2x config
+static constexpr FontProblem::SDFConfig SDF_CONFIG = {
+  .sdf_size = 36 * 2,
[…]

### 2021-03-22 lowercase/paper/lowercase.tex
Paper almost done

@@
 
+\newcommand\upsigbovik{
+  \includegraphics[height=1em]{upsigbovik}
+}
+
 \begin{document}
@@
 Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
-Appears in SIGBOVIK~2021 with the
+Appears in \upsigbovik~2021 with the
 OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
@@
 
-% shift-reduce conflict
-
 \subsection{Induction}
@@
 also categorizing them as serif, sans-serif, decorative, techno, etc.,
-which I never used). In doing this I noticed another extremely common
-problem, which was that many fonts had the same letter shapes for
-uppercase and lowercase letters. This would not do for the current
-application!
-
-\begin{figure*}[ht]
-\centering
-  \includegraphics[width=0.9 \linewidth]{sortition}
+which classifications I never used). In doing this I noticed another
+extremely common problem, which was that many fonts had the same
+letter shapes for uppercase and lowercase letters. This would not do
+for the current application!
+
+\begin{figure}[tp]
+\centering
+  \includegraphics[width=0.98 \linewidth]{sortition}
 \caption{ The interactive font data-cleaning UI. A seemingly endless
@@
 
-\begin{figure}[ht]
+\begin{figure}[tp]
 \centering
@@
 letterforms fit within that budget. Letterforms can be made of
-multiple contours (e.g. a stacked \letterform{g} typically has two
+multiple contours (e.g.~a stacked \letterform{g} typically has two
 holes in it, and \letterform{j} has two disjoint parts). I found that
@@
 place because C is the third letter. We don't need these outputs for
-the problem itself (e.g. to lowercase new letter shapes), but there
+the problem itself (e.g.~to lowercase new letter shapes), but there
 are several ideas behind this. First, the lowercasing function we're
@@
 MS~\cite{comicsans} lowercase using the model (with the exception of the
-\letterform{\%} character, which has too many contours---{\it five}!).
+\letterform{{\sf \%}} character, which has too many contours---{\it five}!).
 Mostly this model produces small, non-confident scrawls, like little
@@
   and ``Sphinx of black quartz, judge my vow!'' Although the output
-  barely resembles letters, it does have a certain dynamic
+  barely resembles letters, it does have a certain wispy
   Rorschach aesthetic, like a collection of delicate moths pinned
@@
 I also output each contour's points in a normalized order, starting
-from the point closest to the origin. This removes one needless
-degree of freedom.\footnote{We can see how this manifests in
-  the biases on the output layer, which are a proxy for the
-  ``average prediction''. In the first model, because of the
-  unstructured order, these are mostly near 0.5 (center of the
-  character) or 0.0 (degenerate, unused contours). In this new model,
-  the distribution of biases is much more flat; it can learn that
-  ``the first point tends to be near 0.25,0.25'' and ``the seventh
-  point tends to be near 0.64,0.3.''}
+from the point closest to the origin. This removes one needless degree
+of freedom.\footnote{We can see (well, it's not pictured since I have
+  far exceeded a reasonable number of figures in this paper, but {\em
+    I} can see) how this manifests in the biases on the output layer,
+  which are a proxy for the ``average prediction''. In the first
+  model, because of the unstructured order, these are mostly near 0.5
+  (center of the character) or 0.0 (degenerate, unused contours). In
+  this new model, the distribution of biases is much more flat; it can
+  learn that ``the first point tends to be near 0.25,0.25'' and ``the
+  seventh point tends to be near 0.64,0.3.''}
 
@@
 generated {\bf Futurda} (Figure~\ref{fig:futurda}).
-%
-% \begin{figure}[ht]
-% \centering
-%   \includegraphics[width=0.5 \linewidth]{rotatej}
-% \caption{ Example alignment (from debug UI) of predicted shape (blue)
-%   to expected shape (green). We require each point to be mapped (red)
-%   to a point from the expected contour in a monotonic order (but
-%   several can be mapped to the same one), so that we can attribute
-%   error to each point. I hope it was fun to write this code because
-%   it was too slow to do pretty much anything but generate this figure!
-% } \label{fig:rotatej}
-% \end{figure}
-%
 
@@
 
-% the sdf problem
-
 \section{SDFs}
@@
   far superior despite being derived from a $36 \times 36$ pixel
-  bitmap. Typically this process is performed at an even higher
+  image. Typically this process is performed at an even higher
   scale and then downsampled to produce an anti-aliased image.
@@
 value'' of 0.862 (because much more of the SDF is outside the letter
-than inside) and the distance is scaled as 0.059 units per pixel (so
-that pixels on the outer edge often have non-zero values). Compared to
-the first version, I was somewhat more permissive in what fonts I
-trained on, since there was no inherent limit to the number of
+than inside) and the distance is scaled as 0.059 units per pixel
+(chosen so that pixels on the outer edge often have non-zero values).
+Compared to the first version, I was somewhat more permissive in what
+fonts I trained on, since there was no inherent limit to the number of
 contours or their complexity. I did exclude fonts whose rasterizations
@@
 ``inversion'' examples (Figure~\ref{fig:cotraining}), but this is just
-a matter of programming, thank God. I should remember to do projects
+a matter of programming, thank god. I should remember to do projects
 that are mostly a matter of programming. Each round, $25\%$ of the
@@
   almost never activated (Section~\ref{sec:fiddly}). These momentarily
-  increase error but it is easily fine-tuned away (e.g. by learning
+  increase error but it is easily fine-tuned away (e.g.~by learning
   new biases). The peak at around 1.4M rounds is when I added a new
@@
     although there are some oddities (probably because it found some
-    other, better correlates). Nodes on this new layer have dense
+    other, better correlates). All nodes on this new layer have dense
     references to these 26 predictions on the previous; this means
-    that the bottom row kind of represents biases for each of the 26
+    that the bottom 26 rows kind of represent biases for each of the 26
     letters (what does an average 'e' look like?). I also included a
@@
 waiting). Since I am using OpenCL (whoa, yeah, stop me right there, I
-know) I found a good trick was to generate different OpenCL code with
+know) I found a good technique was to generate different OpenCL code with
 constants baked for each layer (for example their size and
-\verb+indices_per_node+); this allows the cmopiler to use faster
+\verb+indices_per_node+); this allows the compiler to use faster
 tricks in inner loops for e.g.~multiplication by a compile-time constant
@@
   ANGRY}. I will say in my defense that at least I felt disappointed
-at the time that it didn't seem to make a difference!
+at the time that it didn't seem to make a difference! (The dark
+wizard of superstitious fiddling nods sagely.)
 
 % you propagate my back, I'll propagate yours
+
+Ultimately, each of the two models was trained for over 2 million
+rounds, which corresponds to 510 million training examples. Each
+model is about 24 megabytes.
 
@@
 we'd actually like to have vector representations so that we can
-use them as TrueType fonts for our desktop publishing needs.
+download the TTF files and clog up our fonts menu forever.
 
@@
 and see where it reaches the edge value on the SDF; this point (a
-float) is output into the contour. I'm guaranteed to get back to where
-I started. I recurse by negating the SDF (outside becomes inside) and
-bitmap (land becomes sea), and reverse the winding order of the result
-of recursion.
+float) is output into the contour. The process is guaranteed to return
+to where we started. I recurse by negating the SDF (outside
+becomes inside) and bitmap (land becomes sea), and reverse the winding
+order of the result of recursion.
 
@@
 Since I output points at every pixel, sometimes these points are
-inefficient (i.e. a series of colinear points on a straight line),
+inefficient (e.g.~a series of colinear points on a straight line),
 and sometimes they reflect sharp corners that are not aesthetic. So
@@
 
+\begin{figure*}[tp]
+\centering
+\includegraphics[width=0.9 \linewidth]{heavenica} \\[0.5in]
+\includegraphics[width=0.9 \linewidth]{spezialhellvetica} \\
+\caption{
+  Type specimens for the generated font {\bf Heavenica} (top)
+  and {\bf Spezial Hellvetica} (bottom). The uppercase letters
+  in Heavenica are \makeuppercase\ applied to uppercase letters
+  from Helvetica, and the lowercase are \makelowercase\ applied
+  to those. These lower-upperercase letters resemble regular
+  uppercase letters, as they should; this gives you some idea
+  of the quality of the model. Spezial Hellvetica is the
+  symmetric thing (its lowercase letters are \makelowercase\ 
+  of Helvetica's lowercase). The sample text in this latter
+  case is ``Quartz jock vends BMW glyph fix. Twelve ziggurats
+  quickly jumped a finch box.''
+} \label{fig:heavenica}
+\end{figure*}
+
+\begin{figure*}[tp]
+\centering
+\includegraphics[width=0.9 \linewidth]{expnewroman} \\[0.5in]
+\includegraphics[width=0.9 \linewidth]{plusnewroman} \\
+\caption{
+  Type specimens for the generated font {\bf Exponential New Roman} (top)
+  and {\bf Plus New Roman} (bottom). These were produced with
+  the same procedure as in Figure~\ref{fig:heavenica}, but starting
+  with Times New Roman. The letterforms are clearly different, so
+  it's not as though the models are (just) memorizing a shape for
+  each letter. Notably, smudgy serifs reappear when the upperercase letters
+  are re-lowercased, as desired.
+  Sample text here is ``Amazingly few discotheques provide jukeboxes.
+  Those that don't MAKE ME QUITE ANGRY.'' and 
+  ``By Jove, my quirky study of lexicography won a prize! (the prize
+  was a crappy font)''
+} \label{fig:roman}
+\end{figure*}
+
+\medskip
+Now we're all set up to take an input shape (e.g.~from an existing
+font), run the \makeuppercase\ or \makelowercase\ model(s) on it,
+maybe multiple times, and trace the resulting SDF into a vector form
+that can be used in a font. I did this on the canonical sans serif
+font Helvetica~\cite{helvetica} and serif font Times New
+Roman~\cite{timesnewroman}. Each font is produced by a symmetric
+process; for example, to make a font ``more lowercase,'' I take the
+input font's lowercase alphabet and run \makelowercase\ on it (this
+becomes the output font's lowercase letters), and then run
+\makeuppercase\ on {\em those} to produce the output font's uppercase
+letters. These letters are usually recognizable as the ``normal''
+lowercase letters even though they've been through both neural
+networks. Before tracing, I do some automatic gamma adjustment of the
+SDFs (e.g.~at least 5\% of the pixels should be above the threshold),
+as the unadjusted letters seemed a bit too light. These fonts can
+also be downloaded from \url{http://tom7.org/lowercase/} for your
+corporate PowerPoint needs.
+
+Helvetica means ``of Hell'', so making the font more uppercase give us
+{\bf Heavenica} (Figure~\ref{fig:heavenica}), since Heaven is ``up''
+from Hell. What's lower than Hell? Spezial Hell,\footnote{
+  I first learned about Spezial Hell from a Rugen Br\"au beer that I
+  drank in the Alps in Grindelwald, Switzerland ({\it la
+    Conf\'ed\'eration H\'elvetique}).
+  % Conf\oe{}deratio Helvetica
+}
+%
+as in ``There's a Spezial Hell for the scalpers and cryptocurrency
+environmental terrorists stockpiling GeForce 3000 series GPUs so that
+I can {\em not} just get {\em one} darn card at a reasonable price for
+my important SIGBOVIK experiments.'' So the extra-lowercase version of
+Helvetica is {\bf Spezial Hellvetica} (also Figure~\ref{fig:heavenica}).
+
+Times New Roman refers to the multiplication operator in algebra,
+which has a natural uppercase in exponentiation. Thus the uppercase
+version of Times New Roman is {\bf Exponential New Roman}. Computing
+Tetration New Roman or $\uparrow\uparrow\uparrow\uparrow$ New
+Roman~\cite{knuth1976coping} is straightforward, but extremely
+punctilious SIGBOVIK page limits preclude showing them here. Of course
+the lowercase version is {\bf Plus New Roman} (and similarly implies
+Successor New Roman). Both fonts are shown in Figure~\ref{fig:roman}.
+
+The vector-based {\bf Futurda} font (Figure~\ref{fig:futurda}) is in
+some ways more readable than these, but for the sake of comparison,
+note that these fonts are actually doing something more interesting,
+as they are built with both the \makeuppercase\ and
+\makelowercase\ models. Futurda's \letterform{A}--\letterform{Z} are
+just the lowercase of existing uppercase letters, which already has a
+correct solution and which a ML model can simply learn through
+memorization. In contrast, none of the letterforms in Heavenica can
+come through memorization of a training example (at worst, memorizing
+an inversion example generated by the other model after {\em it}
+memorized something). Subjectively, the fonts are not very readable
+and only slightly interesting, but the two models did demonstrate a
+reasonable ability to invert one another's behavior.
+
 \section{Perfect letters, hallucinated} \label{sec:hallucination}
 
-All right, but why stop there??
-
-... the fonts in the previous section were created by modifying the
-case of existing letterforms, with mixed success. We can also do this
-to any letter-like shape. I built a UI for drawing letters and seeing
+Oh, you don't like letters that look {\em bad}? Instead you want
+letters that look {\em good}? How about {\em best}?
+
+The fonts in the previous section were created by modifying the case
+of existing letterforms, with mixed success. We can also do this to
+any letter-like shape. I built a UI for drawing letters and seeing
 them uppercased and lowercased (and then re-lowercased and
-re-uppercased) live, but it's impossible to demonstrate in paper
-form. It's pretty much what you'd expect.
+re-uppercased) live, but it's impossible to demonstrate in paper form.
+It's pretty much what you'd expect.
 
@@
 the difference between its letter predictions and the
-$[0, 0, \ldots 1, \ldots 0]$ vector corresponding to the letter
+$[0, 0, \ldots 1, \ldots 0]$ vector selecting the letter
 we are ``playing as.'' \deterministic \asymmetric
 
-The players all perform fairly badly. Letters \letterform{E} and
-\letterform{S} are the worst (agreed on \letterform{S} being the worst,
-thank you very much!), even worse than the ``No, I insist!'' strategy
-that tries to force its opponent to capture its pieces. Letters
-\letterform{K} and \letterform{T} (eponymous! yeah!) perform the best,
-but still worse than random. The numeric players like $\pi$ and $e$
-are categorically better than the alphabetical ones, but this is not
+After tens of thousands of games each, it is clear that the
+letter-based players are all bad at chess. Letters \letterform{E} and
+\letterform{S} perform the worst (agreed on \letterform{S} being the
+worst, thank you very much!), even worse than the ``No, I insist!''
+strategy that tries to force its opponent to capture its pieces. The
+letter \letterform{T} (eponymous! yeah!) performs best, but still
+worse than random. The numeric players like $\pi$ and $e$ are
+categorically better than the alphabetical ones, but this is not
 surprising because chess is more of a mathematical game than a
@@
 
-% figure? drop it?
+The abbreviated tournament results:
 {
-  \tiny
-\begin{tabular}{lrrrr}
-name & elo & win & loss & draw \\
+\scriptsize
+\begin{center}
+
+\tablefirsthead{%
+  {\bf name} & {\bf elo} & {\bf wins} & {\bf losses} & {\bf draws}   \\  
+  \hline
+  }
+\tablehead{%
+  \ldots & {\bf elo} & {\bf wins} & {\bf losses} & {\bf draws}   \\    
+  \hline
+}
+\tabletail{%
+  \multicolumn{5}{c}{\ldots} \\
+  % \hline
+}
+\tablelasttail{%
+  % nothing
+}
+% \begin{tabular}{lrrrr}
+\begin{supertabular}{lrrrr}
+% name & elo & wins & losses & draws \\
+% \hline  
+worstfish & 395.95 & 10 & 44406 & 18584 \\
+\ldots \\
+letter e &  602.27 &  1717 & 22566 & 38717 \\
+letter s &  602.92 &  1310 & 22020 & 39670 \\
 no i insist & 605.47 & 0 & 20405 & 42595 \\
@@
 chessmaster.nes lv1 & 1014.74 & 37302 & 13992 & 11706 \\
-\ldots
+\ldots \\
 stockfish1m & 2831.66 & 60167 & 493 & 2340 \\
-\end{tabular}
+\end{supertabular}
+% \end{tabular}
+\end{center}
 }
@@
 
-\section{To infinity, but let's stop there}
-
-Wow, this project is pretty involved, huh? Let's just add another
-dimension to it.
-
-% lowercase uppercase letters etc.
-% iterate this a lot of times?
-% "to infinity, but let's stop there"
-
-% run it to infinity to see the uppestcase, etc.
-% if these vanish, then we have a fixed point.
-% if not, or perhaps by doing some auto-gamma, then
-% it's like a fractal zoomer, and we can set the
-% record on the "most uppercase letter" by running
-% a jazillion times.
-
-% lowestcase is just dust, not surprising
-% uppestcase is actually a fixed point (not FixedSys)
-% ... so it's like a 3D manifold
+\section{Other Applications}
+
+People often stop me on the street to ask, Tom, Why did you spend so
+much time and energy on this useless SIGBOVIK project? To which I say,
+Ha! At least I am not wasting my time {\it reading} SIGBOVIK papers or
+talking to strangers on the street! and run off. Although the main
+purpose of SIGBOVIK is to confound bibliometrics with ambiguously
+good-faith and high-quality research published in a clearly satirical
+but superficially decorous venue associated with a traditionally
+esteemed university, it is also possible for such work to have
+practical applications in arts and spycraft. You simply need to give
+it some thought.
+
+For example, it is well known to {\sf internet troIIs} and {\sf DOMAlN
+  NAME PHlSHERS} that the uppercase \letterform{i} and loweracse
+\letterform{L} are indistinguishable in many fonts, allowing for
+various Tomfoolery.\footnote{I actually wrote ``troiis'' and ``domaln
+  name phlshers'', hehe!} This can also be used for
+steganography---hiding messages inside text without the use of em
+dashes---by selectively replacing letters with their alternates. Each
+replacement only encodes one bit of information, however. With the
+generic ability to uppercase and lowercase letterforms, we can exploit
+this ambiguity to generate a large variety of letterforms that can be
+used like this.
+
+For example, the following sequence of distinct letters are hard to
+distinguish from one another and could all be used in place of a
+lowercase \letterform{L} or uppercase \letterform{i}:
+
+\begin{center}
+\includegraphics[width=0.25 \textwidth]{stegtaller}
+\end{center}
+
+The letterforms are generated by repeated application of the
+\makeuppercase\ ($\uparrow$) and \makelowercase\ ($\downarrow$)
+networks to the lowercase \letterform{l} from Helvetica. They
+are, from left to right: $\letterform{l}$,
+$\uparrow \downarrow \letterform{l}$,
+$\downarrow\uparrow \uparrow \downarrow \letterform{l}$,
+$\uparrow \downarrow\uparrow \downarrow \letterform{l}$,
+$\uparrow \uparrow \downarrow\downarrow \letterform{l}$,
+$\downarrow\downarrow\uparrow \uparrow \downarrow \letterform{l}$,
+$\downarrow\uparrow \uparrow \downarrow\downarrow \letterform{l}$,
+$\uparrow \uparrow \uparrow \downarrow\downarrow \letterform{l}$,
+$\downarrow\uparrow \downarrow\uparrow \uparrow \downarrow \letterform{l}$,
+$\downarrow\uparrow \uparrow \downarrow\uparrow \downarrow \letterform{l}$,
+$\downarrow\uparrow \uparrow \uparrow \downarrow\downarrow \letterform{l}$,
+$\uparrow \downarrow\downarrow\uparrow \uparrow \downarrow \letterform{l}$.
+
+% letters in figure:
+% adlnotwERTUZ
+
+% a:
+% b: l
+% c: ll
+% d: ul
+% e: lll
+% f: lul
+% g: ull
+% h: uul
+% i: llll
+% j: llul
+% k: lull
+% l: luul
+% m: ulll
+% n: ulul
+% o: uull
+% p: uuul
+% q: lllll
+% r: lllul
+% s: llull
+% t: lluul
+% u: lulll
+% v: lulul
+% w: luull
+% x: luuul
+% y: ullll
+% z: ullul
+% A: ulull
+% B: uluul
+% C: uulll
+% D: uulul
+% E: uuull
+% F: uuuul
+% G: llllll
+% H: llllul
+% I: lllull
+% J: llluul
+% K: llulll
+% L: llulul
+% M: lluull
+% N: lluuul
+% O: lullll
+% P: lullul
+% Q: lulull
+% R: luluul
+% S: luulll
+% T: luulul
+% U: luuull
+% V: luuuul
+% W: ulllll
+% X: ulllul
+% Y: ullull
+% Z: ulluul
+
+In this way we can encode much longer bit strings in a single
+character, even thousands of iterations deep if there is a reasonable
+balance of uppercasing and lowercasing operations. (Too many in a row
+will get us stuck; see Section~\ref{sec:lowestcase}.) Not all
+sequences lead to a shape like this (commonly they resemble
+\letterform{i} or \letterform{L} when starting from \letterform{l}),
+but we can easily create a codebook of ones that do. Maybe I have even
+hidden an intricate message in this paper for you to discover? (I
+didn't.)
+
+\section{To infinity, but let's stop there} \label{sec:lowestcase}
+
+\begin{figure}[t]
+\centering
+  \includegraphics[width=0.80 \linewidth]{lowestcaseo}
+  \caption{ The first 27 iterations of the \makelowercase\ model on
+    the letter \letterform{o}. The lowercase model generally makes
+    letters get smaller and smaller until they disappear. There is
+    still energy in the SDF (left column) but no pixels exceed the
+    threshold so the rasterization is empty (right column) for about
+    16 iterations. Finally it reaches a stable state, a tiny piece of
+    colon-shaped dust, easily mistaken for a printing error. All
+    letters (from the test font Helvetica) converge to this shape,
+    except mysteriously the letter \letterform{v}. Is \letterform{v} a
+    different alien species, masquerading as a normal letter for
+    millions of years? } \label{fig:lowestcaseo}
+\end{figure}
 
@@
   \makeuppercase\ model iteratively uppercases a letter. (Shown here
-  is the input \lowercase{q} from iteration 245--377, but they all
+  is the input \letterform{q} from iteration 245--377, but they all
   converge to this same periodic shape.) Slices through this shape give a
@@
 
-% bristol stool chart
-
-\section{Acknowledgements}
-
+Wow, this project is pretty involved, huh? Let's just add another
+dimension to it!
+
+We looked at what we get when we run \makelowercase\ on an existing
+lowercase letter, making it lowerercase. Of course, we can run the
+model again, and get an even lowererercase letter. The process can be
+repeated indefinitely. As it turns out, lowercasing tends to make
+letters smaller and smaller (makes sense) and they eventually just
+turn to dust (Figure~\ref{fig:lowestcaseo}) and stay that way (makes
+sense; ``dust to dust''~\cite{commonprayerburial}).
+
+It's possible to make the results a bit more interesting by injecting
+additional energy after each iteration, either by adjusting the gamma
+or ``zooming'' into the active area. This seems pretty arbitrary,
+though. I think it is right to conclude that the lowestcase versions
+of letters are canonically specks of dust.
+
+On the other hand, repeatedly uppercasing produces more interesting
+results. Uppercasing usually increases the scale of a letter, but this
+effect is limited by the finiteness of the SDF and the fact that the
+outer edges are very unlikely to have high values. Moreover, although
+uppercase letters are large, they also have a lot of internal space.
+So iterations do not simply grow in size or fill the space, but
+repeatedly grow and deteriorate like an organism in the Game of
+Life~\cite{games1970fantastic}. Animating the SDF under iterations of
+the model is reminiscent of a flickering candle, recalling another
+well-known approach to increasing the emphasis of text,
+\url{flamingtext.com}. Is this eternal flame itself the uppestcase
+letter? Alas, such an effect is not possible in print (paper is too
+flammable).
+
+Iterating the \makeuppercase\ model with 32-bit floats does not form
+any cycles in 25,000,000 iterations, nor does it if the intermediates
+are quantized to 8-bit ints. This is curious because under visual
+observation the sequence does appear periodic, and in fact seems to be
+the same characteristic loop reached by all letters. Presumably there
+are some oscillations with long, relatively prime periods or even some
+pixels that are monotonically growing or shrinking, but very slowly.
+However, if I render the SDF as a 1-bit bitmap at $2\times$ scale, I
+get a perfect cycle of 132 frames. This appears to be two passes
+through the main characteristic loop, but slightly different each
+time. This is not a strange property for a letter to have; for example
+a typical \letterform{B} has two copies of the same basic idea in it.
+
+None of these alone could be considered the uppestcase letter, but
+perhaps together they are? A natural way to include them all in one
+shape is to stack them in 3D, as if each iteration is an MRI slice of
+the brachial plexus.
+
+To generate a 3D mesh, I stack the 132 SDFs in the z direction, and
+this natrually yields a three-dimensional signed distance field
+(trilinear interpolation). The classic ``marching cubes''
+algorithm~\cite{lorensen1987marching} for mesh generation works
+natively on such a field, and come to think of it, I probably could
+have used that in two dimensions to trace the SDFs to vectors. Oh,
+well. Fortunately there is enough spatial similarity between the
+slices that it makes a reasonable 3D shape even without interpolation,
+but sampled at a decent resolution and then cleaned up, it looks quite
+nice; much better than the lowercase colon. Speaking of colons, it
+rates approximately a 2.5 on the Bristol stool
+scale~\cite{lewis1997bristol}. A 2D projection of the manifold is in
+Figure~\ref{fig:uppestcase3d}. I 3D-printed it and am currently
+working on a way of embedding it as an unusually tall key on my
+keyboard that I can press whenever I wanted to express ULTIMATE ANGER.
+
+\section{Conclusion}
+We performed an exhaustive case analysis, exploring cases both more
+upper- and lower- than ever been seen before. Sideways case was not
+considered, as that is nonsense. We had modest success generating
+upperercase and lowerercase letterforms through two neural network
+models simultaneously trained to be each other's inverse, although
+frankly it was much faster and more aesthetic to just do it by hand.
+After really following through on the downloadables and taking some
+needless excursions for effect, we saw that these models have
+limits (at least informally). The lowestcase letter is already
+on your keyboard; it is the ASCII eyeballs character \letterform{:}.
+The uppestcase letter is not so easily typed, and perhaps that
+is for the best.
+
+\paragraph{Futura Work.}
+I think I pretty much beat this one to death, honestly.
+
+\paragraph{Acknowledgements.}
 For this project I used \verb+stb_truetype.h+ from the excellent stb
-collection of single-file libraries.~\cite{stb} I did push its limits
+collection of single-file libraries~\cite{stb}. I did push its limits
 somewhat and ``automatically discovered'' assertion failures and other
-crashes (e.g. during blackbox optimization on all 100k fonts), but it
+crashes (e.g.~during blackbox optimization on all 100k fonts), but it
 saved a lot of time. This library only helps with reading the fonts

### 2021-03-31 trunk/lowercase/paper/lowercase.tex
Various fixes / tools for video

@@
 
-\title{Lowestcase and Uppestcase letters: Advances in Derp Learning}
+\title{Lowestcase and uppestcase letters: Advances in derp learning}
 \author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
@@
 optimization algorithms are worth considering. Here, the
-interface\footnote{ Here a simplfied wrapper around
+interface\footnote{ Here a simplified wrapper around
   BiteOpt~\cite{biteopt} in my {\tt cc-lib} library. See
@@
 train and seemed like a complicated version of an bad, oversimplified
-approach. So, I control-C'd that thing and threw this whole endeavour
+approach. So, I control-C'd that thing and threw this whole endeavor
 in the trash! But I must have confused the Recycle~Bin icon with the
@@
 % problems or successes?
-
-% the sdf problem
 
@@
 I also adapted my weird home-grown software to train the
-\makeuppercase\ and \makelowercase\ models simultanously. Two models
+\makeuppercase\ and \makelowercase\ models simultaneously. Two models
 fit easily in GPU memory, with plenty of space for a stream of
@@
   from one another} (redundant or even partially redundant features
-are less valuable). It was nice to satisfy my supersition, and the
+are less valuable). It was nice to satisfy my superstition, and the
 dark wizard of superstitious fiddling with neural networks in the
@@
 prediction for the target letter, and the lowest possible prediction
-for the rest. This preduced results that are fully bonkers
+for the rest. This produced results that are fully bonkers
 (Figure~\ref{fig:bonkersf}).
@@
 
-% figure? drop it?
-
 The abbreviated tournament results:
@@
 good-faith and high-quality research published in a clearly satirical
-but superfically decorous venue associated with a traditionally
+but superficially decorous venue associated with a traditionally
 esteemed university, it is also possible for such work to have
@@
 For example, it is well known to {\sf internet troIIs} and {\sf DOMAlN
-  NAME PHlSHERS} that the uppercase \letterform{i} and loweracse
+  NAME PHlSHERS} that the uppercase \letterform{i} and lowercase
 \letterform{L} are indistinguishable in many fonts, allowing for
@@
 
+\begin{figure}[t]
+\centering
+  \includegraphics[width=0.80 \linewidth]{lowestcaseo}
+  \caption{ The first 27 iterations of the \makelowercase\ model on
+    the letter \letterform{o}. The lowercase model generally makes
+    letters get smaller and smaller until they disappear. There is
+    still energy in the SDF (left column) but no pixels exceed the
+    threshold so the rasterization is empty (right column) for about
+    16 iterations. Finally it reaches a stable state, a tiny piece of
+    colon-shaped dust, easily mistaken for a printing error. All
+    letters (from the test font Helvetica) converge to this shape,
+    except mysteriously the letter \letterform{v}. Is \letterform{v} a
+    different alien species, masquerading as a normal letter for
+    millions of years? } \label{fig:lowestcaseo}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+  \includegraphics[width=0.7 \linewidth]{uppestcase3d}
+\caption{ 3D manifold showing a section of the repeating loop as the
+  \makeuppercase\ model iteratively uppercases a letter. (Shown here
+  is the input \letterform{q} from iteration 245--377, but they all
+  converge to this same periodic shape.) Slices through this shape give a
+  letterform's outline (or usually, a linear interpolation between two
+  of them). The bottom of the shape is its ``beginning'' but it
+  appears to repeat like this forever.
+} \label{fig:uppestcase3d}
+\end{figure}
+
 Wow, this project is pretty involved, huh? Let's just add another
@@
 However, if I render the SDF as a 1-bit bitmap at $2\times$ scale, I
-get a perfect cycle of 132 frames, which appears to be two passes
+get a perfect cycle of 132 frames. This appears to be two passes
 through the main characteristic loop, but slightly different each
@@
 
-.... XXX more here ...
-
-\begin{figure}[t]
-\centering
-  \includegraphics[width=0.80 \linewidth]{lowestcaseo}
-  \caption{ The first 27 iterations of the \makelowercase\ model on
-    the letter \letterform{o}. The lowercase model generally makes
-    letters get smaller and smaller until they disappear. There is
-    still energy in the SDF (left column) but no pixels exceed the
-    threshold so the rasterization is empty (right column) for about
-    16 iterations. Finally it reaches a stable state, a tiny piece of
-    colon-shaped dust, easily mistaken for a printing error. All
-    letters (from the test font Helvetica) converge to this shape,
-    except mysteriously the letter \letterform{v}. Is \letterform{v} a
-    different alien species, masquerading as a normal letter for
-    millions of years? } \label{fig:lowestcaseo}
-\end{figure}
-
-% lowercase uppercase letters etc.
-% iterate this a lot of times?
-% "to infinity, but let's stop there"
-
-% run it to infinity to see the uppestcase, etc.
-% if these vanish, then we have a fixed point.
-% if not, or perhaps by doing some auto-gamma, then
-% it's like a fractal zoomer, and we can set the
-% record on the "most uppercase letter" by running
-% a jazillion times.
-
-% lowestcase is just dust, not surprising
-% uppestcase is actually a fixed point (not FixedSys)
-% ... so it's like a 3D manifold
-
-\begin{figure}[t]
-\centering
-  \includegraphics[width=0.7 \linewidth]{uppestcase3d}
-\caption{ 3D manifold showing a section of the repeating loop as the
-  \makeuppercase\ model iteratively uppercases a letter. (Shown here
-  is the input \letterform{q} from iteration 245--377, but they all
-  converge to this same periodic shape.) Slices through this shape give a
-  letterform's outline (or usually, a linear interpolation between two
-  of them). The bottom of the shape is its ``beginning'' but it
-  appears to repeat like this forever.
-} \label{fig:uppestcase3d}
-\end{figure}
-
-% bristol stool chart
+None of these alone could be considered the uppestcase letter, but
+perhaps together they are? A natural way to include them all in one
+shape is to stack them in 3D, as if each iteration is an MRI slice of
+the brachial plexus.
+
+To generate a 3D mesh, I stack the 132 SDFs in the z direction, and
+this naturally yields a three-dimensional signed distance field
+(trilinear interpolation). The classic ``marching cubes''
+algorithm~\cite{lorensen1987marching} for mesh generation works
+natively on such a field, and come to think of it, I probably could
+have used that in two dimensions to trace the SDFs to vectors. Oh,
+well. Fortunately there is enough spatial similarity between the
+slices that it makes a reasonable 3D shape even without interpolation,
+but sampled at a decent resolution and then cleaned up, it looks quite
+nice; much better than the lowercase colon. Speaking of colons, it
+rates approximately a 2.5 on the Bristol stool
+scale~\cite{lewis1997bristol}. A 2D projection of the manifold is in
+Figure~\ref{fig:uppestcase3d}. I 3D-printed it and am currently
+working on a way of embedding it as an unusually tall key on my
+keyboard that I can press whenever I wanted to express ULTIMATE ANGER.
 
 \section{Conclusion}
-... XXX ...
-This fixed point
-suggests that the lowestcase letter is already on your keyboard,
-but it is not a point, it is just \letterform{:}.
-
-\paragraph{Futura Work}
+We performed an exhaustive case analysis, exploring cases both more
+upper- and lower- than ever been seen before. Sideways case was not
+considered, as that is nonsense. We had modest success generating
+upperercase and lowerercase letterforms through two neural network
+models simultaneously trained to be each other's inverse, although
+frankly it was much faster and more aesthetic to just do it by hand.
+After really following through on the downloadables and taking some
+needless excursions for effect, we saw that these models have
+limits (at least informally). The lowestcase letter is already
+on your keyboard; it is the ASCII eyeballs character \letterform{:}.
+The uppestcase letter is not so easily typed, and perhaps that
+is for the best.
+
+\paragraph{Futura Work.}
 I think I pretty much beat this one to death, honestly.
 
-\paragraph{Acknowledgements}
+\paragraph{Acknowledgements.}
 For this project I used \verb+stb_truetype.h+ from the excellent stb

### 2021-03-31 lowercase/resequence.cc
Various fixes / tools for video

@@ -1,0 +1,93 @@
+
+// For numbered files in a directory, copy them to a new directory
+// but with consecutive frame numbers (e.g. for VirtualDub).
+
+#include <vector>
+#include <string>
+#include <algorithm>
+#include <cstdint>
+#include <memory>
+
+#include "timer.h"
+
+#include "threadutil.h"
+#include "base/stringprintf.h"
+#include "util.h"
+#include "base/logging.h"
+
+using namespace std;
+
+using uint8 = uint8_t;
+using uint32 = uint32_t;
+using int64 = int64_t;
+
+static string Frontslash(const string &s) {
+  string ret;
+  for (const char c : s)
+    ret += (c == '\\' ? '/' : c);
+
+  if (ret.find("d:/") == 0) {
+    ret[0] = '/';
+    ret[1] = 'd';
+  } else if (ret.find("c:/") == 0) {
+    ret[0] = '/';
+    ret[1] = 'c';
+  }
+
+  return ret;
+}
+
+static string Backslash(const string &s) {
+  string ret;
+  for (const char c : s)
+    ret += (c == '/' ? '\\' : c);
+  return ret;
+}
+
+static void AddAllFiles(const string &dir, vector<string> *all_files) {
+  for (const string &f : Util::ListFiles(dir)) {
+    const string filename = Util::dirplus(dir, f);
+    // printf("%s + %s = %s\n", dir.c_str(), f.c_str(), filename.c_str());
+    if (!Util::isdir(filename)) {
+      if (!filename.empty() &&
+          // Should probably delete emacs backups..?
+          filename[filename.size() - 1] != '#' &&
+          filename[filename.size() - 1] != '~') {
+        all_files->push_back(Backslash(filename));
+      }
+    }
+  }
+}
+
+static void Resequence(const string &in_dir, const string &out_dir) {
+  CHECK(out_dir.find("\\") == string::npos);
+  CHECK(out_dir.find("/") == string::npos);  
+  vector<string> files;
+  AddAllFiles(in_dir, &files);
+  std::sort(files.begin(), files.end(),
+            [](const string &a, const string &b) {
+              return Util::natural_compare(a, b) < 0;
+            });
+  printf("%lld files\n", (int64)files.size());
+  vector<pair<string, string>> remapped;
+  remapped.reserve(files.size());
+  for (int i = 0; i < files.size(); i++) {
+    remapped.emplace_back(files[i],
[…]
